### 2019-12-12 0_TittleAndAbastract.tex
First full draft

\title{Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations}

\author[1]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
\author[1]{Greg Chance}
\author[2]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
\author[2]{Tony Pipe}
\author[1]{Kerstin Eder}

\affil[1]{Trustworthy Systems Lab (TSL), Department of Computer Science, University of Bristol}
\affil[2]{Bristol Robotics Lab (BRL), University of West of England}

\maketitle

\begin{abstract}
\noindent The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation. It is important for these engines to be deterministic in order for one to carry out these different processes. However, such engines are inherently non-deterministic. We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic. We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.          
\end{abstract}

\begin{IEEEkeywords}
Autonomous Driving, Determinism, Physics Engines, Connected Autonomous Vehicles (CAV), Verification and Validation(V\&V)
\end{IEEEkeywords}
\IEEEpeerreviewmaketitle

### 2019-12-12 1_Intro.tex
First full draft

\section{Introduction}
\noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (CAVs) on roads is a delicate if not utterly fascinating subject.
Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}. 
Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care. 
It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
\noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
\noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system. 
Similarly, the use of simulation for traffic modelling is well established. 
Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate CAVs\cite{FrameworkAndChallenges}.\\\\
While traffic-level simulators can model road layouts, they do not require as much realism or detail as CAVs require. On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for CAV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure. 
Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
Consequently, one of the major routes CAVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate CAVs \cite{FrameworkAndChallenges}.  
Robopilot, Capri, Carla, Apollo, Airsim, Udacity etc. are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}. \\\\
Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of CAV software.\\\\
\TODO{Remove this paragraph?} Determinism of gaming (physics) engines has not attracted much attention by engines' developers nor researchers. This previously was not important since in gaming, determinism is not critical. 
Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
This paper aims to investigate how deterministic physics engines are for usage in V\&V for CAV simulations; and will cover the list of following points:
\begin{itemize}[leftmargin=*]
    \item Necessary background on gaming engines used in CAV simulations.
    \item Review of sources of non-determinism in such engines.
    \item Provide a method of how the level of non-determinism can be investigated and defined.
    \item A case study showing the implementation of this methodology in an engine widely used in CAVs simulations (Unreal Engine).
\end{itemize}

### 2019-12-12 4_CaseStudy.tex
First full draft

\section{Case Study} \label{CaseStudy}
\noindent Given that our interest is on CAV simulations, this study presents a number of experiments run on typical agents used in CAV simulations and their interactions with each other. These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals). \\\\
Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests. We will point out at each stage where we sit in Fig \ref{method_diagram} as we go through this case study.

\subsection{Problem And Hypothesis}
\noindent We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the level of non-determinism we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.\\\\
Our hypothesis is that since UE4 is a gaming engine it will not be deterministic, however, the level of non-determinism would be very low to acceptable limits that allows verification engineers to use it confidently.
This behaviour is expected not to hold as the level of computing stress increases.

\subsection{Tests Description And Technicalities}\label{TestsDescriptionAndTechnicalities}
\noindent The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints. 
Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.\\\\ 
\noindent Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments. 
The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents. 
These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.

\begin{figure}[h] 
\label{TestsDiagrams}
  \centering
  \subfloat[]{\includegraphics[width=0.25\textwidth]{Other/Figures/TestCasesDiagram_a.pdf}\label{Test_a}}
  \subfloat[]{\includegraphics[width=0.25\textwidth]{Other/Figures/TestCasesDiagram_b.pdf}\label{Test_b}}
  \caption{Shows the setup of the different tests in Table \ref{TableOfExperiments} (a)Test IDs 1 to 4 (b)Test IDs 5 to 8.}
\end{figure}

\noindent Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them. The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other. \\\\ 
Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.\\\\
Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Fig \ref{Test_b}). 
Pedestrians have the functionality of dynamic path planning i.e. they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other. 
Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.\\\\ 
In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the level of non-determinism of the simulator.\\\\
Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Fig \ref{Test_a}.\\\\
Proceeding to step 2; by doing various inspections around the different internal settings of this simulator we did not notice any improvement or worsening in the results. 
On the other hand the external settings (step 3) did show significant variations on the results as will be discussed in the next section. \\\\
Different levels of stress were applied to the computer; using the ------ to stress the CPU and ------ for the GPU.\\\\ 
In terms of running the experiments (step 4), \textbf{i)} the specs of the computer used are -----; \textbf{ii)} ------ was used to monitor the levels of GPU and CPU stress before and while the experiment was executed; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.

\subsection{Results And Discussion}
\noindent We start by analysing (step 5) the summary from all of the tests plotted in Fig \ref{ExperimentsStressSummary}. 
Note the dotted red line in the plot shows the millimeter (mm) level. 
Below this level we assume that the level of non-determinism is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic. 
This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests are assumed to be sufficiently deterministic, but just for the sake of argument we chose it to be the millimeter level.\\\\
First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows to be deterministic for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e. with in the centimeter level). This clearly shows how the load on a computer (i.e. stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles. \\\\
Next, given the pedestrians were showing a very robust and low level of non-determinism behaviour, we ran tests for pedestrians colliding (i.e. with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummary.pdf}
    \caption{Summary of mean deviation for the different tests runs plotted against the different CPU+GPU stresses.}

    \label{ExperimentsStressSummary}
\end{figure}

\begin{figure*}[h] 
  \centering
  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsCollisionCG25.pdf}\label{CarsCollisionCG25}}
  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsCollisionPrePost.pdf}\label{CarsCollisionPrePost}}
  \caption{Test ID 2 (a)Mean deviation vs simulation time for 25\% stress. (b)Pre and post collision mean deviation at different CPU+GPU stresses.}
\end{figure*}

\begin{figure*}[h] 
  \centering
  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsPeopleCollsionsCG25.pdf}\label{CarsPeopleCollsionsCG25}}
  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsPeopleCollisionPrePost.pdf}\label{CarsPeopleCollisionPrePost}}
  \caption{Test ID 4 (a)Mean deviation vs simulation time for 25\% stress. (b)Pre and post collision mean deviation at different CPU+GPU stresses.}
\end{figure*}

\noindent Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively. 
We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects. \\\\
\noindent Thus, we resorted to plotting the mean deviation of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Fig. \ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}). 
Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Fig \ref{ExperimentsStressSummary} to always surpass our red line.\\\\
\noindent Plotting Figs \ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge. 
However, in V\&V of CAV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated. Thus, the post collision effect should not be of much concern in the context of CAVs.\\\\
\noindent It is important to note, however, that in test 4 a delayed effect of increase in the level of non determinism was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents. The importance of this outcome is that it shows even if the ego vehicle is not involved in a collision, but other agents are, this will still have effects on the determinism of the ego vehicle. Thus failing to stay below our red line and the test should be terminated.

% \noindent\underline{\textit{No collisions:}}
% First, the tests with no collsions are considered, these represnt tests' ID 1, 3 and 5. ...
% --General deterministic behaviour of all of the agnets whent there are no collisions (tests 1,3 and 5)\\\\
% \noindent\underline{\textit{Pedestrians paths intersection:}}
% Pedestrians do not just blindly follow their trajectories, as already iterated in section \ref{TestsDescriptionAndTechnicalities}; thus it it 
% --Pathintersections bewteen pedestrains (tests 6,7 and 8)\\\\
% \noindent\underline{\textit{Collision between vehicles:}}
% Test 2 where the vehilcles collide together...
% --Collsion between vehicles alone (test 2)\\\\

% \noindent\underline{\textit{Collision between vehicles and pedestrians:}}
% --collsion between vehicles and pedestrains(test 4)
% -- coparison between variance before and after collsion

% - after a collsion the level of non determinsim increases significantly 
% - However In V\&V of CAV simulations, if a collsion is detected then the test should end

\TODO{We might need to change the y axes nameing betweent he summary plot (fig 4) and the fig 5 and 6, to show that one is mean deviation of all runs and thourgh out simulation time and the other is only between the different runs.}

\noindent In summary, from the tests carried out we can deduce that for a computer of the following specifications,-----..--, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.

### 2019-12-12 paper/2_Background.tex
organised some files

\section{Background}
\noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in CAV simulations. 
Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
\subsection{Overview of a Game Loop}\label{GameLoopSection}
\noindent The game loop is responsible for the interaction between the math, physics and rendering engines\cite{GameEngineArchBook}.
Figure \ref{GameEngineLoopDiagram} depicts a very basic diagram showing the flow in a game loop. 
A traditional game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).
\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
\caption{Game engine loop block diagram}
\label{GameEngineLoopDiagram}
\end{figure}
The first part of the game loop is to process the user input. Then comes update, which advances the AI and physics engine. Finally the rendering engine draws the game\cite{GameProgPatternsBook}.
It is crucial for this paper to understand how the rendering and physics engines operate together.
Briefly, the game loop operates as follows:\cite{GameProgPatternsBook} (see Figure \ref{GameEngineLoopDiagram})
\begin{itemize}[leftmargin=*]
    \item At the beginning of each tick (frame), the lag between the game clock and the real world is updated based on how much real time passed. This measures how far the game's clock is behind compared to the real world.
    \item Then the user inputs are processed.
    \item There is then an inner loop to update the physics engine, one fixed step at a time until it catches up with the real world. 
    The physics engine uses a fixed time step, because it makes everything simpler and more stable for physics and AI. 
    The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, as will be explained in section \ref{nondeterminisimSources}. 
    Thus, ideally the time step should be as short as possible, so that simulations run with high fidelity on fast machines. 
    However, if the fixed time step is too short i.e. less than the time it takes to process an ``Update" inner loop on some (slow) machines, then the simulation will simply never catch up on these slow machines.
    \item Rendering occurs once the physics engine catches up. The process then starts again.
\end{itemize}

\noindent This process allows a game to simulate across a range of hardware, but the rendering will become of jerky quality on slower machines. \\\\
These engines update at fixed intervals and render whenever they can, which is not steady. 
This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates. 
In this case, the engine will use extrapolation techniques to give a rough estimate of where it thinks the object should be. 
This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
\subsection{Sources of non-determinism}\label{nondeterminisimSources}
\noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic. 
The main reasons that could cause non-determinism are discussed below. \\\\
\noindent\underline{\textit{Floating Points:}}
A generic issue with computers are floating points precision. 
Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}. 
Translating to the usage of CAV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.\\\\ 
\noindent For instance, consider a simulation where the world is infinite and progressively  generated in the physics engine. 
After a few hundred meters, precision issues will start to occur, and will get progressively worse the further from the origin the simulation gets. 
This is due to the values of the coordinates being used in calculations are getting bigger and hence will not cope with small values that are critical for V\&V testing purposes. 
A solution to this problem is perhaps every time the simulations moves 100 meters away from the origin the entire world will move by 100m in the opposite direction. Thus, avoiding getting into floating point issues.\cite{FloatingPointArithmeticArticle}\\\\
\noindent There are many scenarios that could be addressed, and solutions to them will entirely depend on the characteristics of that specific scenario. There is no generalised solution for the floating problem, regardless of how good computers get there will always be a limitation. However, floating points as they stand could just be sufficient for V\&V testing purposes. Whenever they are not, smart solutions can be found to accommodate for these limitations.\\\\
\begin{figure*}[b]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagramCropped.pdf}
    \caption{Shows a flow diagram of the methodology proposed to define the level of non-determinism of a physics engine.}

    \label{method_diagram}
\end{figure*}
\noindent\underline{\textit{Navigation meshes and AI:}}
Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic\cite{CARLABenchmark}. 
The validity of this depends on what are the built in AI algorithms being used by the engine. 
Most of them tend to use the A* algorithm\cite{AStarBook}, which is an algorithm with deterministic behaviour if the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}. 
Therefore the determinism of the AI depends on the determinism of the engine and how its navigation meshes are created or modelled. It is interesting to note that changes can occur to meshes every time the simulation is loaded. This is mainly resorted to the different scheduling of threads. \\\\
\noindent\underline{\textit{Physics and Rendering clocks:}}
The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic. 
A follow up from section \ref{GameLoopSection}, these engines use a variable frame rate, which is good for hardware scalability but creates a challenge for the physics engine which works best with small fixed time steps. 
The problem of having a variable frame rate could also be a reason for the non-deterministic behaviour of these engines. This issue could perhaps be of negligible significance if the time steps are small enough. \\

\noindent\underline{\textit{Scheduling of threads:}} Scheduling of threads is believed to be the main reason for the non deterministic behaviour of these engines. Given the same hardware the output should always be the same, unless scheduling of the threads change, this would then cause non-deterministic behaviours. To solve this one would have to control all of the threads, which is a very involved and to achieve that one would need to replace or control the whole run-time system to allocate the same threads in the same order.

### 2019-12-12 paper/3_Methodology.tex
organised some files

\section{Methodology} \label{method}
\noindent A method for determining the level of non-determinism of an engine is presented in this section and the working flow of the method is shown in Figure \ref{method_diagram}. Each of the steps in the method is elaborated in the following subsections.

\subsection{Design experiment}
\noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic? ii) If not, how does its level of non-determinism vary?\\\\
In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination. 
Thus, given that an engine is non-deterministic one needs to set up an experiment that would stress the engine in order to determine how its level of non-determinism vary. This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision. Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
It is crucial to also make sure that any randomisation factors are eliminated in the experiment.

\subsection{Internal settings}
\noindent There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the level of non-determinism. \\\\
\noindent Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of non-determinism since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive. However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests. \\\\
\noindent If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of non-determinism as a result of allowing the AI to navigate in a more well defined space.  \\\\
\noindent Disabling rendering or running in headless mode is a way of attempting to decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests. 
The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.  \\\\
\noindent Running in editor or deployed mode can sometimes cause massive differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.

\begin{table*}[b]
\centering
\begin{tabular}{clclcc}
\toprule
Test ID & Test description (See Fig \ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
\bottomrule
\end{tabular}
\caption{Set of experiments}
\label{TableOfExperiments}
\end{table*}

\subsection{External settings}
\noindent Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.\\\\ 
\noindent Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing. Note, however, that during running these tests computers should be left alone not be used for any other purposes apart from running the experiments.

\subsection{Running experiment}
\noindent When it comes to running the experiment itself, things that one should consider  are the problem space exploration; the number of runs that should be executed to get reliable results; the frequency of logging data of actors; and possibly having a program to monitor the hardware utilisation.
% Problem exploration, there is too many variables that one can play with. do statistical analysis some.... some of them get problematic...

\subsection{Output and analysis}
\noindent Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs. Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.\\\\
This whole process is repeated for various internal and external settings. 
Then various plots can be created to identify and engine's level of non-determinsim. 
Thereon, it would be up to the verification engineers to determine where they want to define the line below which the level of non-determinism would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.

### 2019-12-12 paper/6_Conclusions.tex
organised some files

\section{Conclusions}
\noindent We have presented a method by which one can determine the level of non-determinism of a physics engine used in V\&V application of CAV simulations. The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance and finally focuses on running the experiment and analysing the output. \\\\
A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA. It was found that for the computer used (with specs of----) the level of non determinism is sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected. %since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
% \TODO{Add the conclusion mentioned by Thales}

### 2019-12-12 paper_arxiv/1_Intro.tex
Added archives version but needs modifications

--- 2019-12-12 1_Intro.tex
@@
 \section{Introduction}
-\noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (CAVs) on roads is a delicate if not utterly fascinating subject.
+\noindent The evolution and inevitable implementation of connected and autonomous vehicles (CAVs) on roads is a delicate if not utterly fascinating subject.
 Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}. 

### 2019-12-12 paper_arxiv/4_CaseStudy.tex
Added archives version but needs modifications

--- 2019-12-12 4_CaseStudy.tex
@@
 \begin{figure}[h]
-\label{TestsDiagrams}
     \centering
-  \subfloat[]{\includegraphics[width=0.25\textwidth]{Other/Figures/TestCasesDiagram_a.pdf}\label{Test_a}}
-  \subfloat[]{\includegraphics[width=0.25\textwidth]{Other/Figures/TestCasesDiagram_b.pdf}\label{Test_b}}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagram_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagram_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
     \caption{Shows the setup of the different tests in Table \ref{TableOfExperiments} (a)Test IDs 1 to 4 (b)Test IDs 5 to 8.}
@@
     \centering
-  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsCollisionCG25.pdf}\label{CarsCollisionCG25}}
-  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsCollisionPrePost.pdf}\label{CarsCollisionPrePost}}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25.pdf}
+        \caption{}
+        \label{CarsCollisionCG25}
+    \end{subfigure}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePost.pdf}
+        \caption{}
+        \label{CarsCollisionPrePost}
+    \end{subfigure}
     \caption{Test ID 2 (a)Mean deviation vs simulation time for 25\% stress. (b)Pre and post collision mean deviation at different CPU+GPU stresses.}
@@
     \centering
-  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsPeopleCollsionsCG25.pdf}\label{CarsPeopleCollsionsCG25}}
-  \subfloat[]{\includegraphics[width=0.48\textwidth]{Other/Figures/CarsPeopleCollisionPrePost.pdf}\label{CarsPeopleCollisionPrePost}}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25.pdf}
+        \caption{}
+        \label{CarsPeopleCollsionsCG25}
+    \end{subfigure}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePost.pdf}
+        \caption{}
+        \label{CarsPeopleCollisionPrePost}
+    \end{subfigure}
     \caption{Test ID 4 (a)Mean deviation vs simulation time for 25\% stress. (b)Pre and post collision mean deviation at different CPU+GPU stresses.}

### 2019-12-12 paper_arxiv/main_arxiv.tex
Added archives version but needs modifications

\documentclass[runningheads,a4paper]{llncs}
\usepackage{etex}
\reserveinserts{28}
\pdfoutput=1

\setcounter{tocdepth}{3}

\usepackage[linesnumbered,lined,boxed,commentsnumbered,ruled]{algorithm2e}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}
\usepackage{ textcomp }
\usepackage{ comment }
\usepackage[color=yellow!60,textsize=footnotesize,obeyDraft,draft]{todonotes}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{capt-of}
\usepackage{makecell}
\usepackage{textcomp }
\usepackage{multirow}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[font=footnotesize,labelfont=bf]{subcaption}
\usepackage[hidelinks]{hyperref}
\usepackage[capitalize]{cleveref}
\captionsetup{compatibility=false}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[para,online,flushleft]{threeparttable}
\usepackage{multirow}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{enumitem}
\usetikzlibrary{arrows}
% \usepackage{subfig}
\usepackage{subcaption}
\usepackage{breakcites}
\setcounter{secnumdepth}{3}

% Make tables a little bit easier to read
\renewcommand{\arraystretch}{1.15}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NOTES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcounter{mnotei}
\setcounter{mnotei}{0}
\newcommand{\mnote}[1]{%
  {\scriptsize\textsf{$^{\textcolor{red}{[n.\themnotei]}}$}}%
  \marginpar{\scriptsize\textsf{\textcolor{red}{[n.\themnotei]: #1}}}%
  % \marginpar{\scriptsize\textsf{n.\themnotei: #1}}%
  \stepcounter{mnotei}
}
%\renewcommand{\mnote}[1]{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Provide paragraph-style table columns
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

% Give more marginpar
\setlength{\marginparwidth}{25mm}

% Give more marginpar
\setlength{\marginparwidth}{25mm}

% Prompt people to generate graphs
\newcommand{\includegraphicsmaybe}[2]{
    \IfFileExists{#2}{\includegraphics[#1]{#2}}{
    \detokenize{File #2 is missing, maybe you need to run plots.py?}
}}

\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}

\begin{document}
\bibliographystyle{ieeetr}
\mainmatter

\title{Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations}

\titlerunning{Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations}

\author{Abanoub Ghobrial\inst{1,3}, Greg Chance\inst{1,3}, Severin Lemaignan\inst{2,3}, \\Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}

\authorrunning{A. Ghobrial et al.}
\institute{University of Bristol, Bristol, UK \and University of the West of England, Bristol, UK \and Bristol Robotics Laboratory, University of the West of England, Bristol, UK}
\tocauthor{Authors' Instructions}
\maketitle

\makeatletter
\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
                       {4\p@ \@plus 2\p@ \@minus 2\p@}%
                       {\normalfont\normalsize\bfseries\boldmath
                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
\makeatother

\begin{abstract}
\noindent The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation. It is important for these engines to be deterministic in order for one to carry out these different processes. However, such engines are inherently non-deterministic. We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic. We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.  
\end{abstract}

\input{1_Intro.tex}

\input{2_Background.tex}

\input{3_Methodology.tex}

\input{4_CaseStudy.tex}

% \input{5_ResultsAndDiscussion.tex}

\input{6_Conclusions.tex}

\section{Acknowledgement}
This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
projects are part-funded by the Centre for Connected and Autonomous
Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
103703 (CAPRI) and 103288 (ROBOPILOT), respectively.

\bibliography{IROS.bib}

\end{document}

\grid

### 2019-12-12 paper/0_TittleAndAbastract.tex
arxivs template setup

--- 2019-12-12 0_TittleAndAbastract.tex
@@
 
-\author[1]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
-\author[1]{Greg Chance}
-\author[2]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-\author[2]{Tony Pipe}
-\author[1]{Kerstin Eder}
+\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
+\author[1,3]{Greg Chance}
+\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
+\author[2,3]{Tony Pipe}
+\author[1,3]{Kerstin Eder}
 
-\affil[1]{Trustworthy Systems Lab (TSL), Department of Computer Science, University of Bristol}
-\affil[2]{Bristol Robotics Lab (BRL), University of West of England}
+\affil[1]{University of Bristol, Bristol, UK}
+\affil[2]{University of West of England, Bristol, UK}
+\affil[3]{Bristol Robotics Lab, Bristol, UK}
 

### 2019-12-12 paper_arxiv/2_Background.tex
Added archives version but needs modifications
arxivs template setup

--- 2019-12-12 paper/2_Background.tex
@@
 Whenever they are not, smart solutions can be found to accommodate for these limitations.\\\\
-\begin{figure*}[b]
-    \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagramCropped.pdf}
-    \caption{Shows a flow diagram of the methodology proposed to define the level of non-determinism of a physics engine.}
-
-    \label{method_diagram}
-\end{figure*}
 \noindent\underline{\textit{Navigation meshes and AI:}}

### 2019-12-12 paper_arxiv/3_Methodology.tex
arxivs template setup

--- 2019-12-12 paper/3_Methodology.tex
@@
 Each of the steps in the method is elaborated in the following subsections.
+\begin{figure*}[h!]
+    \centering
+    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagramCropped.pdf}
+    \caption{Shows a flow diagram of the methodology proposed to define the level of non-determinism of a physics engine.}
 
+    \label{method_diagram}
+\end{figure*}
 \subsection{Design experiment}
@@
 
-\begin{table*}[b]
-\centering
-\begin{tabular}{clclcc}
-\toprule
-Test ID & Test description (See Fig \ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
-1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
-2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
-3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
-4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
-5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
-6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
-7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
-8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
-\bottomrule
-\end{tabular}
-\caption{Set of experiments}
-\label{TableOfExperiments}
-\end{table*}
-
 \subsection{External settings}

### 2019-12-12 paper_arxiv/4_CaseStudy.tex
arxivs template setup

@@
 \section{Case Study} \label{CaseStudy}
+\begin{table*}[h!]
+\centering
+\resizebox{1\textwidth}{!}{
+\begin{tabular}{clclcc}
+\toprule
+Test ID & Test description (See Fig \ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
+1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
+2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
+3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
+4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
+5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
+6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
+7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
+8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
+\bottomrule
+\end{tabular}
+}
+\caption{Set of experiments}
+\label{TableOfExperiments}
+\end{table*}
+
 \noindent Given that our interest is on CAV simulations, this study presents a number of experiments run on typical agents used in CAV simulations and their interactions with each other.
@@
 This behaviour is expected not to hold as the level of computing stress increases.
-
-\subsection{Tests Description And Technicalities}\label{TestsDescriptionAndTechnicalities}
-\noindent The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints. 
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.\\\\ 
-\noindent Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments. 
-The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents. 
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
 
@@
 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummary.pdf}
+    \caption{Summary of mean deviation for the different tests runs plotted against the different CPU+GPU stresses.}
+
+    \label{ExperimentsStressSummary}
+\end{figure}
+\subsection{Tests Description And Technicalities}\label{TestsDescriptionAndTechnicalities}
+\noindent The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints. 
+Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.\\\\ 
+\noindent Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments. 
+The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents. 
+These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.\\\\
 \noindent Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
@@
 \subsection{Results And Discussion}
-\noindent We start by analysing (step 5) the summary from all of the tests plotted in Fig \ref{ExperimentsStressSummary}. 
-Note the dotted red line in the plot shows the millimeter (mm) level. 
-Below this level we assume that the level of non-determinism is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic. 
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests are assumed to be sufficiently deterministic, but just for the sake of argument we chose it to be the millimeter level.\\\\
-First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows to be deterministic for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
-with in the centimeter level).
-This clearly shows how the load on a computer (i.e.
-stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
-This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
-\\\\
-Next, given the pedestrians were showing a very robust and low level of non-determinism behaviour, we ran tests for pedestrians colliding (i.e.
-with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummary.pdf}
-    \caption{Summary of mean deviation for the different tests runs plotted against the different CPU+GPU stresses.}
-
-    \label{ExperimentsStressSummary}
-\end{figure}
-
 \begin{figure*}[h]
@@
 \end{figure*}
+\noindent We start by analysing (step 5) the summary from all of the tests plotted in Fig \ref{ExperimentsStressSummary}. 
+Note the dotted red line in the plot shows the millimeter (mm) level. 
+Below this level we assume that the level of non-determinism is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic. 
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests are assumed to be sufficiently deterministic, but just for the sake of argument we chose it to be the millimeter level.\\\\
+First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
+Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows to be deterministic for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
+with in the centimeter level).
+This clearly shows how the load on a computer (i.e.
+stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
+This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
+\\\\
+Next, given the pedestrians were showing a very robust and low level of non-determinism behaviour, we ran tests for pedestrians colliding (i.e.
+with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
 

### 2019-12-12 paper_arxiv/main_arxiv.tex
arxivs template setup

@@
-\documentclass[runningheads,a4paper]{llncs}
+\documentclass[runningheads,twocolumn,a4paper,10pt]{llncs}
 \usepackage{etex}
@@
 
+\usepackage[a4paper, total={7in, 10in}]{geometry}
 \usepackage[linesnumbered,lined,boxed,commentsnumbered,ruled]{algorithm2e}
@@
 \usepackage{breakcites}
+% \usepackage{multicol}
 \setcounter{secnumdepth}{3}
@@
 
-\titlerunning{Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations}
+\titlerunning{Determinism of Physics Engines for CAV}
 
-\author{Abanoub Ghobrial\inst{1,3}, Greg Chance\inst{1,3}, Severin Lemaignan\inst{2,3}, \\Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}
+\author{Abanoub Ghobrial\inst{1,3}, Greg Chance\inst{1,3}, Severin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}
 
 \authorrunning{A. Ghobrial et al.}
-\institute{University of Bristol, Bristol, UK \and University of the West of England, Bristol, UK \and Bristol Robotics Laboratory, University of the West of England, Bristol, UK}
+\institute{University of Bristol, Bristol, UK \and University of the West of England, Bristol, UK \and Bristol Robotics Laboratory, Bristol, UK}
 \tocauthor{Authors' Instructions}
@@
 
-\begin{abstract}
-\noindent The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
+% \begin{multicols}{2}
+
+% \begin{abstract}
+\textbf{\textit{Abstract}--The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
 It is important for these engines to be deterministic in order for one to carry out these different processes.
@@
 We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic.
-We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.
-\end{abstract}
+We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.}
+% \end{abstract}
 
@@
 \input{4_CaseStudy.tex}
-
-% \input{5_ResultsAndDiscussion.tex}
 
@@
 
+% \end{multicols}
 \end{document}

### 2019-12-14 paper/cav_determinisim_main.tex
Got rid of unnecessary files

\documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
\input{main.config.tex}
\pagenumbering{gobble} %suppresses page numbers
%-------------------------------------------------------------------------------

\begin{document}

\pagenumbering{arabic}

\input{0_TittleAndAbastract.tex}

\input{1_Intro.tex}

\input{2_Background.tex}

\input{3_Methodology.tex}

\input{4_CaseStudy.tex}

% \input{5_ResultsAndDiscussion.tex}

\input{6_Conclusions.tex}

\input{9_FundingAndReferences.tex}

\end{document}

% \begin{table*}
% \centering
% \caption{Feature importance given in rank order for prediction of elbow position ($\Phi$) for univariate and multivariate response. $\Sigma(NI)$ indicates sum of normalized importance for the top 5 features.}
% \label{elbow_posn}
% \begin{tabular}{|c||c|c|c||c|}
% \hline
% Rank & $\Phi(x)$ & $\Phi(y)$ & $\Phi(z)$ & $\Phi(xyz)$ \\
% \hline
% 1&left shoulder ($13P_x$) &left shoulder ($13P_y$)&left shoulder ($13P_z$) &right hip ($20P_x$)\\
% 2&right hip ($20P_x$)     &right hip ($20P_x$)    &head ($7P_z$)           &left shoulder ($13P_z$)\\
% 3&head ($7P_x$)           &right elbow ($10P_y$)  &right hand ($11P_z$)    &left shoulder ($13P_x$)\\
% 4&right hand ($11P_x$)    &right hand ($11P_x$)   &head ($7P_x$)		   &right hand ($11P_x$)\\
% 5&left shoulder ($13P_y$) &right elbow ($10P_x$)  &neck ($6P_z$)		   &head ($7P_x$)\\
% \hline 
% $\Sigma(NI)$ & 0.683      &0.684                  &0.723 					&0.570 \\
% \hline
% %RMSE-test ($m$) & 0.011& 0.017& 0.007& 0.024\\
% %RMSE-unseen ($m$) & 0.086& 0.060& 0.038& 0.063\\
% %\hline
% \end{tabular}
% \end{table*}

### 2019-12-17 paper/0_TittleAndAbastract.tex
GC change abstract

@@
 \begin{abstract}
-\noindent The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
+\noindent 
+The connected and autonomous vehicle (CAV) community are adopting the use of gaming engines to develop vehicle control systems and testing environments using simulation.
+These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V).
+This work highlights the requirements these engines need (to fulfil this purpose/for CAV development) and proposes a method to determine \textit{simulation variance} for specific configurations and resource utilisation.
+A case study is used to illustrate the proposed method and results from a small scenario indicate deterministic operational domains.
+% The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
 It is important for these engines to be deterministic in order for one to carry out these different processes.

### 2019-12-17 paper/0_TittleAndAbastract.tex
GC change intro bib

@@
-\title{Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations}
+\title{Investigating Determinism of Gaming Engines for Autonomous Vehicle Verification}
 
@@
 This work highlights the requirements these engines need (to fulfil this purpose/for CAV development) and proposes a method to determine \textit{simulation variance} for specific configurations and resource utilisation.
-A case study is used to illustrate the proposed method and results from a small scenario indicate deterministic operational domains.
+A case study is used to illustrate the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
 % The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.

### 2019-12-17 paper/3_Methodology.tex
Finished changes to Methodology,still not done with case study changes. Need to add devaition equation, change y-axes titles and add more details to step 2 in case study

@@
 \section{Methodology} \label{method}
-\noindent A method for determining the level of non-determinism of an engine is presented in this section and the working flow of the method is shown in Figure \ref{method_diagram}.
+\noindent A method for determining the level of simulation variance of an engine is presented in this section and the working flow of the method is shown in Figure \ref{method_diagram}.
 Each of the steps in the method is elaborated in the following subsections.
@@
 \subsection{Design experiment}
-\noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic? ii) If not, how does its level of non-determinism vary?\\\\
+\noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic? ii) If not, how does its simulation variance vary?\\\\
 In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination. 
-Thus, given that an engine is non-deterministic one needs to set up an experiment that would stress the engine in order to determine how its level of non-determinism vary.
+Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary. 
 This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
@@
 \subsection{Internal settings}
-\noindent There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the level of non-determinism. \\\\
-\noindent Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of non-determinism since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
+\noindent There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine. \\\\
+\noindent Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
 However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
 \\\\
-\noindent If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of non-determinism as a result of allowing the AI to navigate in a more well defined space.
+\noindent If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of simulation variance as a result of allowing the AI to navigate in a more well defined space.
 \\\\
-\noindent Disabling rendering or running in headless mode is a way of attempting to decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
+\noindent Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
 The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.  \\\\
@@
 \noindent Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
-Note, however, that during running these tests computers should be left alone not be used for any other purposes apart from running the experiments.
+Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.
 
@@
 Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.\\\\
-This whole process is repeated for various internal and external settings. 
-Then various plots can be created to identify and engine's level of non-determinsim. 
-Thereon, it would be up to the verification engineers to determine where they want to define the line below which the level of non-determinism would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
+This whole process is repeated for various internal and external settings (if needed). 
+Then various plots can be created to draw the engine's simulation variance. 
+Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.

### 2019-12-17 paper/1_Intro.tex

--- 2019-12-12 paper_arxiv/1_Intro.tex
@@
-\section{Introduction}
-\noindent The evolution and inevitable implementation of connected and autonomous vehicles (CAVs) on roads is a delicate if not utterly fascinating subject.
+\section{Introduction} \label{s:introduction}
+
+% What is verification
+\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}.
+Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it.
+These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988}.
+Simulation is an obvious domain choice to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving} where the rare event can be made to happen more often~\cite{Koopman2018} such as unpredictable actor behaviour~\cite{RobustnessAutonomy}.
+
+% Requirements of simulation based verification SBV
+For games engines to offer a suitable testing environment they must: operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors), be realistic enough \cite{koopman realistic enough} including suitable rendering for perception stack testing, be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015.}, include the provision for sensor analogues to a suitable accuracy (e.g.
+video, LIDAR, IR), allow hardware-in-the-loop development and provide a suitable real-time test-bed for cyber-security testing.
+%
+% Game engines for SBV
+Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements but many challenges still exist.
+%
+% Verification and Determinism 
+For games engines to be useful as a verification tool then they must be deterministic. 
+%
+% What is a deterministic system
+Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
+%
+% Why is determinism important for verification
+Determinism is important for verification.
+Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
+Furthermore, defects not found on the verification testing system may arise on deployed system with a different hardware architecture resulting in untested and potentially unsafe vehicle behaviour.
+
+% Challenges of SBV
+\noindent The research questions in this work are:
+\begin{itemize}[leftmargin=*]
+    \item To what extent are gaming engines deterministic so that they can provide a reliable and repeatable testing environment for CAV verification?
+    \item Can this be done whilst providing the realism required for the physical and perceptual requirements of CAV development?
+    \item Will a game engine allowing real-time testing for cyber security analysis? %\todo{What will this be, minimum 30FPS?}
+\end{itemize}
+
+% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (CAVs) on roads is a delicate if not utterly fascinating subject.
 Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.
+
 Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.

### 2019-12-18 paper/1_Intro.tex
finish intro and updated bib

@@
 These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988}.
+%
+\noindent There have been a number of fatalities with autonomous vehicles which could be attributed to a lack of verification and validation (V\&V), e.g.\cite{FatalityExample}.
+%
 Simulation is an obvious domain choice to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving} where the rare event can be made to happen more often~\cite{Koopman2018} such as unpredictable actor behaviour~\cite{RobustnessAutonomy}.
+%
 
@@
 For games engines to offer a suitable testing environment they must: operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors), be realistic enough \cite{koopman realistic enough} including suitable rendering for perception stack testing, be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015.}, include the provision for sensor analogues to a suitable accuracy (e.g.
-video, LIDAR, IR), allow hardware-in-the-loop development and provide a suitable real-time test-bed for cyber-security testing.
+video, LIDAR, IR), allow hardware-in-the-loop development and provide a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}.
+%, e.g. minimum rate of frames per second (fps).
 %
@@
 Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
-Furthermore, defects not found on the verification testing system may arise on deployed system with a different hardware architecture resulting in untested and potentially unsafe vehicle behaviour.
+Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe vehicle behaviour.
 
 % Challenges of SBV
-\noindent The research questions in this work are:
-\begin{itemize}[leftmargin=*]
-    \item To what extent are gaming engines deterministic so that they can provide a reliable and repeatable testing environment for CAV verification?
-    \item Can this be done whilst providing the realism required for the physical and perceptual requirements of CAV development?
-    \item Will a game engine allowing real-time testing for cyber security analysis? %\todo{What will this be, minimum 30FPS?}
-\end{itemize}
+Our main research question is to what extent are gaming engines deterministic so that they can provide a reliable and repeatable testing environment for CAV verification? 
+%
+The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.
+
+This paper is structured as follows.
+In Section~\ref{s:background} a background on gaming engines and sources of potential non-determinism are explored.
+Section~\ref{s:methodology} presents a method to design and execute an experiment to assess suitability of a system for CAV verification.
+A case study to illustrate the level of simulation variance for a number of scenarios is explored in Section~\ref{s:case-study}.
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.
 
@@
 
-Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.
-It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
-\noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
-As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
-\noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system. 
+% Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.
+
+% It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
+
+% \noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
+
+% As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
+
+% \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system.
 Similarly, the use of simulation for traffic modelling is well established.
 Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate CAVs\cite{FrameworkAndChallenges}.\\\\
-While traffic-level simulators can model road layouts, they do not require as much realism or detail as CAVs require.
+
+% move to background---->>>?
+% While traffic-level simulators can model road layouts, they do not require as much realism or detail as CAVs require.
 On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for CAV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
-Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
-Consequently, one of the major routes CAVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate CAVs \cite{FrameworkAndChallenges}.
-Robopilot, Capri, Carla, Apollo, Airsim, Udacity etc.
+% Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
+% Consequently, one of the major routes CAVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate CAVs \cite{FrameworkAndChallenges}.
+% Robopilot, Capri, Carla, Apollo, Airsim, Udacity etc.
 are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}.
 \\\\
-Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
-In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of CAV software.\\\\
-\TODO{Remove this paragraph?} Determinism of gaming (physics) engines has not attracted much attention by engines' developers nor researchers.
+% Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
+% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of CAV software.\\\\
+% \TODO{Remove this paragraph?} Determinism of gaming (physics) engines has not attracted much attention by engines' developers nor researchers.
 This previously was not important since in gaming, determinism is not critical.
-Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
-On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
-This paper aims to investigate how deterministic physics engines are for usage in V\&V for CAV simulations; and will cover the list of following points:
-\begin{itemize}[leftmargin=*]
-    \item Necessary background on gaming engines used in CAV simulations.
-    \item Review of sources of non-determinism in such engines.
-    \item Provide a method of how the level of non-determinism can be investigated and defined.
-    \item A case study showing the implementation of this methodology in an engine widely used in CAVs simulations (Unreal Engine).
-\end{itemize}
+% Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
+% On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
+
+% This paper aims to investigate how deterministic physics engines are for usage in V\&V for CAV simulations; and will cover the list of following points:
+% \begin{itemize}[leftmargin=*]
+%     \item Necessary background on gaming engines used in CAV simulations.
+%     \item Review of sources of non-determinism in such engines.
+%     \item Provide a method of how the level of non-determinism can be investigated and defined.
+%     \item A case study showing the implementation of this methodology in an engine widely used in CAVs simulations (Unreal Engine).
+% \end{itemize}

### 2019-12-18 paper/4_CaseStudy.tex

--- 2019-12-12 paper_arxiv/4_CaseStudy.tex
@@
 \subsection{Problem And Hypothesis}
-\noindent We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the level of non-determinism we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.\\\\
-Our hypothesis is that since UE4 is a gaming engine it will not be deterministic, however, the level of non-determinism would be very low to acceptable limits that allows verification engineers to use it confidently.
+\noindent We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the level of simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.\\\\
+Our hypothesis is that since UE4 is a gaming engine it is not deterministic, however, the level of simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
 This behaviour is expected not to hold as the level of computing stress increases.
@@
 Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.\\\\
-In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the level of non-determinism of the simulator.\\\\
+In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.\\\\
 Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Fig \ref{Test_a}.\\\\
@@
 On the other hand the external settings (step 3) did show significant variations on the results as will be discussed in the next section. \\\\
-Different levels of stress were applied to the computer; using the ------ to stress the CPU and ------ for the GPU.\\\\ 
-In terms of running the experiments (step 4), \textbf{i)} the specs of the computer used are -----; \textbf{ii)} ------ was used to monitor the levels of GPU and CPU stress before and while the experiment was executed; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.\\\\
+In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
 
@@
 Note the dotted red line in the plot shows the millimeter (mm) level. 
-Below this level we assume that the level of non-determinism is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic. 
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests are assumed to be sufficiently deterministic, but just for the sake of argument we chose it to be the millimeter level.\\\\
+Below this level we assume that the level of simulation variance is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. 
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic, but just for the sake of argument we chose it to be the millimeter level.\\\\
 First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows to be deterministic for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
+Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
 with in the centimeter level).
@@
 \\\\
-Next, given the pedestrians were showing a very robust and low level of non-determinism behaviour, we ran tests for pedestrians colliding (i.e.
+Next, given the pedestrians were showing a very robust and low level of simulation variance, we ran tests for pedestrians colliding (i.e.
 with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
@@
 Thus, the post collision effect should not be of much concern in the context of CAVs.\\\\
-\noindent It is important to note, however, that in test 4 a delayed effect of increase in the level of non determinism was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
-The importance of this outcome is that it shows even if the ego vehicle is not involved in a collision, but other agents are, this will still have effects on the determinism of the ego vehicle.
+\noindent It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
+The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
 Thus failing to stay below our red line and the test should be terminated.

### 2019-12-18 paper/2_Background.tex
GC background

@@
-\section{Background}
-\noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in CAV simulations. 
+\section{Background} \label{s:background}
+
+There are numerous game engines with their associated development environments which could be suitable for CAV development, e.g.
+Unreal, Unity, CryEngine.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.
+Carla, AirSim, Apollo, and some have been developed specifically, e.g.
+the could based Nvidia Drive Constellation simulator.
+\TODO{ref each of these}
+%
+Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution.
+However, when these engines are used for CAV development then deterministic behaviour is required.
+This section gives an overview of a gaming engine and what sources or settings in the engine may affect \textit{simulation variance}.
+%
+% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in CAV simulations.
 Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
-\subsection{Overview of a Game Loop}\label{GameLoopSection}
-\noindent The game loop is responsible for the interaction between the math, physics and rendering engines\cite{GameEngineArchBook}.
-Figure \ref{GameEngineLoopDiagram} depicts a very basic diagram showing the flow in a game loop. 
-A traditional game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).
+
+% ======= Overview of a Game Loop
+\subsection{Overview of a Game Loop} \label{GameLoopSection}
+The game loop is responsible for the interaction between the maths, physics and rendering engines~\cite{GameEngineArchBook}.
+Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop.
+A traditional \TODO{For AG: how is it traditional, what is the alternative?} game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).
+
 \begin{figure}[h]
@@
 \end{figure}
-The first part of the game loop is to process the user input. Then comes update, which advances the AI and physics engine. Finally the rendering engine draws the game\cite{GameProgPatternsBook}.
-It is crucial for this paper to understand how the rendering and physics engines operate together.
+The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment.
+\TODO{For AG: need to expand, e.g what are the inputs?} \TODO{For AG: does game logic need to be mentioned?}
+%
+The update interprets the intended actions of the AV along with all other dynamic actors in the scene and performs any necessary physics calculations and updates the actor states.
+\TODO{For AG: what does the clock symbolise, need to explain it?}
+%
+The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
+\TODO{For AG: you cite this book 4 times which might be too much as a general reference, is there something specific you want to draw the reader towards in the book?
+What specifically about the rendering part of this cycle does the book help to illustrate the point?}
+
+It is crucial for this paper to understand how the rendering and physics engines operate together. \TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
 Briefly, the game loop operates as follows:\cite{GameProgPatternsBook} (see Figure \ref{GameEngineLoopDiagram})
 \begin{itemize}[leftmargin=*]
-    \item At the beginning of each tick (frame), the lag between the game clock and the real world is updated based on how much real time passed.
+    \item At the beginning of each tick (scene), the lag between the game clock and the real world is updated based on how much real time passed.
 This measures how far the game's clock is behind compared to the real world.
     \item Then the user inputs are processed.
-    \item There is then an inner loop to update the physics engine, one fixed step at a time until it catches up with the real world. 
+    \item There is then an inner loop to update the physics engine (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at fixed step until the game clock is equal to the real world.
 The physics engine uses a fixed time step, because it makes everything simpler and more stable for physics and AI.
-    The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, as will be explained in section \ref{nondeterminisimSources}.
+The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, as will be explained in section \ref{s:nondeterminisimSources}.
 Thus, ideally the time step should be as short as possible, so that simulations run with high fidelity on fast machines.
@@
 
-\noindent This process allows a game to simulate across a range of hardware, but the rendering will become of jerky quality on slower machines. \\\\
+The time step, $dt$, in any physics calculation is important.
+To use a fixed physics time step the user's display refresh rate needs to be known in advance.
+This requires an update loop to take less than one frame of real world time~\cite{gaffer}.
+Given the different capabilities of game players hardware a variable delta time can be implemented taking the previous frame time as the next $dt$.
+However, variable $dt$ can lead to different physical results and in some cases unrealistic physics.
+Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware.
+Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}.
+If the engine tries to render between updates \textit{residual lag} can occur \cite{GameEngineArchBook}\cite{GameProgPatternsBook} and extrapolation between frames is performed to smooth transition between scenes.
+
+% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines.
 These engines update at fixed intervals and render whenever they can, which is not steady.
 This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates.
-In this case, the engine will use extrapolation techniques to give a rough estimate of where it thinks the object should be. 
+In this case, the engine will use extrapolation techniques to give an estimate of where the object should be.
 This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
-\subsection{Sources of non-determinism}\label{nondeterminisimSources}
+
+% ======= Sources of non-determinism
+\subsection{Sources of non-determinism} \label{s:nondeterminisimSources}
 \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic. 
 The main reasons that could cause non-determinism are discussed below. \\\\
+
 \noindent\underline{\textit{Floating Points:}}

### 2019-12-18 paper/1_Intro.tex

@@
 % Requirements of simulation based verification SBV
-For games engines to offer a suitable testing environment they must: operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors), be realistic enough \cite{koopman realistic enough} including suitable rendering for perception stack testing, be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015.}, include the provision for sensor analogues to a suitable accuracy (e.g.
+For games engines to offer a suitable testing environment they must: operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors), be realistic enough \cite{Koopman2018} including suitable rendering for perception stack testing, be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015}, include the provision for sensor analogues to a suitable accuracy (e.g.
 video, LIDAR, IR), allow hardware-in-the-loop development and provide a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}.
@@
 Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
+%
+Herein the term \textit{simulation variance} is used to refer to how simulation runs vary when repeated many times and specifically the variance of agent paths within a given scenario.
+We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
+Dynamic elements are the elements in a scene whose actions or behaviour may change over time, these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals.
+The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.
+A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
 % 
@@
 
+% ======================
+% This is all from "FrameworkAndChallenges" without using quotes 
 % \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system.
@@
 
-% move to background---->>>?
 % While traffic-level simulators can model road layouts, they do not require as much realism or detail as CAVs require.
@@
 % Consequently, one of the major routes CAVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate CAVs \cite{FrameworkAndChallenges}.
+% ======================
+
 % Robopilot, Capri, Carla, Apollo, Airsim, Udacity etc.

### 2019-12-18 paper/3_Methodology.tex
GC method para1

@@
 \section{Methodology} \label{method}
-\noindent A method for determining the level of simulation variance of an engine is presented in this section and the working flow of the method is shown in Figure \ref{method_diagram}.
+\noindent A method for determining the level of simulation variance of an engine is presented in this section and the working flow of the method is shown in Figure~\ref{method_diagram}.
 Each of the steps in the method is elaborated in the following subsections.
@@
 \subsection{Design experiment}
-\noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic? ii) If not, how does its simulation variance vary?\\\\
+The experiment should be a simple scenario that can be easily repeated with the same initial conditions.
+Logs should be used to record the actor positions at fixed time intervals throughout the simulation.
+For each actor the variance in trajectory can be determined and the deviation from the mean, $S_t$, of the trajectory monitored at each time step.
+Averages over time and for different actors can be used to condense results down to single figures which is done for some results.
+
+We use game engine actors to simulate the AV and pedestrians.
+Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.
+A* search algorithm.
+% 
+% Resource - Realism
+To simulate the demands of realistic rendering and complex physics calculations in a simple scenario, stress routines are used to artificially occupy system resources. 
+% Collision callbacks
+Further to this, the experiment should be designed to create collision callbacks to the physics engine by allowing actors to collide during the scenario. 
+%
+The user should also design the experiment to monitor and record system CPU and GPU utilisation.
+Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
+%
+Where possible a fixed $dt$ should be used and any random numbers used should be seed controlled.
+
+% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
+ii) If not, how does its simulation variance vary?
 In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
 Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
-This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
+% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
 Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\

### 2019-12-18 paper/4_CaseStudy.tex

@@
 Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Fig \ref{Test_a}.\\\\
-Proceeding to step 2; by doing various inspections around the different internal settings of this simulator we did not notice any improvement or worsening in the results. 
-On the other hand the external settings (step 3) did show significant variations on the results as will be discussed in the next section. \\\\
+Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal setting option was disregarded.\\\\
+On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
 Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.\\\\
@@
 
-\subsection{Results And Discussion}
-\noindent We start by analysing (step 5) the summary from all of the tests plotted in Fig \ref{ExperimentsStressSummary}. 
+\subsection{Results And Discussion}\label{ResultsSection}
+\noindent We start by analysing (step 5) the summary from all of the tests plotted in Fig \ref{ExperimentsStressSummary}; where the x-axis represents the percentage of CPU + GPU utilisation and the y-axis shows the total mean devaition.
+This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.\\\\
 Note the dotted red line in the plot shows the millimeter (mm) level. 
@@
     \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummary.pdf}
-    \caption{Summary of mean deviation for the different tests runs plotted against the different CPU+GPU stresses.}
+    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.}
 
@@
     \end{subfigure}
-    \caption{Test ID 2 (a)Mean deviation vs simulation time for 25\% stress. (b)Pre and post collision mean deviation at different CPU+GPU stresses.}
+    \caption{Test ID 2 (a)Mean deviation over runs no. vs simulation time for 25\% stress. (b)Pre and post collision mean deviation over runs no. for different CPU + GPU stresses.}
 \end{figure*}
@@
     \end{subfigure}
-    \caption{Test ID 4 (a)Mean deviation vs simulation time for 25\% stress. (b)Pre and post collision mean deviation at different CPU+GPU stresses.}
+    \caption{Test ID 4 (a)Mean deviation over runs no. vs simulation time for 25\% stress. (b)Pre and post collision mean deviation over runs no. for different CPU + GPU stresses.}
 \end{figure*}
@@
 We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects. \\\\
-\noindent Thus, we resorted to plotting the mean deviation of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Fig.
+\noindent Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Fig.
 \ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
@@
 The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
-Thus failing to stay below our red line and the test should be terminated.
-
-% \noindent\underline{\textit{No collisions:}}
-% First, the tests with no collsions are considered, these represnt tests' ID 1, 3 and 5. ...
-% --General deterministic behaviour of all of the agnets whent there are no collisions (tests 1,3 and 5)\\\\
-% \noindent\underline{\textit{Pedestrians paths intersection:}}
-% Pedestrians do not just blindly follow their trajectories, as already iterated in section \ref{TestsDescriptionAndTechnicalities}; thus it it 
-% --Pathintersections bewteen pedestrains (tests 6,7 and 8)\\\\
-% \noindent\underline{\textit{Collision between vehicles:}}
-% Test 2 where the vehilcles collide together...
-% --Collsion between vehicles alone (test 2)\\\\
-
-% \noindent\underline{\textit{Collision between vehicles and pedestrians:}}
-% --collsion between vehicles and pedestrains(test 4)
-% -- coparison between variance before and after collsion
-
-% - after a collsion the level of non determinsim increases significantly 
-% - However In V\&V of CAV simulations, if a collsion is detected then the test should end
-
-\TODO{We might need to change the y axes nameing betweent he summary plot (fig 4) and the fig 5 and 6, to show that one is mean deviation of all runs and thourgh out simulation time and the other is only between the different runs.}
-
-\noindent In summary, from the tests carried out we can deduce that for a computer of the following specifications,-----..--, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
+Thus failing to stay below our red line and the test should be terminated.\\\\
+\noindent In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.

### 2019-12-19 paper/1_Intro.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

@@
 %
-\noindent There have been a number of fatalities with autonomous vehicles which could be attributed to a lack of verification and validation (V\&V), e.g.\cite{FatalityExample}.
+There have been a number of fatalities with autonomous vehicles which could be attributed to a lack of verification and validation (V\&V), e.g.\cite{FatalityExample}.
 %
 Simulation is an obvious domain choice to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving} where the rare event can be made to happen more often~\cite{Koopman2018} such as unpredictable actor behaviour~\cite{RobustnessAutonomy}.
-%
 
@@
 % Game engines for SBV
-Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements but many challenges still exist.
-%
+Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements, as opposed to vehicle dynamics simulators and traffic-level simulators used my manufacturers~\cite{FrameworkAndChallenges}; but many challenges still exist.
+
 % Verification and Determinism 
-For games engines to be useful as a verification tool then they must be deterministic. 
+For game engines to be useful as a verification tool then they must be deterministic. 
 %

### 2019-12-19 paper/2_Background.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

@@
 \TODO{ref each of these}
-%
+
 Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution.
@@
 \subsection{Overview of a Game Loop} \label{GameLoopSection}
-The game loop is responsible for the interaction between the maths, physics and rendering engines~\cite{GameEngineArchBook}.
+The game loop is responsible for the interaction between the maths, physics and rendering engines.
 Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop.
-A traditional \TODO{For AG: how is it traditional, what is the alternative?} game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).
+A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).~\cite{GameEngineArchBook}
 
@@
 The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment.
-\TODO{For AG: need to expand, e.g what are the inputs?} \TODO{For AG: does game logic need to be mentioned?}
+For example, the throttle and braking inputs from the PID controller controlling the speed of an AV.
+%\TODO{For AG: need to expand, e.g what are the inputs?} %\TODO{For AG: does game logic need to be mentioned?}
 %
-The update interprets the intended actions of the AV along with all other dynamic actors in the scene and performs any necessary physics calculations and updates the actor states.
-\TODO{For AG: what does the clock symbolise, need to explain it?}
+The update interprets the intended actions of all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states.
+%This update process can still proceed independelty of the main game loop as symbolised by the clock in the upper loop in Figure~\ref{GameEngineLoopDiagram}.
+% \TODO{For AG: what does the clock symbolise, need to explain it?}
 %
 The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
-\TODO{For AG: you cite this book 4 times which might be too much as a general reference, is there something specific you want to draw the reader towards in the book?
+%\TODO{For AG: you cite this book 4 times which might be too much as a general reference, is there something specific you want to draw the reader towards in the book?
 What specifically about the rendering part of this cycle does the book help to illustrate the point?}
 
-It is crucial for this paper to understand how the rendering and physics engines operate together. \TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
-Briefly, the game loop operates as follows:\cite{GameProgPatternsBook} (see Figure \ref{GameEngineLoopDiagram})
+It is necessary to have an understanding of how rendering and physics engines operate together in order to grasp some of the sources of non-determinsim that will be mentioned.
+%\TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
+The game loop operates as follows: (see Figure \ref{GameEngineLoopDiagram})
 \begin{itemize}[leftmargin=*]
@@
     \item Then the user inputs are processed.
-    \item There is then an inner loop to update the physics engine (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at fixed step until the game clock is equal to the real world.
+    \item There is then an inner loop to update the physics engine (clock symbol in Figure~\ref{GameEngineLoopDiagram}), incrementing at fixed step until the game clock is equal to the real world.
 The physics engine uses a fixed time step, because it makes everything simpler and more stable for physics and AI.
-The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, as will be explained in section \ref{s:nondeterminisimSources}.
+The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
 Thus, ideally the time step should be as short as possible, so that simulations run with high fidelity on fast machines.
@@
 Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}.
-If the engine tries to render between updates \textit{residual lag} can occur \cite{GameEngineArchBook}\cite{GameProgPatternsBook} and extrapolation between frames is performed to smooth transition between scenes.
+If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes.%\cite{GameEngineArchBook}\cite{GameProgPatternsBook}
 
@@
 
-% ======= Sources of non-determinism
-\subsection{Sources of non-determinism} \label{s:nondeterminisimSources}
-\noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic. 
-The main reasons that could cause non-determinism are discussed below. \\\\
-
-\noindent\underline{\textit{Floating Points:}}
-A generic issue with computers are floating points precision. 
-Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
-Translating to the usage of CAV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.\\\\ 
-\noindent For instance, consider a simulation where the world is infinite and progressively  generated in the physics engine. 
-After a few hundred meters, precision issues will start to occur, and will get progressively worse the further from the origin the simulation gets. 
-This is due to the values of the coordinates being used in calculations are getting bigger and hence will not cope with small values that are critical for V\&V testing purposes. 
-A solution to this problem is perhaps every time the simulations moves 100 meters away from the origin the entire world will move by 100m in the opposite direction.
-Thus, avoiding getting into floating point issues.\cite{FloatingPointArithmeticArticle}\\\\
-\noindent There are many scenarios that could be addressed, and solutions to them will entirely depend on the characteristics of that specific scenario.
-There is no generalised solution for the floating problem, regardless of how good computers get there will always be a limitation.
-However, floating points as they stand could just be sufficient for V\&V testing purposes.
-Whenever they are not, smart solutions can be found to accommodate for these limitations.\\\\
 \begin{figure*}[b]
@@
 \end{figure*}
-\noindent\underline{\textit{Navigation meshes and AI:}}
-Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic\cite{CARLABenchmark}. 
-The validity of this depends on what are the built in AI algorithms being used by the engine. 
-Most of them tend to use the A* algorithm\cite{AStarBook}, which is an algorithm with deterministic behaviour if the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
-Therefore the determinism of the AI depends on the determinism of the engine and how its navigation meshes are created or modelled.
-It is interesting to note that changes can occur to meshes every time the simulation is loaded.
-This is mainly resorted to the different scheduling of threads.
-\\\\
+
+% ======= Sources of non-determinism
+\subsection{Sources of non-determinism} \label{s:nondeterminisimSources}
+\noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic. 
+The main reasons that we think are causes of these engines' non-determinism are discussed below. \\\\
+%
+\noindent\underline{\textit{Floating Points:}}
+A generic issue with computers are floating points precision. 
+Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
+
+Translating to the usage of CAV simulations in physics engines, this could result in many precision issues; for instance consider a simulation where the world is infinite and progressively  generated in the physics engine.
+After a few hundred meters, precision issues will start to occur due to the coordinates used in arithmetic cacluations getting bigger, and will get progressively worse the further from the origin the simulation gets.
+Such reasons would often make one rapidly jump to conclude that floating points are the cause of non-determinism of these engines. 
+However, it is important to note that these precision issues should be referred to as incorrecctness and not non-determinisim. 
+Since hardware is deterministic, then the same hardware should produce the same exact output for the same inputted data i.e.
+even if the output is incorrect due to floating point precision it should always be the same, given the same hardware.\\\\
+%
+% A generic issue with computers are floating points precision. 
+%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
+%Translating to the usage of CAV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.
+%
+% For instance, consider a simulation where the world is infinite and progressively  generated in the physics engine. 
+% After a few hundred meters, precision issues will start to occur, and will get progressively worse the further from the origin the simulation gets. 
+% This is due to the values of the coordinates being used in calculations are getting bigger and hence will not cope with small values that are critical for V\&V testing purposes. 
+% A solution to this problem is perhaps every time the simulations moves 100 meters away from the origin the entire world will move by 100m in the opposite direction.
+Thus, avoiding getting into floating point issues.\cite{FloatingPointArithmeticArticle}
+%
+% There are many scenarios that could be addressed, and solutions to them will entirely depend on the characteristics of that specific scenario.
+There is no generalised solution for the floating problem, regardless of how good computers get there will always be a limitation.
+However, floating points as they stand could just be sufficient for V\&V testing purposes.
+Whenever they are not, smart solutions can be found to accommodate for these limitations.\\\\
+%
+\noindent\underline{\textit{Scheduling of threads:}} Scheduling of threads is believed to be the main reason for non-deterministic behaviour of game engines.
+Given the same hardware the output should always be the same, unless scheduling of the threads change i.e.
+the utilisation of hardware is manipulated differently each time.
+This would then result in non-deterministic behaviours especailly if it is combined with the floating point problem discussed earlier.
+To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
+To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order.
+This is beyond the sope of our work and will not be covered in this paper.\\\\
+%
 \noindent\underline{\textit{Physics and Rendering clocks:}}
@@
 A follow up from section \ref{GameLoopSection}, these engines use a variable frame rate, which is good for hardware scalability but creates a challenge for the physics engine which works best with small fixed time steps.
-The problem of having a variable frame rate could also be a reason for the non-deterministic behaviour of these engines.
-This issue could perhaps be of negligible significance if the time steps are small enough.
-\\
-
-\noindent\underline{\textit{Scheduling of threads:}} Scheduling of threads is believed to be the main reason for the non deterministic behaviour of these engines.
-Given the same hardware the output should always be the same, unless scheduling of the threads change, this would then cause non-deterministic behaviours.
-To solve this one would have to control all of the threads, which is a very involved and to achieve that one would need to replace or control the whole run-time system to allocate the same threads in the same order.
+The problem of having a variable frame rate could also contribute to the non-deterministic behaviour of these engines, especially if put together with the scheduling of threads problem.
+That is to say, if the frame rate is allowed to vary then it has to be controlled to give the same variablity through out a given test in order for the test results to be repeatable.
+This issue might be of negligible significance if the frame rate is consistent and the time steps are small enough.
+\\\\
+%
+\noindent\underline{\textit{Navigation meshes and AI:}}
+Game engines' often use navigation meshes for AI path planning. Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic\cite{CARLABenchmark}. 
+The validity of this depends on what are the built in AI algorithms being used by the engine. 
+Most of them tend to use the A* algorithm\cite{AStarBook}, which is an algorithm with deterministic behaviour if the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
+Therefore, the determinism of the AI depends on the determinism of the engine and how its navigation meshes are created/modelled, their shape, granularity etc.
+It is interesting to note that changes can occur to meshes every time the simulation is loaded or if dynamically loading these navigation meshes as the simulation proceeds.
+This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes everytime they are loaded.

### 2019-12-19 paper/3_Methodology.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

@@
-\section{Methodology} \label{method}
+\section{Methodology} \label{s:methodology}
 \noindent A method for determining the level of simulation variance of an engine is presented in this section and the working flow of the method is shown in Figure~\ref{method_diagram}.
 Each of the steps in the method is elaborated in the following subsections.
-
+%
 \subsection{Design experiment}
@@
 For each actor the variance in trajectory can be determined and the deviation from the mean, $S_t$, of the trajectory monitored at each time step.
-Averages over time and for different actors can be used to condense results down to single figures which is done for some results.
+Averages over time and for different actors can be used to condense results down to single figures.
 
@@
 Where possible a fixed $dt$ should be used and any random numbers used should be seed controlled.
-
 % \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
@@
 It is crucial to also make sure that any randomisation factors are eliminated in the experiment.
+%
+\subsection{Internal settings}
+There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine.
 
-\subsection{Internal settings}
-\noindent There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine. \\\\
-\noindent Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
+Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
 However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
-\\\\
-\noindent If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of simulation variance as a result of allowing the AI to navigate in a more well defined space.
-\\\\
-\noindent Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
-The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.  \\\\
-\noindent Running in editor or deployed mode can sometimes cause massive differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
+
+If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of simulation variance as a result of allowing the AI to navigate in a more well defined space.
+
+Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
+The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.
+
+Running in editor or deployed mode can sometimes cause massive differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
 
@@
 \toprule
-Test ID & Test description (See Fig \ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
+Test ID & Test description (See Figures~\ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
 1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
@@
 \subsection{External settings}
-\noindent Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.\\\\
-\noindent Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
+Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.
+
+Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
 Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.
@@
 \subsection{Output and analysis}
-\noindent Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
-Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.\\\\
+Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
+Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
+
 This whole process is repeated for various internal and external settings (if needed). 

### 2019-12-19 paper/4_CaseStudy.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

@@
-\section{Case Study} \label{CaseStudy}
-\noindent Given that our interest is on CAV simulations, this study presents a number of experiments run on typical agents used in CAV simulations and their interactions with each other.
+\section{Case Study} \label{s:case-study}
+Given that our interest is on CAV simulations, this study presents a number of experiments run on typical agents used in CAV simulations and their interactions with each other.
 These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-\\\\
+
 Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Fig \ref{method_diagram} as we go through this case study.
+We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
 
 \subsection{Problem And Hypothesis}
-\noindent We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the level of simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.\\\\
+We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the level of simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+
 Our hypothesis is that since UE4 is a gaming engine it is not deterministic, however, the level of simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
@@
 \subsection{Tests Description And Technicalities}\label{TestsDescriptionAndTechnicalities}
-\noindent The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints. 
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.\\\\ 
-\noindent Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments. 
+The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints. 
+Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
+
+Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments. 
 The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents. 
@@
 
-\noindent Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
 The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-\\\\
-Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.\\\\
-Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Fig \ref{Test_b}). 
+
+Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+
+Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
 Pedestrians have the functionality of dynamic path planning i.e. they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other. 
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.\\\\
-In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.\\\\
-Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Fig \ref{Test_a}.\\\\
+Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
+
+In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+
+Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+
 Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal setting option was disregarded.\\\\
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+
 On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.\\\\
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
+
 In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
@@
 \subsection{Results And Discussion}\label{ResultsSection}
-\noindent We start by analysing (step 5) the summary from all of the tests plotted in Fig \ref{ExperimentsStressSummary}; where the x-axis represents the percentage of CPU + GPU utilisation and the y-axis shows the total mean devaition.
-This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.\\\\
-Note the dotted red line in the plot shows the millimeter (mm) level. 
-Below this level we assume that the level of simulation variance is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. 
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic, but just for the sake of argument we chose it to be the millimeter level.\\\\
-First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
-with in the centimeter level).
-This clearly shows how the load on a computer (i.e.
-stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
-This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
-\\\\
-Next, given the pedestrians were showing a very robust and low level of simulation variance, we ran tests for pedestrians colliding (i.e.
-with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
-
+We start by analysing (step 5) the summary from all of the tests plotted in Figure~\ref{ExperimentsStressSummary}; where the x-axis represents the percentage of CPU + GPU utilisation and the y-axis shows the total mean devaition.
+This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
 \begin{figure}[h]
@@
 \end{figure*}
+Note the dotted red line in the plot shows the millimeter (mm) level. 
+Below this level we assume that the level of simulation variance is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. 
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic, but just for the sake of argument we chose it to be the millimeter level.
 
-\noindent Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively. 
-We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects. \\\\
-\noindent Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Fig.
-\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
-Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Fig \ref{ExperimentsStressSummary} to always surpass our red line.\\\\
-\noindent Plotting Figs \ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
+First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
+Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
+with in the centimeter level).
+This clearly shows how the load on a computer (i.e.
+stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
+This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
+
+Next, given the pedestrians were showing a very robust and low level of simulation variance, we ran tests for pedestrians colliding (i.e.
+with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
+
+Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively. 
+We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
+
+Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
+Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
+
+Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
 However, in V\&V of CAV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
-Thus, the post collision effect should not be of much concern in the context of CAVs.\\\\
-\noindent It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
+Thus, the post collision effect should not be of much concern in the context of CAVs.
+It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
 The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
-Thus failing to stay below our red line and the test should be terminated.\\\\
-\noindent In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
+Thus failing to stay below our red line and the test should be terminated.
+In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.

### 2019-12-19 paper/6_Conclusions.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

@@
-\section{Conclusions}
-\noindent We have presented a method by which one can determine the level of non-determinism of a physics engine used in V\&V application of CAV simulations.
-The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance and finally focuses on running the experiment and analysing the output.
-\\\\
+\section{Conclusions}\label{s:conclusion}
+We have provided a summary of the sources of non-determinisim in game engines.
+This was followed by presenting a method by which one can determine the level of simulation variance of a gaming engine used in V\&V application of CAV simulations.
+The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
+
 A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
-It was found that for the computer used (with specs of----) the level of non determinism is sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
+It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the level of non determinism is sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
 %since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
-% \TODO{Add the conclusion mentioned by Thales}
+
+In highlight of obtaining a fully determinstic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
+Alternatively, another approach would be to create a new engine that is designed to be deterministic with minimal attention given to performance.

### 2019-12-19 paper/4_CaseStudy.tex
nothing much

@@
 By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
 
@@
 \subsection{Results And Discussion}\label{ResultsSection}
-We start by analysing (step 5) the summary from all of the tests plotted in Figure~\ref{ExperimentsStressSummary}; where the x-axis represents the percentage of CPU + GPU utilisation and the y-axis shows the total mean devaition.
+We start by analysing (step 5) the summary from all of the tests plotted in Figure~\ref{ExperimentsStressSummary}; where the $x$-axis represents the percentage of CPU + GPU utilisation and the $y$-axis shows the total mean devaition.
 This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
@@
 Note the dotted red line in the plot shows the millimeter (mm) level. 
-Below this level we assume that the level of simulation variance is very low for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. 
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic, but just for the sake of argument we chose it to be the millimeter level.
+Below this level we assume that the level of simulation variance is low enough that it would not affect our assertion checking for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
 

### 2019-12-19 paper/1_Intro.tex
GC BG day 2

@@
 Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it.
-These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988}.
+These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving assertions which must not be violated.
+Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules through assertion checking.
 %
@@
 % Requirements of simulation based verification SBV
-For games engines to offer a suitable testing environment they must: operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors), be realistic enough \cite{Koopman2018} including suitable rendering for perception stack testing, be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015}, include the provision for sensor analogues to a suitable accuracy (e.g.
-video, LIDAR, IR), allow hardware-in-the-loop development and provide a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}.
+For games engines to offer a suitable testing environment they must operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors) and be realistic enough \cite{Koopman2018} to produce suitable rendering of scenes for perception stack testing.
+
+The simulation must be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015} and include a provision for sensor analogues to a suitable degree of accuracy (e.g.
+video, LIDAR, IR).
+Some researchers may also require support for hardware-in-the-loop development or a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}.
 %, e.g. minimum rate of frames per second (fps).
@@
 % Game engines for SBV
-Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements, as opposed to vehicle dynamics simulators and traffic-level simulators used my manufacturers~\cite{FrameworkAndChallenges}; but many challenges still exist.
+Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements, as opposed to vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}; but many challenges still exist.
 
@@
 Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe vehicle behaviour.
+%
+Also, if the simulation is non-deterministic, variance in actor positions may lead to subsequent assessment errors, e.g. assertion checking.
 

### 2019-12-19 paper/2_Background.tex
nothing much
GC BG day 2

@@
 There are numerous game engines with their associated development environments which could be suitable for CAV development, e.g.
-Unreal, Unity, CryEngine.
+Unreal~\footnote{\url{https://www.unrealengine.com/}}, Unity~\footnote{\url{https://unity.com/}}, CryEngine~\footnote{\url{https://www.cryengine.com/}}.
 Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.
-Carla, AirSim, Apollo, and some have been developed specifically, e.g.
-the could based Nvidia Drive Constellation simulator.
-\TODO{ref each of these}
+Carla~\footnote{http://carla.org/}, AirSim~\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo~\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.
+Nvidia Drive Constellation~\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
 
@@
 
-The time step, $dt$, in any physics calculation is important.
+The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
+If $dt$ is too large or irregularly spaced then the rendered motion may not be smooth and interpolation may be required at additional computational cost.
 To use a fixed physics time step the user's display refresh rate needs to be known in advance.

### 2019-12-19 paper/2_Background.tex
some changes to game loop

--- 2019-12-19 paper/2_Background.tex
@@
 \includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
-\caption{Game engine loop block diagram}
+\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}}
 \label{GameEngineLoopDiagram}
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagramCropped.pdf}
-    \caption{Shows a flow diagram of the methodology proposed to define the level of non-determinism of a physics engine.}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
+    \caption{Shows a flow diagram of the methodology proposed to define the level of simulation variance of a gaming engine for usage in CAV simulations.}
 

### 2019-12-19 paper/4_CaseStudy.tex
some changes to game loop

@@
 We start by analysing (step 5) the summary from all of the tests plotted in Figure~\ref{ExperimentsStressSummary}; where the $x$-axis represents the percentage of CPU + GPU utilisation and the $y$-axis shows the total mean devaition.
-This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
 \begin{figure}[h]
@@
 \end{figure*}
+This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
+
 Note the dotted red line in the plot shows the millimeter (mm) level. 

### 2019-12-19 paper/2_Background.tex
GC BG day 2

--- 2019-12-19 paper/2_Background.tex
@@
 Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.
-Carla~\footnote{http://carla.org/}, AirSim~\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo~\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.
+Carla~\footnote{\url{http://carla.org/}}, AirSim~\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo~\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.
 Nvidia Drive Constellation~\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
@@
 What specifically about the rendering part of this cycle does the book help to illustrate the point?}
+% 
+% It is necessary to have an understanding of how rendering and physics engines operate together in order to grasp some of the sources of non-determinsim that will be mentioned.
+%\TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
+% 
+It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below. 
+% 
+The key aspects of the game loop needed for this discussion are as follows (initialisation, game logic and decommissioning are removed from this diagram for brevity), see Fig.~\ref{GameEngineLoopDiagram}:
 
-It is necessary to have an understanding of how rendering and physics engines operate together in order to grasp some of the sources of non-determinsim that will be mentioned.
-%\TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
-The game loop operates as follows: (see Figure \ref{GameEngineLoopDiagram})
 \begin{itemize}[leftmargin=*]
@@
     \item Then the user inputs are processed.
-    \item There is then an inner loop to update the physics engine (clock symbol in Figure~\ref{GameEngineLoopDiagram}), incrementing at fixed step until the game clock is equal to the real world.
-The physics engine uses a fixed time step, because it makes everything simpler and more stable for physics and AI.
-The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
-Thus, ideally the time step should be as short as possible, so that simulations run with high fidelity on fast machines.
-However, if the fixed time step is too short i.e.
-less than the time it takes to process an ``Update" inner loop on some (slow) machines, then the simulation will simply never catch up on these slow machines.
-    \item Rendering occurs once the physics engine catches up. The process then starts again.
+    \item There is then an inner loop to update the physics processes (clock symbol in Figure~\ref{GameEngineLoopDiagram}), incrementing at a fixed delta time step, $dt$, until the game clock is equal to the real world.
+    \item Rendering occurs once the game clock has caught up with the real world clock. 
+    \item The process then repeats.
 \end{itemize}
 
-The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
+The physics engine uses a fixed time step and %, because it makes everything simpler and more stable for physics and AI.
+the shorter this fixed time step is, the smoother the interpretation of the physical dynamics will be.
+%processing time it takes to catch up to game clock.
+% and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
+%
+To use a fixed physics time step the user's display refresh rate needs to be known in advance.
+This requires an update loop to take less than one render tick (one frame of real world time).
+Given the range of different hardware capabilities a variable delta time can be implemented taking the previous frame time as the next $dt$.
+However, variable $dt$ can lead to different physical results and in some cases unrealistic physical representations~\cite{gaffer}.
+Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware.
+Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}}.
+If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes.
+%\cite{GameEngineArchBook}\cite{GameProgPatternsBook}
+%
+% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
+In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
+and the simulation will exhibit delays between input commands and rendered states resulting in unsychronised and unrealistic behaviour.
+
+% The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
 If $dt$ is too large or irregularly spaced then the rendered motion may not be smooth and interpolation may be required at additional computational cost.
-To use a fixed physics time step the user's display refresh rate needs to be known in advance.
-This requires an update loop to take less than one frame of real world time~\cite{gaffer}.
-Given the different capabilities of game players hardware a variable delta time can be implemented taking the previous frame time as the next $dt$.
-However, variable $dt$ can lead to different physical results and in some cases unrealistic physics.
-Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware.
-Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}.
-If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes.%\cite{GameEngineArchBook}\cite{GameProgPatternsBook}
 
@@
 % ======= Sources of non-determinism
-\subsection{Sources of non-determinism} \label{s:nondeterminisimSources}
-\noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic. 
-The main reasons that we think are causes of these engines' non-determinism are discussed below. \\\\
-%
+\subsection{Sources of Non-Determinism} \label{s:nondeterminisimSources}
+
+The following points form a hypothesis around potential sources of non-determinism.
+% \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
+The main reasons that we think are causes of these engines' non-determinism are discussed below.
+\\\\
+
 \noindent\underline{\textit{Floating Points:}}
-A generic issue with computers are floating points precision. 
-Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
+Floating point representation of real numbers is limited with fixed register sizes resulting in errors of \textit{rounding} where loss of bits occur and are rounded or truncated and \textit{cancellation} which happens where two similar values are subtracted
+% A generic issue with computers is floating points precision.
+Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.
+~\cite{FloatingPointsBook}\cite{goldberg1991every}.
 
@@
 Such reasons would often make one rapidly jump to conclude that floating points are the cause of non-determinism of these engines. 
-However, it is important to note that these precision issues should be referred to as incorrecctness and not non-determinisim. 
+However, it is important to note that these precision issues should be referred to as incorrectness and not non-determinism. 
 Since hardware is deterministic, then the same hardware should produce the same exact output for the same inputted data i.e.

### 2019-12-19 paper/2_Background.tex
GC BG day 2

@@
 The game loop is responsible for the interaction between the maths, physics and rendering engines.
-Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop.
+Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop (initialisation, game logic and decommissioning are removed from this diagram for brevity).
 A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).~\cite{GameEngineArchBook}
@@
 The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment.
-For example, the throttle and braking inputs from the PID controller controlling the speed of an AV.
-%\TODO{For AG: need to expand, e.g what are the inputs?} %\TODO{For AG: does game logic need to be mentioned?}
+For example, the throttle and steering inputs from the vehicle controller of an AV
 %
 The update interprets the intended actions of all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states. 
-%This update process can still proceed independelty of the main game loop as symbolised by the clock in the upper loop in Figure~\ref{GameEngineLoopDiagram}.
-% \TODO{For AG: what does the clock symbolise, need to explain it?}
 %
 The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}. 
-%\TODO{For AG: you cite this book 4 times which might be too much as a general reference, is there something specific you want to draw the reader towards in the book?
-What specifically about the rendering part of this cycle does the book help to illustrate the point?}
-% 
-% It is necessary to have an understanding of how rendering and physics engines operate together in order to grasp some of the sources of non-determinsim that will be mentioned.
-%\TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
 % 
@@
 % 
-The key aspects of the game loop needed for this discussion are as follows (initialisation, game logic and decommissioning are removed from this diagram for brevity), see Fig.~\ref{GameEngineLoopDiagram}:
+The key aspects of the game loop needed for this discussion are as follows, see Fig.~\ref{GameEngineLoopDiagram}:
 
@@
 
-The following points form a hypothesis around potential sources of non-determinism.
+The following points form a hypothesis around potential sources of non-determinism. \\\\
 % \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
@@
 ~\cite{FloatingPointsBook}\cite{goldberg1991every}. 
-
-Translating to the usage of CAV simulations in physics engines, this could result in many precision issues; for instance consider a simulation where the world is infinite and progressively  generated in the physics engine.
-After a few hundred meters, precision issues will start to occur due to the coordinates used in arithmetic cacluations getting bigger, and will get progressively worse the further from the origin the simulation gets.
-Such reasons would often make one rapidly jump to conclude that floating points are the cause of non-determinism of these engines. 
-However, it is important to note that these precision issues should be referred to as incorrectness and not non-determinism. 
-Since hardware is deterministic, then the same hardware should produce the same exact output for the same inputted data i.e.
+%
+% Translating to the usage of CAV simulations in physics engines,
+This could result in precision issues of, for example, actor position within the environment and falsely satisfying or failing assertions.
+%; for instance consider a simulation where the world is infinite and progressively  generated in the physics engine. 
+% After a few hundred meters, precision issues will start to occur due to the coordinates used in arithmetic cacluations getting bigger, and will get progressively worse the further from the origin the simulation gets.
+Such reasons would often make one jump to conclude that floating points are the cause of non-determinism, but these precision issues should be referred to as \textit{incorrectness} and not non-determinism.
+Since hardware is assumed deterministic, then the same hardware should produce the same \textit{incorrect} output for the same initial conditions.
+%, i.e.
 even if the output is incorrect due to floating point precision it should always be the same, given the same hardware.\\\\
@@
 However, floating points as they stand could just be sufficient for V\&V testing purposes.
-Whenever they are not, smart solutions can be found to accommodate for these limitations.\\\\
+Whenever they are not, smart solutions can be found to accommodate for these limitations.
+\\\\
+\noindent\underline{\textit{Multi-Thread Scheduling:}} Scheduling of threads may be a reason for non-deterministic simulation behaviour in game engines.
+Given the same hardware the output should always be the same, unless scheduling of the threads change.
+% i.e.
+the utilisation of hardware is manipulated differently each time.
+This would then result in non-deterministic behaviours especially if it is combined with the floating point problem discussed earlier.
 %
-\noindent\underline{\textit{Scheduling of threads:}} Scheduling of threads is believed to be the main reason for non-deterministic behaviour of game engines.
-Given the same hardware the output should always be the same, unless scheduling of the threads change i.e.
-the utilisation of hardware is manipulated differently each time.
-This would then result in non-deterministic behaviours especailly if it is combined with the floating point problem discussed earlier.
-To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
+For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
+Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available.
+This may lead to events happening out of order depending on the computational load and hence non-deterministic.
+%
+% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
 To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order.
-This is beyond the sope of our work and will not be covered in this paper.\\\\
+This is beyond the scope of our work and will not be covered in this paper.
+\\\\
+\noindent\underline{\textit{Physics and Rendering clocks:}}
+% The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
+Some game engines use a variable frame rate, which is good for hardware scalability but creates a challenge for the physics engine which works best with small fixed time steps. %
 %
-\noindent\underline{\textit{Physics and Rendering clocks:}}
-The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic. 
-A follow up from section \ref{GameLoopSection}, these engines use a variable frame rate, which is good for hardware scalability but creates a challenge for the physics engine which works best with small fixed time steps.
-The problem of having a variable frame rate could also contribute to the non-deterministic behaviour of these engines, especially if put together with the scheduling of threads problem.
-That is to say, if the frame rate is allowed to vary then it has to be controlled to give the same variablity through out a given test in order for the test results to be repeatable.
+The problem of having a variable frame rate could also contribute to the non-deterministic behaviour of these engines, especially combined with loss of thread scheduling synchrony. 
+%
+That is to say, if the frame rate is allowed to vary then it has to be controlled to give the same variability through out a given test in order for the test results to be repeatable.
 This issue might be of negligible significance if the frame rate is consistent and the time steps are small enough.
 \\\\
+\noindent\underline{\textit{Actor Navigation:}}
+Game engines often use navigation meshes for actor path planning, which should be considered as a potential source of non-deterministic behaviour, e.g.%.
+Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic
+~\cite{CARLABenchmark}. 
+This may depend on the algorithms used by the engine which typically use the A* algorithm\cite{AStarBook} and should give deterministic outcomes as long as the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
+Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
 %
-\noindent\underline{\textit{Navigation meshes and AI:}}
-Game engines' often use navigation meshes for AI path planning. Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic\cite{CARLABenchmark}. 
-The validity of this depends on what are the built in AI algorithms being used by the engine. 
-Most of them tend to use the A* algorithm\cite{AStarBook}, which is an algorithm with deterministic behaviour if the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
-Therefore, the determinism of the AI depends on the determinism of the engine and how its navigation meshes are created/modelled, their shape, granularity etc.
-It is interesting to note that changes can occur to meshes every time the simulation is loaded or if dynamically loading these navigation meshes as the simulation proceeds.
-This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes everytime they are loaded.
+It is interesting to note that changes can occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps~\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/index.html}}.
+%This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes every time they are loaded.

### 2019-12-20 paper/3_Methodology.tex
nothing much
GC BG day 3

@@
 \section{Methodology} \label{s:methodology}
-\noindent A method for determining the level of simulation variance of an engine is presented in this section and the working flow of the method is shown in Figure~\ref{method_diagram}.
-Each of the steps in the method is elaborated in the following subsections.
+
+% \noindent 
+In this section a method for determining the level of \textit{simulation variance} is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
+The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may influence thread scheduling.
+To replicate high computational loads, and hence game loop update rates, that may be experienced in high realism simulations, resource utilisation software is used at runtime for both central and graphical processing units.
+
+ % in this section and the working flow of the method is shown in Figure~\ref{method_diagram}. Each of the steps in the method is elaborated in the following subsections.
 %
+
 \subsection{Design experiment}
-The experiment should be a simple scenario that can be easily repeated with the same initial conditions.
+The experiment should be a short scenario that can be easily repeated with the same initial conditions.
 Logs should be used to record the actor positions at fixed time intervals throughout the simulation.
-For each actor the variance in trajectory can be determined and the deviation from the mean, $S_t$, of the trajectory monitored at each time step.
-Averages over time and for different actors can be used to condense results down to single figures.
+For each actor the variance in path trajectory should be determined and the deviation from the mean of that trajectory recorded at each time step.
+Average deviation over time and for different actors can be used to condense results down to single figures.
+Tests should be repeated enough of times to ensure statistical power~\cite{cohen2013statistical} reveals any low probability errors.
+If only a small number of tests can be performed  then maximum, rather than average deviation should be considered
 
-We use game engine actors to simulate the AV and pedestrians.
+We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
 Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.

### 2019-12-20 paper/2_Background.tex
.....
Did some minor tweaks

@@
 The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment.
-For example, the throttle and steering inputs from the vehicle controller of an AV
+For example, the throttle and steering inputs from the vehicle controller of an AV.
 %
@@
     \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
-    \caption{Shows a flow diagram of the methodology proposed to define the level of simulation variance of a gaming engine for usage in CAV simulations.}
+    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a gaming engine for usage in CAV simulations.}
 

### 2019-12-20 paper/6_Conclusions.tex
Did some minor tweaks

@@
 We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the level of simulation variance of a gaming engine used in V\&V application of CAV simulations.
+This was followed by presenting a method by which one can determine the simulation variance of a gaming engine used in V\&V application of CAV simulations.
 The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
@@
 A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
-It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the level of non determinism is sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
+It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
 %since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.

### 2019-12-23 paper/2_Background.tex
GC BG day 4

--- 2019-12-19 paper/2_Background.tex
@@
 The game loop is responsible for the interaction between the maths, physics and rendering engines.
-Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop (initialisation, game logic and decommissioning are removed from this diagram for brevity).
+Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop (initialisation, game logic and decommissioning are removed from this diagram for brevity\footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}).
 A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).~\cite{GameEngineArchBook}

### 2019-12-23 paper/3_Methodology.tex

@@
 % \noindent 
-In this section a method for determining the level of \textit{simulation variance} is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
+In this section a method for determining the \textit{simulation variance} is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
 The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may influence thread scheduling.
@@
 
-Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
+Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
 However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
 
-If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of simulation variance as a result of allowing the AI to navigate in a more well defined space.
+If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the simulation variance as a result of allowing the AI to navigate in a more well defined space.
 
@@
 
-Running in editor or deployed mode can sometimes cause massive differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
+Running in editor or deployed mode can sometimes cause differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
 

### 2019-12-23 paper/4_CaseStudy.tex

@@
 \subsection{Problem And Hypothesis}
-We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the level of simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
 
-Our hypothesis is that since UE4 is a gaming engine it is not deterministic, however, the level of simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
+Our hypothesis is that since UE4 is a gaming engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
 This behaviour is expected not to hold as the level of computing stress increases.
@@
 Note the dotted red line in the plot shows the millimeter (mm) level. 
-Below this level we assume that the level of simulation variance is low enough that it would not affect our assertion checking for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
+Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
 This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
@@
 
-Next, given the pedestrians were showing a very robust and low level of simulation variance, we ran tests for pedestrians colliding (i.e.
+Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
 with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.

### 2020-01-17 paper/3_Methodology.tex
gc changes method threshold

@@
 The experiment should be a short scenario that can be easily repeated with the same initial conditions.
-Logs should be used to record the actor positions at fixed time intervals throughout the simulation.
+Data logging should be used to record the actor positions at fixed time intervals throughout the simulation.
 For each actor the variance in path trajectory should be determined and the deviation from the mean of that trajectory recorded at each time step.
-Average deviation over time and for different actors can be used to condense results down to single figures.
+The average deviation over time can be calculated and this can also be averaged across different actors to condense results down to single figures.
 Tests should be repeated enough of times to ensure statistical power~\cite{cohen2013statistical} reveals any low probability errors.
@@
 It is crucial to also make sure that any randomisation factors are eliminated in the experiment.
-%
+
+The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
+For the majority of CAV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
+
+Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
+This is to ensure that an outlier data-point is not hidden in the average of $k$-tests
+
 \subsection{Internal settings}

### 2020-01-22 paper/3_Methodology.tex
GC paper updates

@@
 
-% \noindent 
 In this section a method for determining the \textit{simulation variance} is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
@@
 
- % in this section and the working flow of the method is shown in Figure~\ref{method_diagram}. Each of the steps in the method is elaborated in the following subsections.
-%
+\subsection{Experimental Aims}
+The aim of the experiment is to determine if the simulation is deterministic.
+This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories.
+Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.
+The user must specify the level of variance sufficient for their application which is used as a threshold for deterministic execution.
+In our case study, Section~\ref{s:case-study}, a threshold of $1cm$ was chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
+This value is sufficiently small to allow accurate \textit{assertion checking}[ref] in urban environments but a larger value may be acceptable for non-urban environments, e.g.
+motorway/highway.
+Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
+$1\mu m$.
 
@@
 The experiment should be a short scenario that can be easily repeated with the same initial conditions.
-Data logging should be used to record the actor positions at fixed time intervals throughout the simulation.
-For each actor the variance in path trajectory should be determined and the deviation from the mean of that trajectory recorded at each time step.
-The average deviation over time can be calculated and this can also be averaged across different actors to condense results down to single figures.
-Tests should be repeated enough of times to ensure statistical power~\cite{cohen2013statistical} reveals any low probability errors.
-If only a small number of tests can be performed  then maximum, rather than average deviation should be considered
+Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
+The maximum deviation in actor path with respect to the mean for $n$-repeated tests will indicate at what dimensional scale the game engine is deterministic.
+For example, if 100 tests are executed and the maximum deviation in actor path is $<1cm$, then this scenario could be interpreted as being deterministic above this scale.
 
-We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
+Tests should be sufficiently repeated to ensure that statistical power~\cite{cohen2013statistical} reveals any outlier errors that occur with low probability.
+Initial testing with the Unreal Engine indicated an actor path deviation of $1\times10^{-13} cm$ for 998 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
+Executing 100 repeats may seem sufficient but in this case would fail to observe these events that occur with low probability, inferring a false confidence in the results.
+
+Game engine actors can be used to simulate the AV and other vehicles and pedestrians.
+For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
+Hard-coded waypoint values ensure all actors are given the same initial conditions and any variation comes from execution of the simulator at runtime.
+%
+% \todo{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
+
+% \todo{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
+%
+% For the vehicles in the simulation, waypoints are generated and interpolated in the same way but execution at runtime is slightly different where a PID controller is used to generate a velocity profile and steering angle requests to reach the destination targets.}
+
+% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
 Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.
 A* search algorithm.
-% 
+
 % Resource - Realism
-To simulate the demands of realistic rendering and complex physics calculations in a simple scenario, stress routines are used to artificially occupy system resources. 
+To simulate the demands of realistic rendering and complex physics calculations in an otherwise simple scenario, stress programs are used to artificially occupy system resources. 
 % Collision callbacks

### 2020-02-07 paper/3_Methodology.tex
update to para 1-3

@@
 In this section a method for determining the \textit{simulation variance} is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
-The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may influence thread scheduling.
-To replicate high computational loads, and hence game loop update rates, that may be experienced in high realism simulations, resource utilisation software is used at runtime for both central and graphical processing units.
+The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may tax different aspects of the engine such as actor navigation and collision callbacks.
+To replicate high computational loads that may be experienced in high realism simulations, resource utilisation software is used at runtime for both central and graphical processing units.
+The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
 
-\subsection{Experimental Aims}
 The aim of the experiment is to determine if the simulation is deterministic.
@@
 Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.
+
+\subsection{Deterministic Threshold}
+We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($sigma^2$), which is quoted in terms of deviation ($sigma$) for ease of interpretation.
+This threshold is analogous to the analogue level associated with a binary state in a logical processor.
+At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.
+
 The user must specify the level of variance sufficient for their application which is used as a threshold for deterministic execution.
 In our case study, Section~\ref{s:case-study}, a threshold of $1cm$ was chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
-This value is sufficiently small to allow accurate \textit{assertion checking}[ref] in urban environments but a larger value may be acceptable for non-urban environments, e.g.
+This value is sufficiently small to allow accurate \textit{assertion checking} of autonomous vehicles in urban environments [ref] but a larger value may be acceptable for non-urban environments, e.g.
 motorway/highway.
@@
 
-Tests should be sufficiently repeated to ensure that statistical power~\cite{cohen2013statistical} reveals any outlier errors that occur with low probability.
-Initial testing with the Unreal Engine indicated an actor path deviation of $1\times10^{-13} cm$ for 998 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
-Executing 100 repeats may seem sufficient but in this case would fail to observe these events that occur with low probability, inferring a false confidence in the results.
+Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.
+Initial testing\footnote{Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
+While executing 100 repeats may seem sufficient, in this case the sample size would fail to observe these events that occur with low probability, inferring a false confidence in the results.
 

### 2020-02-11 paper/1_Intro.tex
update to scheduling in background

@@
 %
-Herein the term \textit{simulation variance} is used to refer to how simulation runs vary when repeated many times and specifically the variance of agent paths within a given scenario.
+Herein the term \textit{simulation variance} is used to refer to how simulation outputs vary when repeated many times and specifically the variance of actor paths within a given scenario.
 We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
@@
 A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
-% 
+
 % Why is determinism important for verification
@@
 Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
-Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe vehicle behaviour.
+Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour.
+%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
 %
@@
 Section~\ref{s:methodology} presents a method to design and execute an experiment to assess suitability of a system for CAV verification.
-A case study to illustrate the level of simulation variance for a number of scenarios is explored in Section~\ref{s:case-study}.
+A case study to illustrate how the simulation variance vary for a number of scenarios is explored in Section~\ref{s:case-study}.
 We conclude in Section~\ref{s:conclusion} and give an outlook on future work.

### 2020-03-05 paper/cav_determinisim_main.tex
trying out gitignore is working
.....
GC BG day 3
GC changes to section 1-4

@@
 \pagenumbering{gobble} %suppresses page numbers
+\addbibresource{IROS.bib}
+\usepackage{balance} %balance columns
+
 %-------------------------------------------------------------------------------
@@
 \begin{document}
+% %
+% \pagenumbering{arabic}
+% %
+% \input{0_TittleAndAbastract.tex}
+% %
+% \input{1_Intro.tex}
+% %
+% \input{2_Background.tex}
+% %
+% \input{3_Methodology.tex}
+% %
+% \input{4_CaseStudy.tex}
+% %
+% % \input{5_ResultsAndDiscussion.tex}
+% %
+% \input{6_Conclusions.tex}
+% %
+% \input{9_FundingAndReferences.tex}
+% %
 
-\pagenumbering{arabic}
+% ***************************************************
+%  TITLE
+% ***************************************************
+\title{Investigating Determinism of Gaming Engines for Autonomous Vehicle Verification}
+\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
+\author[1,3]{Greg Chance}
+\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
+\author[2,3]{Tony Pipe}
+\author[1,3]{Kerstin Eder}
+\affil[1]{Trustworthy Systems Lab, University of Bristol, Bristol, UK}
+\affil[2]{University of West of England, Bristol, UK}
+\affil[3]{Bristol Robotics Lab, Bristol, UK}
+\maketitle
 
-\input{0_TittleAndAbastract.tex}
+\begin{abstract}
+\noindent 
+The connected and autonomous vehicle (CAV) community are adopting the use of gaming engines to develop vehicle control systems and testing environments using simulation.
+These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V).
+This work highlights the requirements these engines need (to fulfil this purpose/for CAV development) and proposes a method to determine \textit{simulation variance} for specific configurations and resource utilisation.
+A case study is used to illustrate the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
+% The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
+It is important for these engines to be deterministic in order for one to carry out these different processes.
+However, such engines are inherently non-deterministic.
+We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic.
+We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.
+\end{abstract}
 
-\input{1_Intro.tex}
+\begin{IEEEkeywords}
+Autonomous Driving, Determinism, Physics Engines, Connected Autonomous Vehicles (CAV), Verification and Validation(V\&V)
+\end{IEEEkeywords}
+\IEEEpeerreviewmaketitle
 
-\input{2_Background.tex}
+% ***************************************************
+%  INTRODUCTION
+% ***************************************************
+\section{Introduction} \label{s:introduction}
+% What is verification
+\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}.
+Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it.
+These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving assertions which must not be violated.
+Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules through assertion checking.
+There have been a number of fatalities with autonomous vehicles which could be attributed to a lack of verification and validation (V\&V), e.g.\cite{FatalityExample}.
+Simulation is an obvious domain choice to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving} where the rare event can be made to happen more often~\cite{Koopman2018} such as unpredictable actor behaviour~\cite{RobustnessAutonomy}.
 
-\input{3_Methodology.tex}
+% Requirements of simulation based verification SBV
+For games engines to offer a suitable testing environment they must operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors) and be realistic enough \cite{Koopman2018} to produce suitable rendering of scenes for perception stack testing.
+The simulation must be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015} and include a provision for sensor analogues to a suitable degree of accuracy (e.g.
+video, LIDAR, IR).
+Some researchers may also require support for hardware-in-the-loop development or a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}.
+% Game engines for SBV
+Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements, as opposed to vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}; but many challenges still exist.
 
-\input{4_CaseStudy.tex}
+% Verification and Determinism 
+For game engines to be useful as a verification tool then they must be deterministic. 
+%
+% What is a deterministic system
+Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
+%
+Herein the term \textit{simulation variance} is used to refer to how simulation outputs vary when repeated many times and specifically the variance of actor paths within a given scenario.
+We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
+Dynamic elements are the elements in a scene whose actions or behaviour may change over time, these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals.
+The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.
+A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
 
-% \input{5_ResultsAndDiscussion.tex}
+% Why is determinism important for verification
+Determinism is important for verification.
+Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
+Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour.
+%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
+%
+Also, if the simulation is non-deterministic, variance in actor positions may lead to subsequent assessment errors, e.g. assertion checking.
 
-\input{6_Conclusions.tex}
+% Challenges of SBV
+Our main research question is to what extent are gaming engines deterministic so that they can provide a reliable and repeatable testing environment for CAV verification?
+The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.
 
-\input{9_FundingAndReferences.tex}
+This paper is structured as follows.
+In Section~\ref{s:background} a background on gaming engines and sources of potential non-determinism are explored.
+Section~\ref{s:methodology} presents a method to design and execute an experiment to assess suitability of a system for CAV verification.
+A case study to illustrate how the simulation variance vary for a number of scenarios is explored in Section~\ref{s:case-study}.
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.
+
+% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (CAVs) on roads is a delicate if not utterly fascinating subject.
+Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.
+
+% Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.
+
+% It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
+
+% \noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
+
+% As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
+
+% ======================
+% This is all from "FrameworkAndChallenges" without using quotes 
+% \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system.
+Similarly, the use of simulation for traffic modelling is well established.
+Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate CAVs\cite{FrameworkAndChallenges}.\\\\
+
+% While traffic-level simulators can model road layouts, they do not require as much realism or detail as CAVs require.
+On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for CAV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
+% Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
+% Consequently, one of the major routes CAVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate CAVs \cite{FrameworkAndChallenges}.
+% ======================
+
+% Robopilot, Capri, Carla, Apollo, Airsim, Udacity etc.
+are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}.
+\\\\
+% Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
+% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of CAV software.\\\\
+% \TODO{Remove this paragraph?} Determinism of gaming (physics) engines has not attracted much attention by engines' developers nor researchers.
+This previously was not important since in gaming, determinism is not critical.
+% Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
+% On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
+
+% This paper aims to investigate how deterministic physics engines are for usage in V\&V for CAV simulations; and will cover the list of following points:
+% \begin{itemize}[leftmargin=*]
+%     \item Necessary background on gaming engines used in CAV simulations.
+%     \item Review of sources of non-determinism in such engines.
+%     \item Provide a method of how the level of non-determinism can be investigated and defined.
+%     \item A case study showing the implementation of this methodology in an engine widely used in CAVs simulations (Unreal Engine).
+% \end{itemize}
+
+% ***************************************************
+%  BACKGROUND
+% ***************************************************
+\section{Background} \label{s:background}
+
+There are numerous game engines with their associated development environments which could be suitable for CAV development, e.g.
+Unreal~\footnote{\url{https://www.unrealengine.com/}}, Unity~\footnote{\url{https://unity.com/}}, CryEngine~\footnote{\url{https://www.cryengine.com/}}.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.
+Carla~\footnote{\url{http://carla.org/}}, AirSim~\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo~\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.
+Nvidia Drive Constellation~\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
+
+Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution.
+%However, when these engines are used for CAV development then deterministic behaviour is required.
+This section gives an overview of a gaming engine and what sources or settings in the engine may affect \textit{simulation variance}.
+%
+% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in CAV simulations.
+Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
+
+% ======= Overview of a Game Loop
+\subsection{Overview of a Game Loop} \label{GameLoopSection}
+The game loop is responsible for the interaction between the physics and rendering engines.
+Fig.~\ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop (initialisation, game logic and decommissioning are removed from this diagram for brevity\footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}).
+A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).~\cite{GameEngineArchBook}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
+\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}}
+\label{GameEngineLoopDiagram}
+\end{figure}
+The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment.
+For example, the throttle and steering inputs from the vehicle controller of an AV.
+The update interprets the intended actions of all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states.
+The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
+% 
+% It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below. 
+% 
+The key aspects of the game loop needed for this discussion are as follows:
+
+\begin{itemize}[leftmargin=*]
+    \item At the beginning of each tick, the lag between the game clock and the real world is updated based on how much real time passed.
+This measures how far the game's clock is behind compared to the real world.
+    \item Then the user inputs are processed.
+    \item There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
+    \item Rendering occurs once the game clock has caught up with the real world clock. 
+    \item The process then repeats.
+\end{itemize}
+
+If the physics engine uses a fixed time step ($dt$), %, because it makes everything simpler and more stable for physics and AI.
+the shorter this time step is the smoother the interpretation of the physical dynamics will be.
+%processing time it takes to catch up to game clock.
+% and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
+%
+To use a fixed physics time step the user's display refresh rate needs to be known in advance.
+This requires an update loop to take less than one render tick (one frame of real world time).
+Given the range of different hardware capabilities a variable delta time is often implemented for game playing taking the previous frame time as the next $dt$.
+However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}.
+Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware.
+Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}}.
+If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes, see lower part of Fig.~\ref{GameEngineLoopDiagram}.
+%\cite{GameEngineArchBook}\cite{GameProgPatternsBook}
+%
+% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
+In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
+and the simulation will exhibit lag between input commands and rendered states resulting in unsynchronised and unrealistic behaviour.
+
+% The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
+If $dt$ is too large or irregularly spaced then the rendered motion may not be smooth and interpolation may be required at additional computational cost.
+
+% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines.
+These engines update at fixed intervals and render whenever they can, which is not steady.
+This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates.
+In this case, the engine will use extrapolation techniques to give an estimate of where the object should be.
+This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
+
+\begin{figure*}[b]
+    \centering
+    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
+    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a gaming engine for usage in CAV simulations.}
+
+    \label{method_diagram}
+\end{figure*}
+
+% ======= Sources of non-determinism
+\subsection{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}
+
+The following points form a hypothesis around the potential sources of non-determinism that could occur in game engines. \\\\
+% \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
+The main reasons that we think are causes of these engines' non-determinism are discussed below.
+\\\\
+
+\noindent\underline{\textit{Floating Point Representation:}}
+Floating point representation of real numbers is limited with fixed register sizes resulting in errors of \textit{rounding} where loss of bits occur which are rounded or truncated and \textit{cancellation} which happens where two similar values are subtracted~\cite{FloatingPointsBook,goldberg1991every}.
+%
+% Translating to the usage of CAV simulations in physics engines,
+This could result in precision issues of, for example, actor position within the environment and falsely satisfying or failing assertions.
+Such reasons could  make one jump to the conclusion that floating point errors could result in non-deterministic behaviour of repeated tests, but these precision issues should be referred to as \textit{incorrectness} and not be confused with determinism.
+Assuming hardware is deterministic and a rounding error occurs, then the same hardware should produce the same \textit{incorrect} output for the same initial conditions.
+So, even if the output is incorrect due to floating point precision it should always produce the same \textit{incorrect} output, given the same logical registers which conform to the same IEEE floating point standard~\cite{8766229}.
+Therefore, it is unlikely that floating point representation will contribute to \textit{simulation variance} for repeated tests on the same hardware.
+%
+% A generic issue with computers are floating points precision. 
+%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
+%Translating to the usage of CAV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.
+%
+% For instance, consider a simulation where the world is infinite and progressively  generated in the physics engine. 
+% After a few hundred meters, precision issues will start to occur, and will get progressively worse the further from the origin the simulation gets. 
+% This is due to the values of the coordinates being used in calculations are getting bigger and hence will not cope with small values that are critical for V\&V testing purposes. 
+% A solution to this problem is perhaps every time the simulations moves 100 meters away from the origin the entire world will move by 100m in the opposite direction.
+Thus, avoiding getting into floating point issues.\cite{FloatingPointArithmeticArticle}
+%
+% There are many scenarios that could be addressed, and solutions to them will entirely depend on the characteristics of that specific scenario.
+There is no generalised solution for the floating problem, regardless of how good computers get there will always be a limitation.
+However, floating points as they stand could just be sufficient for V\&V testing purposes.
+Whenever they are not, smart solutions can be found to accommodate for these limitations.
+\\\\
+\noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
+ECC Memory can detect and correct for single bit errors in DRAM memory.
+Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
+If single bit errors go uncorrected then subsequent computational processing will produce incorrect and non-deterministic results.
+Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit} but CAV verification tasks are likely to incur significant CPU hours and therefore increase exposure to such risks.
+\\\\
+\noindent\underline{\textit{Game Engine Code:}}
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
+%
+For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/index.html}} command can be used to determine processing bottlenecks by indicating the total time spent generating a frame, executing the game loop or render thread.
+\\\\
+\noindent\underline{\textit{Scheduling:}} %
+% info at:
+% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
+% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
+% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
+% https://en.wikipedia.org/wiki/Concurrent_computing
+% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
+Scheduling is a resource management method for sharing computational resources between tasks of different or competing priority where tasks are executed  dependent on the scheduler policy.
+A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
+%
+If during simulation the scheduling policy was updated or thread priority was inconsistently handled then a high variance in output may be possible.
+It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.
+%
+
+Similar to thread scheduling, the CPU scheduling policy determines which processor core to execute processes in the ready queue and this is maybe ordered by factors such as throughput, latency and CPU utilisation.
+If the same single-threaded script executes multiple times across different physical cores due to the CPU utilisation policy then should you expect deterministic behaviour?
+Processor core designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist.
+End users expect such minor variations to be controlled for in architecture tolerances and factory testing.
+
+Scheduling of cores or threads may be a reason for increased \textit{simulation variance} but controlling for this falls outside of the scope of this work and no attempt to control these parameter have been made here.
+However, users should be aware of this potential issue and consider methods to move towards a fixed scheduling sequence if possible.
+\\\\
+\noindent\underline{\textit{Concurrency:}}
+Another issue to consider is concurrency, where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program.
+This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}.
+A game engine may use such techniques to improve performance although monitoring and detecting such events falls outside of the scope for this work.
+\\\\
+\noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
+For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the memory location relative to the processor.
+Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
+The question here is will changes in latency between repeated tests impact the game engine to operate non-deterministically or simply with a increased data transfer cost?
+This may only be answered by determining the scheduling policy of the game engine, but by binding a process to a specific core the variations in data transfer time can be minimised, e.g.
+by using numactl\footnote{url{https://linux.die.net/man/8/numactl}}.
+\\\\
+% these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
+% i.e.
+the utilisation of hardware is manipulated differently each time.
+This would then result in non-deterministic behaviours especially if it is combined with the floating point problem discussed earlier.
+%
+% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
+Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available.
+This may lead to events happening out of order depending on the computational load and hence non-deterministic.
+%
+% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
+To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order.
+This is beyond the scope of our work and will not be covered in this paper.
+\noindent\underline{\textit{Physics and Rendering clocks:}}
+I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.
+% The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
+Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.
+in Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
+This is good for hardware scalability but creates a challenge for the physics calculations which provides the most consistent results with a small fixed time step.
+%
+%
+A variable frame rate could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.
+Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
+Repeated tests on the same system should result in similar frame rates unless there are additional competition for resources.
+%
+CAV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhered to this rate.
+%
+That is to say, if the frame rate is allowed to vary then it has to be controlled to give the same variability through out a given test in order for the test results to be repeatable.
+This issue might be of negligible significance if the frame rate is consistent and the time steps are small enough.
+\\\\
+\noindent\underline{\textit{Actor Navigation:}}
+Navigation meshes are used in game engines for actor path planning, and could be considered as a potential source of non-deterministic behaviour, e.g.%.
+Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic
+~\cite{CARLABenchmark}. 
+This may depend on the algorithms used by the engine which typically use the A* algorithm\cite{AStarBook} and should give deterministic outcomes as long as the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
+Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
+%
+It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps~\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/index.html}}.
+%This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes every time they are loaded.
+
+% ***************************************************
+%  METHOD
+% ***************************************************
+\section{Methodology} \label{s:methodology}
+In this section a method for determining \textit{simulation variance} of an end-user system is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
+The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may tax different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
+To replicate high computational loads that may be experienced in high realism simulations, software that artificially utilises resources is used at runtime for both central and graphical processing units.
+% The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
+The aim of the experiment is to determine if the simulation is deterministic.
+This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories.
+Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.
+
+\subsection{Deterministic Threshold}
+We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($sigma^2$), which is quoted in terms of deviation ($sigma$) for ease of interpretation.
+This threshold is analogous to the analogue level associated with a binary state in a logical processor.
+At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.
+
+The user must specify the level of variance acceptable for their application which is used as a threshold for deterministic execution.
+In our case study, Section~\ref{s:case-study}, a threshold of $1cm$ was chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
+This value is sufficiently small to allow accurate \textit{assertion checking} of autonomous vehicles in urban environments [ref] but a larger value may be acceptable for non-urban environments, e.g.
+motorway or highway.
+Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
+$1\mu m$.
+
+Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
+This is to ensure that an outlier data-point is not hidden in the average of $n$-tests
+
+\subsection{Scenario Description}
+The experiment should be a short scenario that can be easily repeated with the same initial conditions.
+Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
+The maximum deviation in actor path with respect to the mean for $n$-repeated tests will indicate at what dimensional scale deterministic results are achieved.
+For example, if 100 tests are executed and the maximum deviation in actor path is $<1cm$, then this scenario could be interpreted as being deterministic above this scale.
+
+Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.
+Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
+While executing 100 repeats may seem sufficient, in this case the sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
+To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
+
+Game engine actors can be used to simulate the AV and other vehicles and pedestrians.
+For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
+In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
+Initial testing also indicated that when actors collide there is a significant change in \textit{simulation variance}, therefore the experiment should include some scenarios where collision callbacks are activated.
+
+It is also important to control any pseudo-random numbers that are called in any external during the simulation with fixed seeds to ensure repeatable testing, e.g.
+external vehicle control programs or other clients connected to the simulator.
+
+% \todo{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
+
+% \todo{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
+%
+% For the vehicles in the simulation, waypoints are generated and interpolated in the same way but execution at runtime is slightly different where a PID controller is used to generate a velocity profile and steering angle requests to reach the destination targets.}
+
+% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
+Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.
+A* search algorithm.
+
+\subsection{System Resources}
+To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
+Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will  contribute towards increased \textit{simulation variance}.
+CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system.
+This can be used to tune to total load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
+
+% The user should also monitor and record system CPU and GPU utilisation rates.
+Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
+
+% Collision callbacks
+% \subsection{Other Considerations}
+% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
+ii) If not, how does its simulation variance vary?
+In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
+Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
+% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
+Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
+% The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
+For the majority of CAV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
+
+\subsection{Internal settings}
+Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
+If using Carla a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}} and a good description of the interplay between physics and render clock is given here~\cite{JohnAustinUnity}.
+In Carla, further settings can be applied to allow synchronous communication to external controllers to ensure no sensor data are passed out of order to the simulator~\cite{carla_sim_config}.
+
+% There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine.
+% Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
+However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
+% If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the simulation variance as a result of allowing the AI to navigate in a more well defined space.
+% Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
+% The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.
+% Running in editor or deployed mode can sometimes cause differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
+
+% \begin{table*}[b]
+% \centering
+% \begin{tabular}{clclcc}
+% \toprule
+% Test ID & Test description (See Figures~\ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
+% 1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
+% 2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
+% 3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
+% 4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
+% 5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
+% 6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
+% 7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
+% 8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
+% \bottomrule
+% \end{tabular}
+% \caption{Set of experiments}
+% \label{TableOfExperiments}
+% \end{table*}
+
+\begin{table*}[b]
+\centering
+\begin{tabular}{clclccc}
+\toprule
+Test 		& Actors 						 & Collision & Collision Type 			 & $d_l$ (m) & $n$  &$\max\sigma$ (m) \\ \midrule
+1       & Two vehicles                   & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
+2       & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 2 		 & 1000 & 1000 \\
+3       & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
+4       & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 2 		 & 1000 & 1000 \\
+5       & Two pedestrians                & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
+6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 0.4 		 & 1000 & 1000 \\
+7       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 2 		 & 1000 & 1000 \\
+8       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 20 		 & 1000 & 1000 \\
+\bottomrule
+\end{tabular}
+\caption{Set of experiments where $d_l$ is the \textit{look ahead distance}, $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.}
+\label{TableOfExperiments}
+\end{table*}
+
+% \subsection{External settings}
+% Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.
+Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
+Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.
+
+% \subsection{Running experiment}
+% \noindent When it comes to running the experiment itself, things that one should consider are the problem space exploration; the number of runs that should be executed to get reliable results; the frequency of logging data of actors; and possibly having a program to monitor the hardware utilisation.
+% Problem exploration, there is too many variables that one can play with. do statistical analysis some.... some of them get problematic...
+
+\subsection{Output and Analysis}
+The output for each experiment can be summarised as the maximum variance of any actor over the repeated tests.
+This value can be analysed against the different scenarios or settings that have been identified as potential sources of simulation variance.
+These values can then be assessed against the user-defined \textit{deterministic threshold} to determine if the simulation is deterministic or not.
+
+% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
+Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
+
+% This whole process is repeated for various internal and external settings (if needed). 
+% Then various plots can be created to draw the engine's simulation variance. 
+% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
+
+% ***************************************************
+%  CASE STUDY
+% ***************************************************
+\section{Case Study} \label{s:case-study}
+We present a case study to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction}.
+The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance $\sigma^2$ for any actor should be less than a user-defined threshold.
+Results are shown in terms of the maximum deviation ($\sigma$) and the deviation threshold in our case is $1cm$, where any value higher than this is considered non-deterministic.
+
+In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation.
+From previous work it was found that additional loads on either the CPU or GPU could increase simulation variance by equivalent levels\footnote{see footnote 14}.
+The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
+
+Tests 5\&6 involve only pedestrians where they do not collide (test 5) and where they do (test 6).
+Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
+In test 7 $d_l$ is creased from $0.4m$ to $2.0m$ and in test 8 to $20m$.
+
+Abanoub: can you change the diagram to be a staggered t-junction as the DMR is just confusing for this example.
+Also can we zoom in on the lower right area.
+Tests 5 and 6 have different look ahead values, which is correct?
+
+% Given that our interest is on CAV simulations, this study presents a number of experiments run on typical agents used in CAV simulations and their interactions with each other.
+These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
+Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
+We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
+
+% \subsection{Problem And Hypothesis}
+% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+
+% Our hypothesis is that since UE4 is a gaming engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
+This behaviour is expected not to hold as the level of computing stress increases.
+
+% \subsection{Tests Description And Technicalities}\label{TestsDescriptionAndTechnicalities}
+\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
+The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover a range of interactions between actor types.
+Test 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at 0.1m intervals.
+Vehicle actors were programmed to use a PID controller to follow the waypoints targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints that are within $d_l$ tend to cause improper navigation with the implemented PID controller}.
+
+In test 3 two vehicles dive without collision and a pedestrian walks across the road at a crossing point.
+In test 4 this pedestrian collides with one of the vehicles at the crossing, see Fig~\ref{Test_a}.
+Pedestrians navigate to a set of regularly space waypoints at 0.1m intervals using the A* algorithm which are the default settings in the Caral/Unreal distribution.
+
+% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
+Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
+
+Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
+
+The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents. 
+These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
+
+\begin{figure}[h]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagram_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagram_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Shows the setup of the different tests in Table \ref{TableOfExperiments} (a)Test IDs 1 to 4 (b)Test IDs 5 to 8.}
+\end{figure}
+
+Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
+
+Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+
+Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
+Pedestrians have the functionality of dynamic path planning i.e. they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other. 
+Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
+
+In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+
+Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+
+Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
+
+On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
+
+In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
+
+% ***************************************************
+%  RESULTS
+% ***************************************************
+\subsection{Results And Discussion}\label{ResultsSection}
+We start by analysing (step 5) the summary from all of the tests plotted in Figure~\ref{ExperimentsStressSummary}; where the $x$-axis represents the percentage of CPU + GPU utilisation and the $y$-axis shows the total mean devaition.
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummary.pdf}
+    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.}
+
+    \label{ExperimentsStressSummary}
+\end{figure}
+
+\begin{figure*}[h]
+    \centering
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25.pdf}
+        \caption{}
+        \label{CarsCollisionCG25}
+    \end{subfigure}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePost.pdf}
+        \caption{}
+        \label{CarsCollisionPrePost}
+    \end{subfigure}
+    \caption{Test ID 2 (a)Mean deviation over runs no. vs simulation time for 25\% stress. (b)Pre and post collision mean deviation over runs no. for different CPU + GPU stresses.}
+\end{figure*}
+
+\begin{figure*}[h]
+    \centering
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25.pdf}
+        \caption{}
+        \label{CarsPeopleCollsionsCG25}
+    \end{subfigure}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePost.pdf}
+        \caption{}
+        \label{CarsPeopleCollisionPrePost}
+    \end{subfigure}
+    \caption{Test ID 4 (a)Mean deviation over runs no. vs simulation time for 25\% stress. (b)Pre and post collision mean deviation over runs no. for different CPU + GPU stresses.}
+\end{figure*}
+This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
+
+Note the dotted red line in the plot shows the millimeter (mm) level. 
+Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
+
+First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
+Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
+with in the centimeter level).
+This clearly shows how the load on a computer (i.e.
+stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
+This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
+
+Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
+with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
+
+Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively. 
+We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
+
+Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
+Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
+
+Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
+However, in V\&V of CAV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
+Thus, the post collision effect should not be of much concern in the context of CAVs.
+It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
+The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
+Thus failing to stay below our red line and the test should be terminated.
+In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
+
+% ***************************************************
+%  CONCLUSION
+% ***************************************************
+\section{Conclusions}\label{s:conclusion}
+We have provided a summary of the sources of non-determinisim in game engines.
+This was followed by presenting a method by which one can determine the simulation variance of a gaming engine used in V\&V application of CAV simulations.
+The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
+
+A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
+It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
+%since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
+
+In highlight of obtaining a fully deterministic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
+Alternatively, another approach would be to create a new engine that is designed to be deterministic with minimal attention given to performance.
+
+% ***************************************************
+%  FUNDING, REFS
+% ***************************************************
+\section*{Acknowledgements}
+This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
+projects are part-funded by the Centre for Connected and Autonomous
+Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
+103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
+\balance
+\printbibliography
 

### 2020-03-06 paper/6_Conclusions.tex

@@
 \section{Conclusions}\label{s:conclusion}
+
+\todo{KIE: Mention our ResiCAV experiments and the importance to recognize that once physics is involved then a single experiment is not sufficient to determine the outcome / impact of a test.
+Instead, at this stage several thousands of tests may need to be executed to obtain reliable statistics to estimate the worst/best case.}
+
 We have provided a summary of the sources of non-determinisim in game engines.

### 2020-03-09 paper/cav_determinisim_main.tex
GC changes to section 1-4
GC put all into single file

@@
 \input{main.config.tex}
-\pagenumbering{gobble} %suppresses page numbers
+% \pagenumbering{gobble} %suppresses page numbers
 \addbibresource{IROS.bib}
@@
 CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system.
-This can be used to tune to total load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
+This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
+For GPU utilisation \texttt{gpu-burn} was employed using the fur test at different resolutions and multiple instances can be used to generate high utilisation levels~\cite{GPU_stress}.
 
@@
 In Carla, further settings can be applied to allow synchronous communication to external controllers to ensure no sensor data are passed out of order to the simulator~\cite{carla_sim_config}.
+
+Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
+Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.
+\textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.
+\textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
+Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/index.html}}.
+However, none of these options improved simulation variance significantly beyond the default settings, see our previous investigation~\cite{TSLUnrealEngineTesting}.
+
+% By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronous mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
 
@@
 5       & Two pedestrians                & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
-6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 0.4 		 & 1000 & 1000 \\
-7       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 2 		 & 1000 & 1000 \\
-8       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 20 		 & 1000 & 1000 \\
+6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 2 		 & 1000 & 1000 \\
+% 6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 0.4 		 & 1000 & 1000 \\
+% 8       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 20 		 & 1000 & 1000 \\
 \bottomrule
 \end{tabular}
-\caption{Set of experiments where $d_l$ is the \textit{look ahead distance}, $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.}
+\caption{Test scenarios where $d_l$ is the \textit{look ahead distance}, $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
+CONSIDER removing entries 6 and 8 if they do not contribute towards outcomes and remove references to $d_l$.
+\textbf{remove all reference to $d_l$}}
 \label{TableOfExperiments}
@@
 In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation.
-From previous work it was found that additional loads on either the CPU or GPU could increase simulation variance by equivalent levels\footnote{see footnote 14}.
-The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
-
-Tests 5\&6 involve only pedestrians where they do not collide (test 5) and where they do (test 6).
-Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is creased from $0.4m$ to $2.0m$ and in test 8 to $20m$.
+Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
+Resource utilisation figures reported here should be considered approximate values.
+% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
 
@@
 Tests 5 and 6 have different look ahead values, which is correct?
+
+These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
+To ensure reliability of results, each test was repeated 1000 times.
 
@@
 The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover a range of interactions between actor types.
-Test 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at 0.1m intervals.
-Vehicle actors were programmed to use a PID controller to follow the waypoints targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints that are within $d_l$ tend to cause improper navigation with the implemented PID controller}.
+Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1m$ intervals using a PID controller.
+In test 3 a mixture of different actor types are introduced where two vehicles dive without collision and a pedestrian walks across the road at a crossing point.
+In test 4 this pedestrian collides with one of the vehicles at the crossing triggering a collision callback, see Fig~\ref{Test_a}.
+Similar to vehicles, pedestrians navigate via a set of regularly space waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
+Tests 5 \& 6 involve only pedestrians where they do not collide (test 5) and where they do (test 6), see Fig.~\ref{Test_b}.
 
-In test 3 two vehicles dive without collision and a pedestrian walks across the road at a crossing point.
-In test 4 this pedestrian collides with one of the vehicles at the crossing, see Fig~\ref{Test_a}.
-Pedestrians navigate to a set of regularly space waypoints at 0.1m intervals using the A* algorithm which are the default settings in the Caral/Unreal distribution.
+ % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
+To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
+(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
+
+% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
+In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
 
@@
 
-Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
+% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
 
-The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents. 
+% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
 These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
@@
     \end{subfigure}
-    \caption{Shows the setup of the different tests in Table \ref{TableOfExperiments} (a)Test IDs 1 to 4 (b)Test IDs 5 to 8.}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
 \end{figure}
 
-Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
 The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
 
-Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
 
-Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-Pedestrians have the functionality of dynamic path planning i.e. they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other. 
+% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
 Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
 
-In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+% In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
 
-Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
 
-Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
 By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
@@
 
-On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
+% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
 Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
 
-In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
+% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
 
@@
 \subsection{Results And Discussion}\label{ResultsSection}
-We start by analysing (step 5) the summary from all of the tests plotted in Figure~\ref{ExperimentsStressSummary}; where the $x$-axis represents the percentage of CPU + GPU utilisation and the $y$-axis shows the total mean devaition.
+The summary of all the tests are shown in the final column of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels.
+From these summary results it is clear that scenarios with only pedestrian actors not involved in collisions are those that are consistently deterministic over all resource utilisation levels \textbf{(is this true, any other cases?)}.
+However, resource utlisation level has a significant impact on \textit{simulation variance}.
+
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU) and the $y$-axis shows the maximum deviation ($\max\sigma$).
+In this figure, anything above the $cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level based on our verification requirements.
+A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.
+
+% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
+Note the dotted red line in the plot shows the millimeter (mm) level.
+Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
+
 \begin{figure}[!t]
@@
     \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummary.pdf}
-    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.}
+    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.\textbf{ Change to "CPU \& GPU Resource Utilisation" and "y axis to $\max\sigma$ (m)" Change to $cm$ level in the graphs.
+Cannot see Test 6 in the results.
+Remove test 6 and 8 and rename 7 to test 6.
+Move 0 stress level data to 5\% to reflect simulation resources required.}}
 
@@
     \end{subfigure}
-    \caption{Test ID 2 (a)Mean deviation over runs no. vs simulation time for 25\% stress. (b)Pre and post collision mean deviation over runs no. for different CPU + GPU stresses.}
+    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.
+\textbf{Change Y axis value to $\max\sigma (m)$}}
 \end{figure*}
@@
     \end{subfigure}
-    \caption{Test ID 4 (a)Mean deviation over runs no. vs simulation time for 25\% stress. (b)Pre and post collision mean deviation over runs no. for different CPU + GPU stresses.}
+    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation.
+\textbf{Change Y axis value to $\max\sigma (m)$}}
 \end{figure*}
-This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
-
-Note the dotted red line in the plot shows the millimeter (mm) level. 
-Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
 
@@
 \end{document}
-
-% \begin{table*}
-% \centering
-% \caption{Feature importance given in rank order for prediction of elbow position ($\Phi$) for univariate and multivariate response.
-$\Sigma(NI)$ indicates sum of normalized importance for the top 5 features.}
-% \label{elbow_posn}
-% \begin{tabular}{|c||c|c|c||c|}
-% \hline
-% Rank & $\Phi(x)$ & $\Phi(y)$ & $\Phi(z)$ & $\Phi(xyz)$ \\
-% \hline
-% 1&left shoulder ($13P_x$) &left shoulder ($13P_y$)&left shoulder ($13P_z$) &right hip ($20P_x$)\\
-% 2&right hip ($20P_x$)     &right hip ($20P_x$)    &head ($7P_z$)           &left shoulder ($13P_z$)\\
-% 3&head ($7P_x$)           &right elbow ($10P_y$)  &right hand ($11P_z$)    &left shoulder ($13P_x$)\\
-% 4&right hand ($11P_x$)    &right hand ($11P_x$)   &head ($7P_x$)		   &right hand ($11P_x$)\\
-% 5&left shoulder ($13P_y$) &right elbow ($10P_x$)  &neck ($6P_z$)		   &head ($7P_x$)\\
-% \hline 
-% $\Sigma(NI)$ & 0.683      &0.684                  &0.723 					&0.570 \\
-% \hline
-% %RMSE-test ($m$) & 0.011& 0.017& 0.007& 0.024\\
-% %RMSE-unseen ($m$) & 0.086& 0.060& 0.038& 0.063\\
-% %\hline
-% \end{tabular}
-% \end{table*}

### 2020-03-09 paper/cav_determinisim_main.tex
GC put all into single file

@@
 From these summary results it is clear that scenarios with only pedestrian actors not involved in collisions are those that are consistently deterministic over all resource utilisation levels \textbf{(is this true, any other cases?)}.
-However, resource utlisation level has a significant impact on \textit{simulation variance}.
+However, resource utilisation level has a significant impact on \textit{simulation variance}.
 
@@
 A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.
+
+Scenarios that include both collisions and vehicle actors were not deterministic.
+However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
+This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any results after this point are not reliable.
 

### 2020-03-11 paper/cav_determinisim_main.tex
GC changes to results

@@
 From these summary results it is clear that scenarios with only pedestrian actors not involved in collisions are those that are consistently deterministic over all resource utilisation levels \textbf{(is this true, any other cases?)}.
-However, resource utilisation level has a significant impact on \textit{simulation variance}.
+However, for other scenarios involving vehicles or mixed actors resource utilisation level has a significant impact on \textit{simulation variance}.
 
 Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU) and the $y$-axis shows the maximum deviation ($\max\sigma$).
-In this figure, anything above the $cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level based on our verification requirements.
+In this figure, anything above the $cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
 A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.
+
+However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are an overly pessimistic account of the situation.
+Examining these specific results as a function of simulation time reveals further information about the simulation variance before and after an actor collision callback.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision and is non-deterministic afterwards.
+This pattern of deterministic behaviour prior to collision is maintained up to resource utilisation levels of approximately 50\%, Fig.~\ref{CarsCollisionPrePost}.
+
+This time series examination is repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+\textbf{What is the summary data for this scenario?} Similarly to vehicle collisions, the variation of $\max\sigma$ for pedestrian collisions with resource utilisation indicates deterministic behaviour below 95\% resource utilisation.
+
+\textbf{Strange gap post-collision for few seconds}
+
+\textbf{Should we plot a pre-collision only summary chart for all scenarios?}
+
+\textbf{Should we rename 'test' to 'scenario???'}
 
@@
 \section{Conclusions}\label{s:conclusion}
+
+**** These results are not meant as a universal guide but merely to illustrate our method for a single hardware configuration.
+It is probable that the values quoted here will not translate to individual end-users unless the system configuration is very similar.
+This paper should be used as a method for users to determine what scenarios and system setting are appropriate for their particular verification needs.
+
 We have provided a summary of the sources of non-determinisim in game engines.

### 2020-03-11 paper/cav_determinisim_main.tex
GC conclusion

@@
 The question here is will changes in latency between repeated tests impact the game engine to operate non-deterministically or simply with a increased data transfer cost?
-This may only be answered by determining the scheduling policy of the game engine, but by binding a process to a specific core the variations in data transfer time can be minimised, e.g.
-by using numactl\footnote{url{https://linux.die.net/man/8/numactl}}.
+This may only be answered by determining the scheduling policy of the game engine, but by binding a process to a specific core the variations in data transfer time can be minimised.
 \\\\
@@
 \subsection{Deterministic Threshold}
-We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($sigma^2$), which is quoted in terms of deviation ($sigma$) for ease of interpretation.
-This threshold is analogous to the analogue level associated with a binary state in a logical processor.
+We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
+This threshold is analogous to the analogue signal level associated with a binary state in a logical processor.
 At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.
@@
 
-Abanoub: can you change the diagram to be a staggered t-junction as the DMR is just confusing for this example.
+\textbf{Abanoub: can you change the diagram to be a staggered t-junction as the DMR is just confusing for this example.
 Also can we zoom in on the lower right area.
-Tests 5 and 6 have different look ahead values, which is correct?
+Tests 5 and 6 have different look ahead values, which is correct?}
 
+\subsection{System Configuration and Screening}
 These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
 To ensure reliability of results, each test was repeated 1000 times.
+Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$.
+To ensure our experiment did not fall foul of some of the possible sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
+These were:
+
+\begin{itemize}[leftmargin=*]
+    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
+    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
+    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} \textbf{we need to run these test to look for $\max\sigma$ variations}
+\end{itemize}
 
@@
 
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU) and the $y$-axis shows the maximum deviation ($\max\sigma$).
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU).
 In this figure, anything above the $cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
@@
 
-However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are an overly pessimistic account of the situation.
-Examining these specific results as a function of simulation time reveals further information about the simulation variance before and after an actor collision callback.
+However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are an overly pessimistic/conservative account of the situation.
+Examining these specific results (test 2, 4 \& 6??) as a function of simulation time reveals further information about the simulation variance before and after an actor collision callback.
 Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision and is non-deterministic afterwards.
 This pattern of deterministic behaviour prior to collision is maintained up to resource utilisation levels of approximately 50\%, Fig.~\ref{CarsCollisionPrePost}.
+This time series examination is repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+\textbf{What is the summary data for this scenario?} Similarly to vehicle collisions, the variation of $\max\sigma$ for pedestrian collisions indicates deterministic behaviour below 95\% resource utilisation, Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any results after this point are not reliable.
 
-This time series examination is repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-\textbf{What is the summary data for this scenario?} Similarly to vehicle collisions, the variation of $\max\sigma$ for pedestrian collisions with resource utilisation indicates deterministic behaviour below 95\% resource utilisation.
+The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
+In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is not involved in the collision.
+There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as describe above.
+Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour even after the collision \textbf{is this because the walker is killed and removed from the game?}.
+And, thirdly we observed a delayed effect on Car ID 1 showing high simulation variance post-collision but with a $5s$ delayed effect \textit{even though this vehicle was not involved in the collision}.
+This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors will affect the simulation variance of all actors in the simulation and could potentially result in false-positive assertion violations.
 
@@
 
-Scenarios that include both collisions and vehicle actors were not deterministic.
+\textbf{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
+
+% Scenarios that include both collisions and vehicle actors were not deterministic.
 However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
-This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any results after this point are not reliable.
 
@@
 
-First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
+% First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
 Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
@@
 
-Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
+% Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
 with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
 
-Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
+% Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
 We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
 
-Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
+% Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
 Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
 
-Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
+% Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
 However, in V\&V of CAV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
 Thus, the post collision effect should not be of much concern in the context of CAVs.
-It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
+
+% It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
 The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
@@
 \section{Conclusions}\label{s:conclusion}
+A detailed description of how practitioners in this field can resolve the non-deterministic operational domains of their simulation is presented.
+Potential sources of non-deterministic simulation behaviour are discussed and some of these are resolved, such as floating point error, and other are left unanswered such as scheduling policy.
+%
+A case study was used to ascertain the boundaries of non-deterministic operation of the Carla driving simulator which seem to exist at collision events and when resource utilisation levels exceeded 50\% for the particular system used in this study.
+A collision is not necessarily a problem as long as the simulation is stopped as soon as it is detected as even actors not involved directly will behave non-deterministically.
 
-**** These results are not meant as a universal guide but merely to illustrate our method for a single hardware configuration.
+% We have provided a summary of the sources of non-determinisim in game engines.
+This was followed by presenting a method by which one can determine the simulation variance of a gaming engine used in V\&V application of CAV simulations.
+The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
+
+% A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
+It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
+%since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
+
+% **** These results are not meant as a universal guide but merely to illustrate our method for a single hardware configuration.
 It is probable that the values quoted here will not translate to individual end-users unless the system configuration is very similar.
@@
 
-We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the simulation variance of a gaming engine used in V\&V application of CAV simulations.
-The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
+% You must monitor all collisions, not just for the ego vehicle as any collision callback may result in high simulation variance and hence non-deterministic results for your experiments.
 
-A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
-It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
-%since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
-
-In highlight of obtaining a fully deterministic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
+% In highlight of obtaining a fully deterministic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
 Alternatively, another approach would be to create a new engine that is designed to be deterministic with minimal attention given to performance.

### 2020-03-11 Results_matlab_code/CarlaReadLog_v1.m
Added matlab code

@@ -1,0 +1,505 @@
+% close all
+clear all; clc
+
+%% Set the folder
+cd '/home/is18902/git/Robopilot_Carla/PythonAPI/examples/ExperimentResutls'
+
+%% or add results directory to path
+addpath('~/git/Robopilot_Carla/PythonAPI/examples/ExperimentResutls')
+
+% set graph display as option
+display_graphs = 1;
+display_graphs_2 = 0;
+plot_velocity_graph = 0;
+
+%% setup for multiple reads
+nFilep1='Experiment-'; nFilep2='.csv';
+nVelocities = 1;
+fileNumberOffset = 0; %for file ...-5001.csv use = 4, 6001 = 5
+
+file = 'TEST_Cars_OUTPUT.txt';
+data = importfile_data(file);
+
+%% Find the number of exclusive agents & tests
+agentIDs = unique(data.agentID,'stable');
+agentIDs = table2array(agentIDs);
+Agents = unique(data.agentNo,'stable');
+nAgents = length(Agents);
+nRepeats = max(data.repeatNo);
+maxTime = max(data.time);
+timeStep = 0.1;
+
+%% Store raw, non-interpolated data (T,X,Y) and variance
+rawData = zeros(nRepeats,nAgents,3,round(maxTime/timeStep));
+variance = zeros(nRepeats,nAgents,2);
+avgVar = zeros(nRepeats,1);
+
+%% For each agent get (T,X,Y) and interpolate, then variance
+for i=1:nRepeats
+    for j=1:nAgents
+        sel1 = data.repeatNo==i; %select data for each repeat
+        sel2 = data.agentNo==j; %select data for each agent
+        sel = sel1 & sel2;
+        tempData = data(sel,:);
+        tempT = tempData.time;
+        tempX = tempData.x;
+        tempY = tempData.y;
+        rawData(i,j,:,:) = [tempT, tempX, tempY]';
+    end
+end
+
+%% Interpolate data if you need to?
+% % interpolate data
+% regT = min(tempT):0.1:max(tempT);
+% regX=interp1(tempT,tempX,regT,'linear','extrap');
+% regY=interp1(tempT,tempY,regT,'linear','extrap');
+
+%% Get variance per experiment and agent
+for i=1:nRepeats
+    agentVar = zeros(nAgents,1);
+    for j=1:nAgents
+        rawX = squeeze(rawData(:,j,2,:)); % x data
[…]

### 2020-03-11 Results_matlab_code/CarlaReadLog_v1_AG.m

@@ -1,0 +1,321 @@
+% close all
+clear all; close all; clc
+
+%% Set the folder
+% cd '/home/is18902/git/Robopilot_Carla/PythonAPI/examples/ExperimentResutls'
+
+%% or add results directory to path
+% addpath('~/git/Robopilot_Carla/PythonAPI/examples/ExperimentResutls')
+
+% set graph display as option
+display_graphs = 0;
+display_graphs_2 = 0;
+plot_velocity_graph = 0;
+
+%% setup for multiple reads
+nFilep1='Experiment-'; nFilep2='.csv';
+nVelocities = 1;
+fileNumberOffset = 0; %for file ...-5001.csv use = 4, 6001 = 5
+
+% file = "TEST_CarsPeopleCollision_CG25.txt";
+% file = "TEST_CarsCollision_CG25.txt";
+
+% file = "TEST_Cars_CG0.txt";
+% file = "TEST_Cars_CG25.txt";
+% file = "TEST_Cars_CG50.txt";
+% file = "TEST_Cars_CG75.txt";
+% file = "TEST_Cars_CG95.txt";
+% 
+% file = "TEST_CarsCollision_CG0.txt";
+file = "TEST_CarsCollision_CG25.txt";
+% file = "TEST_CarsCollision_CG50.txt";
+% file = "TEST_CarsCollision_CG75.txt";
+% file = "TEST_CarsCollision_CG95.txt";
+% 
+% file = "TEST_CarsPeople_CG0.txt";
+% file = "TEST_CarsPeople_CG25.txt";
+% file = "TEST_CarsPeople_CG50.txt";
+% file = "TEST_CarsPeople_CG75.txt";
+% file = "TEST_CarsPeople_CG95.txt";
+% 
+% file = "TEST_CarsPeopleCollision_CG0.txt";
+% file = "TEST_CarsPeopleCollision_CG25.txt";
+% file = "TEST_CarsPeopleCollision_CG50.txt";
+% file = "TEST_CarsPeopleCollision_CG75.txt";
+% file = "TEST_CarsPeopleCollision_CG95.txt";
+% 
+% file = "TEST_People_CG0.txt";
+% file = "TEST_People_CG25.txt";
+% file = "TEST_People_CG50.txt";
+% file = "TEST_People_CG75.txt";
+% file = "TEST_People_CG95.txt";
+% 
+% file = "TEST_PeopleCollsion_L4_CG0.txt";
[…]

### 2020-03-11 Results_matlab_code/Copy_of_CarlaReadLog_v1_AG.m

@@ -1,0 +1,311 @@
+% close all
+clear all; close all; clc
+
+%% Set the folder
+% cd '/home/is18902/git/Robopilot_Carla/PythonAPI/examples/ExperimentResutls'
+
+%% or add results directory to path
+% addpath('~/git/Robopilot_Carla/PythonAPI/examples/ExperimentResutls')
+
+% set graph display as option
+display_graphs = 0;
+display_graphs_2 = 0;
+plot_velocity_graph = 0;
+
+%% setup for multiple reads
+nFilep1='Experiment-'; nFilep2='.csv';
+nVelocities = 1;
+fileNumberOffset = 0; %for file ...-5001.csv use = 4, 6001 = 5
+
+% file = "TEST_CarsPeopleCollision_CG25.txt";
+% file = "TEST_CarsCollision_CG25.txt";
+
+% file = "TEST_Cars_CG0.txt";
+% file = "TEST_Cars_CG25.txt";
+% file = "TEST_Cars_CG50.txt";
+% file = "TEST_Cars_CG75.txt";
+% file = "TEST_Cars_CG95.txt";
+% 
+% file = "TEST_CarsCollision_CG0.txt";
+file = "TEST_CarsCollision_CG25.txt";
+% file = "TEST_CarsCollision_CG50.txt";
+% file = "TEST_CarsCollision_CG75.txt";
+% file = "TEST_CarsCollision_CG95.txt";
+% 
+% file = "TEST_CarsPeople_CG0.txt";
+% file = "TEST_CarsPeople_CG25.txt";
+% file = "TEST_CarsPeople_CG50.txt";
+% file = "TEST_CarsPeople_CG75.txt";
+% file = "TEST_CarsPeople_CG95.txt";
+% 
+% file = "TEST_CarsPeopleCollision_CG0.txt";
+% file = "TEST_CarsPeopleCollision_CG25.txt";
+% file = "TEST_CarsPeopleCollision_CG50.txt";
+% file = "TEST_CarsPeopleCollision_CG75.txt";
+% file = "TEST_CarsPeopleCollision_CG95.txt";
+% 
+% file = "TEST_People_CG0.txt";
+% file = "TEST_People_CG25.txt";
+% file = "TEST_People_CG50.txt";
+% file = "TEST_People_CG75.txt";
+% file = "TEST_People_CG95.txt";
+% 
+% file = "TEST_PeopleCollsion_L4_CG0.txt";
[…]

### 2020-03-11 Results_matlab_code/Results/Determinisim_Results.m

@@ -1,0 +1,128 @@
+close all
+clear all 
+clc
+
+CarsCollisionPrePostData = readmatrix("CarsCollisionPrePostData.csv");
+
+CarsPeopleCollisionPrePostData = readmatrix("CarsPeopleCollisionPrePostData.csv");
+
+ExperimentsStressSummaryData = readmatrix("ExperimentsStressSummary.csv");
+
+FontSize = 12;
+
+%% Plot for paper (CarsCollisionPrePostData)
+figure(1)
+StressArray                        = CarsCollisionPrePostData(1,2:end);
+AveragePreCarsCollisionDevaition   = CarsCollisionPrePostData(2,2:end);
+AveragePostCarsCollisionDevaition  = CarsCollisionPrePostData(3,2:end);
+
+plot(StressArray,AveragePreCarsCollisionDevaition,'ks','markerfacecolor','k','LineWidth',1,'DisplayName',"Pre Collision"); hold on;
+
+plot(StressArray,AveragePostCarsCollisionDevaition,'ko','markerfacecolor','w','LineWidth',1,'DisplayName',"Post Collision"); hold on;
+
+annotation('doublearrow',[0.13 0.9],[0.518 0.518],"Head1Style","none","Head2Style","none","Color","r","LineStyle","--","LineWidth",2)
+text(1,8*10^-4,"mm level",'FontSize',FontSize,"Color","r")
+
+yticks([10^-6 10^-3 1])
+yticklabels({'10^{-6}','10^{-3}','10^{0}'})
+xlabel("CPU + GPU Stress (%)")
+ylabel("Mean Deviation Over Runs No. (m)")
+set(gca,'FontSize',FontSize)
+set(gca, 'YScale', 'log')
+legend("Location","SouthEast")
+
+
+%% Plot for paper (CarsPeopleCollisionPrePostData)
+figure(2)
+StressArray                              = CarsPeopleCollisionPrePostData(1,2:end);
+AveragePreCarsPeopleCollisionDevaition   = CarsPeopleCollisionPrePostData(2,2:end);
+AveragePostCarsPeopleCollisionDevaition  = CarsPeopleCollisionPrePostData(3,2:end);
+AveragePreCarsPeopleCollisionDevaition(1:4) = 6*10^-15; % Values are very small they are being read as zeros
+
[…]

### 2020-03-11 Results_matlab_code/Results/Determinisim_ResultsV2.m

@@ -1,0 +1,135 @@
+close all
+clear all 
+clc
+
+CarsCollisionPrePostData = readmatrix("CarsCollisionPrePostData_max.csv");
+
+CarsPeopleCollisionPrePostData = readmatrix("CarPeopleCollisionPrePostData_max.csv");
+
+ExperimentsStressSummaryData = readmatrix("ExperimentsStressSummary_max.csv");
+
+FontSize = 12;
+
+%% Plot for paper (CarsCollisionPrePostData)
+figure(1)
+StressArray                        = CarsCollisionPrePostData(1,2:end);
+AveragePreCarsCollisionDevaition   = CarsCollisionPrePostData(2,2:end);
+AveragePostCarsCollisionDevaition  = CarsCollisionPrePostData(3,2:end);
+
+plot(StressArray,AveragePreCarsCollisionDevaition,'ks','markerfacecolor','k','LineWidth',1,'DisplayName',"Pre Collision"); hold on;
+
+plot(StressArray,AveragePostCarsCollisionDevaition,'ko','markerfacecolor','w','LineWidth',1,'DisplayName',"Post Collision"); hold on;
+
+annotation('doublearrow',[0.14 0.9],[0.72 0.72],"Head1Style","none","Head2Style","none","Color","r","LineStyle","--","LineWidth",2)
+text(1,5*10^-4,"mm level",'FontSize',FontSize,"Color","r")
+
+axis([0 100 10^(-15) 10^(1)])
+yticks([10^(-14) 10^(-12) 10^(-10) 10^(-8) 10^(-6) 10^(-4) 10^(-2) 10^(0)])
+
+xlabel("CPU & GPU Resource Utilisation (%)")
+ylabel("max \sigma (m)")
+set(gca,'FontSize',FontSize)
+set(gca, 'YScale', 'log')
+% legend("Location","SouthEast")
+legend("Position",[0.79 0.3 0 0])
+
+%% Plot for paper (CarsPeopleCollisionPrePostData)
+figure(2)
+StressArray                              = CarsPeopleCollisionPrePostData(1,2:end);
+PreCarsPeopleCollisionDevaition   = CarsPeopleCollisionPrePostData(2,2:end);
+PostCarsPeopleCollisionDevaition  = CarsPeopleCollisionPrePostData(3,2:end);
[…]

### 2020-03-11 Results_matlab_code/importfile_data.m

@@ -1,0 +1,46 @@
+function data = importfile(filename, dataLines)
+%IMPORTFILE Import data from a text file
+%  TESTCARSCOLLISIONOUTPUT = IMPORTFILE(FILENAME) reads data from text
+%  file FILENAME for the default selection.  Returns the data as a table.
+%
+%  TESTCARSCOLLISIONOUTPUT = IMPORTFILE(FILE, DATALINES) reads data for
+%  the specified row interval(s) of text file FILENAME. Specify
+%  DATALINES as a positive scalar integer or a N-by-2 array of positive
+%  scalar integers for dis-contiguous row intervals.
+%
+%  Example:
+%  TESTCarsCollisionOUTPUT = importfile("/home/is18902/git/Robopilot_Carla/PythonAPI/examples/ExperimentResutls/TEST_CarsCollision_OUTPUT.txt", [2, Inf]);
+%
+%  See also READTABLE.
+%
+% Auto-generated by MATLAB on 15-Oct-2019 15:39:24
+
+%% Input handling
+
+% If dataLines is not specified, define defaults
+if nargin < 2
+    dataLines = [2, Inf];
+end
+
+%% Setup the Import Options and import the data
+opts = delimitedTextImportOptions("NumVariables", 11);
+
+% Specify range and delimiter
+opts.DataLines = dataLines;
+opts.Delimiter = ",";
+
+% Specify column names and types
+opts.VariableNames = ["repeatNo", "agentNo", "agentID", "agentType", "agentTypeNo", "time", "fps", "x", "y", "z", "yaw"];
+opts.VariableTypes = ["double", "double", "double", "categorical", "double", "double", "double", "double", "double", "double", "double"];
+
+% Specify file level properties
+opts.ExtraColumnsRule = "ignore";
+opts.EmptyLineRule = "read";
+
+% Specify variable properties
+opts = setvaropts(opts, "agentType", "EmptyFieldRule", "auto");
+
+% Import the data
+data = readtable(filename, opts);
+
+end

### 2020-03-12 paper/cav_determinisim_main.tex
Added results matlab code
Updated graphs to show man instead of mean.
Modified figures to show cm level instead of mm
GC checking max sigma values

@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagram_a.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
         \caption{}
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagram_b.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
         \caption{}
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummary.pdf}
-    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.\textbf{ Change to "CPU \& GPU Resource Utilisation" and "y axis to $\max\sigma$ (m)" Change to $cm$ level in the graphs.
-Cannot see Test 6 in the results.
-Remove test 6 and 8 and rename 7 to test 6.
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV2.pdf}
+    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
+Tests 5 and 6 overlap having almost identical results.
+\textbf{ Change to $cm$ level in the graphs.
 Move 0 stress level data to 5\% to reflect simulation resources required.}}
-
     \label{ExperimentsStressSummary}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V2.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePost.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV2.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V2.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePost.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV2.pdf}
         \caption{}

### 2020-03-13 paper/cav_determinisim_main.tex
added restricted results section and changed conclusion

@@
 \toprule
-Test 		& Actors 						 & Collision & Collision Type 			 & $d_l$ (m) & $n$  &$\max\sigma$ (m) \\ \midrule
-1       & Two vehicles                   & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
-2       & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 2 		 & 1000 & 1000 \\
-3       & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
-4       & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 2 		 & 1000 & 1000 \\
-5       & Two pedestrians                & No  		 & N/A 			  			 & 2 		 & 1000 & 1000 \\
-6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 2 		 & 1000 & 1000 \\
-% 6       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 0.4 		 & 1000 & 1000 \\
-% 8       & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 20 		 & 1000 & 1000 \\
+Test & Actors 					 	  & Collision 	 & Collision Type 			 & $n$  & $\max\sigma$ (m) & $\max\sigma$ (m) \\ 
+	 & 								  & 			 & 							 &	    & (unrestricted)  & (restricted)	\\ \midrule
+1    & Two vehicles                   & No  		 & N/A 			  			 & 1000 & 0.03 & $7.0{\times}10^{-3}$ \\
+2    & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 1000 & 0.31 & $9.8{\times}10^{-3}$ \\
+3    & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 1000 & 0.07 & $5.2{\times}10^{-4}$ \\
+4    & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 1000 & 0.59 & $1.5{\times}10^{-12}$ \\
+5    & Two pedestrians                & No  		 & N/A 			  			 & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
+6    & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
 \bottomrule
 \end{tabular}
-\caption{Test scenarios where $d_l$ is the \textit{look ahead distance}, $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
-CONSIDER removing entries 6 and 8 if they do not contribute towards outcomes and remove references to $d_l$.
-\textbf{remove all reference to $d_l$}}
+\caption{A description of the test scenarios showing the test number, the actors included and if a collision occurred and between which actors.
+Where $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
+The term \textit{unrestricted} refers to an unrestricted subset of the results and allowing system states of any resource utilisation whereas \textit{restricted} employs results filtering to pre-collision and limits system resource utilisation levels to 75\%.}
 \label{TableOfExperiments}
@@
 
-\textbf{Abanoub: can you change the diagram to be a staggered t-junction as the DMR is just confusing for this example.
-Also can we zoom in on the lower right area.
-Tests 5 and 6 have different look ahead values, which is correct?}
-
 \subsection{System Configuration and Screening}
@@
     \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
-    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} \textbf{we need to run these test to look for $\max\sigma$ variations}
+    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator to a single core but no improvement in simulation variance was observed and not used for subsequent testing.
 \end{itemize}
@@
 
-\begin{figure}[h]
+\begin{figure}[!t]
     \centering
@@
     \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
+    \caption{Schematic of test scenarios for (a) tests 1-4, (b) tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
+\end{figure}
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV2.pdf}
+    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
+    \label{ExperimentsStressSummary}
 \end{figure}
@@
 The summary of all the tests are shown in the final column of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels.
-From these summary results it is clear that scenarios with only pedestrian actors not involved in collisions are those that are consistently deterministic over all resource utilisation levels \textbf{(is this true, any other cases?)}.
-However, for other scenarios involving vehicles or mixed actors resource utilisation level has a significant impact on \textit{simulation variance}.
+From these summary results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels.
+However, for other scenarios involving vehicles or a mixture of actors types resource utilisation level has a significant impact on \textit{simulation variance}.
 
@@
 
-However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are an overly pessimistic/conservative account of the situation.
-Examining these specific results (test 2, 4 \& 6??) as a function of simulation time reveals further information about the simulation variance before and after an actor collision callback.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision and is non-deterministic afterwards.
-This pattern of deterministic behaviour prior to collision is maintained up to resource utilisation levels of approximately 50\%, Fig.~\ref{CarsCollisionPrePost}.
-This time series examination is repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-\textbf{What is the summary data for this scenario?} Similarly to vehicle collisions, the variation of $\max\sigma$ for pedestrian collisions indicates deterministic behaviour below 95\% resource utilisation, Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any results after this point are not reliable.
+However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation.
+Examining these specific results, tests 2 \& 4, in more detail as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision and is non-deterministic post-collision.
+This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates pre-collision deterministic behaviour with up to 75\% resource utilisation, Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any post-collision results are not reliable.
 
 The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is not involved in the collision.
+In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}.
 There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as describe above.
-Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour even after the collision \textbf{is this because the walker is killed and removed from the game?}.
-And, thirdly we observed a delayed effect on Car ID 1 showing high simulation variance post-collision but with a $5s$ delayed effect \textit{even though this vehicle was not involved in the collision}.
+Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision}.
+And, thirdly we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delayed effect \textit{even though this vehicle was not involved in the collision}.
 This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors will affect the simulation variance of all actors in the simulation and could potentially result in false-positive assertion violations.
 
-\textbf{Strange gap post-collision for few seconds}
+% \textbf{Strange gap post-collision for few seconds}
 
-\textbf{Should we plot a pre-collision only summary chart for all scenarios?}
-
-\textbf{Should we rename 'test' to 'scenario???'}
+% \textbf{Should we rename 'test' to 'scenario???'}
 
@@
 
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV2.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
-Tests 5 and 6 overlap having almost identical results.
-\textbf{ Change to $cm$ level in the graphs.
-Move 0 stress level data to 5\% to reflect simulation resources required.}}
-    \label{ExperimentsStressSummary}
-\end{figure}
-
 \begin{figure*}[h]
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V2.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V3.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV2.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
         \caption{}
@@
     \end{subfigure}
-    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.
-\textbf{Change Y axis value to $\max\sigma (m)$}}
+    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.}
 \end{figure*}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V2.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV2.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV3.pdf}
         \caption{}
@@
     \end{subfigure}
-    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation.
-\textbf{Change Y axis value to $\max\sigma (m)$}}
+    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation.}
 \end{figure*}
@@
 \section{Conclusions}\label{s:conclusion}
-A detailed description of how practitioners in this field can resolve the non-deterministic operational domains of their simulation is presented.
-Potential sources of non-deterministic simulation behaviour are discussed and some of these are resolved, such as floating point error, and other are left unanswered such as scheduling policy.
-%
-A case study was used to ascertain the boundaries of non-deterministic operation of the Carla driving simulator which seem to exist at collision events and when resource utilisation levels exceeded 50\% for the particular system used in this study.
-A collision is not necessarily a problem as long as the simulation is stopped as soon as it is detected as even actors not involved directly will behave non-deterministically.
+Practitioners in this field can resolve the non-deterministic operational domains of their simulation can be found by following the method outlined in this paper.
+Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy.
+A case study was used to ascertain the boundary between deterministic and non-deterministic operation of the Carla driving simulator.
+Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study.
+Furthermore it was found that all actors, even those not involved directly in a collision, will behave non-deterministically if a collision callback to the game engine is made.
+Best operating practice follows that the simulation should be stopped as soon as a collision callback is detected or resource utilisation levels exceed 75\% although this value may be system specific.
 

### 2020-03-13 paper/cav_determinisim_main.tex
numa comment

@@
 % \pagenumbering{gobble} %suppresses page numbers
-\addbibresource{IROS.bib}
+\addbibresource{cav_determinism.bib}
 \usepackage{balance} %balance columns
@@
     \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
-    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator to a single core but no improvement in simulation variance was observed and not used for subsequent testing.
+    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
 \end{itemize}

### 2020-03-16 paper/cav_determinisim_main.tex
changed fig 4
Modified results typo on summary graph

--- 2020-03-12 paper/cav_determinisim_main.tex
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV2.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
-Tests 5 and 6 overlap having almost identical results.
-\textbf{ Change to $cm$ level in the graphs.
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
+    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.\textbf{ Change to "CPU \& GPU Resource Utilisation" and "y axis to $\max\sigma$ (m)" Change to $cm$ level in the graphs.
+Cannot see Test 6 in the results.
+Remove test 6 and 8 and rename 7 to test 6.
 Move 0 stress level data to 5\% to reflect simulation resources required.}}
+
     \label{ExperimentsStressSummary}

### 2020-03-16 paper/cav_determinisim_main.tex
changed 1cm threshold reason

--- 2020-03-13 paper/cav_determinisim_main.tex
@@
 The user must specify the level of variance acceptable for their application which is used as a threshold for deterministic execution.
-In our case study, Section~\ref{s:case-study}, a threshold of $1cm$ was chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
-This value is sufficiently small to allow accurate \textit{assertion checking} of autonomous vehicles in urban environments [ref] but a larger value may be acceptable for non-urban environments, e.g.
+In our case study, Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
+This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g.
 motorway or highway.
@@
 $1\mu m$.
+Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.
 
-Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
-This is to ensure that an outlier data-point is not hidden in the average of $n$-tests
+% Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
+This is to ensure that an outlier data-point is not hidden in the average of $n$-tests.
 
@@
 \subsection{Output and Analysis}
-The output for each experiment can be summarised as the maximum variance of any actor over the repeated tests.
+The output for each experiment can be summarised as the \textit{maximum variance} of any actor over the $n-$repeated tests.
+The maximum variance is chosen to indicate the worst case rather than an average which could hide outlier data points when a high number of repeated tests are performed.
 This value can be analysed against the different scenarios or settings that have been identified as potential sources of simulation variance.
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV2.pdf}
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV3.pdf}
     \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
@@
 This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
+    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.}
 

### 2020-03-16 Results_matlab_code/CarlaReadLog_v1_GC.m
rm old fex files

@@ -30,7 +30,7 @@ PCC=false; CC=false;
 % file = "TEST_CarsCollision_CG0.txt";CC=true;
 % file = "TEST_CarsCollision_CG25.txt";CC=true;
 % file = "TEST_CarsCollision_CG50.txt";CC=true;
-% file = "TEST_CarsCollision_CG75.txt";CC=true;
+file = "TEST_CarsCollision_CG75.txt";CC=true;
 % file = "TEST_CarsCollision_CG95.txt";CC=true;
 % 
 % file = "TEST_CarsPeople_CG0.txt";
@@ -39,7 +39,7 @@ PCC=false; CC=false;
 % file = "TEST_CarsPeople_CG75.txt";
 % file = "TEST_CarsPeople_CG95.txt";
 % 
-file = "TEST_CarsPeopleCollision_CG0.txt";PCC=true;
+% file = "TEST_CarsPeopleCollision_CG0.txt";PCC=true;
 % file = "TEST_CarsPeopleCollision_CG25.txt";PCC=true;
 % file = "TEST_CarsPeopleCollision_CG50.txt";PCC=true;
 % file = "TEST_CarsPeopleCollision_CG75.txt"; PCC=true;

### 2020-03-20 paper/cav_determinisim_main.tex
Fixed issues occured due to multiple commits.
small change to conclusion

@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV3.pdf}
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
     \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
@@
 
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
-    \caption{Summary of total mean deviation for the different tests runs plotted against the different CPU + GPU stresses.}
-
 \begin{figure*}[h]
@@
 \section{Conclusions}\label{s:conclusion}
-Practitioners in this field can resolve the non-deterministic operational domains of their simulation can be found by following the method outlined in this paper.
+Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
 Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy.

### 2020-04-07 paper/cav_determinisim_main.tex
got through introduction

@@
 \usepackage{balance} %balance columns
+\usepackage{todonotes}
 
@@
 \noindent 
-The connected and autonomous vehicle (CAV) community are adopting the use of gaming engines to develop vehicle control systems and testing environments using simulation.
-These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V).
-This work highlights the requirements these engines need (to fulfil this purpose/for CAV development) and proposes a method to determine \textit{simulation variance} for specific configurations and resource utilisation.
-A case study is used to illustrate the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
+The autonomous vehicle (AV) community is adopting the use of gaming engines to develop vehicle control systems and testing environments using simulation.
+Determinism is a key requirement for simulation during AV development and testing.
+In particular, it ensures repeatability of tests, so that software bugs can be found and fixed, and simulation results are trustworthy.
+%
+% KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
+% These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
+%
+However, gaming engines do not need to be deterministic per se. % when used for gaming. 
+%
+This work identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation variance} for specific configurations and utilisation of the software and hardware, respectively.
+%
+A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
+\todo{KIE: review at end}
+
 % The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
@@
 \begin{IEEEkeywords}
-Autonomous Driving, Determinism, Physics Engines, Connected Autonomous Vehicles (CAV), Verification and Validation(V\&V)
+Autonomous Driving, Determinism, Physics Engines, Autonomous Vehicles, Verification and Validation (V\&V), Simulation, Testing
 \end{IEEEkeywords}
@@
 Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it.
-These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving assertions which must not be violated.
-Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules through assertion checking.
-There have been a number of fatalities with autonomous vehicles which could be attributed to a lack of verification and validation (V\&V), e.g.\cite{FatalityExample}.
-Simulation is an obvious domain choice to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving} where the rare event can be made to happen more often~\cite{Koopman2018} such as unpredictable actor behaviour~\cite{RobustnessAutonomy}.
+These simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving assertions that must not be violated.
+
+Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules, e.g.\ through assertion checking during simulation. 
+There have been a number of fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
+Simulation environments offer a means to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving}.
+In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
+
+% The use of game engines for AV testing
+Increasingly, the autonomous vehicle (AV) community is adopting gaming engines as simulation platforms to support the development and testing of vehicle control software. 
+%
+CARLA\footnote{http://carla.org/}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
 
 % Requirements of simulation based verification SBV
-For games engines to offer a suitable testing environment they must operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors) and be realistic enough \cite{Koopman2018} to produce suitable rendering of scenes for perception stack testing.
-The simulation must be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015} and include a provision for sensor analogues to a suitable degree of accuracy (e.g.
-video, LIDAR, IR).
-Some researchers may also require support for hardware-in-the-loop development or a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}.
+The benefit of using state-of-the-art gaming engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes suitable for perception stack testing.
+%
+Furthermore, they are easy to setup and run; they also are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
+%
+For gaming engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
+CARLA currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
+%
+Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required.
+
 % Game engines for SBV
-Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements, as opposed to vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}; but many challenges still exist.
-
-% Verification and Determinism 
-For game engines to be useful as a verification tool then they must be deterministic. 
+Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, gaming engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
+However, several challenges remain; one of these is determinism.
 %
@@
 %
-Herein the term \textit{simulation variance} is used to refer to how simulation outputs vary when repeated many times and specifically the variance of actor paths within a given scenario.
+Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly under identical conditions.
+Specifically, we focus on the variance of actor paths within a given scenario.
+
 We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
-Dynamic elements are the elements in a scene whose actions or behaviour may change over time, these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals.
-The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.
-A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
+Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals.
+The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.\todo{values - Does this mean stimulus or other parameters?} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
 
+% Verification and Determinism 
 % Why is determinism important for verification
-Determinism is important for verification.
-Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
+%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
+Determinism is a key prerequisite for simulation during AV development and testing. 
+%
+A deterministic simulation environment guarantees that tests are repeatable. 
+%
+Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
+%
+%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
 Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour.
@@
 %
-Also, if the simulation is non-deterministic, variance in actor positions may lead to subsequent assessment errors, e.g. assertion checking.
+If the simulation is not deterministic, however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
+In the worst case, bugs that could have been identified in simulation are left undetected, resulting in false confidence in the safety of the AV's control software.
 
 % Challenges of SBV
-Our main research question is to what extent are gaming engines deterministic so that they can provide a reliable and repeatable testing environment for CAV verification?
-The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.
+Gaming engines do not need to be deterministic when used for gaming; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+Because the situation is very different for AV development and testing, our main research question is:
+how can one assess the extent to which a simulation environment is deterministic? 
+%
+In pursuit of an answer to this question we have developed a method to determine the simulation variance for a given simulation environment. 
+%
+Knowing the simulation variance will help developers and testers understand the impact of using a gaming engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.\todo{review and potentially summarize results from case study}
+%
+%The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo{next}
 
 This paper is structured as follows.
-In Section~\ref{s:background} a background on gaming engines and sources of potential non-determinism are explored.
-Section~\ref{s:methodology} presents a method to design and execute an experiment to assess suitability of a system for CAV verification.
-A case study to illustrate how the simulation variance vary for a number of scenarios is explored in Section~\ref{s:case-study}.
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.
+In Section~\ref{s:background} gaming engines and sources of potential non-determinism in gaming engines are investigated.
+Section~\ref{s:methodology} presents a method to assess the suitability of a simulation system for AV verification.
+%
+A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
 
-% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (CAVs) on roads is a delicate if not utterly fascinating subject.
+% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (AVs) on roads is a delicate if not utterly fascinating subject.
 Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.
@@
 Similarly, the use of simulation for traffic modelling is well established.
-Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate CAVs\cite{FrameworkAndChallenges}.\\\\
+Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate AVs\cite{FrameworkAndChallenges}.\\\\
 
-% While traffic-level simulators can model road layouts, they do not require as much realism or detail as CAVs require.
-On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for CAV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
+% While traffic-level simulators can model road layouts, they do not require as much realism or detail as AVs require.
+On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for AV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
 % Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
-% Consequently, one of the major routes CAVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate CAVs \cite{FrameworkAndChallenges}.
+% Consequently, one of the major routes AVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate AVs \cite{FrameworkAndChallenges}.
 % ======================
@@
 % Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
-% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of CAV software.\\\\
+% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of AV software.\\\\
 % \TODO{Remove this paragraph?} Determinism of gaming (physics) engines has not attracted much attention by engines' developers nor researchers.
@@
 
-% This paper aims to investigate how deterministic physics engines are for usage in V\&V for CAV simulations; and will cover the list of following points:
+% This paper aims to investigate how deterministic physics engines are for usage in V\&V for AV simulations; and will cover the list of following points:
 % \begin{itemize}[leftmargin=*]
-%     \item Necessary background on gaming engines used in CAV simulations.
+%     \item Necessary background on gaming engines used in AV simulations.
 %     \item Review of sources of non-determinism in such engines.
 %     \item Provide a method of how the level of non-determinism can be investigated and defined.
-%     \item A case study showing the implementation of this methodology in an engine widely used in CAVs simulations (Unreal Engine).
+%     \item A case study showing the implementation of this methodology in an engine widely used in AVs simulations (Unreal Engine).
 % \end{itemize}
@@
 
-There are numerous game engines with their associated development environments which could be suitable for CAV development, e.g.
+There are numerous game engines with their associated development environments which could be suitable for AV development, e.g.
 Unreal~\footnote{\url{https://www.unrealengine.com/}}, Unity~\footnote{\url{https://unity.com/}}, CryEngine~\footnote{\url{https://www.cryengine.com/}}.
@@
 Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution.
-%However, when these engines are used for CAV development then deterministic behaviour is required.
+%However, when these engines are used for AV development then deterministic behaviour is required.
 This section gives an overview of a gaming engine and what sources or settings in the engine may affect \textit{simulation variance}.
 %
-% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in CAV simulations.
+% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations.
 Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
@@
 \end{figure}
-The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment.
+The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the resultant actions of the autonomous vehicle given the current state of the environment.
 For example, the throttle and steering inputs from the vehicle controller of an AV.
@@
     \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
-    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a gaming engine for usage in CAV simulations.}
+    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a gaming engine for usage in AV simulations.}
 
@@
 %
-% Translating to the usage of CAV simulations in physics engines,
+% Translating to the usage of AV simulations in physics engines,
 This could result in precision issues of, for example, actor position within the environment and falsely satisfying or failing assertions.
@@
 %Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
-%Translating to the usage of CAV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.
+%Translating to the usage of AV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.
 %
@@
 If single bit errors go uncorrected then subsequent computational processing will produce incorrect and non-deterministic results.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit} but CAV verification tasks are likely to incur significant CPU hours and therefore increase exposure to such risks.
+Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit} but AV verification tasks are likely to incur significant CPU hours and therefore increase exposure to such risks.
 \\\\
@@
 %
-CAV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhered to this rate.
+AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhered to this rate.
 %
@@
 Navigation meshes are used in game engines for actor path planning, and could be considered as a potential source of non-deterministic behaviour, e.g.%.
-Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic
+Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
 ~\cite{CARLABenchmark}. 
@@
 Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.
-Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
+Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
 While executing 100 repeats may seem sufficient, in this case the sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
@@
 % The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
-For the majority of CAV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
+For the majority of AV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
 
@@
 % Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
-However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
+However, in such applications of V\&V of AV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
 % If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the simulation variance as a result of allowing the AI to navigate in a more well defined space.
@@
 
-% Given that our interest is on CAV simulations, this study presents a number of experiments run on typical agents used in CAV simulations and their interactions with each other.
+% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
 These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
@@
 % \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of CAV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
 
@@
 
-% In V\&V of CAV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
 
@@
 Note the dotted red line in the plot shows the millimeter (mm) level.
-Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of CAV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
+Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
 This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
@@
 % Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
-However, in V\&V of CAV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
-Thus, the post collision effect should not be of much concern in the context of CAVs.
+However, in V\&V of AV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
+Thus, the post collision effect should not be of much concern in the context of AVs.
 
@@
 % We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the simulation variance of a gaming engine used in V\&V application of CAV simulations.
+This was followed by presenting a method by which one can determine the simulation variance of a gaming engine used in V\&V application of AV simulations.
 The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.

### 2020-04-08 paper/cav_determinisim_main.tex
up to 2.B

@@
 % ***************************************************
-\title{Investigating Determinism of Gaming Engines for Autonomous Vehicle Verification}
+\title{Investigating Determinism of Game Engines for Autonomous Vehicle Verification}
 \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
@@
 \noindent 
-The autonomous vehicle (AV) community is adopting the use of gaming engines to develop vehicle control systems and testing environments using simulation.
+The autonomous vehicle (AV) community is adopting the use of game engines to develop vehicle control systems and testing environments using simulation.
 Determinism is a key requirement for simulation during AV development and testing.
@@
 %
-However, gaming engines do not need to be deterministic per se. % when used for gaming. 
+However, game engines do not need to be deterministic per se. % when used for games. 
 %
@@
 %
-A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
+A case study using Carla, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
 \todo{KIE: review at end}
@@
 % The use of game engines for AV testing
-Increasingly, the autonomous vehicle (AV) community is adopting gaming engines as simulation platforms to support the development and testing of vehicle control software. 
+Increasingly, the autonomous vehicle (AV) community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
 %
-CARLA\footnote{http://carla.org/}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
+Carla\footnote{http://carla.org/}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
 
 % Requirements of simulation based verification SBV
-The benefit of using state-of-the-art gaming engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes suitable for perception stack testing.
+The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes suitable for perception stack testing.
 %
@@
 %
-For gaming engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
-CARLA currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
+For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
+Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
 %
@@
 % Game engines for SBV
-Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, gaming engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
+Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
 However, several challenges remain; one of these is determinism.
@@
 % Challenges of SBV
-Gaming engines do not need to be deterministic when used for gaming; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+Game engines do not need to be deterministic when used for gaming; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
 Because the situation is very different for AV development and testing, our main research question is:
@@
 %
-Knowing the simulation variance will help developers and testers understand the impact of using a gaming engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.\todo{review and potentially summarize results from case study}
+Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.\todo{review and potentially summarize results from case study}
 %
@@
 This paper is structured as follows.
-In Section~\ref{s:background} gaming engines and sources of potential non-determinism in gaming engines are investigated.
+In Section~\ref{s:background} game engines and sources of potential non-determinism in game engines are investigated.
 Section~\ref{s:methodology} presents a method to assess the suitability of a simulation system for AV verification.
 %
-A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
+A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo{point to UE4} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
 
@@
 % In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of AV software.\\\\
-% \TODO{Remove this paragraph?} Determinism of gaming (physics) engines has not attracted much attention by engines' developers nor researchers.
+% \TODO{Remove this paragraph?} Determinism of game (physics) engines has not attracted much attention by engines' developers nor researchers.
 This previously was not important since in gaming, determinism is not critical.
@@
 % \begin{itemize}[leftmargin=*]
-%     \item Necessary background on gaming engines used in AV simulations.
+%     \item Necessary background on game engines used in AV simulations.
 %     \item Review of sources of non-determinism in such engines.
@@
 There are numerous game engines with their associated development environments which could be suitable for AV development, e.g.
-Unreal~\footnote{\url{https://www.unrealengine.com/}}, Unity~\footnote{\url{https://unity.com/}}, CryEngine~\footnote{\url{https://www.cryengine.com/}}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.
-Carla~\footnote{\url{http://carla.org/}}, AirSim~\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo~\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.
-Nvidia Drive Constellation~\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
+Unreal\footnote{\url{https://www.unrealengine.com/}}, Unity\footnote{\url{https://unity.com/}}, CryEngine\footnote{\url{https://www.cryengine.com/}}.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.\ Carla\footnote{\url{http://carla.org/}}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
 
-Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution.
+Investigating the determinism of game engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable execution.
 %However, when these engines are used for AV development then deterministic behaviour is required.
-This section gives an overview of a gaming engine and what sources or settings in the engine may affect \textit{simulation variance}.
+This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
 %
@@
 \subsection{Overview of a Game Loop} \label{GameLoopSection}
+
+Central to a game engine are the main game logic, the artificial intelligence (AI) component, the audio engine, and the physics and rendering engines. For AV simulation, we focus on the latter two.
+%
 The game loop is responsible for the interaction between the physics and rendering engines.
-Fig.~\ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop (initialisation, game logic and decommissioning are removed from this diagram for brevity\footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}).
-A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).~\cite{GameEngineArchBook}
+Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed\footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}.
+A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering)~\cite{GameEngineArchBook}.
 
@@
 \label{GameEngineLoopDiagram}
-\end{figure}
-The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the resultant actions of the autonomous vehicle given the current state of the environment.
-For example, the throttle and steering inputs from the vehicle controller of an AV.
-The update interprets the intended actions of all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states.
+\end{figure}\todo{Fig 1: update rates of which clock?}
+
+The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment.
+For example, the throttle and steering angle from the controller of an AV.
+The update stage interprets the input from all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states.
 The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
@@
 % It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below.
+
+As the game progresses, the game loop runs continuously according to the game clock.\todo{introduce the different clocks and what they govern, real world, game, physics and render}
 %
-The key aspects of the game loop needed for this discussion are as follows:
+At the beginning of each game clock tick, the lag\todo{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
+This measures how far the game's clock is behind compared to the real world.
+%
+Then the user inputs are processed.
+%
+There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
+%
+Rendering occurs once the game clock has caught up with the real world clock. 
+%
+The process then repeats.
 
-\begin{itemize}[leftmargin=*]
-    \item At the beginning of each tick, the lag between the game clock and the real world is updated based on how much real time passed.
-This measures how far the game's clock is behind compared to the real world.
-    \item Then the user inputs are processed.
-    \item There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
-    \item Rendering occurs once the game clock has caught up with the real world clock. 
-    \item The process then repeats.
-\end{itemize}
-
-If the physics engine uses a fixed time step ($dt$), %, because it makes everything simpler and more stable for physics and AI.
-the shorter this time step is the smoother the interpretation of the physical dynamics will be.
+The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
+The shorter this time step is, the smoother the interpretation of the physical dynamics will be.
 %processing time it takes to catch up to game clock.
@@
 %
+Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
+%
 To use a fixed physics time step the user's display refresh rate needs to be known in advance.
 This requires an update loop to take less than one render tick (one frame of real world time).
-Given the range of different hardware capabilities a variable delta time is often implemented for game playing taking the previous frame time as the next $dt$.
+Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$.
 However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}.
+%
 Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware.
-Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}}.
-If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes, see lower part of Fig.~\ref{GameEngineLoopDiagram}.
-%\cite{GameEngineArchBook}\cite{GameProgPatternsBook}
+Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}.
+%
+If the engine tries to render between physics updates, \textit{residual lag} can occur and extrapolation between frames needs to be performed to smooth transition between scenes.
+This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}.
+\cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo{Should this still be cited?}\todo{Why is this part of the figure?
+It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.}
 %
@@
 In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
-and the simulation will exhibit lag between input commands and rendered states resulting in unsynchronised and unrealistic behaviour.
+and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.
 
@@
     \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
-    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a gaming engine for usage in AV simulations.}
+    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a game engine for usage in AV simulations.}
 
@@
 
-% Our hypothesis is that since UE4 is a gaming engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
+% Our hypothesis is that since UE4 is a game engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
 This behaviour is expected not to hold as the level of computing stress increases.
@@
 % We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the simulation variance of a gaming engine used in V\&V application of AV simulations.
+This was followed by presenting a method by which one can determine the simulation variance of a game engine used in V\&V application of AV simulations.
 The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.

### 2020-04-08 paper/cav_determinisim_main.tex
got through the 2nd section just before Method

@@
 \label{GameEngineLoopDiagram}
-\end{figure}\todo{Fig 1: update rates of which clock?}
+\end{figure}
 
@@
 The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
+\todo{Fig 1: update rates of which clock?}\todo{Should there not also be a clock in the overall loop?}
+
 % 
@@
 
-The following points form a hypothesis around the potential sources of non-determinism that could occur in game engines. \\\\
+The following discussion is the result of our investigation into the potential sources of non-determinism that could occur in game engines.
+We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
+
+\medskip
+
 % \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
@@
 \noindent\underline{\textit{Floating Point Representation:}}
-Floating point representation of real numbers is limited with fixed register sizes resulting in errors of \textit{rounding} where loss of bits occur which are rounded or truncated and \textit{cancellation} which happens where two similar values are subtracted~\cite{FloatingPointsBook,goldberg1991every}.
+Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
+Thus, the results of arithmetic operations must fit into the given bit with.
+This is achieved through rounding to the nearest representable value, which gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
+%
+% Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.
 %
 % Translating to the usage of AV simulations in physics engines,
-This could result in precision issues of, for example, actor position within the environment and falsely satisfying or failing assertions.
-Such reasons could  make one jump to the conclusion that floating point errors could result in non-deterministic behaviour of repeated tests, but these precision issues should be referred to as \textit{incorrectness} and not be confused with determinism.
-Assuming hardware is deterministic and a rounding error occurs, then the same hardware should produce the same \textit{incorrect} output for the same initial conditions.
-So, even if the output is incorrect due to floating point precision it should always produce the same \textit{incorrect} output, given the same logical registers which conform to the same IEEE floating point standard~\cite{8766229}.
-Therefore, it is unlikely that floating point representation will contribute to \textit{simulation variance} for repeated tests on the same hardware.
+In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo{state what assertions are (used for)}
 %
+One could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.
+However, these precision issues are better described as \textit{incorrectness}, rather than them being a source of non-determinism.
+%
+It is safe to assume that the hardware on which game engines run is designed to be deterministic. 
+%
+Thus, when a rounding error occurs, then the same hardware should produce the same \textit{incorrect} result given the same operands. 
+%
+So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always 
+be equally \textit{incorrect} as long as the implementation conforms to the IEEE floating-point standard~\cite{8766229}.
+%
+Beyond hardware limitations, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
+However, the same executable should still return the same output for identical input.
+%
+Therefore, it can be concluded that floating point representation does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware.
+%
+% KIE: The text below is not sufficiently rigorous and not correct. 
 % A generic issue with computers are floating points precision. 
@@
 \noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
-ECC Memory can detect and correct for single bit errors in DRAM memory.
+ECC Memory can detect and correct single bit errors in DRAM memory.
 Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
-If single bit errors go uncorrected then subsequent computational processing will produce incorrect and non-deterministic results.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit} but AV verification tasks are likely to incur significant CPU hours and therefore increase exposure to such risks.
+If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
+Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
+AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.\todo{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
+;)}
 \\\\
 \noindent\underline{\textit{Game Engine Code:}}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo{introduce actor navigation mesh}, deterministic ego vehicle\todo{introduce ego vehicle earlier on} controllers and engine texture loading rates especially for perception stack testing.
 %
-For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/index.html}} command can be used to determine processing bottlenecks by indicating the total time spent generating a frame, executing the game loop or render thread.
+For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo{what does indicating mean?} the total time spent generating a frame, executing the game loop or render thread.
 \\\\
-\noindent\underline{\textit{Scheduling:}} %
+\noindent\underline{\textit{Scheduling:}}\todo{is this at runtime?} %
 % info at:
@@
 % https://en.wikipedia.org/wiki/Race_condition#cite_note-1
-Scheduling is a resource management method for sharing computational resources between tasks of different or competing priority where tasks are executed  dependent on the scheduler policy.
+Scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
+\todo{is this the scheduling that is done by the operating system, i.e.
+at runtime?}
 %
-If during simulation the scheduling policy was updated or thread priority was inconsistently handled then a high variance in output may be possible.
-It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.
-%
+If during simulation the scheduling policy was updated or thread priority was inconsistently handled, then a high variance in output may result.
+It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo{justify - WHY?}
+% KIE: If all threads have the same priority, then the scheduler is free to select one, this can give rise to non-det - or not?
 
-Similar to thread scheduling, the CPU scheduling policy determines which processor core to execute processes in the ready queue and this is maybe ordered by factors such as throughput, latency and CPU utilisation.
+Similar to thread scheduling, the CPU scheduling policy determines on which processor core to execute processes in the ready queue and this is maybe ordered by factors such as throughput, latency and CPU utilisation.
 If the same single-threaded script executes multiple times across different physical cores due to the CPU utilisation policy then should you expect deterministic behaviour?
 Processor core designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist.
-End users expect such minor variations to be controlled for in architecture tolerances and factory testing.
+End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo{not clear, what is the difference between scheduling cores and threads - are both done by the operating system?}
 
-Scheduling of cores or threads may be a reason for increased \textit{simulation variance} but controlling for this falls outside of the scope of this work and no attempt to control these parameter have been made here.
-However, users should be aware of this potential issue and consider methods to move towards a fixed scheduling sequence if possible.
+Overall, scheduling of cores or threads at runtime may be a reason for increased \textit{simulation variance}. 
+%
+% KIE: Not relevant here - we are just listing the sources of non-det.
+% but controlling for this falls outside of the scope of this work and no attempt to control 
+% these parameter have been made here. However, users should be aware of this potential issue 
+% and consider methods to move towards a fixed scheduling sequence if possible.
 \\\\
 \noindent\underline{\textit{Concurrency:}}
+\todo{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
+Consider also parallelism, e.g.
+in GPUs.
+Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism.
+Say which parts of a game engine could reasonably be expected to run concurrently.}
 Another issue to consider is concurrency, where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program.
 This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use such techniques to improve performance although monitoring and detecting such events falls outside of the scope for this work.
+A game engine may use such techniques to improve performance.\todo{how?}
+% KIE: We only list the potential sources of non-det, I therefore removed this. 
+% although monitoring and detecting such events falls outside of the scope for this work. 
 \\\\
@@
 Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
-The question here is will changes in latency between repeated tests impact the game engine to operate non-deterministically or simply with a increased data transfer cost?
-This may only be answered by determining the scheduling policy of the game engine, but by binding a process to a specific core the variations in data transfer time can be minimised.
+%
+Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically, or, perhaps, simply with an increased data transfer cost. 
+%
+By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
 \\\\
@@
 \noindent\underline{\textit{Physics and Rendering clocks:}}
-I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.
+\todo{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
 % The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
-Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.
-in Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
-This is good for hardware scalability but creates a challenge for the physics calculations which provides the most consistent results with a small fixed time step.
+Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
+This is good\todo{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
 %
+A variable frame rate\todo{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo{what does the hardware have to do with this?
+Do we not assume the same HW at this stage?} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
+Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources.
 %
-A variable frame rate could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.
-Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
-Repeated tests on the same system should result in similar frame rates unless there are additional competition for resources.
+AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
 %
-AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhered to this rate.
-%
-That is to say, if the frame rate is allowed to vary then it has to be controlled to give the same variability through out a given test in order for the test results to be repeatable.
-This issue might be of negligible significance if the frame rate is consistent and the time steps are small enough.
+That is to say, if the frame rate is allowed to vary, then this variation has to be controlled to give the same variability throughout a given test in order for the test results to be repeatable.
+This issue might be of negligible significance if the frame rate is consistent and the time steps\todo{of which clock} are small enough.
 \\\\
@@
 ~\cite{CARLABenchmark}. 
-This may depend on the algorithms used by the engine which typically use the A* algorithm\cite{AStarBook} and should give deterministic outcomes as long as the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
+This may depend on the algorithms used by the engine which typically use the A* algorithm~\cite{AStarBook} and should give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
 Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
 %
-It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps~\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/index.html}}.
-%This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes every time they are loaded.
+It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.\todo{unclear :(}
+% This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes every time they are loaded.
+
+\todo[inline]{Add a summary/conclusion?}
 
@@
 \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/index.html}}.
+Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}.
 However, none of these options improved simulation variance significantly beyond the default settings, see our previous investigation~\cite{TSLUnrealEngineTesting}.

### 2020-04-09 paper/cav_determinisim_main.tex
todonotes inline

@@
 A case study using Carla, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
-\todo{KIE: review at end}
+\todo[inline]{KIE: review at end}
 
@@
 Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals.
-The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.\todo{values - Does this mean stimulus or other parameters?} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
+The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.\todo[inline]{values - Does this mean stimulus or other parameters?} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
 
@@
 %
-Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.\todo{review and potentially summarize results from case study}
+Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.\todo[inline]{review and potentially summarize results from case study}
 %
-%The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo{next}
+%The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
 
@@
 %
-A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo{point to UE4} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
+A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}
 
@@
 The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
-\todo{Fig 1: update rates of which clock?}\todo{Should there not also be a clock in the overall loop?}
+\todo[inline]{Fig 1: update rates of which clock?}\todo[inline]{Should there not also be a clock in the overall loop?}
 
@@
 
-As the game progresses, the game loop runs continuously according to the game clock.\todo{introduce the different clocks and what they govern, real world, game, physics and render}
+As the game progresses, the game loop runs continuously according to the game clock.\todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
 %
-At the beginning of each game clock tick, the lag\todo{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
+At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
 This measures how far the game's clock is behind compared to the real world.
@@
 This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}.
-\cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo{Should this still be cited?}\todo{Why is this part of the figure?
+\cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited?}\todo[inline]{Why is this part of the figure?
 It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.}
@@
 % Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo{state what assertions are (used for)}
+In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo[inline]{state what assertions are (used for)}
 %
@@
 Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
-AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.\todo{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
+AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
 ;)}
@@
 \noindent\underline{\textit{Game Engine Code:}}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo{introduce actor navigation mesh}, deterministic ego vehicle\todo{introduce ego vehicle earlier on} controllers and engine texture loading rates especially for perception stack testing.
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo[inline]{introduce actor navigation mesh}, deterministic ego vehicle\todo[inline]{introduce ego vehicle earlier on} controllers and engine texture loading rates especially for perception stack testing.
 %
-For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo{what does indicating mean?} the total time spent generating a frame, executing the game loop or render thread.
+For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean?} the total time spent generating a frame, executing the game loop or render thread.
 \\\\
-\noindent\underline{\textit{Scheduling:}}\todo{is this at runtime?} %
+\noindent\underline{\textit{Scheduling:}}\todo[inline]{is this at runtime?} %
 % info at:
@@
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-\todo{is this the scheduling that is done by the operating system, i.e.
+\todo[inline]{is this the scheduling that is done by the operating system, i.e.
 at runtime?}
@@
 If during simulation the scheduling policy was updated or thread priority was inconsistently handled, then a high variance in output may result.
-It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo{justify - WHY?}
+It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo[inline]{justify - WHY?}
 % KIE: If all threads have the same priority, then the scheduler is free to select one, this can give rise to non-det - or not?
@@
 Processor core designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist.
-End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo{not clear, what is the difference between scheduling cores and threads - are both done by the operating system?}
+End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo[inline]{not clear, what is the difference between scheduling cores and threads - are both done by the operating system?}
 
@@
 \noindent\underline{\textit{Concurrency:}}
-\todo{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
+\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
 Consider also parallelism, e.g.
@@
 This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use such techniques to improve performance.\todo{how?}
+A game engine may use such techniques to improve performance.\todo[inline]{how?}
 % KIE: We only list the potential sources of non-det, I therefore removed this. 
@@
 \noindent\underline{\textit{Physics and Rendering clocks:}}
-\todo{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
+\todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
 % The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
 Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
-This is good\todo{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
+This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
 %
-A variable frame rate\todo{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo{what does the hardware have to do with this?
+A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
 Do we not assume the same HW at this stage?} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
@@
 That is to say, if the frame rate is allowed to vary, then this variation has to be controlled to give the same variability throughout a given test in order for the test results to be repeatable.
-This issue might be of negligible significance if the frame rate is consistent and the time steps\todo{of which clock} are small enough.
+This issue might be of negligible significance if the frame rate is consistent and the time steps\todo[inline]{of which clock} are small enough.
 \\\\
@@
 %
-It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.\todo{unclear :(}
+It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.\todo[inline]{unclear :(}
 % This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes every time they are loaded.
@@
 
-% \todo{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
+% \todo[inline]{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
 
-% \todo{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
+% \todo[inline]{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
 %

### 2020-04-09 paper/cav_determinisim_main.tex
\\ removed where needed

@@
 If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
-AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
-;)}
+Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
+;)} AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
 \\\\
@@
 % and consider methods to move towards a fixed scheduling sequence if possible.
-\\\\
+
 \noindent\underline{\textit{Concurrency:}}
@@
 % although monitoring and detecting such events falls outside of the scope for this work.
-\\\\
+
 \noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
@@
 By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
-\\\\
+
 % these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
@@
 This issue might be of negligible significance if the frame rate is consistent and the time steps\todo[inline]{of which clock} are small enough.
-\\\\
+
 \noindent\underline{\textit{Actor Navigation:}}

### 2020-04-09 paper/cav_determinisim_main.tex
minor todo

@@
 The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
-\todo[inline]{Fig 1: update rates of which clock?}\todo[inline]{Should there not also be a clock in the overall loop?}
+\todo[inline]{Fig 1: update rates of which clock?}\todo[inline]{Should there not also be a clock in the overall loop?}\todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.}
 

### 2020-04-15 paper/cav_determinisim_main.tex
added some comments based on our discussion of the different priorities in gaming/game development (wrt user experience in real time coming before realism), and AV V&V where realism is critically important, especially for preception stack testing and visual inspection of accidents or near misses)

@@
 % Requirements of simulation based verification SBV
-The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes suitable for perception stack testing.
+The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
 %
@@
 
-Investigating the determinism of game engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable execution.
+Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.
 %However, when these engines are used for AV development then deterministic behaviour is required.
@@
 In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.
+and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?}
 
@@
 \end{figure*}
+
+\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
+This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer).
+For instance, see below, but I'm not sure whether this is the right place.}
+In addition, providing game players with a responsive real-time experience is often achieved at the cost of realism.\todo[inline]{Do we have a ref for this, Unity vs UE4?}
+%
+In contrast, realism is essential in AV simulation, especially for perception stack testing, but also for visual inspection.
+If necessary, it is acceptable to achieve the required level of realism at the cost of the real-time user experience.
 
@@
 \noindent\underline{\textit{Concurrency:}}
+% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
 \todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.

### 2020-04-15 paper/cav_determinisim_main.tex
still within III.A Determinism Threshold

@@
     \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
-    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a game engine for usage in AV simulations.}
+    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation environment.}
 
@@
 \section{Methodology} \label{s:methodology}
-In this section a method for determining \textit{simulation variance} of an end-user system is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
-The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may tax different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
-To replicate high computational loads that may be experienced in high realism simulations, software that artificially utilises resources is used at runtime for both central and graphical processing units.
+In this section a method for determining the variance of a simulation environment, termed \textit{simulation variance} is presented.
+We have identified a number of experimental parameters that may influence the simulation variance of a game engine, see Fig.~\ref{method_diagram}.
+These need to be considered at different stages in the process.
+%
+The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
+These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
+\todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''.} 
+%
+To replicate and control the high computational loads that may be experienced in high performance simulations, software that artificially utilises resources is executed alongside the simulation on both central and graphical processing units.
+\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW.}
 % The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
@@
 
-\subsection{Deterministic Threshold}
-We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
-This threshold is analogous to the analogue signal level associated with a binary state in a logical processor.
-At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.
+\subsection{Setting a Determinism  Threshold}
+We use the term \textit{determinism threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
+This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
+At some low analogue voltage the signal is interpreted as binary $0$, this may be below 0.5V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above 3.5V.
 
-The user must specify the level of variance acceptable for their application which is used as a threshold for deterministic execution.
-In our case study, Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
+The level of variance acceptable for the simulation must first be set.
+It serves as a threshold for deterministic execution.
+This level depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+%
+\todo{KIE HERE}
+In our case study, presented in Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
 This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g.

### 2020-04-15 paper/cav_determinisim_main.tex
still III.A

@@
 \subsection{Setting a Determinism  Threshold}
+The level of variance acceptable for a simulation must first be set. It serves as a threshold for deterministic execution.
+%
 We use the term \textit{determinism threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. 
+%
 This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
 At some low analogue voltage the signal is interpreted as binary $0$, this may be below 0.5V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above 3.5V.
+This convention enables the abstraction from the physical signal values that are analogue voltages into the digital domain with binary representation.
 
-The level of variance acceptable for the simulation must first be set.
-It serves as a threshold for deterministic execution.
-This level depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+The determinism threshold depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+In particular, this threshold must be sufficiently small to enable accurate assertion checking.
+\todo[inline]{Make sure assertion checking has previously been introduced.}
 %
-\todo{KIE HERE}
+
+\todo[inline]{KIE: Move this to the Case Study section.}
 In our case study, presented in Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.

### 2020-04-18 paper/cav_determinisim_main.tex
gone fully through and added lots of todo notes, suggested restructuring to move case study before the derivation of the general methodology

@@
 
+\todo[inline]{We may need to introduce the Case Study first, then derive the general Methodology from it.
+This way, we can explain the requirements of the case study, how things were set up, what was observed, how the methodology was refined, several iterations even, etc.}
+
 \subsection{Setting a Determinism  Threshold}
@@
 In particular, this threshold must be sufficiently small to enable accurate assertion checking.
-\todo[inline]{Make sure assertion checking has previously been introduced.}
+\todo{Make sure assertion checking has previously been introduced.}
 %
+In practice, it may be necessary to determine this threshold experimentally.
 
-\todo[inline]{KIE: Move this to the Case Study section.}
+\todo[inline]{KIE: Move this to the Case Study section and justify the values.
 In our case study, presented in Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
@@
 Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.
+$1\mu m$.}
+%
+\todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.}\todo{Formal writing, no reference to the group or project, please.}
 
@@
 
-\subsection{Scenario Description}
-The experiment should be a short scenario that can be easily repeated with the same initial conditions.
+\subsection{Scenario Description}\todo[inline]{Select section titles to match the flow diagram.
+In each of these sections, go through the matching phase in the flow diagram, explaining each point listed there and justify recommendations.}
+The experiment should be a short\todo{What does short mean?
+Quantify.} scenario that can be easily repeated with the same initial conditions.
 Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-The maximum deviation in actor path with respect to the mean for $n$-repeated tests will indicate at what dimensional scale deterministic results are achieved.
+The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
 For example, if 100 tests are executed and the maximum deviation in actor path is $<1cm$, then this scenario could be interpreted as being deterministic above this scale.
 
-Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.
-Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
+Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo{Explain what this means.} Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
 While executing 100 repeats may seem sufficient, in this case the sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
@@
 
-Game engine actors can be used to simulate the AV and other vehicles and pedestrians.
+Game engine actors can be used to simulate the AV and other vehicles as well as pedestrians.
 For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
-In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
-Initial testing also indicated that when actors collide there is a significant change in \textit{simulation variance}, therefore the experiment should include some scenarios where collision callbacks are activated.
 
-It is also important to control any pseudo-random numbers that are called in any external during the simulation with fixed seeds to ensure repeatable testing, e.g.
-external vehicle control programs or other clients connected to the simulator.
+\todo[inline]{This is mixing our experiments (which are part of the case study) with the methodology.
+It is probably best to first introduce the case study and then derive the methodology from it, rather than mixing both.} In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
+Initial testing also indicated that when actors collide there is a significant change in \textit{simulation variance}, therefore the experiment should include some scenarios where collision callbacks are activated.\todo[inline]{Make sure collision callbacks have been introduced earlier.}
+
+Randomisation must be fully controlled by fixing the seeds for any random number generation that is used during the experiment.
+It is important to identify and control any external sources that may use randomisation, e.g.\ external vehicle controllers or other clients connected to the simulation environment.
 
@@
 
-\subsection{System Resources}
-To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will  contribute towards increased \textit{simulation variance}.
-CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system.
-This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
-For GPU utilisation \texttt{gpu-burn} was employed using the fur test at different resolutions and multiple instances can be used to generate high utilisation levels~\cite{GPU_stress}.
-
-% The user should also monitor and record system CPU and GPU utilisation rates.
-Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
-
-% Collision callbacks
-% \subsection{Other Considerations}
-% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
-ii) If not, how does its simulation variance vary?
-In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
-Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
-% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
-Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
-% The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
-For the majority of AV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
-
-\subsection{Internal settings}
+\subsection{Internal settings}\todo[inline]{Systematically go through the points listed in the flow diagram, explain and justify.}
 Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
@@
 \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
+
 Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}.
-However, none of these options improved simulation variance significantly beyond the default settings, see our previous investigation~\cite{TSLUnrealEngineTesting}.
+However, none of these options improved simulation variance significantly beyond the default settings, see our previous investigation\footnote{\url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}.
 
@@
 
+\subsection{System Resources}\todo[inline]{This should be ``External Settings'' - from flow diagram?}
+To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
+Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will  contribute towards increased \textit{simulation variance}.
+CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system.
+This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
+For GPU utilisation \texttt{gpu-burn} was employed using the fur test at different resolutions and multiple instances can be used to generate high utilisation levels~\cite{GPU_stress}.\todo[inline]{Again, this sounds as if having the case study first makes more sense.
+That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
+
+% The user should also monitor and record system CPU and GPU utilisation rates.
+Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
+
+% Collision callbacks
+% \subsection{Other Considerations}
+% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
+ii) If not, how does its simulation variance vary?
+In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
+Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
+% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
+Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
+% The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
+For the majority of AV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
+
 % \subsection{External settings}
@@
 \subsection{Output and Analysis}
-The output for each experiment can be summarised as the \textit{maximum variance} of any actor over the $n-$repeated tests.
+\todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
+The output for each experiment can be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests.
 The maximum variance is chosen to indicate the worst case rather than an average which could hide outlier data points when a high number of repeated tests are performed.
@@
 % ***************************************************
-\section{Case Study} \label{s:case-study}
+\section{Case Study} \label{s:case-study}\todo[inline]{Put this section before Methodology.}
 We present a case study to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction}.
@@
 
+\todo[inline]{Introduce the Scenario - see my comment later.}
+
+\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+
+\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
+
+\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
 In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation.
@@
 % \subsection{Tests Description And Technicalities}\label{TestsDescriptionAndTechnicalities}
-\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
-The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover a range of interactions between actor types.
+\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}\todo[inline]{This should go first.}
+\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+Then say how this has been implemented, referring to test numbers and the }
+The tests listed in Table~\ref{TableOfExperiments} were chosen to cover a range of interactions between actor types.
 Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
 In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1m$ intervals using a PID controller.
-In test 3 a mixture of different actor types are introduced where two vehicles dive without collision and a pedestrian walks across the road at a crossing point.
-In test 4 this pedestrian collides with one of the vehicles at the crossing triggering a collision callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly space waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
-Tests 5 \& 6 involve only pedestrians where they do not collide (test 5) and where they do (test 6), see Fig.~\ref{Test_b}.
+In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
+In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a collision callback, see Fig~\ref{Test_a}.
+Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
+Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
 
@@
 % ***************************************************
-\subsection{Results And Discussion}\label{ResultsSection}
+\subsection{Results and Discussion}\label{ResultsSection}
 The summary of all the tests are shown in the final column of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels.
-From these summary results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels.
-However, for other scenarios involving vehicles or a mixture of actors types resource utilisation level has a significant impact on \textit{simulation variance}.
+From these  results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels.
+However, for other scenarios involving vehicles or a mixture of actors types, the resource utilisation level has a significant impact on \textit{simulation variance}.
 
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU).
-In this figure, anything above the $cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
-A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU).
+In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.\todo[inline]{Where are the verification requirements listed?
+They should be part of the scenario description.} A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s).}
 
@@
 Examining these specific results, tests 2 \& 4, in more detail as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision and is non-deterministic post-collision.
-This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, Fig.~\ref{CarsCollisionPrePost}.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision.
+This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
 This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates pre-collision deterministic behaviour with up to 75\% resource utilisation, Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work and suggests that verification engineers should consider this as an end condition for terminating tests as any post-collision results are not reliable.
+Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism.
+\todo{Why ``not reliable''.}
 
@@
 In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}.
-There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as describe above.
-Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision}.
-And, thirdly we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delayed effect \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors will affect the simulation variance of all actors in the simulation and could potentially result in false-positive assertion violations.
+There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as described above.
+Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
+Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delay \textit{even though this vehicle was not involved in the collision}.
+This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  false-positive assertion violations.\todo[inline]{Make sure that assertion checking has been introduced before.}
 
@@
 
-\textbf{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
+\todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
 
@@
 This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
+
+\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
+This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
+Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
 
@@
 % ***************************************************
-\section{Conclusions}\label{s:conclusion}
+\section{Conclusions and Future Work}\label{s:conclusion}
+
+\todo[inline]{Say what the paper investigated and what we aimed to achieve, why this is important and how we went about it.
+Then point out the derivation of a re-usable methodology based on the knowledge and understanding we gained from the case study.}
+
+\todo[inline]{Review/rewrite this to match the earlier changes in the paper.
 Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
@@
 Best operating practice follows that the simulation should be stopped as soon as a collision callback is detected or resource utilisation levels exceed 75\% although this value may be system specific.
+}
+
+\todo[inline]{Point out (again) that all data and scripts etc are on github.}
+
+\todo[inline]{Future work...}
 

### 2020-04-20 paper/cav_determinisim_main.tex
New branch with AG's updates

--- 2020-04-15 paper/cav_determinisim_main.tex
@@
 % Requirements of simulation based verification SBV
-The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
+The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes suitable for perception stack testing.
 %
@@
 
-Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.
+Investigating the determinism of game engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable execution.
 %However, when these engines are used for AV development then deterministic behaviour is required.
@@
 In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?}
+and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.
 
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
-    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation environment.}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v2.pdf}
+    \caption{Shows a flow diagram of the methodology proposed to define the simulation variance of a game engine for usage in AV simulations.}
 
@@
 \end{figure*}
-
-\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
-This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer).
-For instance, see below, but I'm not sure whether this is the right place.}
-In addition, providing game players with a responsive real-time experience is often achieved at the cost of realism.\todo[inline]{Do we have a ref for this, Unity vs UE4?}
-%
-In contrast, realism is essential in AV simulation, especially for perception stack testing, but also for visual inspection.
-If necessary, it is acceptable to achieve the required level of realism at the cost of the real-time user experience.
 
@@
 \noindent\underline{\textit{Concurrency:}}
-% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
 \todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
@@
 \section{Methodology} \label{s:methodology}
-In this section a method for determining the variance of a simulation environment, termed \textit{simulation variance} is presented.
-We have identified a number of experimental parameters that may influence the simulation variance of a game engine, see Fig.~\ref{method_diagram}.
-These need to be considered at different stages in the process.
-%
-The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
-These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
-\todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''.} 
-%
-To replicate and control the high computational loads that may be experienced in high performance simulations, software that artificially utilises resources is executed alongside the simulation on both central and graphical processing units.
-\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW.}
+In this section a method for determining \textit{simulation variance} of an end-user system is presented along with experimental parameters that may influence it, see Fig.~\ref{method_diagram}.
+The method includes scenarios that exercise different actor types and functional callbacks of the game engine that may tax different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
+To replicate high computational loads that may be experienced in high realism simulations, software that artificially utilises resources is used at runtime for both central and graphical processing units.
 % The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
@@
 
-\subsection{Setting a Determinism  Threshold}
-We use the term \textit{determinism threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
-This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
-At some low analogue voltage the signal is interpreted as binary $0$, this may be below 0.5V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above 3.5V.
+% \subsection{Deterministic Threshold}
+% We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
+This threshold is analogous to the analogue signal level associated with a binary state in a logical processor.
+At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.
 
-The level of variance acceptable for the simulation must first be set.
-It serves as a threshold for deterministic execution.
-This level depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-%
-\todo{KIE HERE}
-In our case study, presented in Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
+% The user must specify the level of variance acceptable for their application which is used as a threshold for deterministic execution.
+In our case study, Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
 This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g.
@@
 A* search algorithm.
-
-\subsection{System Resources}
+\subsection{Execution} 
+\subsubsection{System Resources}
 To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
@@
 
-\subsection{Internal settings}
+\subsubsection{Internal settings}
 Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
@@
 \subsection{Output and Analysis}
+% \subsection{Deterministic Threshold}
+We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
+This threshold is analogous to the analogue signal level associated with a binary state in a logical processor.
+At some low analogue voltage the signal is interpreted as binary $0$ which may be below 0.5V and is considered binary $1$ when the same signal is above 3.5v.
+
+The user must specify the level of variance acceptable for their application which is used as a threshold for deterministic execution.
+In our case study, Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
+This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g.
+motorway or highway.
+Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
+$1\mu m$.
+Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.
+
 The output for each experiment can be summarised as the \textit{maximum variance} of any actor over the $n-$repeated tests.

### 2020-04-20 paper/cav_determinisim_main.tex
GC comments

--- 2020-04-18 paper/cav_determinisim_main.tex
@@
 % The use of game engines for AV testing
-Increasingly, the autonomous vehicle (AV) community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
+Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
 %
@@
 %
-Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly under identical conditions.
+Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly under identical initial conditions.
 Specifically, we focus on the variance of actor paths within a given scenario.
@@
 We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
-Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals.
-The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.\todo[inline]{values - Does this mean stimulus or other parameters?} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
+Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals.
+The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.\todo[inline]{values - Does this mean stimulus or other parameters?
+GC - changed to parameters} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
 
@@
 %
-Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.\todo[inline]{review and potentially summarize results from case study}
+Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.
+
+\todo[inline]{review and potentially summarize results from case study GC - added see below.}
+A case study to highlight the non-deterministic effects in a game engine are presented where scenarios between pedestrian and vehicle actors are used to illustrate conditions that result in high simulation variance.
+By limiting the the level of system utilisation to less than 75\% and terminating scenarios when any collision callback is made, a deterministic operational domain was found.
 %
@@
 %
-A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}
+A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4 GC - ??} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}
 
@@
 \includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
-\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}}
+\caption{Game engine loop block diagram~\cite{GameProgPatternsBook} \todo[inline]{AG: remove the clock symbol and replace with $dt$, possibly remove the ticks chart as this is confusing given they are not ticks but event markers.}}
 \label{GameEngineLoopDiagram}
@@
 The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
-\todo[inline]{Fig 1: update rates of which clock?}\todo[inline]{Should there not also be a clock in the overall loop?}\todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.}
+
+\todo[inline]{Fig 1: update rates of which clock?
+GC-is a clock confusing as it does not run on a 'clock' in computer science terms, rather it is a timer which is used as reference to compare to a 'wall clock'}
+
+\todo[inline]{Should there not also be a clock in the overall loop? GC-yes the wall clock but it may get confusing as that is a real clock and the game loop clock is a timer?}
+
+\todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.
+Unity page on multi-threading \footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and a plugin for multi-thread management \footnote{\url{https://forum.unity.com/threads/task-for-unity-scripting-threading.462971/?_ga=2.228561631.331624405.1587372521-1435703724.1587372521}}}
 
@@
 
-As the game progresses, the game loop runs continuously according to the game clock.\todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
+\todo[inline]{We should agree if the 'game clock' really is a clock or just times the computation for each game loop.}
+
+\todo[inline]{For AG: We use the phase 'game clock tick'- where did the come from as I think it is misleading.}
+
+As the game progresses, the game loop runs continuously according to the game clock. \todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
 %
@@
 This measures how far the game's clock is behind compared to the real world.
+
+\todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors.
+Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
+The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}.
+When the frame is rendered the game loop cycle returns to processing inputs.}
 %
@@
 This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}.
-\cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited?}\todo[inline]{Why is this part of the figure?
-It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.}
+\cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited?
+GC-No, it is cited elsewhere.}
+
+\todo[inline]{Why is this part of the figure?
+It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.
+GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
 %
@@
 In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?}
+and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.
+
+\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?
+GC-suggest remove this part of the diagram, too confusing as there are not really any clocks or ticks - we should remove all mention of ticks too.}
 
@@
 % Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo[inline]{state what assertions are (used for)}
+In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo[inline]{state what assertions are (used for) GC-see below} In the field of autonomous vehicle verification, assertions are the immutable rules and regulations that describe safe and legal driving.
 %
@@
 % A generic issue with computers are floating points precision. 
-%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
+%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.increasingl\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
 %Translating to the usage of AV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.
@@
 Whenever they are not, smart solutions can be found to accommodate for these limitations.
-\\\\
+
 \noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
@@
 Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
-;)} AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
-\\\\
+;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.} AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
+
+\noindent\underline{\textit{Actor Navigation:}}
+There is existing evidence to suggests actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
+% 
+Unreal Engine is the underlying framework for the Carla simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
+%
+The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}. 
+%
+Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.%.
+Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
+%
+However, the environment management in Carla implements fixed binary files for navigation meshes that are linked to each driving scene and cannot be unknowingly modified and therefore can be considered fixed.
+%
+Unless other sources that can alter the fixed environment exist, actor navigation should not be considered a source of non-determinism
+%
+The navigation mesh is a fixed area of the environment where actors are free to navigate. Typically the navigation mesh is fixed for a specific environment or road scene and therefore the A* algorithm
+ should give deterministic results.
+
+% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
+%
+% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
+\todo[inline]{unclear :( GC- re-written}
+
 \noindent\underline{\textit{Game Engine Code:}}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo[inline]{introduce actor navigation mesh}, deterministic ego vehicle\todo[inline]{introduce ego vehicle earlier on} controllers and engine texture loading rates especially for perception stack testing.
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}, deterministic ego vehicle\todo[inline]{introduce ego vehicle earlier on GC added line 99} controllers and engine texture loading rates especially for perception stack testing.
 %
-For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean?} the total time spent generating a frame, executing the game loop or render thread.
+For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean?
+GC-displaying, returning or showing} the total time spent generating a frame, executing the game loop or render thread.
 \\\\
@@
 % https://en.wikipedia.org/wiki/Race_condition#cite_note-1
-Scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy.
+Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
 \todo[inline]{is this the scheduling that is done by the operating system, i.e.
-at runtime?}
+at runtime?
+GC - runtime, added to 1st para.}
 %
 If during simulation the scheduling policy was updated or thread priority was inconsistently handled, then a high variance in output may result.
-It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo[inline]{justify - WHY?}
+It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo[inline]{justify - WHY?
+GC- see below} Given that hardware is considered deterministic and if the scheduling policy does not change between tests, then for the same hardware and operating system configuration the same output will be given.
+%
+However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and the scheduling policy is set to equal priority for all threads then a background process may interrupt the otherwise deterministic sequence of events which may, for example, alter the number of physics calculations that can be performed within the game loop and hence variance in the simulation.
 % KIE: If all threads have the same priority, then the scheduler is free to select one, this can give rise to non-det - or not?
@@
 Processor core designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist.
-End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo[inline]{not clear, what is the difference between scheduling cores and threads - are both done by the operating system?}
+End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo[inline]{not clear, what is the difference between scheduling cores and threads - are both done by the operating system?
+GC - does this come from a software request?
+i.e.
+request to use a core or just a thread}
 
@@
 Say which parts of a game engine could reasonably be expected to run concurrently.}
-Another issue to consider is concurrency, where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program.
+
+Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
 This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use such techniques to improve performance.\todo[inline]{how?}
+A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.\todo[inline]{how?
+GC-see above.}
 % KIE: We only list the potential sources of non-det, I therefore removed this. 
@@
 \noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
-For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the memory location relative to the processor.
+For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
 Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
@@
 A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
-Do we not assume the same HW at this stage?} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
+Do we not assume the same HW at this stage?
+GC- we are just listing sources at this point so should be hardware agnostic.
+also, yes, hardware is the same for us but we cannot control for resource utilisation and background processes.} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
 Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources.
@@
 AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
-%
-That is to say, if the frame rate is allowed to vary, then this variation has to be controlled to give the same variability throughout a given test in order for the test results to be repeatable.
-This issue might be of negligible significance if the frame rate is consistent and the time steps\todo[inline]{of which clock} are small enough.
-
-\noindent\underline{\textit{Actor Navigation:}}
-Navigation meshes are used in game engines for actor path planning, and could be considered as a potential source of non-deterministic behaviour, e.g.%.
-Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
-~\cite{CARLABenchmark}. 
-This may depend on the algorithms used by the engine which typically use the A* algorithm~\cite{AStarBook} and should give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}.
-Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
-%
-It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.\todo[inline]{unclear :(}
-% This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes every time they are loaded.
 
 \todo[inline]{Add a summary/conclusion?}
+
+\noindent\underline{\textit{Summary:}}
+Floating point as long as h/w is det any deterministically-incorrect
+
+ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software
+
+Actor navigation not a problem if navigation meshes are fixed
+
+Game engine, ensuring fixed $dt$ and control of random numners
+
+Core or thread scheduling will not be issue unless changing scheduling policy or migrating bewtween machines that have different policies.
+
+Concurrency is really down to the development of the simulation software and reputable companies should have good control over these potential issues following good software development practice.
+
+NUMA should only affect interprocessor data transfer cost, at wrst only make computations longer.
+relative times beyween access different caches is likely to be small although may impact on high throughput, e.g.
+HPC.
+
+VFR - want to remove this section.
 

### 2020-04-24 paper/cav_determinisim_main.tex
GC updates, change case study location

@@
 Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals.
-The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.\todo[inline]{values - Does this mean stimulus or other parameters?
-GC - changed to parameters} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
+The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.
+A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
+
+% \todo[inline]{values - Does this mean stimulus or other parameters? GC - changed to parameters}
 
@@
 
-\todo[inline]{review and potentially summarize results from case study GC - added see below.}
+% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
 A case study to highlight the non-deterministic effects in a game engine are presented where scenarios between pedestrian and vehicle actors are used to illustrate conditions that result in high simulation variance.
@@
 \includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
-\caption{Game engine loop block diagram~\cite{GameProgPatternsBook} \todo[inline]{AG: remove the clock symbol and replace with $dt$, possibly remove the ticks chart as this is confusing given they are not ticks but event markers.}}
+\caption{Game engine loop block diagram~\cite{GameProgPatternsBook} \todo[inline]{AG: remove the clock symbol, possibly remove the ticks chart as this is confusing given they are not ticks but event markers.}}
 \label{GameEngineLoopDiagram}
@@
 
-The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment.
+% The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment.
 For example, the throttle and steering angle from the controller of an AV.
@@
 
-\todo[inline]{Fig 1: update rates of which clock?
+% \todo[inline]{Fig 1: update rates of which clock?
 GC-is a clock confusing as it does not run on a 'clock' in computer science terms, rather it is a timer which is used as reference to compare to a 'wall clock'}
 
-\todo[inline]{Should there not also be a clock in the overall loop? GC-yes the wall clock but it may get confusing as that is a real clock and the game loop clock is a timer?}
-
-\todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.
+% \todo[inline]{Should there not also be a clock in the overall loop? GC-yes the wall clock but it may get confusing as that is a real clock and the game loop clock is a timer?}
+
+% \todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.
 Unity page on multi-threading \footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and a plugin for multi-thread management \footnote{\url{https://forum.unity.com/threads/task-for-unity-scripting-threading.462971/?_ga=2.228561631.331624405.1587372521-1435703724.1587372521}}}
@@
 
-\todo[inline]{We should agree if the 'game clock' really is a clock or just times the computation for each game loop.}
-
-\todo[inline]{For AG: We use the phase 'game clock tick'- where did the come from as I think it is misleading.}
-
-As the game progresses, the game loop runs continuously according to the game clock. \todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
-%
-At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
+% \todo[inline]{We should agree if the 'game clock' really is a clock or just times the computation for each game loop.}
+
+% \todo[inline]{For AG: We use the phase 'game clock tick'- where did the come from as I think it is misleading.}
+
+% As the game progresses, the game loop runs continuously according to the game clock. \todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
+%
+% At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
 This measures how far the game's clock is behind compared to the real world.
 
-\todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors.
+The game loop cycle starts with initialising the scene and actors.
+Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate (loop in physics update in Fig.~\ref{GameEngineLoopDiagram}).
+The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}.
+When the frame is rendered the game loop cycle returns to processing inputs.
+
+% \todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors.
 Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
@@
 %
-Then the user inputs are processed.
-%
+% Then the user inputs are processed.
 There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
-%
 Rendering occurs once the game clock has caught up with the real world clock.
-%
 The process then repeats.
@@
 This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}.
-\cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited?
-GC-No, it is cited elsewhere.}
-
-\todo[inline]{Why is this part of the figure?
+% \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
+
+% \todo[inline]{Why is this part of the figure?
 It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.
@@
 and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.
-
-\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?
+\todo[inline]{Explain how this lag may affect perception stack testing}
+
+% \todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?
 GC-suggest remove this part of the diagram, too confusing as there are not really any clocks or ticks - we should remove all mention of ticks too.}
@@
 
+% ======= Floating Point
 \noindent\underline{\textit{Floating Point Representation:}}
@@
 % Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.
-%
+
 % Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo[inline]{state what assertions are (used for) GC-see below} In the field of autonomous vehicle verification, assertions are the immutable rules and regulations that describe safe and legal driving.
+In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.
+% \todo[inline]{state what assertions are (used for) GC-see below} 
+In the field of autonomous vehicle verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
 %
@@
 Whenever they are not, smart solutions can be found to accommodate for these limitations.
-
+\\\\
+% ======= Floating Point
 \noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
@@
 Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
-;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.} AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
+;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.}
+
+Any hardware simulation that runs for 1000's hours, such as AV verification tasks, are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
 
@@
  should give deterministic results.
-
 % Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
@@
 % It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
-\todo[inline]{unclear :( GC- re-written}
-
+% \todo[inline]{unclear :( GC- re-written}
+\\\\
 \noindent\underline{\textit{Game Engine Code:}}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}, deterministic ego vehicle\todo[inline]{introduce ego vehicle earlier on GC added line 99} controllers and engine texture loading rates especially for perception stack testing.
-%
-For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean?
-GC-displaying, returning or showing} the total time spent generating a frame, executing the game loop or render thread.
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh 
+% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
+, deterministic ego vehicle
+% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
+ controllers and engine texture loading rates especially for perception stack testing. %
+For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{game} which reports
+% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
+the total time spent generating a frame, executing the game loop or render thread.
 \\\\
-\noindent\underline{\textit{Scheduling:}}\todo[inline]{is this at runtime?} %
+\noindent\underline{\textit{Scheduling:}}
+% \todo[inline]{is this at runtime?} %
 % info at:
@@
 % https://en.wikipedia.org/wiki/Race_condition#cite_note-1
-Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy.
+Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy by the operating system.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-\todo[inline]{is this the scheduling that is done by the operating system, i.e.
-at runtime?
-GC - runtime, added to 1st para.}
+% \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
 %
 If during simulation the scheduling policy was updated or thread priority was inconsistently handled, then a high variance in output may result.
-It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo[inline]{justify - WHY?
-GC- see below} Given that hardware is considered deterministic and if the scheduling policy does not change between tests, then for the same hardware and operating system configuration the same output will be given.
+It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.
+% \todo[inline]{justify - WHY? GC- see below} 
+Given that hardware is considered deterministic and if the scheduling policy does not change between tests, then for the same hardware and operating system configuration the same output will be given.
 %
@@
 request to use a core or just a thread}
-
+%
 Overall, scheduling of cores or threads at runtime may be a reason for increased \textit{simulation variance}. 
@@
 % and consider methods to move towards a fixed scheduling sequence if possible.
-
+\\\\
 \noindent\underline{\textit{Concurrency:}}
@@
 Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
-This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.\todo[inline]{how?
-GC-see above.}
+This technique can lead to issues such as shared resource access conflicts due to mis-managed timing and RACE conditions~\cite{huffman1954synthesis}.
+A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.
+% \todo[inline]{how? GC-see above.}
 % KIE: We only list the potential sources of non-det, I therefore removed this. 
 % although monitoring and detecting such events falls outside of the scope for this work. 
-
+\\\\
 \noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
@@
 By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
-
+\\\\
+\noindent\underline{\textit{Ego Vehicle Controller:}}
+The ego vehicle controller may be a source of non-determinism if using complex control algorithms such as neural networks for visual processing and perception.
+A neural network can only be deterministic within in the bounds of the training data that was used to construct it.
+If the network is exposed to unseen data it may incorrectly classify perfectly legitimate scenes due to lack of training data.
+In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically.
+In this sense, control over weather and lighting in the game engine should be considered and controlled as well as any other dynamic elements to the environment, e.g.
+reflections from surface water.
 % these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
@@
 This is beyond the scope of our work and will not be covered in this paper.
-\noindent\underline{\textit{Physics and Rendering clocks:}}
-\todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
-% The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
-Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
+% \\\\
+% \noindent\underline{\textit{Physics and Rendering clocks:}}
+% \todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
+% % The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
+% Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
 This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
-%
-A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
+% %
+% A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
 Do we not assume the same HW at this stage?
@@
 Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources.
-%
-AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
-
-\todo[inline]{Add a summary/conclusion?}
-
+% %
+% AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
+% \todo[inline]{Add a summary/conclusion?}
+\\\\
 \noindent\underline{\textit{Summary:}}
-Floating point as long as h/w is det any deterministically-incorrect
-
-ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software
-
-Actor navigation not a problem if navigation meshes are fixed
-
-Game engine, ensuring fixed $dt$ and control of random numners
-
-Core or thread scheduling will not be issue unless changing scheduling policy or migrating bewtween machines that have different policies.
-
+The potential sources of non-determinism described above are summarised here along with their potential effects on simulation variance.
+Assuming deterministic hardware, there is no contribution to simulation variance due to floating point representation and any incorrect rounding errors due to byte size limits should be repeatable.
+Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
+%
+The physics setting $dt$ must be fixed and along with any actor navigation meshes and random number seeds to ensure accurate and repeatable simulation data.  
+%
+Thread scheduling should not affect the simulation determinism unless changing scheduling policy or migrating between machines that have different policies.
+However, should new and unexpected threads start during the simulation where an equal priority policy is implemented, the interruption to execution order or additional resource demand may affect timing of subsequent steps reducing the number of physics updates within a game loop and potentially give non-deterministic results.
+%
+The issue of concurrency should be minimised if good software development practice has been followed during the development of the simulation software. 
+%
+NUMA should only affect interprocessor data transfer cost and only make computations longer.
+relative times between access different caches is likely to be small although may impact on high throughput, e.g.
+HPC.
+%
+The ego vehicle controller can be considered deterministic if dynamic elements to the simulation environment are controlled, such as weather and lighting that may be on random cycles. 
+%
+% ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software.
+Actor navigation not a problem if navigation meshes are fixed.
+Game engine, ensuring fixed $dt$ and control of random numbers.
+Core or thread scheduling will not be issue unless changing scheduling policy or migrating between machines that have different policies.
 Concurrency is really down to the development of the simulation software and reputable companies should have good control over these potential issues following good software development practice.
-
-NUMA should only affect interprocessor data transfer cost, at wrst only make computations longer.
-relative times beyween access different caches is likely to be small although may impact on high throughput, e.g.
+NUMA should only affect interprocessor data transfer cost, at wrt only make computations longer.
+relative times between access different caches is likely to be small although may impact on high throughput, e.g.
 HPC.
-
 VFR - want to remove this section.
+EGO vehicle deterministic?
+Can't control library change, perception stack
+
+% ***************************************************
+%  CASE STUDY
+% ***************************************************
+\section{Empirical Investigation} \label{s:case-study}
+We present an empirical investigation to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction}.
+The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance %$\sigma^2$
+for any actor should be less than a user-defined threshold.
+
+% \todo[inline]{Introduce the Scenario - see my comment later.}
+\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+
+Autonomous vehicles operating as mobility and last mile delivery solutions will predominantly operate within urban rather than highway or motorway environments.
+To determine if the ego vehicle performs correctly, i.e.
+does not violate assertions, the precision of the simulated actor states must be repeatable to within a defined tolerance or threshold.
+
+% \todo[inline]{KIE: Move this to the Case Study section and justify the values.}
+This threshold, termed here as the \textit{deterministic threshold}, must be chosen to reflect the scale at which repeatability of simulation execution is required, see Section~\ref{s:threshold}.
+This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment.
+%
+% but a larger value may be acceptable for non-urban environments, e.g.
+motorway or highway.
+Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
+$1\mu m$.
+%
+A deterministic threshold of $1m$ is insufficient when considering the spacial resolution required to resolve scenarios such as potentially dangerous close passing of cyclists and pedestrians, and a smaller value, e.g.
+$1mm$, may result in false positives due to the simulation \textit{noise floor}\footnote{see Fig.~\ref{CarsCollisionCG25}.} Therefore a threshold of $1cm$ is chosen.
+%
+% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
+
+Results are shown in terms of the maximum deviation, $\max\sigma$, over the entire simulation history %and the deviation threshold in this investigation is $1cm$, 
+where any value higher than the deterministic threshold is considered non-deterministic.
+
+% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
+
+\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
+
+\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
+A T-junction scene was used to create a number of scenarios involving pedestrians and vehicles.
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding or avoiding with one another.
+These scenarios were created in order to identify any changes in the actor trajectories over repeated tests under different conditions and hence observe any change in \textit{simulation variance}.
+
+The tests listed in Table~\ref{TableOfExperiments} were chosen to cover a range of interactions between actor types.
+Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1m$ intervals using a PID controller.
+In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
+In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a collision callback in the game engine, see Fig~\ref{Test_a}.
+Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
+Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
+
+\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
+
+\subsection{Internal Settings}
+
+Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
+Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.
+\textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.
+\textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
+
+Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}.
+However, none of these options improved simulation variance significantly and all internal setting were set to the default values.
+Details on this previous investigation can be found on the Trustworthy Systems github\footnote{\url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}.
+
+Within the Carla development environment, there are setting for fixed $dt$ and synchronous mode which were both enabled.
+
+\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
+In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation.
+Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
+Resource utilisation figures reported here should be considered approximate values.
+% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
+
+\subsection{System Configuration and Screening}
+These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
+To ensure reliability of results, each test was repeated 1000 times.
+Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$.
+To ensure our experiment did not fall foul of some of the possible sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
+These were:
+
+\begin{itemize}[leftmargin=*]
+    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
+    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
+    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
+\end{itemize}
+
+% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
+These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
+Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
+We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
+
+% \subsection{Problem And Hypothesis}
+% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+
+% Our hypothesis is that since UE4 is a game engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
+This behaviour is expected not to hold as the level of computing stress increases.
+
+ % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
+To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
+(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
+
+% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
+In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
+
+% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
+Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
+
+% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
+
+% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
+These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
+
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) tests 1-4, (b) tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
+\end{figure}
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
+    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
+    \label{ExperimentsStressSummary}
+\end{figure}
+
+% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
+
+% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+
+% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
+Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
+
+% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+
+% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+
+% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
+
+% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
+
+% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
 
@@
 
-\subsection{Setting a Determinism  Threshold}
-The level of variance acceptable for a simulation must first be set. It serves as a threshold for deterministic execution.
-%
-We use the term \textit{determinism threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. 
-%
+\subsection{Setting a Deterministic Threshold} \label{s:threshold}
+The level of variance acceptable for a simulation must be set by the user appropriate to their needs.
+It serves as a threshold for deterministic execution.
+We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
 This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
-At some low analogue voltage the signal is interpreted as binary $0$, this may be below 0.5V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above 3.5V.
+At some low analogue voltage the signal is interpreted as binary $0$, this may be any voltage below 0.5V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above 3.5V.
 This convention enables the abstraction from the physical signal values that are analogue voltages into the digital domain with binary representation.
 
-The determinism threshold depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+The deterministic threshold that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
 In particular, this threshold must be sufficiently small to enable accurate assertion checking.
 \todo{Make sure assertion checking has previously been introduced.}
-%
 In practice, it may be necessary to determine this threshold experimentally.
-
-\todo[inline]{KIE: Move this to the Case Study section and justify the values.
-In our case study, presented in Section~\ref{s:case-study}, a threshold must be chosen for our given application which reflects the scale at which repeatability of simulation execution is required.
-This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.}
-%
-\todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.}\todo{Formal writing, no reference to the group or project, please.}
 
@@
 In Carla, further settings can be applied to allow synchronous communication to external controllers to ensure no sensor data are passed out of order to the simulator~\cite{carla_sim_config}.
-
-Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
-Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.
-\textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.
-\textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}.
-However, none of these options improved simulation variance significantly beyond the default settings, see our previous investigation\footnote{\url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}.
 
@@
 
-\subsection{Output and Analysis}
+\subsection{Results and Analysis}\label{s:results}
 \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
@@
 % Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
-
-% ***************************************************
-%  CASE STUDY
-% ***************************************************
-\section{Case Study} \label{s:case-study}\todo[inline]{Put this section before Methodology.}
-We present a case study to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction}.
-The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance $\sigma^2$ for any actor should be less than a user-defined threshold.
-Results are shown in terms of the maximum deviation ($\sigma$) and the deviation threshold in our case is $1cm$, where any value higher than this is considered non-deterministic.
-
-\todo[inline]{Introduce the Scenario - see my comment later.}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
-
-\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
-In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation.
-Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
-Resource utilisation figures reported here should be considered approximate values.
-% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
-
-\subsection{System Configuration and Screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
-To ensure reliability of results, each test was repeated 1000 times.
-Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$.
-To ensure our experiment did not fall foul of some of the possible sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
-These were:
-
-\begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
-    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
-\end{itemize}
-
-% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
-These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
-
-% \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
-
-% Our hypothesis is that since UE4 is a game engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
-This behaviour is expected not to hold as the level of computing stress increases.
-
-% \subsection{Tests Description And Technicalities}\label{TestsDescriptionAndTechnicalities}
-\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}\todo[inline]{This should go first.}
-\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the }
-The tests listed in Table~\ref{TableOfExperiments} were chosen to cover a range of interactions between actor types.
-Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1m$ intervals using a PID controller.
-In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
-In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a collision callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
-Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
-
- % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
-To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
-(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
-
-% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
-
-% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
-
-% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
-
-% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) tests 1-4, (b) tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
-\end{figure}
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
-    \label{ExperimentsStressSummary}
-\end{figure}
-
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
 

### 2020-04-24 paper/cav_determinisim_main.tex
GC updates, change case study location
GC fig5 caption update

@@
 \centering
-\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
+\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoopv2.pdf}
 \caption{Game engine loop block diagram~\cite{GameProgPatternsBook} \todo[inline]{AG: remove the clock symbol, possibly remove the ticks chart as this is confusing given they are not ticks but event markers.}}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V3.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
         \caption{}
@@
     \end{subfigure}
-    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.}
+    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.
+The simulation noise floor is shown in (a) which is the empirical lower limit of deviation in this configuration.}
 \end{figure*}

### 2020-04-27 paper/cav_determinisim_main.tex
move results to investigation section

@@
 %
-A deterministic threshold of $1m$ is insufficient when considering the spacial resolution required to resolve scenarios such as potentially dangerous close passing of cyclists and pedestrians, and a smaller value, e.g.
-$1mm$, may result in false positives due to the simulation \textit{noise floor}\footnote{see Fig.~\ref{CarsCollisionCG25}.} Therefore a threshold of $1cm$ is chosen.
+A deterministic threshold of $1m$ is insufficient when considering the spacial resolution required to resolve scenarios such as potentially dangerous close passing of cyclists and pedestrians, and a very small value, e.g.
+$1\times10^{-12}$, may result in false positives due to the simulation \textit{noise floor}\footnote{see Fig.~\ref{CarsCollisionCG25}.} Therefore a threshold of $1cm$ is chosen which is sufficient for assertion checking in urban environments.
 %
@@
 A T-junction scene was used to create a number of scenarios involving pedestrians and vehicles.
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding or avoiding with one another.
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding or avoiding with other actors.
 These scenarios were created in order to identify any changes in the actor trajectories over repeated tests under different conditions and hence observe any change in \textit{simulation variance}.
@@
 
+\subsection{Experiment Description}\label{s:results}
+\todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
+For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time.
+Repeated traces are compared at the same time index to provide a value for the variance of actor $a$ with respect to time $t$, $\sigma_a(t)$.
+The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation,
+\begin{equation}
+\max_{a,t}\sigma_a(t), 
+\end{equation}
+herein referred to as $\max\sigma$ for brevity.
+% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
+% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
+The $\max\sigma$ value was analysed against the different scenarios and settings that were identified as potential sources of simulation variance, and compared against the \textit{deterministic threshold} to indicate if the simulation was deterministic or not.
+%
+% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
+\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
+
+% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
+Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
+
+% This whole process is repeated for various internal and external settings (if needed). 
+% Then various plots can be created to draw the engine's simulation variance. 
+% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
+
 \subsection{Internal Settings}
@@
 Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}.
-However, none of these options improved simulation variance significantly and all internal setting were set to the default values.
+
+\todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
+
+An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
+The results were compared to a baseline of the default engine settings.
+%
+However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
 Details on this previous investigation can be found on the Trustworthy Systems github\footnote{\url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}.
 
-Within the Carla development environment, there are setting for fixed $dt$ and synchronous mode which were both enabled.
-
 \subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
-In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation.
+
+After discovering that internal engine settings could not achieve a simulation variance that met the required \textit{deterministic threshold}, settings external to the game engine were explored. 
+%
+Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
+%
+For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
+This hypothesis was supported by our initial work performed on the unreal engine\footnote{see footnote 14} and was explored more fully in this work.
+%
+To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation.
+%
+% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
 Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
@@
 \subsection{System Configuration and Screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
+These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
 To ensure reliability of results, each test was repeated 1000 times.
 Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$.
-To ensure our experiment did not fall foul of some of the possible sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
+To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
 These were:
@@
 
+% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
+
+% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+
+% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
+Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
+
+% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+
+% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+
+% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
+
+% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
+
+% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
+
+\begin{table*}[b]
+\centering
+\begin{tabular}{clclccc}
+\toprule
+Test & Actors 					 	  & Collision 	 & Collision Type 			 & $n$  & $\max\sigma$ (m) & $\max\sigma$ (m) \\ 
+	 & 								  & 			 & 							 &	    & (unrestricted)  & (restricted)	\\ \midrule
+1    & Two vehicles                   & No  		 & N/A 			  			 & 1000 & 0.03 & $7.0{\times}10^{-3}$ \\
+2    & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 1000 & 0.31 & $9.8{\times}10^{-3}$ \\
+3    & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 1000 & 0.07 & $5.2{\times}10^{-4}$ \\
+4    & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 1000 & 0.59 & $1.5{\times}10^{-12}$ \\
+5    & Two pedestrians                & No  		 & N/A 			  			 & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
+6    & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
+\bottomrule
+\end{tabular}
+\caption{A description of the test scenarios showing the test number, the actors included, if a collision occurred and between which actors.
+Where $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
+The term \textit{unrestricted} refers to an unrestricted subset of the results and allowing system states of any resource utilisation whereas \textit{restricted} employs results filtering to pre-collision and limits system resource utilisation levels to 75\%.}
+\label{TableOfExperiments}
+\end{table*}
+
+% ***************************************************
+%  RESULTS
+% ***************************************************
+\subsection{Results and Discussion}\label{ResultsSection}
+The summary of all the tests are shown in the final two columns of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels.
+From these  results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels.
+However, for other scenarios involving vehicles or a mixture of actors types, the resource utilisation level has a significant impact on \textit{simulation variance}.
+
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU).
+In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.\todo[inline]{Where are the verification requirements listed?
+They should be part of the scenario description.} A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s).}
+
+However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation.
+Examining these specific results, tests 2 \& 4, in more detail as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision.
+This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism.
+\todo{Why ``not reliable''.}
+
+The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
+In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}.
+There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as described above.
+Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
+Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delay \textit{even though this vehicle was not involved in the collision}.
+This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  false-positive assertion violations.\todo[inline]{Make sure that assertion checking has been introduced before.}
+
+% \textbf{Strange gap post-collision for few seconds}
+
 \begin{figure}[!t]
@@
 
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was left alone and not used for any other tasks while the runs were executed.
+% \textbf{Should we rename 'test' to 'scenario???'}
+
+\todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
+
+% Scenarios that include both collisions and vehicle actors were not deterministic.
+However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
+
+% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
+Note the dotted red line in the plot shows the millimeter (mm) level.
+Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
+
+\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
+This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
+Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
+
+\begin{figure*}[h]
+    \centering
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
+        \caption{}
+        \label{CarsCollisionCG25}
+    \end{subfigure}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
+        \caption{}
+        \label{CarsCollisionPrePost}
+    \end{subfigure}
+    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.
+The simulation noise floor is shown in (a) which is the empirical lower limit of deviation in this configuration.}
+\end{figure*}
+
+\begin{figure*}[h]
+    \centering
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
+        \caption{}
+        \label{CarsPeopleCollsionsCG25}
+    \end{subfigure}
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV3.pdf}
+        \caption{}
+        \label{CarsPeopleCollisionPrePost}
+    \end{subfigure}
+    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation.}
+\end{figure*}
+
+% First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
+Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
+with in the centimeter level).
+This clearly shows how the load on a computer (i.e.
+stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
+This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
+
+% Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
+with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
+
+% Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
+We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
+
+% Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
+Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
+
+% Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
+However, in V\&V of AV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
+Thus, the post collision effect should not be of much concern in the context of AVs.
+
+% It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
+The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
+Thus failing to stay below our red line and the test should be terminated.
+In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
+left alone and not used for any other tasks while the runs were executed.
 
@@
 %
-To replicate and control the high computational loads that may be experienced in high performance simulations, software that artificially utilises resources is executed alongside the simulation on both central and graphical processing units.
-\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW.}
 % The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
@@
 
-\begin{table*}[b]
-\centering
-\begin{tabular}{clclccc}
-\toprule
-Test & Actors 					 	  & Collision 	 & Collision Type 			 & $n$  & $\max\sigma$ (m) & $\max\sigma$ (m) \\ 
-	 & 								  & 			 & 							 &	    & (unrestricted)  & (restricted)	\\ \midrule
-1    & Two vehicles                   & No  		 & N/A 			  			 & 1000 & 0.03 & $7.0{\times}10^{-3}$ \\
-2    & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 1000 & 0.31 & $9.8{\times}10^{-3}$ \\
-3    & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 1000 & 0.07 & $5.2{\times}10^{-4}$ \\
-4    & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 1000 & 0.59 & $1.5{\times}10^{-12}$ \\
-5    & Two pedestrians                & No  		 & N/A 			  			 & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-6    & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-\bottomrule
-\end{tabular}
-\caption{A description of the test scenarios showing the test number, the actors included and if a collision occurred and between which actors.
-Where $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
-The term \textit{unrestricted} refers to an unrestricted subset of the results and allowing system states of any resource utilisation whereas \textit{restricted} employs results filtering to pre-collision and limits system resource utilisation levels to 75\%.}
-\label{TableOfExperiments}
-\end{table*}
-
 \subsection{System Resources}\todo[inline]{This should be ``External Settings'' - from flow diagram?}
 To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will  contribute towards increased \textit{simulation variance}.
 CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system.
@@
 
-\subsection{Results and Analysis}\label{s:results}
-\todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-The output for each experiment can be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests.
-The maximum variance is chosen to indicate the worst case rather than an average which could hide outlier data points when a high number of repeated tests are performed.
-This value can be analysed against the different scenarios or settings that have been identified as potential sources of simulation variance.
-These values can then be assessed against the user-defined \textit{deterministic threshold} to determine if the simulation is deterministic or not.
-
-% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
-Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
-
-% This whole process is repeated for various internal and external settings (if needed). 
-% Then various plots can be created to draw the engine's simulation variance. 
-% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
-
-% ***************************************************
-%  RESULTS
-% ***************************************************
-\subsection{Results and Discussion}\label{ResultsSection}
-The summary of all the tests are shown in the final column of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels.
-From these  results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels.
-However, for other scenarios involving vehicles or a mixture of actors types, the resource utilisation level has a significant impact on \textit{simulation variance}.
-
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU).
-In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.\todo[inline]{Where are the verification requirements listed?
-They should be part of the scenario description.} A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s).}
-
-However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation.
-Examining these specific results, tests 2 \& 4, in more detail as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision.
-This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism.
-\todo{Why ``not reliable''.}
-
-The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}.
-There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as described above.
-Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
-Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  false-positive assertion violations.\todo[inline]{Make sure that assertion checking has been introduced before.}
-
-% \textbf{Strange gap post-collision for few seconds}
-
-% \textbf{Should we rename 'test' to 'scenario???'}
-
-\todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
-
-% Scenarios that include both collisions and vehicle actors were not deterministic.
-However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
-
-% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
-Note the dotted red line in the plot shows the millimeter (mm) level.
-Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
-
-\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
-This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
-Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
-
-\begin{figure*}[h]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
-        \caption{}
-        \label{CarsCollisionCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
-        \caption{}
-        \label{CarsCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.
-The simulation noise floor is shown in (a) which is the empirical lower limit of deviation in this configuration.}
-\end{figure*}
-
-\begin{figure*}[h]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
-        \caption{}
-        \label{CarsPeopleCollsionsCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV3.pdf}
-        \caption{}
-        \label{CarsPeopleCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation.}
-\end{figure*}
-
-% First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
-with in the centimeter level).
-This clearly shows how the load on a computer (i.e.
-stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
-This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
-
-% Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
-with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
-
-% Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
-We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
-
-% Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
-Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
-
-% Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
-However, in V\&V of AV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
-Thus, the post collision effect should not be of much concern in the context of AVs.
-
-% It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
-The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
-Thus failing to stay below our red line and the test should be terminated.
-In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
-
 % ***************************************************

### 2020-04-27 paper/cav_determinisim_main.tex
move results to investigation section

@@
 % ***************************************************
-%  CASE STUDY
+%  Empirical Investigation
 % ***************************************************
@@
 
+% ***************************************************
+%  Scenario Description
+% ***************************************************
 \subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
 A T-junction scene was used to create a number of scenarios involving pedestrians and vehicles. 
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding or avoiding with other actors.
+%
 These scenarios were created in order to identify any changes in the actor trajectories over repeated tests under different conditions and hence observe any change in \textit{simulation variance}.
-
-The tests listed in Table~\ref{TableOfExperiments} were chosen to cover a range of interactions between actor types.
+%
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding with or avoiding other actors.
+
+% ***************************************************
+%  Actor Collisions
+% ***************************************************
+\subsection{Actor Collisions}\label{S:Actor_Collisions}
+Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Collision/Overview/index.html}}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}.
+%
+Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.
+
+The tests are listed in Table~\ref{TableOfExperiments} and were chosen to cover a range of interactions between actor types. 
+%
 Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
@@
 
+% ***************************************************
+%  Experiment Description
+% ***************************************************
 \subsection{Experiment Description}\label{s:results}
@@
 The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation,
-\begin{equation}
+\begin{equation} \label{eq:max_sigma}
 \max_{a,t}\sigma_a(t), 
@@
 
+% ***************************************************
+%  Internal Settings
+% ***************************************************
 \subsection{Internal Settings}
@@
 However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
-Details on this previous investigation can be found on the Trustworthy Systems github\footnote{\url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}.
-
-\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
-
-After discovering that internal engine settings could not achieve a simulation variance that met the required \textit{deterministic threshold}, settings external to the game engine were explored. 
-%
-Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
-%
-For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
-This hypothesis was supported by our initial work performed on the unreal engine\footnote{see footnote 14} and was explored more fully in this work.
-%
-To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation.
-%
-% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
-Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
-Resource utilisation figures reported here should be considered approximate values.
-% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
-
-\subsection{System Configuration and Screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
-To ensure reliability of results, each test was repeated 1000 times.
-Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$.
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
-These were:
-
-\begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
-    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
-\end{itemize}
-
-% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
-These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
-
-% \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
-
-% Our hypothesis is that since UE4 is a game engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
-This behaviour is expected not to hold as the level of computing stress increases.
-
- % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
-To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
-(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
-
-% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
-
-% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
-
-% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
-
-% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) tests 1-4, (b) tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
-\end{figure}
-
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
+Details on this previous investigation can be found on the Trustworthy Systems github~\cite{TSLUnrealEngineTesting}.
 
@@
 % ***************************************************
-%  RESULTS
+%  External Settings
+% ***************************************************
+\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
+
+After discovering that internal engine settings could not achieve a simulation variance that met the required \textit{deterministic threshold}, settings external to the game engine were explored. 
+%
+Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
+%
+For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
+This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the Carla platform.
+
+To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation.
+%
+% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
+Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
+Resource utilisation figures reported here should be considered approximate values.
+% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
+
+% ***************************************************
+%  System Configuration and Screening
+% ***************************************************
+\subsection{System Configuration and Screening}
+These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
+To ensure reliability of results, each test was repeated 1000 times.
+Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$.
+To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
+These were:
+
+\begin{itemize}[leftmargin=*]
+    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
+    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
+    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
+\end{itemize}
+
+% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
+These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
+Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
+We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
+
+% \subsection{Problem And Hypothesis}
+% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+
+% Our hypothesis is that since UE4 is a game engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
+This behaviour is expected not to hold as the level of computing stress increases.
+
+ % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
+To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
+(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
+
+% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
+In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
+
+% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
+Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
+
+% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
+
+% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
+These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
+
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) tests 1-4, (b) tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
+\end{figure}
+
+% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
+
+% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+
+% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
+Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
+
+% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+
+% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+
+% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
+
+% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
+
+% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
+
+% ***************************************************
+%  Results and Discussion
 % ***************************************************
 \subsection{Results and Discussion}\label{ResultsSection}
-The summary of all the tests are shown in the final two columns of Table~\ref{TableOfExperiments} where $\max\sigma$ is the maximum deviation across all resource utilisation levels.
-From these  results it is clear that scenarios with only pedestrian actors are those that are consistently deterministic over all resource utilisation levels.
-However, for other scenarios involving vehicles or a mixture of actors types, the resource utilisation level has a significant impact on \textit{simulation variance}.
-
+A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted) where the value reported is the maximum across all resource utilisation levels, i.e.
+the worst case for a given scenario.
+%
+From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) are those that are consistently deterministic over all resource utilisation levels with or without a collision.
+However, for the other scenarios involving vehicles or a mixture of actors types, the all breach the deterministic threshold that has been set, with some deviation in actor path as large as 59cm.
+
+Resource utilisation level was found to have a significant impact on \textit{simulation variance}.
+%
 Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU).
-In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.\todo[inline]{Where are the verification requirements listed?
-They should be part of the scenario description.} A general pattern in the results indicates that some scenarios fail to produce deterministic results consistently against any resource utilisation level and that some cases only fail at utilisation levels above 50\%.\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s).}
-
-However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation.
-Examining these specific results, tests 2 \& 4, in more detail as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision.
-This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism.
-\todo{Why ``not reliable''.}
-
-The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}.
-There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as described above.
-Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
-Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  false-positive assertion violations.\todo[inline]{Make sure that assertion checking has been introduced before.}
-
-% \textbf{Strange gap post-collision for few seconds}
+In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
+%
+\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
+%
+A general pattern in the results indicates that; some scenarios fail to produce deterministic results consistently against any resource utilisation level (see Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed $cm$ threshold), some are consistently below the threshold (test ID 5 \& 6) and that some cases only fail at higher utilisation levels, breaching the threshold above 75\% resource utilisation (Test ID 1 \& 3).
+
+\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
 
@@
 \end{figure}
-
-% \textbf{Should we rename 'test' to 'scenario???'}
-
-\todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
-
-% Scenarios that include both collisions and vehicle actors were not deterministic.
-However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
-
-% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
-Note the dotted red line in the plot shows the millimeter (mm) level.
-Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
-
-\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
-This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
-Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
 
@@
     \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.
-The simulation noise floor is shown in (a) which is the empirical lower limit of deviation in this configuration.}
+The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
 \end{figure*}
+
+However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation time.
+%
+Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision.
+This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism.
+\todo{Why ``not reliable''.}
 
@@
 \end{figure*}
+
+The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
+In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}.
+There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as described above.
+Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
+Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delay \textit{even though this vehicle was not involved in the collision}.
+This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  false-positive assertion violations.\todo[inline]{Make sure that assertion checking has been introduced before.}
+
+% \textbf{Strange gap post-collision for few seconds}
+
+% \textbf{Should we rename 'test' to 'scenario???'}
+
+\todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
+
+% Scenarios that include both collisions and vehicle actors were not deterministic.
+However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
+
+% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
+Note the dotted red line in the plot shows the millimeter (mm) level.
+Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
+This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
+
+\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
+This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
+Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
 

### 2020-04-28 paper/cav_determinisim_main.tex
changes to method section

@@
 \usepackage{todonotes}
+\usepackage{siunitx}
 
@@
 When the frame is rendered the game loop cycle returns to processing inputs.
+And a good description of the interplay between the physics and render cycles are given here~\cite{JohnAustinUnity}.
 
@@
 This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
-
-\begin{figure*}[b]
-    \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
-    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation environment.}
-
-    \label{method_diagram}
-\end{figure*}
 
@@
 %
-A deterministic threshold of $1m$ is insufficient when considering the spacial resolution required to resolve scenarios such as potentially dangerous close passing of cyclists and pedestrians, and a very small value, e.g.
-$1\times10^{-12}$, may result in false positives due to the simulation \textit{noise floor}\footnote{see Fig.~\ref{CarsCollisionCG25}.} Therefore a threshold of $1cm$ is chosen which is sufficient for assertion checking in urban environments.
+A deterministic threshold of $1$m is insufficient when considering the spacial resolution required to resolve scenarios such as potentially dangerous close passing of cyclists and pedestrians, and a very small value, e.g.
+$1\times10^{-12}$, may result in false positives due to the simulation \textit{noise floor}\footnote{see Fig.~\ref{CarsCollisionCG25}.} Therefore a threshold of $1$cm is chosen which is sufficient for assertion checking in urban environments.
 %
@@
 Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1m$ intervals using a PID controller.
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m intervals using a PID controller.
 In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
 In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a collision callback in the game engine, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1m$ intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
+Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
 Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
@@
 % ***************************************************
-\subsection{Experiment Description}\label{s:results}
+\subsection{Experiment Description}\label{s:Experiment_Description}
 \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
@@
 Where $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
-The term \textit{unrestricted} refers to an unrestricted subset of the results and allowing system states of any resource utilisation whereas \textit{restricted} employs results filtering to pre-collision and limits system resource utilisation levels to 75\%.}
+The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation, whereas \textit{restricted} restricts simulation time to pre-collision and limits system resource utilisation levels to 75\%.}
 \label{TableOfExperiments}
@@
 % In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
-Resource utilisation was artificially increased for both CPU and GPU devices (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
+Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0-95\% (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
 Resource utilisation figures reported here should be considered approximate values.
@@
 % ***************************************************
-\subsection{System Configuration and Screening}
+\subsection{System Configuration and Screening}\label{screening}
 These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
+Operating systems was Linux Ubuntu 18.04.4 LTS.
 To ensure reliability of results, each test was repeated 1000 times.
-Tests were carried out in Carla using synchronous mode with a fixed $dt$ of $0.05s$.
+Tests were carried out in Carla (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
 To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
@@
     \end{subfigure}
-    \caption{Schematic of test scenarios for (a) tests 1-4, (b) tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
 \end{figure}
@@
 \subsection{Results and Discussion}\label{ResultsSection}
-A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted) where the value reported is the maximum across all resource utilisation levels, i.e.
+A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted), where the value reported is the maximum deviation across all resource utilisation levels, i.e.
 the worst case for a given scenario.
 %
-From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) are those that are consistently deterministic over all resource utilisation levels with or without a collision.
-However, for the other scenarios involving vehicles or a mixture of actors types, the all breach the deterministic threshold that has been set, with some deviation in actor path as large as 59cm.
+From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) are consistently deterministic over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the deterministic threshold that has been set, with some deviation in actor path as large as $59$cm.
+%
+Clearly a deviation in results of such a large amount is unacceptable if simulation is to be used as a reliable verification tool.
 
@@
 %
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (CPU \& GPU).
-In this figure, anything above the $1cm$ threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
+In this figure, anything above the $1$cm threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
 %
@@
 %
-A general pattern in the results indicates that; some scenarios fail to produce deterministic results consistently against any resource utilisation level (see Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed $cm$ threshold), some are consistently below the threshold (test ID 5 \& 6) and that some cases only fail at higher utilisation levels, breaching the threshold above 75\% resource utilisation (Test ID 1 \& 3).
-
-\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
 
@@
 \end{figure}
+
+A general pattern in the results indicates that; some scenarios fail to produce deterministic results consistently against any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm threshold), some are consistently below the threshold (Test ID 5 \& 6 with pedestrians only), and that some cases only fail at higher utilisation levels, breaching the threshold above 75\% resource utilisation (Test ID 1 \& 3).
+
+\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
 
@@
 Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results cannot be relied upon due to the observed non-determinism.
-\todo{Why ``not reliable''.}
+This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-deterministic.
+% \todo{Why ``not reliable''.}
 
@@
 The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (test 4), there is a collision between Car ID 2 (solid line) and Ped ID 3 (dot dash line) which occurs at a simulation time of approximately $6s$ and the other vehicle, Car ID 1, is \textit{not involved in the collision}.
+In this scenario (test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
 There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as described above.
 Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
-Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5s$ delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  false-positive assertion violations.\todo[inline]{Make sure that assertion checking has been introduced before.}
+Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
+This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  erroneous assertion checking.
+
+% \todo[inline]{Make sure that assertion checking has been introduced before.}
 
@@
 
-\todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
+% \todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
 
@@
 This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
+%
+\begin{figure*}[b]
+    \centering
+    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
+    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}
+    \label{method_diagram}
+\end{figure*}
 
@@
 Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
+
+The main findings of this work suggest a working practice for this specific hardware configuration that would minimise the non-deterministic effects observed in this investigation.
+By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the deterministic threshold of $1$cm is achievable as shown in the restricted column in Table.~\ref{TableOfExperiments}.
+
+\subsection{Investigation Summary} \label{s:empirical_summary}
+
+% \todo[inline]{GC-need to reference the 'restricted' set of results here too}
+
+This empirical investigation has highlighted the shortcomings of using a games engine for simulation based verification and advice for best working practice.
+However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly.
+Therefore we present a general methodology that practitioners can follow to find the deterministic domains of operation for any hardware configuration.
 
@@
 \section{Methodology} \label{s:methodology}
-In this section a method for determining the variance of a simulation environment, termed \textit{simulation variance} is presented.
-We have identified a number of experimental parameters that may influence the simulation variance of a game engine, see Fig.~\ref{method_diagram}.
-These need to be considered at different stages in the process.
-%
-The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
+In this section a method for determining the variance of simulated actor path trajectories, termed \textit{simulation variance}, and resolving the deterministic operational domains of the simulation are presented.
+%
+% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
+This methodology has been refined several times over the course of our investigations into both Unreal Engine and Carla and is presented here as a work flow, see Fig.~\ref{method_diagram}.
+%These need to be considered at different stages in the process.
+%
+\subsection{Design Experiment}\label{s:design_experiment}
+
+\subsubsection{Simulated Actors} \label{s:actors}
+
+Different actor types may be handled differntly by the game engine and our investigation indicated that Carla pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
+However the introduction of Carla vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
+As such, all actors that could be included in any future simulation should be tested, including any non-standard Carla or bespoke actors.
+%
+Actor paths should hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
+
+\subsubsection{Setting a Deterministic Threshold} \label{s:threshold}
+The verification engineer should set the \textit{deterministic threshold} based on their specific requirements which refers to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
+This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
+At some low analogue voltage the signal is interpreted as binary $0$, this may be any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
+This convention enables the abstraction from the physical signal values that are analogue voltages into the digital domain with binary representation.
+%
+The deterministic threshold that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+In particular, this threshold must be sufficiently small to enable accurate assertion checking.
+In our empirical investigation we set a threshold of $1$cm which is sufficient for urban scenario assertion checking.
+In practice, it may be necessary to determine this threshold experimentally.
+
+% \todo[inline]{Make sure assertion checking has previously been introduced.}
+
+% The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
 These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
-\todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''.} 
-%
+
+% \todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''. GC change callback to 'event hit'}
+
 % The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
@@
 
-\todo[inline]{We may need to introduce the Case Study first, then derive the general Methodology from it.
+% \todo[inline]{We may need to introduce the Case Study first, then derive the general Methodology from it.
 This way, we can explain the requirements of the case study, how things were set up, what was observed, how the methodology was refined, several iterations even, etc.}
 
-\subsection{Setting a Deterministic Threshold} \label{s:threshold}
-The level of variance acceptable for a simulation must be set by the user appropriate to their needs.
-It serves as a threshold for deterministic execution.
-We use the term \textit{deterministic threshold} to refer to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
-This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
-At some low analogue voltage the signal is interpreted as binary $0$, this may be any voltage below 0.5V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above 3.5V.
-This convention enables the abstraction from the physical signal values that are analogue voltages into the digital domain with binary representation.
-
-The deterministic threshold that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-In particular, this threshold must be sufficiently small to enable accurate assertion checking.
-\todo{Make sure assertion checking has previously been introduced.}
-In practice, it may be necessary to determine this threshold experimentally.
+\subsubsection{Simulation Length} 
+The simulation time should be sufficient to record interactions between actors but not so long that 100's or 1000's of repeated tests take a considerable time to complete.
+In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
 
@@
 
-\subsection{Scenario Description}\todo[inline]{Select section titles to match the flow diagram.
+\subsection{Internal Settings}
+The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
+If using Carla a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}}.
+In Carla, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
+%
+The use of random numbers must be controlled through the use of fixed seeds which might be used to control variations of background effects, e.g.
+weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment.
+%
+Actors that navigate through the environment should use a fixed navigation mesh.
+
+The version number of the Carla and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}, so ensuring consistent version number throughout testing is also important.
+
+% \todo[inline]{Select section titles to match the flow diagram.
 In each of these sections, go through the matching phase in the flow diagram, explaining each point listed there and justify recommendations.}
-The experiment should be a short\todo{What does short mean?
-Quantify.} scenario that can be easily repeated with the same initial conditions.
+% The experiment should be a short\todo[inline]{What does short mean? Quantify.} scenario that can be easily repeated with the same initial conditions.
+
+\subsection{External Settings}
+
+\subsubsection{Resource Utilisation}
+% To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
+CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program can be used spawn workers on any number of cores or virtual threads on a system.
+This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
+For GPU utilisation \texttt{gpu-burn} was employed using the \texttt{fur test} where different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
+
+Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU respectively.
+These values can also be written into the data logging for completeness.
+
+% \todo[inline]{Again, this sounds as if having the case study first makes more sense. That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
+
+% \todo[inline]{This should be ``External Settings'' - from flow diagram?}
+
+\subsubsection{Memory Testing}
+
+See~\ref{screening}
+\texttt{memtest86}\
+\texttt{cuda\_memtest}
+
+\subsubsection{Sheduling Policy}
+-thread sceduling
+-reduce uncecessary tasks
+
+\subsubsection{NUMA Control}
+
+\subsubsection{Ego Vehicle Controller}
+
+\subsection{Execution}
+
+\subsubsection{Sample Size}
+
+Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
+While executing 100 repeats may seem sufficient, in this case the sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
+To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
+
+\subsubsection{Data Logging}
+
 Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
@@
 
-Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo{Explain what this means.} Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
-While executing 100 repeats may seem sufficient, in this case the sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
-To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
-
-Game engine actors can be used to simulate the AV and other vehicles as well as pedestrians.
+Regular time for easier results analysis.
+
+Link to TSL github for our data logging code?
+
+\subsection{Analysis}
+
+% Game engine actors can be used to simulate the AV and other vehicles as well as pedestrians.
 For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
 
-\todo[inline]{This is mixing our experiments (which are part of the case study) with the methodology.
+% \todo[inline]{This is mixing our experiments (which are part of the case study) with the methodology.
 It is probably best to first introduce the case study and then derive the methodology from it, rather than mixing both.} In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
@@
 
-Randomisation must be fully controlled by fixing the seeds for any random number generation that is used during the experiment.
+% Randomisation must be fully controlled by fixing the seeds for any random number generation that is used during the experiment.
 It is important to identify and control any external sources that may use randomisation, e.g.\ external vehicle controllers or other clients connected to the simulation environment.
@@
 
-\subsection{Internal settings}\todo[inline]{Systematically go through the points listed in the flow diagram, explain and justify.}
-Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
-If using Carla a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}} and a good description of the interplay between physics and render clock is given here~\cite{JohnAustinUnity}.
-In Carla, further settings can be applied to allow synchronous communication to external controllers to ensure no sensor data are passed out of order to the simulator~\cite{carla_sim_config}.
+% \subsection{Internal settings}\todo[inline]{Systematically go through the points listed in the flow diagram, explain and justify.}
+% Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
 
@@
 
-\subsection{System Resources}\todo[inline]{This should be ``External Settings'' - from flow diagram?}
-To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program that can be used spawn workers on any number of cores or virtual threads on a system.
-This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
-For GPU utilisation \texttt{gpu-burn} was employed using the fur test at different resolutions and multiple instances can be used to generate high utilisation levels~\cite{GPU_stress}.\todo[inline]{Again, this sounds as if having the case study first makes more sense.
-That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
-
 % The user should also monitor and record system CPU and GPU utilisation rates.

### 2020-04-29 paper/cav_determinisim_main.tex
changes to method

@@
 %
+\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
+%
 For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
@@
 Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
-However, several challenges remain; one of these is determinism.
+%
+\todo[inline]{GC Replace: However, several challenges remain; one of these is determinism.
+with: However, game engines are designed principally for user experience whereas the requirements for CAV verification is not user experience but determinism.}
 %
@@
 % ***************************************************
-\subsection{System Configuration and Screening}\label{screening}
+\subsection{System Configuration and Screening}\label{s:screening}
 These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
@@
 
-The main findings of this work suggest a working practice for this specific hardware configuration that would minimise the non-deterministic effects observed in this investigation.
+The main findings of this work suggest a working practice that would minimise the non-deterministic effects observed in this investigation.
 By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the deterministic threshold of $1$cm is achievable as shown in the restricted column in Table.~\ref{TableOfExperiments}.
@@
 % \todo[inline]{GC-need to reference the 'restricted' set of results here too}
+
+\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
+The results indicate that actor type, resource utilisation and collision events all affect the determinism in the Carla simulator.
+Collision events may utilise more computational resources than simulations without such events, i.e.
+the physics loop takes longer to process as there are complex physical interactions to resolve.
+This may coincide with the second observation that higher resource levels also lead to non-deterministic results.
+We hypothesise that both these observations are due to improper management of thread scheduling within the engine.
+When resources are occupied there may be a delay in processing critical calculations for the simulator, which may arrive out of an expected sequence or the engine compensates for this delay by reducing the number of physics loops that can be performed within a render cycle.
+(is this 2nd one true, or does carla ignore render requests?)}
 
@@
 % To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
+The resources available to the simulator have been shown to have a significant effect on the determinism of simulated vehicles and exploring this as a variable the analysis is required. 
 CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program can be used spawn workers on any number of cores or virtual threads on a system.
@@
 For GPU utilisation \texttt{gpu-burn} was employed using the \texttt{fur test} where different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
-
 Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU respectively.
@@
 
+Alternatively, in place of artificial resource utilisation multiple instances of the simulation could be executed simultaneously. However, the granularity of control with this approach may be reduced.
+
 % \todo[inline]{Again, this sounds as if having the case study first makes more sense. That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
@@
 \subsubsection{Memory Testing}
-
-See~\ref{screening}
-\texttt{memtest86}\
-\texttt{cuda\_memtest}
-
-\subsubsection{Sheduling Policy}
--thread sceduling
--reduce uncecessary tasks
+Prior to experimental execution the system hardware should be tested for memory conformity and to ensure no single bit errors are occurring, see~\ref{s:screening}.
+For mainboard memory \texttt{memtest86}\ can be used on most platforms to run a series of pre-defined memory test patterns.
+This memory testing software can also be used for ECC enabled hardware.
+Similarly to test memory on Nvidia based graphical adaptors \texttt{cuda\_memtest} can be used to ensure no memory errors exist.
+
+\subsubsection{Scheduling Policy}
+We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
+However, fine control over the scheduling policy and thread execution order may be non-trivial.
+Such policies are defined by the operating system and may execute threads with equal priority.
+In such a case, all tasks non-essential to the simulator should be terminated to prevent interruption to the simulator.
+
+% If the policy or operating system cannot be changed then s
+Setting the simulator program to a higher priority may help to alleviate conflicting tasks that cannot be shutdown, such as \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}} or \texttt{nice} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.
 
 \subsubsection{NUMA Control}
+Control over Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multiprocessors with shared memory.
+This control allows the simulator program to be fixed on a single core reducing memory access time.
+%
+Initial screening tests done with NUMA control gave moderate improvements in simulation variance of the order of a few percent, see Section~\ref{s:screening}.
+This control may assist if simulation variance is borderline to the deterministic threshold but not seen as essential.
 
 \subsubsection{Ego Vehicle Controller}
+No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
 
@@
 \subsubsection{Sample Size}
-
-Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} Initial testing\footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}} indicated an actor path deviation of $1\times10^{-13} cm$ for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10cm$.
-While executing 100 repeats may seem sufficient, in this case the sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
-To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
+% Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} 
+% Histogram to show how sampling with position binning varies?
+% \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
+Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
+While executing 100 repeats may seem sufficient, this sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
+It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the deterministic threshold.
+% To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
 
 \subsubsection{Data Logging}
-
 Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
-For example, if 100 tests are executed and the maximum deviation in actor path is $<1cm$, then this scenario could be interpreted as being deterministic above this scale.
-
-Regular time for easier results analysis.
-
-Link to TSL github for our data logging code?
+Additional information can also be logged such as the CPU and GPU utilisation levels.
+For the subsequent analysis, the actor should have an identifier and the repeat number and simulation time should also be captured.
+% The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
+For example, if 100 tests are executed and the maximum deviation in actor path is $<1$cm, then this scenario could be interpreted as being deterministic above this scale.
 
 \subsection{Analysis}
+The maximum value of simulation variance over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control for deterministic operation.
+Finding these boundaries will identify the deterministic operational domain for the user specific hardware.
 

### 2020-04-29 paper/cav_determinisim_main.tex
method flowchart figure

@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagram.pdf}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v4.pdf}
     \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}
@@
 
-\subsubsection{Simulated Actors} \label{s:actors}
+\subsubsection{Actors} \label{s:actors}
 
@@
 %
+\subsubsection{Trajectories and Collisions} 
 Actor paths should hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
@@
 
-\subsection{Internal Settings}
+\subsection{Simulator Settings}
 The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
@@
 
-\subsubsection{Scheduling Policy}
+\subsubsection{Scheduling}
 We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.

### 2020-05-12 paper/cav_determinisim_main.tex
added new method diagram
changes  up to 2B

@@
 Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
+\todo[inline]{GC: what is ray-case?}
 %
@@
 A case study to highlight the non-deterministic effects in a game engine are presented where scenarios between pedestrian and vehicle actors are used to illustrate conditions that result in high simulation variance.
-By limiting the the level of system utilisation to less than 75\% and terminating scenarios when any collision callback is made, a deterministic operational domain was found.
+By limiting the the level of system utilisation to less than 75\% and terminating scenarios when any collision is detected, a deterministic operational domain was found.
 %
@@
 
-This paper is structured as follows.
-In Section~\ref{s:background} game engines and sources of potential non-determinism in game engines are investigated.
+This paper is structured as follows. In Section~\ref{s:background} game engines and sources of potential non-determinism in game engines are investigated. 
+%
+A case study to illustrate the observed simulation variance for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4 GC - ??} 
+%
 Section~\ref{s:methodology} presents a method to assess the suitability of a simulation system for AV verification. 
 %
-A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4 GC - ??} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}
 
@@
 \includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoopv2.pdf}
-\caption{Game engine loop block diagram~\cite{GameProgPatternsBook} \todo[inline]{AG: remove the clock symbol, possibly remove the ticks chart as this is confusing given they are not ticks but event markers.}}
+\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}.}
 \label{GameEngineLoopDiagram}
 \end{figure}
+
+% \todo[inline]{AG: remove the clock symbol, possibly remove the ticks chart as this is confusing given they are not ticks but event markers}
 
@@
 %
-Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
+% Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
 %
@@
 If the engine tries to render between physics updates, \textit{residual lag} can occur and extrapolation between frames needs to be performed to smooth transition between scenes.
-This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}.
+This residual lag may result in frames arriving with a delay to the simulated physics which could be result in errors for perception stack testing.
+% This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
 % \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
@@
 and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour. 
-\todo[inline]{Explain how this lag may affect perception stack testing}
+% \todo[inline]{Explain how this lag may affect perception stack testing}
 
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v4.pdf}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v3.pdf}
     \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}

### 2020-05-13 paper/cav_determinisim_main.tex
changes  up to conclusion

@@
 
-Any hardware simulation that runs for 1000's hours, such as AV verification tasks, are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
+Any simulation hardware not using ECC memory that runs for 1000's hours, typical in AV verification, are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
 
@@
 %
-Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.%.
+Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
+%.
 Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
@@
 \noindent\underline{\textit{Game Engine Code:}}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh 
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
+%
 % \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
-, deterministic ego vehicle
 % \todo[inline]{introduce ego vehicle earlier on GC added line 99}
- controllers and engine texture loading rates especially for perception stack testing. %
 For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{game} which reports
@@
 If the network is exposed to unseen data it may incorrectly classify perfectly legitimate scenes due to lack of training data.
-In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically.
+In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically, assuming no \textit{residual lag}.
 In this sense, control over weather and lighting in the game engine should be considered and controlled as well as any other dynamic elements to the environment, e.g.
@@
 %
-NUMA should only affect interprocessor data transfer cost and only make computations longer.
-relative times between access different caches is likely to be small although may impact on high throughput, e.g.
+NUMA should only affect interprocessor data transfer cost and without control measures will only make the computation cycle longer.
+Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.
 HPC.
@@
 \section{Empirical Investigation} \label{s:case-study}
-We present an empirical investigation to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction}.
+We present an empirical investigation to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction} and including the potential sources of non-determinism from Section~\ref{s:nondeterminisimSources}.
 The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance %$\sigma^2$
@@
 %
-Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a collision callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m intervals using a PID controller.
+Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a \textit{event hit} callback in the game engine (test 2).
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m \todo[inline]{AG: is this value correct or should it be 2.0m?} intervals using a PID controller.
 In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
-In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a collision callback in the game engine, see Fig~\ref{Test_a}.
+In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a \textit{event hit} callback, see Fig~\ref{Test_a}.
 Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
@@
 \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-
 Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}.
-
-\todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
-
+% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
 An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
@@
 
-However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation time.
+However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
 %
@@
 Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision.
-This pattern of deterministic behaviour prior to collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This pattern of deterministic behaviour prior to collision and non-deterministic post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
 This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-deterministic.
+Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-deterministic.
 % \todo{Why ``not reliable''.}
@@
 In this scenario (test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays non-deterministic actor position post-collision as described above.
+There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision as described above.
 Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
 Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that any collision between actors can affect the simulation variance of all actors in the simulation and could potentially result in  erroneous assertion checking.
+This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that \textit{any collision between actors can affect the determinism of the entire actor population} and could potentially result in  erroneous assertion checking.
 
@@
 However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly.
-Therefore we present a general methodology that practitioners can follow to find the deterministic domains of operation for any hardware configuration.
+Therefore we present a general methodology that practitioners can follow to find the \textit{deterministic domains of operation} for any hardware configuration.
 
@@
 
-Different actor types may be handled differntly by the game engine and our investigation indicated that Carla pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
+Different actor types may be handled differently by the game engine and our investigation indicated that Carla pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
 However the introduction of Carla vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
@@
 This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
-At some low analogue voltage the signal is interpreted as binary $0$, this may be any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
+At some low analogue voltage the signal is interpreted as binary $0$, this may be, for example, any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
 This convention enables the abstraction from the physical signal values that are analogue voltages into the digital domain with binary representation.
@@
 \subsubsection{Simulation Length} 
-The simulation time should be sufficient to record interactions between actors but not so long that 100's or 1000's of repeated tests take a considerable time to complete.
+The simulation time should be sufficient to record interactions between actors but not so long that 100's or 1000's of repeated tests take an inconvenient time to complete.
 In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
@@
 CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program can be used spawn workers on any number of cores or virtual threads on a system.
-This can be used to artificially increased the load on the system to explore at what point your system may become unstable leading to \textit{simulation variance}.
+This can be used to artificially increased the load on the system to explore at what point your system may become susceptible to \textit{simulation variance}.
 For GPU utilisation \texttt{gpu-burn} was employed using the \texttt{fur test} where different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
@@
 The maximum value of simulation variance over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control for deterministic operation.
-Finding these boundaries will identify the deterministic operational domain for the user specific hardware.
+Finding these boundaries will identify the \textit{deterministic operational domain} for the user specific hardware.
 

### 2020-05-15 paper/cav_determinisim_main.tex
Replied to todo questions.

@@
 Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a \textit{event hit} callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m \todo[inline]{AG: is this value correct or should it be 2.0m?} intervals using a PID controller.
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m \todo[inline]{AG: is this value correct or should it be 2.0m?
+AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
+actors would ignore trajectory waypoints that are less than 2.0m away from them.} intervals using a PID controller.
 In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.

### 2020-05-26 paper/cav_determinisim_main.tex
added citation to definition of determinism

--- 2020-05-13 paper/cav_determinisim_main.tex
@@
 % What is a deterministic system
-Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
+Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
 %

### 2020-06-01 paper/cav_determinisim_main.tex
Written conclusion, to fit the latest modifications made by the team

@@
 \todo[inline]{Future work...}
+In this paper we presented an investigation (based on Carla driving simulator) to show whether game engines are deterministic or not. 
+Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs. 
+We found two main outcomes from our investigation: 
+1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}. 
+2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.
+Thus, by altering these settings one can achieve sufficiently low levels of simulation varince that can allow users to assume the simulation is deterministic. 
+We phrased the level below which the simulation is to be assumed determinisitic as the \textit{determinisitic threshold}, which is up to the user to define where it exactly lies. 
+This allows for one to use these game engines knowing that their simulation will be sufficiently deterministic provided that they bound their simulation within a certain configuration.
+
+The configuration presented in here is specific to the hardware and software used in our investigation. 
+Therefore, we outlined a methodology in this paper to allow practitioners in this field of AVs to resolve the non-deterministic operational domains of their simulation.
+
+Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
+Hence allowing its extensive usgae by both the AVs and the gaming community.
 

### 2020-06-03 paper/cav_determinisim_main.tex
Written conclusion, to fit the latest modifications made by the team

@@
 We found two main outcomes from our investigation: 
-1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}. 
+1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}, e.g. runtime scheduling. 
 2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.

### 2020-06-24 paper/cav_determinisim_main.tex
conclusion update

@@
 
+\todo[inline]{GC: we need to say what we set out to achieve and why that is important, then say how our investigation led us to several theories which resulted in understanding several key parameters which directly affect simulation variance.
+From this we developed a methodology that the community can use to understand the deterministic operational domains of their own specific hardware:  Autonomous vehicle testing is adopting the use of game engines for development and verification of vehicle control systems in simulation.
+Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
+An empirical investigation uncovered several parameters that contributed to simulation variance ...
+explain results ...
+This led to the design of a methodology allowing 'users' to determin the deterministic operational domains for a specific hardware configuration ...
+}
+
 \todo[inline]{Review/rewrite this to match the earlier changes in the paper.
@@
 Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy.
-A case study was used to ascertain the boundary between deterministic and non-deterministic operation of the Carla driving simulator.
+A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the Carla driving simulator.
 Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study.
@@
 }
+
+\todo[inline]{GC: think we should put more emphasis on the outcomes of the method rather than the cast study.}
 

### 2020-06-24 paper/cav_determinisim_main.tex
Conclusion done

@@
 %
-Carla\footnote{http://carla.org/}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
+Carla\footnote{\url{http://carla.org/}}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
 
@@
 Unreal\footnote{\url{https://www.unrealengine.com/}}, Unity\footnote{\url{https://unity.com/}}, CryEngine\footnote{\url{https://www.cryengine.com/}}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.\ Carla\footnote{\url{http://carla.org/}}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.\ Carla\footnote{\url{http://carla.org/}}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
+% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
 
@@
 The game loop is responsible for the interaction between the physics and rendering engines.
-Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed\footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}.
+Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed~\cite{unity_event_execution}.
+% \footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}. 
 A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering)~\cite{GameEngineArchBook}.
@@
 Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate (loop in physics update in Fig.~\ref{GameEngineLoopDiagram}).
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}.
-When the frame is rendered the game loop cycle returns to processing inputs.
-And a good description of the interplay between the physics and render cycles are given here~\cite{JohnAustinUnity}.
+The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\cite{unity_framerates}.
+% \footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}. 
+When the frame is rendered the game loop cycle returns to processing inputs. And a good description of the interplay between the physics and render cycles are given here~\cite{JohnAustinUnity}.
 
@@
 \todo[inline]{Future work...}
-In this paper we presented an investigation (based on Carla driving simulator) to show whether game engines are deterministic or not. 
-Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs. 
-We found two main outcomes from our investigation: 
-1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}, e.g. runtime scheduling. 
-2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.
-Thus, by altering these settings one can achieve sufficiently low levels of simulation varince that can allow users to assume the simulation is deterministic. 
-We phrased the level below which the simulation is to be assumed determinisitic as the \textit{determinisitic threshold}, which is up to the user to define where it exactly lies. 
-This allows for one to use these game engines knowing that their simulation will be sufficiently deterministic provided that they bound their simulation within a certain configuration.
-
-The configuration presented in here is specific to the hardware and software used in our investigation. 
-Therefore, we outlined a methodology in this paper to allow practitioners in this field of AVs to resolve the non-deterministic operational domains of their simulation.
+
+In this paper we explored the determinisim of game engines.
+This is important for autonomous vehicles' development and verification of vehicle control systems in simulation.
+Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
+
+An empirical investigation based on Unreal Engine running Carla driving simulator, uncovered several parameters that contributed to simulation variance.
+These parameters include numerous internal settings based in Unreal Engine, and others that are external mainly dependant on CPU and GPU utilisation.
+
+Changes in CPU and GPU resource utlisation has shown to be the main influencer for the non-deterministic behaviour of game engines.
+This is clearly demonstrated in the results (see section \ref{s:case-study}), where the max deviation between runs increased from the orders of $10^{-12}$m to devaitons of up to 1m as the CPU and GPU resource utilisation increased above 50\%.
+
+The term \textit{deterministic operational domain} was given to the range of system resource utlisations that provides low simulation variances that is insignificant for development and verifcation of autonomous vehicles.
+The deterministic operational domain specified herein is specific to our hardware configuration.
+
+This led to the design of a methodology allowing `users' to determine the deterministic operational domain for a specific hardware configuration.
+This will help guide specialists in the field of AVs to setup their custom experiments for their specific simulators and identify their specific deterministic operational domains.
 
 Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
-Hence allowing its extensive usgae by both the AVs and the gaming community.
+Hence allowing its extensive usgae by both the autonmous vehicles' and the gaming community.
+
+% presented an investigation (based on Carla driving simulator) to show whether game engines are deterministic or not. 
+% Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs.
+
+% An empirical investigation based on Unreal Engine running Carla driving simulator, was used to illustrate the challenges of determinisim faced by game engines. 
+% % It was found that game engines are non-deterministic.
+
+% We reviewed possible sources of non-determinisim in section \ref{s:background} and investigated the effect of each on the simulation variance.
+
+% We found that the sources of non determinsim that had a significant effect on simulation variance were attributed to CPU and GPU resource utilisation. 
+% The case study clearly shows how the max simulation variance changes from very low simulation varicances, below picometer level ($10^-12$), to centimeter level when the CPU and GPU resource utilisation increases above 50\%.
+
+% We found two main outcomes from our investigation: 
+% 1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}, e.g. runtime scheduling. 
+% 2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.
+% Thus, by altering these settings one can achieve sufficiently low levels of simulation varince that can allow users to assume the simulation is deterministic. 
+% We phrased the level below which the simulation is to be assumed determinisitic as the \textit{determinisitic threshold}, which is up to the user to define where it exactly lies. 
+% This allows for one to use these game engines knowing that their simulation will be sufficiently deterministic provided that they bound their simulation within a certain configuration.
+
+% The configuration presented in here is specific to the hardware and software used in our investigation. 
+% Therefore, we outlined a methodology in this paper to allow practitioners in this field of AVs to resolve the non-deterministic operational domains of their simulation.
 

### 2020-06-29 paper/cav_determinisim_main.tex
GC don't think i made any changes

--- 2020-06-24 paper/cav_determinisim_main.tex
@@
 explain results ...
-This led to the design of a methodology allowing 'users' to determin the deterministic operational domains for a specific hardware configuration ...
+This led to the design of a methodology allowing 'users' to determine the deterministic operational domains for a specific hardware/platform configuration ...
 }

### 2020-06-29 paper/cav_determinisim_main.tex
conclusion GC

--- 2020-06-24 paper/cav_determinisim_main.tex
@@
 % ***************************************************
-\section{Conclusions and Future Work}\label{s:conclusion}
-
-\todo[inline]{Say what the paper investigated and what we aimed to achieve, why this is important and how we went about it.
+% \section{Conclusions and Future Work}\label{s:conclusion}
+\section{Conclusions}\label{s:conclusion}
+
+% \todo[inline]{Say what the paper investigated and what we aimed to achieve, why this is important and how we went about it.
 Then point out the derivation of a re-usable methodology based on the knowledge and understanding we gained from the case study.}
 
-\todo[inline]{GC: we need to say what we set out to achieve and why that is important, then say how our investigation led us to several theories which resulted in understanding several key parameters which directly affect simulation variance.
+% \todo[inline]{GC: we need to say what we set out to achieve and why that is important, then say how our investigation led us to several theories which resulted in understanding several key parameters which directly affect simulation variance.
 From this we developed a methodology that the community can use to understand the deterministic operational domains of their own specific hardware:  Autonomous vehicle testing is adopting the use of game engines for development and verification of vehicle control systems in simulation.
@@
 explain results ...
-This led to the design of a methodology allowing 'users' to determin the deterministic operational domains for a specific hardware configuration ...
+This led to the design of a methodology allowing 'users' to determine the deterministic operational domains for a specific hardware/platform configuration ...
 }
 
-\todo[inline]{Review/rewrite this to match the earlier changes in the paper.
+The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification.
+Having a deterministic simulator is required for reliable and repeatable results, to find and fix software bugs and ensure results are trustworthy.
+If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains where reliable results can be found.
+During an investigation into the Carla simulator, a \textit{simulation variance} was observed for repeated tests with the same initial conditions indicating non-deterministic execution.
+During the study we hypothesized and uncovered several parameters that contribute towards greater simulation variance and hence non-deterministic execution, such as actor collisions and system resource utilisation.
+The results of the investigation are hardware and software version specific, so we present a universal methodology that can be used to find the \textit{deterministic operational domain} for any given configuration.
+This methodology will allow the AV verification community using simulation tools to ensure their results are reliable and trustworthy.
+
+% \todo[inline]{Review/rewrite this to match the earlier changes in the paper.
 Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
@@
 
-\todo[inline]{GC: think we should put more emphasis on the outcomes of the method rather than the cast study.}
-
-\todo[inline]{Point out (again) that all data and scripts etc are on github.}
-
-\todo[inline]{Future work...}
-
-In this paper we explored the determinisim of game engines.
+% \todo[inline]{GC: think we should put more emphasis on the outcomes of the method rather than the cast study.}
+% \todo[inline]{Point out (again) that all data and scripts etc are on github.}
+
+% \todo[inline]{Future work...}
+
+% In this paper we explored the determinisim of game engines.
 This is important for autonomous vehicles' development and verification of vehicle control systems in simulation.
@@
 
-An empirical investigation based on Unreal Engine running Carla driving simulator, uncovered several parameters that contributed to simulation variance.
+% An empirical investigation based on Unreal Engine running Carla driving simulator, uncovered several parameters that contributed to simulation variance.
 These parameters include numerous internal settings based in Unreal Engine, and others that are external mainly dependant on CPU and GPU utilisation.
 
-Changes in CPU and GPU resource utlisation has shown to be the main influencer for the non-deterministic behaviour of game engines.
+% Changes in CPU and GPU resource utlisation has shown to be the main influencer for the non-deterministic behaviour of game engines.
 This is clearly demonstrated in the results (see section \ref{s:case-study}), where the max deviation between runs increased from the orders of $10^{-12}$m to devaitons of up to 1m as the CPU and GPU resource utilisation increased above 50\%.
 
-The term \textit{deterministic operational domain} was given to the range of system resource utlisations that provides low simulation variances that is insignificant for development and verifcation of autonomous vehicles.
-The deterministic operational domain specified herein is specific to our hardware configuration.
-
-This led to the design of a methodology allowing `users' to determine the deterministic operational domain for a specific hardware configuration.
+% The term \textit{deterministic operational domain} was given to the range of system resource utlisations that provides low simulation variances that is insignificant for development and verifcation of autonomous vehicles.
+% The deterministic operational domain specified herein is specific to our hardware configuration.
+
+% This led to the design of a methodology allowing `users' to determine the deterministic operational domain for a specific hardware configuration.
 This will help guide specialists in the field of AVs to setup their custom experiments for their specific simulators and identify their specific deterministic operational domains.
 
-Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
+% Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
 Hence allowing its extensive usgae by both the autonmous vehicles' and the gaming community.

### 2020-06-30 paper/cav_determinisim_main.tex
conclusion GC

@@
 Having a deterministic simulator is required for reliable and repeatable results, to find and fix software bugs and ensure results are trustworthy.
-If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains where reliable results can be found.
+If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains for reliable results.
 During an investigation into the Carla simulator, a \textit{simulation variance} was observed for repeated tests with the same initial conditions indicating non-deterministic execution.
 During the study we hypothesized and uncovered several parameters that contribute towards greater simulation variance and hence non-deterministic execution, such as actor collisions and system resource utilisation.
-The results of the investigation are hardware and software version specific, so we present a universal methodology that can be used to find the \textit{deterministic operational domain} for any given configuration.
+The results of the investigation are hardware and software version specific, so we present a universal methodology that can be used to find the \textit{deterministic operational domain} for any given system configuration.
 This methodology will allow the AV verification community using simulation tools to ensure their results are reliable and trustworthy.

### 2020-07-02 paper/cav_determinisim_main.tex
revision of first page and addition of CyRes to text and bib

@@
 % ***************************************************
-\title{Investigating Determinism of Game Engines for Autonomous Vehicle Verification}
+\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
 \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
@@
 \author[1,3]{Kerstin Eder}
-\affil[1]{Trustworthy Systems Lab, University of Bristol, Bristol, UK}
+\affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
 \affil[2]{University of West of England, Bristol, UK}
-\affil[3]{Bristol Robotics Lab, Bristol, UK}
+\affil[3]{Bristol Robotics Laboratory, Bristol, UK}
 \maketitle
@@
 \noindent 
-The autonomous vehicle (AV) community is adopting the use of game engines to develop vehicle control systems and testing environments using simulation.
-Determinism is a key requirement for simulation during AV development and testing.
-In particular, it ensures repeatability of tests, so that software bugs can be found and fixed, and simulation results are trustworthy.
+Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
+%
+Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found, fixed and fixes can be checked, and simulation results are trustworthy.
 %
@@
 %
-This work identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation variance} for specific configurations and utilisation of the software and hardware, respectively.
-%
-A case study using Carla, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from a small scenario indicate the presence of deterministic operational domains.
-\todo[inline]{KIE: review at end}
+This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation variance} for specific configurations and utilisation of the software and hardware, respectively.
+%
+A case study using Carla, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of deterministic operational domains.
+\todo{Review again at end.
+Potentially give away more on results to motivate the reader to continue.}
 
@@
 \begin{IEEEkeywords}
-Autonomous Driving, Determinism, Physics Engines, Autonomous Vehicles, Verification and Validation (V\&V), Simulation, Testing
+Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
 \end{IEEEkeywords}
@@
 % What is verification
-\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}.
-Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it.
-These simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving assertions that must not be violated.
+% Can't use this again as this is how the AI Test paper starts. -> Avoid self-plagiarism!
+%\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. 
+%
+\IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it.
+Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving rules, termed assertions, that must not be violated.
 
@@
 % Requirements of simulation based verification SBV
-The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
+State-of-the-art game engines provide a convenient option for simulation-based testing.
+They offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
 %
@@
 %
-\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
-%
+%\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
+
 For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
 Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
-\todo[inline]{GC: what is ray-case?}
-%
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required.
+\todo{GC: what is ray-case? KIE: Not sure. :(}
+%
+Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013,cyres} may also be required.
 
 % Game engines for SBV
-Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
-%
-\todo[inline]{GC Replace: However, several challenges remain; one of these is determinism.
-with: However, game engines are designed principally for user experience whereas the requirements for CAV verification is not user experience but determinism.}
+Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
+%
+\todo{GC Replace: However, several challenges remain; one of these is determinism. with:}
+%
+However, while game engines are designed primarily for user experience, the requirements for AV verification go beyond that and include determinism.
 %
 % What is a deterministic system
-Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
-%
-Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly under identical initial conditions.
+In this context, deterministic refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
+%
+Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
 Specifically, we focus on the variance of actor paths within a given scenario.
@@
 %For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
-Determinism is a key prerequisite for simulation during AV development and testing. 
+\todo{move to earlier :)}Determinism is a key prerequisite for simulation during AV development and testing. 
 %

### 2020-07-21 paper/cav_determinisim_main.tex
up to Empirical Investigation

@@
 %
-Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found, fixed and fixes can be checked, and simulation results are trustworthy.
+Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
 %
@@
 %
-A case study using Carla, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of deterministic operational domains.
+A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.
 \todo{Review again at end.
@@
 \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it.
-Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving rules, termed assertions, that must not be violated.
+Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of safe and legal driving rules, termed assertions, that must not be violated.
 
@@
 %
-Carla\footnote{\url{http://carla.org/}}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
+CARLA\footnote{\url{http://carla.org/}}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
 
@@
 For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
-Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
+CARLA currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
 \todo{GC: what is ray-case? KIE: Not sure. :(}
@@
 Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013,cyres} may also be required. 
+%
+% The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
 
@@
 %
-However, while game engines are designed primarily for user experience, the requirements for AV verification go beyond that and include determinism.
+However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
 %
@@
 %For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
-\todo{move to earlier :)}Determinism is a key prerequisite for simulation during AV development and testing. 
+Determinism is a key prerequisite for simulation during AV development and testing. 
 %
@@
 If the simulation is not deterministic, however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation are left undetected, resulting in false confidence in the safety of the AV's control software.
+In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
 
 % Challenges of SBV
-Game engines do not need to be deterministic when used for gaming; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
-Because the situation is very different for AV development and testing, our main research question is:
-how can one assess the extent to which a simulation environment is deterministic? 
-%
-In pursuit of an answer to this question we have developed a method to determine the simulation variance for a given simulation environment. 
-%
-Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.
-
+When used for gaming, game engines do not need to be deterministic; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
+However, the situation is very different for AV development and testing.
+Thus, our main research questions is:
+How can one assess the extent to which a simulation environment is deterministic?
+
+In pursuit of an answer to this question we have investigated how non-determinism of a game engine, namely Unreal Engine that powers CARLA, when used for simulation-based AV verification affects the simulation results.
+%
+In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in high simulation variance.
+%
+By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
+
+The insights gained from this case study have motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
+%
+Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
+In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
+%
 % \todo[inline]{review and potentially summarize results from case study GC - added see below.}
-A case study to highlight the non-deterministic effects in a game engine are presented where scenarios between pedestrian and vehicle actors are used to illustrate conditions that result in high simulation variance.
-By limiting the the level of system utilisation to less than 75\% and terminating scenarios when any collision is detected, a deterministic operational domain was found.
-%
-%The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
-
-This paper is structured as follows. In Section~\ref{s:background} game engines and sources of potential non-determinism in game engines are investigated. 
-%
-A case study to illustrate the observed simulation variance for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4 GC - ??} 
-%
-Section~\ref{s:methodology} presents a method to assess the suitability of a simulation system for AV verification. 
-%
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}
+%We present a case study to illustrate the use of this method.
+The study highlights the effects of non-determinism of a game engine on simulation results.
+Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
+%
+%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
+%
+
+This paper is structured as follows.
+%
+Section~\ref{s:background} briefly introduces how game engines work before investigating potential sources of non-determinism in game engines.
+%
+Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
+%
+Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
+%
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
 
@@
 
-% Robopilot, Capri, Carla, Apollo, Airsim, Udacity etc.
+% Robopilot, Capri, CARLA, Apollo, Airsim, Udacity etc.
 are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}.
@@
 
-There are numerous game engines with their associated development environments which could be suitable for AV development, e.g.
-Unreal\footnote{\url{https://www.unrealengine.com/}}, Unity\footnote{\url{https://unity.com/}}, CryEngine\footnote{\url{https://www.cryengine.com/}}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g.\ Carla\footnote{\url{http://carla.org/}}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
+There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g.
+Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, Unity\footnote{\url{https://unity.com/}}, CryEngine\footnote{\url{https://www.cryengine.com/}}.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA\footnote{\url{http://carla.org/}}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
 % \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
@@
 The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate (loop in physics update in Fig.~\ref{GameEngineLoopDiagram}).
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\cite{unity_framerates}.
+Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
+This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}.
+The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}.
 % \footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}. 
-When the frame is rendered the game loop cycle returns to processing inputs. And a good description of the interplay between the physics and render cycles are given here~\cite{JohnAustinUnity}.
+When the frame is rendered the game loop cycle returns to processing inputs.
+An intuitive and  more detailed description of the interplay between the physics and render cycles is given in~\cite{JohnAustinUnity}.
 
@@
 %
-To use a fixed physics time step the user's display refresh rate needs to be known in advance.
+To use a fixed physics time step, the user's display refresh rate needs to be known in advance.
 This requires an update loop to take less than one render tick (one frame of real world time).
@@
 %
-Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware.
+Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware.
 Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}.
 %
-If the engine tries to render between physics updates, \textit{residual lag} can occur and extrapolation between frames needs to be performed to smooth transition between scenes.
-This residual lag may result in frames arriving with a delay to the simulated physics which could be result in errors for perception stack testing.
+If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics.
+%
+Thus, extrapolation between frames may need to be performed to smooth transition between scenes. 
+%
+Note that both residual lag and extrapolation could affect perception stack testing.
 % This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
 % \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
-
+%
 % \todo[inline]{Why is this part of the figure?
@@
 % Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
-In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour. 
+In exceptional cases where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines,
+and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
 % \todo[inline]{Explain how this lag may affect perception stack testing}
@@
 
-\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
+Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences. 
+%
+Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy.
+%
+In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses.
+If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance.
+%
+For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
+This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
+\todo{KIE: ...
+by slowing down the simulation to enable more extensive physics calculations?}
+
+%\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
 This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer).
 For instance, see below, but I'm not sure whether this is the right place.}
-In addition, providing game players with a responsive real-time experience is often achieved at the cost of realism.\todo[inline]{Do we have a ref for this, Unity vs UE4?}
-%
-In contrast, realism is essential in AV simulation, especially for perception stack testing, but also for visual inspection.
-If necessary, it is acceptable to achieve the required level of realism at the cost of the real-time user experience.
 
@@
 
-The following discussion is the result of our investigation into the potential sources of non-determinism that could occur in game engines.
+The following discussion reflects the results of our investigation into the potential sources of non-determinism that could affect game engines.
 We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
@@
 
-% ======= Floating Point
-\noindent\underline{\textit{Floating Point Representation:}}
+% ======= Floating-Point Numbers
+\noindent\underline{\textit{Floating-Point Arithmetic:}}
 Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
-Thus, the results of arithmetic operations must fit into the given bit with.
-This is achieved through rounding to the nearest representable value, which gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
+Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value.
+This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
 %
@@
 % Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.
+In the context of AV simulation, such rounding errors could result in precision issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
 % \todo[inline]{state what assertions are (used for) GC-see below} 
-In the field of autonomous vehicle verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
-%
-One could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.
+%In the field of AV verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
+%
+Thus, one could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.
+
 However, these precision issues are better described as \textit{incorrectness}, rather than them being a source of non-determinism. 
 %
-It is safe to assume that the hardware on which game engines run is designed to be deterministic. 
-%
-Thus, when a rounding error occurs, then the same hardware should produce the same \textit{incorrect} result given the same operands. 
+It is normally reasonable to assume that the processors on which game engines run are designed to be deterministic with respect to floating-point arithmetic. 
+%
+Thus, when a rounding error occurs, then the same processor should produce the same \textit{incorrect} result given the same operands. 
 %
@@
 be equally \textit{incorrect} as long as the implementation conforms to the IEEE floating-point standard~\cite{8766229}.
-%
+
 Beyond hardware limitations, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
@@
 %
-Therefore, it can be concluded that floating point representation does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware.
-%
-% KIE: The text below is not sufficiently rigorous and not correct. 
-% A generic issue with computers are floating points precision. 
-%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.increasingl\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.
-%Translating to the usage of AV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.
-%
-% For instance, consider a simulation where the world is infinite and progressively  generated in the physics engine. 
-% After a few hundred meters, precision issues will start to occur, and will get progressively worse the further from the origin the simulation gets. 
-% This is due to the values of the coordinates being used in calculations are getting bigger and hence will not cope with small values that are critical for V\&V testing purposes. 
-% A solution to this problem is perhaps every time the simulations moves 100 meters away from the origin the entire world will move by 100m in the opposite direction.
-Thus, avoiding getting into floating point issues.\cite{FloatingPointArithmeticArticle}
-%
-% There are many scenarios that could be addressed, and solutions to them will entirely depend on the characteristics of that specific scenario.
-There is no generalised solution for the floating problem, regardless of how good computers get there will always be a limitation.
-However, floating points as they stand could just be sufficient for V\&V testing purposes.
-Whenever they are not, smart solutions can be found to accommodate for these limitations.
-\\\\
-% ======= Floating Point
+Therefore, it can be concluded that floating-point arithmetic does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware.
+
 \noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
-ECC Memory can detect and correct single bit errors in DRAM memory.
+ECC Memory is commonly used in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory.
 Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
 If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask.
+Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
+
+Any simulation hardware not using ECC memory that runs for 1000's of hours, typical in AV verification, is likely to incur significant CPU hours and is therefore subject to increased exposure to these errors.
+To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.\todo{Check this, please.}
+
+%\todo[inline]{That said, HW design simulation is deterministic, uses long simulation runs and thousands of CPU hours, but I've not heard that this is an issue.
+I can ask, though.
 ;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.}
 
-Any simulation hardware not using ECC memory that runs for 1000's hours, typical in AV verification, are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
-
 \noindent\underline{\textit{Actor Navigation:}}
-There is existing evidence to suggests actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-% 
-Unreal Engine is the underlying framework for the Carla simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
-%
-The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}. 
+There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
+% 
+Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
+%
+The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
+
+A navigation mesh is a fixed area of the environment where actors are free to navigate.
 %
@@
 %
-However, the environment management in Carla implements fixed binary files for navigation meshes that are linked to each driving scene and cannot be unknowingly modified and therefore can be considered fixed.
-%
-Unless other sources that can alter the fixed environment exist, actor navigation should not be considered a source of non-determinism
-%
-The navigation mesh is a fixed area of the environment where actors are free to navigate. Typically the navigation mesh is fixed for a specific environment or road scene and therefore the A* algorithm
- should give deterministic results.
+However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment and cannot be unknowingly modified and therefore can be considered fixed.
+
+Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
+%
 % Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
@@
 \\\\
-\noindent\underline{\textit{Game Engine Code:}}
+\noindent\underline{\textit{Game Engine Setup:}}
 The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
@@
 the total time spent generating a frame, executing the game loop or render thread.
+%
+Wrt.\ perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water.
 \\\\
-\noindent\underline{\textit{Scheduling:}}
+\noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
+For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
+Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
+%
+Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
+%
+By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
+\\\\
+\noindent\underline{\textit{Scheduling, Concurrency and Parallelisation:}}
 % \todo[inline]{is this at runtime?} %
@@
 % https://en.wikipedia.org/wiki/Race_condition#cite_note-1
-Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy by the operating system.
+Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy of the operating system.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
@@
 %
-If during simulation the scheduling policy was updated or thread priority was inconsistently handled, then a high variance in output may result.
-It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.
-% \todo[inline]{justify - WHY? GC- see below} 
-Given that hardware is considered deterministic and if the scheduling policy does not change between tests, then for the same hardware and operating system configuration the same output will be given.
-%
-However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and the scheduling policy is set to equal priority for all threads then a background process may interrupt the otherwise deterministic sequence of events which may, for example, alter the number of physics calculations that can be performed within the game loop and hence variance in the simulation.
-% KIE: If all threads have the same priority, then the scheduler is free to select one, this can give rise to non-det - or not?
-
-Similar to thread scheduling, the CPU scheduling policy determines on which processor core to execute processes in the ready queue and this is maybe ordered by factors such as throughput, latency and CPU utilisation.
-If the same single-threaded script executes multiple times across different physical cores due to the CPU utilisation policy then should you expect deterministic behaviour?
-Processor core designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist.
-End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo[inline]{not clear, what is the difference between scheduling cores and threads - are both done by the operating system?
-GC - does this come from a software request?
-i.e.
-request to use a core or just a thread}
-%
-Overall, scheduling of cores or threads at runtime may be a reason for increased \textit{simulation variance}. 
-%
-% KIE: Not relevant here - we are just listing the sources of non-det.
-% but controlling for this falls outside of the scope of this work and no attempt to control 
-% these parameter have been made here. However, users should be aware of this potential issue 
-% and consider methods to move towards a fixed scheduling sequence if possible.
+Changing the scheduling policy and thread priorities may increase simulation variance. 
+%
+It is, however, unlikely that these would occur during repeated controlled tests for the same hardware and operating system configuration.
+Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
+
+However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
+This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
+%
+Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
+
+Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes.
+This may be decided based on factors such as throughput, latency or CPU utilisation.
+%
+If due to the CPU utilisation policy the same single-threaded script executes multiple times across different physical cores of the same CPU type,  then execution should still produce the same output.
+%
+This is because the processor cores are identical and any impurities across the bulk silicon and minor perturbations in the semiconductor processing across the chip that may exist should have been accommodated for in design and manufacturing tolerances.
+%
+However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
+
+Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously.
+While this would be faster than a serial execution, the order of execution arising from program-level concurrency is often not guaranteed.
+
+Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}. 
 \\\\
-\noindent\underline{\textit{Concurrency:}}
+%\noindent\underline{\textit{Concurrency:}}
 % Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
-\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
+%\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
 Consider also parallelism, e.g.
@@
 Say which parts of a game engine could reasonably be expected to run concurrently.}
-
-Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
+%
+%Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
 This technique can lead to issues such as shared resource access conflicts due to mis-managed timing and RACE conditions~\cite{huffman1954synthesis}.
@@
 % although monitoring and detecting such events falls outside of the scope for this work. 
-\\\\
-\noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
-For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
-Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
-%
-Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically, or, perhaps, simply with an increased data transfer cost. 
-%
-By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
-\\\\
-\noindent\underline{\textit{Ego Vehicle Controller:}}
-The ego vehicle controller may be a source of non-determinism if using complex control algorithms such as neural networks for visual processing and perception.
-A neural network can only be deterministic within in the bounds of the training data that was used to construct it.
-If the network is exposed to unseen data it may incorrectly classify perfectly legitimate scenes due to lack of training data.
-In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically, assuming no \textit{residual lag}.
+%
+%\noindent\underline{\textit{Ego Vehicle Controller:}}
+%The ego vehicle controller may be a source of non-determinism if using complex control algorithms such as neural networks for visual processing and perception.
+A neural network can only be deterministic within in the bounds of the training data used to construct it.
+If the network is exposed to unseen data it may incorrectly classify legitimate scenes due to lack of training data.\todo{That alone won't cause non-determinism, i.e.
+the NN will misclassify exactly in the same way each time - not?}
+%
+%In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically, assuming no \textit{residual lag}.
 In this sense, control over weather and lighting in the game engine should be considered and controlled as well as any other dynamic elements to the environment, e.g.
@@
 % %
-% AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
+% AV specific platforms, such as CARLA, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
 % \todo[inline]{Add a summary/conclusion?}
-\\\\
+%\\\\
 \noindent\underline{\textit{Summary:}}
-The potential sources of non-determinism described above are summarised here along with their potential effects on simulation variance.
-Assuming deterministic hardware, there is no contribution to simulation variance due to floating point representation and any incorrect rounding errors due to byte size limits should be repeatable.
-Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
-%
-The physics setting $dt$ must be fixed and along with any actor navigation meshes and random number seeds to ensure accurate and repeatable simulation data.  
-%
-Thread scheduling should not affect the simulation determinism unless changing scheduling policy or migrating between machines that have different policies.
-However, should new and unexpected threads start during the simulation where an equal priority policy is implemented, the interruption to execution order or additional resource demand may affect timing of subsequent steps reducing the number of physics updates within a game loop and potentially give non-deterministic results.
-%
-The issue of concurrency should be minimised if good software development practice has been followed during the development of the simulation software. 
+We have investigated the potential sources of non-determinism affecting game engines 
+and explored the impact they may have on simulation variance. 
+%
+Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. 
+%
+Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.\todo{So, do commercial HPC facilities have ECC memory?}
+%
+To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
 %
 NUMA should only affect interprocessor data transfer cost and without control measures will only make the computation cycle longer.
-Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.
-HPC.
-%
-The ego vehicle controller can be considered deterministic if dynamic elements to the simulation environment are controlled, such as weather and lighting that may be on random cycles. 
+Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
+%
+Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
+%
+However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop.
+Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism.
+%
+
+%
+%The ego vehicle controller can be considered deterministic if dynamic elements to the simulation environment are controlled, such as weather and lighting that may be on random cycles. 
 %
@@
 In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a \textit{event hit} callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the Carla pedestrian actors.
+Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
 Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
@@
 For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
-This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the Carla platform.
+This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the CARLA platform.
 
@@
 To ensure reliability of results, each test was repeated 1000 times.
-Tests were carried out in Carla (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
+Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
 To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
@@
 There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision as described above.
-Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in Carla the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
+Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in CARLA the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
 Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
@@
 \todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
-The results indicate that actor type, resource utilisation and collision events all affect the determinism in the Carla simulator.
+The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator.
 Collision events may utilise more computational resources than simulations without such events, i.e.
@@
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
-This methodology has been refined several times over the course of our investigations into both Unreal Engine and Carla and is presented here as a work flow, see Fig.~\ref{method_diagram}.
+This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}.
 %These need to be considered at different stages in the process.
@@
 
-Different actor types may be handled differently by the game engine and our investigation indicated that Carla pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
-However the introduction of Carla vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
-As such, all actors that could be included in any future simulation should be tested, including any non-standard Carla or bespoke actors.
+Different actor types may be handled differently by the game engine and our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
+However the introduction of CARLA vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
+As such, all actors that could be included in any future simulation should be tested, including any non-standard CARLA or bespoke actors.
 %
@@
 The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
-If using Carla a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}}.
-In Carla, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
+If using CARLA a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}}.
+In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
 %
@@
 
-The version number of the Carla and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}, so ensuring consistent version number throughout testing is also important.
+The version number of the CARLA and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}, so ensuring consistent version number throughout testing is also important.
 
@@
 If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains for reliable results.
-During an investigation into the Carla simulator, a \textit{simulation variance} was observed for repeated tests with the same initial conditions indicating non-deterministic execution.
+During an investigation into the CARLA simulator, a \textit{simulation variance} was observed for repeated tests with the same initial conditions indicating non-deterministic execution.
 During the study we hypothesized and uncovered several parameters that contribute towards greater simulation variance and hence non-deterministic execution, such as actor collisions and system resource utilisation.
@@
 Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy.
-A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the Carla driving simulator.
+A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the CARLA driving simulator.
 Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study.
@@
 
-% An empirical investigation based on Unreal Engine running Carla driving simulator, uncovered several parameters that contributed to simulation variance.
+% An empirical investigation based on Unreal Engine running CARLA driving simulator, uncovered several parameters that contributed to simulation variance.
 These parameters include numerous internal settings based in Unreal Engine, and others that are external mainly dependant on CPU and GPU utilisation.
@@
 
-% presented an investigation (based on Carla driving simulator) to show whether game engines are deterministic or not. 
+% presented an investigation (based on CARLA driving simulator) to show whether game engines are deterministic or not. 
 % Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs.
 
-% An empirical investigation based on Unreal Engine running Carla driving simulator, was used to illustrate the challenges of determinisim faced by game engines. 
+% An empirical investigation based on Unreal Engine running CARLA driving simulator, was used to illustrate the challenges of determinisim faced by game engines. 
 % % It was found that game engines are non-deterministic.

### 2020-08-13 paper/cav_determinisim_main.tex
added definitions to terms precision and tolerance

@@
 %For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
+% KE Text
 Determinism is a key prerequisite for simulation during AV development and testing.
-%
 A deterministic simulation environment guarantees that tests are repeatable.
-%
 Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
 This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
+
+% GC Text
+\todo[inline]{GC Edit: Determinism is a key prerequisite for simulation during AV development and testing.
+A deterministic simulation environment guarantees that tests are repeatable, or more formally, are \textit{precise}.
+The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
+Therefore \textit{tolerance} is the acceptable degree of variability between execution outputs.
+When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
+}
 %
@@
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
+
+\todo[inline]{GC Edit: If the simulation is not deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
+In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
 

### 2020-08-13 paper/cav_determinisim_main.tex
changed entries wrt accuracy

@@
 A deterministic simulation environment guarantees that tests are repeatable, or more formally, are \textit{precise}.
+This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
 The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
@@
 Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-%
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy.
-%
 In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses.
 If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance.
-%
 For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
 This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-\todo{KIE: ...
-by slowing down the simulation to enable more extensive physics calculations?}
+
+\todo[inline]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
+Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
+The gamer neither wishes a faithful representation of reality, i.e.
+can be low accuracy, nor do they require for repeated actions to result in the same outcome to particular tolerance, i.e.
+can be low precision.
+In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
+If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
+For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
+This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
+
+\todo{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
 

### 2020-08-13 paper/cav_determinisim_main.tex
checked link between deterministic threshold and tolerance

@@
 Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
+\todo[inline]{Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
 %
@@
 To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
+\todo[inline]{To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
 %
@@
 The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification.
-Having a deterministic simulator is required for reliable and repeatable results, to find and fix software bugs and ensure results are trustworthy.
+Having a deterministic simulator is required for reliable and repeatable results, \todo[inline]{Having a deterministic simulator is required for precise results,} to find and fix software bugs and ensure results are trustworthy.
 If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains for reliable results.

### 2020-08-18 paper/cav_determinisim_main.tex
various edits

--- 2020-08-13 paper/cav_determinisim_main.tex
@@
 %For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
-% KE Text
 Determinism is a key prerequisite for simulation during AV development and testing. 
+%
 A deterministic simulation environment guarantees that tests are repeatable. 
+%
 Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
 This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-
-% GC Text
-\todo[inline]{GC Edit: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable, or more formally, are \textit{precise}.
-The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
-Therefore \textit{tolerance} is the acceptable degree of variability between execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
 %
@@
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
-
-\todo[inline]{GC Edit: If the simulation is not deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
 
@@
 % ***************************************************
-\section{Empirical Investigation} \label{s:case-study}
-We present an empirical investigation to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction} and including the potential sources of non-determinism from Section~\ref{s:nondeterminisimSources}.
-The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance %$\sigma^2$
-for any actor should be less than a user-defined threshold.
-
-% \todo[inline]{Introduce the Scenario - see my comment later.}
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-Autonomous vehicles operating as mobility and last mile delivery solutions will predominantly operate within urban rather than highway or motorway environments.
-To determine if the ego vehicle performs correctly, i.e.
-does not violate assertions, the precision of the simulated actor states must be repeatable to within a defined tolerance or threshold.
-
-% \todo[inline]{KIE: Move this to the Case Study section and justify the values.}
-This threshold, termed here as the \textit{deterministic threshold}, must be chosen to reflect the scale at which repeatability of simulation execution is required, see Section~\ref{s:threshold}.
-This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment.
+\section{Case Study of Simulation Variance} \label{s:case-study}
+
+We present an empirical investigation into using game engines for
+simulation-based verification of autonomous vehicles with a focus on
+characterising sources of non-determinism in order to understand  the impact
+they have on simulation variance. 
+%
+Ultimately, our objective is to control non-determinisim to minimise simulation variance.
+
+We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.
+%
+\todo{Complete the section structure as this is a very long section.}
+
+\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
+
+% ***************************************************
+%  Scenario Description
+% ***************************************************
+\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}\todo{Need to stick to scene, scenario definitions as in Intro section.}
+
+This case study draws on a setup used to verify an urban last-mile mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
+%
+Thus, the scene for our investigation is the T-junction shown in Figure~\ref{f:test_a_and_b}.
+
+This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor trajectories over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
+%
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding with or avoiding other actors.
+\todo{Decide past or present tense and be consistent throughout.}
+
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
+\end{figure}
+
+\todo{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+
+\subsection{Tolerable Simulation Variance}
+
+\todo[inline]{Terminology:\\
+{\bf Accuracy} is the degree of closeness/conformance to a known value.
+\\
+{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
+\\
+{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
+\\
+See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
+}
+
+To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
+%
+This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment.
+%
+Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
+
+tolerances are controlled to within a small, acceptable degree of variability from one part to another.
+
 %
@@
 $1\mu m$.
-%
-A deterministic threshold of $1$m is insufficient when considering the spacial resolution required to resolve scenarios such as potentially dangerous close passing of cyclists and pedestrians, and a very small value, e.g.
-$1\times10^{-12}$, may result in false positives due to the simulation \textit{noise floor}\footnote{see Fig.~\ref{CarsCollisionCG25}.} Therefore a threshold of $1$cm is chosen which is sufficient for assertion checking in urban environments.
+
+To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
+
+The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments. 
+%
+In particular, in this case study a determinism threshold of $1$m would be insufficient when considering the spacial resolution required to resolve situations such as potentially dangerous close passing of cyclists and pedestrians.
+A very small value, e.g.\ $1\times10^{-12}$,\todo{units for this - m(etres)?} may result in false positives due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} Therefore, for this case study, a threshold of $1$cm has been selected.
+Thus any variation of less than $1$cm is tolerable
+
+considered to be sufficient for assertion checking in urban environments. 
 %
@@
 
-\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
-A T-junction scene was used to create a number of scenarios involving pedestrians and vehicles. 
-%
-These scenarios were created in order to identify any changes in the actor trajectories over repeated tests under different conditions and hence observe any change in \textit{simulation variance}.
-%
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding with or avoiding other actors.
+For the simulation to be considered repeatable, the overall level of simulation variance 
+for any actor should be less than a user-defined threshold for a given scenario and verification objective. 
+%
 
@@
 
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
-\end{figure}
-
 % Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.

### 2020-08-18 paper/cav_determinisim_main.tex

@@
 %For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
+% KE Text
 Determinism is a key prerequisite for simulation during AV development and testing.
-%
 A deterministic simulation environment guarantees that tests are repeatable.
-%
 Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
 This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
+
+% GC Text
+\todo[inline]{GC Edit: Determinism is a key prerequisite for simulation during AV development and testing.
+A deterministic simulation environment guarantees that tests are repeatable, or more formally, are \textit{precise}.
+This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
+The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
+Therefore \textit{tolerance} is the acceptable degree of variability between execution outputs.
+When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
+}
 %

### 2020-08-19 paper/cav_determinisim_main.tex
minor edit
literature added

@@
 The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
-Therefore \textit{tolerance} is the acceptable degree of variability between execution outputs.
+\textit{Tolerance} is the acceptable degree of variability between repeated execution outputs.
 When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
@@
 We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
+\todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
+Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
+But it is important that we demonstrate knowledge of others working on this in other areas.
+{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
+\todo[inline]{We should also refer to literature like:\\
+Making System User Interactive Tests Repeatable: When and What Should We Control?~\cite{when-and-what-should-we-control} and
+Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.
+}}
 
@@
 \noindent\underline{\textit{Floating-Point Arithmetic:}}
+\todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
 Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 

### 2020-08-20 paper/archive/2_Background.tex
gc tidy up folder and tex headers

--- 2019-12-20 paper/2_Background.tex
@@
 Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution.
-However, when these engines are used for CAV development then deterministic behaviour is required.
+%However, when these engines are used for CAV development then deterministic behaviour is required.
 This section gives an overview of a gaming engine and what sources or settings in the engine may affect \textit{simulation variance}.
@@
 \subsection{Overview of a Game Loop} \label{GameLoopSection}
-The game loop is responsible for the interaction between the maths, physics and rendering engines.
-Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop (initialisation, game logic and decommissioning are removed from this diagram for brevity).
+The game loop is responsible for the interaction between the physics and rendering engines.
+Fig.~\ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop (initialisation, game logic and decommissioning are removed from this diagram for brevity\footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}).
 A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).~\cite{GameEngineArchBook}
@@
 % 
-It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below. 
+% It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below. 
 % 
-The key aspects of the game loop needed for this discussion are as follows, see Fig.~\ref{GameEngineLoopDiagram}:
+The key aspects of the game loop needed for this discussion are as follows:
 
 \begin{itemize}[leftmargin=*]
-    \item At the beginning of each tick (scene), the lag between the game clock and the real world is updated based on how much real time passed.
+    \item At the beginning of each tick, the lag between the game clock and the real world is updated based on how much real time passed.
 This measures how far the game's clock is behind compared to the real world.
     \item Then the user inputs are processed.
-    \item There is then an inner loop to update the physics processes (clock symbol in Figure~\ref{GameEngineLoopDiagram}), incrementing at a fixed delta time step, $dt$, until the game clock is equal to the real world.
+    \item There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
     \item Rendering occurs once the game clock has caught up with the real world clock. 
@@
 
-The physics engine uses a fixed time step and %, because it makes everything simpler and more stable for physics and AI.
-the shorter this fixed time step is, the smoother the interpretation of the physical dynamics will be.
+If the physics engine uses a fixed time step ($dt$), %, because it makes everything simpler and more stable for physics and AI.
+the shorter this time step is the smoother the interpretation of the physical dynamics will be.
 %processing time it takes to catch up to game clock.
@@
 This requires an update loop to take less than one render tick (one frame of real world time).
-Given the range of different hardware capabilities a variable delta time can be implemented taking the previous frame time as the next $dt$.
-However, variable $dt$ can lead to different physical results and in some cases unrealistic physical representations~\cite{gaffer}.
+Given the range of different hardware capabilities a variable delta time is often implemented for game playing taking the previous frame time as the next $dt$.
+However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}.
 Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware.
 Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}}.
-If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes.
+If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes, see lower part of Fig.~\ref{GameEngineLoopDiagram}.
 %\cite{GameEngineArchBook}\cite{GameProgPatternsBook}
@@
 In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
-and the simulation will exhibit delays between input commands and rendered states resulting in unsychronised and unrealistic behaviour.
+and the simulation will exhibit lag between input commands and rendered states resulting in unsynchronised and unrealistic behaviour.
 
@@
 
-The following points form a hypothesis around potential sources of non-determinism. \\\\
+The following points form a hypothesis around the potential sources of non-determinism that could occur in game engines. \\\\
 % \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
@@
 
-\noindent\underline{\textit{Floating Points:}}
-Floating point representation of real numbers is limited with fixed register sizes resulting in errors of \textit{rounding} where loss of bits occur and are rounded or truncated and \textit{cancellation} which happens where two similar values are subtracted
-% A generic issue with computers is floating points precision.
-Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.
-~\cite{FloatingPointsBook}\cite{goldberg1991every}. 
+\noindent\underline{\textit{Floating Point Representation:}}
+Floating point representation of real numbers is limited with fixed register sizes resulting in errors of \textit{rounding} where loss of bits occur which are rounded or truncated and \textit{cancellation} which happens where two similar values are subtracted~\cite{FloatingPointsBook,goldberg1991every}.
 %
@@
 This could result in precision issues of, for example, actor position within the environment and falsely satisfying or failing assertions.
-%; for instance consider a simulation where the world is infinite and progressively  generated in the physics engine. 
-% After a few hundred meters, precision issues will start to occur due to the coordinates used in arithmetic cacluations getting bigger, and will get progressively worse the further from the origin the simulation gets.
-Such reasons would often make one jump to conclude that floating points are the cause of non-determinism, but these precision issues should be referred to as \textit{incorrectness} and not non-determinism.
+Such reasons would often make one jump to conclude that floating points are the cause of non-determinism, but these precision issues should be referred to as \textit{incorrectness} and not non-deterministic.
 Since hardware is assumed deterministic, then the same hardware should produce the same \textit{incorrect} output for the same initial conditions.
-%, i.e.
-even if the output is incorrect due to floating point precision it should always be the same, given the same hardware.\\\\
+So, even if the output is incorrect due to floating point precision it should always be the same, given the same logical registers which conform to the same IEEE floating point standard~\cite{8766229}.
+Therefore, it is unlikely that floating point representation will contribute to \textit{simulation variance} for repeated tests on the same hardware.
 %
@@
 \\\\
-\noindent\underline{\textit{Multi-Thread Scheduling:}} Scheduling of threads may be a reason for non-deterministic simulation behaviour in game engines.
+\noindent\underline{\textit{Scheduling:}} %
+Scheduling is a resource management method for sharing computational resources between tasks of different or competing priority where tasks are executed  dependent on the scheduler policy.
+A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
+%
+If during simulation the scheduling policy was updated or thread priority was inconsistently handled then variance in output may be expected.
+It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.
+%
+%
+Similar to thread scheduling, the CPU scheduling policy determines which processor core to execute processes in the ready queue and this is maybe ordered by factors such as throughput, latency and CPU utilisation.
+%
+%
+If the same single-threaded script executes multiple times across different physical cores due to the CPU utilisation policy then should this be considered deterministic?
+Processor designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist.
+End users expect such minor variations to be controlled for in architecture design and factory testing.
+
+Another issue to consider is concurrency, where sequential tasks can be executed out of order to improve speed by utilising parallelisation although it depends on the decomposability of the program.
+This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}.
+
+Scheduling of threads may be a reason for non-deterministic simulation behaviour in game engines.
 Given the same hardware the output should always be the same, unless scheduling of the threads change.
@@
 %
-For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
+% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
 Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available.

### 2020-08-20 paper/cav_determinisim_main.tex
gc tidy up folder and tex headers

--- 2020-08-18 paper/cav_determinisim_main.tex
@@
 \documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
-\input{main.config.tex}
+% \input{main.config.tex}
+\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
+% \addbibresource{IROS.bib}
+\usepackage{array}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{color}
+\usepackage{threeparttable}
+\usepackage{balance} %balance columns
+\hyphenation{op-tical net-works semi-conduc-tor}
+\usepackage{enumitem}
+\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
+% \usepackage{subfig}
+\usepackage{subcaption}
+\usepackage{dblfloatfix}
+\usepackage{booktabs}
+\usepackage{authblk}
+\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
 % \pagenumbering{gobble} %suppresses page numbers
@@
 \begin{document}
-% %
-% \pagenumbering{arabic}
-% %
-% \input{0_TittleAndAbastract.tex}
-% %
-% \input{1_Intro.tex}
-% %
-% \input{2_Background.tex}
-% %
-% \input{3_Methodology.tex}
-% %
-% \input{4_CaseStudy.tex}
-% %
-% % \input{5_ResultsAndDiscussion.tex}
-% %
-% \input{6_Conclusions.tex}
-% %
-% \input{9_FundingAndReferences.tex}
-% %
 
@@
 % ***************************************************
-\section{Case Study of Simulation Variance} \label{s:case-study}
-
-We present an empirical investigation into using game engines for
-simulation-based verification of autonomous vehicles with a focus on
-characterising sources of non-determinism in order to understand  the impact
-they have on simulation variance. 
-%
-Ultimately, our objective is to control non-determinisim to minimise simulation variance.
-
-We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.
-%
-\todo{Complete the section structure as this is a very long section.}
-
-\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}\todo{Need to stick to scene, scenario definitions as in Intro section.}
-
-This case study draws on a setup used to verify an urban last-mile mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-%
-Thus, the scene for our investigation is the T-junction shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor trajectories over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-%
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding with or avoiding other actors.
-\todo{Decide past or present tense and be consistent throughout.}
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
-\end{figure}
-
-\todo{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-\subsection{Tolerable Simulation Variance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-%
-This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment.
-%
-Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
-
-tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-
+\section{Empirical Investigation} \label{s:case-study}
+We present an empirical investigation to illustrate some the challenges of simulation based verification outlined in Section~\ref{s:introduction} and including the potential sources of non-determinism from Section~\ref{s:nondeterminisimSources}.
+The simulation must be deterministic and by our definition this should mean that the overall level of simulation variance %$\sigma^2$
+for any actor should be less than a user-defined threshold.
+
+% \todo[inline]{Introduce the Scenario - see my comment later.}
+\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+
+Autonomous vehicles operating as mobility and last mile delivery solutions will predominantly operate within urban rather than highway or motorway environments.
+To determine if the ego vehicle performs correctly, i.e.
+does not violate assertions, the precision of the simulated actor states must be repeatable to within a defined tolerance or threshold.
+
+% \todo[inline]{KIE: Move this to the Case Study section and justify the values.}
+This threshold, termed here as the \textit{deterministic threshold}, must be chosen to reflect the scale at which repeatability of simulation execution is required, see Section~\ref{s:threshold}.
+This value must be sufficiently small to allow accurate \textit{assertion checking} of the ego vehicle in the environment.
 %
@@
 $1\mu m$.
-
-To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments. 
-%
-In particular, in this case study a determinism threshold of $1$m would be insufficient when considering the spacial resolution required to resolve situations such as potentially dangerous close passing of cyclists and pedestrians.
-A very small value, e.g.\ $1\times10^{-12}$,\todo{units for this - m(etres)?} may result in false positives due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} Therefore, for this case study, a threshold of $1$cm has been selected.
-Thus any variation of less than $1$cm is tolerable
-
-considered to be sufficient for assertion checking in urban environments. 
+%
+A deterministic threshold of $1$m is insufficient when considering the spacial resolution required to resolve scenarios such as potentially dangerous close passing of cyclists and pedestrians, and a very small value, e.g.
+$1\times10^{-12}$, may result in false positives due to the simulation \textit{noise floor}\footnote{see Fig.~\ref{CarsCollisionCG25}.} Therefore a threshold of $1$cm is chosen which is sufficient for assertion checking in urban environments.
 %
@@
 
-For the simulation to be considered repeatable, the overall level of simulation variance 
-for any actor should be less than a user-defined threshold for a given scenario and verification objective. 
-%
+\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
+
+% ***************************************************
+%  Scenario Description
+% ***************************************************
+\subsection{Scenario Description}\label{TestsDescriptionAndTechnicalities}
+A T-junction scene was used to create a number of scenarios involving pedestrians and vehicles. 
+%
+These scenarios were created in order to identify any changes in the actor trajectories over repeated tests under different conditions and hence observe any change in \textit{simulation variance}.
+%
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding with or avoiding other actors.
 
@@
 
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}
+\end{figure}
+
 % Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.

### 2020-08-20 Results_matlab_code/Results/Determinisim_ResultsV2.m

@@ -58,7 +58,7 @@ set(gca, 'YScale', 'log')
 legend("Position",[0.79 0.3 0 0])
 
 %% Plot for paper (ExperimentsSummary)
-figure(3)
+figure(3); clf;
 StressArray               = ExperimentsStressSummaryData(1,2:end);
 ID1_CarsOnly              = ExperimentsStressSummaryData(2,2:end);
 ID2_CarsCollision         = ExperimentsStressSummaryData(3,2:end);
@@ -88,39 +88,62 @@ xlabel("CPU & GPU Resource Utilisation (%)")
 ylabel("max \sigma (m)")
 set(gca,'FontSize',FontSize)
 set(gca, 'YScale', 'log')
-legend("Location","East")
+% legend("Location","East")
 
 
 %% Plotting X-Y paths
-% CarsPeopleCollision_CG0 = readmatrix("TEST_CarsPeopleCollision_CG95.txt");
-% 
-% Agent1DataIndex = CarsPeopleCollision_CG0(:,2) == 1;
-% Agent1Data      = CarsPeopleCollision_CG0(Agent1DataIndex,:);
-% Agent1Type = IdentifyType(Agent1Data(1,5),Agent1Data(1,2));
-% 
-% Agent2DataIndex = CarsPeopleCollision_CG0(:,2) == 2;
-% Agent2Data      = CarsPeopleCollision_CG0(Agent2DataIndex,:);
-% Agent2Type = IdentifyType(Agent2Data(1,5),Agent2Data(1,2));
-% 
-% Agent3DataIndex = CarsPeopleCollision_CG0(:,2) == 3;
-% Agent3Data      = CarsPeopleCollision_CG0(Agent3DataIndex,:);
-% Agent3Type = IdentifyType(Agent3Data(1,5),Agent3Data(1,2));
-% 
-% for i = 1:10:max(Agent1Data(:,1))
-% RunIndexAgent1 = Agent1Data(:,1) == i;
-% RunIndexAgent2 = Agent2Data(:,1) == i;
-% RunIndexAgent3 = Agent3Data(:,1) == i;
-% 
-% figure(1)
-% plot(Agent1Data(RunIndexAgent1,8),Agent1Data(RunIndexAgent1,9),'k.','DisplayName',Agent1Type)
-% hold on
[…]

### 2020-08-27 paper/cav_determinisim_main.tex
literature added and bib - replaced
gc changes to precision

--- 2020-08-19 paper/cav_determinisim_main.tex
@@
 \documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
-\input{main.config.tex}
+% \input{archive/main.config.tex}
+\usepackage[maxnames=6,firstinits=true,doi=false,url=false,isbn=false]{biblatex}
+% \addbibresource{IROS.bib}
+\addbibresource{cav_determinism.bib}
+\usepackage{array}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{color}
+\usepackage{threeparttable}
+\usepackage{balance} %balance columns
+\hyphenation{op-tical net-works semi-conduc-tor}
+\usepackage{enumitem}
+\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
+% \usepackage{subfig}
+\usepackage{subcaption}
+\usepackage{dblfloatfix}
+\usepackage{booktabs}
+\usepackage{authblk}
+\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
 % \pagenumbering{gobble} %suppresses page numbers
-\addbibresource{cav_determinism.bib}
 \usepackage{balance} %balance columns
@@
 \title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
+% \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
 \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
@@
 %
-This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation variance} for specific configurations and utilisation of the software and hardware, respectively.
+This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively.
 %
@@
 Potentially give away more on results to motivate the reader to continue.}
+
+\todo[inline]{A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
+}
+\todo{Review again at end. Potentially give away more on results to motivate the reader to continue.}
 
@@
 For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
-CARLA currently offers various basic types of cameras and ray-case based lidar; extensions are still under development.
-\todo{GC: what is ray-case? KIE: Not sure. :(}
+CARLA currently offers various basic types of cameras and %ray-case based
+lidar; extensions are still under development. 
+% \todo{GC: what is ray-case? KIE: Not sure. :(}
 %
@@
 % 
-Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
+% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
 Specifically, we focus on the variance of actor paths within a given scenario.
+%
+Herein we adopt the engineering terms of \textit{precision} and \textit{tolerance} to describe how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
+%
+%
+As an analogy, the simulator can be thought of as manufacturing a product, producing data traces and if tasked to repeatedly make the same product we must ask how \textit{precise} can it be and what will the \textit{tolerance} be when comparing any differences between products.
+The precision of any computer based task is casually assumed to be absolute, but technically this is equivalent to the representation of numbers within it, e.g.
+64-bit floating point \textbf{[ref]}.
+%
+%
+Real manufacturing can only aspire to achieve such precision and hence the need for tolerances on design specifications to account for the variance in real-world processes.
+If a simulator is deterministic then it will produce results with some degree of floating point precision and zero tolerance, i.e.
+with zero variance on any measure of distinction between simulation output traces.
+%
+This is not to be confused with \textit{accuracy} which describes the extent to which the simulation complies to a known standard, in many cases a simulation is held to the standard of the real world, where any divergence from this standard is termed the \textit{reality gap}.
+
+However, if the simulator is not deterministic then we must firstly ask what is the permissible limit of tolerance on the simulation output traces without affecting the verification process, and secondly how this can be measured to better inform engineers.
+
+We focus on a measure of the \textit{tolerance} given by the statistical variance of actor paths within a given scenario.
+If the variance is zero then the simulation is deterministic to the precision of the numerical representations within it.
+Actor path variance is chosen over other measures as this distance based metric forms the basis for many downstream verification tests such as assertion checking.
 
@@
 \todo[inline]{GC Edit: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable, or more formally, are \textit{precise}.
-This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
+A deterministic simulation environment guarantees that tests are repeatable.
+In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}.
+If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
+Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
+When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.}
+
+% This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
 The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
@@
 %
-If the simulation is not deterministic, however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
+If the simulation is not deterministic or has a non-permissible variance in the tolerance of, for example, actor positions, this may lead to assessment errors, making it difficult to understand and remove bugs.
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
-
-\todo[inline]{GC Edit: If the simulation is not deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
+The term \texit{simulation variance} refers to a measure of tolerance in the simulation and actor position is used as an example throughout this paper.
+
+% \todo[inline]{GC Edit: If the simulation is not deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
@@
 % Challenges of SBV
-When used for gaming, game engines do not need to be deterministic; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible tolerance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
 It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
@@
 How can one assess the extent to which a simulation environment is deterministic? 
+/todo{GC: How can one assess the extent to which a simulation environment is deterministic or has permissible tolerance?}
 
@@
 %
-In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in high simulation variance.
+In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation traces as a result of non-permissible actor position variance.
 %
@@
 Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.
-}}
+}
 

### 2020-08-27 paper/cav_determinisim_main.tex
gc change up to line 175

@@
 \usepackage{siunitx}
+\usepackage{draftwatermark}
 
@@
 %
+
 If the simulation is not deterministic or has a non-permissible variance in the tolerance of, for example, actor positions, this may lead to assessment errors, making it difficult to understand and remove bugs.
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
-The term \texit{simulation variance} refers to a measure of tolerance in the simulation and actor position is used as an example throughout this paper.
+The term \textit{simulation variance} refers to a measure of tolerance in the simulation and actor position is used as an example throughout this paper.
 
@@
 How can one assess the extent to which a simulation environment is deterministic? 
-/todo{GC: How can one assess the extent to which a simulation environment is deterministic or has permissible tolerance?}
+/todo{GC: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
 

### 2020-08-27 paper/cav_determinisim_main.tex
gc change up to line 341 and inc 451

@@
 \usepackage{siunitx}
-\usepackage{draftwatermark}
+% \usepackage{draftwatermark}
 
@@
 %
-In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation traces as a result of non-permissible actor position variance.
+In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation traces by analysis of actor position variance.
 %
 By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
+
+\todo[inline]{GC: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
+However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
 
@@
 In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
+%
+\todo[inline]{Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
+In particular, they can gain a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
+}
 %
@@
 %However, when these engines are used for AV development then deterministic behaviour is required.
+
+\todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
+Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
+But it is important that we demonstrate knowledge of others working on this in other areas.
+{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
+
+Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
+
 This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
@@
 The gamer neither wishes a faithful representation of reality, i.e.
-can be low accuracy, nor do they require for repeated actions to result in the same outcome to particular tolerance, i.e.
+can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
 can be low precision.
@@
 If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
+For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved.
 This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
@@
 We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
-\todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
+
+% MOVED TO INTRO \todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
 Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
@@
 {\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
+
 \todo[inline]{We should also refer to literature like:\\
@@
 Changing the scheduling policy and thread priorities may increase simulation variance.
-%
+
+Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core which avoids data races, as single threads cannot access shared memory.
+This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
+
 It is, however, unlikely that these would occur during repeated controlled tests for the same hardware and operating system configuration.
@@
 This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
+%
+\todo[inline]{add the ref from the google paper on using multiple threads and initialisation ordering for training ML models leads to unpredictable ordering of training data \cite{Sculley2015,Breck2017}}
 %

### 2020-08-28 paper/cav_determinisim_main.tex
gc change up to line 613

@@
 Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
-\todo[inline]{Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
+\todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
 %
@@
 A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.
-\todo{Review again at end.
-Potentially give away more on results to motivate the reader to continue.}
-
-\todo[inline]{A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
+
+\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
 }
-\todo{Review again at end. Potentially give away more on results to motivate the reader to continue.}
+\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}
 
@@
 Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
-%
-\todo{GC Replace: However, several challenges remain; one of these is determinism. with:}
 %
@@
 % GC Text
-\todo[inline]{GC Edit: Determinism is a key prerequisite for simulation during AV development and testing.
+\todo[inline,color=green!40]{GC Alternative: Determinism is a key prerequisite for simulation during AV development and testing.
 A deterministic simulation environment guarantees that tests are repeatable.
@@
 How can one assess the extent to which a simulation environment is deterministic? 
-/todo{GC: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
+\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
 
@@
 
-\todo[inline]{GC: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
+\todo[inline,color=green!40]{GC Addition: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
 However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
@@
 %
-\todo[inline]{Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
+\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
 In particular, they can gain a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
@@
 
-\todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
+\todo[inline]{GC line added to introduction.
+We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
 Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
@@
 
-Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
+\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
+}
 
@@
 
-\todo[inline]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
+\todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
@@
 
-\todo{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
+\todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
 
@@
 Making System User Interactive Tests Repeatable: When and What Should We Control?~\cite{when-and-what-should-we-control} and
-Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.
-}
+Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.}
 
@@
 \noindent\underline{\textit{Floating-Point Arithmetic:}}
-\todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
+
 Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
@@
 This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
+%
+%
+% SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
+Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition which may result in non-deterministic behaviour, and some authors suggest avoiding the use of floating point entirely for assertion testing due to imprecision~\cite{empirical-analysis-of-flaky-tests}.
 %
@@
 %
+\todo[inline,color=red!40]{GC I think all this need re-write now as we should be saying that fp could lead to non-det if issues such as non-associative addition are not controlled....
 Thus, one could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.
@@
 Therefore, it can be concluded that floating-point arithmetic does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware.
+}
 
 \noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
-ECC Memory is commonly used in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory.
+ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}.
 Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
@@
 % \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{game} which reports
+For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{Frame} which reports
 % to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
-the total time spent generating a frame, executing the game loop or render thread.
+the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
 %
@@
 
-Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core which avoids data races, as single threads cannot access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
-
-It is, however, unlikely that these would occur during repeated controlled tests for the same hardware and operating system configuration.
+It is, however, unlikely that thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
 Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
@@
 %
-\todo[inline]{add the ref from the google paper on using multiple threads and initialisation ordering for training ML models leads to unpredictable ordering of training data \cite{Sculley2015,Breck2017}}
+\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
+}
 %
@@
 However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
+
+\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot access shared memory.
+This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
+}
 
@@
 %
-Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. 
-%
-Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.\todo{So, do commercial HPC facilities have ECC memory?}
+\todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable.
+}
+%
+Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
+\todo[inline]{So, do commercial HPC facilities have ECC memory?}
+\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
 %
 To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
-\todo[inline]{To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
+\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
 %
@@
 We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.
-%
-\todo{Complete the section structure as this is a very long section.}
-
-\todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
+%
+\todo[inline]{Complete the section structure as this is a very long section.}
+
+% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
 Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
@@
 % ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}\todo{Need to stick to scene, scenario definitions as in Intro section.}
+\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
+\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
 
@@
 %
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that will result in either colliding with or avoiding other actors.
-\todo{Decide past or present tense and be consistent throughout.}
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
+\todo[inline]{Decide past or present tense and be consistent throughout.}
 
@@
 
-\todo{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
 
@@
 
+% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
+
 To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
+\todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance in the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
 %
@@
 
-tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-
+% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
 %
@@
 $1\mu m$.
-
-To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments. 
-%
-In particular, in this case study a determinism threshold of $1$m would be insufficient when considering the spacial resolution required to resolve situations such as potentially dangerous close passing of cyclists and pedestrians.
-A very small value, e.g.\ $1\times10^{-12}$,\todo{units for this - m(etres)?} may result in false positives due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} Therefore, for this case study, a threshold of $1$cm has been selected.
-Thus any variation of less than $1$cm is tolerable
-
-considered to be sufficient for assertion checking in urban environments. 
+% 
+% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
+
+% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
+%
+For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish situations such as potentially dangerous close passing of a cyclist.
+A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives.
+% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
+Therefore, for this case study, a threshold of $1$cm has been selected. Thus any variation of less than $1$cm is tolerable.
+
+% considered to be sufficient for assertion checking in urban environments. 
 %
@@
 
-Results are shown in terms of the maximum deviation, $\max\sigma$, over the entire simulation history %and the deviation threshold in this investigation is $1cm$, 
-where any value higher than the deterministic threshold is considered non-deterministic.
+\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
+where any value higher than the specified tolerance is considered non-permissible. }
 
@@
 
+\todo[inline,color=green!40]{GC question: should we remove this line now as it is repeating above statement?}
 For the simulation to be considered repeatable, the overall level of simulation variance 

### 2020-09-02 paper/cav_determinisim_main.tex
up to line 873

@@
 
-\todo[inline,color=green!40]{GC question: should we remove this line now as it is repeating above statement?}
-For the simulation to be considered repeatable, the overall level of simulation variance 
-for any actor should be less than a user-defined threshold for a given scenario and verification objective. 
+% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
+For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
 %
@@
 %
-Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering a \textit{event hit} callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m \todo[inline]{AG: is this value correct or should it be 2.0m?
+Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering an \textit{event hit} callback in the game engine (test 2).
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
+% \todo[inline]{AG: is this value correct or should it be 2.0m?
 AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
-actors would ignore trajectory waypoints that are less than 2.0m away from them.} intervals using a PID controller.
+actors would ignore trajectory waypoints that are less than 2.0m away from them.}
+intervals using a PID controller.
 In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
@@
 \subsection{Experiment Description}\label{s:Experiment_Description}
-\todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time.
-Repeated traces are compared at the same time index to provide a value for the variance of actor $a$ with respect to time $t$, $\sigma_a(t)$.
+% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
+\todo[inline,color=green!40]{GC Edit: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
+%
+Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance of actor position $\sigma_a^{2}(t)$. 
+%
+Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
+}
+%
 The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, 
 \begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a(t), 
+\max_{a,t}\sigma_a^{2}(t), 
 \end{equation}
-herein referred to as $\max\sigma$ for brevity.
+and therefore the maximum deviation is simply square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity.
+
 % The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
 % The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
-The $\max\sigma$ value was analysed against the different scenarios and settings that were identified as potential sources of simulation variance, and compared against the \textit{deterministic threshold} to indicate if the simulation was deterministic or not.
+\todo[inline,color=green!40]{GC Edit: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
+}
 %
 % Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
-\todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
+% \todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
 
@@
 % ***************************************************
-\subsection{External Settings}\todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
-
-After discovering that internal engine settings could not achieve a simulation variance that met the required \textit{deterministic threshold}, settings external to the game engine were explored. 
+\subsection{External Settings}
+% \todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
+
+After discovering that internal engine settings could not achieve a simulation variance that met the required 
+% \textit{deterministic threshold}, 
+\textit{permissible tolerance}, 
+settings external to the game engine were explored. 
 %
@@
 \subsection{System Configuration and Screening}\label{s:screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 graphics card with 8GB GDDR6 RAM at 1515 MHz.
+These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz.
 Operating systems was Linux Ubuntu 18.04.4 LTS.
@@
 %
-From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) are consistently deterministic over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the deterministic threshold that has been set, with some deviation in actor path as large as $59$cm.
+\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
+}
 %
@@
 Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, anything above the $1$cm threshold (dashed line) is considered non-deterministic according to our specific threshold level informed from our verification requirements.
+In this figure, anything above the $1$cm threshold (dashed line) is considered non-permissible according to our specific threshold level informed from our verification requirements.
 %
@@
 
-However, the non-deterministic results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-%
-Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from deterministic prior to the vehicle collision to non-deterministic post-collision.
-This pattern of deterministic behaviour prior to collision and non-deterministic post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+However, the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
+%
+\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
+%
+This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
 This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates deterministic pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-deterministic.
+Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
+}
 % \todo{Why ``not reliable''.}
@@
     \end{subfigure}
-    \caption{Vehicle to pedestrian collision (test 4) showing (a) mean deviation against simulation time for 25\% stress and (b) mean deviation pre- and post-collision for different resource utilisation.}
+    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% stress and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
 \end{figure*}
@@
 In this scenario (test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision as described above.
-Secondly, that the pedestrian involved in the collision (Ped ID 3) displays deterministic behaviour throughout the test\footnote{Note that in CARLA the pedestrian object is destroyed post-collision hence the flat line following the collision.}.
+There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.% as described above.
+Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6s$ onwards.}.
 Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that \textit{any collision between actors can affect the determinism of the entire actor population} and could potentially result in  erroneous assertion checking.
+This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous assertion checking.
 
@@
 The main findings of this work suggest a working practice that would minimise the non-deterministic effects observed in this investigation.
-By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the deterministic threshold of $1$cm is achievable as shown in the restricted column in Table.~\ref{TableOfExperiments}.
+By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible tolerance threshold of $1$cm is achievable as shown in the \textit{restricted} column in Table.~\ref{TableOfExperiments}.
+By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98cm$ which is within the target tolerance we set out to achieve.
+%
+\todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
+For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm.
+}
 
@@
 However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly.
-Therefore we present a general methodology that practitioners can follow to find the \textit{deterministic domains of operation} for any hardware configuration.
+Therefore we present a general methodology that practitioners can follow to find the
+% \textit{deterministic domains of operation} 
+\textit{operational domains of permissible tolerance} 
+for any hardware configuration.
 

### 2020-09-03 paper/cav_determinisim_main.tex
added ref to GPU FP

@@
 % SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition which may result in non-deterministic behaviour, and some authors suggest avoiding the use of floating point entirely for assertion testing due to imprecision~\cite{empirical-analysis-of-flaky-tests}.
+\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
+
+Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
+Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
+Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
+%
+Here we are assuming these calculations are performed on a single core and therefore concurrency of multiple threads is not an issue which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.}
+
 %
@@
 %
-\todo[inline,color=red!40]{GC I think all this need re-write now as we should be saying that fp could lead to non-det if issues such as non-associative addition are not controlled....
 Thus, one could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.
@@
 %
-Therefore, it can be concluded that floating-point arithmetic does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware.
-}
+Therefore, it can be concluded that floating-point arithmetic does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware with the same configuration.
 
@@
 
-\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot access shared memory.
+\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
@@
 %
-\todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable.
+% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable.
 }
@@
 
+\todo[inline,color=green!40]{GC Addition (para below):} 
+%Issues around floating point precision and concurrency as previously mentioned notwithstanding, 
+Practitioners should also be away that many libraries for calculating variance itself may require attention to get precise results.
+For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
+Even in \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision\footnote{\url{https://uk.mathworks.com/matlabcentral/answers/317128-does-matlab-use-all-cores-by-default-when-running-a-program}}.
+
 % ***************************************************
@@
 There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.% as described above.
-Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6s$ onwards.}.
+Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}.
 Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
@@
 By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible tolerance threshold of $1$cm is achievable as shown in the \textit{restricted} column in Table.~\ref{TableOfExperiments}.
-By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98cm$ which is within the target tolerance we set out to achieve.
+By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve.
 %
@@
 \section{Methodology} \label{s:methodology}
-In this section a method for determining the variance of simulated actor path trajectories, termed \textit{simulation variance}, and resolving the deterministic operational domains of the simulation are presented.
+In this section a method for determining the variance of simulated actor path trajectories, termed \textit{simulation variance}, and resolving the operational domains of permissible tolerance of the simulation are presented.
 %
@@
 \subsubsection{Trajectories and Collisions} 
-Actor paths should hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
-
-\subsubsection{Setting a Deterministic Threshold} \label{s:threshold}
-The verification engineer should set the \textit{deterministic threshold} based on their specific requirements which refers to the acceptable level of actor path variance ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
-This threshold is analogous to the analogue signal level associated with each of the binary states of the signals in digital circuits.
+Actor paths should be hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
+
+\subsubsection{Setting a Permissible Tolerance} \label{s:threshold}
+The verification engineer should set the \textit{permissible tolerance} based on their specific requirements which refers to the acceptable level of actor path variance.
+% ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
+This tolerance level is analogous to the analogue signal level associated with each of the binary states %of the signals 
+in digital circuits.
 At some low analogue voltage the signal is interpreted as binary $0$, this may be, for example, any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
-This convention enables the abstraction from the physical signal values that are analogue voltages into the digital domain with binary representation.
-%
-The deterministic threshold that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-In particular, this threshold must be sufficiently small to enable accurate assertion checking.
+This convention enables the abstraction from the noisy physical signal values that are analogue voltages into the digital domain with binary representation.
+%
+The permissible tolerance that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+For example, this threshold must be sufficiently small to enable accurate assertion checking.
 In our empirical investigation we set a threshold of $1$cm which is sufficient for urban scenario assertion checking.
@@
 \subsubsection{Simulation Length} 
-The simulation time should be sufficient to record interactions between actors but not so long that 100's or 1000's of repeated tests take an inconvenient time to complete.
+The simulation time should be sufficient to record interactions between actors but not so long that 
+%100's or 1000's 
+a statistically significant number 
+of repeated tests take an inconvenient time to complete.
 In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
@@
 % To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-The resources available to the simulator have been shown to have a significant effect on the determinism of simulated vehicles and exploring this as a variable the analysis is required. 
+The resources available to the simulator have been shown to have a significant effect on the simulation variance of simulated vehicles and exploring this as a variable the analysis is required. 
 CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program can be used spawn workers on any number of cores or virtual threads on a system.

### 2020-09-03 paper/cav_determinisim_main.tex
final few changes

@@
 This memory testing software can also be used for ECC enabled hardware.
-Similarly to test memory on Nvidia based graphical adaptors \texttt{cuda\_memtest} can be used to ensure no memory errors exist.
+Similarly, to test memory on Nvidia based graphical adaptors \texttt{cuda\_memtest} can be used to ensure no memory errors exist.
 
@@
 While executing 100 repeats may seem sufficient, this sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
-It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the deterministic threshold.
+It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the permissible tolerance.
 % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
@@
 Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-Additional information can also be logged such as the CPU and GPU utilisation levels.
+Additional information can also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
+% such as the \texttt{unit} command in Unreal Engine. 
 For the subsequent analysis, the actor should have an identifier and the repeat number and simulation time should also be captured. 
@@
 \subsection{Analysis}
-The maximum value of simulation variance over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control for deterministic operation.
-Finding these boundaries will identify the \textit{deterministic operational domain} for the user specific hardware.
+The maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control for permissible operation.
+Finding these boundaries will identify the \textit{domains of permissible tolerance} for the user specific hardware.
 
@@
 During the study we hypothesized and uncovered several parameters that contribute towards greater simulation variance and hence non-deterministic execution, such as actor collisions and system resource utilisation.
-The results of the investigation are hardware and software version specific, so we present a universal methodology that can be used to find the \textit{deterministic operational domain} for any given system configuration.
+The results of the investigation are hardware and software version specific, so we present a universal methodology that can be used to find the \textit{domains of permissible tolerance} for any given system configuration.
 This methodology will allow the AV verification community using simulation tools to ensure their results are reliable and trustworthy.

### 2020-10-15 paper/cav_determinisim_main.tex
Pushing changes from discussion with Greg.

@@
 \usepackage{todonotes}
+% \usepackage[disable]{todonotes}
 \usepackage{siunitx}
@@
 %
-A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.
+\todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.
+}
+A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
 
 \todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
+}
+\todo[inline,color=yellow!40]{AG Alternative: A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
 }
@@
 %
-Herein we adopt the engineering terms of \textit{precision} and \textit{tolerance} to describe how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
+Herein we adopt the engineering terms of \textit{precision} and \textit{tolerance} to describe how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}.
 %
@@
 % KE Text
+\todo[inline,color=red!40]{old: Determinism is a key prerequisite for simulation during AV development and testing.
+A deterministic simulation environment guarantees that tests are repeatable.
+Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
+}
+%
 Determinism is a key prerequisite for simulation during AV development and testing.
 A deterministic simulation environment guarantees that tests are repeatable.
-Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}.
+If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
+Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
+When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
 This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
@@
 Thus, our main research questions is:
-How can one assess the extent to which a simulation environment is deterministic? 
+\todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
+How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?
 \todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
@@
 %
-By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
-
-\todo[inline,color=green!40]{GC Addition: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
+\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
+}
+%
+By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
+However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
+%
+\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
 However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
@@
 %
-Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
+\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
 In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
-In particular, they can gain a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
+}
+%
+Knowing the simulation variance will help verifcation engineers assess the impact that using a game engine for AV simulation may have.
+In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
+%
+\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verifcation engineers assess the impact that using a game engine for AV simulation may have.
+In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
 }
@@
 %However, when these engines are used for AV development then deterministic behaviour is required.
-
+%
+Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
+%
 \todo[inline]{GC line added to introduction.
@@
 {\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-
+%
 \todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
 }
-
+%
 This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
@@
 This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
-
-Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
+%
+\todo[inline,color=red!40]{old: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy.
@@
 This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-
+}
+%
+Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
+Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
+The gamer neither wishes a faithful representation of reality, i.e.
+can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
+can be low precision.
+In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
+If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
+For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved.
+This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
 \todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
@@
 This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
-
+%
 \todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
@@
 % SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
-
 Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
@@
 %
-Here we are assuming these calculations are performed on a single core and therefore concurrency of multiple threads is not an issue which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.}
-
+If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
+\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
+
+Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
+Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
+Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
+%
+If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
+% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.}
+}
 %
@@
 %
+Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
+%
 \todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
@@
 
+Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
+%
 \todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
@@
 %
-To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
+\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
+To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
 \todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
@@
 
-We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.
+\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
+We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
 \todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
@@
 %
-Thus, the scene for our investigation is the T-junction shown in Figure~\ref{f:test_a_and_b}.
+Thus, the scene for our investigation is the T-junction and the scenarios are shown in Figure~\ref{f:test_a_and_b}.
 
@@
 To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-\todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance in the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
+%
+Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance?
+%
+\todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
 %
@@
 % due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study, a threshold of $1$cm has been selected. Thus any variation of less than $1$cm is tolerable.
+Therefore, for this case study and more broadly our general verifcation requirements, a threshold of $1$cm has been selected. Thus any variation of less than $1$cm is tolerable.
 
@@
 
+% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
+
+\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
+}
+
+% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
+Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
+%
 \todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
 where any value higher than the specified tolerance is considered non-permissible. }
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
-For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-%
 
@@
 % \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-\todo[inline,color=green!40]{GC Edit: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance of actor position $\sigma_a^{2}(t)$. 
+For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
+%
+Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
+%
+Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
+\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
+%
+Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
 %
@@
 
+The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
 % The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
 % The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
-\todo[inline,color=green!40]{GC Edit: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
+\todo[inline,color=green!40]{GC Addition: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
 }
@@
 %
+From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
 \todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
@@
 However, the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-%
+
+Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
+%
+This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
 \todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
@@
 In this scenario (test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three key observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.% as described above.
+There are three observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
+% as described above. 
 Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}.
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v3.pdf}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v4.pdf}
     \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}
@@
 %
+Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application. 
+%
 \todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm.
+%For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm. 
 }
@@
 Initial screening tests done with NUMA control gave moderate improvements in simulation variance of the order of a few percent, see Section~\ref{s:screening}.
-This control may assist if simulation variance is borderline to the deterministic threshold but not seen as essential.
+This control may assist if simulation variance is borderline to the permissible tolerance but not seen as essential.
 
@@
 The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
-Having a deterministic simulator is required for reliable and repeatable results, \todo[inline]{Having a deterministic simulator is required for precise results,} to find and fix software bugs and ensure results are trustworthy.
+%
+\todo[inline,color=red!40]{old:Having a deterministic simulator is required for reliable and repeatable results,}
+Having a deterministic simulator is required for precise results, 
+\todo[inline]{Having a deterministic simulator is required for precise results,} 
+%
+to find and fix software bugs and ensure results are trustworthy. 
+%
 If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains for reliable results.
 During an investigation into the CARLA simulator, a \textit{simulation variance} was observed for repeated tests with the same initial conditions indicating non-deterministic execution.
+%
 During the study we hypothesized and uncovered several parameters that contribute towards greater simulation variance and hence non-deterministic execution, such as actor collisions and system resource utilisation.
+%
 The results of the investigation are hardware and software version specific, so we present a universal methodology that can be used to find the \textit{domains of permissible tolerance} for any given system configuration.
+%
 This methodology will allow the AV verification community using simulation tools to ensure their results are reliable and trustworthy.

### 2020-10-20 paper/cav_determinisim_main.tex
added refs 11 and 40

@@
 %
-Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
+Ensuring software operates deterministically is a non-trivial task and catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
 %
@@
 We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
+%
+\todo[inline,color=green!40]{GC Addition below}%
+%
+A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests} although the CAV simulation domain introduces it's own unique challenges that are not considered in this paper.
 
@@
 %
+\todo[inline,color=green!40]{GC new: line below.}
+%
+Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
+%
 Ultimately, our objective is to control non-determinisim to minimise simulation variance.
@@
 \todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
+%
 We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
+%
 \todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}

### 2020-11-09 paper/cav_determinisim_main.tex
commented out cyres reference
introduction definitions added

@@
 \usepackage{balance} %balance columns
-\usepackage{todonotes}
-% \usepackage[disable]{todonotes}
+% \usepackage{todonotes}
+\usepackage[disable]{todonotes}
 \usepackage{siunitx}
@@
 %
+A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
+%
 Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
@@
 There have been a number of fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-Simulation environments offer a means to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving}.
+Simulation environments offer a means to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
 In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
@@
 %
-CARLA\footnote{\url{http://carla.org/}}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
+CARLA\footnote{\url{http://carla.org/}}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
 
@@
 %
-Furthermore, they are easy to setup and run; they also are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
+Furthermore, they are easy to setup and run with respect to on-road testing and are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
 %
 %\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
-
-For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
+% 
+% GC Removing the following para:
+%For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
 CARLA currently offers various basic types of cameras and %ray-case based
-lidar; extensions are still under development. 
+% lidar; extensions are still under development. 
 % \todo{GC: what is ray-case? KIE: Not sure. :(}
 %
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013,cyres} may also be required. 
+Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. %~\cite{Javaid2013,cyres}
 %
 % The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
-
+%
 % Game engines for SBV
@@
 However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
-%
+
+\todo[inline,color=green!40]{GC New below}
+
+\subsection{Definition: Determinism}
 % What is a deterministic system
-In this context, deterministic refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
+Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
+However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
+Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
 % 
@@
 Specifically, we focus on the variance of actor paths within a given scenario.
-%
-Herein we adopt the engineering terms of \textit{precision} and \textit{tolerance} to describe how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}.
-%
-%
-As an analogy, the simulator can be thought of as manufacturing a product, producing data traces and if tasked to repeatedly make the same product we must ask how \textit{precise} can it be and what will the \textit{tolerance} be when comparing any differences between products.
-The precision of any computer based task is casually assumed to be absolute, but technically this is equivalent to the representation of numbers within it, e.g.
-64-bit floating point \textbf{[ref]}.
-%
-%
-Real manufacturing can only aspire to achieve such precision and hence the need for tolerances on design specifications to account for the variance in real-world processes.
-If a simulator is deterministic then it will produce results with some degree of floating point precision and zero tolerance, i.e.
-with zero variance on any measure of distinction between simulation output traces.
-%
-This is not to be confused with \textit{accuracy} which describes the extent to which the simulation complies to a known standard, in many cases a simulation is held to the standard of the real world, where any divergence from this standard is termed the \textit{reality gap}.
-
-However, if the simulator is not deterministic then we must firstly ask what is the permissible limit of tolerance on the simulation output traces without affecting the verification process, and secondly how this can be measured to better inform engineers.
-
-We focus on a measure of the \textit{tolerance} given by the statistical variance of actor paths within a given scenario.
-If the variance is zero then the simulation is deterministic to the precision of the numerical representations within it.
+
+A simulation can be thought of as the generation or production of experimental data. 
+%
+In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data. 
+%
+If a simulation is deterministic then there will be no variation in the generated output data, i.e. all future states are perfectly reproducible from prior states and actions. 
+%
+However, if a simulation is not deterministic then then there will be a variation in the generated output data.
+
+\subsection{Definitions: Precision, Variance \& Tolerance}
+
+We adopt some terms from the mechanical engineering and statistics domain that best describe this situation~\cite{ADictionaryofMechanicalEngineering}.
+%
+\textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
+%
+\textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
+%
+\textit{Tolerance} is defined as the permissible limit of the variance.
+
+% The engineering terms \textit{precision} and \textit{tolerance} suitably describe how the simulation data vary when the same repeated stimulus is provided to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}.
+%
+%
+% As an analogy, the simulator can be thought of as a manufacturing process, producing data and if tasked to repeatedly execute the same process we must ask how \textit{precise} can this process be and what, if any, will the \textit{variance} be?
+% when comparing the differences between multiple products.
+As an analogy, the simulator can be thought of as a manufacturing process, producing data and to find the value of the precision then the output must be measured for differences.
+Those differences describe the spread or variance in the process output and a hard limit on the variance can be defined, beyond which the output fails to meet the required tolerance, e.g.
+rejected for quality control.
+%
+% In relation to manufacturing, the precision of any computer based task is casually assumed to be absolute, for example one may question the manufactured precision of a physical component but not the computer assisted drawing that it was made from.
+%but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point. %\textbf{[ref]}. %
+%
+Real manufacturing fails to achieve absolute precision and hence the need for tolerances on design specifications to account for the variance in real-world processes.
+
+If a simulator is deterministic then it will produce results with absolute precision or zero variance and hence will be within acceptable tolerance.
+If the simulator is non-deterministic then there will be a variance in the output which can be measured.
+%
+
+\subsection{Definition: Accuracy}
+
+Precision and tolerance are not to be confused with \textit{accuracy} which describes how closely the generated output of a process aligns to a known standard.
+We therefore define accuracy as the difference between the true value (or reference value) and what has been achieved in the simulation or generation process.
+%
+For a driving simulation the reference value, for which accuracy can be measured against, may be the real world that it seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}.
+In most cases such accuracy will not be possible due to computational demands of such an exact replica and in most cases is unnecessary, where some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
+
+\subsection{Definition: Variance Measurement}
+
+If the simulator is not deterministic then how can this variance be measured?
+%
+This can be achieved from the actor state, or more specifically from the statistical variance of actor paths within a given scenario with a fixed set of actions.
 Actor path variance is chosen over other measures as this distance based metric forms the basis for many downstream verification tests such as assertion checking.
+%
+% Within this paper the term tolerance is a synonym for the permissible limit of variance.
+
+\subsection{Definition: Scene, Scenario \& Situation}
 
@@
 %
+
+\subsection{Why is Determinism needed?}
 Determinism is a key prerequisite for simulation during AV development and testing.

### 2020-11-10 paper/cav_determinisim_main.tex
GC revisions to Section 4

@@
 % \input{archive/main.config.tex}
-\usepackage[maxnames=6,firstinits=true,doi=false,url=false,isbn=false]{biblatex}
+\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
 % \addbibresource{IROS.bib}
@@
 %
-However, if a simulation is not deterministic then then there will be a variation in the generated output data.
+However, if a simulation is non-deterministic then then there will be a variation in the generated output data.
 
@@
 
-If the simulator is not deterministic then how can this variance be measured?
+If the simulator is non-deterministic then how can this variance be measured?
 %
@@
 Actor path variance is chosen over other measures as this distance based metric forms the basis for many downstream verification tests such as assertion checking.
+
+The term \textit{simulation variance} refers to a measure of actor state variance in the simulation assuming fixed actions.
+Results are referred to in terms of deviation, rather than variance, as this is a more intuitive dimension to comprehend when considering verification tests, e.g.
+vehicle close-passing.
+Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.
+
 %
@@
 
-\subsection{Why is Determinism needed?}
+\subsection{When is Determinism needed?}
 Determinism is a key prerequisite for simulation during AV development and testing.
 A deterministic simulation environment guarantees that tests are repeatable.
-In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}.
-If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
+Using mechanical engineering terms, a deterministic simulator can be considered to have zero \textit{variance}.
+%
+If the variance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
 Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
@@
 
-If the simulation is not deterministic or has a non-permissible variance in the tolerance of, for example, actor positions, this may lead to assessment errors, making it difficult to understand and remove bugs.
+If the simulation is non-deterministic and has a non-permissible variance in, for example, actor positions, this may lead to assessment errors, making it difficult to understand and remove bugs.
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
-The term \textit{simulation variance} refers to a measure of tolerance in the simulation and actor position is used as an example throughout this paper.
-
-% \todo[inline]{GC Edit: If the simulation is not deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
+%
+
+% \todo[inline]{GC Edit: If the simulation is non-deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
@@
 % Challenges of SBV
-When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible tolerance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
 It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
@@
 Thus, our main research questions is:
+%
 \todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
-How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?
+%
+How can one assess the extent to which a simulation environment is deterministic or has a permissible variance?
 \todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
 
-In pursuit of an answer to this question we have investigated how non-determinism of a game engine, namely Unreal Engine that powers CARLA, when used for simulation-based AV verification affects the simulation results.
-%
-In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation traces by analysis of actor position variance.
+% In pursuit of an answer to this question we
+In this paper we investigated how the non-determinism of Carla, based on the Unreal game engine, when used for simulation-based AV verification affects the simulation results. 
+%
+In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation output by analysis of actor position variance.
 %
@@
 %
-By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
-However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
-%
-\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is not deterministic under certain conditions.
+By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
+% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
+%
+Carla exhibits a permissible level of variance when system utilisation is restricted to 75\% or less and ensuring termination of scenarios once a vehicle collision has been detected.
+%
+\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
 However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
 
-The insights gained from this case study have motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
+The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
 %
@@
 %
-Knowing the simulation variance will help verifcation engineers assess the impact that using a game engine for AV simulation may have.
-In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verifcation engineers assess the impact that using a game engine for AV simulation may have.
+Knowing the simulation variance will help assess the impact that using a game engine for AV simulation may have on verification tasks.
+In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact on verification results.
+%
+\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
 In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
@@
 % ======= Overview of a Game Loop
-\subsection{Overview of a Game Loop} \label{GameLoopSection}
+\subsection{Game Engine Overview} \label{GameLoopSection}
 
@@
 The gamer neither wishes a faithful representation of reality, i.e.
-can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
-can be low precision.
+the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
+can be low \textit{precision}.
+%
 In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
 If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved.
+For example, for perception stack testing the sensors need to get input that is repeatable so that if any software bugs are found they can be re-played and the issue resolved.
 This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
+%
 \todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
@@
 
-The following discussion reflects the results of our investigation into the potential sources of non-determinism that could affect game engines.
+% The following discussion reflects the results of our investigation into the potential sources of non-determinism that could affect game engines.
 We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
+The following review discusses the potential sources of non-determinism that were found in the literature or found as part of our investigation into game engines.
+We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
 %
@@
 %
-A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests} although the CAV simulation domain introduces it's own unique challenges that are not considered in this paper.
+A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests} although the AV simulation domain introduces it's own unique challenges that are not considered in this paper.
 
@@
 %
-If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
+If calculations are performed on a single core, then concurrency of multiple threads is not an issue.
 \todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
@@
 
-Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+Indeed, the developers of the de-bugging program \texttt{RR}\footnote{\url{https://github.com/mozilla/rr}} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
@@
 
-% ***************************************************
-%  Empirical Investigation
-% ***************************************************
-\section{Case Study of Simulation Variance} \label{s:case-study}
-
-We present an empirical investigation into using game engines for
-simulation-based verification of autonomous vehicles with a focus on
-characterising sources of non-determinism in order to understand  the impact
-they have on simulation variance. 
-%
-\todo[inline,color=green!40]{GC new: line below.}
-%
-Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
-%
-Ultimately, our objective is to control non-determinisim to minimise simulation variance.
-
-\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
-%
-We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-%
-\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
-%
-\todo[inline]{Complete the section structure as this is a very long section.}
-
-% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
-
-This case study draws on a setup used to verify an urban last-mile mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-%
-Thus, the scene for our investigation is the T-junction and the scenarios are shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor trajectories over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-%
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-\todo[inline]{Decide past or present tense and be consistent throughout.}
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
-\end{figure}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-\subsection{Tolerable Simulation Variance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
-
-To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-%
-Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance?
-%
-\todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
-%
-This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment.
-%
-Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
-
-% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-%
-% but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-% 
-% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
-%
-For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish situations such as potentially dangerous close passing of a cyclist.
-A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives.
-% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study and more broadly our general verifcation requirements, a threshold of $1$cm has been selected. Thus any variation of less than $1$cm is tolerable.
-
-% considered to be sufficient for assertion checking in urban environments. 
-%
-% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
-Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-%
-\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
-where any value higher than the specified tolerance is considered non-permissible. }
-
-% ***************************************************
-%  Actor Collisions
-% ***************************************************
-\subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Collision/Overview/index.html}}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}.
-%
-Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.
-
-The tests are listed in Table~\ref{TableOfExperiments} and were chosen to cover a range of interactions between actor types. 
-%
-Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering an \textit{event hit} callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
-% \todo[inline]{AG: is this value correct or should it be 2.0m?
-AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
-actors would ignore trajectory waypoints that are less than 2.0m away from them.}
-intervals using a PID controller.
-In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
-In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a \textit{event hit} callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
-Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
-
-\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
-
-% ***************************************************
-%  Experiment Description
-% ***************************************************
-\subsection{Experiment Description}\label{s:Experiment_Description}
-% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-}
-%
-The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, 
-\begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a^{2}(t), 
-\end{equation}
-and therefore the maximum deviation is simply square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity.
-
-The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
-% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
-% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
-\todo[inline,color=green!40]{GC Addition: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
-}
-%
-% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
-% \todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
-
-% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
-Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
-
-% This whole process is repeated for various internal and external settings (if needed). 
-% Then various plots can be created to draw the engine's simulation variance. 
-% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
-
-% ***************************************************
-%  Internal Settings
-% ***************************************************
-\subsection{Internal Settings}
-
-Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
-Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.
-\textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.
-\textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}.
-% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
-An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
-The results were compared to a baseline of the default engine settings.
-%
-However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
-Details on this previous investigation can be found on the Trustworthy Systems github~\cite{TSLUnrealEngineTesting}.
-
 \begin{table*}[b]
@@
 % ***************************************************
+%  Empirical Investigation
+% ***************************************************
+\section{Case Study of Simulation Variance} \label{s:case-study}
+
+We present an empirical investigation into using game engines for
+simulation-based verification of autonomous vehicles with a focus on
+characterising sources of non-determinism in order to understand  the impact
+they have on simulation variance. 
+%
+\todo[inline,color=green!40]{GC new: line below.}
+%
+Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
+%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
+%
+Ultimately, our objective is to control non-determinisim to minimise simulation variance.
+
+\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
+%
+We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
+%
+\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
+%
+\todo[inline]{Complete the section structure as this is a very long section.}
+
+% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
+
+% ***************************************************
+%  Scenario Description
+% ***************************************************
+\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
+\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
+
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
+\end{figure}
+
+This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
+%
+Thus, the scene for our investigation is the T-junction (T-intersection) and the scenarios are shown in Figure~\ref{f:test_a_and_b}.
+
+This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor trajectories over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
+%
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
+\todo[inline]{Decide past or present tense and be consistent throughout.}
+
+\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+
+\subsection{Tolerable Simulation Variance}
+
+\todo[inline]{Terminology:\\
+{\bf Accuracy} is the degree of closeness/conformance to a known value.
+\\
+{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
+\\
+{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
+\\
+See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
+}
+
+% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
+
+To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
+%
+Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
+%
+% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
+%
+This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment.
+%
+Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
+
+% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
+%
+% but a larger value may be acceptable for non-urban environments, e.g.
+motorway or highway.
+Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
+$1\mu m$.
+% 
+% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
+
+% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
+%
+For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish situations such as potentially dangerous close passing of a cyclist.
+A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives.
+% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
+Therefore, for this case study and more broadly our general verifcation requirements, a threshold of $1$cm has been selected. 
+% Thus any variation of less than $1$cm is tolerable. 
+Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
+
+% considered to be sufficient for assertion checking in urban environments. 
+%
+% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
+
+% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
+
+\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
+}
+
+% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
+Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
+%
+\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
+where any value higher than the specified tolerance is considered non-permissible. }
+
+% ***************************************************
+%  Actor Collisions
+% ***************************************************
+
+\subsection{Actor Collisions}\label{S:Actor_Collisions}
+Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Collision/Overview/index.html}}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}.
+%
+Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.
+
+The tests are listed in Table~\ref{TableOfExperiments} and were chosen to cover a range of interactions between actor types. 
+%
+Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering an \textit{event hit} callback in the game engine (test 2).
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
+% \todo[inline]{AG: is this value correct or should it be 2.0m?
+AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
+actors would ignore trajectory waypoints that are less than 2.0m away from them.}
+intervals using a PID controller.
+In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
+In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a \textit{event hit} callback, see Fig~\ref{Test_a}.
+Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
+Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
+
+\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
+
+% ***************************************************
+%  Experiment Description
+% ***************************************************
+\subsection{Experiment Description}\label{s:Experiment_Description}
+% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
+For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
+%
+Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
+%
+Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
+\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
+%
+Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
+%
+Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
+}
+%
+The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, 
+\begin{equation} \label{eq:max_sigma}
+\max_{a,t}\sigma_a^{2}(t), 
+\end{equation}
+and therefore the maximum deviation is simply square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity.
+
+The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
+% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
+% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
+\todo[inline,color=green!40]{GC Addition: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
+}
+%
+% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
+% \todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
+
+% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
+Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
+
+% This whole process is repeated for various internal and external settings (if needed). 
+% Then various plots can be created to draw the engine's simulation variance. 
+% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
+
+% ***************************************************
+%  Internal Settings
+% ***************************************************
+\subsection{Internal Settings}
+
+Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
+Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.
+\textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.
+\textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
+Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking.\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}
+% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
+An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
+The results were compared to a baseline of the default engine settings.
+%
+However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
+Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.
+
+% ***************************************************
 %  External Settings
@@
 %Issues around floating point precision and concurrency as previously mentioned notwithstanding, 
-Practitioners should also be away that many libraries for calculating variance itself may require attention to get precise results.
+Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
 For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
-Even in \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision\footnote{\url{https://uk.mathworks.com/matlabcentral/answers/317128-does-matlab-use-all-cores-by-default-when-running-a-program}}.
-
-% ***************************************************
-%  System Configuration and Screening
-% ***************************************************
-\subsection{System Configuration and Screening}\label{s:screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz.
-Operating systems was Linux Ubuntu 18.04.4 LTS.
-To ensure reliability of results, each test was repeated 1000 times.
-Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
-These were:
-
-\begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
-    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
-\end{itemize}
-
-% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
-These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
-
-% \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
-
-% Our hypothesis is that since UE4 is a game engine it is not deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
-This behaviour is expected not to hold as the level of computing stress increases.
-
- % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
-To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
-(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
-
-% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
-
-% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
-
-% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
-
-% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
-
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
-
-% ***************************************************
-%  Results and Discussion
-% ***************************************************
-\subsection{Results and Discussion}\label{ResultsSection}
-A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted), where the value reported is the maximum deviation across all resource utilisation levels, i.e.
-the worst case for a given scenario.
-%
-From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-}
-%
-Clearly a deviation in results of such a large amount is unacceptable if simulation is to be used as a reliable verification tool.
-
-Resource utilisation level was found to have a significant impact on \textit{simulation variance}.
-%
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, anything above the $1$cm threshold (dashed line) is considered non-permissible according to our specific threshold level informed from our verification requirements.
-%
-\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
-%
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
-    \label{ExperimentsStressSummary}
-\end{figure}
-
-A general pattern in the results indicates that; some scenarios fail to produce deterministic results consistently against any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm threshold), some are consistently below the threshold (Test ID 5 \& 6 with pedestrians only), and that some cases only fail at higher utilisation levels, breaching the threshold above 75\% resource utilisation (Test ID 1 \& 3).
-
-\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
-
-\begin{figure*}[h]
+In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision.\footnote{\url{https://uk.mathworks.com/matlabcentral/answers/317128-does-matlab-use-all-cores-by-default-when-running-a-program}}
+
+\begin{figure*}[b]
     \centering
@@
 
-However, the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per definition \ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
+% ***************************************************
+%  System Configuration and Screening
+% ***************************************************
+\subsection{System Configuration and Screening}\label{s:screening}
+These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz.
+Operating systems was Linux Ubuntu 18.04.4 LTS.
+To ensure reliability of results, each test was repeated 1000 times.
+Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
+To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
+These were:
+
+\begin{itemize}[leftmargin=*]
+    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
+    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
+    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
+\end{itemize}
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
+    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
+    \label{ExperimentsStressSummary}
+\end{figure}
+
+% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
+These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
+Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
+We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
+
+% \subsection{Problem And Hypothesis}
+% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+
+% Our hypothesis is that since UE4 is a game engine it is non-deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
+This behaviour is expected not to hold as the level of computing stress increases.
+
+ % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
+To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
+(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
+
+% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
+In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
+
+% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
+Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
+
+% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
+
+% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
+These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
+
+% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
+
+% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+
+% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
+Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
+
+% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+
+% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+
+% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
+
+% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
+
+% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
+
+% ***************************************************
+%  Results and Discussion
+% ***************************************************
+\subsection{Results and Discussion}\label{ResultsSection}
+A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted), where the value reported is the maximum deviation across all resource utilisation levels, i.e.
+the worst case for a given scenario.
+%
+From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
+\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
+}
+%
+Clearly a deviation in results of such a large amount is unacceptable if simulation is to be used as a reliable verification tool.
+
+Resource utilisation level was found to have a significant impact on \textit{simulation variance}.
+%
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
+In this figure, anything above the $1$cm threshold (dashed line) is considered non-permissible according to our specific threshold level informed from our verification requirements.
+%
+\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
+%
+
+A general pattern in the results indicates that; some scenarios fail to produce deterministic results consistently against any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm threshold), some are consistently below the threshold (Test ID 5 \& 6 with pedestrians only), and that some cases only fail at higher utilisation levels, breaching the threshold above 75\% resource utilisation (Test ID 1 \& 3).
+
+\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
+
+However, the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
 
@@
 
-% Scenarios that include both collisions and vehicle actors were not deterministic.
+% Scenarios that include both collisions and vehicle actors were non-deterministic.
 However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.

### 2020-11-11 paper/cav_determinisim_main.tex
GC revisions KM paper complete

@@
 %
-\todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.
-}
+\todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
+%
 A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
+%
+\textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and advice for best working practice.}
 
@@
 
-%
-% Within this paper the term tolerance is a synonym for the permissible limit of variance.
+\subsection{Definition: Simulation Trace}
+
+A simulation trace is the output log from the simulator which is a time series of all actor positions ($x,y,z$) at regular intervals based on the time-step $dt$.
+This definition could be extended to include other variables, but position is sufficient for this analysis.
 
@@
 If the variance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
+Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulation traces.
 When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
@@
 The gamer neither wishes a faithful representation of reality, i.e.
-the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
+the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance, i.e.
 can be low \textit{precision}.
@@
 Where $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
-The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation, whereas \textit{restricted} restricts simulation time to pre-collision and limits system resource utilisation levels to 75\%.}
+The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation.
+To understand the impact of collisions and high resource utilisation the \textit{restricted} column shows a subset of the results where data post-collision has been removed and where data above a resource utilisation level of 75\% has also been removed.}
 \label{TableOfExperiments}
@@
 %
-This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment.
+This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small as to fail with minor computational perturbations.
 %
@@
 %
-For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish situations such as potentially dangerous close passing of a cyclist.
+For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
+%situations such as potentially dangerous close passing of a cyclist. 
 A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
 % due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study and more broadly our general verifcation requirements, a threshold of $1$cm has been selected. 
+Therefore, for this case study and more broadly our general verification requirements, a tolerance of $1$cm has been selected. 
 % Thus any variation of less than $1$cm is tolerable. 
@@
 
-The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
+The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation was reliable for verification purposes.
 % The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
@@
 % \textit{deterministic threshold}, 
-\textit{permissible tolerance}, 
+\textit{tolerance}, 
 settings external to the game engine were explored. 
@@
 
-\begin{figure*}[b]
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
+    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
+    \label{ExperimentsStressSummary}
+\end{figure}
+
+% ***************************************************
+%  System Configuration and Screening
+% ***************************************************
+\subsection{System Configuration and Screening}\label{s:screening}
+These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz.
+Operating systems was Linux Ubuntu 18.04.4 LTS.
+To ensure reliability of results, each test was repeated 1000 times.
+Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
+To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system.
+These were:
+
+\begin{itemize}[leftmargin=*]
+    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
+    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
+    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
+\end{itemize}
+
+% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
+These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
+Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
+We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
+
+% \subsection{Problem And Hypothesis}
+% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
+
+% Our hypothesis is that since UE4 is a game engine it is non-deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
+This behaviour is expected not to hold as the level of computing stress increases.
+
+ % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
+To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
+(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
+
+% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
+In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
+
+% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
+Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
+
+% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
+
+% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
+These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
+
+% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
+The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
+
+% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
+
+% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
+% Pedestrians have the functionality of dynamic path planning i.e.
+they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
+Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
+
+% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
+
+% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
+
+% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
+By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
+However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
+This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
+
+% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
+Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
+
+% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
+
+\begin{figure*}[t]
     \centering
@@
 % ***************************************************
-%  System Configuration and Screening
-% ***************************************************
-\subsection{System Configuration and Screening}\label{s:screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz.
-Operating systems was Linux Ubuntu 18.04.4 LTS.
-To ensure reliability of results, each test was repeated 1000 times.
-Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests were performed on our system.
-These were:
-
-\begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
-    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
-\end{itemize}
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
-    \label{ExperimentsStressSummary}
-\end{figure}
-
-% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
-These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
-
-% \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
-
-% Our hypothesis is that since UE4 is a game engine it is non-deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
-This behaviour is expected not to hold as the level of computing stress increases.
-
- % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
-To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
-(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
-
-% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
-
-% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
-
-% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
-
-% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
-
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
-
-% ***************************************************
 %  Results and Discussion
@@
 %
-From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
+From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the tolerance that has been set, with some deviation in actor path as large as $59$cm.
 \todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
@@
 Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, anything above the $1$cm threshold (dashed line) is considered non-permissible according to our specific threshold level informed from our verification requirements.
+In this figure, anything above the $1$cm level (dashed line) is considered non-permissible according to our specific tolerance level informed from our verification requirements.
 %
@@
 
-A general pattern in the results indicates that; some scenarios fail to produce deterministic results consistently against any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm threshold), some are consistently below the threshold (Test ID 5 \& 6 with pedestrians only), and that some cases only fail at higher utilisation levels, breaching the threshold above 75\% resource utilisation (Test ID 1 \& 3).
+A general pattern in the results indicates that; some scenarios consistently fail to produce results within permissible levels of variance at any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm level), some are consistently within tolerance (Test ID 5 \& 6 with pedestrians only), and that some cases only fail to meet the tolerance requirement when at higher utilisation levels, breaching the tolerance above 75\% resource utilisation (Test ID 1 \& 3).
 
@@
 %
-\begin{figure*}[b]
+\begin{figure*}[t]
     \centering
@@
 The main findings of this work suggest a working practice that would minimise the non-deterministic effects observed in this investigation.
-By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible tolerance threshold of $1$cm is achievable as shown in the \textit{restricted} column in Table.~\ref{TableOfExperiments}.
+By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table.~\ref{TableOfExperiments}.
 By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve.
@@
 However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly.
-Therefore we present a general methodology that practitioners can follow to find the
+Therefore we now present a general methodology that practitioners can follow to find the
 % \textit{deterministic domains of operation} 
-\textit{operational domains of permissible tolerance} 
+\textit{operational domains of permissible variance} 
 for any hardware configuration.
@@
 \section{Methodology} \label{s:methodology}
-In this section a method for determining the variance of simulated actor path trajectories, termed \textit{simulation variance}, and resolving the operational domains of permissible tolerance of the simulation are presented.
-%
+In this section a method for determining the variance of simulated actor path trajectories, termed \textit{simulation variance}, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.
+%
+In addition a number of recommendations and best practices are suggested that other practitioners can follow to minimise simulation variance. 
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
-This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}.
+% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}.
 %These need to be considered at different stages in the process.
@@
 \subsubsection{Trajectories and Collisions} 
-Actor paths should be hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
-
-\subsubsection{Setting a Permissible Tolerance} \label{s:threshold}
-The verification engineer should set the \textit{permissible tolerance} based on their specific requirements which refers to the acceptable level of actor path variance.
+% Actor paths should be hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
+Actor paths, or the sequence of actions required to generate paths, should be hard-coded to ensure repeatability.
+These paths should include collisions between actors and potentially collisions between actors and static scenery if this is likely to occur in the simulation or as part of the verification process.
+Actor paths without collision are also important to include as these will serve as a baseline to the other tests.
+
+\subsubsection{Setting a Tolerance} \label{s:threshold}
+The verification engineer should set the \textit{tolerance} based on their specific requirements which refers to the acceptable level of actor path variance.
 % ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
@@
 %
-The permissible tolerance that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this threshold must be sufficiently small to enable accurate assertion checking.
-In our empirical investigation we set a threshold of $1$cm which is sufficient for urban scenario assertion checking.
-In practice, it may be necessary to determine this threshold experimentally.
+The tolerance that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+For example, this tolerance must be sufficiently small to enable accurate assertion checking.
+In our empirical investigation we set a tolerance of $1$cm which is sufficient for urban scenario assertion checking.
+In practice, it may be necessary to determine this tolerance experimentally.
 
@@
 \subsubsection{Simulation Length} 
-The simulation time should be sufficient to record interactions between actors but not so long that 
-%100's or 1000's 
-a statistically significant number 
-of repeated tests take an inconvenient time to complete.
+The simulation time should be sufficient to record an interaction between actors but not so long that the testing take an inconvenient time to complete.
 In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
@@
 Initial screening tests done with NUMA control gave moderate improvements in simulation variance of the order of a few percent, see Section~\ref{s:screening}.
-This control may assist if simulation variance is borderline to the permissible tolerance but not seen as essential.
+This control may assist if simulation variance is borderline to the tolerance but not seen as essential.
 
@@
 While executing 100 repeats may seem sufficient, this sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
-It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the permissible tolerance.
+It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the tolerance.
 % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
@@
 The maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control for permissible operation.
-Finding these boundaries will identify the \textit{domains of permissible tolerance} for the user specific hardware.
+Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
+
+This method can be extended to a broader actor states and actions including actor orientation, speed, and any other status indicators that may be of interest and also including appropriate actions that may be useful for verification purposes.
+This broader As such, Equation~\ref{eq:max_sigma} would need to be adjusted to include these new variables.
 
@@
 % \section{Conclusions and Future Work}\label{s:conclusion}
-\section{Conclusions}\label{s:conclusion}
+\section{Conclusions \& Future Work}\label{s:conclusion}
 
@@
 %
-The results of the investigation are hardware and software version specific, so we present a universal methodology that can be used to find the \textit{domains of permissible tolerance} for any given system configuration.
+The results of the investigation are hardware and software version specific, so we proposed a general methodology that can be used to find the \textit{domains of permissible variance} for any given system configuration.
 %
 This methodology will allow the AV verification community using simulation tools to ensure their results are reliable and trustworthy.
+
+\subsection{Future Work}
+A potential avenue of exploration is developing a version of the simulator which has been designed specifically for verification.
+In this version the simulator would be deterministic because the execution is suitably controlled along with sources of randomness and scheduling.
+A similar task was achieved with the \texttt{RR} program.\footnote{\url{https://github.com/mozilla/rr}}
+
+With the advent of more AI systems becoming adaptive, meaning that the system may change it's response to the same stimulus over time, the notion of repeatability will need reassessing in the context of verification.
+The main research question here is what will constitute `the same' which will be important for passing or failing verificatin tests but also for determining coverage.
 

### 2020-11-11 paper/cav_determinisim_main.tex
Added diagrams for variance description and plot for nice analysis

@@
 
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.48\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
+        \caption{}
+        \label{variance_description_a}
+    \end{subfigure}
+
+    \begin{subfigure}{.48\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
+        \caption{}
+        \label{variance_description_b}
+    \end{subfigure}
+    \caption{Demonstration of variance, precision, tolderance and determinism}
+    \label{variance_description}
+\end{figure}
+
 \subsection{Definitions: Precision, Variance \& Tolerance}
@@
 
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary.pdf}
+    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% CPU \& GPU Resource Utilisation.}
+    \label{NICEExperimentStressSummary}
+\end{figure}
+
 The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
@@
 With the advent of more AI systems becoming adaptive, meaning that the system may change it's response to the same stimulus over time, the notion of repeatability will need reassessing in the context of verification.
-The main research question here is what will constitute `the same' which will be important for passing or failing verificatin tests but also for determining coverage.
+The main research question here is what will constitute `the same' which will be important for passing or failing verification tests but also for determining coverage.
 

### 2020-11-13 paper/cav_determinisim_main.tex
Added diagrams for variance description and plot for nice analysis
Added diagrams for variance description, updated clearer plots for pre post and added  plot for nice analysis
Corrected some spelling mistakes
GC added NICE

@@
 \author[1,3]{Greg Chance}
-\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
+\author[2,3]{S\'everin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
 \author[2,3]{Tony Pipe}
@@
         \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
-        \caption{}
+        \caption{Non-deterministic}
         \label{variance_description_a}
@@
         \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
-        \caption{}
+        \caption{Deterministic}
         \label{variance_description_b}
     \end{subfigure}
-    \caption{Demonstration of variance, precision, tolderance and determinism}
+    \caption{Demonstration of variance, precision, tolerance and determinism}
     \label{variance_description}
@@
 % ======= Sources of non-determinism
-\subsection{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}
+\section{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}
 
@@
 % ======= Floating-Point Numbers
-\noindent\underline{\textit{Floating-Point Arithmetic:}}
+\subsection{Floating-Point Arithmetic}
 
@@
 % Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, such rounding errors could result in precision issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
+In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
 % \todo[inline]{state what assertions are (used for) GC-see below} 
@@
 
-\noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
-ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}.
-Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
-If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
-
-Any simulation hardware not using ECC memory that runs for 1000's of hours, typical in AV verification, is likely to incur significant CPU hours and is therefore subject to increased exposure to these errors.
-To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.\todo{Check this, please.}
-
-%\todo[inline]{That said, HW design simulation is deterministic, uses long simulation runs and thousands of CPU hours, but I've not heard that this is an issue.
-I can ask, though.
-;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.}
-
-\noindent\underline{\textit{Actor Navigation:}}
-There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-% 
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
-%
-The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
-
-A navigation mesh is a fixed area of the environment where actors are free to navigate.
-%
-Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
-%.
-Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
-%
-However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment and cannot be unknowingly modified and therefore can be considered fixed.
-
-Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
-%
-% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
-%
-% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
-% \todo[inline]{unclear :( GC- re-written}
-\\\\
-\noindent\underline{\textit{Game Engine Setup:}}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
-%
-% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
-% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{Frame} which reports
-% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
-the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
-%
-Wrt.\ perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water.
-\\\\
-\noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
-For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
-Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
-%
-Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
-%
-By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
-\\\\
-\noindent\underline{\textit{Scheduling, Concurrency and Parallelisation:}}
+\subsection{Scheduling, Concurrency and Parallelisation}
 % \todo[inline]{is this at runtime?} %
@@
 
-It is, however, unlikely that thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
+It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
 Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
@@
 Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}. 
-\\\\
 %\noindent\underline{\textit{Concurrency:}}
@@
 %\\\\
-\noindent\underline{\textit{Summary:}}
+
+\subsection{NUMA Non-Uniform Memory Access}
+For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
+Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
+%
+Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower.
+%
+By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
+
+\subsection{Error Correcting Code (ECC) Memory}
+ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}.
+Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
+If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
+Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
+
+Any simulation hardware not using ECC memory that runs for 1000's of hours, typical in AV verification, is likely to incur significant CPU hours and is therefore subject to increased exposure to these errors.
+To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.\todo{Check this, please.}
+
+%\todo[inline]{That said, HW design simulation is deterministic, uses long simulation runs and thousands of CPU hours, but I've not heard that this is an issue.
+I can ask, though.
+;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.}
+
+\subsection{Game Engine Setup}
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
+%
+%
+% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
+% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
+% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
+For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{Frame} which reports
+% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
+the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
+%
+With respect to perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water, ensuring textures are not randomly generated.
+
+\subsection{Actor Navigation}
+There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
+% 
+Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
+%
+The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
+
+A navigation mesh is a fixed area of the environment where actors are free to navigate.
+%
+Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
+%.
+Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
+%
+However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment and cannot be unknowingly modified and therefore can be considered fixed.
+
+Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
+%
+% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
+%
+% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
+% \todo[inline]{unclear :( GC- re-written}
+
+\subsection{Summary}
 We have investigated the potential sources of non-determinism affecting game engines 
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV3.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV4.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV3.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV4.pdf}
         \caption{}
@@
 }
+
+\subsection{Process Scheduling Priority}
+An investigation into the response of process scheduling on the simulation variance was undertaken.
+The experiment was repeated using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}.
+Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under stress.
+%
+To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.
+
+The results presented in a box and whisker format, Fig.~\ref{NICEExperimentStressSummary}, show that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a 0 NICE value (central plot bars).
+Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation  but this did not account for all the difference in the observed results.
+%
+This unaccounted difference in the variance may be due to the lack of  control that \texttt{NICE} has over the processes~\cite{Abanoub can you cite that quote you found here please}.
 

### 2020-11-13 paper/cav_determinisim_main.tex
Did some updates to plots

@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV4.pdf}
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV5.pdf}
     \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV4.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV4.pdf}
+        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
         \caption{}
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary.pdf}
-    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% CPU \& GPU Resource Utilisation.}
+    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V3.pdf}
+    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% additonal CPU \& GPU resource utilisation.}
     \label{NICEExperimentStressSummary}

### 2020-11-13 paper/cav_determinisim_main.tex
Modified methodology diagram
Reorganised some bits in the intro

@@
 
-\subsection{Definition: Determinism}
+\subsection{Definitions}
+\subsubsection{Determinism}
 % What is a deterministic system
@@
 %
-However, if a simulation is non-deterministic then then there will be a variation in the generated output data.
+However, if a simulation is non-deterministic then then there will be a variation in the generated output data. \\
 
@@
 
-\subsection{Definitions: Precision, Variance \& Tolerance}
+% \subsection{Definitions: Precision, Variance \& Tolerance}
+\subsubsection{Precision, Variance \& Tolerance}
 
@@
 If a simulator is deterministic then it will produce results with absolute precision or zero variance and hence will be within acceptable tolerance.
-If the simulator is non-deterministic then there will be a variance in the output which can be measured.
-%
-
-\subsection{Definition: Accuracy}
-
+If the simulator is non-deterministic then there will be a variance in the output which can be measured.\\
+%
+
+% \subsection{Definition: Accuracy}
+\subsubsection{Accuracy}
 Precision and tolerance are not to be confused with \textit{accuracy} which describes how closely the generated output of a process aligns to a known standard.
@@
 In most cases such accuracy will not be possible due to computational demands of such an exact replica and in most cases is unnecessary, where some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
-
-\subsection{Definition: Variance Measurement}
-
+\\
+
+% \subsection{Definition: Variance Measurement}
+\subsubsection{Variance Measurement}
 If the simulator is non-deterministic then how can this variance be measured?
@@
 vehicle close-passing.
-Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.
-
-\subsection{Definition: Simulation Trace}
-
+Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
+
+% \subsection{Definition: Simulation Trace}
+\subsubsection{Simulation Trace}
 A simulation trace is the output log from the simulator which is a time series of all actor positions ($x,y,z$) at regular intervals based on the time-step $dt$.
-This definition could be extended to include other variables, but position is sufficient for this analysis.
-
-\subsection{Definition: Scene, Scenario \& Situation}
-
+This definition could be extended to include other variables, but position is sufficient for this analysis.\\
+
+% \subsection{Definition: Scene, Scenario \& Situation}
+\subsubsection{Scene, Scenario \& Situation}
 We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v4.pdf}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v5.pdf}
     \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}
@@
 
-\subsection{Future Work}
+% \subsection{Future Work}
 A potential avenue of exploration is developing a version of the simulator which has been designed specifically for verification.

### 2020-11-13 paper/cav_determinisim_main.tex
GC NICE ref

@@
 % If the policy or operating system cannot be changed then s
-Setting the simulator program to a higher priority may help to alleviate conflicting tasks that cannot be shutdown, such as \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}} or \texttt{nice} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.
+Setting the simulator process to a higher priority may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}} or \texttt{NICE} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.
 

### 2020-11-13 paper/cav_determinisim_main.tex
GC abstract

@@
 %
-A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-%
-\textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and advice for best working practice.}
+A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification.
+The proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
+%
+% \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}
 

### 2020-11-13 paper/cav_determinisim_main.tex
Added the setup guide. Done various little touches and corrections.

@@
 %
-CARLA\footnote{\url{http://carla.org/}}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
+CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website} a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
 
@@
 
-\subsection{Definitions}
+\subsection{Definitions (Refer to Fig.~\ref{variance_description} throughout this section)}
 \subsubsection{Determinism}
@@
 There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g.
-Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, Unity\footnote{\url{https://unity.com/}}, CryEngine\footnote{\url{https://www.cryengine.com/}}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA\footnote{\url{http://carla.org/}}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
+Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
 % \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
@@
 % ======= Overview of a Game Loop
-\subsection{Game Engine Overview} \label{GameLoopSection}
+% \subsection{Game Engine Overview} \label{GameLoopSection}
 
@@
 Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware.
-Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}.
+Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}.
 %
@@
 
-However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
+However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
 This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
@@
 
-Indeed, the developers of the de-bugging program \texttt{RR}\footnote{\url{https://github.com/mozilla/rr}} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+Indeed, the developers of the de-bugging program \texttt{RR}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
@@
 % \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{Frame} which reports
+For example, in Unreal Editor the \textit{unit}\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
 % to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
@@
 % 
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
+Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
 %
@@
 \subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Collision/Overview/index.html}}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}.
+Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}.
 %
@@
 \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking.\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}}
+Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
 % \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
@@
 For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
-In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision.\footnote{\url{https://uk.mathworks.com/matlabcentral/answers/317128-does-matlab-use-all-cores-by-default-when-running-a-program}}
+In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision~\cite{matlab_parallel_computing}.
 
@@
 These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
-Operating systems was Linux Ubuntu 18.04.4 LTS.
-To ensure reliability of results, each test was repeated 1000 times.
+%
+Operating systems was Linux Ubuntu 18.04.4 LTS. To ensure reliability of results, each test was repeated 1000 times. 
+%
 Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s. 
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system.
-These were:
+%
+A detailed guide for reproducing these experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
+%
+To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:
 
 \begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
-    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
+    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
+    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}\cite{shi2009testing}.
+    \item Non-uniform memory access: \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
 \end{itemize}
@@
     \end{subfigure}
-    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation.
+    \caption{Vehicle to vehicle collision (test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
 The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
@@
     \end{subfigure}
-    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% stress and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
+    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
 \end{figure*}
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V3.pdf}
+    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
     \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% additonal CPU \& GPU resource utilisation.}
@@
 The experiment was repeated using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}.
-Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under stress.
+Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
 %
@@
 %
-This unaccounted difference in the variance may be due to the lack of  control that \texttt{NICE} has over the processes~\cite{Abanoub can you cite that quote you found here please}.
+This unaccounted difference in the variance may be due to the lack of  control that \texttt{NICE} has over the processes\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}.
+%~\cite{Abanoub can you cite that quote you found here please}.
 
@@
 The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
-If using CARLA a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}}.
+If using CARLA a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$~\cite{MonoBehaviour_unity}.
 In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
@@
 % If the policy or operating system cannot be changed then s
-Setting the simulator process to a higher priority may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}} or \texttt{NICE} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.
+Setting the simulator process to a higher priority may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
 
@@
 In this version the simulator would be deterministic because the execution is suitably controlled along with sources of randomness and scheduling.
-A similar task was achieved with the \texttt{RR} program.\footnote{\url{https://github.com/mozilla/rr}}
+A similar task was achieved with the \texttt{RR} program~\cite{RR_link}.
 
@@
 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-\balance
 \printbibliography

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerCarlaIntegration/Intermediate/Build/Linux/B4D820EA/UE4Editor/Development/RoadRunnerCarlaIntegration/Definitions.RoadRunnerCarlaIntegration.h

@@ -1,0 +1,219 @@
+#undef ROADRUNNERCARLAINTEGRATION_API
+#undef UE_IS_ENGINE_MODULE
+#undef DEPRECATED_FORGAME
+#define DEPRECATED_FORGAME DEPRECATED
+#undef UE_DEPRECATED_FORGAME
+#define UE_DEPRECATED_FORGAME UE_DEPRECATED
+#define IS_PROGRAM 0
+#define UE_EDITOR 1
+#define ENABLE_PGO_PROFILE 0
+#define USE_VORBIS_FOR_STREAMING 1
+#define WITH_DEV_AUTOMATION_TESTS 1
+#define WITH_PERF_AUTOMATION_TESTS 1
+#define UNICODE 1
+#define _UNICODE 1
+#define __UNREAL__ 1
+#define IS_MONOLITHIC 0
+#define WITH_ENGINE 1
+#define WITH_UNREAL_DEVELOPER_TOOLS 1
+#define WITH_APPLICATION_CORE 1
+#define WITH_COREUOBJECT 1
+#define USE_STATS_WITHOUT_ENGINE 0
+#define WITH_PLUGIN_SUPPORT 0
+#define WITH_PERFCOUNTERS 1
+#define USE_LOGGING_IN_SHIPPING 0
+#define WITH_LOGGING_TO_MEMORY 0
+#define USE_CACHE_FREED_OS_ALLOCS 1
+#define USE_CHECKS_IN_SHIPPING 0
+#define WITH_EDITOR 1
+#define WITH_SERVER_CODE 1
+#define WITH_CEF3 1
+#define WITH_LIVE_CODING 0
+#define WITH_XGE_CONTROLLER 0
+#define UBT_MODULE_MANIFEST "UE4Editor.modules"
+#define UBT_MODULE_MANIFEST_DEBUGGAME "UE4Editor-Linux-DebugGame.modules"
+#define UBT_COMPILED_PLATFORM Linux
+#define UBT_COMPILED_TARGET Editor
+#define UE_APP_NAME "UE4Editor"
+#define PLATFORM_LINUX 1
+#define PLATFORM_UNIX 1
+#define LINUX 1
+#define PLATFORM_SUPPORTS_JEMALLOC 1
+#define NDEBUG 1
+#define UE_BUILD_DEVELOPMENT 1
+#define UE_IS_ENGINE_MODULE 0
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerCarlaIntegration/Intermediate/Build/Linux/B4D820EA/UE4Editor/Development/RoadRunnerCarlaIntegration/Module.RoadRunnerCarlaIntegration.cpp

@@ -1,0 +1,3 @@
+// This file is automatically generated at compile-time to include some subset of the user-created cpp files.
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerCarlaIntegration/Source/RoadRunnerCarlaIntegration/Private/RoadRunnerCarlaIntegration.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerCarlaIntegration/Source/RoadRunnerCarlaIntegration/Private/RoadRunnerCarlaIntegrationLog.cpp"

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerCarlaIntegration/Source/RoadRunnerCarlaIntegration/Private/RoadRunnerCarlaIntegration.cpp

@@ -1,0 +1,550 @@
+// This is free and unencumbered software released into the public domain.
+//
+// Anyone is free to copy, modify, publish, use, compile, sell, or
+// distribute this software, either in source code form or as a compiled
+// binary, for any purpose, commercial or non - commercial, and by any
+// means.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+// EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+// MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+// IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+// OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+// ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+// OTHER DEALINGS IN THE SOFTWARE.
+
+#include "RoadRunnerCarlaIntegration.h"
+#include "RoadRunnerCarlaIntegrationLog.h"
+#include "RoadRunnerFbxSceneImportFactory.h"
+#include "RoadRunnerImporterOptions.h"
+#include "RoadRunnerImportUtil.h"
+#include "RoadRunnerImporter.h"
+
+#include <AssetRegistryModule.h>
+#include <Developer/AssetTools/Public/AssetToolsModule.h>
+#include <Developer/AssetTools/Public/IAssetTools.h>
+#include <Editor/UnrealEd/Public/EditorBuildUtils.h>
+#include <Runtime/XmlParser/Public/XmlFile.h>
+#include <ObjectTools.h>
+#include <PackageTools.h>
+#include <PlatformFilemanager.h>
+
+#include <OpenDrive/OpenDriveActor.h>
+
+// Fix for CARLA 0.9.4
+#undef CopyFile
+
+#define LOCTEXT_NAMESPACE "FRoadRunnerCarlaIntegrationModule"
+
+const FString FRoadRunnerCarlaIntegrationModule::BaseMapPath = "/Game/Carla/Maps/BaseMap/BaseMap.BaseMap";
+const FString FRoadRunnerCarlaIntegrationModule::MapDestRelPath = "Carla/ExportedMaps/";
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerCarlaIntegration/Source/RoadRunnerCarlaIntegration/Public/RoadRunnerCarlaIntegration.h

@@ -1,0 +1,61 @@
+// This is free and unencumbered software released into the public domain.
+//
+// Anyone is free to copy, modify, publish, use, compile, sell, or
+// distribute this software, either in source code form or as a compiled
+// binary, for any purpose, commercial or non - commercial, and by any
+// means.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+// EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+// MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+// IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+// OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+// ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+// OTHER DEALINGS IN THE SOFTWARE.
+
+#pragma once
+
+#include "CoreMinimal.h"
+#include "Modules/ModuleManager.h"
+#include <UnrealEd.h>
+
+////////////////////////////////////////////////////////////////////////////////
+// Plugin to streamline the RoadRunner to CARLA process
+//	- Creates new map from BaseMap
+//	- Moves static meshes for segmentation
+//	- Creates materials instanced from CARLA's to work with weather
+//	- Generates routes from OpenDRIVE
+class FRoadRunnerCarlaIntegrationModule : public IModuleInterface
+{
+public:
+	// Constants
+	static const FString BaseMapPath;
+	static const FString MapDestRelPath;
+	static const FString OpenDriveDestRelPath;
+
+	// CARLA segmentation path list
+	static TArray<FString> PathList;
+	// Segmentation enum
+	enum SegmentationType
+	{
+		eRoad,
+		eMarking,
+		eTerrain,
+		eProp,
+		eSidewalk,
+		eSign,
+		eFoliage,
+		eFence,
+		ePole,
+		eVehicle,
+		eWall
+	};
+
+	// Import delegates
+	static void RRCarlaPreImport(UFactory* inFactory, UClass* inClass, UObject* inParent, const FName& name, const TCHAR* type);
+	static void RRCarlaPostImport(UFactory* inFactory, UObject* inCreateObject);
+
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Development/RoadRunnerRuntime/Definitions.RoadRunnerRuntime.h

@@ -1,0 +1,104 @@
+#undef ROADRUNNERRUNTIME_API
+#undef UE_IS_ENGINE_MODULE
+#undef DEPRECATED_FORGAME
+#define DEPRECATED_FORGAME DEPRECATED
+#undef UE_DEPRECATED_FORGAME
+#define UE_DEPRECATED_FORGAME UE_DEPRECATED
+#define IS_PROGRAM 0
+#define UE_GAME 1
+#define ENABLE_PGO_PROFILE 0
+#define USE_VORBIS_FOR_STREAMING 1
+#define WITH_DEV_AUTOMATION_TESTS 1
+#define WITH_PERF_AUTOMATION_TESTS 1
+#define UNICODE 1
+#define _UNICODE 1
+#define __UNREAL__ 1
+#define IS_MONOLITHIC 1
+#define WITH_ENGINE 1
+#define WITH_UNREAL_DEVELOPER_TOOLS 0
+#define WITH_APPLICATION_CORE 1
+#define WITH_COREUOBJECT 1
+#define USE_STATS_WITHOUT_ENGINE 0
+#define WITH_PLUGIN_SUPPORT 0
+#define WITH_PERFCOUNTERS 0
+#define USE_LOGGING_IN_SHIPPING 0
+#define WITH_LOGGING_TO_MEMORY 0
+#define USE_CACHE_FREED_OS_ALLOCS 1
+#define USE_CHECKS_IN_SHIPPING 0
+#define WITH_EDITOR 0
+#define WITH_EDITORONLY_DATA 0
+#define WITH_SERVER_CODE 1
+#define WITH_CEF3 1
+#define WITH_LIVE_CODING 0
+#define WITH_XGE_CONTROLLER 0
+#define UBT_MODULE_MANIFEST "CarlaUE4.modules"
+#define UBT_MODULE_MANIFEST_DEBUGGAME "CarlaUE4-Linux-DebugGame.modules"
+#define UBT_COMPILED_PLATFORM Linux
+#define UBT_COMPILED_TARGET Game
+#define UE_APP_NAME "CarlaUE4"
+#define PLATFORM_LINUX 1
+#define PLATFORM_UNIX 1
+#define LINUX 1
+#define PLATFORM_SUPPORTS_JEMALLOC 1
+#define NDEBUG 1
+#define UE_BUILD_DEVELOPMENT 1
+#define UE_ENGINE_DIRECTORY "../../../../../UnrealEngine_4.22/Engine/"
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Development/RoadRunnerRuntime/Module.RoadRunnerRuntime.cpp

@@ -1,0 +1,3 @@
+// This file is automatically generated at compile-time to include some subset of the user-created cpp files.
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerRuntime/Private/RoadRunnerRuntime.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerRuntime/Private/RoadRunnerTrafficJunction.cpp"

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Development/RoadRunnerRuntime/Module.RoadRunnerRuntime.gen.cpp

@@ -1,0 +1,3 @@
+// This file is automatically generated at compile-time to include some subset of the user-created cpp files.
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Inc/RoadRunnerRuntime/RoadRunnerRuntime.init.gen.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Inc/RoadRunnerRuntime/RoadRunnerTrafficJunction.gen.cpp"

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Inc/RoadRunnerRuntime/RoadRunnerRuntime.init.gen.cpp

@@ -1,0 +1,35 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/GeneratedCppIncludes.h"
+#ifdef _MSC_VER
+#pragma warning (push)
+#pragma warning (disable : 4883)
+#endif
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+void EmptyLinkFunctionForGeneratedCodeRoadRunnerRuntime_init() {}
+	UPackage* Z_Construct_UPackage__Script_RoadRunnerRuntime()
+	{
+		static UPackage* ReturnPackage = nullptr;
+		if (!ReturnPackage)
+		{
+			static const UE4CodeGen_Private::FPackageParams PackageParams = {
+				"/Script/RoadRunnerRuntime",
+				nullptr,
+				0,
+				PKG_CompiledIn | 0x00000000,
+				0x38F58527,
+				0xF91EBF9D,
+				METADATA_PARAMS(nullptr, 0)
+			};
+			UE4CodeGen_Private::ConstructUPackage(ReturnPackage, PackageParams);
+		}
+		return ReturnPackage;
+	}
+PRAGMA_ENABLE_DEPRECATION_WARNINGS
+#ifdef _MSC_VER
+#pragma warning (pop)
+#endif

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Inc/RoadRunnerRuntime/RoadRunnerTrafficJunction.gen.cpp

@@ -1,0 +1,992 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/GeneratedCppIncludes.h"
+#include "RoadRunnerRuntime/Public/RoadRunnerTrafficJunction.h"
+#ifdef _MSC_VER
+#pragma warning (push)
+#pragma warning (disable : 4883)
+#endif
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+void EmptyLinkFunctionForGeneratedCodeRoadRunnerTrafficJunction() {}
+// Cross Module References
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FJunction();
+	UPackage* Z_Construct_UPackage__Script_RoadRunnerRuntime();
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FSignalPhase();
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FLightInterval();
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FSignalState();
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FLightInstanceState();
+	ENGINE_API UClass* Z_Construct_UClass_UMeshComponent_NoRegister();
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FSignalAsset();
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FSignalConfiguration();
+	ROADRUNNERRUNTIME_API UScriptStruct* Z_Construct_UScriptStruct_FLightBulbState();
+	ROADRUNNERRUNTIME_API UClass* Z_Construct_UClass_URoadRunnerTrafficJunction_NoRegister();
+	ROADRUNNERRUNTIME_API UClass* Z_Construct_UClass_URoadRunnerTrafficJunction();
+	ENGINE_API UClass* Z_Construct_UClass_UActorComponent();
+// End Cross Module References
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/CarlaUE4/Inc/RoadRunnerRuntime/RoadRunnerTrafficJunction.generated.h

@@ -1,0 +1,146 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/ObjectMacros.h"
+#include "UObject/ScriptMacros.h"
+
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+#ifdef ROADRUNNERRUNTIME_RoadRunnerTrafficJunction_generated_h
+#error "RoadRunnerTrafficJunction.generated.h already included, missing '#pragma once' in RoadRunnerTrafficJunction.h"
+#endif
+#define ROADRUNNERRUNTIME_RoadRunnerTrafficJunction_generated_h
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerRuntime_Public_RoadRunnerTrafficJunction_h_152_GENERATED_BODY \
+	friend struct Z_Construct_UScriptStruct_FJunction_Statics; \
+	ROADRUNNERRUNTIME_API static class UScriptStruct* StaticStruct();
+
+
+template<> ROADRUNNERRUNTIME_API UScriptStruct* StaticStruct<struct FJunction>();
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerRuntime_Public_RoadRunnerTrafficJunction_h_140_GENERATED_BODY \
+	friend struct Z_Construct_UScriptStruct_FSignalPhase_Statics; \
+	ROADRUNNERRUNTIME_API static class UScriptStruct* StaticStruct();
+
+
+template<> ROADRUNNERRUNTIME_API UScriptStruct* StaticStruct<struct FSignalPhase>();
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerRuntime_Public_RoadRunnerTrafficJunction_h_124_GENERATED_BODY \
+	friend struct Z_Construct_UScriptStruct_FLightInterval_Statics; \
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Development/RoadRunnerImporter/Definitions.RoadRunnerImporter.h

@@ -1,0 +1,206 @@
+#undef ROADRUNNERIMPORTER_API
+#undef UE_IS_ENGINE_MODULE
+#undef DEPRECATED_FORGAME
+#define DEPRECATED_FORGAME DEPRECATED
+#undef UE_DEPRECATED_FORGAME
+#define UE_DEPRECATED_FORGAME UE_DEPRECATED
+#define IS_PROGRAM 0
+#define UE_EDITOR 1
+#define ENABLE_PGO_PROFILE 0
+#define USE_VORBIS_FOR_STREAMING 1
+#define WITH_DEV_AUTOMATION_TESTS 1
+#define WITH_PERF_AUTOMATION_TESTS 1
+#define UNICODE 1
+#define _UNICODE 1
+#define __UNREAL__ 1
+#define IS_MONOLITHIC 0
+#define WITH_ENGINE 1
+#define WITH_UNREAL_DEVELOPER_TOOLS 1
+#define WITH_APPLICATION_CORE 1
+#define WITH_COREUOBJECT 1
+#define USE_STATS_WITHOUT_ENGINE 0
+#define WITH_PLUGIN_SUPPORT 0
+#define WITH_PERFCOUNTERS 1
+#define USE_LOGGING_IN_SHIPPING 0
+#define WITH_LOGGING_TO_MEMORY 0
+#define USE_CACHE_FREED_OS_ALLOCS 1
+#define USE_CHECKS_IN_SHIPPING 0
+#define WITH_EDITOR 1
+#define WITH_SERVER_CODE 1
+#define WITH_CEF3 1
+#define WITH_LIVE_CODING 0
+#define WITH_XGE_CONTROLLER 0
+#define UBT_MODULE_MANIFEST "UE4Editor.modules"
+#define UBT_MODULE_MANIFEST_DEBUGGAME "UE4Editor-Linux-DebugGame.modules"
+#define UBT_COMPILED_PLATFORM Linux
+#define UBT_COMPILED_TARGET Editor
+#define UE_APP_NAME "UE4Editor"
+#define PLATFORM_LINUX 1
+#define PLATFORM_UNIX 1
+#define LINUX 1
+#define PLATFORM_SUPPORTS_JEMALLOC 1
+#define NDEBUG 1
+#define UE_BUILD_DEVELOPMENT 1
+#define UE_IS_ENGINE_MODULE 0
+#define UE_PROJECT_NAME CarlaUE4
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Development/RoadRunnerImporter/Module.RoadRunnerImporter.cpp

@@ -1,0 +1,10 @@
+// This file is automatically generated at compile-time to include some subset of the user-created cpp files.
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerFbxFactory.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerFbxSceneImportFactory.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerImporter.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerImporterLog.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerImporterOptions.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerImporterOptionWindow.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerImportUtil.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerReimportFactory.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerStyle.cpp"

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Development/RoadRunnerImporter/Module.RoadRunnerImporter.gen.cpp

@@ -1,0 +1,6 @@
+// This file is automatically generated at compile-time to include some subset of the user-created cpp files.
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerFbxFactory.gen.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerFbxSceneImportFactory.gen.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerImporter.init.gen.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerImporterOptions.gen.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerReimportFactory.gen.cpp"

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Development/RoadRunnerRuntime/Definitions.RoadRunnerRuntime.h

@@ -1,0 +1,203 @@
+#undef ROADRUNNERRUNTIME_API
+#undef UE_IS_ENGINE_MODULE
+#undef DEPRECATED_FORGAME
+#define DEPRECATED_FORGAME DEPRECATED
+#undef UE_DEPRECATED_FORGAME
+#define UE_DEPRECATED_FORGAME UE_DEPRECATED
+#define IS_PROGRAM 0
+#define UE_EDITOR 1
+#define ENABLE_PGO_PROFILE 0
+#define USE_VORBIS_FOR_STREAMING 1
+#define WITH_DEV_AUTOMATION_TESTS 1
+#define WITH_PERF_AUTOMATION_TESTS 1
+#define UNICODE 1
+#define _UNICODE 1
+#define __UNREAL__ 1
+#define IS_MONOLITHIC 0
+#define WITH_ENGINE 1
+#define WITH_UNREAL_DEVELOPER_TOOLS 1
+#define WITH_APPLICATION_CORE 1
+#define WITH_COREUOBJECT 1
+#define USE_STATS_WITHOUT_ENGINE 0
+#define WITH_PLUGIN_SUPPORT 0
+#define WITH_PERFCOUNTERS 1
+#define USE_LOGGING_IN_SHIPPING 0
+#define WITH_LOGGING_TO_MEMORY 0
+#define USE_CACHE_FREED_OS_ALLOCS 1
+#define USE_CHECKS_IN_SHIPPING 0
+#define WITH_EDITOR 1
+#define WITH_SERVER_CODE 1
+#define WITH_CEF3 1
+#define WITH_LIVE_CODING 0
+#define WITH_XGE_CONTROLLER 0
+#define UBT_MODULE_MANIFEST "UE4Editor.modules"
+#define UBT_MODULE_MANIFEST_DEBUGGAME "UE4Editor-Linux-DebugGame.modules"
+#define UBT_COMPILED_PLATFORM Linux
+#define UBT_COMPILED_TARGET Editor
+#define UE_APP_NAME "UE4Editor"
+#define PLATFORM_LINUX 1
+#define PLATFORM_UNIX 1
+#define LINUX 1
+#define PLATFORM_SUPPORTS_JEMALLOC 1
+#define NDEBUG 1
+#define UE_BUILD_DEVELOPMENT 1
+#define UE_IS_ENGINE_MODULE 0
+#define UE_PROJECT_NAME CarlaUE4
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Development/RoadRunnerRuntime/Module.RoadRunnerRuntime.gen.cpp

@@ -1,0 +1,3 @@
+// This file is automatically generated at compile-time to include some subset of the user-created cpp files.
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerRuntime/RoadRunnerRuntime.init.gen.cpp"
+#include "/home/abanoub_g/carla/Unreal/CarlaUE4/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerRuntime/RoadRunnerTrafficJunction.gen.cpp"

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerFbxFactory.gen.cpp

@@ -1,0 +1,86 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/GeneratedCppIncludes.h"
+#include "RoadRunnerImporter/Public/RoadRunnerFbxFactory.h"
+#ifdef _MSC_VER
+#pragma warning (push)
+#pragma warning (disable : 4883)
+#endif
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+void EmptyLinkFunctionForGeneratedCodeRoadRunnerFbxFactory() {}
+// Cross Module References
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerFbxFactory_NoRegister();
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerFbxFactory();
+	UNREALED_API UClass* Z_Construct_UClass_UFbxFactory();
+	UPackage* Z_Construct_UPackage__Script_RoadRunnerImporter();
+// End Cross Module References
+	void URoadRunnerFbxFactory::StaticRegisterNativesURoadRunnerFbxFactory()
+	{
+	}
+	UClass* Z_Construct_UClass_URoadRunnerFbxFactory_NoRegister()
+	{
+		return URoadRunnerFbxFactory::StaticClass();
+	}
+	struct Z_Construct_UClass_URoadRunnerFbxFactory_Statics
+	{
+		static UObject* (*const DependentSingletons[])();
+#if WITH_METADATA
+		static const UE4CodeGen_Private::FMetaDataPairParam Class_MetaDataParams[];
+#endif
+		static const FCppClassTypeInfoStatic StaticCppClassTypeInfo;
+		static const UE4CodeGen_Private::FClassParams ClassParams;
+	};
+	UObject* (*const Z_Construct_UClass_URoadRunnerFbxFactory_Statics::DependentSingletons[])() = {
+		(UObject* (*)())Z_Construct_UClass_UFbxFactory,
+		(UObject* (*)())Z_Construct_UPackage__Script_RoadRunnerImporter,
+	};
+#if WITH_METADATA
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerFbxFactory.generated.h

@@ -1,0 +1,92 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/ObjectMacros.h"
+#include "UObject/ScriptMacros.h"
+
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+#ifdef ROADRUNNERIMPORTER_RoadRunnerFbxFactory_generated_h
+#error "RoadRunnerFbxFactory.generated.h already included, missing '#pragma once' in RoadRunnerFbxFactory.h"
+#endif
+#define ROADRUNNERIMPORTER_RoadRunnerFbxFactory_generated_h
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerFbxFactory_h_34_RPC_WRAPPERS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerFbxFactory_h_34_RPC_WRAPPERS_NO_PURE_DECLS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerFbxFactory_h_34_INCLASS_NO_PURE_DECLS \
+private: \
+	static void StaticRegisterNativesURoadRunnerFbxFactory(); \
+	friend struct Z_Construct_UClass_URoadRunnerFbxFactory_Statics; \
+public: \
+	DECLARE_CLASS(URoadRunnerFbxFactory, UFbxFactory, COMPILED_IN_FLAGS(0), CASTCLASS_None, TEXT("/Script/RoadRunnerImporter"), NO_API) \
+	DECLARE_SERIALIZER(URoadRunnerFbxFactory)
+
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerFbxFactory_h_34_INCLASS \
+private: \
+	static void StaticRegisterNativesURoadRunnerFbxFactory(); \
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerFbxSceneImportFactory.gen.cpp

@@ -1,0 +1,101 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/GeneratedCppIncludes.h"
+#include "RoadRunnerImporter/Public/RoadRunnerFbxSceneImportFactory.h"
+#ifdef _MSC_VER
+#pragma warning (push)
+#pragma warning (disable : 4883)
+#endif
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+void EmptyLinkFunctionForGeneratedCodeRoadRunnerFbxSceneImportFactory() {}
+// Cross Module References
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerFbxSceneImportFactory_NoRegister();
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerFbxSceneImportFactory();
+	UNREALED_API UClass* Z_Construct_UClass_UFbxSceneImportFactory();
+	UPackage* Z_Construct_UPackage__Script_RoadRunnerImporter();
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerImporterOptions_NoRegister();
+// End Cross Module References
+	void URoadRunnerFbxSceneImportFactory::StaticRegisterNativesURoadRunnerFbxSceneImportFactory()
+	{
+	}
+	UClass* Z_Construct_UClass_URoadRunnerFbxSceneImportFactory_NoRegister()
+	{
+		return URoadRunnerFbxSceneImportFactory::StaticClass();
+	}
+	struct Z_Construct_UClass_URoadRunnerFbxSceneImportFactory_Statics
+	{
+		static UObject* (*const DependentSingletons[])();
+#if WITH_METADATA
+		static const UE4CodeGen_Private::FMetaDataPairParam Class_MetaDataParams[];
+#endif
+#if WITH_METADATA
+		static const UE4CodeGen_Private::FMetaDataPairParam NewProp_RoadRunnerImportOptions_MetaData[];
+#endif
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerFbxSceneImportFactory.generated.h

@@ -1,0 +1,92 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/ObjectMacros.h"
+#include "UObject/ScriptMacros.h"
+
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+#ifdef ROADRUNNERIMPORTER_RoadRunnerFbxSceneImportFactory_generated_h
+#error "RoadRunnerFbxSceneImportFactory.generated.h already included, missing '#pragma once' in RoadRunnerFbxSceneImportFactory.h"
+#endif
+#define ROADRUNNERIMPORTER_RoadRunnerFbxSceneImportFactory_generated_h
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerFbxSceneImportFactory_h_32_RPC_WRAPPERS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerFbxSceneImportFactory_h_32_RPC_WRAPPERS_NO_PURE_DECLS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerFbxSceneImportFactory_h_32_INCLASS_NO_PURE_DECLS \
+private: \
+	static void StaticRegisterNativesURoadRunnerFbxSceneImportFactory(); \
+	friend struct Z_Construct_UClass_URoadRunnerFbxSceneImportFactory_Statics; \
+public: \
+	DECLARE_CLASS(URoadRunnerFbxSceneImportFactory, UFbxSceneImportFactory, COMPILED_IN_FLAGS(0), CASTCLASS_None, TEXT("/Script/RoadRunnerImporter"), NO_API) \
+	DECLARE_SERIALIZER(URoadRunnerFbxSceneImportFactory)
+
+
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerImporter.init.gen.cpp

@@ -1,0 +1,35 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/GeneratedCppIncludes.h"
+#ifdef _MSC_VER
+#pragma warning (push)
+#pragma warning (disable : 4883)
+#endif
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+void EmptyLinkFunctionForGeneratedCodeRoadRunnerImporter_init() {}
+	UPackage* Z_Construct_UPackage__Script_RoadRunnerImporter()
+	{
+		static UPackage* ReturnPackage = nullptr;
+		if (!ReturnPackage)
+		{
+			static const UE4CodeGen_Private::FPackageParams PackageParams = {
+				"/Script/RoadRunnerImporter",
+				nullptr,
+				0,
+				PKG_CompiledIn | 0x00000040,
+				0x617332A9,
+				0xCD3464C7,
+				METADATA_PARAMS(nullptr, 0)
+			};
+			UE4CodeGen_Private::ConstructUPackage(ReturnPackage, PackageParams);
+		}
+		return ReturnPackage;
+	}
+PRAGMA_ENABLE_DEPRECATION_WARNINGS
+#ifdef _MSC_VER
+#pragma warning (pop)
+#endif

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerImporterOptions.gen.cpp

@@ -1,0 +1,125 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/GeneratedCppIncludes.h"
+#include "RoadRunnerImporter/Public/RoadRunnerImporterOptions.h"
+#ifdef _MSC_VER
+#pragma warning (push)
+#pragma warning (disable : 4883)
+#endif
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+void EmptyLinkFunctionForGeneratedCodeRoadRunnerImporterOptions() {}
+// Cross Module References
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerImporterOptions_NoRegister();
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerImporterOptions();
+	COREUOBJECT_API UClass* Z_Construct_UClass_UObject();
+	UPackage* Z_Construct_UPackage__Script_RoadRunnerImporter();
+// End Cross Module References
+	void URoadRunnerImporterOptions::StaticRegisterNativesURoadRunnerImporterOptions()
+	{
+	}
+	UClass* Z_Construct_UClass_URoadRunnerImporterOptions_NoRegister()
+	{
+		return URoadRunnerImporterOptions::StaticClass();
+	}
+	struct Z_Construct_UClass_URoadRunnerImporterOptions_Statics
+	{
+		static UObject* (*const DependentSingletons[])();
+#if WITH_METADATA
+		static const UE4CodeGen_Private::FMetaDataPairParam Class_MetaDataParams[];
+#endif
+#if WITH_METADATA
+		static const UE4CodeGen_Private::FMetaDataPairParam NewProp_ImportSignals_MetaData[];
+#endif
+		static void NewProp_ImportSignals_SetBit(void* Obj);
+		static const UE4CodeGen_Private::FBoolPropertyParams NewProp_ImportSignals;
+#if WITH_METADATA
+		static const UE4CodeGen_Private::FMetaDataPairParam NewProp_OverwriteMaterials_MetaData[];
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerImporterOptions.generated.h

@@ -1,0 +1,96 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/ObjectMacros.h"
+#include "UObject/ScriptMacros.h"
+
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+#ifdef ROADRUNNERIMPORTER_RoadRunnerImporterOptions_generated_h
+#error "RoadRunnerImporterOptions.generated.h already included, missing '#pragma once' in RoadRunnerImporterOptions.h"
+#endif
+#define ROADRUNNERIMPORTER_RoadRunnerImporterOptions_generated_h
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerImporterOptions_h_28_RPC_WRAPPERS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerImporterOptions_h_28_RPC_WRAPPERS_NO_PURE_DECLS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerImporterOptions_h_28_INCLASS_NO_PURE_DECLS \
+private: \
+	static void StaticRegisterNativesURoadRunnerImporterOptions(); \
+	friend struct Z_Construct_UClass_URoadRunnerImporterOptions_Statics; \
+public: \
+	DECLARE_CLASS(URoadRunnerImporterOptions, UObject, COMPILED_IN_FLAGS(0 | CLASS_Config), CASTCLASS_None, TEXT("/Script/RoadRunnerImporter"), ROADRUNNERIMPORTER_API) \
+	DECLARE_SERIALIZER(URoadRunnerImporterOptions) \
+	static const TCHAR* StaticConfigName() {return TEXT("EditorPerProjectUserSettings");} \
+
+
+
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerReimportFactory.gen.cpp

@@ -1,0 +1,86 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/GeneratedCppIncludes.h"
+#include "RoadRunnerImporter/Public/RoadRunnerReimportFactory.h"
+#ifdef _MSC_VER
+#pragma warning (push)
+#pragma warning (disable : 4883)
+#endif
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+void EmptyLinkFunctionForGeneratedCodeRoadRunnerReimportFactory() {}
+// Cross Module References
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerReimportFactory_NoRegister();
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerReimportFactory();
+	ROADRUNNERIMPORTER_API UClass* Z_Construct_UClass_URoadRunnerFbxSceneImportFactory();
+	UPackage* Z_Construct_UPackage__Script_RoadRunnerImporter();
+// End Cross Module References
+	void URoadRunnerReimportFactory::StaticRegisterNativesURoadRunnerReimportFactory()
+	{
+	}
+	UClass* Z_Construct_UClass_URoadRunnerReimportFactory_NoRegister()
+	{
+		return URoadRunnerReimportFactory::StaticClass();
+	}
+	struct Z_Construct_UClass_URoadRunnerReimportFactory_Statics
+	{
+		static UObject* (*const DependentSingletons[])();
+#if WITH_METADATA
+		static const UE4CodeGen_Private::FMetaDataPairParam Class_MetaDataParams[];
+#endif
+		static const FCppClassTypeInfoStatic StaticCppClassTypeInfo;
+		static const UE4CodeGen_Private::FClassParams ClassParams;
+	};
+	UObject* (*const Z_Construct_UClass_URoadRunnerReimportFactory_Statics::DependentSingletons[])() = {
+		(UObject* (*)())Z_Construct_UClass_URoadRunnerFbxSceneImportFactory,
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Intermediate/Build/Linux/B4D820EA/UE4Editor/Inc/RoadRunnerImporter/RoadRunnerReimportFactory.generated.h

@@ -1,0 +1,92 @@
+// Copyright 1998-2019 Epic Games, Inc. All Rights Reserved.
+/*===========================================================================
+	Generated code exported from UnrealHeaderTool.
+	DO NOT modify this manually! Edit the corresponding .h files instead!
+===========================================================================*/
+
+#include "UObject/ObjectMacros.h"
+#include "UObject/ScriptMacros.h"
+
+PRAGMA_DISABLE_DEPRECATION_WARNINGS
+#ifdef ROADRUNNERIMPORTER_RoadRunnerReimportFactory_generated_h
+#error "RoadRunnerReimportFactory.generated.h already included, missing '#pragma once' in RoadRunnerReimportFactory.h"
+#endif
+#define ROADRUNNERIMPORTER_RoadRunnerReimportFactory_generated_h
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerReimportFactory_h_32_RPC_WRAPPERS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerReimportFactory_h_32_RPC_WRAPPERS_NO_PURE_DECLS
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerReimportFactory_h_32_INCLASS_NO_PURE_DECLS \
+private: \
+	static void StaticRegisterNativesURoadRunnerReimportFactory(); \
+	friend struct Z_Construct_UClass_URoadRunnerReimportFactory_Statics; \
+public: \
+	DECLARE_CLASS(URoadRunnerReimportFactory, URoadRunnerFbxSceneImportFactory, COMPILED_IN_FLAGS(0), CASTCLASS_None, TEXT("/Script/RoadRunnerImporter"), NO_API) \
+	DECLARE_SERIALIZER(URoadRunnerReimportFactory)
+
+
+#define CarlaUE4_Plugins_RoadRunnerImporter_Source_RoadRunnerImporter_Public_RoadRunnerReimportFactory_h_32_INCLASS \
+private: \
+	static void StaticRegisterNativesURoadRunnerReimportFactory(); \
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerFbxSceneImportFactory.cpp

@@ -1,0 +1,219 @@
+// This is free and unencumbered software released into the public domain.
+//
+// Anyone is free to copy, modify, publish, use, compile, sell, or
+// distribute this software, either in source code form or as a compiled
+// binary, for any purpose, commercial or non - commercial, and by any
+// means.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+// EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+// MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+// IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+// OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+// ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+// OTHER DEALINGS IN THE SOFTWARE.
+
+#include "RoadRunnerFbxSceneImportFactory.h"
+#include "RoadRunnerImporterLog.h"
+#include "RoadRunnerImporterOptionWindow.h"
+#include "RoadRunnerImporter.h"
+#include "RoadRunnerImporterOptions.h"
+#include "RoadRunnerImportUtil.h"
+
+#include <Factories/Factory.h>
+#include <Editor/UnrealEd/Public/Editor.h>
+#include <Editor/UnrealEd/Public/EditorBuildUtils.h>
+#include <UnrealEd.h>
+#include <Runtime/Engine/Classes/PhysicsEngine/BodySetup.h>
+#include <ObjectTools.h>
+#include <PackageTools.h>
+
+#include <Editor/MainFrame/Public/Interfaces/IMainFrameModule.h>
+
+#define LOCTEXT_NAMESPACE "RoadRunnerFBXSceneImportFactory"
+
+////////////////////////////////////////////////////////////////////////////////
+// Note: The "Import Into Level" button does not currently check for priority,
+// so this factory will only be called through RoadRunnerFbxFactory.
+URoadRunnerFbxSceneImportFactory::URoadRunnerFbxSceneImportFactory(const FObjectInitializer& objectInitializer) : Super(objectInitializer)
+{
+	SupportedClass = UBlueprint::StaticClass();
+	Formats.Add(TEXT("fbx;FBX meshes and animations"));
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerImportUtil.cpp

@@ -1,0 +1,444 @@
+// This is free and unencumbered software released into the public domain.
+//
+// Anyone is free to copy, modify, publish, use, compile, sell, or
+// distribute this software, either in source code form or as a compiled
+// binary, for any purpose, commercial or non - commercial, and by any
+// means.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+// EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+// MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+// IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+// OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+// ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+// OTHER DEALINGS IN THE SOFTWARE.
+
+#include "RoadRunnerImportUtil.h"
+#include "RoadRunnerImporterLog.h"
+
+#include <ObjectTools.h>
+#include <PackageTools.h>
+#include <Runtime/XmlParser/Public/XmlFile.h>
+
+#include <AssetRegistryModule.h>
+
+#include <Developer/AssetTools/Public/AssetToolsModule.h>
+#include <Developer/AssetTools/Public/IAssetTools.h>
+#include <PlatformFilemanager.h>
+
+namespace
+{
+	FString GetTempFolderAbsolute()
+	{
+#if ENGINE_MINOR_VERSION > 17
+		FString pluginsFolder = FPaths::ProjectPluginsDir();
+#else
+		FString pluginsFolder = FPaths::GamePluginsDir();
+#endif
+		FString filesystemTempPath = IFileManager::Get().ConvertToAbsolutePathForExternalAppForRead(*(pluginsFolder + "RoadRunnerImporter/Content/TEMP"));
+
+		return filesystemTempPath;
+	}
+}
+
+////////////////////////////////////////////////////////////////////////////////
+// Helper function to parse the material info from the *.rrdata.xml file
+TArray<FRoadRunnerImportUtil::MaterialInfo> FRoadRunnerImportUtil::ParseMaterialXml(FXmlNode * matList)
+{
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Private/RoadRunnerImporter.cpp

@@ -1,0 +1,766 @@
+// This is free and unencumbered software released into the public domain.
+//
+// Anyone is free to copy, modify, publish, use, compile, sell, or
+// distribute this software, either in source code form or as a compiled
+// binary, for any purpose, commercial or non - commercial, and by any
+// means.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+// EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+// MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+// IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+// OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+// ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+// OTHER DEALINGS IN THE SOFTWARE.
+
+#include "RoadRunnerImporter.h"
+#include "RoadRunnerImporterLog.h"
+#include "RoadRunnerImporterOptions.h"
+#include "RoadRunnerStyle.h"
+#include "RoadRunnerTrafficJunction.h"
+#include "RoadRunnerFbxSceneImportFactory.h"
+#include "RoadRunnerImportUtil.h"
+
+#include <AssetRegistryModule.h>
+#include <Developer/AssetTools/Public/AssetToolsModule.h>
+#include <Developer/AssetTools/Public/IAssetTools.h>
+#include <Editor/MaterialEditor/Public/IMaterialEditor.h>
+#include <Editor/UnrealEd/Public/Editor.h>
+#include <Editor/UnrealEd/Public/Kismet2/KismetEditorUtilities.h>
+#include <Editor/UnrealEd/Public/Layers/ILayers.h>
+#include <Factories/Factory.h>
+#include <Factories/FbxFactory.h>
+#include <unordered_map>
+#include <ObjectTools.h>
+#include <PackageTools.h>
+#include <Runtime/AssetRegistry/Public/ARFilter.h>
+#include <Runtime/Core/Public/Internationalization/Regex.h>
+#include <Runtime/Launch/Resources/Version.h>
+#include <Runtime/XmlParser/Public/XmlFile.h>
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Public/RoadRunnerImportUtil.h

@@ -1,0 +1,65 @@
+// This is free and unencumbered software released into the public domain.
+//
+// Anyone is free to copy, modify, publish, use, compile, sell, or
+// distribute this software, either in source code form or as a compiled
+// binary, for any purpose, commercial or non - commercial, and by any
+// means.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+// EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+// MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+// IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+// OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+// ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+// OTHER DEALINGS IN THE SOFTWARE.
+
+#pragma once
+
+#include "CoreMinimal.h"
+
+class FXmlNode;
+class UTexture;
+class UMaterialInstanceConstant;
+
+////////////////////////////////////////////////////////////////////////////////
+// Helper functions for importing from RoadRunner
+class ROADRUNNERIMPORTER_API FRoadRunnerImportUtil
+{
+public:
+	// Holds strings from the road runner metadata file
+	struct MaterialInfo
+	{
+		FString Name;
+		FString DiffuseMap;
+		FString NormalMap;
+		FString SpecularMap;
+		FString DiffuseColor;
+		FString SpecularColor;
+		FString SpecularFactor;
+		FString TransparencyMap;
+		FString TransparencyFactor;
+		FString Roughness;
+		FString Emission;
+		FString TextureScaleU;
+		FString TextureScaleV;
+		FString TwoSided;
+		FString DrawQueue;
+		FString ShadowCaster;
+		FString IsDecal;
+		FString SegmentationType;
+	};
+
+	static TArray<MaterialInfo> ParseMaterialXml(FXmlNode* matList);
+	static UTexture* ImportTexture(FString absFilePath, FString packagePath, bool setupAsNormalMap);
+	static void SetTextureParameter(UMaterialInstanceConstant* material, const FName& paramName, const FString& baseFilePath, const FString& texturePath, const FString& packagePath, bool isNormal);
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Import_map/Plugins/RoadRunnerImporter/Source/RoadRunnerImporter/Public/RoadRunnerImporter.h

@@ -1,0 +1,43 @@
+// This is free and unencumbered software released into the public domain.
+//
+// Anyone is free to copy, modify, publish, use, compile, sell, or
+// distribute this software, either in source code form or as a compiled
+// binary, for any purpose, commercial or non - commercial, and by any
+// means.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+// EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+// MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+// IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+// OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+// ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+// OTHER DEALINGS IN THE SOFTWARE.
+
+#pragma once
+
+#include "CoreMinimal.h"
+#include "Modules/ModuleManager.h"
+#include <UnrealEd.h>
+
+
+////////////////////////////////////////////////////////////////////////////////
+// Imports FBX files with RoadRunner metadata
+//	- Parses metadata XML lookaside file to set material properties and other attributes
+//	- Material instances are created from the base materials located in the plugin's content folder
+//  - Sets up signal components after importing
+class FRoadRunnerImporterModule : public IModuleInterface
+{
+public:
+	// Asset processing delegates
+	static void RoadRunnerPostProcessing(UFactory* inFactory, UObject* inCreateObject);
+
+	// Public helpers
+	static ANSICHAR* MakeName(const ANSICHAR* Name);
+
+	// IModuleInterface implementation
+	virtual void StartupModule() override;
+	virtual void ShutdownModule() override;
+
+	static const int PluginVersion = 3;
+	static const int32 TransparentRenderQueue = 1000;
+};

### 2020-11-13 CARLA_Tests_setup_guide/Running_tests/ROBOPILOT_Determinisim_V1.py

@@ -1,0 +1,1117 @@
+#!/usr/bin/env python
+
+############################################################################################
+############################ ROBOPILOT UE4 determinisim script  ############################
+############################   Trustworthy Systems Laboratory   ############################ 
+############################       University Of Bristol        ############################     
+############################################################################################
+
+
+import glob
+import os
+import sys
+
+try:
+	sys.path.append(glob.glob('../carla/dist/carla-*%d.%d-%s.egg' % (
+		sys.version_info.major,
+		sys.version_info.minor,
+		'win-amd64' if os.name == 'nt' else 'linux-x86_64'))[0])
+except IndexError:
+	pass
+
+import carla
+import argparse
+import logging
+import random
+import numpy as np
+import math
+import csv
+import time
+import pandas as pd
+import live_plotter as lv   # Custom live plotting library
+
+
+# Script level imports
+sys.path.append('../carla')
+from agents.navigation.roaming_agent import RoamingAgent
+from agents.navigation.basic_agent import BasicAgent
+from numpy import zeros
+
+
+import ROBOPILOT_agentsController
+
+from decimal import *
+
+#=============================================
+#===Declerations 
+#=============================================
+
+
+#===Input directory ==========================
+INPUT_FOLDER             = os.path.dirname(os.path.realpath(__file__)) +\
+							'/ExperimentInputs/'
+
+# INPUT_FILENAME           = 'Determinisim_TEST_Cars.txt'
+
+#===Output directory =========================
+RESULTS_OUTPUT_FOLDER    = os.path.dirname(os.path.realpath(__file__)) +\
+							'/ExperimentResutls/'
+
+# RESULTS_OUTPUT_FILENAME  = 'TEST_ID_001.txt'
+
+#===Ploting configurable parameters ==========
+FIGSIZE_X_INCHES   = 8      # x figure size of feedback in inches
[…]

### 2020-11-13 CARLA_Tests_setup_guide/Running_tests/ROBOPILOT_agentsController.py

@@ -1,0 +1,251 @@
+#!/usr/bin/env python3
+
+"""
+2D Controller for ROBOPILOT MAS agents control 
+"""
+import cutils
+import numpy as np
+
+class Controller2D(object):
+    def __init__(self, waypoints):
+        self.vars                = cutils.CUtils()
+        self._current_x          = 0
+        self._current_y          = 0
+        self._current_yaw        = 0
+        self._current_speed      = 0
+        self._desired_speed      = 0
+        self._current_frame      = 0
+        self._current_timestamp  = 0
+        self._start_control_loop = False
+        self._set_throttle       = 0
+        self._set_brake          = 0
+        self._set_steer          = 0
+        self._waypoints          = waypoints
+        self._conv_rad_to_steer  = 180.0 / 70.0 / np.pi
+        self._pi                 = np.pi
+        self._2pi                = 2.0 * np.pi
+
+    def update_values(self, x, y, yaw, speed, timestamp, frame):
+        self._current_x         = x
+        self._current_y         = y
+        self._current_yaw       = yaw
+        self._current_speed     = speed
+        self._current_timestamp = timestamp
+        self._current_frame     = frame
+        if self._current_frame:
+            self._start_control_loop = True
+
+    def update_desired_speed(self):
+        min_idx       = 0
+        min_dist      = float("inf")
+        desired_speed = 0
+        for i in range(len(self._waypoints)):
+            dist = np.linalg.norm(np.array([
+                    self._waypoints[i][0] - self._current_x,
+                    self._waypoints[i][1] - self._current_y]))
+            if dist < min_dist:
+                min_dist = dist
+                min_idx = i
+        if min_idx < len(self._waypoints)-1:
+            desired_speed = self._waypoints[min_idx][2]
+        else:
+            desired_speed = self._waypoints[-1][2]
+        self._desired_speed = desired_speed
+
+    def update_waypoints(self, new_waypoints):
+        self._waypoints = new_waypoints
+
+    def get_commands(self):
+        return self._set_throttle, self._set_steer, self._set_brake
+
+    def set_throttle(self, input_throttle):
+        # Clamp the throttle command to valid bounds
+        throttle           = np.fmax(np.fmin(input_throttle, 1.0), 0.0)
+        self._set_throttle = throttle
+
+    def set_steer(self, input_steer_in_rad):
+        # Covnert radians to [-1, 1]
[…]

### 2020-11-14 paper/cav_determinisim_main.tex
Added Kevin to authors

@@
 % \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
+\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
 \author[1,3]{Greg Chance}
-\author[2,3]{S\'everin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
+\author[1,3]{Kevin McAreavey}
+\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
 \author[2,3]{Tony Pipe}

### 2020-11-16 paper/cav_determinisim_main.tex
GC read through

@@
 
-\subsection{Definitions (Refer to Fig.~\ref{variance_description} throughout this section)}
+\subsection{Definitions}
+
+A list of definitions are given here of the terms used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
+
 \subsubsection{Determinism}
@@
 
-We adopt some terms from the mechanical engineering and statistics domain that best describe this situation~\cite{ADictionaryofMechanicalEngineering}.
+We adopt some terms from the mechanical engineering and statistics domain that best describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
 %
@@
 % when comparing the differences between multiple products.
-As an analogy, the simulator can be thought of as a manufacturing process, producing data and to find the value of the precision then the output must be measured for differences.
-Those differences describe the spread or variance in the process output and a hard limit on the variance can be defined, beyond which the output fails to meet the required tolerance, e.g.
+As an analogy, the simulator can be thought of as a manufacturing process, producing data and to find the value of the precision then the output must be measured for differences when the process is repeated.
+Those differences describe the spread or variance in the process output and a hard limit on the variance can be defined, Fig.~\ref{variance_description}a, beyond which the output fails to meet the required tolerance, e.g.
 rejected for quality control.
@@
 
-If a simulator is deterministic then it will produce results with absolute precision or zero variance and hence will be within acceptable tolerance.
+If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}b, and hence will be within acceptable tolerance.
 If the simulator is non-deterministic then there will be a variance in the output which can be measured.\\
@@
 \subsubsection{Accuracy}
-Precision and tolerance are not to be confused with \textit{accuracy} which describes how closely the generated output of a process aligns to a known standard.
+Precision and tolerance are not to be confused with \textit{accuracy} which describes how closely the generated output of a process aligns to a known standard or reference value.
 We therefore define accuracy as the difference between the true value (or reference value) and what has been achieved in the simulation or generation process.
@@
 % \subsection{Definition: Variance Measurement}
-\subsubsection{Variance Measurement}
+\subsubsection{Simulation Variance}
 If the simulator is non-deterministic then how can this variance be measured? 
 %
-This can be achieved from the actor state, or more specifically from the statistical variance of actor paths within a given scenario with a fixed set of actions.
-Actor path variance is chosen over other measures as this distance based metric forms the basis for many downstream verification tests such as assertion checking.
-
-The term \textit{simulation variance} refers to a measure of actor state variance in the simulation assuming fixed actions.
-Results are referred to in terms of deviation, rather than variance, as this is a more intuitive dimension to comprehend when considering verification tests, e.g.
+This can be achieved by monitoring any simulated output variables that should be consistent from run to run. 
+% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
+Actor path variance was chosen over other variables as this distance based metric forms the basis for many downstream verification tests such as assertion checking. 
+% 
+Therefore the term \textit{simulation variance} refers to a measure of actor position variance in the simulation with respect to time assuming fixed actions.
+Results are referred to in terms of deviation, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
 vehicle close-passing.
@@
 \subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator which is a time series of all actor positions ($x,y,z$) at regular intervals based on the time-step $dt$.
+A simulation trace is the output log from the simulator which is a time series of all actor positions ($x,y,z$) at regular intervals based on a regular time interval.
 This definition could be extended to include other variables, but position is sufficient for this analysis.\\
@@
 A deterministic simulation environment guarantees that tests are repeatable.
-Using mechanical engineering terms, a deterministic simulator can be considered to have zero \textit{variance}.
-%
+A deterministic simulator can be considered to have zero \textit{variance}.
 If the variance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
@@
 % Challenges of SBV
-When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs related to non-determinism a high priority for games developers.
 It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
 However, the situation is very different for AV development and testing.
-Thus, our main research questions is:
+Thus, our main research question is:
 %
@@
 %
-How can one assess the extent to which a simulation environment is deterministic or has a permissible variance?
+How can one assess the extent to which a simulation environment is deterministic or has a permissible level of variance?
 \todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
@@
 The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation.
-To understand the impact of collisions and high resource utilisation the \textit{restricted} column shows a subset of the results where data post-collision has been removed and where data above a resource utilisation level of 75\% has also been removed.}
+To understand the impact of collisions and high resource utilisation, the \textit{restricted} column shows a subset of the results where post-collision data and experiments above 75\% resource utilisation have been removed.}
 \label{TableOfExperiments}
@@
 An investigation into the response of process scheduling on the simulation variance was undertaken.
-The experiment was repeated using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}.
+The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}.
 Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
@@
 Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation but this did not account for all the difference in the observed results.
-%
-This unaccounted difference in the variance may be due to the lack of  control that \texttt{NICE} has over the processes\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}.
-%~\cite{Abanoub can you cite that quote you found here please}.
+This is demonstrated in the maximum priority setting where the bars in plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
+%
+This unaccounted difference in the variance may be due to the lack of absolute control that \texttt{NICE} has over the process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
 

### 2020-11-17 paper/cav_determinisim_main.tex
AG corrections

@@
 %
-\IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it.
+% Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, benefiting from complete control over (virtual) actors and their environment
+% , offering a relatively inexpensive means to explore test cases while also
+% \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. 
+\IEEEPARstart{S}{}imulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
+%
 Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of safe and legal driving rules, termed assertions, that must not be violated.
@@
 %
-However, if a simulation is non-deterministic then then there will be a variation in the generated output data. \\
+However, if a simulation is non-deterministic then there will be a variation in the generated output data. \\
 
@@
 % \subsection{Definitions: Precision, Variance \& Tolerance}
-\subsubsection{Precision, Variance \& Tolerance}
+\subsubsection{Variance, Precision \& Tolerance}
 
@@
 %
-\textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
-%
-\textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-%
-\textit{Tolerance} is defined as the permissible limit of the variance.
+% \textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
+% %
+% \textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
+%
+\textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
+%
+Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken
+%
+\textit{Tolerance} is defined as the permissible limit of the variance, or in short the \textit{permissible variance}.
 
@@
 \subsubsection{Accuracy}
-Precision and tolerance are not to be confused with \textit{accuracy} which describes how closely the generated output of a process aligns to a known standard or reference value.
+Precision and tolerance should not to be confused with \textit{accuracy} which describes how closely the generated output of a process aligns to a known standard or reference value.
 We therefore define accuracy as the difference between the true value (or reference value) and what has been achieved in the simulation or generation process.
 %
-For a driving simulation the reference value, for which accuracy can be measured against, may be the real world that it seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}.
+For a driving simulation the reference value, for which accuracy can be measured against, may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}.
 In most cases such accuracy will not be possible due to computational demands of such an exact replica and in most cases is unnecessary, where some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
@@
 
+% \subsection{Definition: Simulation Trace}
+\subsubsection{Simulation Trace}
+A simulation trace is the output log from the simulator which, herein, is a time series of all actor positions ($x,y,z$) at regular time intervals.
+This definition could be extended to include other variables.\\%, but position is sufficient for this analysis.\\
+
 % \subsection{Definition: Variance Measurement}
-\subsubsection{Simulation Variance}
+\subsubsection{Simulation Variance and Deviation}
 If the simulator is non-deterministic then how can this variance be measured? 
@@
 % In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
-Actor path variance was chosen over other variables as this distance based metric forms the basis for many downstream verification tests such as assertion checking. 
+Actor path variance, derived from the simulation trace, is chosen over other variables as this distance based metric forms the basis for many downstream verification tests.% such as assertion checking.
 % 
 Therefore the term \textit{simulation variance} refers to a measure of actor position variance in the simulation with respect to time assuming fixed actions.
-Results are referred to in terms of deviation, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
+
+Deviation is the square root of variance, which is more intuitive than variance to comprehend when considering verification tests.
+%Results are referred to in terms of deviation, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
 vehicle close-passing.
 Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
-
-% \subsection{Definition: Simulation Trace}
-\subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator which is a time series of all actor positions ($x,y,z$) at regular intervals based on a regular time interval.
-This definition could be extended to include other variables, but position is sufficient for this analysis.\\
 
@@
 %
-Section~\ref{s:background} briefly introduces how game engines work before investigating potential sources of non-determinism in game engines.
+Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
 %
@@
 %
-A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests} although the AV simulation domain introduces it's own unique challenges that are not considered in this paper.
+A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests} although the AV simulation domain introduces it's own unique challenges that were not considered in that paper.
 
@@
 %
-If calculations are performed on a single core, then concurrency of multiple threads is not an issue.
+%If calculations are performed on a single core, then concurrency of multiple threads is not an issue.
 \todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
@@
 An investigation into the response of process scheduling on the simulation variance was undertaken.
-The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}.
+The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the program \texttt{NICE}.
 Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
@@
 
-The results presented in a box and whisker format, Fig.~\ref{NICEExperimentStressSummary}, show that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a 0 NICE value (central plot bars).
+The results presented in a box and whisker format, Fig.~\ref{NICEExperimentStressSummary}, show that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a NICE value of 0 (central plot bars).
 Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation but this did not account for all the difference in the observed results.

### 2020-11-17 paper/cav_determinisim_main.tex
AG updated abstract

@@
 %
-A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
-%
-Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
-\todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
-%
-% KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
-% These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
-%
-However, game engines do not need to be deterministic per se. % when used for games. 
-%
-This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively.
-%
-\todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
-%
-A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification.
+A key requirement for simulation-based development and verification is determinism.
+%
+Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
+%
+Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
+%
+However, game engines are seldom deterministic.
+
+This paper first reviews and identifies the potential causes of non-determinisitc behaviours in game engines. 
+%
+This is then followed by a case study using CARLA, an open source autonomous driving simulation environment powered by Unreal Engine, which highlightes its inherent shortcomings in providing sufficient precision in experimental results.
+%
+Different configurations and utilisations of the software and hardware are explored to determine an opertional domain where the simulation precision is satisfactory low i.e.
+variance between repeats becomes negligible for development and testing work.
+
+Finally a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
+% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
+% %
+% A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
+% %
+% Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
+% \todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
+% %
+% % KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
+% % These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
+% %
+% However, game engines do not need to be deterministic per se. % when used for games. 
+% %
+% This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively.
+% %
+% \todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
+% %
+% A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification.
 The proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-%
-% \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}
+% %
+% % \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}
 

### 2020-11-17 paper/cav_determinisim_main.tex
AG updated and abstract

@@
 
-This paper first reviews and identifies the potential causes of non-determinisitc behaviours in game engines. 
-%
-This is then followed by a case study using CARLA, an open source autonomous driving simulation environment powered by Unreal Engine, which highlightes its inherent shortcomings in providing sufficient precision in experimental results.
-%
-Different configurations and utilisations of the software and hardware are explored to determine an opertional domain where the simulation precision is satisfactory low i.e.
+This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.
+%
+This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
+%
+Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
 variance between repeats becomes negligible for development and testing work.
 
-Finally a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
+Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
 % Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 

### 2020-11-18 paper/cav_determinisim_main.tex
Changed format to fit ITS guidlines

@@
 % \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
-\author[1,3]{Greg Chance}
-\author[1,3]{Kevin McAreavey}
-\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-\author[2,3]{Tony Pipe}
-\author[1,3]{Kerstin Eder}
-\affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
-\affil[2]{University of West of England, Bristol, UK}
-\affil[3]{Bristol Robotics Laboratory, Bristol, UK}
+% \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
+% \author[1,3]{Greg Chance}
+% \author[1,3]{Kevin McAreavey}
+% \author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
+% \author[2,3]{Tony Pipe}
+% \author[1,3]{Kerstin Eder}
+% \affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
+% \affil[2]{University of West of England, Bristol, UK}
+% \affil[3]{Bristol Robotics Laboratory, Bristol, UK}
+
+% \author{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
+% \author{Greg Chance}
+% \author{Kevin McAreavey}
+% \author{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
+% \author{Tony Pipe}
+
+\author{Abanoub Ghobrial, Greg Chance, Kevin McAreavey, Severin Lemaignan, Tony Pipe, Kerstin Eder 
+\thanks{{\footnotesize
+Manuscript  
+received ...;
+revised ...;  
+accepted.... 
+Date of publication ...;
+date of current version ....
+
+This research has in part been funded by the ROBOPILOT and CAPRI projects.
+Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
+The Associate Editor for this paper was ....(\textit{Corresponding author: Abanoub Ghobrial.})
+
+A.Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
+G.Chance (e-mail: greg.chance@bristol.ac.uk), 
+K.McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
+and 
+K.Eder (e-mail: kerstin.eder@bristol.ac.uk) 
+are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom.
+
+S.Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
+and
+T.Pipe (e-mail: tony.pipe@brl.ac.uk), 
+are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.
+
+Digital Object Identifier ....
+}}}
+% \affil{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
+% \affil{University of West of England, Bristol, UK}
+% \affil{Bristol Robotics Laboratory, Bristol, UK}
+%
+\markboth{IEEE TRANSACTIONS ON INTELLIGENT TRANSPORTATION SYSTEMS VOL.
+..., NO.
+..., date...}{Ghobrial \MakeLowercase{\textit{et al.}}: On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
+%
 \maketitle
-
+%
 \begin{abstract}
@@
 % ***************************************************
-\section*{Acknowledgements}
-This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
-projects are part-funded by the Centre for Connected and Autonomous
-Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
-103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
+% \section*{Acknowledgements}
+% This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
+% projects are part-funded by the Centre for Connected and Autonomous
+% Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
+% 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
 \printbibliography

### 2020-11-18 paper/cav_determinisim_main.tex
Added Tony's suggested abstract modifications

@@
 %
-A key requirement for simulation-based development and verification is determinism.
-%
-Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
-%
-Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
+A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history.
+%
+Thus, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
 %
@@
 Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
+
+% =========================================
+% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
+% %
+% A key requirement for simulation-based development and verification is determinism.
+% %
+% Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
+% %
+% Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
+% %
+% However, game engines are seldom deterministic.
+
+% This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.
+% %
+% This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
+% %
+% Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
+variance between repeats becomes negligible for development and testing work.
+
+% Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
+
+% =========================================
 % Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 

### 2020-11-19 paper/cav_determinisim_main.tex
KM comments by GC

@@
 
-A list of definitions are given here of the terms used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
+A list of definitions are given here which are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
+\\
 
@@
 %
-\textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
+% \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
+\textit{Variance} is used here to define the spread, or distribution, of the generated output data with respect to the mean value.
+\textit{Precision} is synonymous with \textit{variance} although inversely related mathematically.
 %
@@
 
-% \subsection{Definition: Accuracy}
 \subsubsection{Accuracy}
-Precision and tolerance should not to be confused with \textit{accuracy} which describes how closely the generated output of a process aligns to a known standard or reference value.
-We therefore define accuracy as the difference between the true value (or reference value) and what has been achieved in the simulation or generation process.
-%
-For a driving simulation the reference value, for which accuracy can be measured against, may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}.
+Precision and tolerance should not to be confused with \textit{accuracy} which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
+We therefore define accuracy as the difference between the true value, or reference value, and what has been achieved in the simulation or generation process.
+%
+For a driving simulation the reference value 
+% , for which accuracy can be measured against, 
+may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}.
 In most cases such accuracy will not be possible due to computational demands of such an exact replica and in most cases is unnecessary, where some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
@@
 
-% \subsection{Definition: Simulation Trace}
 \subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator which, herein, is a time series of all actor positions ($x,y,z$) at regular time intervals.
-This definition could be extended to include other variables.\\%, but position is sufficient for this analysis.\\
-
-% \subsection{Definition: Variance Measurement}
-\subsubsection{Simulation Variance and Deviation}
-If the simulator is non-deterministic then how can this variance be measured? 
-%
-This can be achieved by monitoring any simulated output variables that should be consistent from run to run. 
+A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) at regular time intervals.
+This definition could be extended to include other variables.
+%, but position is sufficient for this analysis.\\
+A set of simulation traces derived from the same input then forms the experimental data on which variance is calculated. \\
+
+\subsubsection{Simulation Variance \& Deviation}
+If the simulator is non-deterministic then how can this variance be measured? This can be achieved by monitoring any simulated output variables that should be consistent from run to run. 
 % In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
@@
 Therefore the term \textit{simulation variance} refers to a measure of actor position variance in the simulation with respect to time assuming fixed actions. 
-
-Deviation is the square root of variance, which is more intuitive than variance to comprehend when considering verification tests.
-%Results are referred to in terms of deviation, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
+%
+\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive value to comprehend when considering interpretation of verification tests. 
+%Results are referred to in terms of deviation, the square root of variance, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
 vehicle close-passing.
 Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
-
-% \subsection{Definition: Scene, Scenario \& Situation}
+\\
+
 \subsubsection{Scene, Scenario \& Situation}
@@
 An investigation into the response of process scheduling on the simulation variance was undertaken.
-The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the program \texttt{NICE}.
-Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
+The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
 %
@@
 
-The results presented in a box and whisker format, Fig.~\ref{NICEExperimentStressSummary}, show that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a NICE value of 0 (central plot bars).
+The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range, non-outlier limits by the whiskers and a horizontal bar for the median.
+The figure shows that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a NICE value of 0 (central plot bars).
 Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation but this did not account for all the difference in the observed results.

### 2020-12-01 paper_arxiv/arxiv_cav_determinisim_main.tex
Copy of the paper for archives and excel results included

--- 2020-11-19 paper/cav_determinisim_main.tex
@@
-\documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
-% \input{archive/main.config.tex}
-\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
-% \addbibresource{IROS.bib}
-\addbibresource{cav_determinism.bib}
+\documentclass[runningheads,twocolumn,a4paper,10pt]{llncs}
+\usepackage{etex}
+\reserveinserts{28}
+\pdfoutput=1
+
+\setcounter{tocdepth}{3}
+
+\usepackage[a4paper, total={7in, 10in}]{geometry}
+\usepackage[linesnumbered,lined,boxed,commentsnumbered,ruled]{algorithm2e}
+\renewcommand{\algorithmcfname}{ALGORITHM}
+\SetAlFnt{\small}
+\SetAlCapFnt{\small}
+\SetAlCapNameFnt{\small}
+\SetAlCapHSkip{0pt}
+\IncMargin{-\parindent}
+\usepackage{ textcomp }
+\usepackage{ comment }
+% \usepackage[color=yellow!60,textsize=footnotesize,obeyDraft,draft]{todonotes}
+\usepackage{colortbl}
+\usepackage{booktabs}
+\usepackage{capt-of}
+\usepackage{makecell}
+\usepackage{textcomp }
+\usepackage{multirow}
+\usepackage{listings}
+\usepackage{graphicx}
+\usepackage{colortbl}
+\usepackage{booktabs}
+\usepackage{amsmath}
+\usepackage[font=footnotesize,labelfont=bf]{subcaption}
+\usepackage[hidelinks]{hyperref}
+\usepackage[capitalize]{cleveref}
+\captionsetup{compatibility=false}
+\usepackage{tikz}
+\usepackage{pgfplots}
+\usepackage[para,online,flushleft]{threeparttable}
+\usepackage{multirow}
+\usepackage{wasysym}
+\usepackage{amssymb}
+\usepackage{enumitem}
+\usetikzlibrary{arrows}
+% \usepackage{subfig}
+\usepackage{subcaption}
+\usepackage{breakcites}
+
 \usepackage{array}
@@
 \usepackage{authblk}
-\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
 % \pagenumbering{gobble} %suppresses page numbers
@@
 \usepackage{siunitx}
-% \usepackage{draftwatermark}
-
-%-------------------------------------------------------------------------------
+
+% \usepackage{multicol}
+\setcounter{secnumdepth}{3}
+
+% Make tables a little bit easier to read
+\renewcommand{\arraystretch}{1.15}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% NOTES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcounter{mnotei}
+\setcounter{mnotei}{0}
+\newcommand{\mnote}[1]{%
+  {\scriptsize\textsf{$^{\textcolor{red}{[n.\themnotei]}}$}}%
+  \marginpar{\scriptsize\textsf{\textcolor{red}{[n.\themnotei]: #1}}}%
+  % \marginpar{\scriptsize\textsf{n.\themnotei: #1}}%
+  \stepcounter{mnotei}
+}
+%\renewcommand{\mnote}[1]{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Provide paragraph-style table columns
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+% Give more marginpar
+\setlength{\marginparwidth}{25mm}
+
+% Give more marginpar
+\setlength{\marginparwidth}{25mm}
+
+% Prompt people to generate graphs
+\newcommand{\includegraphicsmaybe}[2]{
+    \IfFileExists{#2}{\includegraphics[#1]{#2}}{
+    \detokenize{File #2 is missing, maybe you need to run plots.py?}
+}}
+
+\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
 
 \begin{document}
-
-% ***************************************************
-%  TITLE
-% ***************************************************
+\bibliographystyle{ieeetr}
+\mainmatter
+
 \title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-% \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-% \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
-% \author[1,3]{Greg Chance}
-% \author[1,3]{Kevin McAreavey}
-% \author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-% \author[2,3]{Tony Pipe}
-% \author[1,3]{Kerstin Eder}
-% \affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
-% \affil[2]{University of West of England, Bristol, UK}
-% \affil[3]{Bristol Robotics Laboratory, Bristol, UK}
-
-% \author{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
-% \author{Greg Chance}
-% \author{Kevin McAreavey}
-% \author{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-% \author{Tony Pipe}
-
-\author{Abanoub Ghobrial, Greg Chance, Kevin McAreavey, Severin Lemaignan, Tony Pipe, Kerstin Eder 
-\thanks{{\footnotesize
-Manuscript  
-received ...;
-revised ...;  
-accepted.... 
-Date of publication ...;
-date of current version ....
-
-This research has in part been funded by the ROBOPILOT and CAPRI projects.
-Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-The Associate Editor for this paper was ....(\textit{Corresponding author: Abanoub Ghobrial.})
-
-A.Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
-G.Chance (e-mail: greg.chance@bristol.ac.uk), 
-K.McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
-and 
-K.Eder (e-mail: kerstin.eder@bristol.ac.uk) 
-are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom.
-
-S.Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
-and
-T.Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.
-
-Digital Object Identifier ....
-}}}
-% \affil{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
-% \affil{University of West of England, Bristol, UK}
-% \affil{Bristol Robotics Laboratory, Bristol, UK}
-%
-\markboth{IEEE TRANSACTIONS ON INTELLIGENT TRANSPORTATION SYSTEMS VOL.
-..., NO.
-..., date...}{Ghobrial \MakeLowercase{\textit{et al.}}: On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-%
+
+\titlerunning{On Determinism of Game Engines... }%used for Simulation-based }%Autonomous Vehicle Verification}
+
+\author{Abanoub Ghobrial\inst{1,3}, Greg Chance\inst{1,3}, Kevin McAreavey\inst{1,3}, Severin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}
+
+\authorrunning{A. Ghobrial et al.}
+\institute{University of Bristol, Bristol, UK \and University of the West of England, Bristol, UK \and Bristol Robotics Laboratory, Bristol, UK}
+\tocauthor{Authors' Instructions}
 \maketitle
-%
-\begin{abstract}
-\noindent 
-Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
+
+\makeatletter
+\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
+                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
+                       {4\p@ \@plus 2\p@ \@minus 2\p@}%
+                       {\normalfont\normalsize\bfseries\boldmath
+                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
+\makeatother
+
+% \begin{multicols}{2}
+
+\textbf{\textit{Abstract}--Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
 %
@@
 %
-However, game engines are seldom deterministic.
-
+However, game engines are seldom deterministic.\\
 This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.  
@@
 variance between repeats becomes negligible for development and testing work.
-
-Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
-% %
-% A key requirement for simulation-based development and verification is determinism.
-% %
-% Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
-% %
-% Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
-% %
-% However, game engines are seldom deterministic.
-
-% This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.
-% %
-% This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
-% %
-% Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
-variance between repeats becomes negligible for development and testing work.
-
-% Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
-% %
-% A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
-% %
-% Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
-% \todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
-% %
-% % KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
-% % These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
-% %
-% However, game engines do not need to be deterministic per se. % when used for games. 
-% %
-% This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively.
-% %
-% \todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
-% %
-% A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification.
-The proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-% %
-% % \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}
-
-\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline,color=yellow!40]{AG Alternative: A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}
-
-% The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
-It is important for these engines to be deterministic in order for one to carry out these different processes.
-However, such engines are inherently non-deterministic.
-We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic.
-We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.
-\end{abstract}
-
-\begin{IEEEkeywords}
-Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
-\end{IEEEkeywords}
-\IEEEpeerreviewmaketitle
+\\
+Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.}
 
@@
 \section{Introduction} \label{s:introduction}
-% What is verification
-% Can't use this again as this is how the AI Test paper starts. -> Avoid self-plagiarism!
-%\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. 
-%
-% Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, benefiting from complete control over (virtual) actors and their environment
-% , offering a relatively inexpensive means to explore test cases while also
-% \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. 
-\IEEEPARstart{S}{}imulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
+Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
 %
@@
 %
-%\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
-% 
-% GC Removing the following para:
-%For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
-CARLA currently offers various basic types of cameras and %ray-case based
-% lidar; extensions are still under development. 
-% \todo{GC: what is ray-case? KIE: Not sure. :(}
-%
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. %~\cite{Javaid2013,cyres}
-%
-% The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
+Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. 
 %
@@
 
-\todo[inline,color=green!40]{GC New below}
-
 \subsection{Definitions}
@@
 A list of definitions are given here which are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
-\\
 
@@
 %
-% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
-Specifically, we focus on the variance of actor paths within a given scenario.
 
@@
 %
-However, if a simulation is non-deterministic then there will be a variation in the generated output data. \\
+However, if a simulation is non-deterministic then there will be a variation in the generated output data.
 
@@
 
-% \subsection{Definitions: Precision, Variance \& Tolerance}
 \subsubsection{Variance, Precision \& Tolerance}
@@
 %
-% \textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
-% %
-% \textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-%
-% \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
 \textit{Variance} is used here to define the spread, or distribution, of the generated output data with respect to the mean value.
@@
 
-% The engineering terms \textit{precision} and \textit{tolerance} suitably describe how the simulation data vary when the same repeated stimulus is provided to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}.
-%
-%
-% As an analogy, the simulator can be thought of as a manufacturing process, producing data and if tasked to repeatedly execute the same process we must ask how \textit{precise} can this process be and what, if any, will the \textit{variance} be?
-% when comparing the differences between multiple products.
 As an analogy, the simulator can be thought of as a manufacturing process, producing data and to find the value of the precision then the output must be measured for differences when the process is repeated.
@@
 rejected for quality control.
-%
-% In relation to manufacturing, the precision of any computer based task is casually assumed to be absolute, for example one may question the manufactured precision of a physical component but not the computer assisted drawing that it was made from.
-%but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point. %\textbf{[ref]}. %
 %
@@
 In most cases such accuracy will not be possible due to computational demands of such an exact replica and in most cases is unnecessary, where some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
-\\
 
@@
 This definition could be extended to include other variables.
-%, but position is sufficient for this analysis.\\
-A set of simulation traces derived from the same input then forms the experimental data on which variance is calculated. \\
+A set of simulation traces derived from the same input then forms the experimental data on which variance is calculated.
 
@@
 If the simulator is non-deterministic then how can this variance be measured? This can be achieved by monitoring any simulated output variables that should be consistent from run to run.
-% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
-Actor path variance, derived from the simulation trace, is chosen over other variables as this distance based metric forms the basis for many downstream verification tests.% such as assertion checking.
+
+Actor path variance, derived from the simulation trace, is chosen over other variables as this distance based metric forms the basis for many downstream verification tests. 
 % 
@@
 \textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive value to comprehend when considering interpretation of verification tests.
-%Results are referred to in terms of deviation, the square root of variance, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
-vehicle close-passing.
-Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
-\\
 
@@
 A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
-
-% \todo[inline]{values - Does this mean stimulus or other parameters? GC - changed to parameters}
-
-% Verification and Determinism 
-% Why is determinism important for verification
-%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
-% KE Text
-\todo[inline,color=red!40]{old: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
 
@@
 
-% GC Text
-\todo[inline,color=green!40]{GC Alternative: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}.
-If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.}
-
-% This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
-The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
-\textit{Tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
-%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
-Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour.
-%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
-%
-
 If the simulation is non-deterministic and has a non-permissible variance in, for example, actor positions, this may lead to assessment errors, making it difficult to understand and remove bugs.
 In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
-%
-
-% \todo[inline]{GC Edit: If the simulation is non-deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
 
@@
 %
-\todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
-%
 How can one assess the extent to which a simulation environment is deterministic or has a permissible level of variance?
-\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
 
@@
 %
-\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
-}
-%
 By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
 %
 Carla exhibits a permissible level of variance when system utilisation is restricted to 75\% or less and ensuring termination of scenarios once a vehicle collision has been detected.
-%
-\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
 
 The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
-%
-\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
-In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
-}
 %
@@
 In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact on verification results.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
-In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
-}
-%
-% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
-%We present a case study to illustrate the use of this method.
-The study highlights the effects of non-determinism of a game engine on simulation results.
-Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
-%
-%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
-%
 
@@
 %
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
-
-% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (AVs) on roads is a delicate if not utterly fascinating subject.
-Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.
-
-% Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.
-
-% It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
-
-% \noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
-
-% As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
-
-% ======================
-% This is all from "FrameworkAndChallenges" without using quotes 
-% \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system.
-Similarly, the use of simulation for traffic modelling is well established.
-Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate AVs\cite{FrameworkAndChallenges}.\\\\
-
-% While traffic-level simulators can model road layouts, they do not require as much realism or detail as AVs require.
-On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for AV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
-% Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
-% Consequently, one of the major routes AVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate AVs \cite{FrameworkAndChallenges}.
-% ======================
-
-% Robopilot, Capri, CARLA, Apollo, Airsim, Udacity etc.
-are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}.
-\\\\
-% Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
-% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of AV software.\\\\
-% \TODO{Remove this paragraph?} Determinism of game (physics) engines has not attracted much attention by engines' developers nor researchers.
-This previously was not important since in gaming, determinism is not critical.
-% Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
-% On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
-
-% This paper aims to investigate how deterministic physics engines are for usage in V\&V for AV simulations; and will cover the list of following points:
-% \begin{itemize}[leftmargin=*]
-%     \item Necessary background on game engines used in AV simulations.
-%     \item Review of sources of non-determinism in such engines.
-%     \item Provide a method of how the level of non-determinism can be investigated and defined.
-%     \item A case study showing the implementation of this methodology in an engine widely used in AVs simulations (Unreal Engine).
-% \end{itemize}
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.
 
@@
 Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
-% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
 
 Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.  
-%However, when these engines are used for AV development then deterministic behaviour is required.
 %
@@
 %
-\todo[inline]{GC line added to introduction.
-We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-%
-\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-}
-%
 This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
-%
-% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations.
-Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
-
-% ======= Overview of a Game Loop
-% \subsection{Game Engine Overview} \label{GameLoopSection}
 
@@
 
-% \todo[inline]{AG: remove the clock symbol, possibly remove the ticks chart as this is confusing given they are not ticks but event markers}
-
-% The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment.
-For example, the throttle and steering angle from the controller of an AV.
-The update stage interprets the input from all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states.
-The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
-
-% \todo[inline]{Fig 1: update rates of which clock?
-GC-is a clock confusing as it does not run on a 'clock' in computer science terms, rather it is a timer which is used as reference to compare to a 'wall clock'}
-
-% \todo[inline]{Should there not also be a clock in the overall loop? GC-yes the wall clock but it may get confusing as that is a real clock and the game loop clock is a timer?}
-
-% \todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.
-Unity page on multi-threading \footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and a plugin for multi-thread management \footnote{\url{https://forum.unity.com/threads/task-for-unity-scripting-threading.462971/?_ga=2.228561631.331624405.1587372521-1435703724.1587372521}}}
-
-% 
-% It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below.
-
-% \todo[inline]{We should agree if the 'game clock' really is a clock or just times the computation for each game loop.}
-
-% \todo[inline]{For AG: We use the phase 'game clock tick'- where did the come from as I think it is misleading.}
-
-% As the game progresses, the game loop runs continuously according to the game clock. \todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
-%
-% At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
-This measures how far the game's clock is behind compared to the real world.
-
 The game loop cycle starts with initialising the scene and actors.
@@
 The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}.
-% \footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}. 
+%  
 When the frame is rendered the game loop cycle returns to processing inputs.
@@
 
-% \todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}.
-When the frame is rendered the game loop cycle returns to processing inputs.}
-%
-% Then the user inputs are processed.
-There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
-Rendering occurs once the game clock has caught up with the real world clock.
-The process then repeats.
-
-The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
+The physics engine operates according to a time step, $dt$. 
+%
 The shorter this time step is, the smoother the interpretation of the physical dynamics will be.
-%processing time it takes to catch up to game clock.
-% and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
-%
-% Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
 %
@@
 Note that both residual lag and extrapolation could affect perception stack testing.
-% This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
-% \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
-%
-% \todo[inline]{Why is this part of the figure?
-It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.
-GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
-%
-% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
-In exceptional cases where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines,
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
-% \todo[inline]{Explain how this lag may affect perception stack testing}
-
-% \todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?
-GC-suggest remove this part of the diagram, too confusing as there are not really any clocks or ticks - we should remove all mention of ticks too.}
-
-% The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
-If $dt$ is too large or irregularly spaced then the rendered motion may not be smooth and interpolation may be required at additional computational cost.
-
-% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines.
-These engines update at fixed intervals and render whenever they can, which is not steady.
-This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates.
-In this case, the engine will use extrapolation techniques to give an estimate of where the object should be.
-This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
-%
-\todo[inline,color=red!40]{old: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy.
-In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-}
+%
+In exceptional cases where computational resources are scarce, the fixed time step can be greater than the time between render ticks and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
 %
@@
 This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-%
-\todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-The gamer neither wishes a faithful representation of reality, i.e.
-can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
-can be low precision.
-In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
-%
-\todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
-
-%\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
-This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer).
-For instance, see below, but I'm not sure whether this is the right place.}
-
-% ======= Sources of non-determinism
+
+% ***************************************************
+%  Sources of non-determinism
+% *************************************************** 
 \section{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}
 
-% The following discussion reflects the results of our investigation into the potential sources of non-determinism that could affect game engines.
-We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
 The following review discusses the potential sources of non-determinism that were found in the literature or found as part of our investigation into game engines.
@@
 %
-\todo[inline,color=green!40]{GC Addition below}%
-%
 A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests} although the AV simulation domain introduces it's own unique challenges that were not considered in that paper.
-
-% MOVED TO INTRO \todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-
-\todo[inline]{We should also refer to literature like:\\
-Making System User Interactive Tests Repeatable: When and What Should We Control?~\cite{when-and-what-should-we-control} and
-Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.}
-
-\medskip
-
-% \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
-The main reasons that we think are causes of these engines' non-determinism are discussed below.
-\\\\
 
@@
 \subsection{Floating-Point Arithmetic}
-
 Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
@@
 %
-% SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
 Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
@@
 Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-%
-%If calculations are performed on a single core, then concurrency of multiple threads is not an issue.
-\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
-
-Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
-Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-%
-If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
-% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.}
-}
-%
-% Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.
-
-% Translating to the usage of AV simulations in physics engines,
+
 In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
-% \todo[inline]{state what assertions are (used for) GC-see below} 
-%In the field of AV verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
 %
@@
 \subsection{Scheduling, Concurrency and Parallelisation}
-% \todo[inline]{is this at runtime?} %
-% info at:
-% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
-% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
-% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
-% https://en.wikipedia.org/wiki/Concurrent_computing
-% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
 Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy of the operating system.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-% \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
 %
@@
 %
-\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-}
-%
 Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
@@
 %
-\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
-}
 
@@
 Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}.
-%\noindent\underline{\textit{Concurrency:}}
-% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
-%\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
-Consider also parallelism, e.g.
-in GPUs.
-Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism.
-Say which parts of a game engine could reasonably be expected to run concurrently.}
-%
-%Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
-This technique can lead to issues such as shared resource access conflicts due to mis-managed timing and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.
-% \todo[inline]{how? GC-see above.}
-% KIE: We only list the potential sources of non-det, I therefore removed this. 
-% although monitoring and detecting such events falls outside of the scope for this work. 
-%
-%\noindent\underline{\textit{Ego Vehicle Controller:}}
-%The ego vehicle controller may be a source of non-determinism if using complex control algorithms such as neural networks for visual processing and perception.
-A neural network can only be deterministic within in the bounds of the training data used to construct it.
-If the network is exposed to unseen data it may incorrectly classify legitimate scenes due to lack of training data.\todo{That alone won't cause non-determinism, i.e.
-the NN will misclassify exactly in the same way each time - not?}
-%
-%In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically, assuming no \textit{residual lag}.
-In this sense, control over weather and lighting in the game engine should be considered and controlled as well as any other dynamic elements to the environment, e.g.
-reflections from surface water.
-% these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
-% i.e.
-the utilisation of hardware is manipulated differently each time.
-This would then result in non-deterministic behaviours especially if it is combined with the floating point problem discussed earlier.
-%
-% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
-Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available.
-This may lead to events happening out of order depending on the computational load and hence non-deterministic.
-%
-% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
-To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order.
-This is beyond the scope of our work and will not be covered in this paper.
-% \\\\
-% \noindent\underline{\textit{Physics and Rendering clocks:}}
-% \todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
-% % The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
-% Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
-This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
-% %
-% A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
-Do we not assume the same HW at this stage?
-GC- we are just listing sources at this point so should be hardware agnostic.
-also, yes, hardware is the same for us but we cannot control for resource utilisation and background processes.} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
-Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources.
-% %
-% AV specific platforms, such as CARLA, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
-% \todo[inline]{Add a summary/conclusion?}
-%\\\\
 
@@
 Any simulation hardware not using ECC memory that runs for 1000's of hours, typical in AV verification, is likely to incur significant CPU hours and is therefore subject to increased exposure to these errors.
-To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.\todo{Check this, please.}
-
-%\todo[inline]{That said, HW design simulation is deterministic, uses long simulation runs and thousands of CPU hours, but I've not heard that this is an issue.
-I can ask, though.
-;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.}
+To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.
 
@@
 %
-%
-% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
-% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
-% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
-% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
-the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
+For example, in Unreal Editor the \textit{unit}\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time.
 %
@@
 Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
-%.
-Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
 %
@@
 Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
-%
-% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
-%
-% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
-% \todo[inline]{unclear :( GC- re-written}
 
@@
 Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
-\todo[inline]{So, do commercial HPC facilities have ECC memory?}
-\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
-%
-\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
+%
 To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
-\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
 %
@@
 Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism.
-%
-
-%
-%The ego vehicle controller can be considered deterministic if dynamic elements to the simulation environment are controlled, such as weather and lighting that may be on random cycles. 
-%
-% ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software.
-Actor navigation not a problem if navigation meshes are fixed.
-Game engine, ensuring fixed $dt$ and control of random numbers.
-Core or thread scheduling will not be issue unless changing scheduling policy or migrating between machines that have different policies.
-Concurrency is really down to the development of the simulation software and reputable companies should have good control over these potential issues following good software development practice.
-NUMA should only affect interprocessor data transfer cost, at wrt only make computations longer.
-relative times between access different caches is likely to be small although may impact on high throughput, e.g.
-HPC.
-VFR - want to remove this section.
-EGO vehicle deterministic?
-Can't control library change, perception stack
 
@@
 %
-\todo[inline,color=green!40]{GC new: line below.}
-%
 Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
@@
 
-\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
-%
 We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-%
-\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
-%
-\todo[inline]{Complete the section structure as this is a very long section.}
 
@@
 \subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
 
@@
 The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-\todo[inline]{Decide past or present tense and be consistent throughout.}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
 
 \subsection{Tolerable Simulation Variance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
-
 To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
@@
 %
-% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
-%
 This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small as to fail with minor computational perturbations.
@@
 
-% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-%
-% but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-% 
-% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
-%
 For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
@@
 
-% considered to be sufficient for assertion checking in urban environments. 
-%
-% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
 Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-%
-\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
-where any value higher than the specified tolerance is considered non-permissible. }
 
@@
 % ***************************************************
-
 \subsection{Actor Collisions}\label{S:Actor_Collisions}
@@
 Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering an \textit{event hit} callback in the game engine (test 2).
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
-% \todo[inline]{AG: is this value correct or should it be 2.0m?
-AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
-actors would ignore trajectory waypoints that are less than 2.0m away from them.}
-intervals using a PID controller.
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m intervals using a PID controller.
 In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
@@
 
-\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
-
 % ***************************************************
@@
 \subsection{Experiment Description}\label{s:Experiment_Description}
-% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
 For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
@@
 Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-}
 %
@@
 The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation was reliable for verification purposes.
-% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
-% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
-\todo[inline,color=green!40]{GC Addition: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
-}
-%
-% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
-% \todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
-
-% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
-Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
-
-% This whole process is repeated for various internal and external settings (if needed). 
-% Then various plots can be created to draw the engine's simulation variance. 
-% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
 
@@
 \subsection{External Settings}
-% \todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
 
@@
 %
-% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
 Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0-95\% (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
 Resource utilisation figures reported here should be considered approximate values.
-% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
-
-\todo[inline,color=green!40]{GC Addition (para below):} 
-%Issues around floating point precision and concurrency as previously mentioned notwithstanding, 
+
 Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
@@
 \end{itemize}
-
-% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
-These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
-
-% \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
-
-% Our hypothesis is that since UE4 is a game engine it is non-deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
-This behaviour is expected not to hold as the level of computing stress increases.
-
- % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
-To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
-(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
-
-% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
-
-% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
-
-% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
-
-% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
-
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
 
@@
 However, all other scenarios involving vehicles or a mixture of actor types breach the tolerance that has been set, with some deviation in actor path as large as $59$cm.
-\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-}
 %
@@
 In this figure, anything above the $1$cm level (dashed line) is considered non-permissible according to our specific tolerance level informed from our verification requirements.
-%
-\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
-%
 
 A general pattern in the results indicates that; some scenarios consistently fail to produce results within permissible levels of variance at any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm level), some are consistently within tolerance (Test ID 5 \& 6 with pedestrians only), and that some cases only fail to meet the tolerance requirement when at higher utilisation levels, breaching the tolerance above 75\% resource utilisation (Test ID 1 \& 3).
-
-\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
 
@@
 This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-}
-% \todo{Why ``not reliable''.}
 
@@
 
-% \todo[inline]{Make sure that assertion checking has been introduced before.}
-
-% \textbf{Strange gap post-collision for few seconds}
-
-% \textbf{Should we rename 'test' to 'scenario???'}
-
-% \todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
-
-% Scenarios that include both collisions and vehicle actors were non-deterministic.
-However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
-
-% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
-Note the dotted red line in the plot shows the millimeter (mm) level.
-Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
-%
 \begin{figure*}[t]
@@
 
-\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
-This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
-Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
-
 The main findings of this work suggest a working practice that would minimise the non-deterministic effects observed in this investigation.
@@
 Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-%
-\todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-%For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm. 
-}
 
@@
 
-% First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
-with in the centimeter level).
-This clearly shows how the load on a computer (i.e.
-stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
-This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
-
-% Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
-with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
-
-% Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
-We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
-
-% Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
-Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
-
-% Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
-However, in V\&V of AV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
-Thus, the post collision effect should not be of much concern in the context of AVs.
-
-% It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
-The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
-Thus failing to stay below our red line and the test should be terminated.
-In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
-left alone and not used for any other tasks while the runs were executed.
-
 % ***************************************************
@@
 In addition a number of recommendations and best practices are suggested that other practitioners can follow to minimise simulation variance.
-% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
-% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}.
-%These need to be considered at different stages in the process.
-%
+
 \subsection{Design Experiment}\label{s:design_experiment}
@@
 \subsubsection{Trajectories and Collisions} 
-% Actor paths should be hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
 Actor paths, or the sequence of actions required to generate paths, should be hard-coded to ensure repeatability.
@@
 The verification engineer should set the \textit{tolerance} based on their specific requirements which refers to the acceptable level of actor path variance. 
-% ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation.
+%
 This tolerance level is analogous to the analogue signal level associated with each of the binary states %of the signals 
@@
 
-% \todo[inline]{Make sure assertion checking has previously been introduced.}
-
-% The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
-These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
-
-% \todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''. GC change callback to 'event hit'}
-
-% The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
-The aim of the experiment is to determine if the simulation is deterministic.
-This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories.
-Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.
-
-% \todo[inline]{We may need to introduce the Case Study first, then derive the general Methodology from it.
-This way, we can explain the requirements of the case study, how things were set up, what was observed, how the methodology was refined, several iterations even, etc.}
-
 \subsubsection{Simulation Length} 
@@
 In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-
-% Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
-This is to ensure that an outlier data-point is not hidden in the average of $n$-tests.
 
@@
 
-% \todo[inline]{Select section titles to match the flow diagram.
-In each of these sections, go through the matching phase in the flow diagram, explaining each point listed there and justify recommendations.}
-% The experiment should be a short\todo[inline]{What does short mean? Quantify.} scenario that can be easily repeated with the same initial conditions.
-
 \subsection{External Settings}
@@
 \subsubsection{Resource Utilisation}
-% To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
 The resources available to the simulator have been shown to have a significant effect on the simulation variance of simulated vehicles and exploring this as a variable the analysis is required. 
@@
 
-% \todo[inline]{Again, this sounds as if having the case study first makes more sense. That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
-
-% \todo[inline]{This should be ``External Settings'' - from flow diagram?}
-
 \subsubsection{Memory Testing}
@@
 \subsubsection{Sample Size}
-% Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} 
-% Histogram to show how sampling with position binning varies?
-% \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
 Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
@@
 It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the tolerance.
-% To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
 
@@
 Additional information can also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
-% such as the \texttt{unit} command in Unreal Engine. 
+%
 For the subsequent analysis, the actor should have an identifier and the repeat number and simulation time should also be captured.
-% The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
-For example, if 100 tests are executed and the maximum deviation in actor path is $<1$cm, then this scenario could be interpreted as being deterministic above this scale.
 
@@
 
-% Game engine actors can be used to simulate the AV and other vehicles as well as pedestrians.
-For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
-
-% \todo[inline]{This is mixing our experiments (which are part of the case study) with the methodology.
-It is probably best to first introduce the case study and then derive the methodology from it, rather than mixing both.} In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
-Initial testing also indicated that when actors collide there is a significant change in \textit{simulation variance}, therefore the experiment should include some scenarios where collision callbacks are activated.\todo[inline]{Make sure collision callbacks have been introduced earlier.}
-
-% Randomisation must be fully controlled by fixing the seeds for any random number generation that is used during the experiment.
-It is important to identify and control any external sources that may use randomisation, e.g.\ external vehicle controllers or other clients connected to the simulation environment.
-
-% \todo[inline]{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
-
-% \todo[inline]{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
-%
-% For the vehicles in the simulation, waypoints are generated and interpolated in the same way but execution at runtime is slightly different where a PID controller is used to generate a velocity profile and steering angle requests to reach the destination targets.}
-
-% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
-Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.
-A* search algorithm.
-
-% \subsection{Internal settings}\todo[inline]{Systematically go through the points listed in the flow diagram, explain and justify.}
-% Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
-
-% By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronous mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine.
-% Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
-However, in such applications of V\&V of AV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
-% If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the simulation variance as a result of allowing the AI to navigate in a more well defined space.
-% Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
-% The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.
-% Running in editor or deployed mode can sometimes cause differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
-
-% \begin{table*}[b]
-% \centering
-% \begin{tabular}{clclcc}
-% \toprule
-% Test ID & Test description (See Figures~\ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
-% 1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
-% 2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
-% 3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
-% 4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
-% 5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
-% 6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
-% 7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
-% 8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
-% \bottomrule
-% \end{tabular}
-% \caption{Set of experiments}
-% \label{TableOfExperiments}
-% \end{table*}
-
-% The user should also monitor and record system CPU and GPU utilisation rates.
-Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
-
-% Collision callbacks
-% \subsection{Other Considerations}
-% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
-ii) If not, how does its simulation variance vary?
-In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
-Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
-% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
-Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
-% The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
-For the majority of AV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
-
-% \subsection{External settings}
-% Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.
-Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
-Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.
-
-% \subsection{Running experiment}
-% \noindent When it comes to running the experiment itself, things that one should consider are the problem space exploration; the number of runs that should be executed to get reliable results; the frequency of logging data of actors; and possibly having a program to monitor the hardware utilisation.
-% Problem exploration, there is too many variables that one can play with. do statistical analysis some.... some of them get problematic...
-
 % ***************************************************
@@
 % ***************************************************
-% \section{Conclusions and Future Work}\label{s:conclusion}
 \section{Conclusions \& Future Work}\label{s:conclusion}
-
-% \todo[inline]{Say what the paper investigated and what we aimed to achieve, why this is important and how we went about it.
-Then point out the derivation of a re-usable methodology based on the knowledge and understanding we gained from the case study.}
-
-% \todo[inline]{GC: we need to say what we set out to achieve and why that is important, then say how our investigation led us to several theories which resulted in understanding several key parameters which directly affect simulation variance.
-From this we developed a methodology that the community can use to understand the deterministic operational domains of their own specific hardware:  Autonomous vehicle testing is adopting the use of game engines for development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-An empirical investigation uncovered several parameters that contributed to simulation variance ...
-explain results ...
-This led to the design of a methodology allowing 'users' to determine the deterministic operational domains for a specific hardware/platform configuration ...
-}
-
 The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
 %
-\todo[inline,color=red!40]{old:Having a deterministic simulator is required for reliable and repeatable results,}
-Having a deterministic simulator is required for precise results, 
-\todo[inline]{Having a deterministic simulator is required for precise results,} 
-%
-to find and fix software bugs and ensure results are trustworthy. 
+Having a deterministic simulator is required for precise results, to find and fix software bugs and ensure results are trustworthy. 
 %
@@
 
-% \todo[inline]{Review/rewrite this to match the earlier changes in the paper.
-Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
-Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy.
-A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the CARLA driving simulator.
-Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study.
-Furthermore it was found that all actors, even those not involved directly in a collision, will behave non-deterministically if a collision callback to the game engine is made.
-Best operating practice follows that the simulation should be stopped as soon as a collision callback is detected or resource utilisation levels exceed 75\% although this value may be system specific.
-}
-
-% \todo[inline]{GC: think we should put more emphasis on the outcomes of the method rather than the cast study.}
-% \todo[inline]{Point out (again) that all data and scripts etc are on github.}
-
-% \todo[inline]{Future work...}
-
-% In this paper we explored the determinisim of game engines.
-This is important for autonomous vehicles' development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, uncovered several parameters that contributed to simulation variance.
-These parameters include numerous internal settings based in Unreal Engine, and others that are external mainly dependant on CPU and GPU utilisation.
-
-% Changes in CPU and GPU resource utlisation has shown to be the main influencer for the non-deterministic behaviour of game engines.
-This is clearly demonstrated in the results (see section \ref{s:case-study}), where the max deviation between runs increased from the orders of $10^{-12}$m to devaitons of up to 1m as the CPU and GPU resource utilisation increased above 50\%.
-
-% The term \textit{deterministic operational domain} was given to the range of system resource utlisations that provides low simulation variances that is insignificant for development and verifcation of autonomous vehicles.
-% The deterministic operational domain specified herein is specific to our hardware configuration.
-
-% This led to the design of a methodology allowing `users' to determine the deterministic operational domain for a specific hardware configuration.
-This will help guide specialists in the field of AVs to setup their custom experiments for their specific simulators and identify their specific deterministic operational domains.
-
-% Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
-Hence allowing its extensive usgae by both the autonmous vehicles' and the gaming community.
-
-% presented an investigation (based on CARLA driving simulator) to show whether game engines are deterministic or not. 
-% Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, was used to illustrate the challenges of determinisim faced by game engines. 
-% % It was found that game engines are non-deterministic.
-
-% We reviewed possible sources of non-determinisim in section \ref{s:background} and investigated the effect of each on the simulation variance.
-
-% We found that the sources of non determinsim that had a significant effect on simulation variance were attributed to CPU and GPU resource utilisation. 
-% The case study clearly shows how the max simulation variance changes from very low simulation varicances, below picometer level ($10^-12$), to centimeter level when the CPU and GPU resource utilisation increases above 50\%.
-
-% We found two main outcomes from our investigation: 
-% 1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}, e.g. runtime scheduling. 
-% 2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.
-% Thus, by altering these settings one can achieve sufficiently low levels of simulation varince that can allow users to assume the simulation is deterministic. 
-% We phrased the level below which the simulation is to be assumed determinisitic as the \textit{determinisitic threshold}, which is up to the user to define where it exactly lies. 
-% This allows for one to use these game engines knowing that their simulation will be sufficiently deterministic provided that they bound their simulation within a certain configuration.
-
-% The configuration presented in here is specific to the hardware and software used in our investigation. 
-% Therefore, we outlined a methodology in this paper to allow practitioners in this field of AVs to resolve the non-deterministic operational domains of their simulation.
-
-% We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the simulation variance of a game engine used in V\&V application of AV simulations.
-The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
-
-% A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
-It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
-%since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
-
-% **** These results are not meant as a universal guide but merely to illustrate our method for a single hardware configuration.
-It is probable that the values quoted here will not translate to individual end-users unless the system configuration is very similar.
-This paper should be used as a method for users to determine what scenarios and system setting are appropriate for their particular verification needs.
-
-% You must monitor all collisions, not just for the ego vehicle as any collision callback may result in high simulation variance and hence non-deterministic results for your experiments.
-
-% In highlight of obtaining a fully deterministic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
-Alternatively, another approach would be to create a new engine that is designed to be deterministic with minimal attention given to performance.
-
-% ***************************************************
-%  FUNDING, REFS
-% ***************************************************
-% \section*{Acknowledgements}
-% This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
-% projects are part-funded by the Centre for Connected and Autonomous
-% Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
-% 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-\printbibliography
-
+\section*{Acknowledgement}
+\textit{Abanoub Ghobrial and Greg Chance contributed equally to this paper.}
+
+This research has in part been funded by the ROBOPILOT and CAPRI projects.
+Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
+
+\bibliography{cav_determinism.bib}
+
+% \end{multicols}
 \end{document}
+
+\grid

### 2021-02-15 paper/cav_determinisim_main.tex

@@
 
+\textit{Abanoub Ghobrial and Greg Chance contributed equally to this paper and are both corresponding authors.}
+%
+
 This research has in part been funded by the ROBOPILOT and CAPRI projects.
 Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-The Associate Editor for this paper was ....(\textit{Corresponding author: Abanoub Ghobrial.})
-
-A.Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
-G.Chance (e-mail: greg.chance@bristol.ac.uk), 
-K.McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
+% The Associate Editor for this paper was ....
+
+Abanoub Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
+Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
+Kevin McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
 and 
-K.Eder (e-mail: kerstin.eder@bristol.ac.uk) 
+Kerstin Eder (e-mail: kerstin.eder@bristol.ac.uk) 
 are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom.
 
-S.Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
+Severin Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
 and
-T.Pipe (e-mail: tony.pipe@brl.ac.uk), 
+Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
 are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.

### 2021-02-15 paper/cav_determinisim_main.tex
sorted out the floating point section

@@
 \usepackage{balance} %balance columns
-% \usepackage{todonotes}
+%\usepackage{todonotes}
 \usepackage[disable]{todonotes}
@@
 % \usepackage{draftwatermark}
+\usepackage{flushend}
 
@@
 %
-Thus, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
+Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
 %
@@
 %
-This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
-%
-Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
-variance between repeats becomes negligible for development and testing work.
+A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is then presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
+%
+Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
 
@@
 %
-CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website} a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
+CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
 
@@
 
+\section{Preliminaries}
+
 \subsection{Definitions}
 
-A list of definitions are given here which are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
+Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
 \\
@@
 
-A simulation can be thought of as the generation or production of experimental data. 
+A simulation can be thought of as the process of generating or producing experimental data. 
 %
@@
 %
-If a simulation is deterministic then there will be no variation in the generated output data, i.e. all future states are perfectly reproducible from prior states and actions. 
-%
-However, if a simulation is non-deterministic then there will be a variation in the generated output data. \\
+If a simulation is deterministic, Fig.~\ref{variance_description}~(b), then there will be no variation in the generated output data, i.e.
+all future states are perfectly reproducible from prior states and actions.
+%
+However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data. \\
 
@@
 
-We adopt some terms from the mechanical engineering and statistics domain that best describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
+We adopt terminology from the mechanical engineering and statistics domains to describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
 %
@@
 %
-Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken
+Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed under the same conditions and actions.
 %
@@
 % when comparing the differences between multiple products.
-As an analogy, the simulator can be thought of as a manufacturing process, producing data and to find the value of the precision then the output must be measured for differences when the process is repeated.
-Those differences describe the spread or variance in the process output and a hard limit on the variance can be defined, Fig.~\ref{variance_description}a, beyond which the output fails to meet the required tolerance, e.g.
-rejected for quality control.
+As an analogy, the simulator can be thought of as a manufacturing process that produces data.
+To determine the precision of this process, the output must be measured and analysed for differences when the process is repeated.
+%
+Those differences describe the spread or variance in the process output.
+A hard limit on the variance can then be defined, Fig.~\ref{variance_description}~(a), beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
 %
@@
 %
-Real manufacturing fails to achieve absolute precision and hence the need for tolerances on design specifications to account for the variance in real-world processes.
-
-If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}b, and hence will be within acceptable tolerance.
-If the simulator is non-deterministic then there will be a variance in the output which can be measured.\\
+Real manufacturing fails to achieve absolute precision. Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.
+
+If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within acceptable tolerance by design.
+If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
 %
@@
 \subsubsection{Accuracy}
-Precision and tolerance should not to be confused with \textit{accuracy} which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
-We therefore define accuracy as the difference between the true value, or reference value, and what has been achieved in the simulation or generation process.
+Precision and tolerance should not to be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
+We therefore define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation.
 %
@@
 may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
-In most cases such accuracy will not be possible due to computational demands of such an exact replica and in most cases is unnecessary, where some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
+%
+In practice full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
+In most cases, in fact, it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
 \\
@@
 \subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) at regular time intervals.
+A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) in a 3D environment recorded at regular time intervals.
 This definition could be extended to include other variables.
 %, but position is sufficient for this analysis.\\
-A set of simulation traces derived from the same input then forms the experimental data on which variance is calculated. \\
+A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run. \\
 
 \subsubsection{Simulation Variance \& Deviation}
-If the simulator is non-deterministic then how can this variance be measured? This can be achieved by monitoring any simulated output variables that should be consistent from run to run. 
+If the simulator is non-deterministic then how can the simulation variance be measured?
+This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run.
 % In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
-Actor path variance, derived from the simulation trace, is chosen over other variables as this distance based metric forms the basis for many downstream verification tests.% such as assertion checking.
-% 
-Therefore the term \textit{simulation variance} refers to a measure of actor position variance in the simulation with respect to time assuming fixed actions. 
-%
-\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive value to comprehend when considering interpretation of verification tests. 
-%Results are referred to in terms of deviation, the square root of variance, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
+For example, actor position variance is a distance-based metric that can be derived from simulation traces.
+The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained.
+% 
+Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor position variance in the simulation with respect to time, assuming fixed actions.
+{\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
+%
+%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
+%
+Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
+%
+% more intuitive to comprehend when considering verification tests, e.g.
 vehicle close-passing.
@@
 \subsubsection{Scene, Scenario \& Situation}
-We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
-Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals.
+We adopt the terminology defined for automated driving in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
+Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV, or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals.
 The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.
@@
 \subsection{When is Determinism needed?}
-Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-A deterministic simulator can be considered to have zero \textit{variance}.
-If the variance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulation traces.
+Determinism is a key requirement for simulation during AV development and testing.
+A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
+%
+Thus, a deterministic simulator has zero \textit{variance}.
+
+A simulator with non-zero variance can no longer be considered deterministic. 
+%
+Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}. 
+%
+Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations.
 When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+%
 This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
@@
 
-If the simulation is non-deterministic and has a non-permissible variance in, for example, actor positions, this may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
+If the simulation is non-deterministic, e.g.\ 
+% KIE No, whether the variance is or is not permissible is not what matters for non-det, it should be zero!
+%e.g. it has a non-permissible variance in,
+it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to re-produce, understand and remove bugs and rendering verification results unstable.
+In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
 %
@@
 % Challenges of SBV
-When used for gaming, game engines do not need to be deterministic nor do they even have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs related to non-determinism a high priority for games developers.
+When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs % related to non-determinism
+a high priority for games developers.
 It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
 However, the situation is very different for AV development and testing.
-Thus, our main research question is:
+Thus, our main research questions are:
 %
@@
 %
-How can one assess the extent to which a simulation environment is deterministic or has a permissible level of variance?
+{\em How can one assess whether a simulation environment is deterministic?} and 
+{\em How can one determine and control the simulation variance?}
 \todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
@@
 % In pursuit of an answer to this question we
-In this paper we investigated how the non-determinism of Carla, based on the Unreal game engine, when used for simulation-based AV verification affects the simulation results. 
-%
-In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation output by analysis of actor position variance.
+In this paper we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
+%
+In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
+% KIE: This is covered in the next sentence:
+% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
 %
@@
 %
-By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
+% By analysing actor position variance 
+We find that the CARLA simulator is non-deterministic under certain conditions. {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
 % However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
 %
-Carla exhibits a permissible level of variance when system utilisation is restricted to 75\% or less and ensuring termination of scenarios once a vehicle collision has been detected.
+CARLA exhibits a permissible level of variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
+{\color{red} Why ``vehicle'' collision and not just ``collision''?}
 %
@@
 %
-Knowing the simulation variance will help assess the impact that using a game engine for AV simulation may have on verification tasks.
+Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
+%impact that using a game engine for AV simulation may have.
+% on verification tasks. 
 In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact on verification results.
@@
 
-There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g.
-Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
+There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
 % \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
@@
 %
-Ensuring software operates deterministically is a non-trivial task and catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
+Ensuring software operates deterministically is a non-trivial task.
+Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
 %
@@
 % Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
-In exceptional cases where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines,
+In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines,
 and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
@@
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-The gamer neither wishes a faithful representation of reality, i.e.
-the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance, i.e.
-can be low \textit{precision}.
-%
-In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
+The gamer neither needs a faithful representation of reality, i.e.\ the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance.
+%, i.e.\ can be low \textit{precision}.
+%
+In contrast, the precision required for AV development and testing is very high, especially for perception stack testing, but also, more generally, for obtaining reliable verification results and for visual inspection of accidents or near misses.
+If necessary, it is acceptable to achieve the required tolerance %on the precision
+at the cost of real-time performance.
 For example, for perception stack testing the sensors need to get input that is repeatable so that if any software bugs are found they can be re-played and the issue resolved.
@@
 %
-A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests} although the AV simulation domain introduces it's own unique challenges that were not considered in that paper.
+A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests}, although the AV simulation domain introduces it's own unique challenges that were not considered in that paper.
 
@@
 %
-%
-% SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
-Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
+
+% SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
+{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
+and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }
+
+% As a consequence, floating-point arithmetic is not associative.
+Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, a change of execution order, e.g.\ from $x+(y+z)$ to $(x+y)+z$, can lead to different results~\cite{Kapre2007}.
+%
+As a consequence, floating-point arithmetic is not associative.
+Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}.
+This is because rounding is being performed at different stages in the computation, depending on the respective execution order.
+%
+Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
+%
+Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}. 
+%
+%{\color{red} Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.}
 %
@@
 
-Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
-Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-%
-If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
-% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.}
+% Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
+Calculation results may differ due to the nature of non-associative floating-point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
+Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
+%
+% If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
+% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.
 }
@@
 %
-Thus, one could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.
-
-However, these precision issues are better described as \textit{incorrectness}, rather than them being a source of non-determinism. 
-%
-It is normally reasonable to assume that the processors on which game engines run are designed to be deterministic with respect to floating-point arithmetic. 
-%
-Thus, when a rounding error occurs, then the same processor should produce the same \textit{incorrect} result given the same operands. 
-%
-So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always 
-be equally \textit{incorrect} as long as the implementation conforms to the IEEE floating-point standard~\cite{8766229}.
-
-Beyond hardware limitations, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
+Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability.
+In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.}
+
+In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
+%
+It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
+%
+Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern.
+%, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
+%
+So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
+
+However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating point operations producing different results.
+Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
+
+Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
 However, the same executable should still return the same output for identical input.
-%
-Therefore, it can be concluded that floating-point arithmetic does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware with the same configuration.
+
+In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.
 
@@
 the utilisation of hardware is manipulated differently each time.
-This would then result in non-deterministic behaviours especially if it is combined with the floating point problem discussed earlier.
+This would then result in non-deterministic behaviours especially if it is combined with the floating-point problem discussed earlier.
 %
@@
 \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
+Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
 % \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
@@
 \todo[inline,color=green!40]{GC Addition (para below):} 
-%Issues around floating point precision and concurrency as previously mentioned notwithstanding, 
+%Issues around floating-point precision and concurrency as previously mentioned notwithstanding, 
 Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
@@
 Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
-Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy.
+Potential sources of non-deterministic simulation behaviour have been discussed such as floating-point rounding error and scheduling policy.
 A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the CARLA driving simulator.

### 2021-02-15 paper/cav_determinisim_main.tex
got to just before System Configuration and Screening

@@
 Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability.
-In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.}
+In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
 
@@
 %
-Changing the scheduling policy and thread priorities may increase simulation variance.
+In principle, changing the scheduling policy and thread priorities may increase simulation variance.
 
@@
 Indeed, the developers of the de-bugging program \texttt{RR}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
+This allowed control over the scheduling and execution order of threads promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
 %
@@
 \subsection{Game Engine Setup}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
+The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps, ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
 %
@@
 % \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
+For example, in Unreal Editor the \textit{unit}~\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
 % to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
@@
 %
-However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment and cannot be unknowingly modified and therefore can be considered fixed.
+However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment.
+These cannot be unknowingly modified and therefore can be considered fixed.
 
@@
 %
-\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
-To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
+\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
+To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
 \todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
 %
-NUMA should only affect interprocessor data transfer cost and without control measures will only make the computation cycle longer.
+NUMA should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
 Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
+{\color{red} Not sure this is correct, see the NUMA section above.
+Why ``only''?
+If one computation out of many takes longer, then their exec order can be affected, or not?}
 %
@@
 \end{tabular}
-\caption{A description of the test scenarios showing the test number, the actors included, if a collision occurred and between which actors.
-Where $n$ is the number of repeats and $\max\sigma$ is the maximum simulation deviation.
+\caption{A description of the test scenarios showing the test number, the actors included, whether a collision occurred and if so then between which actors.
+$n$, the number of repeats is set to 1000 and $\max\sigma$ is the maximum simulation deviation.
 The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation.
@@
 %
-Thus, the scene for our investigation is the T-junction (T-intersection) and the scenarios are shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor trajectories over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
+Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
+
+This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
 %
@@
 %
-This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small as to fail with minor computational perturbations.
+The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small as to fail with minor computational perturbations.
+{\color{red} {\sc What would fail?}}
 %
@@
 % due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study and more broadly our general verification requirements, a tolerance of $1$cm has been selected. 
+Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected. 
 % Thus any variation of less than $1$cm is tolerable. 
@@
 % \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
-Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
+In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
 %
@@
 \subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}.
-%
-Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.
-
-The tests are listed in Table~\ref{TableOfExperiments} and were chosen to cover a range of interactions between actor types. 
-%
-Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering an \textit{event hit} callback in the game engine (test 2).
+Previous investigations into the Unreal Engine indicated that collisions between dynamic actors {\color{red} Why DYNAMIC?} and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}.
+{\color{red} NOT ACCESSIBLE}
+%
+Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were identified as potentially key aspects to the investigation into simulation variance.
+
+The tests used for the case study are listed in Table~\ref{TableOfExperiments}. They cover a range of interactions between actor types. 
+%
+Tests 1 \& 2 involve two vehicles meeting at a junction where they either do not collide (Test 1) and where they do collide (Test 2), thereby triggering an \textit{event hit} callback in the game engine.
 In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
@@
 intervals using a PID controller. 
-In test 3 a mixture of different actor types are introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
-In test 4 this pedestrian collides with one of the vehicles at the crossing, triggering a \textit{event hit} callback, see Fig~\ref{Test_a}.
+%
+In Test 3 a mixture of different actor types is introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point. 
+%
+In Test 4 this pedestrian collides with one of the vehicles at the crossing, triggering an \textit{event hit} callback, see Fig~\ref{Test_a}.
 Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
-Tests 5 \& 6 involve only pedestrians that do not collide (test 5) and that do collide (test 6), see Fig.~\ref{Test_b}.
+Tests 5 \& 6 involve only pedestrians that either do not collide (Test 5) and that do collide (Test 6), see Fig.~\ref{Test_b}.
 
@@
 % ***************************************************
-\subsection{Experiment Description}\label{s:Experiment_Description}
+\subsection{Evaluation Metric}\label{s:Experiment_Description}
 % \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
+For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time. 
+%
+The logs from repeated tests are compared at the same time index $t$ for each actor $a$ to provide a value for the variance of actor $a$ at time index $t$, $\sigma_a^{2}(t)$, at each point in time.
+{\color{red} This gives a variance function over time for each actor.}
+
+{\color{blue} The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.}
+
+Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
+%
+\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
+%
+Repeated traces are compared at the same time index $t$ for each actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
 %
 Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-}
-%
-The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, 
+}
+
+The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, as given in Equation~\ref{eq:max_sigma}.
+{\color{red} This is still a function over time - should it be?}
 \begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a^{2}(t), 
+\max_{a,t}\sigma_a^{2}(t)
 \end{equation}
-and therefore the maximum deviation is simply square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity.
+The maximum deviation is the square root of the maximum variance {\color{red} The absolute value of the square root?} and herein referred to as ${\max\sigma}$ for brevity.
+{\color{red} So, ``a'' is any actor or a specific actor?
+Likewise, ``t'' is any time point or a specific time point?
+NOTE that any actor any time means a in A and t in T, where A is the set of actors and T is the set of all time points.}
 
@@
 Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
-Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.
-\textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.
-\textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
+Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.\ \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.\ \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
 Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
@@
 Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.
+{\color{red} NOT ACCESSIBLE}
 
@@
 % \textit{deterministic threshold}, 
-\textit{tolerance}, 
+\textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED?}
 settings external to the game engine were explored. 
@@
 This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the CARLA platform.
+{\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented.
+Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}
 
@@
 % In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
-Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0-95\% (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
+Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
 Resource utilisation figures reported here should be considered approximate values.
@@
 For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
-In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision~\cite{matlab_parallel_computing}.
+In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency-induced imprecision~\cite{matlab_parallel_computing}.
 

### 2021-02-17 paper/cav_determinisim_main.tex
half-way house, not fully completed and some parts half touched

@@
 In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
+{\color{red} Should this be IMPRECISION rather than incorrectness?}
 %
@@
 
-\subsection{NUMA Non-Uniform Memory Access}
+\subsection{Non-Uniform Memory Access (NUMA)}
 For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
@@
 %
-NUMA should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
+Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
 Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
@@
 We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
+{\color{red} and then..?
+Please cover the entire section - consider splitting setup from evaluation.}
 %
@@
 
+\subsection{Initial Results and Discussion}\label{ResultsSection}
+
+{\color{red} This is for unrestricted, so this does not come as a surprise later.}
+
 % ***************************************************
@@
 % \textit{deterministic threshold}, 
-\textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED?}
+\textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
 settings external to the game engine were explored. 
@@
 % The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
+%
+{\color{red} Somwhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
 
@@
 % ***************************************************
-\subsection{System Configuration and Screening}\label{s:screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
-%
-Operating systems was Linux Ubuntu 18.04.4 LTS. To ensure reliability of results, each test was repeated 1000 times. 
+\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
+The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
+%
+The operating systems was Linux Ubuntu 18.04.4 LTS. 
 %
@@
 %
-A detailed guide for reproducing these experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
+Each test was repeated 1000 times. 
+%
+A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
 %
@@
     \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}\cite{shi2009testing}.
-    \item Non-uniform memory access: \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
+    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
+    \item Non-uniform memory access (NUMA): \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed.
+Therefore this was not used for subsequent testing.
+{\color{red} But if we aim to min the sources of non-det, then that should be considered/used?}
 \end{itemize}
@@
     \end{subfigure}
-    \caption{Vehicle to vehicle collision (test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
+    \caption{Vehicle to vehicle collision (Test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
 The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
@@
 % ***************************************************
-\subsection{Results and Discussion}\label{ResultsSection}
-A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted), where the value reported is the maximum deviation across all resource utilisation levels, i.e.
-the worst case for a given scenario.
-%
-From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the tolerance that has been set, with some deviation in actor path as large as $59$cm.
+\subsection{Final Results and Discussion}\label{ResultsSection}
+
+{\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}
+
+A summary of all the results is shown in Table~\ref{TableOfExperiments}.
+In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
+%
+From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+%
+However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
+%
 \todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
@@
 %
-Clearly a deviation in results of such a large amount is unacceptable if simulation is to be used as a reliable verification tool.
-
-Resource utilisation level was found to have a significant impact on \textit{simulation variance}.
+Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
+
+Resource utilisation was found to have a significant impact on \textit{simulation variance}.
 %
 Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, anything above the $1$cm level (dashed line) is considered non-permissible according to our specific tolerance level informed from our verification requirements.
+In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
+%
+Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
 %
@@
 
-A general pattern in the results indicates that; some scenarios consistently fail to produce results within permissible levels of variance at any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm level), some are consistently within tolerance (Test ID 5 \& 6 with pedestrians only), and that some cases only fail to meet the tolerance requirement when at higher utilisation levels, breaching the tolerance above 75\% resource utilisation (Test ID 1 \& 3).
-
-\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
-
-However, the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-
-Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
+{\color{red} ???
+about 85\% based on the figure - or not???}
+
+Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision.
+%
+The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\% {\color{red} Why 75\%?}, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
 Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
+This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
+
 \todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
@@
     \end{subfigure}
-    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
+    \caption{Vehicle to pedestrian collision (Test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
 \end{figure*}
@@
     \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
-    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% additonal CPU \& GPU resource utilisation.}
+    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
     \label{NICEExperimentStressSummary}
@@
 The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
+In this scenario (Test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
+There are three observations; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
 % as described above. 
@@
 Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous assertion checking.
+This final point should be of particular concern to verification engineers, developers and researchers in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous simulation results.
+% not just assertion checking
 
@@
     \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v5.pdf}
-    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}
+    \caption{Stages of the methodology proposed to determine the variance of a simulation.}
     \label{method_diagram}
@@
 
-The main findings of this work suggest a working practice that would minimise the non-deterministic effects observed in this investigation.
-By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table.~\ref{TableOfExperiments}.
+To conclude, the main findings of this case study suggest a working practice that would minimise the factors that give rise to the non-deterministic effects observed in this investigation. 
+%
+By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table~\ref{TableOfExperiments}.
 By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve.
@@
 \subsection{Process Scheduling Priority}
-An investigation into the response of process scheduling on the simulation variance was undertaken.
-The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
+An investigation into the impact of process scheduling on the simulation variance was undertaken {\color{red} say why}.
+The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
 %
@@
 
-The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range, non-outlier limits by the whiskers and a horizontal bar for the median.
-The figure shows that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a NICE value of 0 (central plot bars).
-Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation but this did not account for all the difference in the observed results.
-This is demonstrated in the maximum priority setting where the bars in plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
-%
-This unaccounted difference in the variance may be due to the lack of absolute control that \texttt{NICE} has over the process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
-
-\subsection{Investigation Summary} \label{s:empirical_summary}
+The results are presented in Fig.~\ref{NICEExperimentStressSummary} {\color{red} What is on the y axis?} where the box denotes the inter-quartile range {\color{red} of???}, non-outlier limits by the whiskers and a horizontal bar for the median.
+The figure shows that decreasing the priority {\color{red} of the simulator process} (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot).
+Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results.
+This can be seen in the maximum priority setting where the bars in the plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
+%
+The remaining difference in the variance between the two resource utilisation levels may be due to the lack of absolute control that \texttt{NICE} has over process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
+
+\subsection{Investigation Summary} \label{s:empirical_summary} {\color{red} Factor into overall conclusion of the case study}
 
@@
 
-\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
+%\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
 The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator.
@@
 However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly.
-Therefore we now present a general methodology that practitioners can follow to find the
+%
+Therefore we have derived a general methodology that practitioners can follow to find the 
 % \textit{deterministic domains of operation} 
 \textit{operational domains of permissible variance} 
-for any hardware configuration.
+for %any hardware configuration.
+a game-engine-based simulation environment. This methodology is presented in the next section.
 
@@
 % ***************************************************
-\section{Methodology} \label{s:methodology}
-In this section a method for determining the variance of simulated actor path trajectories, termed \textit{simulation variance}, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.
-%
-In addition a number of recommendations and best practices are suggested that other practitioners can follow to minimise simulation variance. 
+\section{Methodology {\color{red} for what???}} \label{s:methodology}
+In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
+This method can be used to determine the operational domains of permissible variance of a simulation environment.
+%
+{\color{red} MAKES NO SENSE: {\sc In this section a method for determining the variance of simulated actor path trajectories (KIE: a trajectory is a path), termed simulation variance, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.}}
+%
+In addition, recommendations and best practice guidelines for minimising simulation variance are suggested. 
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
@@
 %
-\subsection{Design Experiment}\label{s:design_experiment}
+
+The method consists of a sequence of five stages; experimental design, simulator configuration, external configuration, execution and analysis.
+In the following, each stage is described in detail with reference to the items listed for each stage in Fig.~\ref{method_diagram}.
+{\color{red} NOTE: renamed stages.}
+
+\subsection{Experimental Design}\label{s:design_experiment}
 

### 2021-03-04 paper/cav_determinisim_main.tex
reviewed/revised/commented throughout

@@
 
-Indeed, the developers of the de-bugging program \texttt{RR}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
 %
-\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{rr} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
@@
 
-\subsection{Tolerable Simulation Variance}
+\subsection{Tolerable Simulation Variance}\label{s:tolerance}
 
@@
 
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
+Instead of using ..., {\color{red} \sc please complete} herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
 %
@@
 
-The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, as given in Equation~\ref{eq:max_sigma}.
-{\color{red} This is still a function over time - should it be?}
+The maximum variance over the entire set of $n$ repeated simulations is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
+%{\color{red} This is still a function over time - should it be?}
 \begin{equation} \label{eq:max_sigma}
@@
 
-The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation was reliable for verification purposes.
+The maximum deviation, $\max\sigma$, can be analysed for the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation is sufficiently accurate for verification purposes.
 % The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
@@
 %
-Each test was repeated 1000 times. 
+Each test was repeated 1000 times. {\color{red} Justify why 1000 - see my later comment in the Method section.}
 %
@@
     \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
-    \item Non-uniform memory access (NUMA): \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed.
+    \item Non-uniform memory access (NUMA): \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores, but only a minor (2\%) improvement in simulation variance was observed.
 Therefore this was not used for subsequent testing.
@@
     \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v5.pdf}
-    \caption{Stages of the methodology proposed to determine the variance of a simulation.}
+    \caption{Stages of the method proposed to determine the variance of a simulation. {\color{red} \sc please update to match text}}
     \label{method_diagram}
@@
 % ***************************************************
-\section{Methodology {\color{red} for what???}} \label{s:methodology}
+\section{Method to determine the Variance of a Simulation{\color{red} for what???}} \label{s:methodology}
 In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
@@
 %
-In addition, recommendations and best practice guidelines for minimising simulation variance are suggested. 
+In addition, recommendations and best practice guidelines for minimising simulation variance are suggested. {\color{red} \sc Highlight these in the text.}
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
@@
 
+{\color{red} \sc Surely, a series of systematically designed experiments, each repeated X times, should be set up, where each experiment varies one of the suspected sources of non-determinism?
+Then, the analysis of the data from each experiment should reveal the actual sources of non-determinism for the given setting?
+In other words, we are aiming to fix as much as we can control (easily), and then we intentionally vary (ideally) single variables, e.g.\ the resource utilisation, to find out what makes a difference.
+We need a clear definition of the objectives of the experimental design, i.e.\ what one should vary and what one wants to ``nail down''.}
+
 \subsubsection{Actors} \label{s:actors}
 
-Different actor types may be handled differently by the game engine and our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
-However the introduction of CARLA vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
-As such, all actors that could be included in any future simulation should be tested, including any non-standard CARLA or bespoke actors.
+Different actor types may be handled differently by the game engine.
+For instance, our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
+However, the introduction of CARLA vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
+{\color{red} \sc This needs rephrasing as it appears to contradict the statement above.} As such, all actors that could be included in any future simulation should be tested, including any non-standard CARLA or bespoke actors.
+{\color{red} \sc ...
+such as the ego vehicle...?
+[see section on ``External Settings'' last point - Ego Controller.]}
 %
@@
 Actor paths without collision are also important to include as these will serve as a baseline to the other tests.
-
-\subsubsection{Setting a Tolerance} \label{s:threshold}
-The verification engineer should set the \textit{tolerance} based on their specific requirements which refers to the acceptable level of actor path variance.
+{\color{red} \sc What baseline?}
+
+\subsubsection{Permissible Variance} \label{s:threshold}
+The verification engineer should set the \textit{permissible variance}, also termed \textit{tolerance}.
+% which refers to the acceptable level of actor path variance. 
 % ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. 
-This tolerance level is analogous to the analogue signal level associated with each of the binary states %of the signals 
-in digital circuits.
+%
+The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection.
+In Section~\ref{s:tolerance} a tolerance of $1$cm was considered sufficient for urban scenario assertion checking.
+In practice, it may be necessary to determine this tolerance experimentally.
+
+%The permissible variance is analogous to the analogue signal level associated with each of the binary states %of the signals in digital circuits.
 At some low analogue voltage the signal is interpreted as binary $0$, this may be, for example, any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
 This convention enables the abstraction from the noisy physical signal values that are analogue voltages into the digital domain with binary representation.
-%
-The tolerance that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this tolerance must be sufficiently small to enable accurate assertion checking.
-In our empirical investigation we set a tolerance of $1$cm which is sufficient for urban scenario assertion checking.
-In practice, it may be necessary to determine this tolerance experimentally.
+<- KIE: meant only to illustrate that other areas also have defined tolerances.
+%
 
@@
 
-\subsubsection{Simulation Length} 
-The simulation time should be sufficient to record an interaction between actors but not so long that the testing take an inconvenient time to complete.
+\subsubsection{Simulation Time} {\color{red} \sc Should this not be part of the Execution section?}
+The simulation time should be sufficient to record interactions between actors but not so long that the testing take an inconvenient time to complete.
 In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+{\color{red} \sc Point to the section that specified that.}
 
@@
 The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
-If using CARLA a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$~\cite{MonoBehaviour_unity}.
+If using CARLA, a small fixed value, say $0.05s$, can be set by using \texttt{setting.fixed\_delta\_seconds = 0.05}.
+In Unity, the default fixed time step is set to $0.02s$~\cite{MonoBehaviour_unity}.
+
 In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
-%
-The use of random numbers must be controlled through the use of fixed seeds which might be used to control variations of background effects, e.g.
-weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment.
+
+The use of random numbers must be controlled through fixed seeds, resulting in pseudo-randomness.
+Random numbers might be used in the simulation environment to control variations of background effects, e.g.\ weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment.
 %
 Actors that navigate through the environment should use a fixed navigation mesh. 
-
-The version number of the CARLA and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}, so ensuring consistent version number throughout testing is also important.
+%
+The version number of the CARLA and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}.
+Therefore, ensuring a consistent version number throughout testing is also important.
 
@@
 % To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-The resources available to the simulator have been shown to have a significant effect on the simulation variance of simulated vehicles and exploring this as a variable the analysis is required. 
-CPU utilisation software, such as the linux \texttt{stress} tool which is a workload generator program can be used spawn workers on any number of cores or virtual threads on a system.
-This can be used to artificially increased the load on the system to explore at what point your system may become susceptible to \textit{simulation variance}.
-For GPU utilisation \texttt{gpu-burn} was employed using the \texttt{fur test} where different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
-Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU respectively.
-These values can also be written into the data logging for completeness.
-
-Alternatively, in place of artificial resource utilisation multiple instances of the simulation could be executed simultaneously. However, the granularity of control with this approach may be reduced.
+The resources available to the simulator have been shown to have a significant effect on the variance of the path of simulated vehicles. %and exploring this as a variable the analysis is required. 
+%
+Thus, it is important to understand at what level of resource utilisation the system running the simulation becomes susceptible to \textit{simulation variance}.
+
+CPU utilisation software, such as the linux \texttt{stress} tool, which is a workload generator program, can be used to spawn workers on any number of cores or virtual threads on a system.
+This can be used to artificially increase the load on the system.
+For GPU utilisation, \texttt{gpu-burn} can be employed using the \texttt{fur test}.
+Different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
+Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU, respectively.
+These values should be added to the data logs.
+%
+Alternatively, in place of artificial resource utilisation, multiple instances of the simulation could be executed simultaneously.
+However, the granularity of control with this approach may be reduced.
 
@@
 \subsubsection{Memory Testing}
-Prior to experimental execution the system hardware should be tested for memory conformity and to ensure no single bit errors are occurring, see~\ref{s:screening}.
+Prior to experimental execution the system hardware should be tested for memory conformity and to ensure no single bit errors are occurring, see Section~\ref{s:screening}.
 For mainboard memory \texttt{memtest86}\ can be used on most platforms to run a series of pre-defined memory test patterns.
@@
 We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
-However, fine control over the scheduling policy and thread execution order may be non-trivial.
-Such policies are defined by the operating system and may execute threads with equal priority.
-In such a case, all tasks non-essential to the simulator should be terminated to prevent interruption to the simulator.
+However, gaining fine control over the scheduling policy and thread execution order is known to not be trivial~\cite{acm-q-rr-interview}.
+
+The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority.
+Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation.
 
 % If the policy or operating system cannot be changed then s
-Setting the simulator process to a higher priority may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
+Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
 
 \subsubsection{NUMA Control}
-Control over Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multiprocessors with shared memory.
-This control allows the simulator program to be fixed on a single core reducing memory access time.
-%
-Initial screening tests done with NUMA control gave moderate improvements in simulation variance of the order of a few percent, see Section~\ref{s:screening}.
-This control may assist if simulation variance is borderline to the tolerance but not seen as essential.
-
-\subsubsection{Ego Vehicle Controller}
-No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
+Control over a Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multiprocessors with shared memory.
+This control allows the simulator program to be fixed on a single core, reducing and unifying memory access time.
+%
+Note that initial screening tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}. 
+%
+This control may assist if simulation variance is borderline to the tolerance but not seen as essential. {\color{red} \sc Unclear: what is not essential?}
+
+\subsubsection{Ego Vehicle Controller {\color{red} \sc Why is this not covered under actors? [see my note in the Actors section]}}
+{\color{red} \sc Clearly separate the investigation from the Method.
+The fact that the case study did not have an ego vehicle should be mentioned as a limitation of the case study, rather than elaborated on here, but why does this make a difference?}
+The impact of the ego vehicle  should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
+{\color{red} \sc Long and cumbersome sentence with strange grammar: No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
+}
 
@@
 % \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
-Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
-While executing 100 repeats may seem sufficient, this sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
-It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the tolerance.
+%
+Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
+While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results.
+{\color{red} \sc Again, separate the study from the method.
+This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.} It is therefore recommended that the sample size is determined empirically dependent on the number of results that exceed the tolerance.
+{\color{red} \sc How would one know?}
 % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
@@
 \subsubsection{Data Logging}
-Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-Additional information can also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
+Unique identifiers should be assigned to each experiment, each repeat and each individual actor. 
+The time-stamped actor positions should then be recorded at fixed time intervals throughout the simulation in order to determine the variance in actor path.
+Additional information should also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
 % such as the \texttt{unit} command in Unreal Engine. 
-For the subsequent analysis, the actor should have an identifier and the repeat number and simulation time should also be captured. 
+%For the subsequent analysis, each actor should have an identifier. The repeat number and simulation time should also be captured. 
 % The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
@@
 \subsection{Analysis}
-The maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control for permissible operation.
-Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
-
-This method can be extended to a broader actor states and actions including actor orientation, speed, and any other status indicators that may be of interest and also including appropriate actions that may be useful for verification purposes.
+For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment.
+%Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
+% KIE: Why now point to HW? And, which "boundaries"?
+
+{\color{red} \sc Unclear how this relates to Analysis.
+Is this a summary of the method?
+If so, it should be moved to the front or to the conclusion.} This method can be extended to a broader actor states and actions including actor orientation, speed, and any other status indicators that may be of interest and also including appropriate actions that may be useful for verification purposes.
 This broader As such, Equation~\ref{eq:max_sigma} would need to be adjusted to include these new variables.
+{\color{red} \sc Something is missing in this sentence - please correct before moving to a different place, or removing.}
 
@@
 
-The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
-%
-\todo[inline,color=red!40]{old:Having a deterministic simulator is required for reliable and repeatable results,}
-Having a deterministic simulator is required for precise results, 
-\todo[inline]{Having a deterministic simulator is required for precise results,} 
-%
-to find and fix software bugs and ensure results are trustworthy. 
-%
-If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains for reliable results.
-During an investigation into the CARLA simulator, a \textit{simulation variance} was observed for repeated tests with the same initial conditions indicating non-deterministic execution.
-%
-During the study we hypothesized and uncovered several parameters that contribute towards greater simulation variance and hence non-deterministic execution, such as actor collisions and system resource utilisation.
-%
-The results of the investigation are hardware and software version specific, so we proposed a general methodology that can be used to find the \textit{domains of permissible variance} for any given system configuration.
-%
-This methodology will allow the AV verification community using simulation tools to ensure their results are reliable and trustworthy.
-
+%The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
+Game engines offer simulation environments that are used for the development and verification of autonomous driving functions. 
+%
+Determinism of a simulator is required to achieve repeatability, which is essential to find and fix software bugs efficiently, and to ensure simulation results are trustworthy. 
+%
+If a simulator is non-deterministic then practitioners should at least be aware of, and know how to find, the operational domains where \textit{simulation variance} is tolerable.
+
+An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation.
+%
+We then researched, identified and discussed potential sources for non-determinism in this context.
+%
+A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution.
+In particular, actor collisions and system-level resource utilisation were identified as key contributors.
+%
+{\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.
+This seems to have been commented out but was here before.}
+
+A general method to assess the actor path variance of a game-engine-based simulation environment is then proposed.
+The method can be used to find the \textit{domains of permissible variance} of a simulation environment for a given system configuration.
+This can give AV developers and verification engineers increased confidence in the simulation results and reduce debug time.
+% 
 % \subsection{Future Work}
-A potential avenue of exploration is developing a version of the simulator which has been designed specifically for verification.
-In this version the simulator would be deterministic because the execution is suitably controlled along with sources of randomness and scheduling.
-A similar task was achieved with the \texttt{RR} program~\cite{RR_link}.
-
-With the advent of more AI systems becoming adaptive, meaning that the system may change it's response to the same stimulus over time, the notion of repeatability will need reassessing in the context of verification.
+%
+As future work the method can be extended to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
+
+An ambitious avenue for future work is the development of a deterministic simulator for AV development and verification.
+This requires controlling all potential sources of non-determinism, including randomness and scheduling, very similar to the development of the record-and-reply debugger \texttt{rr}~\cite{RR_link}, originally developed to catch low-frequency non-deterministically failing tests at Mozilla~\cite{acm-q-rr-interview}.
+
+%With the advent of more AI systems becoming adaptive, meaning that the system may change it's response to the same stimulus over time, the notion of repeatability will need reassessing in the context of verification.
 The main research question here is what will constitute `the same' which will be important for passing or failing verification tests but also for determining coverage.

### 2021-03-04 paper/cav_determinisim_main.tex
spell checked

@@
 
+{\color{red} \sc Ensure past/present is used consistently and appropriately throughout.}
+
 % =========================================
@@
 
-If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within acceptable tolerance by design.
+If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within tolerance by design.
 If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
@@
 %
-CARLA exhibits a permissible level of variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
+CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
 {\color{red} Why ``vehicle'' collision and not just ``collision''?}
@@
 \todo[inline]{GC line added to introduction.
-We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
+We need a paragraph on how hard it can be to make a computation fully deterministic, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
 Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
@@
 %
-Ultimately, our objective is to control non-determinisim to minimise simulation variance.
+Ultimately, our objective is to control non-determinism to minimise simulation variance.
 
@@
 %
-{\color{red} Somwhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
+{\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
 

### 2021-03-04 paper/cav_determinisim_main.tex
overall worst case added to max variance def

@@
 
-The maximum variance over the entire set of $n$ repeated simulations is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
+The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
 %{\color{red} This is still a function over time - should it be?}
@@
 
-\subsection{Initial Results and Discussion}\label{ResultsSection}
+\subsection{Initial Results and Discussion}\label{InitialResultsSection}
 
@@
 % ***************************************************
-\subsection{Final Results and Discussion}\label{ResultsSection}
+\subsection{Final Results and Discussion}\label{FinalResultsSection}
 

### 2021-03-04 paper/cav_determinisim_main.tex
edits during review meeting

@@
 CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
-{\color{red} Why ``vehicle'' collision and not just ``collision''?}
 %
@@
 % SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
-and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }
+%{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
+%and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }
 
@@
 In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-{\color{red} Should this be IMPRECISION rather than incorrectness?}
+%{\color{red} Should this be IMPRECISION rather than incorrectness?}
 %
@@
 \subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between dynamic actors {\color{red} Why DYNAMIC?} and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}.
-{\color{red} NOT ACCESSIBLE}
+Previous investigations into the Unreal Engine indicated that collisions between actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}.
 %
@@
 %
-The logs from repeated tests are compared at the same time index $t$ for each actor $a$ to provide a value for the variance of actor $a$ at time index $t$, $\sigma_a^{2}(t)$, at each point in time.
-{\color{red} This gives a variance function over time for each actor.}
-
-{\color{blue} The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.}
+The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.
 
@@
 \end{equation}
-The maximum deviation is the square root of the maximum variance {\color{red} The absolute value of the square root?} and herein referred to as ${\max\sigma}$ for brevity.
-{\color{red} So, ``a'' is any actor or a specific actor?
+The maximum deviation is the absolute value of the square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity. 
+% {\color{red} So, ``a'' is any actor or a specific actor?
 Likewise, ``t'' is any time point or a specific time point?
@@
 Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.
-{\color{red} NOT ACCESSIBLE}
 
@@
 A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
-{\color{red} ???
-about 85\% based on the figure - or not???}
 
@@
 %
-The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\% {\color{red} Why 75\%?}, see Fig.~\ref{CarsCollisionPrePost}.
+The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
 This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
@@
 % ***************************************************
-\section{Method to determine the Variance of a Simulation{\color{red} for what???}} \label{s:methodology}
+\section{Method to determine the Variance of a Simulation {\color{red} Say for what???}} \label{s:methodology}
 In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
@@
 % KIE: Why now point to HW? And, which "boundaries"?
-
-{\color{red} \sc Unclear how this relates to Analysis.
-Is this a summary of the method?
-If so, it should be moved to the front or to the conclusion.} This method can be extended to a broader actor states and actions including actor orientation, speed, and any other status indicators that may be of interest and also including appropriate actions that may be useful for verification purposes.
-This broader As such, Equation~\ref{eq:max_sigma} would need to be adjusted to include these new variables.
-{\color{red} \sc Something is missing in this sentence - please correct before moving to a different place, or removing.}
 

### 2021-03-05 paper/cav_determinisim_main.tex
Done changes to diagrams
Updated diagrams

@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV5.pdf}
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV6.pdf}
     \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v5.pdf}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v6.pdf}
     \caption{Stages of the method proposed to determine the variance of a simulation. {\color{red} \sc please update to match text}}
@@
 These paths should include collisions between actors and potentially collisions between actors and static scenery if this is likely to occur in the simulation or as part of the verification process.
-Actor paths without collision are also important to include as these will serve as a baseline to the other tests.
-{\color{red} \sc What baseline?}
+Actor paths without collision are also important to include as these will serve as a baseline to the other tests; to see if deviations between runs in actor paths increases with collisions.
+{\color{red} \sc What baseline?} {\color{green} \sc Done}
 

### 2021-03-11 paper/cav_determinisim_main.tex
GC up to Sec 7

@@
 % 
-Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor position variance in the simulation with respect to time, assuming fixed actions.
-{\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
+Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions.
+% {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
 %
@@
 
+\subsection{Contribution}
 % In pursuit of an answer to this question we
@@
 % By analysing actor position variance 
-We find that the CARLA simulator is non-deterministic under certain conditions. {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
+We find that the CARLA simulator is non-deterministic.
+Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
+We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
+% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
 % However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
@@
 
-\subsection{Non-Uniform Memory Access (NUMA)}
+\subsection{Non-Uniform Memory Access (NUMA)} \label{s:sources_numa}
 For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
@@
 Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
-{\color{red} Not sure this is correct, see the NUMA section above.
-Why ``only''?
-If one computation out of many takes longer, then their exec order can be affected, or not?}
+If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.
+
+% {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
 %
@@
 Can't control library change, perception stack
+
+% ***************************************************
+%  Empirical Investigation
+% ***************************************************
+\section{Case Study of Simulation Variance} \label{s:case-study}
+
+We present an empirical investigation into using game engines for
+simulation-based verification of autonomous vehicles with a focus on
+characterising sources of non-determinism in order to understand  the impact
+they have on simulation variance. 
+%
+\todo[inline,color=green!40]{GC new: line below.}
+%
+Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
+%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
+%
+Ultimately, our objective is to control non-determinism to minimise simulation variance.
+
+\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
+%
+We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
+{\color{red} and then..?
+Please cover the entire section - consider splitting setup from evaluation.}
+%
+\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
+%
+\todo[inline]{Complete the section structure as this is a very long section.}
+
+% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
+
+% ***************************************************
+%  Scenario Description
+% ***************************************************
+\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
+\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
+
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
+\end{figure}
+
+This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
+%
+Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
+
+This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
+%
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
+\todo[inline]{Decide past or present tense and be consistent throughout.}
+
+\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+
+\subsection{Tolerable Simulation Variance}\label{s:tolerance}
+
+\todo[inline]{Terminology:\\
+{\bf Accuracy} is the degree of closeness/conformance to a known value.
+\\
+{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
+\\
+{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
+\\
+See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
+}
+
+% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
+
+To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
+%
+Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
+%
+% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
+%
+The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small such that assertion checking would fail with minor computational perturbations.
+% {\color{red} {\sc What would fail?}}
+%
+Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
+
+% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
+%
+% but a larger value may be acceptable for non-urban environments, e.g.
+motorway or highway.
+Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
+$1\mu m$.
+% 
+% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
+
+% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
+%
+For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
+%situations such as potentially dangerous close passing of a cyclist. 
+A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
+% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
+Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected. 
+% Thus any variation of less than $1$cm is tolerable. 
+Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
+
+% considered to be sufficient for assertion checking in urban environments. 
+%
+% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
+
+% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
+
+\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
+}
+
+% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
+In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
+%
+\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
+where any value higher than the specified tolerance is considered non-permissible. }
+
+% ***************************************************
+%  Actor Collisions
+% ***************************************************
 
@@
 
-% ***************************************************
-%  Empirical Investigation
-% ***************************************************
-\section{Case Study of Simulation Variance} \label{s:case-study}
-
-We present an empirical investigation into using game engines for
-simulation-based verification of autonomous vehicles with a focus on
-characterising sources of non-determinism in order to understand  the impact
-they have on simulation variance. 
-%
-\todo[inline,color=green!40]{GC new: line below.}
-%
-Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
-%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
-%
-Ultimately, our objective is to control non-determinism to minimise simulation variance.
-
-\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
-%
-We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-{\color{red} and then..?
-Please cover the entire section - consider splitting setup from evaluation.}
-%
-\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
-%
-\todo[inline]{Complete the section structure as this is a very long section.}
-
-% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
-\end{figure}
-
-This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-%
-Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-%
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-\todo[inline]{Decide past or present tense and be consistent throughout.}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-\subsection{Tolerable Simulation Variance}\label{s:tolerance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
-
-To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-%
-Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
-%
-% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
-%
-The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small as to fail with minor computational perturbations.
-{\color{red} {\sc What would fail?}}
-%
-Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
-
-% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-%
-% but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-% 
-% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
-%
-For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
-%situations such as potentially dangerous close passing of a cyclist. 
-A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
-% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected. 
-% Thus any variation of less than $1$cm is tolerable. 
-Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-
-% considered to be sufficient for assertion checking in urban environments. 
-%
-% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
-In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-%
-\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
-where any value higher than the specified tolerance is considered non-permissible. }
-
-% ***************************************************
-%  Actor Collisions
-% ***************************************************
-
 \subsection{Actor Collisions}\label{S:Actor_Collisions}
@@
 
-Instead of using ..., {\color{red} \sc please complete} herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
+Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
 %
@@
 % ***************************************************
-\subsection{Internal Settings}
+\subsection{Simulator Settings}
 
@@
 
-\subsection{Initial Results and Discussion}\label{InitialResultsSection}
-
-{\color{red} This is for unrestricted, so this does not come as a surprise later.}
+% \subsection{Initial Results and Discussion}\label{InitialResultsSection}
+
+% {\color{red} This is for unrestricted, so this does not come as a surprise later.}
 
@@
 
-After discovering that internal engine settings could not achieve a simulation variance that met the required 
+% After discovering that internal engine settings could not achieve a simulation variance that met the required 
 % \textit{deterministic threshold}, 
-\textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
-settings external to the game engine were explored. 
+% \textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
+After examining controls the internal game engine, external settings were explored. 
 %
@@
 For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
-This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the CARLA platform.
-{\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented.
+
+Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load.
+This initial observation was explored more fully in this work using the CARLA platform.
+
+% {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented.
 Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}
@@
 Resource utilisation figures reported here should be considered approximate values.
+%
+Resource utilisation was capped to 75\% in some parts of the results and referred to as \textit{restricted}.
+This was done to limit the load on the test system and hence limiting simulation variance.
+The term \textit{unrestricted} places no such limit on the system load.
+
+% {\color{red} To understand the impact of collisions and high resource utilisation, the restricted column shows a subset of the results where post-collision data and experiments above 75 resource utilisation have been removed}
+%
 % The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
 %
-{\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
+% {\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
 
@@
 Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
-%
-Each test was repeated 1000 times. {\color{red} Justify why 1000 - see my later comment in the Method section.}
+
+% Each test was repeated 1000 times. 
+Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
+While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results.
+Therefore each test was repeated 1000 times to provide a sufficient sample size.
+% {\color{red} Justify why 1000 - see my later comment in the Method section.}
 %
@@
     \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
-    \item Non-uniform memory access (NUMA): \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores, but only a minor (2\%) improvement in simulation variance was observed.
-Therefore this was not used for subsequent testing.
-{\color{red} But if we aim to min the sources of non-det, then that should be considered/used?}
 \end{itemize}
@@
 % In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
+
+% ***************************************************
+%  Results and Discussion
+% ***************************************************
+\section{Results and Discussion}\label{FinalResultsSection}
 
@@
 
-% ***************************************************
-%  Results and Discussion
-% ***************************************************
-\subsection{Final Results and Discussion}\label{FinalResultsSection}
-
-{\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}
-
-A summary of all the results is shown in Table~\ref{TableOfExperiments}.
-In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
-%
-From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-%
-However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
-%
-\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-}
-%
-Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
-
-Resource utilisation was found to have a significant impact on \textit{simulation variance}.
-%
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
-%
-Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-%
-\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
-%
-
-A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
-
-Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision.
-%
-The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-
-\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-}
-% \todo{Why ``not reliable''.}
-
 \begin{figure*}[h]
@@
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
-    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
-    \label{NICEExperimentStressSummary}
-\end{figure}
+% {\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}
+
+A summary of all the results is shown in Table~\ref{TableOfExperiments}.
+In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
+%
+From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+%
+However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
+%
+\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
+}
+%
+Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
+
+Resource utilisation was found to have a significant impact on \textit{simulation variance}.
+%
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
+In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
+%
+Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
+%
+\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
+%
+
+A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
+
+Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision.
+%
+The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
+
+\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
+%
+This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
+}
+% \todo{Why ``not reliable''.}
 
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v6.pdf}
-    \caption{Stages of the method proposed to determine the variance of a simulation. {\color{red} \sc please update to match text}}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v7.pdf}
+    \caption{Stages of the method proposed to determine the variance of a simulation.}
     \label{method_diagram}
@@
 \subsection{Process Scheduling Priority}
-An investigation into the impact of process scheduling on the simulation variance was undertaken {\color{red} say why}.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
+    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
+    \label{NICEExperimentStressSummary}
+\end{figure}
+
+An investigation into the impact of process scheduling on the simulation variance was undertaken following the observations of increased simulation variance with increased resource utilisation.
 The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
@@
 
-The results are presented in Fig.~\ref{NICEExperimentStressSummary} {\color{red} What is on the y axis?} where the box denotes the inter-quartile range {\color{red} of???}, non-outlier limits by the whiskers and a horizontal bar for the median.
-The figure shows that decreasing the priority {\color{red} of the simulator process} (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot).
+The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range of deviation, non-outlier limits by the whiskers and a horizontal bar for the median.
+The figure shows that decreasing the priority of the simulator process (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot).
 Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results.
@@
 
-\subsection{Investigation Summary} \label{s:empirical_summary} {\color{red} Factor into overall conclusion of the case study}
-
+\subsection{Non-uniform memory access (NUMA)}
+An additional investigation into memory access was undertaken given the potential impact that data transfer cost and execution order may have on simulation variance, see Section~\ref{s:sources_numa}.
+%
+The program \texttt{numactl}~\cite{numactl_NUMA} allows a process to run with a specified memory placement policy, essentially allowing the process to be bound to a particular CPU core. %
+\texttt{numactl} was used to fix the simulator and test script to single cores, and a (2\%) improvement in simulation variance was observed. %
+This was considered minor comparative to the changes in simulation variance observed in other aspects of the case study, for example the pre- and post-collision between a vehicle and pedestrian saw a change of $10^{14}$.
+Therefore \texttt{numactl} was not used for subsequent testing.
+
+However, practitioners that are aiming to minimise simulation variance should use a technique such as this to minimise data transfer cost and potentially minimise any potential effect on process execution order that may lead to increased simulation variance.
+
+\subsection{Investigation Summary} \label{s:empirical_summary} 
+% {\color{red} Factor into overall conclusion of the case study}
 % \todo[inline]{GC-need to reference the 'restricted' set of results here too}
@@
 (is this 2nd one true, or does carla ignore render requests?)}
-
-This empirical investigation has highlighted the shortcomings of using a games engine for simulation based verification and advice for best working practice.
+These empirical investigations have highlighted the shortcomings of using a games engine for simulation based verification and suggests advice for best working practice. 
+%
+It was observed that resource utilisation positively correlates with simulation variance and that specific simulation events, such as vehicle collisions, can also lead to a breach in permissible tolerance.
+%
+The investigation found that the effect of higher simulation variance as a result of increased resource utilisation can be reduced, but not omitted entirely, by controlling the scheduling policy. 
+%
+The investigation into specifying memory placement on the simulation process did improve simulation variance but only by a minor amount of 2\%.
+
 However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly. 
@@
 % ***************************************************
-\section{Method to determine the Variance of a Simulation {\color{red} Say for what???}} \label{s:methodology}
-In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
+\section{Method to determine the Variance of a Simulation} \label{s:methodology}
+% In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
 This method can be used to determine the operational domains of permissible variance of a simulation environment.
 %
-{\color{red} MAKES NO SENSE: {\sc In this section a method for determining the variance of simulated actor path trajectories (KIE: a trajectory is a path), termed simulation variance, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.}}
-%
-In addition, recommendations and best practice guidelines for minimising simulation variance are suggested. {\color{red} \sc Highlight these in the text.}
+In this section a method for determining the simulation variance of actor paths and resolving the operational domains of permissible variance is presented here as a work flow, see Fig.~\ref{method_diagram}.
+%
+In addition, recommendations and best practice guidelines for minimising simulation variance are suggested, such as fixing the physics time step in \textit{Simulator Settings}. 
+% {\color{red} \sc Highlight these in the text.}
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
@@
 %
-Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
-While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results.
-{\color{red} \sc Again, separate the study from the method.
-This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.} It is therefore recommended that the sample size is determined empirically dependent on the number of results that exceed the tolerance.
-{\color{red} \sc How would one know?}
+% {\color{red} \sc Again, separate the study from the method.
+This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.}
+It is recommended that the chosen sample size is determined empirically. By increasing the sample size and observing the simulation variance does not change
+...
+...
+ dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
 % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
@@
 For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment.
+
 %Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.

### 2021-03-11 paper/cav_determinisim_main.tex
GC up to Sec 8

@@
 % ***************************************************
-\section{Results and Discussion}\label{FinalResultsSection}
+\section{Results and Discussion}\label{s:FinalResultsSection}
 
@@
 %
-In addition, recommendations and best practice guidelines for minimising simulation variance are suggested, such as fixing the physics time step in \textit{Simulator Settings}. 
-% {\color{red} \sc Highlight these in the text.}
+In addition, recommendations and best practice guidelines for minimising simulation variance are suggested.
+% , such as fixing the physics time step in \textit{Simulator Settings}. 
+{\color{red} \sc Highlight these in the text.}
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
@@
 
-The method consists of a sequence of five stages; experimental design, simulator configuration, external configuration, execution and analysis.
+The method consists of a sequence of five stages; experimental design, simulator settings, external settings, execution and analysis.
 In the following, each stage is described in detail with reference to the items listed for each stage in Fig.~\ref{method_diagram}.
-{\color{red} NOTE: renamed stages.}
 
@@
 
+{\color{green!99} \sc The experiment design is based around a series of carefully selected tests that varies one the sources of non-determinism whilst keeping all others parameters constant.
+Each test is repeated $n$ times and the analysis of the results provide a confidence around a degree of statistical certainty.
+By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and prioritised.
+}
+
 \subsubsection{Actors} \label{s:actors}
 
-Different actor types may be handled differently by the game engine.
+% Different actor types may be handled differently by the game engine.
 For instance, our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
-However, the introduction of CARLA vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
-{\color{red} \sc This needs rephrasing as it appears to contradict the statement above.} As such, all actors that could be included in any future simulation should be tested, including any non-standard CARLA or bespoke actors.
-{\color{red} \sc ...
-such as the ego vehicle...?
-[see section on ``External Settings'' last point - Ego Controller.]}
+However, the introduction of CARLA vehicles increased the simulation variance of the pedestrian-vehicle test by 12 orders of magnitude.
+{\color{red} \sc This needs rephrasing as it appears to contradict the statement above.}
+%
+As such, all actors that could be included in any future simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see section on ``External Settings'' last point - Ego Controller.
+% {\color{red} \sc ... such as the ego vehicle...? [see section on ``External Settings'' last point - Ego Controller.]}
 %
@@
 Actor paths without collision are also important to include as these will serve as a baseline to the other tests; to see if deviations between runs in actor paths increases with collisions.
-{\color{red} \sc What baseline?} {\color{green} \sc Done}
+% {\color{red} \sc What baseline?} {\color{green} \sc Done}
 
@@
 
-\subsubsection{Simulation Time} {\color{red} \sc Should this not be part of the Execution section?}
+\subsubsection{Simulation Time} 
+% {\color{red} \sc Should this not be part of the Execution section?} 
+The simulation time will depend on the actor paths and terminating conditions.
 The simulation time should be sufficient to record interactions between actors but not so long that the testing take an inconvenient time to complete.
-In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-{\color{red} \sc Point to the section that specified that.}
+In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25} in Section~\ref{s:FinalResultsSection}.
+The termination conditions of the simulation can be set by, for example, actors reaching their final trajectory waypoints.
+ % {\color{red} \sc Point to the section that specified that.}
 
@@
 \subsubsection{NUMA Control}
-Control over a Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multiprocessors with shared memory.
-This control allows the simulator program to be fixed on a single core, reducing and unifying memory access time.
-%
-Note that initial screening tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}. 
-%
-This control may assist if simulation variance is borderline to the tolerance but not seen as essential. {\color{red} \sc Unclear: what is not essential?}
-
-\subsubsection{Ego Vehicle Controller {\color{red} \sc Why is this not covered under actors? [see my note in the Actors section]}}
-{\color{red} \sc Clearly separate the investigation from the Method.
+Control over a Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multi-core processors with shared memory.
+This control allows the simulator process to be fixed on a single core, reducing and unifying memory access time.
+%
+Investigations in tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}. 
+%
+% Using \texttt{numactl} allows a process to run with a specified memory placement policy, fixing the process to a single core.
+This will reduce the data transfer cost and the opportunity for disrupted execution order that may be responsible for increase simulation variance.
+%
+Using a fixed memory placement policy may assist if simulation variance is borderline to the tolerance but only of the order of a few percent from our observations.
+
+\subsubsection{Ego Vehicle Controller}
+% {\color{red} \sc Why is this not covered under actors? [see my note in the Actors section]}
+% {\color{red} \sc Clearly separate the investigation from the Method.
 The fact that the case study did not have an ego vehicle should be mentioned as a limitation of the case study, rather than elaborated on here, but why does this make a difference?}
-The impact of the ego vehicle  should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
+%
+An ego vehicle was not used in this study, but the impact of introducing this to the simulator can be considered here.
+The ego vehicle is seen as another actor in the simulation but care must be taken to ensure that any control algorithms, machine learning modules and processing pipelines are deterministic.
+We recommended this be treated as a separate source of non-determinism and handled accordingly.
+
+% ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
 {\color{red} \sc Long and cumbersome sentence with strange grammar: No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
@@
 This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.}
-It is recommended that the chosen sample size is determined empirically. By increasing the sample size and observing the simulation variance does not change
-...
-...
- dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
-% To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
+It is recommended that the chosen sample size is determined empirically. 
+%
+{\color{green} We recommend monitoring variance while increasing the sample size in orders of magnitude
+until there is confidence that $\max\sigma$ will not exceed the specified tolerance for the verification process. 
+%
+Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.}
+
+% By increasing the sample size and observing that the simulation variance is stable, one can infer that a sufficient number of samples have been collected.
+If the variance changes with a change in sample size then a stable condition has not been found.
+% ...
+% ...
+%  dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
+% % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
 

### 2021-03-11 paper/cav_determinisim_main.tex
Modified Section III. Tried to go through past/present consistency as much as I can.

@@
 
-This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
-%
-A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is then presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
-%
-Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
-
-Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
+This paper reviewed and identified the potential causes of non-deterministic behaviours in game engines. 
+%
+A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
+%
+Different configurations and utilisations of the software and hardware were explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
+
+Finally, a method of a general nature was proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
 
@@
 Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules, e.g.\ through assertion checking during simulation. 
-There have been a number of fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-Simulation environments offer a means to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
+There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
+Simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
 In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
@@
 %
-Furthermore, they are easy to setup and run with respect to on-road testing and are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
+Furthermore, they are easy to set up and run compared to on-road testing and are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
 %
@@
 % \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-\textit{Variance} is used here to define the spread, or distribution, of the generated output data with respect to the mean value.
+\textit{Variance} is used here to define the spread or distribution, of the generated output data with respect to the mean value.
 \textit{Precision} is synonymous with \textit{variance} although inversely related mathematically.
@@
 \subsubsection{Accuracy}
-Precision and tolerance should not to be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
-We therefore define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation.
-%
-For a driving simulation the reference value 
+Precision and tolerance should not be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
+Therefore, we define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation.
+%
+For a driving simulation, the reference value 
 % , for which accuracy can be measured against, 
@@
 %
-In practice full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
-In most cases, in fact, it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
+In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
+In most cases it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
 \\
@@
 %e.g. it has a non-permissible variance in,
-it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to re-produce, understand and remove bugs and rendering verification results unstable.
+it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
 In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
@@
 % In pursuit of an answer to this question we
-In this paper we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
+In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
 %
@@
 % on verification tasks. 
-In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact on verification results.
+In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
 %
@@
 
-There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
+There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
 Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
@@
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-The gamer neither needs a faithful representation of reality, i.e.\ the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance.
+% 
+The gamer neither needs a faithful representation of reality (i.e.
+gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
+gamer accepts low precision).
+In contrast, high accuracy and precision are necessary for AV development, testing and verification.
+% The gamer neither needs a faithful representation of reality, i.e.\ the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance.
 %, i.e.\ can be low \textit{precision}.
-%
-In contrast, the precision required for AV development and testing is very high, especially for perception stack testing, but also, more generally, for obtaining reliable verification results and for visual inspection of accidents or near misses.
+% %
+% In contrast, the precision required for AV development and testing is very high, especially for perception stack testing, but also, more generally, for obtaining reliable verification results and for visual inspection of accidents or near misses.
 If necessary, it is acceptable to achieve the required tolerance %on the precision
-at the cost of real-time performance.
+% at the cost of real-time performance.
 For example, for perception stack testing the sensors need to get input that is repeatable so that if any software bugs are found they can be re-played and the issue resolved.
@@
 
-However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating point operations producing different results.
+However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
 Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
@@
 % https://en.wikipedia.org/wiki/Race_condition#cite_note-1
-Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy of the operating system.
+Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.

### 2021-03-12 coveringLetter/coveringLetter.tex
GC friday

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% LaTeX Template
% Version 2.0 (12/2/17)
%
% This template originates from:
% http://www.LaTeXTemplates.com
%
% Authors:
% Brian Moses
% Vel (vel@LaTeXTemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt, a4paper]{letter} % Set the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc)

\input{structure.tex} % Include the file that specifies the document structure

\longindentation=0pt % Un-commenting this line will push the closing "Sincerely," and date to the left of the page

%----------------------------------------------------------------------------------------
%	YOUR INFORMATION
%----------------------------------------------------------------------------------------

\Who{Dr. Greg Chance} % Your name

\Title{, PhD, CEng, MInstP} % Your title, leave blank for no title

\authordetails{
	Trustworthy Systems Lab\\ % Your department/institution
	University of Bristol\\
	1 Cathedral Square\\ % Your address
	Bristol, BS1 5DD, UK\\ % Your city, zip code, country, etc
	Email: greg.chance@bristol.ac.uk\\ % Your email address
	URL: \url{bristol.ac.uk}\\ % Your URL
	% URL: \url{bristol.ac.uk/engineering/research/trustworthy-systems-laboratory/}\\ % Your URL
	% YouTube: \url{https://youtu.be/1emO3-OYn5k}
}

%----------------------------------------------------------------------------------------
%	HEADER CONTENTS
%----------------------------------------------------------------------------------------

\logo{logo.png} % Logo filename, your logo should have square dimensions (i.e. roughly the same width and height), if it does not, you will need to adjust spacing within the HEADER STRUCTURE block in structure.tex (read the comments carefully!)

\headerlineone{} % Top header line, leave blank if you only want the bottom line
\headerlinetwo{} % Bottom header line

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TO ADDRESS
%----------------------------------------------------------------------------------------

\begin{letter}{
	Prof. Chen\\
	Director of TASI\\
	Electrical and Computer Engineering\\
	Purdue School of Engineering \& Technology\\
	Indianapolis, IN 46202
}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

\opening{Dear Prof Yaobin Chen,}

We are pleased to submit an original research article entitled “On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification” for consideration of publishing in IEEE Transactions on Intelligent Transportation Systems.

We confirm that this work is original and has not been published elsewhere and is not currently under consideration for publication elsewhere. We have no conflicts of interest to disclose.

Thank you for your consideration.

\closing{Sincerely,}

% \begin{figure}
% 	\includegraphics[width=1in]{signature.png}
% \end{figure}

%----------------------------------------------------------------------------------------
%	OPTIONAL FOOTNOTE
%----------------------------------------------------------------------------------------

% Uncomment the 4 lines below to print a footnote with custom text
% \def\thefootnote{}
% \def\footnoterule{\hrule}
% \footnotetext{\hspace*{\fill}{\footnotesize\em Footnote text}}
% \def\thefootnote{\arabic{footnote}}

%----------------------------------------------------------------------------------------

\end{letter}

\end{document}

### 2021-03-12 coveringLetter/structure.tex
GC friday

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% Structure Specification File
% Version 1.0 (12/2/17)
%
% This file originates from:
% http://www.LaTeXTemplates.com
%
% Authors:	
% Brian Moses
% Vel (vel@LaTeXTemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\usepackage{graphicx} % Required for including pictures

\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage{gfsdidot} % Use the GFS Didot font: http://www.tug.dk/FontCatalogue/gfsdidot/
\usepackage{microtype} % Improves typography

\pagestyle{empty} % Suppress headers and footers

\setlength\parindent{1cm} % Paragraph indentation

% Create a new command for the horizontal rule in the document which allows thickness specification
\makeatletter
\newcommand{\vhrulefill}[1]{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
\makeatother

%----------------------------------------------------------------------------------------
%	DOCUMENT MARGINS
%----------------------------------------------------------------------------------------

\usepackage{geometry} % Required for adjusting page dimensions

\geometry{
	top=1cm, % Top margin
	bottom=1.5cm, % Bottom margin
	left=3cm, % Left margin
	right=3cm, % Right margin
	%showframe, % Uncomment to show how the type block is set on the page
}

%----------------------------------------------------------------------------------------
%	DEFINE CUSTOM COMMANDS
%----------------------------------------------------------------------------------------

\newcommand{\logo}[1]{\renewcommand{\logo}{#1}}

\newcommand{\Who}[1]{\renewcommand{\Who}{#1}}
\newcommand{\Title}[1]{\renewcommand{\Title}{#1}}

\newcommand{\headerlineone}[1]{\renewcommand{\headerlineone}{#1}}
\newcommand{\headerlinetwo}[1]{\renewcommand{\headerlinetwo}{#1}}

\newcommand{\authordetails}[1]{\renewcommand{\authordetails}{#1}}

%----------------------------------------------------------------------------------------
%	AUTHOR DETAILS STRUCTURE
%----------------------------------------------------------------------------------------

\newcommand{\authordetailsblock}{
	\hspace{\fill} % Move the author details to the far right
	\parbox[t]{0.48\textwidth}{ % Box holding the author details; width value specifies where it starts and ends, increase to move details left
		\footnotesize % Use a smaller font size for the details
		\Who\\ % Author name
		\textit{\authordetails} % The author details text, all italicised
	}
}

%----------------------------------------------------------------------------------------
%	HEADER STRUCTURE
%----------------------------------------------------------------------------------------

\address{
	\includegraphics[width=2.1in]{\logo} % Include the logo of author institution
	\hspace{0.62\textwidth} % Position of the institution logo, increase to move left, decrease to move right
	\vskip -0.1\textheight~\\ % Position of the large header text in relation to the institution logo, increase to move down, decrease to move up
	% \Large\hspace{0.2\textwidth}\headerlineone\hfill ~\\[0.006\textheight] % First line of institution name, adjust hspace if your logo is wide
	\Large\hspace{0.4\textwidth}\headerlineone\hfill ~\\[0.006\textheight] % First line of institution name, adjust hspace if your logo is wide
	\hspace{0.4\textwidth}\headerlinetwo\hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
	\makebox[0ex][r]{\textbf{\Who\Title}}\hspace{0.01\textwidth} % Print author name and title with a little whitespace to the right
	~\\[-0.01\textheight] % Reduce the whitespace above the horizontal rule
	\hspace{0.4\textwidth}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
	\authordetailsblock % Include the letter author's details on the right side of the page under the horizontal rule
	\hspace{-0.25\textwidth} % Horizontal position of the author details block, increase to move left, decrease to move right
	\vspace{-0.1\textheight} % Move the date and letter content up for a more compact look
}

%----------------------------------------------------------------------------------------
%	COMPOSE THE ENTIRE HEADER
%----------------------------------------------------------------------------------------

\renewcommand{\opening}[1]{
	{\centering\fromaddress\vspace{0.05\textheight} \\ % Print the header and from address here, add whitespace to move date down
	\hspace*{\longindentation}\today\hspace*{\fill}\par} % Print today's date, remove \today to not display it
	{\raggedright \toname \\ \toaddress \par} % Print the to name and address
	\vspace{1cm} % White space after the to address
	\noindent #1 % Print the opening line
}

%----------------------------------------------------------------------------------------
%	SIGNATURE STRUCTURE
%----------------------------------------------------------------------------------------

\signature{\Who\Title} % The signature is a combination of the author's name and title

\renewcommand{\closing}[1]{
	\vspace{2.5mm} % Some whitespace after the letter content and before the signature
	\noindent % Stop paragraph indentation
	\hspace*{\longindentation} % Move the signature right to the value of \longindentation
	% \hspace*{-6.5cm}\includegraphics[width=1in]{signature.png}
	\parbox{\indentedwidth}{
		\raggedright
		#1 % Print the signature text
		\vskip 1.65cm % Whitespace between the closing text and author's name for a physical signature
		\fromsig % Prints the value of \signature{}, i.e. author name and title
	}
}

### 2021-03-12 paper/cav_determinisim_main.tex
GC friday part 2

@@
 
-This paper reviewed and identified the potential causes of non-deterministic behaviours in game engines. 
+This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
 %
@@
 %
-Different configurations and utilisations of the software and hardware were explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
-
-Finally, a method of a general nature was proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
+Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
+
+Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
 
@@
 
+% \subsection{Contribution}
+% In pursuit of an answer to this question we
+In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
+%
+In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
+% KIE: This is covered in the next sentence:
+% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
+%
+\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
+}
+%
+% By analysing actor position variance 
+We find that the CARLA simulator is non-deterministic.
+Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
+We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
+% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
+% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
+%
+CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
+%
+\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
+However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
+
+The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
+%
+\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
+In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
+}
+%
+Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
+%impact that using a game engine for AV simulation may have.
+% on verification tasks. 
+In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
+%
+\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
+In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
+}
+%
+% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
+%We present a case study to illustrate the use of this method.
+The study highlights the effects of non-determinism of a game engine on simulation results.
+Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
+%
+%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
+%
+
+This paper is structured as follows.
+%
+Section~\ref{s:prelim} defines terms used throughout the paper and identifies when determinism is needed. 
+Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
+%
+Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
+%
+Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
+%
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
+
 \todo[inline,color=green!40]{GC New below}
 
-\section{Preliminaries}
+\section{Preliminaries} \label{s:prelim}
 
@@
 \todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
-
-\subsection{Contribution}
-% In pursuit of an answer to this question we
-In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
-%
-In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
-% KIE: This is covered in the next sentence:
-% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
-%
-\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
-}
-%
-% By analysing actor position variance 
-We find that the CARLA simulator is non-deterministic.
-Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
-We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
-% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
-% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
-%
-CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
-%
-\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
-
-The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
-%
-\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
-In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
-}
-%
-Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
-%impact that using a game engine for AV simulation may have.
-% on verification tasks. 
-In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
-In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
-}
-%
-% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
-%We present a case study to illustrate the use of this method.
-The study highlights the effects of non-determinism of a game engine on simulation results.
-Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
-%
-%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
-%
-
-This paper is structured as follows.
-%
-Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
-%
-Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
-%
-Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
-%
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
 
@@
 We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-{\color{red} and then..?
-Please cover the entire section - consider splitting setup from evaluation.}
+A discussion on the internal and external settings of the simulation is included, along with system configuration and pre-screening sections.
+% {\color{red} and then..? Please cover the entire section - consider splitting setup from evaluation.}
 %
@@
 
-A summary of all the results is shown in Table~\ref{TableOfExperiments}.
+A summary of main the results are shown in Table~\ref{TableOfExperiments}.
 In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
@@
 Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
+
+A correlation between resource utilisation and simulation variance has been observed in these results which may be due to process scheduling.
+Rather than liming utilisation of the whole system, another approach may be to promote the scheduling priority of the simulation process which is explored in Section~\ref{r:process_scheduling}.
+Furthermore, we were keen to investigate the impact of memory access and include a brief investigation of this in Section~\ref{r:numa}.
 %
@@
 
-\subsection{Process Scheduling Priority}
+\subsection{Process Scheduling Priority} \label{r:process_scheduling}
 
@@
 
-\subsection{Non-uniform memory access (NUMA)}
+\subsection{Non-uniform memory access (NUMA)} \label{r:numa}
 An additional investigation into memory access was undertaken given the potential impact that data transfer cost and execution order may have on simulation variance, see Section~\ref{s:sources_numa}.
@@
 
-{\color{green!99} \sc The experiment design is based around a series of carefully selected tests that varies one the sources of non-determinism whilst keeping all others parameters constant.
-Each test is repeated $n$ times and the analysis of the results provide a confidence around a degree of statistical certainty.
-By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and prioritised.
+{\color{green!99} \sc The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
+Each test is repeated $n$ times and the analysis of the results provide a confidence with a degree of statistical certainty.
+By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and addressed.
 }
@@
 %
-As such, all actors that could be included in any future simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see section on ``External Settings'' last point - Ego Controller.
+All actors that could be included in the simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see Section~\ref{ego_actor}. 
 % {\color{red} \sc ... such as the ego vehicle...? [see section on ``External Settings'' last point - Ego Controller.]}
@@
 
-\subsection{External Settings}
+\subsection{External Settings} \label{ego_actor}
 
@@
 {\color{green} We recommend monitoring variance while increasing the sample size in orders of magnitude
-until there is confidence that $\max\sigma$ will not exceed the specified tolerance for the verification process. 
+until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process. 
 %
@@
 %
+{\color{green} To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
+}
+%
 {\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.

### 2021-03-19 paper/cav_determinisim_main.tex
typos

@@
 
-A summary of main the results are shown in Table~\ref{TableOfExperiments}.
+A summary of the main the results are shown in Table~\ref{TableOfExperiments}.
 In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
@@
 A correlation between resource utilisation and simulation variance has been observed in these results which may be due to process scheduling.
-Rather than liming utilisation of the whole system, another approach may be to promote the scheduling priority of the simulation process which is explored in Section~\ref{r:process_scheduling}.
+Rather than limiting utilisation of the whole system, another approach may be to promote the scheduling priority of the simulation process which is explored in Section~\ref{r:process_scheduling}.
 Furthermore, we were keen to investigate the impact of memory access and include a brief investigation of this in Section~\ref{r:numa}.

### 2021-03-22 paper/cav_determinisim_main.tex
typos
change lead author

@@
 
-\author{Abanoub Ghobrial, Greg Chance, Kevin McAreavey, Severin Lemaignan, Tony Pipe, Kerstin Eder 
+\author{Greg Chance, Abanoub Ghobrial, Kevin McAreavey, Severin Lemaignan, Tony Pipe, Kerstin Eder 
 \thanks{{\footnotesize
@@
 
-\textit{Abanoub Ghobrial and Greg Chance contributed equally to this paper and are both corresponding authors.}
+\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper and are both corresponding authors.}
 %
@@
 
+Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
 Abanoub Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
-Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
 Kevin McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
@@
     \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV6.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
+    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
+Tests 5 and 6 overlap having almost identical results.
+Note that the lines between data points are only a guide.}
     \label{ExperimentsStressSummary}

### 2021-03-22 paper/cav_determinisim_main.tex
removing ighlights

@@
 
-{\color{red} \sc Ensure past/present is used consistently and appropriately throughout.}
+% {\color{red} \sc Ensure past/present is used consistently and appropriately throughout.}
 
@@
 
-{\color{red} \sc Surely, a series of systematically designed experiments, each repeated X times, should be set up, where each experiment varies one of the suspected sources of non-determinism?
+% {\color{red} \sc Surely, a series of systematically designed experiments, each repeated X times, should be set up, where each experiment varies one of the suspected sources of non-determinism?
 Then, the analysis of the data from each experiment should reveal the actual sources of non-determinism for the given setting?
@@
 
-{\color{green!99} \sc The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
+% {\color{green!99} \sc }
+The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
 Each test is repeated $n$ times and the analysis of the results provide a confidence with a degree of statistical certainty.
 By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and addressed.
-}
 
@@
 %
-{\color{green} We recommend monitoring variance while increasing the sample size in orders of magnitude
+We recommend monitoring variance while increasing the sample size in orders of magnitude
 until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process. 
 %
-Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.}
+Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
 

### 2021-03-22 paper/cav_determinisim_main.tex
removing highlights

@@
 However, game engines are seldom deterministic. 
-
+%
 This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
@@
 Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
-
+%
 Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
@@
 % , such as fixing the physics time step in \textit{Simulator Settings}. 
-{\color{red} \sc Highlight these in the text.}
+% {\color{red} \sc Highlight these in the text.}
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
@@
 %
-{\color{green} To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
-}
-%
-{\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.
+To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
+%
+% {\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.
 This seems to have been commented out but was here before.}

### 2021-03-24 paper/cav_determinisim_main.tex
updadting for arxiv submission

@@
 %
-Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
+Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.
 

### 2021-03-24 paper_arxiv/arxiv_cav_determinisim_main.tex
Rearrangments to fit guidelines
updadting for arxiv submission

--- 2020-11-19 paper/cav_determinisim_main.tex
@@
 \usepackage{siunitx}
-% \usepackage{draftwatermark}
-
-%-------------------------------------------------------------------------------
+
+% \usepackage{multicol}
+\setcounter{secnumdepth}{3}
+
+% Make tables a little bit easier to read
+\renewcommand{\arraystretch}{1.15}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% NOTES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcounter{mnotei}
+\setcounter{mnotei}{0}
+\newcommand{\mnote}[1]{%
+  {\scriptsize\textsf{$^{\textcolor{red}{[n.\themnotei]}}$}}%
+  \marginpar{\scriptsize\textsf{\textcolor{red}{[n.\themnotei]: #1}}}%
+  % \marginpar{\scriptsize\textsf{n.\themnotei: #1}}%
+  \stepcounter{mnotei}
+}
+%\renewcommand{\mnote}[1]{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Provide paragraph-style table columns
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+% Give more marginpar
+\setlength{\marginparwidth}{25mm}
+
+% Give more marginpar
+\setlength{\marginparwidth}{25mm}
+
+% Prompt people to generate graphs
+\newcommand{\includegraphicsmaybe}[2]{
+    \IfFileExists{#2}{\includegraphics[#1]{#2}}{
+    \detokenize{File #2 is missing, maybe you need to run plots.py?}
+}}
+
+\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
 
 \begin{document}
-
-% ***************************************************
-%  TITLE
-% ***************************************************
+\bibliographystyle{ieeetr}
+\mainmatter
+
 \title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-% \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-% \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
-% \author[1,3]{Greg Chance}
-% \author[1,3]{Kevin McAreavey}
-% \author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-% \author[2,3]{Tony Pipe}
-% \author[1,3]{Kerstin Eder}
-% \affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
-% \affil[2]{University of West of England, Bristol, UK}
-% \affil[3]{Bristol Robotics Laboratory, Bristol, UK}
-
-% \author{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
-% \author{Greg Chance}
-% \author{Kevin McAreavey}
-% \author{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-% \author{Tony Pipe}
-
-\author{Abanoub Ghobrial, Greg Chance, Kevin McAreavey, Severin Lemaignan, Tony Pipe, Kerstin Eder 
-\thanks{{\footnotesize
-Manuscript  
-received ...;
-revised ...;  
-accepted.... 
-Date of publication ...;
-date of current version ....
-
-This research has in part been funded by the ROBOPILOT and CAPRI projects.
-Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-The Associate Editor for this paper was ....(\textit{Corresponding author: Abanoub Ghobrial.})
-
-A.Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
-G.Chance (e-mail: greg.chance@bristol.ac.uk), 
-K.McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
-and 
-K.Eder (e-mail: kerstin.eder@bristol.ac.uk) 
-are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom.
-
-S.Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
-and
-T.Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.
-
-Digital Object Identifier ....
-}}}
-% \affil{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
-% \affil{University of West of England, Bristol, UK}
-% \affil{Bristol Robotics Laboratory, Bristol, UK}
-%
-\markboth{IEEE TRANSACTIONS ON INTELLIGENT TRANSPORTATION SYSTEMS VOL.
-..., NO.
-..., date...}{Ghobrial \MakeLowercase{\textit{et al.}}: On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-%
+
+\titlerunning{On Determinism of Game Engines... }%used for Simulation-based }%Autonomous Vehicle Verification}
+
+\author{Greg Chance\inst{1,3}, Abanoub Ghobrial\inst{1,3}, Kevin McAreavey\inst{1,3}, Severin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}
+
+\authorrunning{A. Ghobrial et al.}
+\institute{University of Bristol, Bristol, UK \and University of the West of England, Bristol, UK \and Bristol Robotics Laboratory, Bristol, UK}
+\tocauthor{Authors' Instructions}
 \maketitle
-%
-\begin{abstract}
-\noindent 
-Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
+
+\makeatletter
+\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
+                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
+                       {4\p@ \@plus 2\p@ \@minus 2\p@}%
+                       {\normalfont\normalsize\bfseries\boldmath
+                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
+\makeatother
+
+% \begin{multicols}{2}
+
+\textbf{\textit{Abstract}--Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
 %
@@
 %
-Thus, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
+Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
 %
 However, game engines are seldom deterministic. 
-
-This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
-%
-This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
-%
-Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
-variance between repeats becomes negligible for development and testing work.
-
-Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
-% %
-% A key requirement for simulation-based development and verification is determinism.
-% %
-% Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
-% %
-% Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
-% %
-% However, game engines are seldom deterministic.
-
-% This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.
-% %
-% This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
-% %
-% Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
-variance between repeats becomes negligible for development and testing work.
-
-% Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
-% %
-% A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
-% %
-% Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
-% \todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
-% %
-% % KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
-% % These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
-% %
-% However, game engines do not need to be deterministic per se. % when used for games. 
-% %
-% This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively.
-% %
-% \todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
-% %
-% A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification.
-The proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-% %
-% % \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}
-
-\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline,color=yellow!40]{AG Alternative: A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}
-
-% The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
-It is important for these engines to be deterministic in order for one to carry out these different processes.
-However, such engines are inherently non-deterministic.
-We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic.
-We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.
-\end{abstract}
-
-\begin{IEEEkeywords}
-Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
-\end{IEEEkeywords}
-\IEEEpeerreviewmaketitle
+%
+This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
+%
+A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
+%
+Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
+%
+Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.}
 
@@
 % \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. 
-\IEEEPARstart{S}{}imulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
+Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
 %
@@
 
+% \subsection{Contribution}
+% In pursuit of an answer to this question we
+In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
+%
+In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
+% KIE: This is covered in the next sentence:
+% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
+%
+\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
+}
+%
+% By analysing actor position variance 
+We find that the CARLA simulator is non-deterministic.
+Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
+We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
+% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
+% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
+%
+CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
+%
+\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
+However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
+
+The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
+%
+\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
+In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
+}
+%
+Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
+%impact that using a game engine for AV simulation may have.
+% on verification tasks. 
+In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
+%
+\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
+In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
+}
+%
+% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
+%We present a case study to illustrate the use of this method.
+The study highlights the effects of non-determinism of a game engine on simulation results.
+Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
+%
+%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
+%
+
+This paper is structured as follows.
+%
+Section~\ref{s:prelim} defines terms used throughout the paper and identifies when determinism is needed. 
+Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
+%
+Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
+%
+Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
+%
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
+
 \todo[inline,color=green!40]{GC New below}
 
+\section{Preliminaries} \label{s:prelim}
+
 \subsection{Definitions}
 
-A list of definitions are given here which are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
+Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
 \\
@@
 % \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-\textit{Variance} is used here to define the spread, or distribution, of the generated output data with respect to the mean value.
+\textit{Variance} is used here to define the spread or distribution, of the generated output data with respect to the mean value.
 \textit{Precision} is synonymous with \textit{variance} although inversely related mathematically.
 %
-Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken
+Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed under the same conditions and actions.
 %
@@
 %
-Real manufacturing fails to achieve absolute precision and hence the need for tolerances on design specifications to account for the variance in real-world processes.
-
-If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}b, and hence will be within acceptable tolerance.
-If the simulator is non-deterministic then there will be a variance in the output which can be measured.\\
+Real manufacturing fails to achieve absolute precision. Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.
+
+If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within tolerance by design.
+If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
 %
@@
 \subsubsection{Accuracy}
-Precision and tolerance should not to be confused with \textit{accuracy} which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
-We therefore define accuracy as the difference between the true value, or reference value, and what has been achieved in the simulation or generation process.
-%
-For a driving simulation the reference value 
+Precision and tolerance should not be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
+Therefore, we define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation.
+%
+For a driving simulation, the reference value 
 % , for which accuracy can be measured against, 
 may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
-In most cases such accuracy will not be possible due to computational demands of such an exact replica and in most cases is unnecessary, where some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
+%
+In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
+In most cases it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
 \\
@@
 \subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) at regular time intervals.
+A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) in a 3D environment recorded at regular time intervals.
 This definition could be extended to include other variables.
 %, but position is sufficient for this analysis.\\
-A set of simulation traces derived from the same input then forms the experimental data on which variance is calculated. \\
+A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run. \\
 
 \subsubsection{Simulation Variance \& Deviation}
-If the simulator is non-deterministic then how can this variance be measured? This can be achieved by monitoring any simulated output variables that should be consistent from run to run. 
+If the simulator is non-deterministic then how can the simulation variance be measured?
+This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run.
 % In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
-Actor path variance, derived from the simulation trace, is chosen over other variables as this distance based metric forms the basis for many downstream verification tests.% such as assertion checking.
-% 
-Therefore the term \textit{simulation variance} refers to a measure of actor position variance in the simulation with respect to time assuming fixed actions. 
-%
-\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive value to comprehend when considering interpretation of verification tests. 
-%Results are referred to in terms of deviation, the square root of variance, rather than variance, as this is more intuitive to comprehend when considering verification tests, e.g.
+For example, actor position variance is a distance-based metric that can be derived from simulation traces.
+The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained.
+% 
+Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions.
+% {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
+%
+%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
+%
+Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
+%
+% more intuitive to comprehend when considering verification tests, e.g.
 vehicle close-passing.
@@
 
-If the simulation is non-deterministic and has a non-permissible variance in, for example, actor positions, this may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.
+If the simulation is non-deterministic, e.g.\ 
+% KIE No, whether the variance is or is not permissible is not what matters for non-det, it should be zero!
+%e.g. it has a non-permissible variance in,
+it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
+In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
 %
@@
 %
-How can one assess the extent to which a simulation environment is deterministic or has a permissible level of variance?
+{\em How can one assess whether a simulation environment is deterministic?} and 
+{\em How can one determine and control the simulation variance?}
 \todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
-
-% In pursuit of an answer to this question we
-In this paper we investigated how the non-determinism of Carla, based on the Unreal game engine, when used for simulation-based AV verification affects the simulation results. 
-%
-In our case study, scenarios between pedestrian and vehicle actors have been analysed to identify conditions that result in non-deterministic simulation output by analysis of actor position variance.
-%
-\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
-}
-%
-By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
-%
-Carla exhibits a permissible level of variance when system utilisation is restricted to 75\% or less and ensuring termination of scenarios once a vehicle collision has been detected.
-%
-\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
-
-The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
-%
-\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
-In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
-}
-%
-Knowing the simulation variance will help assess the impact that using a game engine for AV simulation may have on verification tasks.
-In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact on verification results.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
-In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
-}
-%
-% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
-%We present a case study to illustrate the use of this method.
-The study highlights the effects of non-determinism of a game engine on simulation results.
-Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
-%
-%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
-%
-
-This paper is structured as follows.
-%
-Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
-%
-Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
-%
-Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
-%
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
 
@@
 
-There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g.
-Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
+There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
 % \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
@@
 %
-Ensuring software operates deterministically is a non-trivial task and catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
+Ensuring software operates deterministically is a non-trivial task.
+Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
 %
 \todo[inline]{GC line added to introduction.
-We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
+We need a paragraph on how hard it can be to make a computation fully deterministic, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
 Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
@@
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-The gamer neither wishes a faithful representation of reality, i.e.
-the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance, i.e.
-can be low \textit{precision}.
-%
-In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
+% 
+The gamer neither needs a faithful representation of reality (i.e.
+gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
+gamer accepts low precision).
+In contrast, high accuracy and precision are necessary for AV development, testing and verification.
+% The gamer neither needs a faithful representation of reality, i.e.\ the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance.
+%, i.e.\ can be low \textit{precision}.
+% %
+% In contrast, the precision required for AV development and testing is very high, especially for perception stack testing, but also, more generally, for obtaining reliable verification results and for visual inspection of accidents or near misses.
+If necessary, it is acceptable to achieve the required tolerance %on the precision
+% at the cost of real-time performance.
 For example, for perception stack testing the sensors need to get input that is repeatable so that if any software bugs are found they can be re-played and the issue resolved.
@@
 %
-%
-% SEE BELOW \todo[inline]{Check the points on floating point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
-Calculation results may differ due to the nature of non-associative floating point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-Some authors even suggest avoiding the use of floating point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
+
+% SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
+%{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
+%and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }
+
+% As a consequence, floating-point arithmetic is not associative.
+Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, a change of execution order, e.g.\ from $x+(y+z)$ to $(x+y)+z$, can lead to different results~\cite{Kapre2007}.
+%
+As a consequence, floating-point arithmetic is not associative.
+Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}.
+This is because rounding is being performed at different stages in the computation, depending on the respective execution order.
+%
+Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
+%
+Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}. 
+%
+%{\color{red} Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.}
 %
@@
 %
-Thus, one could jump to the conclusion that floating-point errors cause non-deterministic behaviour, resulting in loss of repeatability.
-
-However, these precision issues are better described as \textit{incorrectness}, rather than them being a source of non-determinism. 
-%
-It is normally reasonable to assume that the processors on which game engines run are designed to be deterministic with respect to floating-point arithmetic. 
-%
-Thus, when a rounding error occurs, then the same processor should produce the same \textit{incorrect} result given the same operands. 
-%
-So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always 
-be equally \textit{incorrect} as long as the implementation conforms to the IEEE floating-point standard~\cite{8766229}.
-
-Beyond hardware limitations, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
+Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability.
+In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
+
+In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
+%{\color{red} Should this be IMPRECISION rather than incorrectness?}
+%
+It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
+%
+Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern.
+%, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
+%
+So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
+
+However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
+Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
+
+Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
 However, the same executable should still return the same output for identical input.
-%
-Therefore, it can be concluded that floating-point arithmetic does not contribute to \textit{simulation variance} for repeated tests using the same executable on the same hardware with the same configuration.
+
+In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.
 
@@
 
-Indeed, the developers of the de-bugging program \texttt{RR}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
-%
-\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+This allowed control over the scheduling and execution order of threads promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
+%
+\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{rr} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
@@
 
-\subsection{NUMA Non-Uniform Memory Access}
+\subsection{Non-Uniform Memory Access (NUMA)} \label{s:sources_numa}
 For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
@@
 %
-\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
-To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.
+\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
+To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
 \todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
 %
-NUMA should only affect interprocessor data transfer cost and without control measures will only make the computation cycle longer.
+Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
 Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
+If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.
+
+% {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
 %
@@
 Can't control library change, perception stack
+
+% ***************************************************
+%  Empirical Investigation
+% ***************************************************
+\section{Case Study of Simulation Variance} \label{s:case-study}
+
+We present an empirical investigation into using game engines for
+simulation-based verification of autonomous vehicles with a focus on
+characterising sources of non-determinism in order to understand  the impact
+they have on simulation variance. 
+%
+\todo[inline,color=green!40]{GC new: line below.}
+%
+Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
+%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
+%
+Ultimately, our objective is to control non-determinism to minimise simulation variance.
+
+\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
+%
+We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
+A discussion on the internal and external settings of the simulation is included, along with system configuration and pre-screening sections.
+% {\color{red} and then..? Please cover the entire section - consider splitting setup from evaluation.}
+%
+\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
+%
+\todo[inline]{Complete the section structure as this is a very long section.}
+
+% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
+Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
+
+% ***************************************************
+%  Scenario Description
+% ***************************************************
+\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
+\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
+
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \caption{}
+        \label{Test_a}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \caption{}
+        \label{Test_b}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
+\end{figure}
+
+This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
+%
+Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
+
+This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
+%
+The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
+\todo[inline]{Decide past or present tense and be consistent throughout.}
+
+\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
+
+\subsection{Tolerable Simulation Variance}\label{s:tolerance}
+
+\todo[inline]{Terminology:\\
+{\bf Accuracy} is the degree of closeness/conformance to a known value.
+\\
+{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
+\\
+{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
+\\
+See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
+}
+
+% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
+
+To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
+%
+Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
+%
+% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
+%
+The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small such that assertion checking would fail with minor computational perturbations.
+% {\color{red} {\sc What would fail?}}
+%
+Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
+
+% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
+%
+% but a larger value may be acceptable for non-urban environments, e.g.
+motorway or highway.
+Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
+$1\mu m$.
+% 
+% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
+
+% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
+%
+For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
+%situations such as potentially dangerous close passing of a cyclist. 
+A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
+% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
+Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected. 
+% Thus any variation of less than $1$cm is tolerable. 
+Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
+
+% considered to be sufficient for assertion checking in urban environments. 
+%
+% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
+
+% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
+Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
+
+\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
+}
+
+% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
+In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
+%
+\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
+where any value higher than the specified tolerance is considered non-permissible. }
+
+% ***************************************************
+%  Actor Collisions
+% ***************************************************
 
@@
 
-% ***************************************************
-%  Empirical Investigation
-% ***************************************************
-\section{Case Study of Simulation Variance} \label{s:case-study}
-
-We present an empirical investigation into using game engines for
-simulation-based verification of autonomous vehicles with a focus on
-characterising sources of non-determinism in order to understand  the impact
-they have on simulation variance. 
-%
-\todo[inline,color=green!40]{GC new: line below.}
-%
-Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
-%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
-%
-Ultimately, our objective is to control non-determinisim to minimise simulation variance.
-
-\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
-%
-We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-%
-\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
-%
-\todo[inline]{Complete the section structure as this is a very long section.}
-
-% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
-\end{figure}
-
-This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-%
-Thus, the scene for our investigation is the T-junction (T-intersection) and the scenarios are shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor trajectories over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-%
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-\todo[inline]{Decide past or present tense and be consistent throughout.}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-\subsection{Tolerable Simulation Variance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
-
-To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-%
-Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
-%
-% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
-%
-This tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small as to fail with minor computational perturbations.
-%
-Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
-
-% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-%
-% but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-% 
-% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
-%
-For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
-%situations such as potentially dangerous close passing of a cyclist. 
-A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
-% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study and more broadly our general verification requirements, a tolerance of $1$cm has been selected. 
-% Thus any variation of less than $1$cm is tolerable. 
-Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-
-% considered to be sufficient for assertion checking in urban environments. 
-%
-% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
-Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-%
-\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
-where any value higher than the specified tolerance is considered non-permissible. }
-
-% ***************************************************
-%  Actor Collisions
-% ***************************************************
-
 \subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}.
-%
-Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.
-
-The tests are listed in Table~\ref{TableOfExperiments} and were chosen to cover a range of interactions between actor types. 
-%
-Tests 1 \& 2 involve 2 vehicles meeting at a junction where they do not collide (test 1) and when they do tiggering an \textit{event hit} callback in the game engine (test 2).
+Previous investigations into the Unreal Engine indicated that collisions between actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}.
+%
+Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were identified as potentially key aspects to the investigation into simulation variance.
+
+The tests used for the case study are listed in Table~\ref{TableOfExperiments}. They cover a range of interactions between actor types. 
+%
+Tests 1 \& 2 involve two vehicles meeting at a junction where they either do not collide (Test 1) and where they do collide (Test 2), thereby triggering an \textit{event hit} callback in the game engine.
 In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
@@
 % ***************************************************
-\subsection{Experiment Description}\label{s:Experiment_Description}
+\subsection{Evaluation Metric}\label{s:Experiment_Description}
 % \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
+For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time. 
+%
+The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.
+
+Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
+%
+\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
+%
+Repeated traces are compared at the same time index $t$ for each actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
 %
 Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-}
-%
-The maximum variance over the entire set of $n$ simulations is therefore defined as the largest variance of any actor at any time in the simulation, 
+}
+
+The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
+%{\color{red} This is still a function over time - should it be?}
 \begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a^{2}(t), 
+\max_{a,t}\sigma_a^{2}(t)
 \end{equation}
-and therefore the maximum deviation is simply square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity.
-
-The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation was reliable for verification purposes.
+The maximum deviation is the absolute value of the square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity. 
+% {\color{red} So, ``a'' is any actor or a specific actor?
+Likewise, ``t'' is any time point or a specific time point?
+NOTE that any actor any time means a in A and t in T, where A is the set of actors and T is the set of all time points.}
+
+The maximum deviation, $\max\sigma$, can be analysed for the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation is sufficiently accurate for verification purposes.
 % The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
@@
 % ***************************************************
-\subsection{Internal Settings}
+\subsection{Simulator Settings}
 
 Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
-Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.
-\textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.
-\textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
+Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.\ \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.\ \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
+Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
 % \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
@@
 
+% \subsection{Initial Results and Discussion}\label{InitialResultsSection}
+
+% {\color{red} This is for unrestricted, so this does not come as a surprise later.}
+
 % ***************************************************
@@
 
-After discovering that internal engine settings could not achieve a simulation variance that met the required 
+% After discovering that internal engine settings could not achieve a simulation variance that met the required 
 % \textit{deterministic threshold}, 
-\textit{tolerance}, 
-settings external to the game engine were explored. 
+% \textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
+After examining controls the internal game engine, external settings were explored. 
 %
@@
 For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
-This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the CARLA platform.
+
+Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load.
+This initial observation was explored more fully in this work using the CARLA platform.
+
+% {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented.
+Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}
 
@@
 % In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
-Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0-95\% (see Section \ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
+Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
 Resource utilisation figures reported here should be considered approximate values.
+%
+Resource utilisation was capped to 75\% in some parts of the results and referred to as \textit{restricted}.
+This was done to limit the load on the test system and hence limiting simulation variance.
+The term \textit{unrestricted} places no such limit on the system load.
+
+% {\color{red} To understand the impact of collisions and high resource utilisation, the restricted column shows a subset of the results where post-collision data and experiments above 75 resource utilisation have been removed}
+%
 % The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
+%
+% {\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
 
 \todo[inline,color=green!40]{GC Addition (para below):} 
-%Issues around floating point precision and concurrency as previously mentioned notwithstanding, 
+%Issues around floating-point precision and concurrency as previously mentioned notwithstanding, 
 Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
 For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
-In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision~\cite{matlab_parallel_computing}.
+In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency-induced imprecision~\cite{matlab_parallel_computing}.
 
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV5.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels. Tests 5 and 6 overlap having almost identical results.}
+    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV6.pdf}
+    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
+Tests 5 and 6 overlap having almost identical results.
+Note that the lines between data points are only a guide.}
     \label{ExperimentsStressSummary}
@@
 % ***************************************************
-\subsection{System Configuration and Screening}\label{s:screening}
-These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
-%
-Operating systems was Linux Ubuntu 18.04.4 LTS. To ensure reliability of results, each test was repeated 1000 times. 
+\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
+The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
+%
+The operating systems was Linux Ubuntu 18.04.4 LTS. 
 %
 Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
-%
-A detailed guide for reproducing these experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
+
+% Each test was repeated 1000 times. 
+Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
+While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results.
+Therefore each test was repeated 1000 times to provide a sufficient sample size.
+% {\color{red} Justify why 1000 - see my later comment in the Method section.}
+%
+A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
 %
@@
     \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}\cite{shi2009testing}.
-    \item Non-uniform memory access: \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
+    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
 \end{itemize}
@@
 % In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
+
+% ***************************************************
+%  Results and Discussion
+% ***************************************************
+\section{Results and Discussion}\label{s:FinalResultsSection}
 
@@
     \end{subfigure}
-    \caption{Vehicle to vehicle collision (test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
+    \caption{Vehicle to vehicle collision (Test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
 The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
 \end{figure*}
-
-% ***************************************************
-%  Results and Discussion
-% ***************************************************
-\subsection{Results and Discussion}\label{ResultsSection}
-A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted), where the value reported is the maximum deviation across all resource utilisation levels, i.e.
-the worst case for a given scenario.
-%
-From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the tolerance that has been set, with some deviation in actor path as large as $59$cm.
-\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-}
-%
-Clearly a deviation in results of such a large amount is unacceptable if simulation is to be used as a reliable verification tool.
-
-Resource utilisation level was found to have a significant impact on \textit{simulation variance}.
-%
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, anything above the $1$cm level (dashed line) is considered non-permissible according to our specific tolerance level informed from our verification requirements.
-%
-\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
-%
-
-A general pattern in the results indicates that; some scenarios consistently fail to produce results within permissible levels of variance at any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm level), some are consistently within tolerance (Test ID 5 \& 6 with pedestrians only), and that some cases only fail to meet the tolerance requirement when at higher utilisation levels, breaching the tolerance above 75\% resource utilisation (Test ID 1 \& 3).
-
-\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}
-
-However, the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-
-Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-}
-% \todo{Why ``not reliable''.}
 
@@
     \end{subfigure}
-    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
+    \caption{Vehicle to pedestrian collision (Test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
 \end{figure*}
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
-    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% additonal CPU \& GPU resource utilisation.}
-    \label{NICEExperimentStressSummary}
-\end{figure}
+% {\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}
+
+A summary of the main the results are shown in Table~\ref{TableOfExperiments}.
+In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
+%
+From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+%
+However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
+%
+\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
+However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
+}
+%
+Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
+
+Resource utilisation was found to have a significant impact on \textit{simulation variance}.
+%
+Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
+In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
+%
+Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
+%
+\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
+%
+
+A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
+
+Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision.
+%
+The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
+
+\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
+Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
+%
+This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
+This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
+This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
+}
+% \todo{Why ``not reliable''.}
 
 The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
+In this scenario (Test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
+There are three observations; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
 % as described above. 
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v5.pdf}
-    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v7.pdf}
+    \caption{Stages of the method proposed to determine the variance of a simulation.}
     \label{method_diagram}
@@
 Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
+
+A correlation between resource utilisation and simulation variance has been observed in these results which may be due to process scheduling.
+Rather than limiting utilisation of the whole system, another approach may be to promote the scheduling priority of the simulation process which is explored in Section~\ref{r:process_scheduling}.
+Furthermore, we were keen to investigate the impact of memory access and include a brief investigation of this in Section~\ref{r:numa}.
 %
@@
 
-\subsection{Process Scheduling Priority}
-An investigation into the response of process scheduling on the simulation variance was undertaken.
-The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
+\subsection{Process Scheduling Priority} \label{r:process_scheduling}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
+    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
+    \label{NICEExperimentStressSummary}
+\end{figure}
+
+An investigation into the impact of process scheduling on the simulation variance was undertaken following the observations of increased simulation variance with increased resource utilisation.
+The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
 %
@@
 
-The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range, non-outlier limits by the whiskers and a horizontal bar for the median.
-The figure shows that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a NICE value of 0 (central plot bars).
-Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation but this did not account for all the difference in the observed results.
-This is demonstrated in the maximum priority setting where the bars in plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
-%
-This unaccounted difference in the variance may be due to the lack of absolute control that \texttt{NICE} has over the process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
+The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range of deviation, non-outlier limits by the whiskers and a horizontal bar for the median.
+The figure shows that decreasing the priority of the simulator process (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot).
+Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results.
+This can be seen in the maximum priority setting where the bars in the plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
+%
+The remaining difference in the variance between the two resource utilisation levels may be due to the lack of absolute control that \texttt{NICE} has over process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
+
+\subsection{Non-uniform memory access (NUMA)} \label{r:numa}
+An additional investigation into memory access was undertaken given the potential impact that data transfer cost and execution order may have on simulation variance, see Section~\ref{s:sources_numa}.
+%
+The program \texttt{numactl}~\cite{numactl_NUMA} allows a process to run with a specified memory placement policy, essentially allowing the process to be bound to a particular CPU core. %
+\texttt{numactl} was used to fix the simulator and test script to single cores, and a (2\%) improvement in simulation variance was observed. %
+This was considered minor comparative to the changes in simulation variance observed in other aspects of the case study, for example the pre- and post-collision between a vehicle and pedestrian saw a change of $10^{14}$.
+Therefore \texttt{numactl} was not used for subsequent testing.
+
+However, practitioners that are aiming to minimise simulation variance should use a technique such as this to minimise data transfer cost and potentially minimise any potential effect on process execution order that may lead to increased simulation variance.
 
 \subsection{Investigation Summary} \label{s:empirical_summary} 
-
+% {\color{red} Factor into overall conclusion of the case study}
 % \todo[inline]{GC-need to reference the 'restricted' set of results here too}
 
-\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
+%\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
 The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator.
@@
 (is this 2nd one true, or does carla ignore render requests?)}
-
-This empirical investigation has highlighted the shortcomings of using a games engine for simulation based verification and advice for best working practice.
+These empirical investigations have highlighted the shortcomings of using a games engine for simulation based verification and suggests advice for best working practice. 
+%
+It was observed that resource utilisation positively correlates with simulation variance and that specific simulation events, such as vehicle collisions, can also lead to a breach in permissible tolerance.
+%
+The investigation found that the effect of higher simulation variance as a result of increased resource utilisation can be reduced, but not omitted entirely, by controlling the scheduling policy. 
+%
+The investigation into specifying memory placement on the simulation process did improve simulation variance but only by a minor amount of 2\%.
+
 However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly. 
-Therefore we now present a general methodology that practitioners can follow to find the
+%
+Therefore we have derived a general methodology that practitioners can follow to find the 
 % \textit{deterministic domains of operation} 
 \textit{operational domains of permissible variance} 
-for any hardware configuration.
+for %any hardware configuration.
+a game-engine-based simulation environment. This methodology is presented in the next section.
 
@@
 % ***************************************************
-\section{Methodology} \label{s:methodology}
-In this section a method for determining the variance of simulated actor path trajectories, termed \textit{simulation variance}, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.
-%
-In addition a number of recommendations and best practices are suggested that other practitioners can follow to minimise simulation variance. 
+\section{Method to determine the Variance of a Simulation} \label{s:methodology}
+% In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
+This method can be used to determine the operational domains of permissible variance of a simulation environment.
+%
+In this section a method for determining the simulation variance of actor paths and resolving the operational domains of permissible variance is presented here as a work flow, see Fig.~\ref{method_diagram}.
+%
+In addition, recommendations and best practice guidelines for minimising simulation variance are suggested.
+% , such as fixing the physics time step in \textit{Simulator Settings}. 
+% {\color{red} \sc Highlight these in the text.}
 % A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
@@
 %
-\subsection{Design Experiment}\label{s:design_experiment}
+
+The method consists of a sequence of five stages; experimental design, simulator settings, external settings, execution and analysis.
+In the following, each stage is described in detail with reference to the items listed for each stage in Fig.~\ref{method_diagram}.
+
+\subsection{Experimental Design}\label{s:design_experiment}
+
+% {\color{red} \sc Surely, a series of systematically designed experiments, each repeated X times, should be set up, where each experiment varies one of the suspected sources of non-determinism?
+Then, the analysis of the data from each experiment should reveal the actual sources of non-determinism for the given setting?
+In other words, we are aiming to fix as much as we can control (easily), and then we intentionally vary (ideally) single variables, e.g.\ the resource utilisation, to find out what makes a difference.
+We need a clear definition of the objectives of the experimental design, i.e.\ what one should vary and what one wants to ``nail down''.}
+
+% {\color{green!99} \sc }
+The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
+Each test is repeated $n$ times and the analysis of the results provide a confidence with a degree of statistical certainty.
+By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and addressed.
 
@@
 
-Different actor types may be handled differently by the game engine and our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
-However the introduction of CARLA vehicles increased the simulation variance of the pedestrian-only test by 12 orders of magnitude.
-As such, all actors that could be included in any future simulation should be tested, including any non-standard CARLA or bespoke actors.
+% Different actor types may be handled differently by the game engine.
+For instance, our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
+However, the introduction of CARLA vehicles increased the simulation variance of the pedestrian-vehicle test by 12 orders of magnitude.
+{\color{red} \sc This needs rephrasing as it appears to contradict the statement above.}
+%
+All actors that could be included in the simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see Section~\ref{ego_actor}. 
+% {\color{red} \sc ... such as the ego vehicle...? [see section on ``External Settings'' last point - Ego Controller.]}
 %
@@
 These paths should include collisions between actors and potentially collisions between actors and static scenery if this is likely to occur in the simulation or as part of the verification process.
-Actor paths without collision are also important to include as these will serve as a baseline to the other tests.
-
-\subsubsection{Setting a Tolerance} \label{s:threshold}
-The verification engineer should set the \textit{tolerance} based on their specific requirements which refers to the acceptable level of actor path variance.
+Actor paths without collision are also important to include as these will serve as a baseline to the other tests; to see if deviations between runs in actor paths increases with collisions.
+% {\color{red} \sc What baseline?} {\color{green} \sc Done}
+
+\subsubsection{Permissible Variance} \label{s:threshold}
+The verification engineer should set the \textit{permissible variance}, also termed \textit{tolerance}.
+% which refers to the acceptable level of actor path variance. 
 % ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. 
-This tolerance level is analogous to the analogue signal level associated with each of the binary states %of the signals 
-in digital circuits.
+%
+The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
+For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection.
+In Section~\ref{s:tolerance} a tolerance of $1$cm was considered sufficient for urban scenario assertion checking.
+In practice, it may be necessary to determine this tolerance experimentally.
+
+%The permissible variance is analogous to the analogue signal level associated with each of the binary states %of the signals in digital circuits.
 At some low analogue voltage the signal is interpreted as binary $0$, this may be, for example, any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
 This convention enables the abstraction from the noisy physical signal values that are analogue voltages into the digital domain with binary representation.
-%
-The tolerance that the user must set depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this tolerance must be sufficiently small to enable accurate assertion checking.
-In our empirical investigation we set a tolerance of $1$cm which is sufficient for urban scenario assertion checking.
-In practice, it may be necessary to determine this tolerance experimentally.
+<- KIE: meant only to illustrate that other areas also have defined tolerances.
+%
 
@@
 
-\subsubsection{Simulation Length} 
-The simulation time should be sufficient to record an interaction between actors but not so long that the testing take an inconvenient time to complete.
-In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
+\subsubsection{Simulation Time} 
+% {\color{red} \sc Should this not be part of the Execution section?} 
+The simulation time will depend on the actor paths and terminating conditions.
+The simulation time should be sufficient to record interactions between actors but not so long that the testing take an inconvenient time to complete.
+In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25} in Section~\ref{s:FinalResultsSection}.
+The termination conditions of the simulation can be set by, for example, actors reaching their final trajectory waypoints.
+ % {\color{red} \sc Point to the section that specified that.}
 
@@
 We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
-However, fine control over the scheduling policy and thread execution order may be non-trivial.
-Such policies are defined by the operating system and may execute threads with equal priority.
-In such a case, all tasks non-essential to the simulator should be terminated to prevent interruption to the simulator.
+However, gaining fine control over the scheduling policy and thread execution order is known to not be trivial~\cite{acm-q-rr-interview}.
+
+The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority.
+Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation.
 
 % If the policy or operating system cannot be changed then s
-Setting the simulator process to a higher priority may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
+Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
 
 \subsubsection{NUMA Control}
-Control over Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multiprocessors with shared memory.
-This control allows the simulator program to be fixed on a single core reducing memory access time.
-%
-Initial screening tests done with NUMA control gave moderate improvements in simulation variance of the order of a few percent, see Section~\ref{s:screening}.
-This control may assist if simulation variance is borderline to the tolerance but not seen as essential.
+Control over a Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multi-core processors with shared memory.
+This control allows the simulator process to be fixed on a single core, reducing and unifying memory access time.
+%
+Investigations in tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}. 
+%
+% Using \texttt{numactl} allows a process to run with a specified memory placement policy, fixing the process to a single core.
+This will reduce the data transfer cost and the opportunity for disrupted execution order that may be responsible for increase simulation variance.
+%
+Using a fixed memory placement policy may assist if simulation variance is borderline to the tolerance but only of the order of a few percent from our observations.
 
 \subsubsection{Ego Vehicle Controller} 
-No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
+% {\color{red} \sc Why is this not covered under actors? [see my note in the Actors section]}
+% {\color{red} \sc Clearly separate the investigation from the Method.
+The fact that the case study did not have an ego vehicle should be mentioned as a limitation of the case study, rather than elaborated on here, but why does this make a difference?}
+%
+An ego vehicle was not used in this study, but the impact of introducing this to the simulator can be considered here.
+The ego vehicle is seen as another actor in the simulation but care must be taken to ensure that any control algorithms, machine learning modules and processing pipelines are deterministic.
+We recommended this be treated as a separate source of non-determinism and handled accordingly.
+
+% ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
+{\color{red} \sc Long and cumbersome sentence with strange grammar: No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
+}
 
@@
 % \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
-Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests but with 3 tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
-While executing 100 repeats may seem sufficient, this sample size may fail to observe these events that occur with low probability, inferring a false confidence in the results.
-It is therefore recommended that the sample size is found empirically dependent on the number of results that exceed the tolerance.
-% To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
+%
+% {\color{red} \sc Again, separate the study from the method.
+This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.}
+It is recommended that the chosen sample size is determined empirically. 
+%
+We recommend monitoring variance while increasing the sample size in orders of magnitude
+until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process. 
+%
+Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
+
+% By increasing the sample size and observing that the simulation variance is stable, one can infer that a sufficient number of samples have been collected.
+If the variance changes with a change in sample size then a stable condition has not been found.
+% ...
+% ...
+%  dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
+% % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
 
 \subsubsection{Data Logging}
-Data logging should be used to record the actor positions at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-Additional information can also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
+Unique identifiers should be assigned to each experiment, each repeat and each individual actor. 
+The time-stamped actor positions should then be recorded at fixed time intervals throughout the simulation in order to determine the variance in actor path.
+Additional information should also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
 % such as the \texttt{unit} command in Unreal Engine. 
-For the subsequent analysis, the actor should have an identifier and the repeat number and simulation time should also be captured. 
+%For the subsequent analysis, each actor should have an identifier. The repeat number and simulation time should also be captured. 
 % The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
@@
 \subsection{Analysis}
-The maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control for permissible operation.
-Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
-
-This method can be extended to a broader actor states and actions including actor orientation, speed, and any other status indicators that may be of interest and also including appropriate actions that may be useful for verification purposes.
-This broader As such, Equation~\ref{eq:max_sigma} would need to be adjusted to include these new variables.
+For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment.
+
+%Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
+% KIE: Why now point to HW? And, which "boundaries"?
 
@@
 
-The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
-%
-\todo[inline,color=red!40]{old:Having a deterministic simulator is required for reliable and repeatable results,}
-Having a deterministic simulator is required for precise results, 
-\todo[inline]{Having a deterministic simulator is required for precise results,} 
-%
-to find and fix software bugs and ensure results are trustworthy. 
-%
-If a simulator is non-deterministic then practitioners should at least be aware of, and how to find, the operational domains for reliable results.
-During an investigation into the CARLA simulator, a \textit{simulation variance} was observed for repeated tests with the same initial conditions indicating non-deterministic execution.
-%
-During the study we hypothesized and uncovered several parameters that contribute towards greater simulation variance and hence non-deterministic execution, such as actor collisions and system resource utilisation.
-%
-The results of the investigation are hardware and software version specific, so we proposed a general methodology that can be used to find the \textit{domains of permissible variance} for any given system configuration.
-%
-This methodology will allow the AV verification community using simulation tools to ensure their results are reliable and trustworthy.
-
+%The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
+Game engines offer simulation environments that are used for the development and verification of autonomous driving functions. 
+%
+Determinism of a simulator is required to achieve repeatability, which is essential to find and fix software bugs efficiently, and to ensure simulation results are trustworthy. 
+%
+If a simulator is non-deterministic then practitioners should at least be aware of, and know how to find, the operational domains where \textit{simulation variance} is tolerable.
+
+An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation.
+%
+We then researched, identified and discussed potential sources for non-determinism in this context.
+%
+A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution.
+In particular, actor collisions and system-level resource utilisation were identified as key contributors.
+%
+To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
+%
+% {\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.
+This seems to have been commented out but was here before.}
+
+A general method to assess the actor path variance of a game-engine-based simulation environment is then proposed.
+The method can be used to find the \textit{domains of permissible variance} of a simulation environment for a given system configuration.
+This can give AV developers and verification engineers increased confidence in the simulation results and reduce debug time.
+% 
 % \subsection{Future Work}
-A potential avenue of exploration is developing a version of the simulator which has been designed specifically for verification.
-In this version the simulator would be deterministic because the execution is suitably controlled along with sources of randomness and scheduling.
-A similar task was achieved with the \texttt{RR} program~\cite{RR_link}.
-
-With the advent of more AI systems becoming adaptive, meaning that the system may change it's response to the same stimulus over time, the notion of repeatability will need reassessing in the context of verification.
-The main research question here is what will constitute `the same' which will be important for passing or failing verification tests but also for determining coverage.
-
-% \todo[inline]{Review/rewrite this to match the earlier changes in the paper.
-Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
-Potential sources of non-deterministic simulation behaviour have been discussed such as floating point rounding error and scheduling policy.
-A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the CARLA driving simulator.
-Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study.
-Furthermore it was found that all actors, even those not involved directly in a collision, will behave non-deterministically if a collision callback to the game engine is made.
-Best operating practice follows that the simulation should be stopped as soon as a collision callback is detected or resource utilisation levels exceed 75\% although this value may be system specific.
-}
-
-% \todo[inline]{GC: think we should put more emphasis on the outcomes of the method rather than the cast study.}
-% \todo[inline]{Point out (again) that all data and scripts etc are on github.}
-
-% \todo[inline]{Future work...}
-
-% In this paper we explored the determinisim of game engines.
-This is important for autonomous vehicles' development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, uncovered several parameters that contributed to simulation variance.
-These parameters include numerous internal settings based in Unreal Engine, and others that are external mainly dependant on CPU and GPU utilisation.
-
-% Changes in CPU and GPU resource utlisation has shown to be the main influencer for the non-deterministic behaviour of game engines.
-This is clearly demonstrated in the results (see section \ref{s:case-study}), where the max deviation between runs increased from the orders of $10^{-12}$m to devaitons of up to 1m as the CPU and GPU resource utilisation increased above 50\%.
-
-% The term \textit{deterministic operational domain} was given to the range of system resource utlisations that provides low simulation variances that is insignificant for development and verifcation of autonomous vehicles.
-% The deterministic operational domain specified herein is specific to our hardware configuration.
-
-% This led to the design of a methodology allowing `users' to determine the deterministic operational domain for a specific hardware configuration.
-This will help guide specialists in the field of AVs to setup their custom experiments for their specific simulators and identify their specific deterministic operational domains.
-
-% Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
-Hence allowing its extensive usgae by both the autonmous vehicles' and the gaming community.
-
-% presented an investigation (based on CARLA driving simulator) to show whether game engines are deterministic or not. 
-% Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, was used to illustrate the challenges of determinisim faced by game engines. 
-% % It was found that game engines are non-deterministic.
-
-% We reviewed possible sources of non-determinisim in section \ref{s:background} and investigated the effect of each on the simulation variance.
-
-% We found that the sources of non determinsim that had a significant effect on simulation variance were attributed to CPU and GPU resource utilisation. 
-% The case study clearly shows how the max simulation variance changes from very low simulation varicances, below picometer level ($10^-12$), to centimeter level when the CPU and GPU resource utilisation increases above 50\%.
-
-% We found two main outcomes from our investigation: 
-% 1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}, e.g. runtime scheduling. 
-% 2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.
-% Thus, by altering these settings one can achieve sufficiently low levels of simulation varince that can allow users to assume the simulation is deterministic. 
-% We phrased the level below which the simulation is to be assumed determinisitic as the \textit{determinisitic threshold}, which is up to the user to define where it exactly lies. 
-% This allows for one to use these game engines knowing that their simulation will be sufficiently deterministic provided that they bound their simulation within a certain configuration.
-
-% The configuration presented in here is specific to the hardware and software used in our investigation. 
-% Therefore, we outlined a methodology in this paper to allow practitioners in this field of AVs to resolve the non-deterministic operational domains of their simulation.
-
-% We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the simulation variance of a game engine used in V\&V application of AV simulations.
-The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
-
-% A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
-It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
-%since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
-
-% **** These results are not meant as a universal guide but merely to illustrate our method for a single hardware configuration.
-It is probable that the values quoted here will not translate to individual end-users unless the system configuration is very similar.
-This paper should be used as a method for users to determine what scenarios and system setting are appropriate for their particular verification needs.
-
-% You must monitor all collisions, not just for the ego vehicle as any collision callback may result in high simulation variance and hence non-deterministic results for your experiments.
-
-% In highlight of obtaining a fully deterministic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
-Alternatively, another approach would be to create a new engine that is designed to be deterministic with minimal attention given to performance.
-
-% ***************************************************
-%  FUNDING, REFS
-% ***************************************************
-% \section*{Acknowledgements}
-% This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
-% projects are part-funded by the Centre for Connected and Autonomous
-% Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
-% 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-\printbibliography
-
+%
+As future work the method can be extended to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
+
+An ambitious avenue for future work is the development of a deterministic simulator for AV development and verification.
+This requires controlling all potential sources of non-determinism, including randomness and scheduling, very similar to the development of the record-and-reply debugger \texttt{rr}~\cite{RR_link}, originally developed to catch low-frequency non-deterministically failing tests at Mozilla~\cite{acm-q-rr-interview}.
+
+\section*{Acknowledgement}
+This research has in part been funded by the ROBOPILOT and CAPRI projects.
+Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
+%
+\textit{Abanoub Ghobrial and Greg Chance contributed equally to this paper.}
+\bibliography{cav_determinism.bib}
+
+% \end{multicols}
 \end{document}
+
+\grid

### 2021-03-25 paper/cav_determinisim_main.tex
authors names

@@
 
-\author{Greg Chance, Abanoub Ghobrial, Kevin McAreavey, Severin Lemaignan, Tony Pipe, Kerstin Eder 
+\author{Greg Chance, Abanoub Ghobrial, Kevin McAreavey, S\'everin Lemaignan, Tony Pipe, Kerstin Eder 
 \thanks{{\footnotesize
@@
 
-Severin Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
+S\'everin Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
 and

### 2021-03-25 paper_arxiv/arxiv_cav_determinisim_main.tex
updating text, authorship

@@
 
-\author{Greg Chance\inst{1,3}, Abanoub Ghobrial\inst{1,3}, Kevin McAreavey\inst{1,3}, Severin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}
-
-\authorrunning{A. Ghobrial et al.}
+\author{Greg Chance\inst{1,3}, Abanoub Ghobrial\inst{1,3}, Kevin McAreavey\inst{1,3}, S\'everin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}
+
+\authorrunning{G. Chance, A. Ghobrial et al.}
 \institute{University of Bristol, Bristol, UK \and University of the West of England, Bristol, UK \and Bristol Robotics Laboratory, Bristol, UK}

### 2021-03-25 paper/cav_determinisim_main.tex
assertion lower limit

@@
 The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection.
+For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection, but not so small for assertion results to differ for repeated runs.
 In Section~\ref{s:tolerance} a tolerance of $1$cm was considered sufficient for urban scenario assertion checking.
 In practice, it may be necessary to determine this tolerance experimentally.
-
+%
 %The permissible variance is analogous to the analogue signal level associated with each of the binary states %of the signals in digital circuits.

### 2021-03-25 paper_arxiv/arxiv_cav_determinisim_main.tex
updating text, authorship
assertion lower limit

@@
+\RequirePackage{etex} %% **************** add this for ARXIV submission
 \documentclass[runningheads,twocolumn,a4paper,10pt]{llncs}
+% \documentclass[runningheads,a4paper]{llncs}
 \usepackage{etex}
@@
 \pdfoutput=1
-
 \setcounter{tocdepth}{3}
-
 \usepackage[a4paper, total={7in, 10in}]{geometry}
@@
 \usepackage{makecell}
-\usepackage{textcomp }
 \usepackage{multirow}
@@
 \usepackage{graphicx}
-\usepackage{colortbl}
-\usepackage{booktabs}
 \usepackage{amsmath}
@@
 \usepackage[para,online,flushleft]{threeparttable}
-\usepackage{multirow}
 \usepackage{wasysym}
@@
 % \usepackage{subfig}
-\usepackage{subcaption}
 \usepackage{breakcites}
-
 \usepackage{array}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
 \usepackage{color}
-\usepackage{threeparttable}
 \usepackage{balance} %balance columns
 \hyphenation{op-tical net-works semi-conduc-tor}
-\usepackage{enumitem}
-\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
-% \usepackage{subfig}
-\usepackage{subcaption}
 \usepackage{dblfloatfix}
-\usepackage{booktabs}
 \usepackage{authblk}
-% \pagenumbering{gobble} %suppresses page numbers
-\usepackage{balance} %balance columns
-% \usepackage{todonotes}
 \usepackage[disable]{todonotes}
 \usepackage{siunitx}
+\usepackage{xurl} %for long URLs
 
@@
 The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection.
+For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection, but not so small for assertion results to differ for repeated runs.
 In Section~\ref{s:tolerance} a tolerance of $1$cm was considered sufficient for urban scenario assertion checking.

### 2021-03-25 paper/cav_determinisim_main.tex
added changes to cuda, footnote author contrib, email addresses

@@
     \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
+    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, no failures on all 11 tests~\cite{shi2009testing}.
 \end{itemize}
@@
 
-A summary of the main the results are shown in Table~\ref{TableOfExperiments}.
+A summary of the main results are shown in Table~\ref{TableOfExperiments}.
 In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.

### 2021-03-25 paper_arxiv/arxiv_cav_determinisim_main.tex
the main the
added changes to cuda, footnote author contrib, email addresses

@@
 \tocauthor{Authors' Instructions}
+
 \maketitle
+\let\thefootnote\relax\footnotetext{\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper and are both corresponding authors.}
+Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
+Abanoub Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
+Kevin McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
+and 
+Kerstin Eder (e-mail: kerstin.eder@bristol.ac.uk) 
+are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom. 
+%
+S\'everin Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
+and
+Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
+are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom. 
+}.
 

### 2021-03-29 paper/cav_determinisim_main.tex
Added CHANCE to headnote on the IEEE paper's version
modified method diagram

@@
 ..., NO.
-..., date...}{Ghobrial \MakeLowercase{\textit{et al.}}: On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
+..., date...}{Chance, Ghobrial \MakeLowercase{\textit{et al.}}: On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
 %
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v7.pdf}
+    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v8.pdf}
     \caption{Stages of the method proposed to determine the variance of a simulation.}

### 2021-04-07 paper_arxiv/arxiv_cav_determinisim_main.tex
removing comments from paper and tidy dir

@@
 \usepackage{ comment }
-% \usepackage[color=yellow!60,textsize=footnotesize,obeyDraft,draft]{todonotes}
+\usepackage[color=yellow!60,textsize=footnotesize,obeyDraft,draft]{todonotes}
 \usepackage{colortbl}
@@
 \usepackage{authblk}
-\usepackage[disable]{todonotes}
+% \usepackage[disable]{todonotes}
 \usepackage{siunitx}
@@
 
-% Give more marginpar
-\setlength{\marginparwidth}{25mm}
-
-% Prompt people to generate graphs
-\newcommand{\includegraphicsmaybe}[2]{
-    \IfFileExists{#2}{\includegraphics[#1]{#2}}{
-    \detokenize{File #2 is missing, maybe you need to run plots.py?}
-}}
-
-\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
+% % Prompt people to generate graphs
+% \newcommand{\includegraphicsmaybe}[2]{
+%     \IfFileExists{#2}{\includegraphics[#1]{#2}}{
+%     \detokenize{File #2 is missing, maybe you need to run plots.py?}
+% }}
+
+% \newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
 
@@
 
-\author{Greg Chance\inst{1,3}, Abanoub Ghobrial\inst{1,3}, Kevin McAreavey\inst{1,3}, S\'everin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,2}}
+\author{Greg Chance\inst{1,3}, Abanoub Ghobrial\inst{1,3}, Kevin McAreavey\inst{1,3}, S\'everin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,3}}
 
@@
 Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom. 
-}.
-
+are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom. }.
 \makeatletter
@@
 \section{Introduction} \label{s:introduction}
-% What is verification
-% Can't use this again as this is how the AI Test paper starts. -> Avoid self-plagiarism!
-%\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. 
-%
-% Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, benefiting from complete control over (virtual) actors and their environment
-% , offering a relatively inexpensive means to explore test cases while also
-% \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. 
 Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
@@
 
-% The use of game engines for AV testing
 Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
@@
 
-% Requirements of simulation based verification SBV
 State-of-the-art game engines provide a convenient option for simulation-based testing.
@@
 %
-%\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
-% 
-% GC Removing the following para:
-%For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
-CARLA currently offers various basic types of cameras and %ray-case based
-% lidar; extensions are still under development. 
-% \todo{GC: what is ray-case? KIE: Not sure. :(}
-%
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. %~\cite{Javaid2013,cyres}
-%
-% The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
-%
-% Game engines for SBV
+Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. 
+%
 Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
@@
 
-% \subsection{Contribution}
-% In pursuit of an answer to this question we
 In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
@@
 In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs.
-% KIE: This is covered in the next sentence:
-% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
-%
-\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
-}
-%
-% By analysing actor position variance 
 We find that the CARLA simulator is non-deterministic.
@@
 We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
-% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
-% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
-%
 CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
 %
-\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
 
@@
 %
-\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
-In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
-}
-%
 Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
-%impact that using a game engine for AV simulation may have.
-% on verification tasks. 
 In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
-In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
-}
-%
-% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
-%We present a case study to illustrate the use of this method.
-The study highlights the effects of non-determinism of a game engine on simulation results.
-Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
-%
-%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
 %
@@
 %
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
-
-\todo[inline,color=green!40]{GC New below}
+We conclude in Section~\ref{s:conclusion} and give an outlook on future work.
 
 \section{Preliminaries} \label{s:prelim}
-
 \subsection{Definitions}
-
 Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
@@
 \subsubsection{Determinism}
-% What is a deterministic system
 Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
@@
 Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
-% 
-% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
-Specifically, we focus on the variance of actor paths within a given scenario.
 
 A simulation can be thought of as the process of generating or producing experimental data.
-%
 In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data.
-%
 If a simulation is deterministic, Fig.~\ref{variance_description}~(b), then there will be no variation in the generated output data, i.e.
 all future states are perfectly reproducible from prior states and actions.
-%
-However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data. \\
+However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data.
+\\
 
@@
 
-% \subsection{Definitions: Precision, Variance \& Tolerance}
 \subsubsection{Variance, Precision \& Tolerance}
-
 We adopt terminology from the mechanical engineering and statistics domains to describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
-%
-% \textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
-% %
-% \textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-%
-% \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
 \textit{Variance} is used here to define the spread or distribution, of the generated output data with respect to the mean value.
 \textit{Precision} is synonymous with \textit{variance} although inversely related mathematically.
-%
 Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed under the same conditions and actions.
-%
 \textit{Tolerance} is defined as the permissible limit of the variance, or in short the \textit{permissible variance}.
 
-% The engineering terms \textit{precision} and \textit{tolerance} suitably describe how the simulation data vary when the same repeated stimulus is provided to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}.
-%
-%
-% As an analogy, the simulator can be thought of as a manufacturing process, producing data and if tasked to repeatedly execute the same process we must ask how \textit{precise} can this process be and what, if any, will the \textit{variance} be?
-% when comparing the differences between multiple products.
 As an analogy, the simulator can be thought of as a manufacturing process that produces data.
 To determine the precision of this process, the output must be measured and analysed for differences when the process is repeated.
-%
 Those differences describe the spread or variance in the process output.
 A hard limit on the variance can then be defined, Fig.~\ref{variance_description}~(a), beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
-%
-% In relation to manufacturing, the precision of any computer based task is casually assumed to be absolute, for example one may question the manufactured precision of a physical component but not the computer assisted drawing that it was made from.
-%but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point. %\textbf{[ref]}. %
-%
-Real manufacturing fails to achieve absolute precision. Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.
+Real manufacturing fails to achieve absolute precision.
+Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.
 
@@
 %
-For a driving simulation, the reference value 
-% , for which accuracy can be measured against, 
-may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
+For a driving simulation, the reference value may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
 %
@@
 This definition could be extended to include other variables.
-%, but position is sufficient for this analysis.\\
-A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run. \\
+A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run.
+\\
 
@@
 This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run.
-% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
 For example, actor position variance is a distance-based metric that can be derived from simulation traces.
@@
 Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions.
-% {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
-%
-%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
-%
 Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
-%
-% more intuitive to comprehend when considering verification tests, e.g.
-vehicle close-passing.
-Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
 \\
@@
 
-% \todo[inline]{values - Does this mean stimulus or other parameters? GC - changed to parameters}
-
-% Verification and Determinism 
-% Why is determinism important for verification
-%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
-% KE Text
-\todo[inline,color=red!40]{old: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
-
 \subsection{When is Determinism needed?}
@@
 A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
-%
 Thus, a deterministic simulator has zero \textit{variance}.
@@
 A simulator with non-zero variance can no longer be considered deterministic.
-%
 Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}.
-%
 Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations.
 When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-%
 This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
 
-% GC Text
-\todo[inline,color=green!40]{GC Alternative: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}.
-If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.}
-
-% This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
-The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
-\textit{Tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
-%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
-Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour.
-%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
-%
-
-If the simulation is non-deterministic, e.g.\ 
-% KIE No, whether the variance is or is not permissible is not what matters for non-det, it should be zero!
-%e.g. it has a non-permissible variance in,
-it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
+If the simulation is non-deterministic, e.g.\ it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
 In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
-%
-
-% \todo[inline]{GC Edit: If the simulation is non-deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
-
-% Challenges of SBV
-When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs % related to non-determinism
-a high priority for games developers.
+
+When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
 It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
@@
 Thus, our main research questions are:
-%
-\todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
-%
 {\em How can one assess whether a simulation environment is deterministic?} and 
 {\em How can one determine and control the simulation variance?}
-\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
-
-% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (AVs) on roads is a delicate if not utterly fascinating subject.
-Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.
-
-% Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.
-
-% It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
-
-% \noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
-
-% As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
-
-% ======================
-% This is all from "FrameworkAndChallenges" without using quotes 
-% \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system.
-Similarly, the use of simulation for traffic modelling is well established.
-Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate AVs\cite{FrameworkAndChallenges}.\\\\
-
-% While traffic-level simulators can model road layouts, they do not require as much realism or detail as AVs require.
-On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for AV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
-% Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
-% Consequently, one of the major routes AVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate AVs \cite{FrameworkAndChallenges}.
-% ======================
-
-% Robopilot, Capri, CARLA, Apollo, Airsim, Udacity etc.
-are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}.
-\\\\
-% Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
-% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of AV software.\\\\
-% \TODO{Remove this paragraph?} Determinism of game (physics) engines has not attracted much attention by engines' developers nor researchers.
-This previously was not important since in gaming, determinism is not critical.
-% Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
-% On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
-
-% This paper aims to investigate how deterministic physics engines are for usage in V\&V for AV simulations; and will cover the list of following points:
-% \begin{itemize}[leftmargin=*]
-%     \item Necessary background on game engines used in AV simulations.
-%     \item Review of sources of non-determinism in such engines.
-%     \item Provide a method of how the level of non-determinism can be investigated and defined.
-%     \item A case study showing the implementation of this methodology in an engine widely used in AVs simulations (Unreal Engine).
-% \end{itemize}
 
@@
 Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
-% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
 
 Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.
-%However, when these engines are used for AV development then deterministic behaviour is required.
-%
 Ensuring software operates deterministically is a non-trivial task.
 Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-%
-\todo[inline]{GC line added to introduction.
-We need a paragraph on how hard it can be to make a computation fully deterministic, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-%
-\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-}
-%
 This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
-%
-% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations.
-Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
 
 % ======= Overview of a Game Loop
-% \subsection{Game Engine Overview} \label{GameLoopSection}
-
-Central to a game engine are the main game logic, the artificial intelligence (AI) component, the audio engine, and the physics and rendering engines. For AV simulation, we focus on the latter two.
-%
+Central to a game engine are the main game logic, the artificial intelligence (AI) component, the audio engine, and the physics and rendering engines.
+For AV simulation, we focus on the latter two.
 The game loop is responsible for the interaction between the physics and rendering engines.
 Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed~\cite{unity_event_execution}.
-% \footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}. 
 A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering)~\cite{GameEngineArchBook}.
@@
 The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}.
-% \footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}. 
 When the frame is rendered the game loop cycle returns to processing inputs.
@@
 
-% \todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}.
-When the frame is rendered the game loop cycle returns to processing inputs.}
-%
-% Then the user inputs are processed.
-There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
-Rendering occurs once the game clock has caught up with the real world clock.
-The process then repeats.
-
-The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
+The physics engine operates according to a time step, $dt$.
 The shorter this time step is, the smoother the interpretation of the physical dynamics will be.
-%processing time it takes to catch up to game clock.
-% and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
-%
-% Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
-%
 To use a fixed physics time step, the user's display refresh rate needs to be known in advance.
@@
 Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}.
-%
 If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics.
-%
 Thus, extrapolation between frames may need to be performed to smooth transition between scenes.
-%
 Note that both residual lag and extrapolation could affect perception stack testing.
-% This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
-% \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
-%
-% \todo[inline]{Why is this part of the figure?
-It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.
-GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
-%
-% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
-In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines,
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
-% \todo[inline]{Explain how this lag may affect perception stack testing}
-
-% \todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?
-GC-suggest remove this part of the diagram, too confusing as there are not really any clocks or ticks - we should remove all mention of ticks too.}
-
-% The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
-If $dt$ is too large or irregularly spaced then the rendered motion may not be smooth and interpolation may be required at additional computational cost.
-
-% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines.
-These engines update at fixed intervals and render whenever they can, which is not steady.
-This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates.
-In this case, the engine will use extrapolation techniques to give an estimate of where the object should be.
-This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
-%
-\todo[inline,color=red!40]{old: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy.
-In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-}
-%
+In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
+
 Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
@@
 In contrast, high accuracy and precision are necessary for AV development, testing and verification.
-% The gamer neither needs a faithful representation of reality, i.e.\ the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance.
-%, i.e.\ can be low \textit{precision}.
-% %
-% In contrast, the precision required for AV development and testing is very high, especially for perception stack testing, but also, more generally, for obtaining reliable verification results and for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required tolerance %on the precision
-% at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is repeatable so that if any software bugs are found they can be re-played and the issue resolved.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-%
-\todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-The gamer neither wishes a faithful representation of reality, i.e.
-can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
-can be low precision.
-In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
-%
-\todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
-
-%\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
-This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer).
-For instance, see below, but I'm not sure whether this is the right place.}
 
@@
 
-% The following discussion reflects the results of our investigation into the potential sources of non-determinism that could affect game engines.
-We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
 The following review discusses the potential sources of non-determinism that were found in the literature or found as part of our investigation into game engines.
@@
 %
-\todo[inline,color=green!40]{GC Addition below}%
-%
 A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests}, although the AV simulation domain introduces it's own unique challenges that were not considered in that paper.
 
-% MOVED TO INTRO \todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-
-\todo[inline]{We should also refer to literature like:\\
-Making System User Interactive Tests Repeatable: When and What Should We Control?~\cite{when-and-what-should-we-control} and
-Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.}
-
 \medskip
-
-% \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
-The main reasons that we think are causes of these engines' non-determinism are discussed below.
-\\\\
 
@@
 
-% SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-%{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
-%and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }
-
-% As a consequence, floating-point arithmetic is not associative.
-Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, a change of execution order, e.g.\ from $x+(y+z)$ to $(x+y)+z$, can lead to different results~\cite{Kapre2007}.
-%
 As a consequence, floating-point arithmetic is not associative.
@@
 Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-%
-%{\color{red} Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.}
-%
-%If calculations are performed on a single core, then concurrency of multiple threads is not an issue.
-\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
-
-% Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
-Calculation results may differ due to the nature of non-associative floating-point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-%
-% If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
-% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.
-}
-%
-% Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.
 
@@
 In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
-% \todo[inline]{state what assertions are (used for) GC-see below} 
-%In the field of AV verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
 %
@@
 In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-%{\color{red} Should this be IMPRECISION rather than incorrectness?}
 %
@@
 Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. 
-%, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
 %
@@
 \subsection{Scheduling, Concurrency and Parallelisation}
-% \todo[inline]{is this at runtime?} %
-% info at:
-% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
-% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
-% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
-% https://en.wikipedia.org/wiki/Concurrent_computing
-% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
 Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-% \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
 %
@@
 %
-\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-}
-%
+
 Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
@@
 %
-\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{rr} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
-}
 
@@
 Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}.
-%\noindent\underline{\textit{Concurrency:}}
-% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
-%\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
-Consider also parallelism, e.g.
-in GPUs.
-Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism.
-Say which parts of a game engine could reasonably be expected to run concurrently.}
-%
-%Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
-This technique can lead to issues such as shared resource access conflicts due to mis-managed timing and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.
-% \todo[inline]{how? GC-see above.}
-% KIE: We only list the potential sources of non-det, I therefore removed this. 
-% although monitoring and detecting such events falls outside of the scope for this work. 
-%
-%\noindent\underline{\textit{Ego Vehicle Controller:}}
-%The ego vehicle controller may be a source of non-determinism if using complex control algorithms such as neural networks for visual processing and perception.
-A neural network can only be deterministic within in the bounds of the training data used to construct it.
-If the network is exposed to unseen data it may incorrectly classify legitimate scenes due to lack of training data.\todo{That alone won't cause non-determinism, i.e.
-the NN will misclassify exactly in the same way each time - not?}
-%
-%In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically, assuming no \textit{residual lag}.
-In this sense, control over weather and lighting in the game engine should be considered and controlled as well as any other dynamic elements to the environment, e.g.
-reflections from surface water.
-% these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
-% i.e.
-the utilisation of hardware is manipulated differently each time.
-This would then result in non-deterministic behaviours especially if it is combined with the floating-point problem discussed earlier.
-%
-% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
-Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available.
-This may lead to events happening out of order depending on the computational load and hence non-deterministic.
-%
-% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
-To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order.
-This is beyond the scope of our work and will not be covered in this paper.
-% \\\\
-% \noindent\underline{\textit{Physics and Rendering clocks:}}
-% \todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
-% % The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
-% Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
-This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
-% %
-% A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
-Do we not assume the same HW at this stage?
-GC- we are just listing sources at this point so should be hardware agnostic.
-also, yes, hardware is the same for us but we cannot control for resource utilisation and background processes.} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
-Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources.
-% %
-% AV specific platforms, such as CARLA, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
-% \todo[inline]{Add a summary/conclusion?}
-%\\\\
 
@@
 %
-%
-% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
-% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
-% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}~\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
-% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
-the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
+For example, in Unreal Editor the \textit{unit}~\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time.
 %
@@
 There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-% 
 Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
-%
 The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
-
 A navigation mesh is a fixed area of the environment where actors are free to navigate.
-%
 Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
-%.
-Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
-%
 However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment.
@@
 Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
-%
-% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
-%
-% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
-% \todo[inline]{unclear :( GC- re-written}
 
 \subsection{Summary}
-We have investigated the potential sources of non-determinism affecting game engines 
-and explored the impact they may have on simulation variance. 
-%
-% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable.
-}
-%
+We have investigated the potential sources of non-determinism affecting game engines and explored the impact they may have on simulation variance.
 Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
-\todo[inline]{So, do commercial HPC facilities have ECC memory?}
-\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
-%
-\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
 To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
-\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
-%
 Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
@@
 
-% {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
-%
 Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups.
-%
 However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop.
 Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism.
-%
-
-%
-%The ego vehicle controller can be considered deterministic if dynamic elements to the simulation environment are controlled, such as weather and lighting that may be on random cycles. 
-%
-% ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software.
-Actor navigation not a problem if navigation meshes are fixed.
-Game engine, ensuring fixed $dt$ and control of random numbers.
-Core or thread scheduling will not be issue unless changing scheduling policy or migrating between machines that have different policies.
-Concurrency is really down to the development of the simulation software and reputable companies should have good control over these potential issues following good software development practice.
-NUMA should only affect interprocessor data transfer cost, at wrt only make computations longer.
-relative times between access different caches is likely to be small although may impact on high throughput, e.g.
-HPC.
-VFR - want to remove this section.
-EGO vehicle deterministic?
-Can't control library change, perception stack
 
@@
 %
-\todo[inline,color=green!40]{GC new: line below.}
-%
 Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
-%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
-%
 Ultimately, our objective is to control non-determinism to minimise simulation variance.
 
-\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
-%
 We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
 A discussion on the internal and external settings of the simulation is included, along with system configuration and pre-screening sections.
-% {\color{red} and then..? Please cover the entire section - consider splitting setup from evaluation.}
-%
-\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
-%
-\todo[inline]{Complete the section structure as this is a very long section.}
-
-% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
 
@@
 \subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
-
 \begin{figure}[!t]
@@
 This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-%
 Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
@@
 This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-%
 The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-\todo[inline]{Decide past or present tense and be consistent throughout.}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
 
 \subsection{Tolerable Simulation Variance}\label{s:tolerance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
-
 To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-%
 Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
-%
-% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
-%
 The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small such that assertion checking would fail with minor computational perturbations.
-% {\color{red} {\sc What would fail?}}
-%
 Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
 
-% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-%
-% but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-% 
-% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
-%
 For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event.
-%situations such as potentially dangerous close passing of a cyclist. 
 A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives.
-% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
 Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected.
-% Thus any variation of less than $1$cm is tolerable. 
-Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-
-% considered to be sufficient for assertion checking in urban environments. 
-%
-% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
+Thus any variance of less than $1$cm is permissible.
+To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
+
 In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-%
-\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
-where any value higher than the specified tolerance is considered non-permissible. }
 
@@
 Tests 1 \& 2 involve two vehicles meeting at a junction where they either do not collide (Test 1) and where they do collide (Test 2), thereby triggering an \textit{event hit} callback in the game engine.
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
-% \todo[inline]{AG: is this value correct or should it be 2.0m?
-AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
-actors would ignore trajectory waypoints that are less than 2.0m away from them.}
-intervals using a PID controller. 
+In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m intervals using a PID controller.
 %
@@
 \subsection{Evaluation Metric}\label{s:Experiment_Description}
-% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
 For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time.
-%
 The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.
@@
 Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
-%
-\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for each actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-}
-
 The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
-%{\color{red} This is still a function over time - should it be?}
 \begin{equation} \label{eq:max_sigma}
@@
 The maximum deviation is the absolute value of the square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity.
-% {\color{red} So, ``a'' is any actor or a specific actor?
-Likewise, ``t'' is any time point or a specific time point?
-NOTE that any actor any time means a in A and t in T, where A is the set of actors and T is the set of all time points.}
 
 The maximum deviation, $\max\sigma$, can be analysed for the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation is sufficiently accurate for verification purposes.
-% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
-% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
-\todo[inline,color=green!40]{GC Addition: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
-}
-%
-% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
-% \todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
-
-% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
-Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
-
-% This whole process is repeated for various internal and external settings (if needed). 
-% Then various plots can be created to draw the engine's simulation variance. 
-% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
 
@@
 Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
-% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
 An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
 The results were compared to a baseline of the default engine settings.
-%
 However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
@@
 
-% \subsection{Initial Results and Discussion}\label{InitialResultsSection}
-
-% {\color{red} This is for unrestricted, so this does not come as a surprise later.}
-
 % ***************************************************
@@
 \subsection{External Settings}
-% \todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
-
-% After discovering that internal engine settings could not achieve a simulation variance that met the required 
-% \textit{deterministic threshold}, 
-% \textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
+
 After examining controls the internal game engine, external settings were explored.
-%
 Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
-%
 For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
@@
 
-% {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented.
-Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}
-
 To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation.
-%
-% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
 Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
 Resource utilisation figures reported here should be considered approximate values.
-%
 Resource utilisation was capped to 75\% in some parts of the results and referred to as \textit{restricted}.
@@
 
-% {\color{red} To understand the impact of collisions and high resource utilisation, the restricted column shows a subset of the results where post-collision data and experiments above 75 resource utilisation have been removed}
-%
-% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
-%
-% {\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
-
-\todo[inline,color=green!40]{GC Addition (para below):} 
-%Issues around floating-point precision and concurrency as previously mentioned notwithstanding, 
 Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
@@
 The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz.
-%
 The operating systems was Linux Ubuntu 18.04.4 LTS.
-%
 Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
 
-% Each test was repeated 1000 times. 
 Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
@@
 Therefore each test was repeated 1000 times to provide a sufficient sample size.
-% {\color{red} Justify why 1000 - see my later comment in the Method section.}
-%
 A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
-%
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:
+To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system.
+These were:
 
@@
 
-% {\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}
-
 A summary of the main results are shown in Table~\ref{TableOfExperiments}.
 In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
-%
 From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-%
 However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
-%
-\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-}
-%
 Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
@@
 Resource utilisation was found to have a significant impact on \textit{simulation variance}.
-%
 Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
 In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
-%
 Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-%
-\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
-%
 
@@
 
-\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-}
-% \todo{Why ``not reliable''.}
-
 The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
@@
 There are three observations; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
-% as described above. 
 Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}.
@@
 
-\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
-This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
-Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
-
 To conclude, the main findings of this case study suggest a working practice that would minimise the factors that give rise to the non-deterministic effects observed in this investigation.
-%
 By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table~\ref{TableOfExperiments}.
@@
 Furthermore, we were keen to investigate the impact of memory access and include a brief investigation of this in Section~\ref{r:numa}.
-%
-\todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-%For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm. 
-}
 
 \subsection{Process Scheduling Priority} \label{r:process_scheduling}
-
 \begin{figure}[t]
@@
 \subsection{Investigation Summary} \label{s:empirical_summary} 
-% {\color{red} Factor into overall conclusion of the case study}
-% \todo[inline]{GC-need to reference the 'restricted' set of results here too}
-
-%\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
-The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator.
-Collision events may utilise more computational resources than simulations without such events, i.e.
-the physics loop takes longer to process as there are complex physical interactions to resolve.
-This may coincide with the second observation that higher resource levels also lead to non-deterministic results.
-We hypothesise that both these observations are due to improper management of thread scheduling within the engine.
-When resources are occupied there may be a delay in processing critical calculations for the simulator, which may arrive out of an expected sequence or the engine compensates for this delay by reducing the number of physics loops that can be performed within a render cycle.
-(is this 2nd one true, or does carla ignore render requests?)}
 These empirical investigations have highlighted the shortcomings of using a games engine for simulation based verification and suggests advice for best working practice.
-%
 It was observed that resource utilisation positively correlates with simulation variance and that specific simulation events, such as vehicle collisions, can also lead to a breach in permissible tolerance.
-%
 The investigation found that the effect of higher simulation variance as a result of increased resource utilisation can be reduced, but not omitted entirely, by controlling the scheduling policy.
-%
 The investigation into specifying memory placement on the simulation process did improve simulation variance but only by a minor amount of 2\%.
@@
 However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly.
-%
-Therefore we have derived a general methodology that practitioners can follow to find the 
-% \textit{deterministic domains of operation} 
-\textit{operational domains of permissible variance} 
-for %any hardware configuration.
-a game-engine-based simulation environment. This methodology is presented in the next section.
-
-% First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
-with in the centimeter level).
-This clearly shows how the load on a computer (i.e.
-stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
-This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
-
-% Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
-with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
-
-% Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
-We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
-
-% Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
-Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
-
-% Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
-However, in V\&V of AV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
-Thus, the post collision effect should not be of much concern in the context of AVs.
-
-% It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
-The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
-Thus failing to stay below our red line and the test should be terminated.
-In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
-left alone and not used for any other tasks while the runs were executed.
+Therefore we have derived a general methodology that practitioners can follow to find the \textit{operational domains of permissible variance} for a game-engine-based simulation environment.
+This methodology is presented in the next section.
 
@@
 \section{Method to determine the Variance of a Simulation} \label{s:methodology}
-% In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
-This method can be used to determine the operational domains of permissible variance of a simulation environment.
-%
 In this section a method for determining the simulation variance of actor paths and resolving the operational domains of permissible variance is presented here as a work flow, see Fig.~\ref{method_diagram}.
-%
 In addition, recommendations and best practice guidelines for minimising simulation variance are suggested.
-% , such as fixing the physics time step in \textit{Simulator Settings}. 
-% {\color{red} \sc Highlight these in the text.}
-% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
-% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}.
-%These need to be considered at different stages in the process.
-%
 
@@
 
-% {\color{red} \sc Surely, a series of systematically designed experiments, each repeated X times, should be set up, where each experiment varies one of the suspected sources of non-determinism?
-Then, the analysis of the data from each experiment should reveal the actual sources of non-determinism for the given setting?
-In other words, we are aiming to fix as much as we can control (easily), and then we intentionally vary (ideally) single variables, e.g.\ the resource utilisation, to find out what makes a difference.
-We need a clear definition of the objectives of the experimental design, i.e.\ what one should vary and what one wants to ``nail down''.}
-
-% {\color{green!99} \sc }
 The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
@@
 
-% Different actor types may be handled differently by the game engine.
-For instance, our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
-However, the introduction of CARLA vehicles increased the simulation variance of the pedestrian-vehicle test by 12 orders of magnitude.
-{\color{red} \sc This needs rephrasing as it appears to contradict the statement above.}
-%
 All actors that could be included in the simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see Section~\ref{ego_actor}.
-% {\color{red} \sc ... such as the ego vehicle...? [see section on ``External Settings'' last point - Ego Controller.]}
-%
+
 \subsubsection{Trajectories and Collisions} 
-% Actor paths should be hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
 Actor paths, or the sequence of actions required to generate paths, should be hard-coded to ensure repeatability.
@@
 Actor paths without collision are also important to include as these will serve as a baseline to the other tests; to see if deviations between runs in actor paths increases with collisions.
-% {\color{red} \sc What baseline?} {\color{green} \sc Done}
 
@@
 The verification engineer should set the \textit{permissible variance}, also termed \textit{tolerance}.
-% which refers to the acceptable level of actor path variance. 
-% ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. 
-%
 The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
@@
 
-%The permissible variance is analogous to the analogue signal level associated with each of the binary states %of the signals in digital circuits.
-At some low analogue voltage the signal is interpreted as binary $0$, this may be, for example, any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
-This convention enables the abstraction from the noisy physical signal values that are analogue voltages into the digital domain with binary representation.
-<- KIE: meant only to illustrate that other areas also have defined tolerances.
-%
-
-% \todo[inline]{Make sure assertion checking has previously been introduced.}
-
-% The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
-These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
-
-% \todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''. GC change callback to 'event hit'}
-
-% The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
-The aim of the experiment is to determine if the simulation is deterministic.
-This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories.
-Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.
-
-% \todo[inline]{We may need to introduce the Case Study first, then derive the general Methodology from it.
-This way, we can explain the requirements of the case study, how things were set up, what was observed, how the methodology was refined, several iterations even, etc.}
-
 \subsubsection{Simulation Time} 
-% {\color{red} \sc Should this not be part of the Execution section?} 
 The simulation time will depend on the actor paths and terminating conditions.
@@
 The termination conditions of the simulation can be set by, for example, actors reaching their final trajectory waypoints.
- % {\color{red} \sc Point to the section that specified that.}
-
-% Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
-This is to ensure that an outlier data-point is not hidden in the average of $n$-tests.
 
@@
 Random numbers might be used in the simulation environment to control variations of background effects, e.g.\ weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment.
-%
 Actors that navigate through the environment should use a fixed navigation mesh.
-%
 The version number of the CARLA and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}.
@@
 
-% \todo[inline]{Select section titles to match the flow diagram.
-In each of these sections, go through the matching phase in the flow diagram, explaining each point listed there and justify recommendations.}
-% The experiment should be a short\todo[inline]{What does short mean? Quantify.} scenario that can be easily repeated with the same initial conditions.
-
 \subsection{External Settings} \label{ego_actor}
@@
 \subsubsection{Resource Utilisation}
-% To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-The resources available to the simulator have been shown to have a significant effect on the variance of the path of simulated vehicles. %and exploring this as a variable the analysis is required. 
-%
+The resources available to the simulator have been shown to have a significant effect on the variance of the path of simulated vehicles.
 Thus, it is important to understand at what level of resource utilisation the system running the simulation becomes susceptible to \textit{simulation variance}.
@@
 These values should be added to the data logs.
-%
 Alternatively, in place of artificial resource utilisation, multiple instances of the simulation could be executed simultaneously.
 However, the granularity of control with this approach may be reduced.
-
-% \todo[inline]{Again, this sounds as if having the case study first makes more sense. That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
-
-% \todo[inline]{This should be ``External Settings'' - from flow diagram?}
 
@@
 
-% If the policy or operating system cannot be changed then s
 Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
@@
 This control allows the simulator process to be fixed on a single core, reducing and unifying memory access time.
-%
 Investigations in tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}.
-%
-% Using \texttt{numactl} allows a process to run with a specified memory placement policy, fixing the process to a single core.
-This will reduce the data transfer cost and the opportunity for disrupted execution order that may be responsible for increase simulation variance.
-%
 Using a fixed memory placement policy may assist if simulation variance is borderline to the tolerance but only of the order of a few percent from our observations.
@@
 \subsubsection{Ego Vehicle Controller} 
-% {\color{red} \sc Why is this not covered under actors? [see my note in the Actors section]}
-% {\color{red} \sc Clearly separate the investigation from the Method.
-The fact that the case study did not have an ego vehicle should be mentioned as a limitation of the case study, rather than elaborated on here, but why does this make a difference?}
-%
 An ego vehicle was not used in this study, but the impact of introducing this to the simulator can be considered here.
@@
 
-% ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
-{\color{red} \sc Long and cumbersome sentence with strange grammar: No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
-}
-
 \subsection{Execution}
@@
 \subsubsection{Sample Size}
-% Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} 
-% Histogram to show how sampling with position binning varies?
-% \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
-%
-% {\color{red} \sc Again, separate the study from the method.
-This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.}
 It is recommended that the chosen sample size is determined empirically.
-%
-We recommend monitoring variance while increasing the sample size in orders of magnitude
-until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process. 
-%
+We recommend monitoring variance while increasing the sample size in orders of magnitude until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process.
 Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
-
-% By increasing the sample size and observing that the simulation variance is stable, one can infer that a sufficient number of samples have been collected.
-If the variance changes with a change in sample size then a stable condition has not been found.
-% ...
-% ...
-%  dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
-% % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
 
@@
 Additional information should also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
-% such as the \texttt{unit} command in Unreal Engine. 
-%For the subsequent analysis, each actor should have an identifier. The repeat number and simulation time should also be captured. 
-% The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
-For example, if 100 tests are executed and the maximum deviation in actor path is $<1$cm, then this scenario could be interpreted as being deterministic above this scale.
 
@@
 
-%Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
-% KIE: Why now point to HW? And, which "boundaries"?
-
-% Game engine actors can be used to simulate the AV and other vehicles as well as pedestrians.
-For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
-
-% \todo[inline]{This is mixing our experiments (which are part of the case study) with the methodology.
-It is probably best to first introduce the case study and then derive the methodology from it, rather than mixing both.} In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
-Initial testing also indicated that when actors collide there is a significant change in \textit{simulation variance}, therefore the experiment should include some scenarios where collision callbacks are activated.\todo[inline]{Make sure collision callbacks have been introduced earlier.}
-
-% Randomisation must be fully controlled by fixing the seeds for any random number generation that is used during the experiment.
-It is important to identify and control any external sources that may use randomisation, e.g.\ external vehicle controllers or other clients connected to the simulation environment.
-
-% \todo[inline]{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
-
-% \todo[inline]{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
-%
-% For the vehicles in the simulation, waypoints are generated and interpolated in the same way but execution at runtime is slightly different where a PID controller is used to generate a velocity profile and steering angle requests to reach the destination targets.}
-
-% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
-Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.
-A* search algorithm.
-
-% \subsection{Internal settings}\todo[inline]{Systematically go through the points listed in the flow diagram, explain and justify.}
-% Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
-
-% By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronous mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine.
-% Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
-However, in such applications of V\&V of AV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
-% If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the simulation variance as a result of allowing the AI to navigate in a more well defined space.
-% Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
-% The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.
-% Running in editor or deployed mode can sometimes cause differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
-
-% \begin{table*}[b]
-% \centering
-% \begin{tabular}{clclcc}
-% \toprule
-% Test ID & Test description (See Figures~\ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
-% 1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
-% 2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
-% 3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
-% 4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
-% 5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
-% 6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
-% 7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
-% 8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
-% \bottomrule
-% \end{tabular}
-% \caption{Set of experiments}
-% \label{TableOfExperiments}
-% \end{table*}
-
-% The user should also monitor and record system CPU and GPU utilisation rates.
-Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
-
-% Collision callbacks
-% \subsection{Other Considerations}
-% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
-ii) If not, how does its simulation variance vary?
-In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
-Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
-% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
-Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
-% The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
-For the majority of AV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
-
-% \subsection{External settings}
-% Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.
-Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
-Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.
-
-% \subsection{Running experiment}
-% \noindent When it comes to running the experiment itself, things that one should consider are the problem space exploration; the number of runs that should be executed to get reliable results; the frequency of logging data of actors; and possibly having a program to monitor the hardware utilisation.
-% Problem exploration, there is too many variables that one can play with. do statistical analysis some.... some of them get problematic...
-
 % ***************************************************
@@
 % ***************************************************
-% \section{Conclusions and Future Work}\label{s:conclusion}
 \section{Conclusions \& Future Work}\label{s:conclusion}
 
-% \todo[inline]{Say what the paper investigated and what we aimed to achieve, why this is important and how we went about it.
-Then point out the derivation of a re-usable methodology based on the knowledge and understanding we gained from the case study.}
-
-% \todo[inline]{GC: we need to say what we set out to achieve and why that is important, then say how our investigation led us to several theories which resulted in understanding several key parameters which directly affect simulation variance.
-From this we developed a methodology that the community can use to understand the deterministic operational domains of their own specific hardware:  Autonomous vehicle testing is adopting the use of game engines for development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-An empirical investigation uncovered several parameters that contributed to simulation variance ...
-explain results ...
-This led to the design of a methodology allowing 'users' to determine the deterministic operational domains for a specific hardware/platform configuration ...
-}
-
-%The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
 Game engines offer simulation environments that are used for the development and verification of autonomous driving functions.
-%
 Determinism of a simulator is required to achieve repeatability, which is essential to find and fix software bugs efficiently, and to ensure simulation results are trustworthy.
-%
 If a simulator is non-deterministic then practitioners should at least be aware of, and know how to find, the operational domains where \textit{simulation variance} is tolerable.
@@
 An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation.
-%
 We then researched, identified and discussed potential sources for non-determinism in this context.
-%
 A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution.
 In particular, actor collisions and system-level resource utilisation were identified as key contributors.
-%
 To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
-%
-% {\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.
-This seems to have been commented out but was here before.}
 
@@
 This can give AV developers and verification engineers increased confidence in the simulation results and reduce debug time.
-% 
-% \subsection{Future Work}
-%
 As future work the method can be extended to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
@@
 %
-\textit{Abanoub Ghobrial and Greg Chance contributed equally to this paper.}
+\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper.}
 \bibliography{cav_determinism.bib}
-
-% \end{multicols}
 \end{document}
-
 \grid

### 2021-04-08 paper_arxiv/arxiv_cav_determinisim_main.tex
minor change in .tex and several in bib file

@@
 \usepackage[font=footnotesize,labelfont=bf]{subcaption}
-\usepackage[hidelinks]{hyperref}
 \usepackage[capitalize]{cleveref}
@@
 \usepackage{siunitx}
-\usepackage{xurl} %for long URLs
+\usepackage{xurl} % for long URLs
+\usepackage[hidelinks]{hyperref} % use hyperrefs for urls
+
+\usepackage{draftwatermark}
+\SetWatermarkText{Preprint \today}
+\SetWatermarkScale{3}
+
+\usepackage{flushend} % even out last page
 
@@
 % 
-The gamer neither needs a faithful representation of reality (i.e.
-gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
-gamer accepts low precision).
+The gamer neither needs a faithful representation of reality (i.e.\ gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.\ gamer accepts low precision).
 In contrast, high accuracy and precision are necessary for AV development, testing and verification.
@@
 %
-A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests}, although the AV simulation domain introduces it's own unique challenges that were not considered in that paper.
+A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests}, although the AV simulation domain introduces its own unique challenges that were not considered in that paper.
 
@@
 There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
+Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{newton2016unreal}.
 The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.

### 2021-04-08 paper/cav_determinisim_main.tex
minor typo and ref correction in .tex and several edits in .bib, capitalisation, urls as well as removing a duplicate entry

@@
 % 
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
+Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{newton2016unreal}.
 %

### 2021-04-09 paper/cav_determinisim_main.tex
greg and kevin bios added

@@
 
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{Other/mugs/greg.png}}]{Greg Chance}
+Greg Chance received the B.Sc.
+(Hons.) and Ph.D.
+degrees in physics from the University of Bath, Bath, U.K., in 2001 and 2005, respectively.
+He was previously at the Bristol Robotics Lab and is currently a Senior Research Associate with the Trustworthy Systems Laboratory, University of Bristol, Bristol, U.K.
+His current research interests are simulation-based verification for autonomous systems and cybersecurity.
+He has ten years industrial experience researching for Oxford Instruments, U.K., and BAE Systems, U.K.
+Dr Chance is a Chartered Engineer and a member of the Institute of Physics and holds a patent for a novel plasma control method.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}{Abanoub Ghobrial}
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{Other/mugs/kevin.jpg}}]{Kevin McAreavey}
+Kevin McAreavey received a BSc (Hons) in computer science from Queen's University Belfast (QUB) in 2010 followed by a PhD in AI in 2014.
+He is currently a Senior Research Associate with the Intelligent Systems Laboratory at the University of Bristol (UOB), working on explainable AI for AI-enabled domestic life.
+Previously he has worked with the Trustworthy Systems Laboratory at UOB and as a Research Fellow at QUB.
+His research interests are broadly in the areas of knowledge representation and reasoning, agents and multi-agent systems, and automated planning.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}{Severin Lemaignan}
+\end{IEEEbiography}
+
+\begin{IEEEbiography}{Tony Pipe}
+\end{IEEEbiography}
+
+\begin{IEEEbiography}{Kerstin Eder}
+\end{IEEEbiography}
+
 \end{document}

### 2021-04-09 paper_arxiv/arxiv_cav_determinisim_main.tex
fixed URL dropouts in bib
balanced last column

--- 2021-04-07 paper_arxiv/arxiv_cav_determinisim_main.tex
@@
 \documentclass[runningheads,twocolumn,a4paper,10pt]{llncs}
+
+\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
+\addbibresource{cav_determinism.bib}
+
 % \documentclass[runningheads,a4paper]{llncs}
@@
 \usepackage{balance} %balance columns
+\usepackage{lastpage}
 \hyphenation{op-tical net-works semi-conduc-tor}
@@
 \begin{document} 
-\bibliographystyle{ieeetr}
+% \bibliographystyle{ieeetr}
 \mainmatter
@@
 \textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper.}
-\bibliography{cav_determinism.bib}
+% \bibliography{cav_determinism.bib}
+\balance
+\printbibliography
 \end{document}

### 2021-04-12 document/arxiv/arxiv_template.tex
new document build structuring with common source

--- 2021-04-09 paper_arxiv/arxiv_cav_determinisim_main.tex
@@
 \usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
-\addbibresource{cav_determinism.bib}
+\addbibresource{../source/cav_determinism.bib}
 
@@
 
-\textbf{\textit{Abstract}--Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
-%
-A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history.
-%
-Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
-%
-However, game engines are seldom deterministic. 
-%
-This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
-%
-A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
-%
-Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
-%
-Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.}
+\textbf{\textit{Abstract}\input{../source/abstract.tex}}
 
 % ***************************************************
-%  INTRODUCTION
+%  Main Body
 % ***************************************************
-\section{Introduction} \label{s:introduction}
-Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
-%
-Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of safe and legal driving rules, termed assertions, that must not be violated.
-
-Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules, e.g.\ through assertion checking during simulation. 
-There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-Simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
-In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
-
-Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
-%
-CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
-
-State-of-the-art game engines provide a convenient option for simulation-based testing.
-They offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
-%
-Furthermore, they are easy to set up and run compared to on-road testing and are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
-%
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. 
-%
-Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
-%
-However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
-
-In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
-%
-In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs.
-We find that the CARLA simulator is non-deterministic.
-Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
-We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
-CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
-%
-
-The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
-%
-Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
-In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
-%
-
-This paper is structured as follows.
-%
-Section~\ref{s:prelim} defines terms used throughout the paper and identifies when determinism is needed. 
-Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
-%
-Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
-%
-Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
-%
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.
-
-\section{Preliminaries} \label{s:prelim}
-\subsection{Definitions}
-Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
-\\
-
-\subsubsection{Determinism}
-Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
-However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
-Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
-
-A simulation can be thought of as the process of generating or producing experimental data.
-In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data.
-If a simulation is deterministic, Fig.~\ref{variance_description}~(b), then there will be no variation in the generated output data, i.e.
-all future states are perfectly reproducible from prior states and actions.
-However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data.
-\\
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
-        \caption{Non-deterministic}
-        \label{variance_description_a}
-    \end{subfigure}
-
-    \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
-        \caption{Deterministic}
-        \label{variance_description_b}
-    \end{subfigure}
-    \caption{Demonstration of variance, precision, tolerance and determinism}
-    \label{variance_description}
-\end{figure}
-
-\subsubsection{Variance, Precision \& Tolerance}
-We adopt terminology from the mechanical engineering and statistics domains to describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
-\textit{Variance} is used here to define the spread or distribution, of the generated output data with respect to the mean value.
-\textit{Precision} is synonymous with \textit{variance} although inversely related mathematically.
-Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed under the same conditions and actions.
-\textit{Tolerance} is defined as the permissible limit of the variance, or in short the \textit{permissible variance}.
-
-As an analogy, the simulator can be thought of as a manufacturing process that produces data.
-To determine the precision of this process, the output must be measured and analysed for differences when the process is repeated.
-Those differences describe the spread or variance in the process output.
-A hard limit on the variance can then be defined, Fig.~\ref{variance_description}~(a), beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
-Real manufacturing fails to achieve absolute precision.
-Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.
-
-If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within tolerance by design.
-If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
-%
-
-\subsubsection{Accuracy}
-Precision and tolerance should not be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
-Therefore, we define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation.
-%
-For a driving simulation, the reference value may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
-%
-In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
-In most cases it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
-\\
-
-\subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) in a 3D environment recorded at regular time intervals.
-This definition could be extended to include other variables.
-A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run.
-\\
-
-\subsubsection{Simulation Variance \& Deviation}
-If the simulator is non-deterministic then how can the simulation variance be measured?
-This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run.
-For example, actor position variance is a distance-based metric that can be derived from simulation traces.
-The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained.
-% 
-Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions.
-Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
-\\
-
-\subsubsection{Scene, Scenario \& Situation}
-We adopt the terminology defined for automated driving in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
-Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV, or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals.
-The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.
-A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
-
-\subsection{When is Determinism needed?}
-Determinism is a key requirement for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
-Thus, a deterministic simulator has zero \textit{variance}.
-
-A simulator with non-zero variance can no longer be considered deterministic.
-Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}.
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-
-If the simulation is non-deterministic, e.g.\ it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
-In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
-
-When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
-It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
-However, the situation is very different for AV development and testing.
-Thus, our main research questions are:
-{\em How can one assess whether a simulation environment is deterministic?} and 
-{\em How can one determine and control the simulation variance?}
+\input{../source/body.tex}
 
 % ***************************************************
-%  BACKGROUND
+%  Thanks
 % ***************************************************
-\section{Background} \label{s:background}
-
-There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
-
-Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.
-Ensuring software operates deterministically is a non-trivial task.
-Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
-
-% ======= Overview of a Game Loop
-Central to a game engine are the main game logic, the artificial intelligence (AI) component, the audio engine, and the physics and rendering engines.
-For AV simulation, we focus on the latter two.
-The game loop is responsible for the interaction between the physics and rendering engines.
-Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed~\cite{unity_event_execution}.
-A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering)~\cite{GameEngineArchBook}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoopv2.pdf}
-\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}.}
-\label{GameEngineLoopDiagram}
-\end{figure}
-
-The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
-This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}.
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}.
-When the frame is rendered the game loop cycle returns to processing inputs.
-An intuitive and  more detailed description of the interplay between the physics and render cycles is given in~\cite{JohnAustinUnity}.
-
-The physics engine operates according to a time step, $dt$.
-The shorter this time step is, the smoother the interpretation of the physical dynamics will be.
-To use a fixed physics time step, the user's display refresh rate needs to be known in advance.
-This requires an update loop to take less than one render tick (one frame of real world time).
-Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$.
-However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}.
-%
-Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware.
-Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}.
-If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics.
-Thus, extrapolation between frames may need to be performed to smooth transition between scenes.
-Note that both residual lag and extrapolation could affect perception stack testing.
-In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
-
-Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-% 
-The gamer neither needs a faithful representation of reality (i.e.
-gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
-gamer accepts low precision).
-In contrast, high accuracy and precision are necessary for AV development, testing and verification.
-
-% ======= Sources of non-determinism
-\section{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}
-
-The following review discusses the potential sources of non-determinism that were found in the literature or found as part of our investigation into game engines.
-We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
-%
-A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests}, although the AV simulation domain introduces its own unique challenges that were not considered in that paper.
-
-\medskip
-
-% ======= Floating-Point Numbers
-\subsection{Floating-Point Arithmetic}
-
-Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
-Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value.
-This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
-%
-
-As a consequence, floating-point arithmetic is not associative.
-Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}.
-This is because rounding is being performed at different stages in the computation, depending on the respective execution order.
-%
-Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
-%
-Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-
-% Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
-%
-Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability.
-In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-
-In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-%
-It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
-%
-Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. 
-%
-So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
-
-However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
-Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
-
-Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
-However, the same executable should still return the same output for identical input.
-
-In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.
-
-\subsection{Scheduling, Concurrency and Parallelisation}
-Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy.
-A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-%
-In principle, changing the scheduling policy and thread priorities may increase simulation variance.
-
-It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
-Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
-
-However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
-This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
-%
-Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-%
-
-Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
-
-Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes.
-This may be decided based on factors such as throughput, latency or CPU utilisation.
-%
-If due to the CPU utilisation policy the same single-threaded script executes multiple times across different physical cores of the same CPU type,  then execution should still produce the same output.
-%
-This is because the processor cores are identical and any impurities across the bulk silicon and minor perturbations in the semiconductor processing across the chip that may exist should have been accommodated for in design and manufacturing tolerances.
-%
-However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
-
-Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
-%
-
-Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously.
-While this would be faster than a serial execution, the order of execution arising from program-level concurrency is often not guaranteed.
-
-Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}.
-
-\subsection{Non-Uniform Memory Access (NUMA)} \label{s:sources_numa}
-For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
-Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
-%
-Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower.
-%
-By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
-
-\subsection{Error Correcting Code (ECC) Memory}
-ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}.
-Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
-If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
-
-Any simulation hardware not using ECC memory that runs for 1000's of hours, typical in AV verification, is likely to incur significant CPU hours and is therefore subject to increased exposure to these errors.
-To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.
-
-\subsection{Game Engine Setup}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps, ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
-%
-For example, in Unreal Editor the \textit{unit}~\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time.
-%
-With respect to perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water, ensuring textures are not randomly generated.
-
-\subsection{Actor Navigation}
-There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
-The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
-A navigation mesh is a fixed area of the environment where actors are free to navigate.
-Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
-However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment.
-These cannot be unknowingly modified and therefore can be considered fixed.
-
-Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
-
-\subsection{Summary}
-We have investigated the potential sources of non-determinism affecting game engines and explored the impact they may have on simulation variance.
-Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
-To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
-Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
-Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
-If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.
-
-Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups.
-However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop.
-Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism.
-
-% ***************************************************
-%  Empirical Investigation
-% ***************************************************
-\section{Case Study of Simulation Variance} \label{s:case-study}
-
-We present an empirical investigation into using game engines for
-simulation-based verification of autonomous vehicles with a focus on
-characterising sources of non-determinism in order to understand  the impact
-they have on simulation variance. 
-%
-Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
-Ultimately, our objective is to control non-determinism to minimise simulation variance.
-
-We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-A discussion on the internal and external settings of the simulation is included, along with system configuration and pre-screening sections.
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
-\end{figure}
-
-This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-
-\subsection{Tolerable Simulation Variance}\label{s:tolerance}
-To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
-The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small such that assertion checking would fail with minor computational perturbations.
-Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
-
-For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event.
-A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives.
-Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected.
-Thus any variance of less than $1$cm is permissible.
-To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-
-In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-
-% ***************************************************
-%  Actor Collisions
-% ***************************************************
-
-\begin{table*}[b]
-\centering
-\begin{tabular}{clclccc}
-\toprule
-Test & Actors               & Collision    & Collision Type        & $n$  & $\max\sigma$ (m) & $\max\sigma$ (m) \\ 
-   &                  &        &               &      & (unrestricted)  & (restricted)  \\ \midrule
-1    & Two vehicles                   & No       & N/A               & 1000 & 0.03 & $7.0{\times}10^{-3}$ \\
-2    & Two vehicles                   & Yes      & Vehicle and Vehicle     & 1000 & 0.31 & $9.8{\times}10^{-3}$ \\
-3    & Two vehicles and a pedestrian  & No       & N/A               & 1000 & 0.07 & $5.2{\times}10^{-4}$ \\
-4    & Two vehicles and a pedestrian  & Yes      & Vehicle and Pedestrian    & 1000 & 0.59 & $1.5{\times}10^{-12}$ \\
-5    & Two pedestrians                & No       & N/A               & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-6    & Two pedestrians                & Yes      & Pedestrian and Pedestrian & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-\bottomrule
-\end{tabular}
-\caption{A description of the test scenarios showing the test number, the actors included, whether a collision occurred and if so then between which actors.
-$n$, the number of repeats is set to 1000 and $\max\sigma$ is the maximum simulation deviation.
-The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation.
-To understand the impact of collisions and high resource utilisation, the \textit{restricted} column shows a subset of the results where post-collision data and experiments above 75\% resource utilisation have been removed.}
-\label{TableOfExperiments}
-\end{table*}
-
-\subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}.
-%
-Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were identified as potentially key aspects to the investigation into simulation variance.
-
-The tests used for the case study are listed in Table~\ref{TableOfExperiments}. They cover a range of interactions between actor types. 
-%
-Tests 1 \& 2 involve two vehicles meeting at a junction where they either do not collide (Test 1) and where they do collide (Test 2), thereby triggering an \textit{event hit} callback in the game engine.
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m intervals using a PID controller.
-%
-In Test 3 a mixture of different actor types is introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point. 
-%
-In Test 4 this pedestrian collides with one of the vehicles at the crossing, triggering an \textit{event hit} callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
-Tests 5 \& 6 involve only pedestrians that either do not collide (Test 5) and that do collide (Test 6), see Fig.~\ref{Test_b}.
-
-% ***************************************************
-%  Experiment Description
-% ***************************************************
-\subsection{Evaluation Metric}\label{s:Experiment_Description}
-For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time.
-The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.
-
-Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
-The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
-\begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a^{2}(t)
-\end{equation}
-The maximum deviation is the absolute value of the square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity.
-
-The maximum deviation, $\max\sigma$, can be analysed for the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation is sufficiently accurate for verification purposes.
-
-% ***************************************************
-%  Internal Settings
-% ***************************************************
-\subsection{Simulator Settings}
-
-Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
-Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.\ \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.\ \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
-An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
-The results were compared to a baseline of the default engine settings.
-However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
-Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.
-
-% ***************************************************
-%  External Settings
-% ***************************************************
-\subsection{External Settings}
-
-After examining controls the internal game engine, external settings were explored.
-Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
-For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
-
-Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load.
-This initial observation was explored more fully in this work using the CARLA platform.
-
-To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation.
-Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
-Resource utilisation figures reported here should be considered approximate values.
-Resource utilisation was capped to 75\% in some parts of the results and referred to as \textit{restricted}.
-This was done to limit the load on the test system and hence limiting simulation variance.
-The term \textit{unrestricted} places no such limit on the system load.
-
-Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
-For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
-In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency-induced imprecision~\cite{matlab_parallel_computing}.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV6.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
-Tests 5 and 6 overlap having almost identical results.
-Note that the lines between data points are only a guide.}
-    \label{ExperimentsStressSummary}
-\end{figure}
-
-% ***************************************************
-%  System Configuration and Screening
-% ***************************************************
-\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
-The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz.
-The operating systems was Linux Ubuntu 18.04.4 LTS.
-Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
-
-Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
-While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results.
-Therefore each test was repeated 1000 times to provide a sufficient sample size.
-A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system.
-These were:
-
-\begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, no failures on all 11 tests~\cite{shi2009testing}.
-\end{itemize}
-
-% ***************************************************
-%  Results and Discussion
-% ***************************************************
-\section{Results and Discussion}\label{s:FinalResultsSection}
-
-\begin{figure*}[t]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
-        \caption{}
-        \label{CarsCollisionCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
-        \caption{}
-        \label{CarsCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to vehicle collision (Test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
-The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
-\end{figure*}
-
-\begin{figure*}[h]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
-        \caption{}
-        \label{CarsPeopleCollsionsCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
-        \caption{}
-        \label{CarsPeopleCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to pedestrian collision (Test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
-\end{figure*}
-
-A summary of the main results are shown in Table~\ref{TableOfExperiments}.
-In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
-From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
-Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
-
-Resource utilisation was found to have a significant impact on \textit{simulation variance}.
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
-Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-
-A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
-
-Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision.
-%
-The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-
-The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (Test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three observations; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
-Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}.
-Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers, developers and researchers in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous simulation results.
-% not just assertion checking
-
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v7.pdf}
-    \caption{Stages of the method proposed to determine the variance of a simulation.}
-    \label{method_diagram}
-\end{figure*}
-
-To conclude, the main findings of this case study suggest a working practice that would minimise the factors that give rise to the non-deterministic effects observed in this investigation.
-By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table~\ref{TableOfExperiments}.
-By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve.
-%
-Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-
-A correlation between resource utilisation and simulation variance has been observed in these results which may be due to process scheduling.
-Rather than limiting utilisation of the whole system, another approach may be to promote the scheduling priority of the simulation process which is explored in Section~\ref{r:process_scheduling}.
-Furthermore, we were keen to investigate the impact of memory access and include a brief investigation of this in Section~\ref{r:numa}.
-
-\subsection{Process Scheduling Priority} \label{r:process_scheduling}
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
-    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
-    \label{NICEExperimentStressSummary}
-\end{figure}
-
-An investigation into the impact of process scheduling on the simulation variance was undertaken following the observations of increased simulation variance with increased resource utilisation.
-The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
-%
-To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.
-
-The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range of deviation, non-outlier limits by the whiskers and a horizontal bar for the median.
-The figure shows that decreasing the priority of the simulator process (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot).
-Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results.
-This can be seen in the maximum priority setting where the bars in the plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
-%
-The remaining difference in the variance between the two resource utilisation levels may be due to the lack of absolute control that \texttt{NICE} has over process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
-
-\subsection{Non-uniform memory access (NUMA)} \label{r:numa}
-An additional investigation into memory access was undertaken given the potential impact that data transfer cost and execution order may have on simulation variance, see Section~\ref{s:sources_numa}.
-%
-The program \texttt{numactl}~\cite{numactl_NUMA} allows a process to run with a specified memory placement policy, essentially allowing the process to be bound to a particular CPU core. %
-\texttt{numactl} was used to fix the simulator and test script to single cores, and a (2\%) improvement in simulation variance was observed. %
-This was considered minor comparative to the changes in simulation variance observed in other aspects of the case study, for example the pre- and post-collision between a vehicle and pedestrian saw a change of $10^{14}$.
-Therefore \texttt{numactl} was not used for subsequent testing.
-
-However, practitioners that are aiming to minimise simulation variance should use a technique such as this to minimise data transfer cost and potentially minimise any potential effect on process execution order that may lead to increased simulation variance.
-
-\subsection{Investigation Summary} \label{s:empirical_summary} 
-These empirical investigations have highlighted the shortcomings of using a games engine for simulation based verification and suggests advice for best working practice.
-It was observed that resource utilisation positively correlates with simulation variance and that specific simulation events, such as vehicle collisions, can also lead to a breach in permissible tolerance.
-The investigation found that the effect of higher simulation variance as a result of increased resource utilisation can be reduced, but not omitted entirely, by controlling the scheduling policy.
-The investigation into specifying memory placement on the simulation process did improve simulation variance but only by a minor amount of 2\%.
-
-However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly.
-Therefore we have derived a general methodology that practitioners can follow to find the \textit{operational domains of permissible variance} for a game-engine-based simulation environment.
-This methodology is presented in the next section.
-
-% ***************************************************
-%  METHOD
-% ***************************************************
-\section{Method to determine the Variance of a Simulation} \label{s:methodology}
-In this section a method for determining the simulation variance of actor paths and resolving the operational domains of permissible variance is presented here as a work flow, see Fig.~\ref{method_diagram}.
-In addition, recommendations and best practice guidelines for minimising simulation variance are suggested.
-
-The method consists of a sequence of five stages; experimental design, simulator settings, external settings, execution and analysis.
-In the following, each stage is described in detail with reference to the items listed for each stage in Fig.~\ref{method_diagram}.
-
-\subsection{Experimental Design}\label{s:design_experiment}
-
-The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
-Each test is repeated $n$ times and the analysis of the results provide a confidence with a degree of statistical certainty.
-By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and addressed.
-
-\subsubsection{Actors} \label{s:actors}
-
-All actors that could be included in the simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see Section~\ref{ego_actor}.
-
-\subsubsection{Trajectories and Collisions} 
-Actor paths, or the sequence of actions required to generate paths, should be hard-coded to ensure repeatability.
-These paths should include collisions between actors and potentially collisions between actors and static scenery if this is likely to occur in the simulation or as part of the verification process.
-Actor paths without collision are also important to include as these will serve as a baseline to the other tests; to see if deviations between runs in actor paths increases with collisions.
-
-\subsubsection{Permissible Variance} \label{s:threshold}
-The verification engineer should set the \textit{permissible variance}, also termed \textit{tolerance}.
-The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection, but not so small for assertion results to differ for repeated runs.
-In Section~\ref{s:tolerance} a tolerance of $1$cm was considered sufficient for urban scenario assertion checking.
-In practice, it may be necessary to determine this tolerance experimentally.
-
-\subsubsection{Simulation Time} 
-The simulation time will depend on the actor paths and terminating conditions.
-The simulation time should be sufficient to record interactions between actors but not so long that the testing take an inconvenient time to complete.
-In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25} in Section~\ref{s:FinalResultsSection}.
-The termination conditions of the simulation can be set by, for example, actors reaching their final trajectory waypoints.
-
-\subsection{Simulator Settings}
-The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
-If using CARLA, a small fixed value, say $0.05s$, can be set by using \texttt{setting.fixed\_delta\_seconds = 0.05}.
-In Unity, the default fixed time step is set to $0.02s$~\cite{MonoBehaviour_unity}.
-
-In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
-
-The use of random numbers must be controlled through fixed seeds, resulting in pseudo-randomness.
-Random numbers might be used in the simulation environment to control variations of background effects, e.g.\ weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment.
-Actors that navigate through the environment should use a fixed navigation mesh.
-The version number of the CARLA and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}.
-Therefore, ensuring a consistent version number throughout testing is also important.
-
-\subsection{External Settings} \label{ego_actor}
-
-\subsubsection{Resource Utilisation}
-The resources available to the simulator have been shown to have a significant effect on the variance of the path of simulated vehicles.
-Thus, it is important to understand at what level of resource utilisation the system running the simulation becomes susceptible to \textit{simulation variance}.
-
-CPU utilisation software, such as the linux \texttt{stress} tool, which is a workload generator program, can be used to spawn workers on any number of cores or virtual threads on a system.
-This can be used to artificially increase the load on the system.
-For GPU utilisation, \texttt{gpu-burn} can be employed using the \texttt{fur test}.
-Different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
-Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU, respectively.
-These values should be added to the data logs.
-Alternatively, in place of artificial resource utilisation, multiple instances of the simulation could be executed simultaneously.
-However, the granularity of control with this approach may be reduced.
-
-\subsubsection{Memory Testing}
-Prior to experimental execution the system hardware should be tested for memory conformity and to ensure no single bit errors are occurring, see Section~\ref{s:screening}.
-For mainboard memory \texttt{memtest86}\ can be used on most platforms to run a series of pre-defined memory test patterns.
-This memory testing software can also be used for ECC enabled hardware.
-Similarly, to test memory on Nvidia based graphical adaptors \texttt{cuda\_memtest} can be used to ensure no memory errors exist.
-
-\subsubsection{Scheduling}
-We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
-However, gaining fine control over the scheduling policy and thread execution order is known to not be trivial~\cite{acm-q-rr-interview}.
-
-The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority.
-Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation.
-
-Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
-
-\subsubsection{NUMA Control}
-Control over a Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multi-core processors with shared memory.
-This control allows the simulator process to be fixed on a single core, reducing and unifying memory access time.
-Investigations in tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}.
-Using a fixed memory placement policy may assist if simulation variance is borderline to the tolerance but only of the order of a few percent from our observations.
-
-\subsubsection{Ego Vehicle Controller} 
-An ego vehicle was not used in this study, but the impact of introducing this to the simulator can be considered here.
-The ego vehicle is seen as another actor in the simulation but care must be taken to ensure that any control algorithms, machine learning modules and processing pipelines are deterministic.
-We recommended this be treated as a separate source of non-determinism and handled accordingly.
-
-\subsection{Execution}
-
-\subsubsection{Sample Size}
-It is recommended that the chosen sample size is determined empirically.
-We recommend monitoring variance while increasing the sample size in orders of magnitude until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process.
-Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
-
-\subsubsection{Data Logging}
-Unique identifiers should be assigned to each experiment, each repeat and each individual actor. 
-The time-stamped actor positions should then be recorded at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-Additional information should also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
-
-\subsection{Analysis}
-For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment.
-
-% ***************************************************
-%  CONCLUSION
-% ***************************************************
-\section{Conclusions \& Future Work}\label{s:conclusion}
-
-Game engines offer simulation environments that are used for the development and verification of autonomous driving functions.
-Determinism of a simulator is required to achieve repeatability, which is essential to find and fix software bugs efficiently, and to ensure simulation results are trustworthy.
-If a simulator is non-deterministic then practitioners should at least be aware of, and know how to find, the operational domains where \textit{simulation variance} is tolerable.
-
-An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation.
-We then researched, identified and discussed potential sources for non-determinism in this context.
-A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution.
-In particular, actor collisions and system-level resource utilisation were identified as key contributors.
-To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
-
-A general method to assess the actor path variance of a game-engine-based simulation environment is then proposed.
-The method can be used to find the \textit{domains of permissible variance} of a simulation environment for a given system configuration.
-This can give AV developers and verification engineers increased confidence in the simulation results and reduce debug time.
-As future work the method can be extended to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
-
-An ambitious avenue for future work is the development of a deterministic simulator for AV development and verification.
-This requires controlling all potential sources of non-determinism, including randomness and scheduling, very similar to the development of the record-and-reply debugger \texttt{rr}~\cite{RR_link}, originally developed to catch low-frequency non-deterministically failing tests at Mozilla~\cite{acm-q-rr-interview}.
-
 \section*{Acknowledgement}
@@
 \textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper.}
-% \bibliography{cav_determinism.bib}
 \balance
+
+% ***************************************************
+%  Bib
+% ***************************************************
 \printbibliography
+
 \end{document}

### 2021-04-12 document/ieee_trans/ieeetrans_template.tex
new document build structuring with common source

--- 2021-04-09 paper/cav_determinisim_main.tex
@@
 % \addbibresource{IROS.bib}
-\addbibresource{cav_determinism.bib}
+\addbibresource{../source/cav_determinism.bib}
 \usepackage{array}
@@
 \maketitle
-%
+
+% ***************************************************
+%  Abstract
+% ***************************************************
 \begin{abstract}
 \noindent 
-Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
-%
-A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history.
-%
-Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
-%
-However, game engines are seldom deterministic. 
-%
-This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
-%
-A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
-%
-Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
-%
-Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.
-
-% {\color{red} \sc Ensure past/present is used consistently and appropriately throughout.}
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
-% %
-% A key requirement for simulation-based development and verification is determinism.
-% %
-% Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
-% %
-% Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
-% %
-% However, game engines are seldom deterministic.
-
-% This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.
-% %
-% This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
-% %
-% Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
-variance between repeats becomes negligible for development and testing work.
-
-% Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
-% %
-% A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
-% %
-% Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
-% \todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
-% %
-% % KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
-% % These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
-% %
-% However, game engines do not need to be deterministic per se. % when used for games. 
-% %
-% This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively.
-% %
-% \todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
-% %
-% A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification.
-The proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-% %
-% % \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}
-
-\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline,color=yellow!40]{AG Alternative: A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}
-
-% The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
-It is important for these engines to be deterministic in order for one to carry out these different processes.
-However, such engines are inherently non-deterministic.
-We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic.
-We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.
+\input{../source/abstract.tex}
 \end{abstract}
@@
 % ***************************************************
-%  INTRODUCTION
+%  Main Body
 % ***************************************************
-\section{Introduction} \label{s:introduction}
-% What is verification
-% Can't use this again as this is how the AI Test paper starts. -> Avoid self-plagiarism!
-%\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. 
-%
-% Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, benefiting from complete control over (virtual) actors and their environment
-% , offering a relatively inexpensive means to explore test cases while also
-% \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. 
-\IEEEPARstart{S}{}imulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
-%
-Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of safe and legal driving rules, termed assertions, that must not be violated.
-
-Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules, e.g.\ through assertion checking during simulation. 
-There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-Simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
-In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
-
-% The use of game engines for AV testing
-Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
-%
-CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
-
-% Requirements of simulation based verification SBV
-State-of-the-art game engines provide a convenient option for simulation-based testing.
-They offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
-%
-Furthermore, they are easy to set up and run compared to on-road testing and are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
-%
-%\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
-% 
-% GC Removing the following para:
-%For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
-CARLA currently offers various basic types of cameras and %ray-case based
-% lidar; extensions are still under development. 
-% \todo{GC: what is ray-case? KIE: Not sure. :(}
-%
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. %~\cite{Javaid2013,cyres}
-%
-% The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
-%
-% Game engines for SBV
-Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
-%
-However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
-
-% \subsection{Contribution}
-% In pursuit of an answer to this question we
-In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
-%
-In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
-% KIE: This is covered in the next sentence:
-% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
-%
-\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
-}
-%
-% By analysing actor position variance 
-We find that the CARLA simulator is non-deterministic.
-Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
-We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
-% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
-% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
-%
-CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
-%
-\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
-
-The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
-%
-\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
-In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
-}
-%
-Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
-%impact that using a game engine for AV simulation may have.
-% on verification tasks. 
-In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
-In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
-}
-%
-% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
-%We present a case study to illustrate the use of this method.
-The study highlights the effects of non-determinism of a game engine on simulation results.
-Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
-%
-%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
-%
-
-This paper is structured as follows.
-%
-Section~\ref{s:prelim} defines terms used throughout the paper and identifies when determinism is needed. 
-Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
-%
-Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
-%
-Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
-%
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
-
-\todo[inline,color=green!40]{GC New below}
-
-\section{Preliminaries} \label{s:prelim}
-
-\subsection{Definitions}
-
-Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
-\\
-
-\subsubsection{Determinism}
-% What is a deterministic system
-Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
-However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
-Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
-% 
-% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
-Specifically, we focus on the variance of actor paths within a given scenario.
-
-A simulation can be thought of as the process of generating or producing experimental data. 
-%
-In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data. 
-%
-If a simulation is deterministic, Fig.~\ref{variance_description}~(b), then there will be no variation in the generated output data, i.e.
-all future states are perfectly reproducible from prior states and actions.
-%
-However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data. \\
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
-        \caption{Non-deterministic}
-        \label{variance_description_a}
-    \end{subfigure}
-
-    \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
-        \caption{Deterministic}
-        \label{variance_description_b}
-    \end{subfigure}
-    \caption{Demonstration of variance, precision, tolerance and determinism}
-    \label{variance_description}
-\end{figure}
-
-% \subsection{Definitions: Precision, Variance \& Tolerance}
-\subsubsection{Variance, Precision \& Tolerance}
-
-We adopt terminology from the mechanical engineering and statistics domains to describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
-%
-% \textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
-% %
-% \textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-%
-% \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-\textit{Variance} is used here to define the spread or distribution, of the generated output data with respect to the mean value.
-\textit{Precision} is synonymous with \textit{variance} although inversely related mathematically.
-%
-Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed under the same conditions and actions.
-%
-\textit{Tolerance} is defined as the permissible limit of the variance, or in short the \textit{permissible variance}.
-
-% The engineering terms \textit{precision} and \textit{tolerance} suitably describe how the simulation data vary when the same repeated stimulus is provided to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}.
-%
-%
-% As an analogy, the simulator can be thought of as a manufacturing process, producing data and if tasked to repeatedly execute the same process we must ask how \textit{precise} can this process be and what, if any, will the \textit{variance} be?
-% when comparing the differences between multiple products.
-As an analogy, the simulator can be thought of as a manufacturing process that produces data.
-To determine the precision of this process, the output must be measured and analysed for differences when the process is repeated.
-%
-Those differences describe the spread or variance in the process output.
-A hard limit on the variance can then be defined, Fig.~\ref{variance_description}~(a), beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
-%
-% In relation to manufacturing, the precision of any computer based task is casually assumed to be absolute, for example one may question the manufactured precision of a physical component but not the computer assisted drawing that it was made from.
-%but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point. %\textbf{[ref]}. %
-%
-Real manufacturing fails to achieve absolute precision. Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.
-
-If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within tolerance by design.
-If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
-%
-
-\subsubsection{Accuracy}
-Precision and tolerance should not be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
-Therefore, we define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation.
-%
-For a driving simulation, the reference value 
-% , for which accuracy can be measured against, 
-may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
-%
-In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
-In most cases it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
-\\
-
-\subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) in a 3D environment recorded at regular time intervals.
-This definition could be extended to include other variables.
-%, but position is sufficient for this analysis.\\
-A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run. \\
-
-\subsubsection{Simulation Variance \& Deviation}
-If the simulator is non-deterministic then how can the simulation variance be measured?
-This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run.
-% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
-For example, actor position variance is a distance-based metric that can be derived from simulation traces.
-The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained.
-% 
-Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions.
-% {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
-%
-%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
-%
-Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
-%
-% more intuitive to comprehend when considering verification tests, e.g.
-vehicle close-passing.
-Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
-\\
-
-\subsubsection{Scene, Scenario \& Situation}
-We adopt the terminology defined for automated driving in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
-Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV, or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals.
-The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.
-A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
-
-% \todo[inline]{values - Does this mean stimulus or other parameters? GC - changed to parameters}
-
-% Verification and Determinism 
-% Why is determinism important for verification
-%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
-% KE Text
-\todo[inline,color=red!40]{old: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
-
-\subsection{When is Determinism needed?}
-Determinism is a key requirement for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
-%
-Thus, a deterministic simulator has zero \textit{variance}.
-
-A simulator with non-zero variance can no longer be considered deterministic. 
-%
-Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}. 
-%
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-%
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-
-% GC Text
-\todo[inline,color=green!40]{GC Alternative: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}.
-If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.}
-
-% This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
-The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
-\textit{Tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
-%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
-Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour.
-%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
-%
-
-If the simulation is non-deterministic, e.g.\ 
-% KIE No, whether the variance is or is not permissible is not what matters for non-det, it should be zero!
-%e.g. it has a non-permissible variance in,
-it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
-In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
-%
-
-% \todo[inline]{GC Edit: If the simulation is non-deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
-
-% Challenges of SBV
-When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs % related to non-determinism
-a high priority for games developers.
-It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
-However, the situation is very different for AV development and testing.
-Thus, our main research questions are:
-%
-\todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
-%
-{\em How can one assess whether a simulation environment is deterministic?} and 
-{\em How can one determine and control the simulation variance?}
-\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
-
-% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (AVs) on roads is a delicate if not utterly fascinating subject.
-Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.
-
-% Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.
-
-% It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
-
-% \noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
-
-% As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
-
-% ======================
-% This is all from "FrameworkAndChallenges" without using quotes 
-% \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system.
-Similarly, the use of simulation for traffic modelling is well established.
-Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate AVs\cite{FrameworkAndChallenges}.\\\\
-
-% While traffic-level simulators can model road layouts, they do not require as much realism or detail as AVs require.
-On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for AV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
-% Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
-% Consequently, one of the major routes AVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate AVs \cite{FrameworkAndChallenges}.
-% ======================
-
-% Robopilot, Capri, CARLA, Apollo, Airsim, Udacity etc.
-are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}.
-\\\\
-% Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
-% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of AV software.\\\\
-% \TODO{Remove this paragraph?} Determinism of game (physics) engines has not attracted much attention by engines' developers nor researchers.
-This previously was not important since in gaming, determinism is not critical.
-% Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
-% On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
-
-% This paper aims to investigate how deterministic physics engines are for usage in V\&V for AV simulations; and will cover the list of following points:
-% \begin{itemize}[leftmargin=*]
-%     \item Necessary background on game engines used in AV simulations.
-%     \item Review of sources of non-determinism in such engines.
-%     \item Provide a method of how the level of non-determinism can be investigated and defined.
-%     \item A case study showing the implementation of this methodology in an engine widely used in AVs simulations (Unreal Engine).
-% \end{itemize}
+\input{../source/body.tex}
+\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper.}
 
 % ***************************************************
-%  BACKGROUND
+%  Bib
 % ***************************************************
-\section{Background} \label{s:background}
-
-There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
-% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
-
-Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.
-%However, when these engines are used for AV development then deterministic behaviour is required.
-%
-Ensuring software operates deterministically is a non-trivial task.
-Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-%
-\todo[inline]{GC line added to introduction.
-We need a paragraph on how hard it can be to make a computation fully deterministic, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-%
-\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-}
-%
-This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
-%
-% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations.
-Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
-
-% ======= Overview of a Game Loop
-% \subsection{Game Engine Overview} \label{GameLoopSection}
-
-Central to a game engine are the main game logic, the artificial intelligence (AI) component, the audio engine, and the physics and rendering engines. For AV simulation, we focus on the latter two.
-%
-The game loop is responsible for the interaction between the physics and rendering engines.
-Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed~\cite{unity_event_execution}.
-% \footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}. 
-A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering)~\cite{GameEngineArchBook}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoopv2.pdf}
-\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}.}
-\label{GameEngineLoopDiagram}
-\end{figure}
-
-% \todo[inline]{AG: remove the clock symbol, possibly remove the ticks chart as this is confusing given they are not ticks but event markers}
-
-% The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment.
-For example, the throttle and steering angle from the controller of an AV.
-The update stage interprets the input from all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states.
-The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
-
-% \todo[inline]{Fig 1: update rates of which clock?
-GC-is a clock confusing as it does not run on a 'clock' in computer science terms, rather it is a timer which is used as reference to compare to a 'wall clock'}
-
-% \todo[inline]{Should there not also be a clock in the overall loop? GC-yes the wall clock but it may get confusing as that is a real clock and the game loop clock is a timer?}
-
-% \todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.
-Unity page on multi-threading \footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and a plugin for multi-thread management \footnote{\url{https://forum.unity.com/threads/task-for-unity-scripting-threading.462971/?_ga=2.228561631.331624405.1587372521-1435703724.1587372521}}}
-
-% 
-% It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below.
-
-% \todo[inline]{We should agree if the 'game clock' really is a clock or just times the computation for each game loop.}
-
-% \todo[inline]{For AG: We use the phase 'game clock tick'- where did the come from as I think it is misleading.}
-
-% As the game progresses, the game loop runs continuously according to the game clock. \todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
-%
-% At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
-This measures how far the game's clock is behind compared to the real world.
-
-The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
-This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}.
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}.
-% \footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}. 
-When the frame is rendered the game loop cycle returns to processing inputs.
-An intuitive and  more detailed description of the interplay between the physics and render cycles is given in~\cite{JohnAustinUnity}.
-
-% \todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}.
-When the frame is rendered the game loop cycle returns to processing inputs.}
-%
-% Then the user inputs are processed.
-There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
-Rendering occurs once the game clock has caught up with the real world clock.
-The process then repeats.
-
-The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
-The shorter this time step is, the smoother the interpretation of the physical dynamics will be.
-%processing time it takes to catch up to game clock.
-% and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
-%
-% Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
-%
-To use a fixed physics time step, the user's display refresh rate needs to be known in advance.
-This requires an update loop to take less than one render tick (one frame of real world time).
-Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$.
-However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}.
-%
-Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware.
-Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}.
-%
-If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics.
-%
-Thus, extrapolation between frames may need to be performed to smooth transition between scenes. 
-%
-Note that both residual lag and extrapolation could affect perception stack testing.
-% This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
-% \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
-%
-% \todo[inline]{Why is this part of the figure?
-It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.
-GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
-%
-% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
-In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines,
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
-% \todo[inline]{Explain how this lag may affect perception stack testing}
-
-% \todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?
-GC-suggest remove this part of the diagram, too confusing as there are not really any clocks or ticks - we should remove all mention of ticks too.}
-
-% The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
-If $dt$ is too large or irregularly spaced then the rendered motion may not be smooth and interpolation may be required at additional computational cost.
-
-% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines.
-These engines update at fixed intervals and render whenever they can, which is not steady.
-This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates.
-In this case, the engine will use extrapolation techniques to give an estimate of where the object should be.
-This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
-%
-\todo[inline,color=red!40]{old: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy.
-In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-}
-%
-Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-% 
-The gamer neither needs a faithful representation of reality (i.e.
-gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
-gamer accepts low precision).
-In contrast, high accuracy and precision are necessary for AV development, testing and verification.
-% The gamer neither needs a faithful representation of reality, i.e.\ the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance.
-%, i.e.\ can be low \textit{precision}.
-% %
-% In contrast, the precision required for AV development and testing is very high, especially for perception stack testing, but also, more generally, for obtaining reliable verification results and for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required tolerance %on the precision
-% at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is repeatable so that if any software bugs are found they can be re-played and the issue resolved.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-%
-\todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-The gamer neither wishes a faithful representation of reality, i.e.
-can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
-can be low precision.
-In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
-%
-\todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
-
-%\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
-This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer).
-For instance, see below, but I'm not sure whether this is the right place.}
-
-% ======= Sources of non-determinism
-\section{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}
-
-% The following discussion reflects the results of our investigation into the potential sources of non-determinism that could affect game engines.
-We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
-The following review discusses the potential sources of non-determinism that were found in the literature or found as part of our investigation into game engines.
-We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
-%
-\todo[inline,color=green!40]{GC Addition below}%
-%
-A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests}, although the AV simulation domain introduces its own unique challenges that were not considered in that paper.
-
-% MOVED TO INTRO \todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-
-\todo[inline]{We should also refer to literature like:\\
-Making System User Interactive Tests Repeatable: When and What Should We Control?~\cite{when-and-what-should-we-control} and
-Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.}
-
-\medskip
-
-% \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
-The main reasons that we think are causes of these engines' non-determinism are discussed below.
-\\\\
-
-% ======= Floating-Point Numbers
-\subsection{Floating-Point Arithmetic}
-
-Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
-Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value.
-This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
-%
-
-% SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-%{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
-%and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }
-
-% As a consequence, floating-point arithmetic is not associative.
-Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, a change of execution order, e.g.\ from $x+(y+z)$ to $(x+y)+z$, can lead to different results~\cite{Kapre2007}.
-%
-As a consequence, floating-point arithmetic is not associative.
-Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}.
-This is because rounding is being performed at different stages in the computation, depending on the respective execution order.
-%
-Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
-%
-Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}. 
-%
-%{\color{red} Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.}
-%
-%If calculations are performed on a single core, then concurrency of multiple threads is not an issue.
-\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
-
-% Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
-Calculation results may differ due to the nature of non-associative floating-point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-%
-% If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
-% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.
-}
-%
-% Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.
-
-% Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
-% \todo[inline]{state what assertions are (used for) GC-see below} 
-%In the field of AV verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
-%
-Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability.
-In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-
-In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-%{\color{red} Should this be IMPRECISION rather than incorrectness?}
-%
-It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
-%
-Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern.
-%, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
-%
-So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
-
-However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
-Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
-
-Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
-However, the same executable should still return the same output for identical input.
-
-In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.
-
-\subsection{Scheduling, Concurrency and Parallelisation}
-% \todo[inline]{is this at runtime?} %
-% info at:
-% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
-% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
-% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
-% https://en.wikipedia.org/wiki/Concurrent_computing
-% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
-Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy.
-A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-% \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
-%
-In principle, changing the scheduling policy and thread priorities may increase simulation variance.
-
-It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
-Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
-
-However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
-This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
-%
-Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-%
-\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-}
-%
-Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
-
-Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes.
-This may be decided based on factors such as throughput, latency or CPU utilisation.
-%
-If due to the CPU utilisation policy the same single-threaded script executes multiple times across different physical cores of the same CPU type,  then execution should still produce the same output.
-%
-This is because the processor cores are identical and any impurities across the bulk silicon and minor perturbations in the semiconductor processing across the chip that may exist should have been accommodated for in design and manufacturing tolerances.
-%
-However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
-
-Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
-%
-\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{rr} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
-}
-
-Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously.
-While this would be faster than a serial execution, the order of execution arising from program-level concurrency is often not guaranteed.
-
-Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}. 
-%\noindent\underline{\textit{Concurrency:}}
-% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
-%\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
-Consider also parallelism, e.g.
-in GPUs.
-Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism.
-Say which parts of a game engine could reasonably be expected to run concurrently.}
-%
-%Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
-This technique can lead to issues such as shared resource access conflicts due to mis-managed timing and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.
-% \todo[inline]{how? GC-see above.}
-% KIE: We only list the potential sources of non-det, I therefore removed this. 
-% although monitoring and detecting such events falls outside of the scope for this work. 
-%
-%\noindent\underline{\textit{Ego Vehicle Controller:}}
-%The ego vehicle controller may be a source of non-determinism if using complex control algorithms such as neural networks for visual processing and perception.
-A neural network can only be deterministic within in the bounds of the training data used to construct it.
-If the network is exposed to unseen data it may incorrectly classify legitimate scenes due to lack of training data.\todo{That alone won't cause non-determinism, i.e.
-the NN will misclassify exactly in the same way each time - not?}
-%
-%In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically, assuming no \textit{residual lag}.
-In this sense, control over weather and lighting in the game engine should be considered and controlled as well as any other dynamic elements to the environment, e.g.
-reflections from surface water.
-% these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
-% i.e.
-the utilisation of hardware is manipulated differently each time.
-This would then result in non-deterministic behaviours especially if it is combined with the floating-point problem discussed earlier.
-%
-% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
-Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available.
-This may lead to events happening out of order depending on the computational load and hence non-deterministic.
-%
-% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
-To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order.
-This is beyond the scope of our work and will not be covered in this paper.
-% \\\\
-% \noindent\underline{\textit{Physics and Rendering clocks:}}
-% \todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
-% % The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
-% Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
-This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
-% %
-% A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
-Do we not assume the same HW at this stage?
-GC- we are just listing sources at this point so should be hardware agnostic.
-also, yes, hardware is the same for us but we cannot control for resource utilisation and background processes.} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
-Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources.
-% %
-% AV specific platforms, such as CARLA, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
-% \todo[inline]{Add a summary/conclusion?}
-%\\\\
-
-\subsection{Non-Uniform Memory Access (NUMA)} \label{s:sources_numa}
-For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
-Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
-%
-Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower.
-%
-By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
-
-\subsection{Error Correcting Code (ECC) Memory}
-ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}.
-Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
-If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
-
-Any simulation hardware not using ECC memory that runs for 1000's of hours, typical in AV verification, is likely to incur significant CPU hours and is therefore subject to increased exposure to these errors.
-To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.\todo{Check this, please.}
-
-%\todo[inline]{That said, HW design simulation is deterministic, uses long simulation runs and thousands of CPU hours, but I've not heard that this is an issue.
-I can ask, though.
-;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.}
-
-\subsection{Game Engine Setup}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps, ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
-%
-%
-% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
-% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
-% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}~\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
-% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
-the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
-%
-With respect to perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water, ensuring textures are not randomly generated.
-
-\subsection{Actor Navigation}
-There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-% 
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{newton2016unreal}.
-%
-The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
-
-A navigation mesh is a fixed area of the environment where actors are free to navigate.
-%
-Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
-%.
-Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
-%
-However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment.
-These cannot be unknowingly modified and therefore can be considered fixed.
-
-Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
-%
-% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
-%
-% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
-% \todo[inline]{unclear :( GC- re-written}
-
-\subsection{Summary}
-We have investigated the potential sources of non-determinism affecting game engines 
-and explored the impact they may have on simulation variance. 
-%
-% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable.
-}
-%
-Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
-\todo[inline]{So, do commercial HPC facilities have ECC memory?}
-\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
-%
-\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
-To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
-\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
-%
-Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
-Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
-If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.
-
-% {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
-%
-Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
-%
-However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop.
-Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism.
-%
-
-%
-%The ego vehicle controller can be considered deterministic if dynamic elements to the simulation environment are controlled, such as weather and lighting that may be on random cycles. 
-%
-% ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software.
-Actor navigation not a problem if navigation meshes are fixed.
-Game engine, ensuring fixed $dt$ and control of random numbers.
-Core or thread scheduling will not be issue unless changing scheduling policy or migrating between machines that have different policies.
-Concurrency is really down to the development of the simulation software and reputable companies should have good control over these potential issues following good software development practice.
-NUMA should only affect interprocessor data transfer cost, at wrt only make computations longer.
-relative times between access different caches is likely to be small although may impact on high throughput, e.g.
-HPC.
-VFR - want to remove this section.
-EGO vehicle deterministic?
-Can't control library change, perception stack
+\printbibliography
 
 % ***************************************************
-%  Empirical Investigation
+%  Bio
 % ***************************************************
-\section{Case Study of Simulation Variance} \label{s:case-study}
-
-We present an empirical investigation into using game engines for
-simulation-based verification of autonomous vehicles with a focus on
-characterising sources of non-determinism in order to understand  the impact
-they have on simulation variance. 
-%
-\todo[inline,color=green!40]{GC new: line below.}
-%
-Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
-%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
-%
-Ultimately, our objective is to control non-determinism to minimise simulation variance.
-
-\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
-%
-We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-A discussion on the internal and external settings of the simulation is included, along with system configuration and pre-screening sections.
-% {\color{red} and then..? Please cover the entire section - consider splitting setup from evaluation.}
-%
-\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
-%
-\todo[inline]{Complete the section structure as this is a very long section.}
-
-% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
-\end{figure}
-
-This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-%
-Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-%
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-\todo[inline]{Decide past or present tense and be consistent throughout.}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-\subsection{Tolerable Simulation Variance}\label{s:tolerance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
-
-To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-%
-Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
-%
-% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
-%
-The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small such that assertion checking would fail with minor computational perturbations.
-% {\color{red} {\sc What would fail?}}
-%
-Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
-
-% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-%
-% but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-% 
-% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
-%
-For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
-%situations such as potentially dangerous close passing of a cyclist. 
-A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
-% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected. 
-% Thus any variation of less than $1$cm is tolerable. 
-Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-
-% considered to be sufficient for assertion checking in urban environments. 
-%
-% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
-In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-%
-\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
-where any value higher than the specified tolerance is considered non-permissible. }
-
-% ***************************************************
-%  Actor Collisions
-% ***************************************************
-
-\begin{table*}[b]
-\centering
-\begin{tabular}{clclccc}
-\toprule
-Test & Actors 					 	  & Collision 	 & Collision Type 			 & $n$  & $\max\sigma$ (m) & $\max\sigma$ (m) \\ 
-	 & 								  & 			 & 							 &	    & (unrestricted)  & (restricted)	\\ \midrule
-1    & Two vehicles                   & No  		 & N/A 			  			 & 1000 & 0.03 & $7.0{\times}10^{-3}$ \\
-2    & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 1000 & 0.31 & $9.8{\times}10^{-3}$ \\
-3    & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 1000 & 0.07 & $5.2{\times}10^{-4}$ \\
-4    & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 1000 & 0.59 & $1.5{\times}10^{-12}$ \\
-5    & Two pedestrians                & No  		 & N/A 			  			 & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-6    & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-\bottomrule
-\end{tabular}
-\caption{A description of the test scenarios showing the test number, the actors included, whether a collision occurred and if so then between which actors.
-$n$, the number of repeats is set to 1000 and $\max\sigma$ is the maximum simulation deviation.
-The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation.
-To understand the impact of collisions and high resource utilisation, the \textit{restricted} column shows a subset of the results where post-collision data and experiments above 75\% resource utilisation have been removed.}
-\label{TableOfExperiments}
-\end{table*}
-
-\subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}.
-%
-Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were identified as potentially key aspects to the investigation into simulation variance.
-
-The tests used for the case study are listed in Table~\ref{TableOfExperiments}. They cover a range of interactions between actor types. 
-%
-Tests 1 \& 2 involve two vehicles meeting at a junction where they either do not collide (Test 1) and where they do collide (Test 2), thereby triggering an \textit{event hit} callback in the game engine.
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
-% \todo[inline]{AG: is this value correct or should it be 2.0m?
-AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
-actors would ignore trajectory waypoints that are less than 2.0m away from them.}
-intervals using a PID controller. 
-%
-In Test 3 a mixture of different actor types is introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point. 
-%
-In Test 4 this pedestrian collides with one of the vehicles at the crossing, triggering an \textit{event hit} callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
-Tests 5 \& 6 involve only pedestrians that either do not collide (Test 5) and that do collide (Test 6), see Fig.~\ref{Test_b}.
-
-\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
-
-% ***************************************************
-%  Experiment Description
-% ***************************************************
-\subsection{Evaluation Metric}\label{s:Experiment_Description}
-% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time. 
-%
-The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.
-
-Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
-%
-\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for each actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-}
-
-The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
-%{\color{red} This is still a function over time - should it be?}
-\begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a^{2}(t)
-\end{equation}
-The maximum deviation is the absolute value of the square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity. 
-% {\color{red} So, ``a'' is any actor or a specific actor?
-Likewise, ``t'' is any time point or a specific time point?
-NOTE that any actor any time means a in A and t in T, where A is the set of actors and T is the set of all time points.}
-
-The maximum deviation, $\max\sigma$, can be analysed for the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation is sufficiently accurate for verification purposes.
-% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
-% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
-\todo[inline,color=green!40]{GC Addition: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
-}
-%
-% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
-% \todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
-
-% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
-Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
-
-% This whole process is repeated for various internal and external settings (if needed). 
-% Then various plots can be created to draw the engine's simulation variance. 
-% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
-
-% ***************************************************
-%  Internal Settings
-% ***************************************************
-\subsection{Simulator Settings}
-
-Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
-Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.\ \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.\ \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
-% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
-An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
-The results were compared to a baseline of the default engine settings.
-%
-However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
-Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.
-
-% \subsection{Initial Results and Discussion}\label{InitialResultsSection}
-
-% {\color{red} This is for unrestricted, so this does not come as a surprise later.}
-
-% ***************************************************
-%  External Settings
-% ***************************************************
-\subsection{External Settings}
-% \todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
-
-% After discovering that internal engine settings could not achieve a simulation variance that met the required 
-% \textit{deterministic threshold}, 
-% \textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
-After examining controls the internal game engine, external settings were explored. 
-%
-Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
-%
-For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
-
-Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load.
-This initial observation was explored more fully in this work using the CARLA platform.
-
-% {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented.
-Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}
-
-To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation.
-%
-% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
-Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
-Resource utilisation figures reported here should be considered approximate values.
-%
-Resource utilisation was capped to 75\% in some parts of the results and referred to as \textit{restricted}.
-This was done to limit the load on the test system and hence limiting simulation variance.
-The term \textit{unrestricted} places no such limit on the system load.
-
-% {\color{red} To understand the impact of collisions and high resource utilisation, the restricted column shows a subset of the results where post-collision data and experiments above 75 resource utilisation have been removed}
-%
-% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
-%
-% {\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
-
-\todo[inline,color=green!40]{GC Addition (para below):} 
-%Issues around floating-point precision and concurrency as previously mentioned notwithstanding, 
-Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
-For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
-In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency-induced imprecision~\cite{matlab_parallel_computing}.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV6.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
-Tests 5 and 6 overlap having almost identical results.
-Note that the lines between data points are only a guide.}
-    \label{ExperimentsStressSummary}
-\end{figure}
-
-% ***************************************************
-%  System Configuration and Screening
-% ***************************************************
-\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
-The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
-%
-The operating systems was Linux Ubuntu 18.04.4 LTS. 
-%
-Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
-
-% Each test was repeated 1000 times. 
-Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
-While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results.
-Therefore each test was repeated 1000 times to provide a sufficient sample size.
-% {\color{red} Justify why 1000 - see my later comment in the Method section.}
-%
-A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
-%
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:
-
-\begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, no failures on all 11 tests~\cite{shi2009testing}.
-\end{itemize}
-
-% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
-These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
-
-% \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
-
-% Our hypothesis is that since UE4 is a game engine it is non-deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
-This behaviour is expected not to hold as the level of computing stress increases.
-
- % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
-To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
-(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
-
-% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
-
-% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
-
-% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
-
-% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
-
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
-
-% ***************************************************
-%  Results and Discussion
-% ***************************************************
-\section{Results and Discussion}\label{s:FinalResultsSection}
-
-\begin{figure*}[t]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
-        \caption{}
-        \label{CarsCollisionCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
-        \caption{}
-        \label{CarsCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to vehicle collision (Test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
-The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
-\end{figure*}
-
-\begin{figure*}[h]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
-        \caption{}
-        \label{CarsPeopleCollsionsCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
-        \caption{}
-        \label{CarsPeopleCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to pedestrian collision (Test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
-\end{figure*}
-
-% {\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}
-
-A summary of the main results are shown in Table~\ref{TableOfExperiments}.
-In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
-%
-From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-%
-However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
-%
-\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-}
-%
-Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
-
-Resource utilisation was found to have a significant impact on \textit{simulation variance}.
-%
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
-%
-Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-%
-\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
-%
-
-A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
-
-Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision.
-%
-The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-
-\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-}
-% \todo{Why ``not reliable''.}
-
-The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (Test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three observations; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
-% as described above. 
-Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}.
-Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers, developers and researchers in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous simulation results.
-% not just assertion checking
-
-% \todo[inline]{Make sure that assertion checking has been introduced before.}
-
-% \textbf{Strange gap post-collision for few seconds}
-
-% \textbf{Should we rename 'test' to 'scenario???'}
-
-% \todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
-
-% Scenarios that include both collisions and vehicle actors were non-deterministic.
-However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
-
-% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
-Note the dotted red line in the plot shows the millimeter (mm) level.
-Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
-%
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v8.pdf}
-    \caption{Stages of the method proposed to determine the variance of a simulation.}
-    \label{method_diagram}
-\end{figure*}
-
-\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
-This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
-Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
-
-To conclude, the main findings of this case study suggest a working practice that would minimise the factors that give rise to the non-deterministic effects observed in this investigation. 
-%
-By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table~\ref{TableOfExperiments}.
-By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve.
-%
-Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-
-A correlation between resource utilisation and simulation variance has been observed in these results which may be due to process scheduling.
-Rather than limiting utilisation of the whole system, another approach may be to promote the scheduling priority of the simulation process which is explored in Section~\ref{r:process_scheduling}.
-Furthermore, we were keen to investigate the impact of memory access and include a brief investigation of this in Section~\ref{r:numa}.
-%
-\todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-%For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm. 
-}
-
-\subsection{Process Scheduling Priority} \label{r:process_scheduling}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
-    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
-    \label{NICEExperimentStressSummary}
-\end{figure}
-
-An investigation into the impact of process scheduling on the simulation variance was undertaken following the observations of increased simulation variance with increased resource utilisation.
-The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
-%
-To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.
-
-The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range of deviation, non-outlier limits by the whiskers and a horizontal bar for the median.
-The figure shows that decreasing the priority of the simulator process (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot).
-Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results.
-This can be seen in the maximum priority setting where the bars in the plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
-%
-The remaining difference in the variance between the two resource utilisation levels may be due to the lack of absolute control that \texttt{NICE} has over process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
-
-\subsection{Non-uniform memory access (NUMA)} \label{r:numa}
-An additional investigation into memory access was undertaken given the potential impact that data transfer cost and execution order may have on simulation variance, see Section~\ref{s:sources_numa}.
-%
-The program \texttt{numactl}~\cite{numactl_NUMA} allows a process to run with a specified memory placement policy, essentially allowing the process to be bound to a particular CPU core. %
-\texttt{numactl} was used to fix the simulator and test script to single cores, and a (2\%) improvement in simulation variance was observed. %
-This was considered minor comparative to the changes in simulation variance observed in other aspects of the case study, for example the pre- and post-collision between a vehicle and pedestrian saw a change of $10^{14}$.
-Therefore \texttt{numactl} was not used for subsequent testing.
-
-However, practitioners that are aiming to minimise simulation variance should use a technique such as this to minimise data transfer cost and potentially minimise any potential effect on process execution order that may lead to increased simulation variance.
-
-\subsection{Investigation Summary} \label{s:empirical_summary} 
-% {\color{red} Factor into overall conclusion of the case study}
-% \todo[inline]{GC-need to reference the 'restricted' set of results here too}
-
-%\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
-The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator.
-Collision events may utilise more computational resources than simulations without such events, i.e.
-the physics loop takes longer to process as there are complex physical interactions to resolve.
-This may coincide with the second observation that higher resource levels also lead to non-deterministic results.
-We hypothesise that both these observations are due to improper management of thread scheduling within the engine.
-When resources are occupied there may be a delay in processing critical calculations for the simulator, which may arrive out of an expected sequence or the engine compensates for this delay by reducing the number of physics loops that can be performed within a render cycle.
-(is this 2nd one true, or does carla ignore render requests?)}
-These empirical investigations have highlighted the shortcomings of using a games engine for simulation based verification and suggests advice for best working practice. 
-%
-It was observed that resource utilisation positively correlates with simulation variance and that specific simulation events, such as vehicle collisions, can also lead to a breach in permissible tolerance.
-%
-The investigation found that the effect of higher simulation variance as a result of increased resource utilisation can be reduced, but not omitted entirely, by controlling the scheduling policy. 
-%
-The investigation into specifying memory placement on the simulation process did improve simulation variance but only by a minor amount of 2\%.
-
-However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly. 
-%
-Therefore we have derived a general methodology that practitioners can follow to find the 
-% \textit{deterministic domains of operation} 
-\textit{operational domains of permissible variance} 
-for %any hardware configuration.
-a game-engine-based simulation environment. This methodology is presented in the next section.
-
-% First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
-with in the centimeter level).
-This clearly shows how the load on a computer (i.e.
-stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
-This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
-
-% Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
-with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
-
-% Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
-We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
-
-% Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
-Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
-
-% Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
-However, in V\&V of AV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
-Thus, the post collision effect should not be of much concern in the context of AVs.
-
-% It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
-The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
-Thus failing to stay below our red line and the test should be terminated.
-In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
-left alone and not used for any other tasks while the runs were executed.
-
-% ***************************************************
-%  METHOD
-% ***************************************************
-\section{Method to determine the Variance of a Simulation} \label{s:methodology}
-% In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
-This method can be used to determine the operational domains of permissible variance of a simulation environment.
-%
-In this section a method for determining the simulation variance of actor paths and resolving the operational domains of permissible variance is presented here as a work flow, see Fig.~\ref{method_diagram}.
-%
-In addition, recommendations and best practice guidelines for minimising simulation variance are suggested.
-% , such as fixing the physics time step in \textit{Simulator Settings}. 
-% {\color{red} \sc Highlight these in the text.}
-% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
-% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}.
-%These need to be considered at different stages in the process.
-%
-
-The method consists of a sequence of five stages; experimental design, simulator settings, external settings, execution and analysis.
-In the following, each stage is described in detail with reference to the items listed for each stage in Fig.~\ref{method_diagram}.
-
-\subsection{Experimental Design}\label{s:design_experiment}
-
-% {\color{red} \sc Surely, a series of systematically designed experiments, each repeated X times, should be set up, where each experiment varies one of the suspected sources of non-determinism?
-Then, the analysis of the data from each experiment should reveal the actual sources of non-determinism for the given setting?
-In other words, we are aiming to fix as much as we can control (easily), and then we intentionally vary (ideally) single variables, e.g.\ the resource utilisation, to find out what makes a difference.
-We need a clear definition of the objectives of the experimental design, i.e.\ what one should vary and what one wants to ``nail down''.}
-
-% {\color{green!99} \sc }
-The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
-Each test is repeated $n$ times and the analysis of the results provide a confidence with a degree of statistical certainty.
-By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and addressed.
-
-\subsubsection{Actors} \label{s:actors}
-
-% Different actor types may be handled differently by the game engine.
-For instance, our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
-However, the introduction of CARLA vehicles increased the simulation variance of the pedestrian-vehicle test by 12 orders of magnitude.
-{\color{red} \sc This needs rephrasing as it appears to contradict the statement above.}
-%
-All actors that could be included in the simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see Section~\ref{ego_actor}. 
-% {\color{red} \sc ... such as the ego vehicle...? [see section on ``External Settings'' last point - Ego Controller.]}
-%
-\subsubsection{Trajectories and Collisions} 
-% Actor paths should be hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
-Actor paths, or the sequence of actions required to generate paths, should be hard-coded to ensure repeatability.
-These paths should include collisions between actors and potentially collisions between actors and static scenery if this is likely to occur in the simulation or as part of the verification process.
-Actor paths without collision are also important to include as these will serve as a baseline to the other tests; to see if deviations between runs in actor paths increases with collisions.
-% {\color{red} \sc What baseline?} {\color{green} \sc Done}
-
-\subsubsection{Permissible Variance} \label{s:threshold}
-The verification engineer should set the \textit{permissible variance}, also termed \textit{tolerance}.
-% which refers to the acceptable level of actor path variance. 
-% ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. 
-%
-The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection, but not so small for assertion results to differ for repeated runs.
-In Section~\ref{s:tolerance} a tolerance of $1$cm was considered sufficient for urban scenario assertion checking.
-In practice, it may be necessary to determine this tolerance experimentally.
-%
-%The permissible variance is analogous to the analogue signal level associated with each of the binary states %of the signals in digital circuits.
-At some low analogue voltage the signal is interpreted as binary $0$, this may be, for example, any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
-This convention enables the abstraction from the noisy physical signal values that are analogue voltages into the digital domain with binary representation.
-<- KIE: meant only to illustrate that other areas also have defined tolerances.
-%
-
-% \todo[inline]{Make sure assertion checking has previously been introduced.}
-
-% The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
-These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
-
-% \todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''. GC change callback to 'event hit'}
-
-% The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
-The aim of the experiment is to determine if the simulation is deterministic.
-This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories.
-Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.
-
-% \todo[inline]{We may need to introduce the Case Study first, then derive the general Methodology from it.
-This way, we can explain the requirements of the case study, how things were set up, what was observed, how the methodology was refined, several iterations even, etc.}
-
-\subsubsection{Simulation Time} 
-% {\color{red} \sc Should this not be part of the Execution section?} 
-The simulation time will depend on the actor paths and terminating conditions.
-The simulation time should be sufficient to record interactions between actors but not so long that the testing take an inconvenient time to complete.
-In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25} in Section~\ref{s:FinalResultsSection}.
-The termination conditions of the simulation can be set by, for example, actors reaching their final trajectory waypoints.
- % {\color{red} \sc Point to the section that specified that.}
-
-% Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
-This is to ensure that an outlier data-point is not hidden in the average of $n$-tests.
-
-\subsection{Simulator Settings}
-The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
-If using CARLA, a small fixed value, say $0.05s$, can be set by using \texttt{setting.fixed\_delta\_seconds = 0.05}.
-In Unity, the default fixed time step is set to $0.02s$~\cite{MonoBehaviour_unity}.
-
-In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
-
-The use of random numbers must be controlled through fixed seeds, resulting in pseudo-randomness.
-Random numbers might be used in the simulation environment to control variations of background effects, e.g.\ weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment.
-%
-Actors that navigate through the environment should use a fixed navigation mesh. 
-%
-The version number of the CARLA and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}.
-Therefore, ensuring a consistent version number throughout testing is also important.
-
-% \todo[inline]{Select section titles to match the flow diagram.
-In each of these sections, go through the matching phase in the flow diagram, explaining each point listed there and justify recommendations.}
-% The experiment should be a short\todo[inline]{What does short mean? Quantify.} scenario that can be easily repeated with the same initial conditions.
-
-\subsection{External Settings} \label{ego_actor}
-
-\subsubsection{Resource Utilisation}
-% To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-The resources available to the simulator have been shown to have a significant effect on the variance of the path of simulated vehicles. %and exploring this as a variable the analysis is required. 
-%
-Thus, it is important to understand at what level of resource utilisation the system running the simulation becomes susceptible to \textit{simulation variance}.
-
-CPU utilisation software, such as the linux \texttt{stress} tool, which is a workload generator program, can be used to spawn workers on any number of cores or virtual threads on a system.
-This can be used to artificially increase the load on the system.
-For GPU utilisation, \texttt{gpu-burn} can be employed using the \texttt{fur test}.
-Different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
-Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU, respectively.
-These values should be added to the data logs.
-%
-Alternatively, in place of artificial resource utilisation, multiple instances of the simulation could be executed simultaneously.
-However, the granularity of control with this approach may be reduced.
-
-% \todo[inline]{Again, this sounds as if having the case study first makes more sense. That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
-
-% \todo[inline]{This should be ``External Settings'' - from flow diagram?}
-
-\subsubsection{Memory Testing}
-Prior to experimental execution the system hardware should be tested for memory conformity and to ensure no single bit errors are occurring, see Section~\ref{s:screening}.
-For mainboard memory \texttt{memtest86}\ can be used on most platforms to run a series of pre-defined memory test patterns.
-This memory testing software can also be used for ECC enabled hardware.
-Similarly, to test memory on Nvidia based graphical adaptors \texttt{cuda\_memtest} can be used to ensure no memory errors exist.
-
-\subsubsection{Scheduling}
-We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
-However, gaining fine control over the scheduling policy and thread execution order is known to not be trivial~\cite{acm-q-rr-interview}.
-
-The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority.
-Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation.
-
-% If the policy or operating system cannot be changed then s
-Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
-
-\subsubsection{NUMA Control}
-Control over a Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multi-core processors with shared memory.
-This control allows the simulator process to be fixed on a single core, reducing and unifying memory access time.
-%
-Investigations in tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}. 
-%
-% Using \texttt{numactl} allows a process to run with a specified memory placement policy, fixing the process to a single core.
-This will reduce the data transfer cost and the opportunity for disrupted execution order that may be responsible for increase simulation variance.
-%
-Using a fixed memory placement policy may assist if simulation variance is borderline to the tolerance but only of the order of a few percent from our observations.
-
-\subsubsection{Ego Vehicle Controller} 
-% {\color{red} \sc Why is this not covered under actors? [see my note in the Actors section]}
-% {\color{red} \sc Clearly separate the investigation from the Method.
-The fact that the case study did not have an ego vehicle should be mentioned as a limitation of the case study, rather than elaborated on here, but why does this make a difference?}
-%
-An ego vehicle was not used in this study, but the impact of introducing this to the simulator can be considered here.
-The ego vehicle is seen as another actor in the simulation but care must be taken to ensure that any control algorithms, machine learning modules and processing pipelines are deterministic.
-We recommended this be treated as a separate source of non-determinism and handled accordingly.
-
-% ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
-{\color{red} \sc Long and cumbersome sentence with strange grammar: No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
-}
-
-\subsection{Execution}
-
-\subsubsection{Sample Size}
-% Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} 
-% Histogram to show how sampling with position binning varies?
-% \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
-%
-% {\color{red} \sc Again, separate the study from the method.
-This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.}
-It is recommended that the chosen sample size is determined empirically. 
-%
-We recommend monitoring variance while increasing the sample size in orders of magnitude
-until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process. 
-%
-Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
-
-% By increasing the sample size and observing that the simulation variance is stable, one can infer that a sufficient number of samples have been collected.
-If the variance changes with a change in sample size then a stable condition has not been found.
-% ...
-% ...
-%  dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
-% % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
-
-\subsubsection{Data Logging}
-Unique identifiers should be assigned to each experiment, each repeat and each individual actor. 
-The time-stamped actor positions should then be recorded at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-Additional information should also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
-% such as the \texttt{unit} command in Unreal Engine. 
-%For the subsequent analysis, each actor should have an identifier. The repeat number and simulation time should also be captured. 
-% The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
-For example, if 100 tests are executed and the maximum deviation in actor path is $<1$cm, then this scenario could be interpreted as being deterministic above this scale.
-
-\subsection{Analysis}
-For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment.
-
-%Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
-% KIE: Why now point to HW? And, which "boundaries"?
-
-% Game engine actors can be used to simulate the AV and other vehicles as well as pedestrians.
-For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
-
-% \todo[inline]{This is mixing our experiments (which are part of the case study) with the methodology.
-It is probably best to first introduce the case study and then derive the methodology from it, rather than mixing both.} In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
-Initial testing also indicated that when actors collide there is a significant change in \textit{simulation variance}, therefore the experiment should include some scenarios where collision callbacks are activated.\todo[inline]{Make sure collision callbacks have been introduced earlier.}
-
-% Randomisation must be fully controlled by fixing the seeds for any random number generation that is used during the experiment.
-It is important to identify and control any external sources that may use randomisation, e.g.\ external vehicle controllers or other clients connected to the simulation environment.
-
-% \todo[inline]{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
-
-% \todo[inline]{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
-%
-% For the vehicles in the simulation, waypoints are generated and interpolated in the same way but execution at runtime is slightly different where a PID controller is used to generate a velocity profile and steering angle requests to reach the destination targets.}
-
-% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
-Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.
-A* search algorithm.
-
-% \subsection{Internal settings}\todo[inline]{Systematically go through the points listed in the flow diagram, explain and justify.}
-% Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
-
-% By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronous mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine.
-% Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
-However, in such applications of V\&V of AV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
-% If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the simulation variance as a result of allowing the AI to navigate in a more well defined space.
-% Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
-% The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.
-% Running in editor or deployed mode can sometimes cause differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
-
-% \begin{table*}[b]
-% \centering
-% \begin{tabular}{clclcc}
-% \toprule
-% Test ID & Test description (See Figures~\ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
-% 1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
-% 2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
-% 3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
-% 4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
-% 5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
-% 6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
-% 7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
-% 8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
-% \bottomrule
-% \end{tabular}
-% \caption{Set of experiments}
-% \label{TableOfExperiments}
-% \end{table*}
-
-% The user should also monitor and record system CPU and GPU utilisation rates.
-Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
-
-% Collision callbacks
-% \subsection{Other Considerations}
-% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
-ii) If not, how does its simulation variance vary?
-In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
-Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
-% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
-Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
-% The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
-For the majority of AV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
-
-% \subsection{External settings}
-% Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.
-Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
-Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.
-
-% \subsection{Running experiment}
-% \noindent When it comes to running the experiment itself, things that one should consider are the problem space exploration; the number of runs that should be executed to get reliable results; the frequency of logging data of actors; and possibly having a program to monitor the hardware utilisation.
-% Problem exploration, there is too many variables that one can play with. do statistical analysis some.... some of them get problematic...
-
-% ***************************************************
-%  CONCLUSION
-% ***************************************************
-% \section{Conclusions and Future Work}\label{s:conclusion}
-\section{Conclusions \& Future Work}\label{s:conclusion}
-
-% \todo[inline]{Say what the paper investigated and what we aimed to achieve, why this is important and how we went about it.
-Then point out the derivation of a re-usable methodology based on the knowledge and understanding we gained from the case study.}
-
-% \todo[inline]{GC: we need to say what we set out to achieve and why that is important, then say how our investigation led us to several theories which resulted in understanding several key parameters which directly affect simulation variance.
-From this we developed a methodology that the community can use to understand the deterministic operational domains of their own specific hardware:  Autonomous vehicle testing is adopting the use of game engines for development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-An empirical investigation uncovered several parameters that contributed to simulation variance ...
-explain results ...
-This led to the design of a methodology allowing 'users' to determine the deterministic operational domains for a specific hardware/platform configuration ...
-}
-
-%The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
-Game engines offer simulation environments that are used for the development and verification of autonomous driving functions. 
-%
-Determinism of a simulator is required to achieve repeatability, which is essential to find and fix software bugs efficiently, and to ensure simulation results are trustworthy. 
-%
-If a simulator is non-deterministic then practitioners should at least be aware of, and know how to find, the operational domains where \textit{simulation variance} is tolerable.
-
-An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation.
-%
-We then researched, identified and discussed potential sources for non-determinism in this context.
-%
-A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution.
-In particular, actor collisions and system-level resource utilisation were identified as key contributors.
-%
-To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
-%
-% {\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.
-This seems to have been commented out but was here before.}
-
-A general method to assess the actor path variance of a game-engine-based simulation environment is then proposed.
-The method can be used to find the \textit{domains of permissible variance} of a simulation environment for a given system configuration.
-This can give AV developers and verification engineers increased confidence in the simulation results and reduce debug time.
-% 
-% \subsection{Future Work}
-%
-As future work the method can be extended to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
-
-An ambitious avenue for future work is the development of a deterministic simulator for AV development and verification.
-This requires controlling all potential sources of non-determinism, including randomness and scheduling, very similar to the development of the record-and-reply debugger \texttt{rr}~\cite{RR_link}, originally developed to catch low-frequency non-deterministically failing tests at Mozilla~\cite{acm-q-rr-interview}.
-
-%With the advent of more AI systems becoming adaptive, meaning that the system may change it's response to the same stimulus over time, the notion of repeatability will need reassessing in the context of verification.
-The main research question here is what will constitute `the same' which will be important for passing or failing verification tests but also for determining coverage.
-
-% \todo[inline]{Review/rewrite this to match the earlier changes in the paper.
-Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
-Potential sources of non-deterministic simulation behaviour have been discussed such as floating-point rounding error and scheduling policy.
-A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the CARLA driving simulator.
-Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study.
-Furthermore it was found that all actors, even those not involved directly in a collision, will behave non-deterministically if a collision callback to the game engine is made.
-Best operating practice follows that the simulation should be stopped as soon as a collision callback is detected or resource utilisation levels exceed 75\% although this value may be system specific.
-}
-
-% \todo[inline]{GC: think we should put more emphasis on the outcomes of the method rather than the cast study.}
-% \todo[inline]{Point out (again) that all data and scripts etc are on github.}
-
-% \todo[inline]{Future work...}
-
-% In this paper we explored the determinisim of game engines.
-This is important for autonomous vehicles' development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, uncovered several parameters that contributed to simulation variance.
-These parameters include numerous internal settings based in Unreal Engine, and others that are external mainly dependant on CPU and GPU utilisation.
-
-% Changes in CPU and GPU resource utlisation has shown to be the main influencer for the non-deterministic behaviour of game engines.
-This is clearly demonstrated in the results (see section \ref{s:case-study}), where the max deviation between runs increased from the orders of $10^{-12}$m to devaitons of up to 1m as the CPU and GPU resource utilisation increased above 50\%.
-
-% The term \textit{deterministic operational domain} was given to the range of system resource utlisations that provides low simulation variances that is insignificant for development and verifcation of autonomous vehicles.
-% The deterministic operational domain specified herein is specific to our hardware configuration.
-
-% This led to the design of a methodology allowing `users' to determine the deterministic operational domain for a specific hardware configuration.
-This will help guide specialists in the field of AVs to setup their custom experiments for their specific simulators and identify their specific deterministic operational domains.
-
-% Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
-Hence allowing its extensive usgae by both the autonmous vehicles' and the gaming community.
-
-% presented an investigation (based on CARLA driving simulator) to show whether game engines are deterministic or not. 
-% Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, was used to illustrate the challenges of determinisim faced by game engines. 
-% % It was found that game engines are non-deterministic.
-
-% We reviewed possible sources of non-determinisim in section \ref{s:background} and investigated the effect of each on the simulation variance.
-
-% We found that the sources of non determinsim that had a significant effect on simulation variance were attributed to CPU and GPU resource utilisation. 
-% The case study clearly shows how the max simulation variance changes from very low simulation varicances, below picometer level ($10^-12$), to centimeter level when the CPU and GPU resource utilisation increases above 50\%.
-
-% We found two main outcomes from our investigation: 
-% 1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}, e.g. runtime scheduling. 
-% 2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.
-% Thus, by altering these settings one can achieve sufficiently low levels of simulation varince that can allow users to assume the simulation is deterministic. 
-% We phrased the level below which the simulation is to be assumed determinisitic as the \textit{determinisitic threshold}, which is up to the user to define where it exactly lies. 
-% This allows for one to use these game engines knowing that their simulation will be sufficiently deterministic provided that they bound their simulation within a certain configuration.
-
-% The configuration presented in here is specific to the hardware and software used in our investigation. 
-% Therefore, we outlined a methodology in this paper to allow practitioners in this field of AVs to resolve the non-deterministic operational domains of their simulation.
-
-% We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the simulation variance of a game engine used in V\&V application of AV simulations.
-The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
-
-% A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
-It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
-%since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
-
-% **** These results are not meant as a universal guide but merely to illustrate our method for a single hardware configuration.
-It is probable that the values quoted here will not translate to individual end-users unless the system configuration is very similar.
-This paper should be used as a method for users to determine what scenarios and system setting are appropriate for their particular verification needs.
-
-% You must monitor all collisions, not just for the ego vehicle as any collision callback may result in high simulation variance and hence non-deterministic results for your experiments.
-
-% In highlight of obtaining a fully deterministic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
-Alternatively, another approach would be to create a new engine that is designed to be deterministic with minimal attention given to performance.
-
-% ***************************************************
-%  FUNDING, REFS
-% ***************************************************
-% \section*{Acknowledgements}
-% This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
-% projects are part-funded by the Centre for Connected and Autonomous
-% Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
-% 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-\printbibliography
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{Other/mugs/greg.png}}]{Greg Chance}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/greg.png}}]{Greg Chance}
 Greg Chance received the B.Sc.
@@
 
-\begin{IEEEbiography}{Abanoub Ghobrial}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/abanoub}}]{Abanoub Ghobrial}
+Abanoub Ghobrial received the MEng degree in mechanical engineering from the University of Bristol, Bristol,  U.K., in 2018.
+He is currently pursuing the Ph.D.
+degree in computer science at the University of Bristol and is a part-time Research Associate with the Trustworthy Systems Lab, Bristol,  U.K.
+From 2018 to 2020, he was a full-time Research Associate with the Trustworthy Systems Lab.
+His current research interests are techniques to allow self-managing of autonomous safety-critical systems via continual learning during operation; and the development of simulation-based verification techniques for autonomous systems.
 \end{IEEEbiography}
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{Other/mugs/kevin.jpg}}]{Kevin McAreavey}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/kevin.jpg}}]{Kevin McAreavey}
 Kevin McAreavey received a BSc (Hons) in computer science from Queen's University Belfast (QUB) in 2010 followed by a PhD in AI in 2014.
@@
 
-\begin{IEEEbiography}{Tony Pipe}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/tony}}]{Tony Pipe}
+Tony Pipe received his BSc (Hons.) degree in Electronic Engineering from Warwick in 1979 and, after an eight-year period in industry that included running a small company, received his PhD in Robotics from the University of the West of England (UWE) in 1997.
+Since 2010 he has been a Professor of Robotics and Autonomous Systems at UWE.
+He was a founding member of the Bristol Robotics Laboratory (BRL), and acted as one of its two Deputy Directors from its inception in 2006 until 2020.
+His research interests focus on safe and correct operation whilst in close-proximity to human beings.
+He has co-authored over 200 international refereed publications, supervised 35 PhDs, and created a BRL income of >£10M since 2013.
+He has acted as BRL-UWE PI on the EPSRC grant ‘Trustworthy Robotic Assistants’, the Programme/Hub grants ‘Robotics for Nuclear Environments’ and ‘National Centre for Nuclear Robotics’, and the Centre for Doctoral Training in Robotics and Autonomous Systems.
+Prof.
+Pipe has been involved with Connected Autonomous Vehicle research and development since 2015 via the Innovate UK/CCAV funded projects VENTURER, FLOURISH, MultiCAV, CAPRI, ROBOPILOT and CAV-Forth.
 \end{IEEEbiography}

### 2021-04-12 document/source/abstract.tex
new document build structuring with common source

--- 2021-03-29 paper/cav_determinisim_main.tex
@@
-\documentclass[letterpaper, 10 pt, journal, twoside]{IEEEtran}
-% \input{archive/main.config.tex}
-\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
-% \addbibresource{IROS.bib}
-\addbibresource{cav_determinism.bib}
-\usepackage{array}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{color}
-\usepackage{threeparttable}
-\usepackage{balance} %balance columns
-\hyphenation{op-tical net-works semi-conduc-tor}
-\usepackage{enumitem}
-\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
-% \usepackage{subfig}
-\usepackage{subcaption}
-\usepackage{dblfloatfix}
-\usepackage{booktabs}
-\usepackage{authblk}
-\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
-% \pagenumbering{gobble} %suppresses page numbers
-\usepackage{balance} %balance columns
-%\usepackage{todonotes}
-\usepackage[disable]{todonotes}
-\usepackage{siunitx}
-% \usepackage{draftwatermark}
-\usepackage{flushend}
-
-%-------------------------------------------------------------------------------
-
-\begin{document}
-
-% ***************************************************
-%  TITLE
-% ***************************************************
-\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-% \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-% \author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
-% \author[1,3]{Greg Chance}
-% \author[1,3]{Kevin McAreavey}
-% \author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-% \author[2,3]{Tony Pipe}
-% \author[1,3]{Kerstin Eder}
-% \affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
-% \affil[2]{University of West of England, Bristol, UK}
-% \affil[3]{Bristol Robotics Laboratory, Bristol, UK}
-
-% \author{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kevin.mcareavey, kerstin.eder\}@bristol.ac.uk}}}
-% \author{Greg Chance}
-% \author{Kevin McAreavey}
-% \author{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
-% \author{Tony Pipe}
-
-\author{Greg Chance, Abanoub Ghobrial, Kevin McAreavey, S\'everin Lemaignan, Tony Pipe, Kerstin Eder 
-\thanks{{\footnotesize
-Manuscript  
-received ...;
-revised ...;  
-accepted.... 
-Date of publication ...;
-date of current version ....
-
-\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper and are both corresponding authors.}
-%
-
-This research has in part been funded by the ROBOPILOT and CAPRI projects.
-Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-% The Associate Editor for this paper was ....
-
-Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
-Abanoub Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
-Kevin McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
-and 
-Kerstin Eder (e-mail: kerstin.eder@bristol.ac.uk) 
-are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom.
-
-S\'everin Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
-and
-Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.
-
-Digital Object Identifier ....
-}}}
-% \affil{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
-% \affil{University of West of England, Bristol, UK}
-% \affil{Bristol Robotics Laboratory, Bristol, UK}
-%
-\markboth{IEEE TRANSACTIONS ON INTELLIGENT TRANSPORTATION SYSTEMS VOL.
-..., NO.
-..., date...}{Chance, Ghobrial \MakeLowercase{\textit{et al.}}: On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-%
-\maketitle
-%
-\begin{abstract}
-\noindent 
 Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
-%
 A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history.
-%
 Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug.
-%
 However, game engines are seldom deterministic.
-%
 This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines.
-%
 A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
-%
 Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
-%
 Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.
-
-% {\color{red} \sc Ensure past/present is used consistently and appropriately throughout.}
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
-% %
-% A key requirement for simulation-based development and verification is determinism.
-% %
-% Whereby, a deterministic process will always produce the same output given the same initial conditions and event history.
-% %
-% Thus, making tests repeatable and yielding simulation results that are trustworthy and straightforward to debug.
-% %
-% However, game engines are seldom deterministic.
-
-% This paper first reviews and identifies the potential causes of non-deterministic behaviours in game engines.
-% %
-% This is then followed by a case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, which highlights its inherent shortcomings in providing sufficient precision in experimental results.
-% %
-% Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is satisfactory low i.e.
-variance between repeats becomes negligible for development and testing work.
-
-% Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engines simulations for any given system configuration.
-
-% =========================================
-% Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
-% %
-% A deterministic process has no randomness and will always produce the same output with the same initial conditions and event history. 
-% %
-% Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.
-% \todo[inline,color=green!40]{GC Alternative: Determinism is a key requirement for simulation-based development and verification; it ensures the simulation is precise and that tests are repeatable, so that software bugs can be found and fixed, and simulation results are trustworthy.}
-% %
-% % KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
-% % These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
-% %
-% However, game engines do not need to be deterministic per se. % when used for games. 
-% %
-% This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation precision} for specific configurations and utilisation of the software and hardware, respectively.
-% %
-% \todo[inline,color=red!40]{old: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of a deterministic operational simulation setup.}
-% %
-% A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine has highlighted the shortcomings of using a games engine for simulation based verification.
-The proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-% %
-% % \textbf{ALTERNATIVE: This empirical investigation using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, has highlighted the shortcomings of using a games engine for simulation based verification and we present advice for best working practice.}
-
-\todo[inline,color=green!40]{GC Alternative: A case study using CARLA, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline,color=yellow!40]{AG Alternative: A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine (game engine), illustrates the proposed method and results from an example scenario demonstrate simulation precision to an acceptable tolerance.
-}
-\todo[inline]{Review again at end. Potentially give away more on results to motivate the reader to continue.}
-
-% The industry and certification bodies for connected autonomous vehicles is adopting the use of physics and gaming engines to develop, train, verify, validate and certify the software of these autonomous systems in simulation.
-It is important for these engines to be deterministic in order for one to carry out these different processes.
-However, such engines are inherently non-deterministic.
-We propose a method by which one can identify the region where they can operate in these engines and guarantee that the level of non determinism is sufficiently low that it can be assumed to be deterministic.
-We also implement this method on a case study and show how we identified the region (encompassing computational utilisation and behaviours in simulation) at which one can guarantee performance to be sufficiently deterministic.
-\end{abstract}
-
-\begin{IEEEkeywords}
-Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
-\end{IEEEkeywords}
-\IEEEpeerreviewmaketitle
-
-% ***************************************************
-%  INTRODUCTION
-% ***************************************************
-\section{Introduction} \label{s:introduction}
-% What is verification
-% Can't use this again as this is how the AI Test paper starts. -> Avoid self-plagiarism!
-%\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. 
-%
-% Simulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, benefiting from complete control over (virtual) actors and their environment
-% , offering a relatively inexpensive means to explore test cases while also
-% \IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. 
-\IEEEPARstart{S}{}imulation-based verification of autonomous driving functionality is a promising counterpart to costly on-road testing, that benefits from complete control over (virtual) actors and their environment.
-%
-Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of safe and legal driving rules, termed assertions, that must not be violated.
-
-Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules, e.g.\ through assertion checking during simulation. 
-There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-Simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
-In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
-
-% The use of game engines for AV testing
-Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
-%
-CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
-
-% Requirements of simulation based verification SBV
-State-of-the-art game engines provide a convenient option for simulation-based testing.
-They offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
-%
-Furthermore, they are easy to set up and run compared to on-road testing and are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}.
-%
-%\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
-% 
-% GC Removing the following para:
-%For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models.
-CARLA currently offers various basic types of cameras and %ray-case based
-% lidar; extensions are still under development. 
-% \todo{GC: what is ray-case? KIE: Not sure. :(}
-%
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. %~\cite{Javaid2013,cyres}
-%
-% The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}
-%
-% Game engines for SBV
-Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
-%
-However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
-
-% \subsection{Contribution}
-% In pursuit of an answer to this question we
-In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
-%
-In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
-% KIE: This is covered in the next sentence:
-% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
-%
-\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified.
-}
-%
-% By analysing actor position variance 
-We find that the CARLA simulator is non-deterministic.
-Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
-We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
-% {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
-% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
-%
-CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
-%
-\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
-However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}
-
-The insights gained from this case study motivated the development of a general step-by-step method for AV developers and verification engineers to determine the simulation variance for a given simulation environment.
-%
-\todo[inline,color=red!40]{old: Knowing the simulation variance will help them assess the impact that using a game engine for AV simulation may have.
-In particular, they can gain a better understanding of how potentially existing non-determinism affects repeatability and reliability of the simulation results.
-}
-%
-Knowing the simulation variance will help assess the suitability of a game engine for AV simulation.
-%impact that using a game engine for AV simulation may have.
-% on verification tasks. 
-In particular, this can give a better understanding of the effects of non-determinism and to what extent simulation precision may impact verification results.
-%
-\todo[inline,color=green!40]{GC Alternative: Knowing the simulation variance will help verification engineers assess the impact that using a game engine for AV simulation may have.
-In particular, this can give a better understanding of how potentially existing non-determinism affects simulation precision and any verification results that are interpreted from this.
-}
-%
-% \todo[inline]{review and potentially summarize results from case study GC - added see below.}
-%We present a case study to illustrate the use of this method.
-The study highlights the effects of non-determinism of a game engine on simulation results.
-Scenarios between pedestrian and vehicle actors are used to identify conditions that result in high simulation variance.
-%
-%By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational domain could be identified.
-%
-
-This paper is structured as follows.
-%
-Section~\ref{s:prelim} defines terms used throughout the paper and identifies when determinism is needed. 
-Section~\ref{s:background} briefly introduces how game engines work before investigating in Section~\ref{s:nondeterminisimSources} the potential sources of non-determinism in game engines.
-%
-Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
-%
-Section~\ref{s:methodology} presents the step-by-step method to assess the suitability of a simulation system for AV verification in general. 
-%
-We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo{review at end}
-
-\todo[inline,color=green!40]{GC New below}
-
-\section{Preliminaries} \label{s:prelim}
-
-\subsection{Definitions}
-
-Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
-\\
-
-\subsubsection{Determinism}
-% What is a deterministic system
-Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
-However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
-Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
-% 
-% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions.
-Specifically, we focus on the variance of actor paths within a given scenario.
-
-A simulation can be thought of as the process of generating or producing experimental data. 
-%
-In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data. 
-%
-If a simulation is deterministic, Fig.~\ref{variance_description}~(b), then there will be no variation in the generated output data, i.e.
-all future states are perfectly reproducible from prior states and actions.
-%
-However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data. \\
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
-        \caption{Non-deterministic}
-        \label{variance_description_a}
-    \end{subfigure}
-
-    \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
-        \caption{Deterministic}
-        \label{variance_description_b}
-    \end{subfigure}
-    \caption{Demonstration of variance, precision, tolerance and determinism}
-    \label{variance_description}
-\end{figure}
-
-% \subsection{Definitions: Precision, Variance \& Tolerance}
-\subsubsection{Variance, Precision \& Tolerance}
-
-We adopt terminology from the mechanical engineering and statistics domains to describe when there is variation in the generated output data~\cite{ADictionaryofMechanicalEngineering}.
-%
-% \textit{Precision} is therefore defined as the degree to which a simulation can repeatedly generate the same result when executed with the same conditions and actions taken. 
-% %
-% \textit{Variance} is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-%
-% \textit{Variance} (or \textit{precision}) is used here to define the spread (distribution) of the generated output data with respect to the mean value. 
-\textit{Variance} is used here to define the spread or distribution, of the generated output data with respect to the mean value.
-\textit{Precision} is synonymous with \textit{variance} although inversely related mathematically.
-%
-Therefore, variance can indicate the degree to which a simulation can repeatedly generate the same result when executed under the same conditions and actions.
-%
-\textit{Tolerance} is defined as the permissible limit of the variance, or in short the \textit{permissible variance}.
-
-% The engineering terms \textit{precision} and \textit{tolerance} suitably describe how the simulation data vary when the same repeated stimulus is provided to the simulation environment under identical initial conditions~\cite{ADictionaryofMechanicalEngineering}.
-%
-%
-% As an analogy, the simulator can be thought of as a manufacturing process, producing data and if tasked to repeatedly execute the same process we must ask how \textit{precise} can this process be and what, if any, will the \textit{variance} be?
-% when comparing the differences between multiple products.
-As an analogy, the simulator can be thought of as a manufacturing process that produces data.
-To determine the precision of this process, the output must be measured and analysed for differences when the process is repeated.
-%
-Those differences describe the spread or variance in the process output.
-A hard limit on the variance can then be defined, Fig.~\ref{variance_description}~(a), beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
-%
-% In relation to manufacturing, the precision of any computer based task is casually assumed to be absolute, for example one may question the manufactured precision of a physical component but not the computer assisted drawing that it was made from.
-%but technically this is equivalent to the representation of numbers within it, e.g. 64-bit floating point. %\textbf{[ref]}. %
-%
-Real manufacturing fails to achieve absolute precision. Hence, there is a need for tolerances to be specified to account for the variance in real-world manufacturing processes.
-
-If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within tolerance by design.
-If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
-%
-
-\subsubsection{Accuracy}
-Precision and tolerance should not be confused with \textit{accuracy}, which describes how closely the mean of the generated output of a process aligns to a known standard or reference value.
-Therefore, we define accuracy as the difference between the true value, or reference value, and what has been achieved in the data generation process or simulation.
-%
-For a driving simulation, the reference value 
-% , for which accuracy can be measured against, 
-may be the real world that the simulation seeks to emulate, where any divergence from this standard is termed the \textit{reality gap}. 
-%
-In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
-In most cases it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
-\\
-
-\subsubsection{Simulation Trace}
-A simulation trace is the output log from the simulator consisting of a time series of all actor positions ($x,y,z$) in a 3D environment recorded at regular time intervals.
-This definition could be extended to include other variables.
-%, but position is sufficient for this analysis.\\
-A set of simulation traces derived from the same input and starting state then forms the experimental data on which variance is calculated for a given simulation run. \\
-
-\subsubsection{Simulation Variance \& Deviation}
-If the simulator is non-deterministic then how can the simulation variance be measured?
-This can be achieved by monitoring the values of any of the recorded output variables that should be consistent from run to run.
-% In this study the actor state was chosen, or more specifically the statistical variance of actor paths within a given scenario using a fixed set of actions. 
-For example, actor position variance is a distance-based metric that can be derived from simulation traces.
-The actor position over time, i.e.\ the actor path, is often used in assertion checking, e.g.\ to determine whether vehicles keep within lanes or whether minimum distances to other vehicles and road users are being maintained.
-% 
-Thus, in the case study presented in this paper, the term \textit{simulation variance}, measured in SI unit $m^2$, refers to a measure of actor path variance in the simulation, assuming fixed actions.
-% {\color{red} Unclear: Why wrt.\ time, that would then be the actor path, or not?}
-%
-%\textit{Deviation} (SI unit $m$) is the square root of variance (SI unit $m^2$), which is a more intuitive measure to comprehend when interpreting test results.
-%
-Case study results are presented using deviation (SI unit $m$), the square root of variance, rather than variance, as this is a more intuitive measure to comprehend when interpreting test results.
-%
-% more intuitive to comprehend when considering verification tests, e.g.
-vehicle close-passing.
-Moreover, the maximum observed value of deviation was chosen over a mean value to ensure sufficient sensitivity.\\
-\\
-
-\subsubsection{Scene, Scenario \& Situation}
-We adopt the terminology defined for automated driving in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements.
-Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV, or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals.
-The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.
-A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
-
-% \todo[inline]{values - Does this mean stimulus or other parameters? GC - changed to parameters}
-
-% Verification and Determinism 
-% Why is determinism important for verification
-%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
-% KE Text
-\todo[inline,color=red!40]{old: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
-
-\subsection{When is Determinism needed?}
-Determinism is a key requirement for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
-%
-Thus, a deterministic simulator has zero \textit{variance}.
-
-A simulator with non-zero variance can no longer be considered deterministic. 
-%
-Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}. 
-%
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-%
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-
-% GC Text
-\todo[inline,color=green!40]{GC Alternative: Determinism is a key prerequisite for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable.
-In conventional engineering terms a deterministic simulator can be considered to have zero \textit{tolerance}.
-If the tolerance is non-zero it can no longer be considered deterministic but this may be sufficient for certain applications as long as it is \textit{within tolerance}.
-Therefore, \textit{tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.}
-
-% This is not to be confused with \textit{accuracy} that describes the extent to which the simulation complies to a known standard, which could be considered a level of realism that faithfully describes the real world.
-The \textit{precision} of the process is the ability of the simulation to repeat the same execution and find no, or a \textit{tolerable}, difference in the outcome.
-\textit{Tolerance} is the acceptable degree of variability between repeated execution outputs.
-When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
-This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
-}
-%
-%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution.
-Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour.
-%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
-%
-
-If the simulation is non-deterministic, e.g.\ 
-% KIE No, whether the variance is or is not permissible is not what matters for non-det, it should be zero!
-%e.g. it has a non-permissible variance in,
-it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
-In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
-%
-
-% \todo[inline]{GC Edit: If the simulation is non-deterministic, it lacks the appropriate precision for this application and the error between output traces exceeds the acceptable tolerance however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs.
-In the worst case, bugs that could have been identified in simulation remain undetected, resulting in false confidence in the safety of the AV's control software.}
-
-% Challenges of SBV
-When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs % related to non-determinism
-a high priority for games developers.
-It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
-However, the situation is very different for AV development and testing.
-Thus, our main research questions are:
-%
-\todo[inline,color=red!40]{old: How can one assess the extent to which a simulation environment is deterministic?  }
-%
-{\em How can one assess whether a simulation environment is deterministic?} and 
-{\em How can one determine and control the simulation variance?}
-\todo[inline,color=green!40]{GC Alternative: How can one assess the extent to which a simulation environment is deterministic or has a permissible tolerance?}
-
-% \noindent \IEEEPARstart{T}{}he evolution and inevitable implementation of connected and autonomous vehicles (AVs) on roads is a delicate if not utterly fascinating subject.
-Such systems operate in environments marked by inaccessibility, unexpected weather conditions, or unpredictable behavior by surrounding humans\cite{RobustnessAutonomy}.
-
-% Public safety is clearly a prime concern for putting autonomous vehicles into play, which means the meticulous and thorough process of developing, training, verifying, validating and certifying the artificial intelligence (AI) and software of the vehicle is a task to be handled with care.
-
-% It requires lots of flexibility and room to iterate to ensure the algorithms are given the sharpest understanding for maximum performance as well as to ensure that they are hazard-free in order for them to be deployed\cite{AirsimUnrealArticle}.\\\\
-
-% \noindent There has been a number of catastrophic fatalities with self-driving vehicles due to lack of verification and validation (V\&V) of software e.g.\cite{FatalityExample}.
-
-% As such, there appears the need for finding ways to V\&V the safety of such software, ranging from AI development to cyber-security, while maintaining minimal risks in testing.\\\\
-
-% ======================
-% This is all from "FrameworkAndChallenges" without using quotes 
-% \noindent Existing vehicle manufacturers use vehicle dynamic simulators to model the physics of each component in a system.
-Similarly, the use of simulation for traffic modelling is well established.
-Despite these proven simulation tools, non of them provide all of the required functionalities together to fully simulate AVs\cite{FrameworkAndChallenges}.\\\\
-
-% While traffic-level simulators can model road layouts, they do not require as much realism or detail as AVs require.
-On the other side, vehicle dynamics simulators include some of the high fidelity physics functionalities needed for AV testing, but they only consider the vehicle itself and perhaps the road surface and gradient; they do not model any of the road layout, signs, traffic/street lights, other street furniture or surrounding infrastructure.
-% Most critically, neither types of simulators model sensors such as cameras, lidars or radars\cite{FrameworkAndChallenges}.\\\\ 
-% Consequently, one of the major routes AVs developers are moving towards is using physics engines, these are often games engine development frameworks, since they tick all of the boxes needed to fully simulate AVs \cite{FrameworkAndChallenges}.
-% ======================
-
-% Robopilot, Capri, CARLA, Apollo, Airsim, Udacity etc.
-are all examples of driving simulator projects that use physics engines to support development, training, and validation of autonomous driving systems software\cite{ListOfSimulators}.
-\\\\
-% Test-retest reliability, is a key factor in V\&V of software. Tests need to be repeatable in order for one to eliminate bugs. 
-% In other words physics engines has to be deterministic, or the level of non-determinism of the engine should be low and clearly defined, for them to be used in the context of V\&V of AV software.\\\\
-% \TODO{Remove this paragraph?} Determinism of game (physics) engines has not attracted much attention by engines' developers nor researchers.
-This previously was not important since in gaming, determinism is not critical.
-% Performance, on the other hand, is a subject of more interest in that field, where developers want games to work on all platforms and still be fast to provide a positive user experience. 
-% On the contrary, determinism becomes vital and performance not when such engines are used in autonomous vehicles' AI development and testing. \\\\
-
-% This paper aims to investigate how deterministic physics engines are for usage in V\&V for AV simulations; and will cover the list of following points:
-% \begin{itemize}[leftmargin=*]
-%     \item Necessary background on game engines used in AV simulations.
-%     \item Review of sources of non-determinism in such engines.
-%     \item Provide a method of how the level of non-determinism can be investigated and defined.
-%     \item A case study showing the implementation of this methodology in an engine widely used in AVs simulations (Unreal Engine).
-% \end{itemize}
-
-% ***************************************************
-%  BACKGROUND
-% ***************************************************
-\section{Background} \label{s:background}
-
-There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
-% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
-
-Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.
-%However, when these engines are used for AV development then deterministic behaviour is required.
-%
-Ensuring software operates deterministically is a non-trivial task.
-Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-%
-\todo[inline]{GC line added to introduction.
-We need a paragraph on how hard it can be to make a computation fully deterministic, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-%
-\todo[inline,color=green!40]{GC Addition: Ensuring software operates deterministically is a non-trivial task and catching intermittent failures in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-}
-%
-This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
-%
-% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in AV simulations.
-Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.
-
-% ======= Overview of a Game Loop
-% \subsection{Game Engine Overview} \label{GameLoopSection}
-
-Central to a game engine are the main game logic, the artificial intelligence (AI) component, the audio engine, and the physics and rendering engines. For AV simulation, we focus on the latter two.
-%
-The game loop is responsible for the interaction between the physics and rendering engines.
-Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed~\cite{unity_event_execution}.
-% \footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}. 
-A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering)~\cite{GameEngineArchBook}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoopv2.pdf}
-\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}.}
-\label{GameEngineLoopDiagram}
-\end{figure}
-
-% \todo[inline]{AG: remove the clock symbol, possibly remove the ticks chart as this is confusing given they are not ticks but event markers}
-
-% The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment.
-For example, the throttle and steering angle from the controller of an AV.
-The update stage interprets the input from all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states.
-The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.
-
-% \todo[inline]{Fig 1: update rates of which clock?
-GC-is a clock confusing as it does not run on a 'clock' in computer science terms, rather it is a timer which is used as reference to compare to a 'wall clock'}
-
-% \todo[inline]{Should there not also be a clock in the overall loop? GC-yes the wall clock but it may get confusing as that is a real clock and the game loop clock is a timer?}
-
-% \todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.
-Unity page on multi-threading \footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and a plugin for multi-thread management \footnote{\url{https://forum.unity.com/threads/task-for-unity-scripting-threading.462971/?_ga=2.228561631.331624405.1587372521-1435703724.1587372521}}}
-
-% 
-% It is necessary to have an understanding of how the game loop operates in order to understand potential sources of non-determinism that are discussed below.
-
-% \todo[inline]{We should agree if the 'game clock' really is a clock or just times the computation for each game loop.}
-
-% \todo[inline]{For AG: We use the phase 'game clock tick'- where did the come from as I think it is misleading.}
-
-% As the game progresses, the game loop runs continuously according to the game clock. \todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
-%
-% At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed.
-This measures how far the game's clock is behind compared to the real world.
-
-The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
-This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}.
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}.
-% \footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}. 
-When the frame is rendered the game loop cycle returns to processing inputs.
-An intuitive and  more detailed description of the interplay between the physics and render cycles is given in~\cite{JohnAustinUnity}.
-
-% \todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors.
-Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate.
-The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}.
-When the frame is rendered the game loop cycle returns to processing inputs.}
-%
-% Then the user inputs are processed.
-There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world.
-Rendering occurs once the game clock has caught up with the real world clock.
-The process then repeats.
-
-The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
-The shorter this time step is, the smoother the interpretation of the physical dynamics will be.
-%processing time it takes to catch up to game clock.
-% and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}.
-%
-% Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
-%
-To use a fixed physics time step, the user's display refresh rate needs to be known in advance.
-This requires an update loop to take less than one render tick (one frame of real world time).
-Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$.
-However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}.
-%
-Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware.
-Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}.
-%
-If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics.
-%
-Thus, extrapolation between frames may need to be performed to smooth transition between scenes. 
-%
-Note that both residual lag and extrapolation could affect perception stack testing.
-% This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
-% \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
-%
-% \todo[inline]{Why is this part of the figure?
-It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.
-GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
-%
-% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
-In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines,
-and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.
-% \todo[inline]{Explain how this lag may affect perception stack testing}
-
-% \todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?
-GC-suggest remove this part of the diagram, too confusing as there are not really any clocks or ticks - we should remove all mention of ticks too.}
-
-% The time step, $dt$, in the physics calculation is important and the smaller the value the more realistic the simulation will be.
-If $dt$ is too large or irregularly spaced then the rendered motion may not be smooth and interpolation may be required at additional computational cost.
-
-% This process allows hardware scalability, but the rendering will become of jerky quality on slower machines.
-These engines update at fixed intervals and render whenever they can, which is not steady.
-This results in what is so called residual lag\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, where the engine is trying to render between two consecutive updates.
-In this case, the engine will use extrapolation techniques to give an estimate of where the object should be.
-This often is sufficient for gaming purposes and unnoticeable to the user, in fact it improves the stuttery motion.
-%
-\todo[inline,color=red!40]{old: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy.
-In contrast, simulation accuracy is essential in AV development and testing, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of accuracy at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is as close as possible to what the real input would be, i.e.\ with no or minimal extrapolation or lag.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-}
-%
-Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-% 
-The gamer neither needs a faithful representation of reality (i.e.
-gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
-gamer accepts low precision).
-In contrast, high accuracy and precision are necessary for AV development, testing and verification.
-% The gamer neither needs a faithful representation of reality, i.e.\ the gamer will accept low \textit{accuracy}, nor do they require for repeated actions to result in the same outcome to within a particular tolerance.
-%, i.e.\ can be low \textit{precision}.
-% %
-% In contrast, the precision required for AV development and testing is very high, especially for perception stack testing, but also, more generally, for obtaining reliable verification results and for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required tolerance %on the precision
-% at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that is repeatable so that if any software bugs are found they can be re-played and the issue resolved.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.
-%
-\todo[inline,color=green!40]{GC Edit: Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
-Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
-The gamer neither wishes a faithful representation of reality, i.e.
-can be low accuracy, nor do they require for repeated actions to result in the same outcome to a particular tolerance, i.e.
-can be low precision.
-In contrast, the precision required for AV development and testing is much higher, especially for perception stack testing, but also for visual inspection of accidents or near misses.
-If necessary, it is acceptable to achieve the required level of tolerance on the precision at the cost of real-time performance.
-For example, for perception stack testing the sensors need to get input that repeatable so that if any software bugs are found they can be re-played and the issue resolved.
-This  may only be realisable by slowing down the rendering to enable more extensive physics calculations.}
-%
-\todo[inline]{KIE: ... by slowing down the simulation to enable more extensive physics calculations?}
-
-%\todo[inline]{THIS IS IMPORTANT: Spell out the conflicting objectives (i.e.\ the fundamental differences) of gaming vs AV V\&V in terms of providing a real-time user experience at the cost of realism (for gaming) vs the need for realism potentially at the cost of losing real-time performance.
-This is particularly important for perception-stack testing (the sensors need to get input that is as close to the real input as possible, ideally without extrapolation or lag) and visual inspection of accidents or near misses (same is true for a human observer).
-For instance, see below, but I'm not sure whether this is the right place.}
-
-% ======= Sources of non-determinism
-\section{Potential Sources of Non-Determinism} \label{s:nondeterminisimSources}
-
-% The following discussion reflects the results of our investigation into the potential sources of non-determinism that could affect game engines.
-We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
-The following review discusses the potential sources of non-determinism that were found in the literature or found as part of our investigation into game engines.
-We have examined hardware- as well as software-borne sources of non-determinism that occur at different layers of abstraction.
-%
-\todo[inline,color=green!40]{GC Addition below}%
-%
-A good analysis of potential sources is given by Strandberg et al.~\cite{intermittently-failing-tests}, although the AV simulation domain introduces it's own unique challenges that were not considered in that paper.
-
-% MOVED TO INTRO \todo[inline]{We need a paragraph on how hard it can be to make a computation fully determinisitc, referring to the ACM queue interview with the {\tt rr} developers~\cite{acm-q-rr-interview}.
-Not sure whether this should be in the introduction, HERE, or in a related work section or in the discussion/conclusion.
-But it is important that we demonstrate knowledge of others working on this in other areas.
-{\tt rr} is somewhat closely related as they need determinism for debugging, and we need it for that purpose too, although we also need verification results to be stable.}
-
-\todo[inline]{We should also refer to literature like:\\
-Making System User Interactive Tests Repeatable: When and What Should We Control?~\cite{when-and-what-should-we-control} and
-Intermittently Failing Tests in the Embedded Systems Domain~\cite{intermittently-failing-tests}.}
-
-\medskip
-
-% \noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic.
-The main reasons that we think are causes of these engines' non-determinism are discussed below.
-\\\\
-
-% ======= Floating-Point Numbers
-\subsection{Floating-Point Arithmetic}
-
-Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
-Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value.
-This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
-%
-
-% SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
-%{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
-%and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }
-
-% As a consequence, floating-point arithmetic is not associative.
-Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, a change of execution order, e.g.\ from $x+(y+z)$ to $(x+y)+z$, can lead to different results~\cite{Kapre2007}.
-%
-As a consequence, floating-point arithmetic is not associative.
-Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}.
-This is because rounding is being performed at different stages in the computation, depending on the respective execution order.
-%
-Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
-%
-Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}. 
-%
-%{\color{red} Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.}
-%
-%If calculations are performed on a single core, then concurrency of multiple threads is not an issue.
-\todo[inline,color=green!40]{GC Addition (added reference on non-associative fp addition Kapre2007, Whitehead2011, empirical-analysis-of-flaky-tests, see para below:)
-
-% Even operations such as the average of an array can result in issues with overflows, underflows and non-associative addition~\cite{Kapre2007} which may result in non-deterministic behaviour.
-Calculation results may differ due to the nature of non-associative floating-point arithmetic if there is a change in compiler, a change in the execution order or even if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-Some authors even suggest avoiding the use of floating-point entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-%
-% If calculations are perofrmed on a single core, then concurrency of multiple threads is not an issue.
-% which is discussed in the section below titled Scheduling, Concurrency and Parallelisation.
-}
-%
-% Floating-point representation of real numbers is limited with fixed register sizes resulting in errors of rounding where loss of bits occur which are rounded or truncated and cancellation which happens where two similar values are subtracted.
-
-% Translating to the usage of AV simulations in physics engines,
-In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
-% \todo[inline]{state what assertions are (used for) GC-see below} 
-%In the field of AV verification, assertions are the immutable rules and regulations that describe safe and legal driving. 
-%
-Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability.
-In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.
-
-In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-%{\color{red} Should this be IMPRECISION rather than incorrectness?}
-%
-It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
-%
-Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern.
-%, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
-%
-So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
-
-However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
-Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
-
-Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
-However, the same executable should still return the same output for identical input.
-
-In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.
-
-\subsection{Scheduling, Concurrency and Parallelisation}
-% \todo[inline]{is this at runtime?} %
-% info at:
-% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
-% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
-% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
-% https://en.wikipedia.org/wiki/Concurrent_computing
-% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
-Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy.
-A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-% \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
-%
-In principle, changing the scheduling policy and thread priorities may increase simulation variance.
-
-It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
-Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
-
-However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
-This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
-%
-Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-%
-\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-}
-%
-Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
-
-Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes.
-This may be decided based on factors such as throughput, latency or CPU utilisation.
-%
-If due to the CPU utilisation policy the same single-threaded script executes multiple times across different physical cores of the same CPU type,  then execution should still produce the same output.
-%
-This is because the processor cores are identical and any impurities across the bulk silicon and minor perturbations in the semiconductor processing across the chip that may exist should have been accommodated for in design and manufacturing tolerances.
-%
-However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
-
-Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
-%
-\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{rr} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
-}
-
-Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously.
-While this would be faster than a serial execution, the order of execution arising from program-level concurrency is often not guaranteed.
-
-Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}. 
-%\noindent\underline{\textit{Concurrency:}}
-% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
-%\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism.
-Consider also parallelism, e.g.
-in GPUs.
-Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism.
-Say which parts of a game engine could reasonably be expected to run concurrently.}
-%
-%Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider.
-This technique can lead to issues such as shared resource access conflicts due to mis-managed timing and RACE conditions~\cite{huffman1954synthesis}.
-A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.
-% \todo[inline]{how? GC-see above.}
-% KIE: We only list the potential sources of non-det, I therefore removed this. 
-% although monitoring and detecting such events falls outside of the scope for this work. 
-%
-%\noindent\underline{\textit{Ego Vehicle Controller:}}
-%The ego vehicle controller may be a source of non-determinism if using complex control algorithms such as neural networks for visual processing and perception.
-A neural network can only be deterministic within in the bounds of the training data used to construct it.
-If the network is exposed to unseen data it may incorrectly classify legitimate scenes due to lack of training data.\todo{That alone won't cause non-determinism, i.e.
-the NN will misclassify exactly in the same way each time - not?}
-%
-%In the simulation domain, if the environment, and hence the data the network is exposed to, is controlled then the perception stack will behave deterministically, assuming no \textit{residual lag}.
-In this sense, control over weather and lighting in the game engine should be considered and controlled as well as any other dynamic elements to the environment, e.g.
-reflections from surface water.
-% these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
-% i.e.
-the utilisation of hardware is manipulated differently each time.
-This would then result in non-deterministic behaviours especially if it is combined with the floating-point problem discussed earlier.
-%
-% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly.
-Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available.
-This may lead to events happening out of order depending on the computational load and hence non-deterministic.
-%
-% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved.
-To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order.
-This is beyond the scope of our work and will not be covered in this paper.
-% \\\\
-% \noindent\underline{\textit{Physics and Rendering clocks:}}
-% \todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
-% % The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
-% Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}.
-This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step.
-% %
-% A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this?
-Do we not assume the same HW at this stage?
-GC- we are just listing sources at this point so should be hardware agnostic.
-also, yes, hardware is the same for us but we cannot control for resource utilisation and background processes.} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur.
-Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources.
-% %
-% AV specific platforms, such as CARLA, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
-% \todo[inline]{Add a summary/conclusion?}
-%\\\\
-
-\subsection{Non-Uniform Memory Access (NUMA)} \label{s:sources_numa}
-For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor.
-Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}.
-%
-Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower.
-%
-By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.
-
-\subsection{Error Correcting Code (ECC) Memory}
-ECC Memory is used ubiquitously in commercial simulation facilities and servers to detect and correct single bit errors in DRAM memory~\cite{Dell1997}.
-Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}.
-If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring.
-Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.
-
-Any simulation hardware not using ECC memory that runs for 1000's of hours, typical in AV verification, is likely to incur significant CPU hours and is therefore subject to increased exposure to these errors.
-To counter this, commercial HPC and simulation facilities typically employ ECC memory as standard.\todo{Check this, please.}
-
-%\todo[inline]{That said, HW design simulation is deterministic, uses long simulation runs and thousands of CPU hours, but I've not heard that this is an issue.
-I can ask, though.
-;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.}
-
-\subsection{Game Engine Setup}
-The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps, ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing.
-%
-%
-% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
-% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
-% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
-For example, in Unreal Editor the \textit{unit}~\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
-% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
-the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
-%
-With respect to perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water, ensuring textures are not randomly generated.
-
-\subsection{Actor Navigation}
-There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-% 
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
-%
-The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
-
-A navigation mesh is a fixed area of the environment where actors are free to navigate.
-%
-Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
-%.
-Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
-%
-However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment.
-These cannot be unknowingly modified and therefore can be considered fixed.
-
-Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
-%
-% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
-%
-% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
-% \todo[inline]{unclear :( GC- re-written}
-
-\subsection{Summary}
-We have investigated the potential sources of non-determinism affecting game engines 
-and explored the impact they may have on simulation variance. 
-%
-% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable.
-}
-%
-Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
-\todo[inline]{So, do commercial HPC facilities have ECC memory?}
-\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
-%
-\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
-To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
-\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
-%
-Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.
-Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC.
-If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.
-
-% {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
-%
-Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
-%
-However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop.
-Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism.
-%
-
-%
-%The ego vehicle controller can be considered deterministic if dynamic elements to the simulation environment are controlled, such as weather and lighting that may be on random cycles. 
-%
-% ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software.
-Actor navigation not a problem if navigation meshes are fixed.
-Game engine, ensuring fixed $dt$ and control of random numbers.
-Core or thread scheduling will not be issue unless changing scheduling policy or migrating between machines that have different policies.
-Concurrency is really down to the development of the simulation software and reputable companies should have good control over these potential issues following good software development practice.
-NUMA should only affect interprocessor data transfer cost, at wrt only make computations longer.
-relative times between access different caches is likely to be small although may impact on high throughput, e.g.
-HPC.
-VFR - want to remove this section.
-EGO vehicle deterministic?
-Can't control library change, perception stack
-
-% ***************************************************
-%  Empirical Investigation
-% ***************************************************
-\section{Case Study of Simulation Variance} \label{s:case-study}
-
-We present an empirical investigation into using game engines for
-simulation-based verification of autonomous vehicles with a focus on
-characterising sources of non-determinism in order to understand  the impact
-they have on simulation variance. 
-%
-\todo[inline,color=green!40]{GC new: line below.}
-%
-Gao et al.~\cite{when-and-what-should-we-control} took a similar approach investigating Java applications, where a set of sources of non-determinism (termed factors) were shown to impact on repeatability of testing.
-%and recommended that results of repeated tests must be quoted in averages and ranges as their testing configuration was not possible to execute deterministically.
-%
-Ultimately, our objective is to control non-determinism to minimise simulation variance.
-
-\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
-%
-We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
-A discussion on the internal and external settings of the simulation is included, along with system configuration and pre-screening sections.
-% {\color{red} and then..? Please cover the entire section - consider splitting setup from evaluation.}
-%
-\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
-%
-\todo[inline]{Complete the section structure as this is a very long section.}
-
-% \todo[inline]{First describe the scenario(s), WHAT happens, HOW does it happen and WHY is this scenario interesting for this investigation, i.e.\ what are the features this scenario has?
-Then say how this has been implemented, referring to test numbers and the GC - think ive done this, take a look at first para below.}
-
-% ***************************************************
-%  Scenario Description
-% ***************************************************
-\subsection{Context, Scene and Scenario}\label{TestsDescriptionAndTechnicalities}
-\todo[inline]{Need to stick to scene, scenario definitions as in Intro section.}
-
-\begin{figure}[!t]
-    \centering
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
-        \caption{}
-        \label{Test_a}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
-        \caption{}
-        \label{Test_b}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
-\end{figure}
-
-This case study draws on a setup used to verify an urban mobility and transport solution, where the primary verification objective is to test the behaviour of an ego vehicle in an urban environment at a T-junction in response to pedestrians and other vehicles.
-%
-Thus, the scene for our investigation is the T-junction with the scenarios as shown in Figure~\ref{f:test_a_and_b}.
-
-This scene was used to create a number of scenarios involving pedestrians and vehicles in order to identify any changes in the actor paths over repeated tests executed under a variety of systematically designed conditions and hence study any simulation variance.
-%
-The vehicles and pedestrians were given trajectories, via pre-defined waypoints, that would result in either colliding with or avoiding other actors.
-\todo[inline]{Decide past or present tense and be consistent throughout.}
-
-\todo[inline]{Set the threshold, explain what a good threshold is, available options and what was selected and why.}
-
-\subsection{Tolerable Simulation Variance}\label{s:tolerance}
-
-\todo[inline]{Terminology:\\
-{\bf Accuracy} is the degree of closeness/conformance to a known value.
-\\
-{\bf Precision} is the ability to get a particular value several times, i.e.\ repeatability of a measurement.
-\\
-{\bf Tolerance} is the allowable or predictable deviation from a known/desired value, e.g.\ $\pm 1cm$. 
-\\
-See {\small\url{https://www.starrapid.com/blog/designing-for-manufacturing-accuracy-precision-and-tolerance/}} for further details.
-}
-
-% \todo[inline,color=green!40]{GC addition: Ideally simulation results should be precise, where repeated tests give exactly the same result.}
-
-To achieve stable verification results over repeated test runs, the simulated actor states must be precise to a specific tolerance.
-%
-Deterministic behaviour would result in zero variance of the simulated actor states but if this cannot be achieved then what is permissible?
-%
-% \todo[inline,color=green!40]{GC addition: Deterministic behaviour would result in zero tolerance of the simulated actor states but if this cannot be achieved then what is a permissible tolerance? }
-%
-The tolerance must be appropriate to allow accurate assertion checking and coverage collection in the simulation environment, but not so small such that assertion checking would fail with minor computational perturbations.
-% {\color{red} {\sc What would fail?}}
-%
-Thus, a tolerance must be defined to reflect the precision at which repeatability of simulation execution is required.
-
-% tolerances are controlled to within a small, acceptable degree of variability from one part to another.
-%
-% but a larger value may be acceptable for non-urban environments, e.g.
-motorway or highway.
-Similarly, the threshold value should not be so small as to fail with minor computational bottlenecks, e.g.
-$1\mu m$.
-% 
-% To determine if the ego vehicle performs correctly, i.e.\ does not violate assertions, t
-
-% The autonomous vehicles operating in the last-mile mobility and delivery sector as considered in this case study will predominantly operate within urban rather than highway or motorway environments.
-%
-For this case study a tolerance on actor position of $1$m would be insufficient when considering the spacial resolution required to distinguish between a collision and a near-miss event. 
-%situations such as potentially dangerous close passing of a cyclist. 
-A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives. 
-% due to the simulation \textit{noise floor}\footnote{See Fig.~\ref{CarsCollisionCG25}.} 
-Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected. 
-% Thus any variation of less than $1$cm is tolerable. 
-Thus any variance of less than $1$cm is permissible. To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-
-% considered to be sufficient for assertion checking in urban environments. 
-%
-% A value larger than this, e.g. $1m$ will not be sufficient to spatially resolve these scenarios and a smaller value, e.g. $1mm$ may result in false positives.
-
-% \todo[inline]{Move to Case Study, rewrite with justification: Two autonomous vehicle projects this group are involved in, Robopilot and Capri, are looking at urban mobility and last mile delivery solutions which are predominantly concerned with urban environments rather than highway or motorway scenarios.
-Because of this, a deterministic threshold of $1m$ is insufficient when considering potentially dangerous close passing of cyclists and pedestrians and therefore a threshold of $1cm$ is more suitable.} \todo[inline]{Formal writing, no reference to the group or project, please.}
-
-\todo[inline,color=red!40]{old: For the simulation to be considered repeatable, the overall level of simulation variance for any actor should be less than a user-defined threshold, or \textit{permissible tolerance}, for a given scenario and verification objective.
-}
-
-% \todo[inline,color=green!40]{GC question: should we remove this line below as it is repeating above statement?}
-In the following, case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history where any value higher than the specified tolerance is considered non-permissible.
-%
-\todo[inline,color=green!40]{GC edit: Case study results are shown in terms of the maximum deviation, $\max\sigma$, from the mean actor path over the entire simulation history %and the deviation threshold in this investigation is $1cm$,
-where any value higher than the specified tolerance is considered non-permissible. }
-
-% ***************************************************
-%  Actor Collisions
-% ***************************************************
-
-\begin{table*}[b]
-\centering
-\begin{tabular}{clclccc}
-\toprule
-Test & Actors 					 	  & Collision 	 & Collision Type 			 & $n$  & $\max\sigma$ (m) & $\max\sigma$ (m) \\ 
-	 & 								  & 			 & 							 &	    & (unrestricted)  & (restricted)	\\ \midrule
-1    & Two vehicles                   & No  		 & N/A 			  			 & 1000 & 0.03 & $7.0{\times}10^{-3}$ \\
-2    & Two vehicles                   & Yes 		 & Vehicle and Vehicle 		 & 1000 & 0.31 & $9.8{\times}10^{-3}$ \\
-3    & Two vehicles and a pedestrian  & No  		 & N/A 			  			 & 1000 & 0.07 & $5.2{\times}10^{-4}$ \\
-4    & Two vehicles and a pedestrian  & Yes 		 & Vehicle and Pedestrian 	 & 1000 & 0.59 & $1.5{\times}10^{-12}$ \\
-5    & Two pedestrians                & No  		 & N/A 			  			 & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-6    & Two pedestrians                & Yes 		 & Pedestrian and Pedestrian & 1000 & $5.6{\times}10^{-13}$ & $5.6{\times}10^{-13}$ \\
-\bottomrule
-\end{tabular}
-\caption{A description of the test scenarios showing the test number, the actors included, whether a collision occurred and if so then between which actors.
-$n$, the number of repeats is set to 1000 and $\max\sigma$ is the maximum simulation deviation.
-The term \textit{unrestricted} refers to an unrestricted account of the results including results of any resource utilisation.
-To understand the impact of collisions and high resource utilisation, the \textit{restricted} column shows a subset of the results where post-collision data and experiments above 75\% resource utilisation have been removed.}
-\label{TableOfExperiments}
-\end{table*}
-
-\subsection{Actor Collisions}\label{S:Actor_Collisions}
-Previous investigations into the Unreal Engine indicated that collisions between actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}.
-%
-Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were identified as potentially key aspects to the investigation into simulation variance.
-
-The tests used for the case study are listed in Table~\ref{TableOfExperiments}. They cover a range of interactions between actor types. 
-%
-Tests 1 \& 2 involve two vehicles meeting at a junction where they either do not collide (Test 1) and where they do collide (Test 2), thereby triggering an \textit{event hit} callback in the game engine.
-In both cases the trajectories of the vehicles are hard-coded to follow a set of waypoints spaced at $0.1$m
-% \todo[inline]{AG: is this value correct or should it be 2.0m?
-AG-reply: The granularity of the trajectory waypoints is 0.1m, but the lookahead distance that the actors use is 2m ...
-actors would ignore trajectory waypoints that are less than 2.0m away from them.}
-intervals using a PID controller. 
-%
-In Test 3 a mixture of different actor types is introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point. 
-%
-In Test 4 this pedestrian collides with one of the vehicles at the crossing, triggering an \textit{event hit} callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
-Tests 5 \& 6 involve only pedestrians that either do not collide (Test 5) and that do collide (Test 6), see Fig.~\ref{Test_b}.
-
-\todo[inline]{Again, match the flow diagram steps in the subsections, but here wrt the case study steps.}
-
-% ***************************************************
-%  Experiment Description
-% ***************************************************
-\subsection{Evaluation Metric}\label{s:Experiment_Description}
-% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
-For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time. 
-%
-The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.
-
-Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
-%
-\todo[inline,color=green!40]{GC Addition: For each test the position of each actor was logged at $0.1s$ intervals providing a trace of that actor's trajectory with respect to simulation time. 
-%
-Repeated traces are compared at the same time index $t$ for each actor $a$ to provide a value for the variance, $\sigma_a^{2}(t)$. 
-%
-Herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor position data relative to the mean and is helpfully in the same units as actor position, $m$, for ease of interpretation.
-}
-
-The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
-%{\color{red} This is still a function over time - should it be?}
-\begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a^{2}(t)
-\end{equation}
-The maximum deviation is the absolute value of the square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity. 
-% {\color{red} So, ``a'' is any actor or a specific actor?
-Likewise, ``t'' is any time point or a specific time point?
-NOTE that any actor any time means a in A and t in T, where A is the set of actors and T is the set of all time points.}
-
-The maximum deviation, $\max\sigma$, can be analysed for the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the limit of \textit{permissible variance} to indicate if the simulation is sufficiently accurate for verification purposes.
-% The output for each experiment can therefore be summarised as the \textit{maximum variance} of any actor over the $n$ repeated tests. 
-% The maximum, rather than average, variance is chosen to indicate the worst case where an average could obscure outlier data points when a high number of repeated tests are performed. 
-\todo[inline,color=green!40]{GC Addition: The maximum deviation, $\max\sigma$, was analysed against the different scenarios and settings that were identified as potential sources of non-determinism, and compared against the threshold of \textit{permissible tolerance} to indicate if the simulation was reliable for verification purposes.
-}
-%
-% Preliminary work\footnote{see footnote 14} suggested that adding stress to the CPU or GPU will contribute towards increased . 
-% \todo[inline]{Introduce the need to use CPU and GPU HW in a game engine or any simulation HW. GC done}
-
-% Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs.
-Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.
-
-% This whole process is repeated for various internal and external settings (if needed). 
-% Then various plots can be created to draw the engine's simulation variance. 
-% Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
-
-% ***************************************************
-%  Internal Settings
-% ***************************************************
-\subsection{Simulator Settings}
-
-Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation.
-Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.\ \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.\ \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}.
-Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
-% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
-An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings.
-The results were compared to a baseline of the default engine settings.
-%
-However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
-Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.
-
-% \subsection{Initial Results and Discussion}\label{InitialResultsSection}
-
-% {\color{red} This is for unrestricted, so this does not come as a surprise later.}
-
-% ***************************************************
-%  External Settings
-% ***************************************************
-\subsection{External Settings}
-% \todo[inline]{This would be ``External Settings'' and can be introduced after finding that internal settings alone did not tame the non-determinism. :(}
-
-% After discovering that internal engine settings could not achieve a simulation variance that met the required 
-% \textit{deterministic threshold}, 
-% \textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
-After examining controls the internal game engine, external settings were explored. 
-%
-Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
-%
-For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}.
-
-Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load.
-This initial observation was explored more fully in this work using the CARLA platform.
-
-% {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented.
-Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}
-
-To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation.
-%
-% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
-Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively.
-Resource utilisation figures reported here should be considered approximate values.
-%
-Resource utilisation was capped to 75\% in some parts of the results and referred to as \textit{restricted}.
-This was done to limit the load on the test system and hence limiting simulation variance.
-The term \textit{unrestricted} places no such limit on the system load.
-
-% {\color{red} To understand the impact of collisions and high resource utilisation, the restricted column shows a subset of the results where post-collision data and experiments above 75 resource utilisation have been removed}
-%
-% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).
-%
-% {\color{red} Somewhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}
-
-\todo[inline,color=green!40]{GC Addition (para below):} 
-%Issues around floating-point precision and concurrency as previously mentioned notwithstanding, 
-Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results.
-For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}.
-In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency-induced imprecision~\cite{matlab_parallel_computing}.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV6.pdf}
-    \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
-Tests 5 and 6 overlap having almost identical results.
-Note that the lines between data points are only a guide.}
-    \label{ExperimentsStressSummary}
-\end{figure}
-
-% ***************************************************
-%  System Configuration and Screening
-% ***************************************************
-\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
-The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
-%
-The operating systems was Linux Ubuntu 18.04.4 LTS. 
-%
-Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s.
-
-% Each test was repeated 1000 times. 
-Initial testing~\cite{TSLUnrealEngineTesting} indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over {\raise.17ex\hbox{$\scriptstyle\sim$}}$10$cm.
-While executing 100 repeats may seem sufficient, this sample size may fail to observe events that occur with low probability, giving false confidence in the results.
-Therefore each test was repeated 1000 times to provide a sufficient sample size.
-% {\color{red} Justify why 1000 - see my later comment in the Method section.}
-%
-A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
-%
-To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:
-
-\begin{itemize}[leftmargin=*]
-    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
-    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, no failures on all 11 tests~\cite{shi2009testing}.
-\end{itemize}
-
-% Given that our interest is on AV simulations, this study presents a number of experiments run on typical agents used in AV simulations and their interactions with each other.
-These agents are classified into two, either vehicles(cars, motorcycles, bikes etc.) or pedestrians (can include animals).
-Here we will walk through our hypothesis, description of our tests and the analysis of results from the different tests.
-We will point out at each stage where we sit in Figure~\ref{method_diagram} as we go through this case study.
-
-% \subsection{Problem And Hypothesis}
-% We want to see whether using Unreal Engine 4 (UE4) along with CARLA \cite{CARLA_paper} for V\&V of AV simulations is deterministic or not; if not then what is the simulation variance that we will operate with, is it acceptable or not, and whether it worsens with stressing GPU and CPU.
-
-% Our hypothesis is that since UE4 is a game engine it is non-deterministic, however, the simulation variance would be very low to acceptable limits that allows verification engineers to use it confidently and assume that tests would be repeatable.
-This behaviour is expected not to hold as the level of computing stress increases.
-
- % targetting the closest waypoint but ignoring any waypoints closer than a fixed \textit{look-ahead distance} ($d_l$)\footnote{Targeting waypoints with $d_l<0.4$ tend to cause improper navigation with the implemented PID controller}.
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid static objects and other actors.
-To ensure they intersect (collide) with each other $d_l$ is altered; the smaller the look ahead distance the less chance the pedestrian will get to path plan around each other, and thus making sure they intersect.
-(This sounds like at some $d_l$ the pedestrians do not collide - is this true?)
-
-% Similar to the vehicle actors, the look ahead range ($d_l$) can be modified so that the actor is forced to navigate on waypoints that are further away.
-In test 7 $d_l$ is increased from $0.4m$ to $2.0m$ and in test 8 to $20m$ which was done to investigate if the path planning algorithm was deterministic when given waypoints further ahead.
-
-% The vehicles in this simulator follow a trajectory of waypoints using a PID controller and they have a look ahead distance that they target for from a given list of trajectory waypoints.
-Similarly with the pedestrians, but they don't have a PID controller and they can do dynamic path planning using A* algorithm to reach their destination or in this case a trajectory waypoint.
-
-% Starting from step 1 in the methodology diagram (Design Experiment); we are using a double mini roundabout for our experiments.
-
-% The list of tests shown in Table \ref{TableOfExperiments} were chosen to cover the mandatory interactions between the different types of agents.
-These are summed up in the following: \textbf{i)}Agents (vehicles and pedestrians) moving with no collisions or interactions with each other \textbf{ii)}Collision between vehicles \textbf{iii)}Collision (intersection) between pedestrians \textbf{iv)}Collision between vehicles and pedestrians.
-
-% Tests 1, 3and 5 are similar and there are no collisions in them, they were done as separate tests (instead of combining them into one) for the sake of simplicity and to use them as base lines for the tests-collision versions of them.
-The tests for collisions and non-collisions differ by a very slight change in the trajectory of the agents just to make them avoid colliding or intersecting with each other.
-
-% Test 2 is a collision between two vehicles and no pedestrians involved and is depicted by \ref{Test_a}, where the two vehicles will approach the roundabout and crash.
-
-% Tests 6,7 and 8 are tests with pedetrians having the same trajectory waypoints but walking in opposite directions to each other (see Figure~\ref{Test_b}). 
-% Pedestrians have the functionality of dynamic path planning i.e.
-they can avoid obstacles to reach their waypoints and so they can avoid bumping into each other.
-Hence, to make sure they intersect (collide) with each other then we  have set tests 6,7 and 8 to be the same but with different look ahead distances; the smaller the look ahead distance the less chance they will get to path plan around each other, and thus making sure they intersect.
-
-% In V\&V of AV simulations, collisions of pedestrians together is not of much interest, nevertheless we decided to include these tests for the sake of completeness and to obtain a full image of the simulation variance of the simulator.
-
-% Test 4, is where a vehicle hits a pedestrian and runs over it on a zebra line, as shown in Figure~\ref{Test_a}.
-
-% Proceeding to step 2; there is about 20 different internal settings related to our context that can be tweaked in UE4; such as \textit{Enable Enhanced Determinsim}, \textit{PhysX Tree Rebuild}, \textit{Max Physics Delta Time} etc.
-By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronus mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% On the other hand the external settings (step 3), where CPU + GPU are stressed, did show significant variations in the results as will be discussed in section \ref{ResultsSection}.
-Different levels of stress were applied to the computer; using the \texttt{stress} function \cite{CPU_stress} to stress the CPU and \texttt{gpu-burn} from github \cite{GPU_stress} to stress the GPU.
-
-% In terms of running the experiments (step 4), \textbf{i)} the computer used is an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM; \textbf{ii)} \texttt{nvidia-smi} and \texttt{htop} were used to monitor the levels of GPU and CPU stress respectively before and while the experiments were executed \cite{monitoring_stresses}; \textbf{iii)} the computer was
-
-% ***************************************************
-%  Results and Discussion
-% ***************************************************
-\section{Results and Discussion}\label{s:FinalResultsSection}
-
-\begin{figure*}[t]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
-        \caption{}
-        \label{CarsCollisionCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
-        \caption{}
-        \label{CarsCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to vehicle collision (Test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation.
-The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
-\end{figure*}
-
-\begin{figure*}[h]
-    \centering
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
-        \caption{}
-        \label{CarsPeopleCollsionsCG25}
-    \end{subfigure}
-    \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
-        \caption{}
-        \label{CarsPeopleCollisionPrePost}
-    \end{subfigure}
-    \caption{Vehicle to pedestrian collision (Test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
-\end{figure*}
-
-% {\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}
-
-A summary of the main results are shown in Table~\ref{TableOfExperiments}.
-In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
-%
-From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-%
-However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
-%
-\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$.
-However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm.
-}
-%
-Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
-
-Resource utilisation was found to have a significant impact on \textit{simulation variance}.
-%
-Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU).
-In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance.
-%
-Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
-%
-\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
-%
-
-A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).
-
-Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision.
-%
-The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-
-\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision.
-Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision.
-%
-This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}.
-This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}.
-Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}.
-This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
-}
-% \todo{Why ``not reliable''.}
-
-The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}.
-In this scenario (Test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}.
-There are three observations; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
-% as described above. 
-Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}.
-Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}.
-This final point should be of particular concern to verification engineers, developers and researchers in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous simulation results.
-% not just assertion checking
-
-% \todo[inline]{Make sure that assertion checking has been introduced before.}
-
-% \textbf{Strange gap post-collision for few seconds}
-
-% \textbf{Should we rename 'test' to 'scenario???'}
-
-% \todo[inline]{Big question why does high-RU influence simulation variance?? Can we tie this to thread scheduling and will NUMAclt change this conclusion?'}
-
-% Scenarios that include both collisions and vehicle actors were non-deterministic.
-However, when observing the deviation in actor paths against simulation time, Fig.~\ref{CarsCollisionCG25}, the simulation is deterministic prior to collision but not afterwards.
-
-% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton.
-Note the dotted red line in the plot shows the millimeter (mm) level.
-Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable.
-This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
-%
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v8.pdf}
-    \caption{Stages of the method proposed to determine the variance of a simulation.}
-    \label{method_diagram}
-\end{figure*}
-
-\todo[inline]{Sum this up and lead on to the Methodology - assuming that this section is placed before the Methodology section.
-This should cover that our results are clearly specific to our HW and provide evidence of non-determinism, but the results are not transferable.
-Thus, we present a general Methodology that others can re-use/apply to determine whether or not their sim is deterministic - in the next section.}
-
-To conclude, the main findings of this case study suggest a working practice that would minimise the factors that give rise to the non-deterministic effects observed in this investigation. 
-%
-By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table~\ref{TableOfExperiments}.
-By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve.
-%
-Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-
-A correlation between resource utilisation and simulation variance has been observed in these results which may be due to process scheduling.
-Rather than limiting utilisation of the whole system, another approach may be to promote the scheduling priority of the simulation process which is explored in Section~\ref{r:process_scheduling}.
-Furthermore, we were keen to investigate the impact of memory access and include a brief investigation of this in Section~\ref{r:numa}.
-%
-\todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application.
-%For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm. 
-}
-
-\subsection{Process Scheduling Priority} \label{r:process_scheduling}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
-    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
-    \label{NICEExperimentStressSummary}
-\end{figure}
-
-An investigation into the impact of process scheduling on the simulation variance was undertaken following the observations of increased simulation variance with increased resource utilisation.
-The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
-%
-To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.
-
-The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range of deviation, non-outlier limits by the whiskers and a horizontal bar for the median.
-The figure shows that decreasing the priority of the simulator process (right hand side of the  plot) has little effect on simulation variance when compared to a default NICE value of 0 (central bars in the plot).
-Increasing priority (left hand side of the plot) significantly reduced the variance for the 75\% resource utilisation experiment, but this does not account for all the difference in the observed results.
-This can be seen in the maximum priority setting where the bars in the plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling.
-%
-The remaining difference in the variance between the two resource utilisation levels may be due to the lack of absolute control that \texttt{NICE} has over process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}
-
-\subsection{Non-uniform memory access (NUMA)} \label{r:numa}
-An additional investigation into memory access was undertaken given the potential impact that data transfer cost and execution order may have on simulation variance, see Section~\ref{s:sources_numa}.
-%
-The program \texttt{numactl}~\cite{numactl_NUMA} allows a process to run with a specified memory placement policy, essentially allowing the process to be bound to a particular CPU core. %
-\texttt{numactl} was used to fix the simulator and test script to single cores, and a (2\%) improvement in simulation variance was observed. %
-This was considered minor comparative to the changes in simulation variance observed in other aspects of the case study, for example the pre- and post-collision between a vehicle and pedestrian saw a change of $10^{14}$.
-Therefore \texttt{numactl} was not used for subsequent testing.
-
-However, practitioners that are aiming to minimise simulation variance should use a technique such as this to minimise data transfer cost and potentially minimise any potential effect on process execution order that may lead to increased simulation variance.
-
-\subsection{Investigation Summary} \label{s:empirical_summary} 
-% {\color{red} Factor into overall conclusion of the case study}
-% \todo[inline]{GC-need to reference the 'restricted' set of results here too}
-
-%\todo[inline]{Do we need to say here why we think certain results are non-det, what the reasons might be??
-The results indicate that actor type, resource utilisation and collision events all affect the determinism in the CARLA simulator.
-Collision events may utilise more computational resources than simulations without such events, i.e.
-the physics loop takes longer to process as there are complex physical interactions to resolve.
-This may coincide with the second observation that higher resource levels also lead to non-deterministic results.
-We hypothesise that both these observations are due to improper management of thread scheduling within the engine.
-When resources are occupied there may be a delay in processing critical calculations for the simulator, which may arrive out of an expected sequence or the engine compensates for this delay by reducing the number of physics loops that can be performed within a render cycle.
-(is this 2nd one true, or does carla ignore render requests?)}
-These empirical investigations have highlighted the shortcomings of using a games engine for simulation based verification and suggests advice for best working practice. 
-%
-It was observed that resource utilisation positively correlates with simulation variance and that specific simulation events, such as vehicle collisions, can also lead to a breach in permissible tolerance.
-%
-The investigation found that the effect of higher simulation variance as a result of increased resource utilisation can be reduced, but not omitted entirely, by controlling the scheduling policy. 
-%
-The investigation into specifying memory placement on the simulation process did improve simulation variance but only by a minor amount of 2\%.
-
-However, these results are specific to the hardware and software used in the study and may not be transferable to other systems directly. 
-%
-Therefore we have derived a general methodology that practitioners can follow to find the 
-% \textit{deterministic domains of operation} 
-\textit{operational domains of permissible variance} 
-for %any hardware configuration.
-a game-engine-based simulation environment. This methodology is presented in the next section.
-
-% First, the tests with no collisions are considered, these represent tests 1, 3 and 5.
-Tests 1 and 3, which involves vehicles only and vehicles with pedestrians respectively, shows very low simulation variance for stress levels below 50\% and as the stress level increases they surpass the red line crossing our defined deterministic level but still being sufficiently low (i.e.
-with in the centimeter level).
-This clearly shows how the load on a computer (i.e.
-stressing CPU and GPU) can have drastic effects on the repeatablitiy of tests.
-This is more specific to vehicle agents rather than pedestrians since the behaviour of pedestrians is consistent and remarkably below the mm level for all of the different levels of stress shown by test 5, where only pedestrians are involved with no collisions, from which it can be deduced that the mean deviation change with the stress level in test 3 was solely due to the vehicles.
-
-% Next, given the pedestrians were showing low simulation variance, we ran tests for pedestrians colliding (i.e.
-with paths intersection) for different look ahead distances to exploit whether that would cause any changes, but yet as it can be seen by lines Test IDs' 6,7 and 8 that their performance is consistent for all of the different tests.
-
-% Finally, we look at the other collisions tests, tests 2 and 4 where the tests correspond to a collision between two vehicles and a collision between a vehicle and a pedestrian respectively.
-We observe that in both of these tests the mean deviation is always above our red line regardless of the stress level, nevertheless still increasing the stress levels shows worsening effects.
-
-% Thus, we resorted to plotting the mean deviation over the number of runs of these two tests vs simulation time with stress level \%25 chosen arbitrarily (see Figure~\ref{CarsCollisionCG25} and \ref{CarsPeopleCollsionsCG25}).
-Doing that we discovered that after collisions occur there is a jump in the mean deviation, which is what causes the averaged mean deviation through the whole simulation time in Figure~\ref{ExperimentsStressSummary} to always surpass our red line.
-
-% Plotting Figures~\ref{CarsCollisionPrePost} and \ref{CarsPeopleCollisionPrePost}, which shows the mean deviation pre and post collision for the different stress levels; it can be seen that there is two discrete levels for the pre and post collison devaition from these plots and as the stress increases these two discrete levels start to merge.
-However, in V\&V of AV simualtions we  are normally interested in the point until a collision occur after which the simulation will be terminated.
-Thus, the post collision effect should not be of much concern in the context of AVs.
-
-% It is important to note, however, that in test 4 a delayed effect of increase in simulation variance was noticed to Car ID 1, which interestingly was not involved in the collision, this was attributed to how the physics engine was programmed to calculate collisions which might cause other obscure behaviours to other agents.
-The importance of this outcome is that it shows even if an ego vehicle is not involved in a collision, but other agents are, this will still have effects on the simulation variance of the ego vehicle.
-Thus failing to stay below our red line and the test should be terminated.
-In summary, from the tests carried out we can deduce that for a computer of the following specifications, an Alienware Area 51 R5 with an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM, that all tests carried out using CARLA in UE4 are gauranteed to be deterministic and repeatable (according to our definition by the red line) as long as the CPU + GPU stress does not exceed 50\% and tests are terminated as soon as a collision of any type is detected in the simulation.
-left alone and not used for any other tasks while the runs were executed.
-
-% ***************************************************
-%  METHOD
-% ***************************************************
-\section{Method to determine the Variance of a Simulation} \label{s:methodology}
-% In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}.
-This method can be used to determine the operational domains of permissible variance of a simulation environment.
-%
-In this section a method for determining the simulation variance of actor paths and resolving the operational domains of permissible variance is presented here as a work flow, see Fig.~\ref{method_diagram}.
-%
-In addition, recommendations and best practice guidelines for minimising simulation variance are suggested.
-% , such as fixing the physics time step in \textit{Simulator Settings}. 
-% {\color{red} \sc Highlight these in the text.}
-% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
-% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}.
-%These need to be considered at different stages in the process.
-%
-
-The method consists of a sequence of five stages; experimental design, simulator settings, external settings, execution and analysis.
-In the following, each stage is described in detail with reference to the items listed for each stage in Fig.~\ref{method_diagram}.
-
-\subsection{Experimental Design}\label{s:design_experiment}
-
-% {\color{red} \sc Surely, a series of systematically designed experiments, each repeated X times, should be set up, where each experiment varies one of the suspected sources of non-determinism?
-Then, the analysis of the data from each experiment should reveal the actual sources of non-determinism for the given setting?
-In other words, we are aiming to fix as much as we can control (easily), and then we intentionally vary (ideally) single variables, e.g.\ the resource utilisation, to find out what makes a difference.
-We need a clear definition of the objectives of the experimental design, i.e.\ what one should vary and what one wants to ``nail down''.}
-
-% {\color{green!99} \sc }
-The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
-Each test is repeated $n$ times and the analysis of the results provide a confidence with a degree of statistical certainty.
-By varying a single parameter and controlling all others, the simulation variance associated with each source of non-determinism can be found and addressed.
-
-\subsubsection{Actors} \label{s:actors}
-
-% Different actor types may be handled differently by the game engine.
-For instance, our investigation indicated that CARLA pedestrians (termed \textit{walkers}) do not suffer from simulation variance under any conditions that were tested.
-However, the introduction of CARLA vehicles increased the simulation variance of the pedestrian-vehicle test by 12 orders of magnitude.
-{\color{red} \sc This needs rephrasing as it appears to contradict the statement above.}
-%
-All actors that could be included in the simulation should be tested, including any non-standard CARLA or bespoke actors including the ego vehicle, see Section~\ref{ego_actor}. 
-% {\color{red} \sc ... such as the ego vehicle...? [see section on ``External Settings'' last point - Ego Controller.]}
-%
-\subsubsection{Trajectories and Collisions} 
-% Actor paths should be hard-coded in a number of different tests that explore; no collisions, collisions between actors, and potentially collisions between actors and static scenery if this is likely to occur in the simulation.
-Actor paths, or the sequence of actions required to generate paths, should be hard-coded to ensure repeatability.
-These paths should include collisions between actors and potentially collisions between actors and static scenery if this is likely to occur in the simulation or as part of the verification process.
-Actor paths without collision are also important to include as these will serve as a baseline to the other tests; to see if deviations between runs in actor paths increases with collisions.
-% {\color{red} \sc What baseline?} {\color{green} \sc Done}
-
-\subsubsection{Permissible Variance} \label{s:threshold}
-The verification engineer should set the \textit{permissible variance}, also termed \textit{tolerance}.
-% which refers to the acceptable level of actor path variance. 
-% ($\sigma^2$), which is quoted in terms of deviation ($\sigma$) for ease of interpretation. 
-%
-The tolerance depends on the objectives of the simulation and the granularity at which the simulation environment operates.
-For example, this tolerance must be sufficiently small to enable accurate assertion checking and coverage collection, but not so small for assertion results to differ for repeated runs.
-In Section~\ref{s:tolerance} a tolerance of $1$cm was considered sufficient for urban scenario assertion checking.
-In practice, it may be necessary to determine this tolerance experimentally.
-%
-%The permissible variance is analogous to the analogue signal level associated with each of the binary states %of the signals in digital circuits.
-At some low analogue voltage the signal is interpreted as binary $0$, this may be, for example, any voltage below $0.5$V, and the same signal is interpreted as binary $1$ at some higher voltage, e.g.\ above $3.5$V.
-This convention enables the abstraction from the noisy physical signal values that are analogue voltages into the digital domain with binary representation.
-<- KIE: meant only to illustrate that other areas also have defined tolerances.
-%
-
-% \todo[inline]{Make sure assertion checking has previously been introduced.}
-
-% The method requires the assessment of scenarios that exercise different actor types and functional callbacks of the game engine.
-These may affect different aspects of the engine as outlined in Section~\ref{s:background}, such as actor navigation and collision callbacks.
-
-% \todo[inline]{Introduce the concept of a ``callback'' as part of the background section, especially ``collision callback''. GC change callback to 'event hit'}
-
-% The following section outlines a guide that may be used to determine the \textit{simulation variance} of an end user system.
-The aim of the experiment is to determine if the simulation is deterministic.
-This is achieved by monitoring actors in repeated tests and examining the deviation in path trajectories.
-Actors are observed in a number of different scenarios involving a mixture of vehicles and pedestrians exercising the demands outlined in Section~\ref{s:background}.
-
-% \todo[inline]{We may need to introduce the Case Study first, then derive the general Methodology from it.
-This way, we can explain the requirements of the case study, how things were set up, what was observed, how the methodology was refined, several iterations even, etc.}
-
-\subsubsection{Simulation Time} 
-% {\color{red} \sc Should this not be part of the Execution section?} 
-The simulation time will depend on the actor paths and terminating conditions.
-The simulation time should be sufficient to record interactions between actors but not so long that the testing take an inconvenient time to complete.
-In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25} in Section~\ref{s:FinalResultsSection}.
-The termination conditions of the simulation can be set by, for example, actors reaching their final trajectory waypoints.
- % {\color{red} \sc Point to the section that specified that.}
-
-% Depending on the number of repeated tests and the average deviation reported from the experiment, a maximum rather than average may be required if the results are close to the user-specified threshold.
-This is to ensure that an outlier data-point is not hidden in the average of $n$-tests.
-
-\subsection{Simulator Settings}
-The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$.
-If using CARLA, a small fixed value, say $0.05s$, can be set by using \texttt{setting.fixed\_delta\_seconds = 0.05}.
-In Unity, the default fixed time step is set to $0.02s$~\cite{MonoBehaviour_unity}.
-
-In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}.
-
-The use of random numbers must be controlled through fixed seeds, resulting in pseudo-randomness.
-Random numbers might be used in the simulation environment to control variations of background effects, e.g.\ weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment.
-%
-Actors that navigate through the environment should use a fixed navigation mesh. 
-%
-The version number of the CARLA and Unreal environment has also be shown to affect results, see~\cite{TSLUnrealEngineTesting}.
-Therefore, ensuring a consistent version number throughout testing is also important.
-
-% \todo[inline]{Select section titles to match the flow diagram.
-In each of these sections, go through the matching phase in the flow diagram, explaining each point listed there and justify recommendations.}
-% The experiment should be a short\todo[inline]{What does short mean? Quantify.} scenario that can be easily repeated with the same initial conditions.
-
-\subsection{External Settings} \label{ego_actor}
-
-\subsubsection{Resource Utilisation}
-% To simulate the demands of taxing simulations, potentially requiring realistic rendering or complex physics calculations in an otherwise simple scenario, stress programs should be used to artificially occupy system resources.
-The resources available to the simulator have been shown to have a significant effect on the variance of the path of simulated vehicles. %and exploring this as a variable the analysis is required. 
-%
-Thus, it is important to understand at what level of resource utilisation the system running the simulation becomes susceptible to \textit{simulation variance}.
-
-CPU utilisation software, such as the linux \texttt{stress} tool, which is a workload generator program, can be used to spawn workers on any number of cores or virtual threads on a system.
-This can be used to artificially increase the load on the system.
-For GPU utilisation, \texttt{gpu-burn} can be employed using the \texttt{fur test}.
-Different resolutions and multiple instances can be used to tune graphical utilisation levels~\cite{GPU_stress}.
-Reported values of resource utilisation can be obtained using the system monitors \texttt{htop} and \texttt{nvidia-smi} for CPU and GPU, respectively.
-These values should be added to the data logs.
-%
-Alternatively, in place of artificial resource utilisation, multiple instances of the simulation could be executed simultaneously.
-However, the granularity of control with this approach may be reduced.
-
-% \todo[inline]{Again, this sounds as if having the case study first makes more sense. That way we can explain wrt the experiments and then generalise to obtain the Methodology.}
-
-% \todo[inline]{This should be ``External Settings'' - from flow diagram?}
-
-\subsubsection{Memory Testing}
-Prior to experimental execution the system hardware should be tested for memory conformity and to ensure no single bit errors are occurring, see Section~\ref{s:screening}.
-For mainboard memory \texttt{memtest86}\ can be used on most platforms to run a series of pre-defined memory test patterns.
-This memory testing software can also be used for ECC enabled hardware.
-Similarly, to test memory on Nvidia based graphical adaptors \texttt{cuda\_memtest} can be used to ensure no memory errors exist.
-
-\subsubsection{Scheduling}
-We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
-However, gaining fine control over the scheduling policy and thread execution order is known to not be trivial~\cite{acm-q-rr-interview}.
-
-The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority.
-Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation.
-
-% If the policy or operating system cannot be changed then s
-Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
-
-\subsubsection{NUMA Control}
-Control over a Non-Uniform Memory Access policy can be achieved using \texttt{numactl} for multi-core processors with shared memory.
-This control allows the simulator process to be fixed on a single core, reducing and unifying memory access time.
-%
-Investigations in tests performed with NUMA control resulted in only minor improvements in simulation variance, see Section~\ref{s:screening}. 
-%
-% Using \texttt{numactl} allows a process to run with a specified memory placement policy, fixing the process to a single core.
-This will reduce the data transfer cost and the opportunity for disrupted execution order that may be responsible for increase simulation variance.
-%
-Using a fixed memory placement policy may assist if simulation variance is borderline to the tolerance but only of the order of a few percent from our observations.
-
-\subsubsection{Ego Vehicle Controller} 
-% {\color{red} \sc Why is this not covered under actors? [see my note in the Actors section]}
-% {\color{red} \sc Clearly separate the investigation from the Method.
-The fact that the case study did not have an ego vehicle should be mentioned as a limitation of the case study, rather than elaborated on here, but why does this make a difference?}
-%
-An ego vehicle was not used in this study, but the impact of introducing this to the simulator can be considered here.
-The ego vehicle is seen as another actor in the simulation but care must be taken to ensure that any control algorithms, machine learning modules and processing pipelines are deterministic.
-We recommended this be treated as a separate source of non-determinism and handled accordingly.
-
-% ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
-{\color{red} \sc Long and cumbersome sentence with strange grammar: No ego vehicle was used in the empirical investigation but this should be considered as an actor but ensuring that any randomness used in the controller is controlled with fixed seeds and that any adaptive learning algorithms should be controlled in the same manner.
-}
-
-\subsection{Execution}
-
-\subsubsection{Sample Size}
-% Tests should be sufficiently repeated using statistical power~\cite{cohen2013statistical} to reveal any outlier errors that occur with low probability.\todo[inline]{Explain what this means.} 
-% Histogram to show how sampling with position binning varies?
-% \footnote{\label{note1}Experiments using the Unreal Engine not reported here for brevity but code and experimental results can be found at \url{https://github.com/TSL-UOB/AV-Determinism/tree/master/UnrealEngineTests}}
-%
-% {\color{red} \sc Again, separate the study from the method.
-This observation should really have been made in the section on ``Experimental System Config and Pre-Screening, to justify the figure of 1000 repeats.}
-It is recommended that the chosen sample size is determined empirically. 
-%
-We recommend monitoring variance while increasing the sample size in orders of magnitude
-until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process. 
-%
-Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
-
-% By increasing the sample size and observing that the simulation variance is stable, one can infer that a sufficient number of samples have been collected.
-If the variance changes with a change in sample size then a stable condition has not been found.
-% ...
-% ...
-%  dependent on the number of results that exceed the tolerance. {\color{red} \sc How would one know?}
-% % To ensure outlier data points are not hidden in the average of a large number of tests, the maximum deviation should be reported rather than the average when summarising.
-
-\subsubsection{Data Logging}
-Unique identifiers should be assigned to each experiment, each repeat and each individual actor. 
-The time-stamped actor positions should then be recorded at fixed time intervals throughout the simulation in order to determine the variance in actor path.
-Additional information should also be logged such as the CPU and GPU utilisation levels and engine specific metrics such as game loop latency.
-% such as the \texttt{unit} command in Unreal Engine. 
-%For the subsequent analysis, each actor should have an identifier. The repeat number and simulation time should also be captured. 
-% The maximum deviation in actor path with respect to the mean for $n$ repeated tests will indicate at what dimensional scale deterministic results are achieved.
-For example, if 100 tests are executed and the maximum deviation in actor path is $<1$cm, then this scenario could be interpreted as being deterministic above this scale.
-
-\subsection{Analysis}
-For each experiment, the maximum value of actor path deviation over all time samples and actors, $\max\sigma$, should be analysed to identify which of the candidate sources of non-determinism require restriction or control to reach the domain of permissible variance within the simulation environment.
-
-%Finding these boundaries will identify the \textit{domains of permissible variance} for the user specific hardware.
-% KIE: Why now point to HW? And, which "boundaries"?
-
-% Game engine actors can be used to simulate the AV and other vehicles as well as pedestrians.
-For the experiment, actors should follow a series of trajectory waypoints which are hard-coded for each test type.
-
-% \todo[inline]{This is mixing our experiments (which are part of the case study) with the methodology.
-It is probably best to first introduce the case study and then derive the methodology from it, rather than mixing both.} In our experiments, hard-coded target location goals regularly spaced at $1m$ intervals were used for pedestrians and vehicles ensuring all actors were given the same initial conditions and any variation comes from execution of the simulator at runtime.
-Initial testing also indicated that when actors collide there is a significant change in \textit{simulation variance}, therefore the experiment should include some scenarios where collision callbacks are activated.\todo[inline]{Make sure collision callbacks have been introduced earlier.}
-
-% Randomisation must be fully controlled by fixing the seeds for any random number generation that is used during the experiment.
-It is important to identify and control any external sources that may use randomisation, e.g.\ external vehicle controllers or other clients connected to the simulation environment.
-
-% \todo[inline]{Abanoub: what is the distance between your hard-coded waypoints, are they regularly spaced? Might need to talk about interpolation code and give link to github}
-
-% \todo[inline]{In our experiment, gross waypoints were interpolated to give a regularly spaced set of coordinates approximately $1m$ apart, and this interpolated set of waypoints were used as "target location goals" (AB: not sure what the correct terminology is here) for the pedestrian actors.
-%
-% For the vehicles in the simulation, waypoints are generated and interpolated in the same way but execution at runtime is slightly different where a PID controller is used to generate a velocity profile and steering angle requests to reach the destination targets.}
-
-% We use game engine actors to simulate the AVs and other agents (vehicles and pedestrians).
-Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g.
-A* search algorithm.
-
-% \subsection{Internal settings}\todo[inline]{Systematically go through the points listed in the flow diagram, explain and justify.}
-% Where possible a fixed $dt$ should be used to give the most reliable physics calculations.
-
-% By doing various inspections around them we did not notice much improvement or worsening in the simulation variance between test runs, therefore the internal settings option was disregarded.
-However, one of the options we used is synchronous mode with a fixed time step that is available in CARLA.
-This would make the simulator proceed to the next tick only when commanded to do so by our script, which makes runs' flow more consistently.
-
-% There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine.
-% Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive.
-However, in such applications of V\&V of AV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.
-% If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the simulation variance as a result of allowing the AI to navigate in a more well defined space.
-% Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests.
-% The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.
-% Running in editor or deployed mode can sometimes cause differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.
-
-% \begin{table*}[b]
-% \centering
-% \begin{tabular}{clclcc}
-% \toprule
-% Test ID & Test description (See Figures~\ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
-% 1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
-% 2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
-% 3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
-% 4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
-% 5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
-% 6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
-% 7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
-% 8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
-% \bottomrule
-% \end{tabular}
-% \caption{Set of experiments}
-% \label{TableOfExperiments}
-% \end{table*}
-
-% The user should also monitor and record system CPU and GPU utilisation rates.
-Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
-
-% Collision callbacks
-% \subsection{Other Considerations}
-% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic?
-ii) If not, how does its simulation variance vary?
-In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination.
-Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary.
-% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision.
-Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
-% The user should set a deviation threshold, beyond which simulation results deemed unacceptable or non-deterministic.
-For the majority of AV navigation on major roads and highways a limit of $1m$ may be acceptable but in denser urban environments a value of $1cm$ may be more appropriate.
-
-% \subsection{External settings}
-% Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.
-Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing.
-Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.
-
-% \subsection{Running experiment}
-% \noindent When it comes to running the experiment itself, things that one should consider are the problem space exploration; the number of runs that should be executed to get reliable results; the frequency of logging data of actors; and possibly having a program to monitor the hardware utilisation.
-% Problem exploration, there is too many variables that one can play with. do statistical analysis some.... some of them get problematic...
-
-% ***************************************************
-%  CONCLUSION
-% ***************************************************
-% \section{Conclusions and Future Work}\label{s:conclusion}
-\section{Conclusions \& Future Work}\label{s:conclusion}
-
-% \todo[inline]{Say what the paper investigated and what we aimed to achieve, why this is important and how we went about it.
-Then point out the derivation of a re-usable methodology based on the knowledge and understanding we gained from the case study.}
-
-% \todo[inline]{GC: we need to say what we set out to achieve and why that is important, then say how our investigation led us to several theories which resulted in understanding several key parameters which directly affect simulation variance.
-From this we developed a methodology that the community can use to understand the deterministic operational domains of their own specific hardware:  Autonomous vehicle testing is adopting the use of game engines for development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-An empirical investigation uncovered several parameters that contributed to simulation variance ...
-explain results ...
-This led to the design of a methodology allowing 'users' to determine the deterministic operational domains for a specific hardware/platform configuration ...
-}
-
-%The autonomous vehicle community are adopting game engine simulators for such purposes as control system development and verification. 
-Game engines offer simulation environments that are used for the development and verification of autonomous driving functions. 
-%
-Determinism of a simulator is required to achieve repeatability, which is essential to find and fix software bugs efficiently, and to ensure simulation results are trustworthy. 
-%
-If a simulator is non-deterministic then practitioners should at least be aware of, and know how to find, the operational domains where \textit{simulation variance} is tolerable.
-
-An investigation into the CARLA simulator revealed a significant simulation variance for repeated tests with the same initial conditions and event history, indicating non-determinism of the simulation.
-%
-We then researched, identified and discussed potential sources for non-determinism in this context.
-%
-A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution.
-In particular, actor collisions and system-level resource utilisation were identified as key contributors.
-%
-To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
-%
-% {\color{red} \sc Say what the recommendations were for the required tolerance of the case study, i.e.\ summarise our results/recommendations here and point to github for data/detail.
-This seems to have been commented out but was here before.}
-
-A general method to assess the actor path variance of a game-engine-based simulation environment is then proposed.
-The method can be used to find the \textit{domains of permissible variance} of a simulation environment for a given system configuration.
-This can give AV developers and verification engineers increased confidence in the simulation results and reduce debug time.
-% 
-% \subsection{Future Work}
-%
-As future work the method can be extended to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
-
-An ambitious avenue for future work is the development of a deterministic simulator for AV development and verification.
-This requires controlling all potential sources of non-determinism, including randomness and scheduling, very similar to the development of the record-and-reply debugger \texttt{rr}~\cite{RR_link}, originally developed to catch low-frequency non-deterministically failing tests at Mozilla~\cite{acm-q-rr-interview}.
-
-%With the advent of more AI systems becoming adaptive, meaning that the system may change it's response to the same stimulus over time, the notion of repeatability will need reassessing in the context of verification.
-The main research question here is what will constitute `the same' which will be important for passing or failing verification tests but also for determining coverage.
-
-% \todo[inline]{Review/rewrite this to match the earlier changes in the paper.
-Practitioners in this field can resolve the non-deterministic operational domains of their simulation by following the method outlined in this paper.
-Potential sources of non-deterministic simulation behaviour have been discussed such as floating-point rounding error and scheduling policy.
-A case study was used to ascertain the boundary between deterministic and non-deterministic operational domains of the CARLA driving simulator.
-Results suggest non-deterministic outcomes when resource utilisation levels exceeded 75\% or when vehicle collisions occur for the particular hardware used in this study.
-Furthermore it was found that all actors, even those not involved directly in a collision, will behave non-deterministically if a collision callback to the game engine is made.
-Best operating practice follows that the simulation should be stopped as soon as a collision callback is detected or resource utilisation levels exceed 75\% although this value may be system specific.
-}
-
-% \todo[inline]{GC: think we should put more emphasis on the outcomes of the method rather than the cast study.}
-% \todo[inline]{Point out (again) that all data and scripts etc are on github.}
-
-% \todo[inline]{Future work...}
-
-% In this paper we explored the determinisim of game engines.
-This is important for autonomous vehicles' development and verification of vehicle control systems in simulation.
-Simulations must be deterministic or practitioners should at least be aware of the operational domains for reliable results and repeatability of tests.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, uncovered several parameters that contributed to simulation variance.
-These parameters include numerous internal settings based in Unreal Engine, and others that are external mainly dependant on CPU and GPU utilisation.
-
-% Changes in CPU and GPU resource utlisation has shown to be the main influencer for the non-deterministic behaviour of game engines.
-This is clearly demonstrated in the results (see section \ref{s:case-study}), where the max deviation between runs increased from the orders of $10^{-12}$m to devaitons of up to 1m as the CPU and GPU resource utilisation increased above 50\%.
-
-% The term \textit{deterministic operational domain} was given to the range of system resource utlisations that provides low simulation variances that is insignificant for development and verifcation of autonomous vehicles.
-% The deterministic operational domain specified herein is specific to our hardware configuration.
-
-% This led to the design of a methodology allowing `users' to determine the deterministic operational domain for a specific hardware configuration.
-This will help guide specialists in the field of AVs to setup their custom experiments for their specific simulators and identify their specific deterministic operational domains.
-
-% Future work; experts in the field of game engines' architecture can use our empirical investigation and the potential sources of non-determisnim outlined in our work to investigate and develop plugins that can switch priority between performance and determinisim in game engines.
-Hence allowing its extensive usgae by both the autonmous vehicles' and the gaming community.
-
-% presented an investigation (based on CARLA driving simulator) to show whether game engines are deterministic or not. 
-% Hence assess their suitablility for usage in development and verification of AVs, where determinisim is key in order for one to eliminate bugs.
-
-% An empirical investigation based on Unreal Engine running CARLA driving simulator, was used to illustrate the challenges of determinisim faced by game engines. 
-% % It was found that game engines are non-deterministic.
-
-% We reviewed possible sources of non-determinisim in section \ref{s:background} and investigated the effect of each on the simulation variance.
-
-% We found that the sources of non determinsim that had a significant effect on simulation variance were attributed to CPU and GPU resource utilisation. 
-% The case study clearly shows how the max simulation variance changes from very low simulation varicances, below picometer level ($10^-12$), to centimeter level when the CPU and GPU resource utilisation increases above 50\%.
-
-% We found two main outcomes from our investigation: 
-% 1) Game engines are non-deterministic and we outlined potential sources for non-determisnitc behaviours in section \ref{s:background}, e.g. runtime scheduling. 
-% 2) The level of simulation variance changes significantly depending on several settings relating internally and externally to the game engine used; the main two factors being CPU and GPU resource utilisation.
-% Thus, by altering these settings one can achieve sufficiently low levels of simulation varince that can allow users to assume the simulation is deterministic. 
-% We phrased the level below which the simulation is to be assumed determinisitic as the \textit{determinisitic threshold}, which is up to the user to define where it exactly lies. 
-% This allows for one to use these game engines knowing that their simulation will be sufficiently deterministic provided that they bound their simulation within a certain configuration.
-
-% The configuration presented in here is specific to the hardware and software used in our investigation. 
-% Therefore, we outlined a methodology in this paper to allow practitioners in this field of AVs to resolve the non-deterministic operational domains of their simulation.
-
-% We have provided a summary of the sources of non-determinisim in game engines.
-This was followed by presenting a method by which one can determine the simulation variance of a game engine used in V\&V application of AV simulations.
-The method consists of five stages, starting with designing the experiment, then suggesting the various settings that can improve or worsen the performance, and finally focuses on running the experiment and analysing the output.
-
-% A case study is then used to show the implementation of the method introduced, where several tests were setup and run to explore the different interactions between the different agents in a simulator called CARLA.
-It was found that for the computer used (with specs of an i9 9960X CPU chip, NVIDIA GeForce RTX 2080 graphics card and 64GB of RAM) the simulation variance was sufficiently low to assume the simulator to be deterministic as long as the CPU + GPU utilisation is below 50\% and that tests are terminated as soon as a collision of any type is detected.
-%since it was noticed that collisions cause the level of non determinism to jump remarkable invalidating the assumption of the simulator is deterministic.
-
-% **** These results are not meant as a universal guide but merely to illustrate our method for a single hardware configuration.
-It is probable that the values quoted here will not translate to individual end-users unless the system configuration is very similar.
-This paper should be used as a method for users to determine what scenarios and system setting are appropriate for their particular verification needs.
-
-% You must monitor all collisions, not just for the ego vehicle as any collision callback may result in high simulation variance and hence non-deterministic results for your experiments.
-
-% In highlight of obtaining a fully deterministic simulator, one would either have to explore methods to control the scheduling of threads in a computer such that if tests are rerun the utilisation of hardware will be the same.
-Alternatively, another approach would be to create a new engine that is designed to be deterministic with minimal attention given to performance.
-
-% ***************************************************
-%  FUNDING, REFS
-% ***************************************************
-% \section*{Acknowledgements}
-% This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
-% projects are part-funded by the Centre for Connected and Autonomous
-% Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
-% 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-\printbibliography
-
-\end{document}

### 2021-04-12 document/source/body.tex
new document build structuring with common source

--- 2021-04-09 paper_arxiv/arxiv_cav_determinisim_main.tex
@@
-\RequirePackage{etex} %% **************** add this for ARXIV submission
-\documentclass[runningheads,twocolumn,a4paper,10pt]{llncs}
-
-\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
-\addbibresource{cav_determinism.bib}
-
-% \documentclass[runningheads,a4paper]{llncs}
-\usepackage{etex}
-\reserveinserts{28}
-\pdfoutput=1
-\setcounter{tocdepth}{3}
-\usepackage[a4paper, total={7in, 10in}]{geometry}
-\usepackage[linesnumbered,lined,boxed,commentsnumbered,ruled]{algorithm2e}
-\renewcommand{\algorithmcfname}{ALGORITHM}
-\SetAlFnt{\small}
-\SetAlCapFnt{\small}
-\SetAlCapNameFnt{\small}
-\SetAlCapHSkip{0pt}
-\IncMargin{-\parindent}
-\usepackage{ textcomp }
-\usepackage{ comment }
-\usepackage[color=yellow!60,textsize=footnotesize,obeyDraft,draft]{todonotes}
-\usepackage{colortbl}
-\usepackage{booktabs}
-\usepackage{capt-of}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{listings}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage[font=footnotesize,labelfont=bf]{subcaption}
-\usepackage[hidelinks]{hyperref}
-\usepackage[capitalize]{cleveref}
-\captionsetup{compatibility=false}
-\usepackage{tikz}
-\usepackage{pgfplots}
-\usepackage[para,online,flushleft]{threeparttable}
-\usepackage{wasysym}
-\usepackage{amssymb}
-\usepackage{enumitem}
-\usetikzlibrary{arrows}
-% \usepackage{subfig}
-\usepackage{breakcites}
-\usepackage{array}
-\usepackage{color}
-\usepackage{balance} %balance columns
-\usepackage{lastpage}
-\hyphenation{op-tical net-works semi-conduc-tor}
-\usepackage{dblfloatfix}
-\usepackage{authblk}
-% \usepackage[disable]{todonotes}
-\usepackage{siunitx}
-\usepackage{xurl} %for long URLs
-
-% \usepackage{multicol}
-\setcounter{secnumdepth}{3}
-
-% Make tables a little bit easier to read
-\renewcommand{\arraystretch}{1.15}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% NOTES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcounter{mnotei}
-\setcounter{mnotei}{0}
-\newcommand{\mnote}[1]{%
-  {\scriptsize\textsf{$^{\textcolor{red}{[n.\themnotei]}}$}}%
-  \marginpar{\scriptsize\textsf{\textcolor{red}{[n.\themnotei]: #1}}}%
-  % \marginpar{\scriptsize\textsf{n.\themnotei: #1}}%
-  \stepcounter{mnotei}
-}
-%\renewcommand{\mnote}[1]{}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Provide paragraph-style table columns
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-% Give more marginpar
-\setlength{\marginparwidth}{25mm}
-
-% % Prompt people to generate graphs
-% \newcommand{\includegraphicsmaybe}[2]{
-%     \IfFileExists{#2}{\includegraphics[#1]{#2}}{
-%     \detokenize{File #2 is missing, maybe you need to run plots.py?}
-% }}
-
-% \newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
-
-\begin{document} 
-% \bibliographystyle{ieeetr}
-\mainmatter
-
-\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-
-\titlerunning{On Determinism of Game Engines... }%used for Simulation-based }%Autonomous Vehicle Verification}
-
-\author{Greg Chance\inst{1,3}, Abanoub Ghobrial\inst{1,3}, Kevin McAreavey\inst{1,3}, S\'everin Lemaignan\inst{2,3}, Tony Pipe\inst{2,3}, Kerstin Eder\inst{1,3}}
-
-\authorrunning{G. Chance, A. Ghobrial et al.}
-\institute{University of Bristol, Bristol, UK \and University of the West of England, Bristol, UK \and Bristol Robotics Laboratory, Bristol, UK}
-\tocauthor{Authors' Instructions}
-
-\maketitle
-\let\thefootnote\relax\footnotetext{\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper and are both corresponding authors.}
-Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
-Abanoub Ghobrial (e-mail: abanoub.ghobrial@bristol.ac.uk), 
-Kevin McAreavey (e-mail: kevin.mcareavey@bristol.ac.uk), 
-and 
-Kerstin Eder (e-mail: kerstin.eder@bristol.ac.uk) 
-are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom. 
-%
-S\'everin Lemaignan (e-mail: severin.lemaignan@brl.ac.uk)
-and
-Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom. }.
-\makeatletter
-\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
-                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
-                       {4\p@ \@plus 2\p@ \@minus 2\p@}%
-                       {\normalfont\normalsize\bfseries\boldmath
-                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
-\makeatother
-
-% \begin{multicols}{2}
-
-\textbf{\textit{Abstract}--Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
-%
-A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history.
-%
-Thus, in a deterministic simulation environment, tests are rendered repeatable and yield simulation results that are trustworthy and straightforward to debug. 
-%
-However, game engines are seldom deterministic. 
-%
-This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines. 
-%
-A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
-%
-Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
-%
-Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.}
-
 % ***************************************************
@@
     \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
         \caption{Non-deterministic}
@@
     \begin{subfigure}{.48\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
         \caption{Deterministic}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoopv2.pdf}
+\includegraphics[width=0.5\textwidth]{../other/figures/GameEngineLoopv2.pdf}
 \caption{Game engine loop block diagram~\cite{GameProgPatternsBook}.}
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_a.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/TestCasesDiagramV2_a.pdf}
         \caption{}
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/TestCasesDiagramV2_b.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/TestCasesDiagramV2_b.pdf}
         \caption{}
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/ExperimentsStressSummaryV6.pdf}
+    \includegraphics[width=0.48\textwidth]{../other/figures/ExperimentsStressSummaryV6.pdf}
     \caption{Summary of results showing maximum deviation for each scenario against different resource utilisation levels.
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/CarsCollisionCG25_V4.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/CarsCollisionPrePostV5.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/CarsPeopleCollsionsCG25_V3.pdf}
         \caption{}
@@
     \begin{subfigure}{.49\textwidth}
-        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/CarsPeopleCollisionPrePostV5.pdf}
         \caption{}
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v7.pdf}
+    \includegraphics[width=0.99\linewidth]{../other/figures/Methodology_Diagram_v7.pdf}
     \caption{Stages of the method proposed to determine the variance of a simulation.}
@@
     \centering
-    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
+    \includegraphics[width=0.48\textwidth]{../other/figures/NICE_analysis_summary_V4.pdf}
     \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
@@
 Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-%
-\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper.}
-% \bibliography{cav_determinism.bib}
-\balance
-\printbibliography
-\end{document}
-\grid

### 2021-04-12 document/arxiv/arxiv_template.tex
TP corrections up to 2.1.3

@@
 \documentclass[runningheads,twocolumn,a4paper,10pt]{llncs}
-
 \usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
 \addbibresource{../source/cav_determinism.bib}
-
 % \documentclass[runningheads,a4paper]{llncs}
@@
 \setcounter{secnumdepth}{3}
-
-% Make tables a little bit easier to read
-\renewcommand{\arraystretch}{1.15}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% NOTES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcounter{mnotei}
-\setcounter{mnotei}{0}
-\newcommand{\mnote}[1]{%
-  {\scriptsize\textsf{$^{\textcolor{red}{[n.\themnotei]}}$}}%
-  \marginpar{\scriptsize\textsf{\textcolor{red}{[n.\themnotei]: #1}}}%
-  % \marginpar{\scriptsize\textsf{n.\themnotei: #1}}%
-  \stepcounter{mnotei}
-}
-%\renewcommand{\mnote}[1]{}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Provide paragraph-style table columns
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-% Give more marginpar
-\setlength{\marginparwidth}{25mm}
-
-% % Prompt people to generate graphs
-% \newcommand{\includegraphicsmaybe}[2]{
-%     \IfFileExists{#2}{\includegraphics[#1]{#2}}{
-%     \detokenize{File #2 is missing, maybe you need to run plots.py?}
-% }}
-
-% \newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
+% % Make tables a little bit easier to read
+% \renewcommand{\arraystretch}{1.15}
+% \newcounter{mnotei}
+% \setcounter{mnotei}{0}
+% \newcommand{\mnote}[1]{%
+%   {\scriptsize\textsf{$^{\textcolor{red}{[n.\themnotei]}}$}}%
+%   \marginpar{\scriptsize\textsf{\textcolor{red}{[n.\themnotei]: #1}}}%
+%   % \marginpar{\scriptsize\textsf{n.\themnotei: #1}}%
+%   \stepcounter{mnotei}
+% }
+% % Provide paragraph-style table columns
+% \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+% \newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+% \newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+% % Give more marginpar
+% \setlength{\marginparwidth}{25mm}
 
 \begin{document} 
-% \bibliographystyle{ieeetr}
-\mainmatter
+% \mainmatter
 
@@
 
-\titlerunning{On Determinism of Game Engines... }%used for Simulation-based }%Autonomous Vehicle Verification}
+\titlerunning{On Determinism of Game Engines... }
 
@@
 Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom. }.
+are with the Bristol Robotics Lab, T Block, University of the West of England, Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.}
 \makeatletter

### 2021-04-12 document/source/abstract.tex
TP corrections up to 2.1.3

@@
 However, game engines are seldom deterministic.
-This paper reviews and identifies the potential causes of non-deterministic behaviours in game engines.
+This paper reviews and identifies the potential causes and effects of non-deterministic behaviours in game engines.
 A case study using CARLA, an open-source autonomous driving simulation environment powered by Unreal Engine, is presented to highlight its inherent shortcomings in providing sufficient precision in experimental results.
-Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently low i.e.\ variance between repeated executions becomes negligible for development and testing work.
+Different configurations and utilisations of the software and hardware are explored to determine an operational domain where the simulation precision is sufficiently high i.e.\ variance between repeated executions becomes negligible for development and testing work.
 Finally, a method of a general nature is proposed, that can be used to find the domains of permissible variance in game engine simulations for any given system configuration.

### 2021-04-12 document/source/body.tex
TP corrections up to 2.1.3

@@
 
-Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules, e.g.\ through assertion checking during simulation. 
+Design confidence is gained when the autonomous vehicle (AV) can be shown to comply with these rules e.g.,\ through assertion checking during simulation.
 There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
@@
 State-of-the-art game engines provide a convenient option for simulation-based testing.
-They offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
+Strong arguments exist that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses.
 %
@@
 %
-Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required. 
+Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be provided if required. 
 %
-Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation.
+Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of the autonomous features of AVs in simulation.
 %
-However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include determinism.
+However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include the issue of determinism.
 
-In this paper, we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
+In this paper, we investigate non-determinism and how it affects simulation results using the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
 %
@@
 We find that the CARLA simulator is non-deterministic.
-Actor path variance was found to be non-zero under certain conditions where the deviation from the mean was observed up to $59$cm.
+Actor path variance was found to be non-zero and, under certain conditions, the deviation from the mean was observed up to $59$cm.
 We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
-CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
+However, we found that CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
 %
@@
 %
-Our case study of simulation variance for a number of scenarios involving pedestrian and vehicle settings using CARLA is presented in Section~\ref{s:case-study}.
+An empirical case study of simulation variance for a number of scenarios involving pedestrian and vehicles is given in Section~\ref{s:case-study} including internal and external setting and system screening tests.
+%
+The results from the case study are presented in Section~\ref{s:FinalResultsSection}. 
 %
@@
 %
-In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing exact replica.
+In practice, full accuracy will often not be achievable due to modelling and associated computational demands of creating and executing an exact replica.
 In most cases it is unnecessary and some authors state that `just the right amount of realism' is required to achieve valid simulation results~\cite{Koopman2018}.
@@
 This requires controlling all potential sources of non-determinism, including randomness and scheduling, very similar to the development of the record-and-reply debugger \texttt{rr}~\cite{RR_link}, originally developed to catch low-frequency non-deterministically failing tests at Mozilla~\cite{acm-q-rr-interview}.
-
-\section*{Acknowledgement}
-This research has in part been funded by the ROBOPILOT and CAPRI projects.
-Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.

### 2021-04-13 document/source/body.tex
up to pg.10 TP comments

@@
 %
-However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include the issue of determinism.
+However, while game engines are designed primarily for performance to achieve a good user experience, the requirements for AV verification go beyond that and include \textit{determinism}.
 
@@
 
-Investigating the determinism of game engines has not attracted much research interest since performance is more critical for game developers than accurate and repeatable execution.
+Investigating the determinism of game engines has not attracted much research interest to date since performance is more critical for game developers than accurate and repeatable execution.
 Ensuring software operates deterministically is a non-trivial task.
@@
 Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
-This allowed control over the scheduling and execution order of threads promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
+This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
 %
@@
 
-\begin{table*}[b]
+\begin{table*}[t]
 \centering
@@
 
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{../other/figures/NICE_analysis_summary_V4.pdf}
+    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
+    \label{NICEExperimentStressSummary}
+\end{figure}
+
 \begin{figure*}[t]
@@
 \subsection{Process Scheduling Priority} \label{r:process_scheduling}
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{../other/figures/NICE_analysis_summary_V4.pdf}
-    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
-    \label{NICEExperimentStressSummary}
-\end{figure}
 

### 2021-04-14 document/source/body.tex
TP comments complete

@@
 
-The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all others parameters constant.
+The experiment design is based around a series of carefully selected tests that varies one of the sources of non-determinism whilst keeping all other parameters constant.
 Each test is repeated $n$ times and the analysis of the results provide a confidence with a degree of statistical certainty.
@@
 The simulation time will depend on the actor paths and terminating conditions.
-The simulation time should be sufficient to record interactions between actors but not so long that the testing take an inconvenient time to complete.
+The simulation time should be sufficient to record interactions between actors but not so long that the testing takes an inconvenient time to complete.
 In our empirical investigation a simulation time of $10-20$s was sufficient to monitor the distinct change in events such as the pre- and post-collision including the delayed effect seen by other actors, shown in Fig.~\ref{CarsPeopleCollsionsCG25} in Section~\ref{s:FinalResultsSection}.
@@
 We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
-However, gaining fine control over the scheduling policy and thread execution order is known to not be trivial~\cite{acm-q-rr-interview}.
+However, gaining fine control over the scheduling policy and thread execution order is non-trivial~\cite{acm-q-rr-interview}.
 
@@
 The ego vehicle is seen as another actor in the simulation but care must be taken to ensure that any control algorithms, machine learning modules and processing pipelines are deterministic.
-We recommended this be treated as a separate source of non-determinism and handled accordingly.
+We recommended that this be treated as a separate source of non-determinism and handled accordingly.
 
@@
 \subsubsection{Sample Size}
-It is recommended that the chosen sample size is determined empirically.
+It is recommended that the chosen sample size, i.e.
+the number of repeated tests, is determined empirically.
 We recommend monitoring variance while increasing the sample size in orders of magnitude until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process.
@@
 
-A general method to assess the actor path variance of a game-engine-based simulation environment is then proposed.
+A general method to assess the actor path variance of a game-engine-based simulation environment was then proposed.
 The method can be used to find the \textit{domains of permissible variance} of a simulation environment for a given system configuration.
 This can give AV developers and verification engineers increased confidence in the simulation results and reduce debug time.
-As future work the method can be extended to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
+As future work, the method can be extended to other simulation platforms and to criteria other than the actor path, e.g.\ actor orientation and any status indicators that may be of interest, also including actions, sequences and timings that may be useful for verification purposes.
 

### 2021-04-16 document/ieee_trans/det_ieeetrans.tex
rename top level files, fixed oreilly ref error

--- 2021-04-12 document/ieee_trans/ieeetrans_template.tex
@@
 \input{../source/body.tex}
-\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper.}
+
+% ***************************************************
+%  Thanks
+% ***************************************************
+% \section*{Acknowledgement}
+% This research has in part been funded by the ROBOPILOT and CAPRI projects.
+Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
 
@@
 
-\begin{IEEEbiography}{Severin Lemaignan}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/severin}}]{S\'everin Lemaignan}
+S\'everin Lemaignan is Associate Professor in Social Robotics and AI at the Bristol Robotics Laboratory, University of the West of England, Bristol.
+He obtained a joint PhD in Cognitive Robotics from the CNRS/LAAS (France) and the Technical University of Munich (Germany) in 2012.
+He then joined the EPFL (Switzerland) and Plymouth University (UK) as post-doc, then lecturer in Robotics until 2018, when he joined the Bristol Robotics Lab.
+His research interest primarily concerns the socio-cognitive aspects of human-robot interaction and he has been focusing his recent experimental work on child-robot interaction and human-in-the-loop machine learning for social robots.
+Pr.
+S\'everin Lemaignan has been involved in several European projects related to social and cognitive robotics (CHRIS, DREAM, L2TOR) and was  awarded in 2015 a EU H2020 Marie Sk\l{}odowska-Curie Individual Fellowship for his project on robots and theory of mind.
 \end{IEEEbiography}
@@
 His research interests focus on safe and correct operation whilst in close-proximity to human beings.
-He has co-authored over 200 international refereed publications, supervised 35 PhDs, and created a BRL income of >£10M since 2013.
+He has co-authored over 200 international refereed publications, supervised 35 PhDs, and created a BRL income of \pounds£10M since 2013.
 He has acted as BRL-UWE PI on the EPSRC grant ‘Trustworthy Robotic Assistants’, the Programme/Hub grants ‘Robotics for Nuclear Environments’ and ‘National Centre for Nuclear Robotics’, and the Centre for Doctoral Training in Robotics and Autonomous Systems.

### 2021-04-16 document/ieee_trans/det_ieeetrans.tex
KIE bio and picture added

@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/greg.png}}]{Greg Chance}
-Greg Chance received the B.Sc.
+received the B.Sc.
 (Hons.) and Ph.D.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/abanoub}}]{Abanoub Ghobrial}
-Abanoub Ghobrial received the MEng degree in mechanical engineering from the University of Bristol, Bristol,  U.K., in 2018.
+received the MEng degree in mechanical engineering from the University of Bristol, Bristol,  U.K., in 2018.
 He is currently pursuing the Ph.D.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/kevin.jpg}}]{Kevin McAreavey}
-Kevin McAreavey received a BSc (Hons) in computer science from Queen's University Belfast (QUB) in 2010 followed by a PhD in AI in 2014.
+received a BSc (Hons) in computer science from Queen's University Belfast (QUB) in 2010 followed by a PhD in AI in 2014.
 He is currently a Senior Research Associate with the Intelligent Systems Laboratory at the University of Bristol (UOB), working on explainable AI for AI-enabled domestic life.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/severin}}]{S\'everin Lemaignan}
-S\'everin Lemaignan is Associate Professor in Social Robotics and AI at the Bristol Robotics Laboratory, University of the West of England, Bristol.
+is Associate Professor in Social Robotics and AI at the Bristol Robotics Laboratory, University of the West of England, Bristol.
 He obtained a joint PhD in Cognitive Robotics from the CNRS/LAAS (France) and the Technical University of Munich (Germany) in 2012.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/tony}}]{Tony Pipe} 
-Tony Pipe received his BSc (Hons.) degree in Electronic Engineering from Warwick in 1979 and, after an eight-year period in industry that included running a small company, received his PhD in Robotics from the University of the West of England (UWE) in 1997.
+received his BSc (Hons.) degree in Electronic Engineering from Warwick in 1979 and, after an eight-year period in industry that included running a small company, received his PhD in Robotics from the University of the West of England (UWE) in 1997.
 Since 2010 he has been a Professor of Robotics and Autonomous Systems at UWE.
@@
 
-\begin{IEEEbiography}{Kerstin Eder}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/kie}}]{{Kerstin Eder}} 
+is Professor of Computer Science and heads the Trustworthy Systems Laboratory at the University of Bristol.
+She also leads the Verification and Validation for Safety in Robots research theme at the  Bristol Robotics Laboratory.
+She has gained extensive experience of verifying complex microelectronic designs while working with leading semiconductor design and Electronic Design Automation companies worldwide.
+Her research is focused on specification, verification and analysis techniques to verify or explore a system's behaviour in terms of functional correctness, safety, security, performance and energy efficiency.
+Prof.\ Eder's most recent contributions include agent-based testing, how to use assertions and theorem proving to verify control system designs, energy modeling of software and static  analysis to predict energy consumption of programs.
+She holds a PhD in Computational Logic and an MSc in Artificial Intelligence from the University  of Bristol, UK, as well as an MEng in Informatics (Dipl.\,Inf.) from the Technical University Dresden, Germany.
+In 2007 she was awarded an ``Excellence in Engineering'' prize from the Royal Academy of Engineering.
+
 \end{IEEEbiography}

### 2021-04-16 document/ieee_trans/det_ieeetrans.tex
a few typos corrected

@@
 Kerstin Eder (e-mail: kerstin.eder@bristol.ac.uk) 
-are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road,  Bristol, BS8 1UQ, United Kingdom.
+are with the Trustworthy Systems Lab, Department of Computer Science, University of Bristol, Merchant Ventures Building, Woodland Road, Bristol, BS8 1UB, United Kingdom.
 
@@
 Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.
+are with the Bristol Robotics Lab, T Block, University of the West of England, Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.
 
@@
 His research interests focus on safe and correct operation whilst in close-proximity to human beings.
-He has co-authored over 200 international refereed publications, supervised 35 PhDs, and created a BRL income of \pounds£10M since 2013.
+He has co-authored over 200 international refereed publications, supervised 35 PhDs, and created a BRL income of \pounds10M since 2013.
 He has acted as BRL-UWE PI on the EPSRC grant ‘Trustworthy Robotic Assistants’, the Programme/Hub grants ‘Robotics for Nuclear Environments’ and ‘National Centre for Nuclear Robotics’, and the Centre for Doctoral Training in Robotics and Autonomous Systems.
-Prof.
-Pipe has been involved with Connected Autonomous Vehicle research and development since 2015 via the Innovate UK/CCAV funded projects VENTURER, FLOURISH, MultiCAV, CAPRI, ROBOPILOT and CAV-Forth.
+Prof.\ Pipe has been involved with Connected Autonomous Vehicle research and development since 2015 via the Innovate UK/CCAV funded projects VENTURER, FLOURISH, MultiCAV, CAPRI, ROBOPILOT and CAV-Forth.
 \end{IEEEbiography}

### 2021-04-19 document/source/body.tex
rename top level files, fixed oreilly ref error
added a-star section

@@
 \subsection{Actor Navigation}
-There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
-The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
+There is existing evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
+Unreal Engine is the underlying framework for the CARLA simulator which uses the A* search algorithm to find optimal paths for actor navigation~\cite{newton2016unreal}.
+The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.
+
+To briefly describe the A* process, at each stage of the search, the algorithm selects one node for expansion from a set of nodes known as the \textit{frontier} (also fringe or open set) using a heuristic to select this node.
+For reasons of efficiency, A* represents the frontier as a priority queue, with nodes prioritised by the heuristic.
+Selection then reduces to a dequeue operation on the frontier.
+In general the heuristic does not guarantee a uniquely optimal node in the frontier and thus ties may be encountered.
+The question of how these ties are broken in a priority queue is related to the broader notion of sorting stability~\cite{Sedgewick2011}.
+An implementation of a priority queue is said to be stable if it preserves insertion order, otherwise it is said to be unstable.
+If an implementation of A* uses a stable priority queue, then it will behave as a deterministic algorithm which always returns the optimal node that was first added to the frontier.
+Stable priority queues are typically less computationally efficient than unstable priority queues, so the requirement on stability is often dropped in practice.
+
+For example, one of the most common implementations of a priority queue relies on a binary heap and is unstable; it will always return the first optimal node according to its internal heap order rather than insertion order.
+The default priority queue implementations in the standard Java and Boost libraries are based on binary heaps~\cite{boost2021,priorityQueue}, while the Unreal documentation and source code suggests that this is also true of CARLA~\cite{FGraphAStar,TArray,GraphAStar}.
+Thus, an \textbf{unstable but deterministic} priority queue is an unstable priority queue that preserves some order, but not necessarily insertion order.
+If an implementation of A* uses an unstable but deterministic priority queue, then it will behave as a deterministic algorithm which always returns the optimal node according the ordering of its priority queue.
+
+Finally, an unstable but deterministic priority queue suggests a third kind of priority queue that we can refer to as an unstable and non-deterministic priority queue; that is, a priority queue that always returns an optimal node selected at random (i.e.
+breaking ties arbitrarily).
+If an implementation of A* uses an unstable and non-deterministic priority queue, then it will behave as a non-deterministic algorithm.
+Overall, an implementation of A* is much more likely to use a priority queue that is unstable but deterministic rather than one that is stable, and is very unlikely to use one that is unstable and non-deterministic.
+
+An implementation of A* will behave as a deterministic algorithm if it uses a stable priority queue or an unstable but deterministic priority queue.
+However, apparent determinism of A* will be sensitive to the ordering preserved by the priority queue (i.e.
+insertion order or e.g.
+heap order).
+If we are not aware of changes to this ordering, then a deterministic implementation of A* may appear to behave as a non-deterministic algorithm, since we will not have accounted for changes to the ordering.
+For example, suppose actions are iterated over in order of hash value (where order of iteration over actions often determines insertion order), but between runs there is some untracked change to the software/hardware stack that changes the calculation of the hash value, then this may change insertion order of nodes to the frontier, which could eventually materialise as a different optimal solution found by A*.
+This suggests that while a simulated environment may behave deterministically, factors outside the simulator may cause changes to the operation of an A* implementation, which would then materialise as non-deterministic runs in the simulator.
+The difference between the use of a stable priority queue or an unstable but deterministic priority is simply that it may be easier to detect such changes under a stable priority queue because insertion order is typically more meaningful with respect to the actual implementation of A*.
+
+% The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
 A navigation mesh is a fixed area of the environment where actors are free to navigate.
-Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
+Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in then environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
 However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment.
@@
 
-Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
+% Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
 
@@
 To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
+%
+%
+Any implementation of the A* search algorithm for actor navigation must use a stable priority queue to ensure deterministic results. 
+%
 Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer.

### 2021-04-19 document/source/body.tex
simplified

@@
 \subsection{Actor Navigation}
-There is existing evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
+There is evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
 Unreal Engine is the underlying framework for the CARLA simulator which uses the A* search algorithm to find optimal paths for actor navigation~\cite{newton2016unreal}.

### 2021-04-20 document/source/body.tex
acks

--- 2021-04-19 document/source/body.tex
@@
 This requires controlling all potential sources of non-determinism, including randomness and scheduling, very similar to the development of the record-and-reply debugger \texttt{rr}~\cite{RR_link}, originally developed to catch low-frequency non-deterministically failing tests at Mozilla~\cite{acm-q-rr-interview}.
+
+% ***************************************************
+%  Thanks
+% ***************************************************
+\section*{Acknowledgement}
+This research has in part been funded by the ROBOPILOT and CAPRI projects.
+Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
+Special thanks to David May, CFMS, Auroch Digital and Fortelix for their discussions and support on this topic.

### 2021-04-20 document/source/body.tex
moved acks
summary for astar

@@
 The difference between the use of a stable priority queue or an unstable but deterministic priority is simply that it may be easier to detect such changes under a stable priority queue because insertion order is typically more meaningful with respect to the actual implementation of A*.
+
+To summarise, documentation suggests that Carla uses a binary heaps for the priority queue~\cite{FGraphAStar,TArray,GraphAStar} which preserves some order of the frontier nodes, but not necessarily the insertion order, and hence is unstable.
+Therefore, determinism of the A* algorithm for actor navigation will be sensitive to the ordering preserved by the priority queue, in this case the heap order.
+This also assumes the simulation environment itself is deterministic and sources of non-determinism that affect the simulator or the system more generally may cause changes to the operation of an A* implementation.
 
@@
 Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-Special thanks to David May, CFMS, Auroch Digital and Fortelix for their discussions and support on this topic.
+Special thanks to David May, Roger Shepherd, CFMS, Auroch Digital and Fortelix for their discussions and support on this topic.

### 2021-04-20 document/arxiv/det_arxiv.tex
rename top level files, fixed oreilly ref error
acknowledgements
single corresponding author

--- 2021-04-12 document/arxiv/arxiv_template.tex
@@
 
+% for caption at top of table
+\usepackage{float}
+\floatstyle{plaintop}
+\restylefloat{table}
+
 % \usepackage{multicol}
@@
 \maketitle
-\let\thefootnote\relax\footnotetext{\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper and are both corresponding authors.}
+\let\thefootnote\relax\footnotetext{\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper, Greg Chance is the corresponding author.}
 Greg Chance (e-mail: greg.chance@bristol.ac.uk), 
@@
 
-% ***************************************************
-%  Thanks
-% ***************************************************
-\section*{Acknowledgement}
-This research has in part been funded by the ROBOPILOT and CAPRI projects.
-Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-%
-\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper.}
 \balance

### 2021-04-20 document/source/body.tex
A* summary small edits

--- 2021-04-20 document/source/body.tex
@@
 The difference between the use of a stable priority queue or an unstable but deterministic priority is simply that it may be easier to detect such changes under a stable priority queue because insertion order is typically more meaningful with respect to the actual implementation of A*.
+
+To summarise, documentation suggests that CARLA uses a binary heap for the priority queue~\cite{FGraphAStar,TArray,GraphAStar} which preserves some order of the frontier nodes, but not necessarily the insertion order, and hence is unstable.
+Therefore, determinism of the A* algorithm for actor navigation will be sensitive to the ordering preserved by the priority queue, in this case the heap order.
+This also assumes the simulation environment itself is deterministic as sources of non-determinism that affect the simulator or the system more generally may cause changes to the operation of an A* implementation.
 

### 2021-04-21 document/source/body.tex
add rFpro to conclusions
diff pdf

@@
 The default priority queue implementations in the standard Java and Boost libraries are based on binary heaps~\cite{boost2021,priorityQueue}, while the Unreal documentation and source code suggests that this is also true of CARLA~\cite{FGraphAStar,TArray,GraphAStar}.
-Thus, an \textbf{unstable but deterministic} priority queue is an unstable priority queue that preserves some order, but not necessarily insertion order.
+Thus, an unstable but deterministic priority queue is an unstable priority queue that preserves some order, but not necessarily insertion order.
 If an implementation of A* uses an unstable but deterministic priority queue, then it will behave as a deterministic algorithm which always returns the optimal node according the ordering of its priority queue.
@@
     \centering
-    \includegraphics[width=0.99\linewidth]{../other/figures/Methodology_Diagram_v7.pdf}
+    \includegraphics[width=0.99\linewidth]{../other/figures/Methodology_Diagram_v8.pdf}
     \caption{Stages of the method proposed to determine the variance of a simulation.}
@@
 To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
+Alternatively there are commercially available driving simulators that claim to be fully deterministic, e.g.
+RFPro~\cite{rFpro2021}, that may be more suitable if using a game engine does not provide a simulation variance sufficient for your verification requirements.
 
@@
 % ***************************************************
-\section*{Acknowledgement}
+\section{Acknowledgement}
 This research has in part been funded by the ROBOPILOT and CAPRI projects.

### 2021-04-26 document/source/body.tex
slight change to A* in sources and case study

@@
 \subsection{Actor Navigation}
-There is evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
+% There is evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
 Unreal Engine is the underlying framework for the CARLA simulator which uses the A* search algorithm to find optimal paths for actor navigation~\cite{newton2016unreal}.
+%
+In game engines, the A* search algorithm is often used for navigation to find optimal paths for actors. 
+%
 The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.
@@
 In Test 3 a mixture of different actor types is introduced where two vehicles drive without collision and a pedestrian walks across the road at a crossing point.
+
+Similar to vehicles, pedestrian actors navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* search algorithm which is the default method to find optimal paths for the CARLA pedestrian actors~\cite{newton2016unreal}.
 %
-In Test 4 this pedestrian collides with one of the vehicles at the crossing, triggering an \textit{event hit} callback, see Fig~\ref{Test_a}.
-Similar to vehicles, pedestrians navigate via a set of regularly spaced waypoints at $0.1$m intervals using the A* algorithm which is the default controller for the CARLA pedestrian actors.
-Tests 5 \& 6 involve only pedestrians that either do not collide (Test 5) and that do collide (Test 6), see Fig.~\ref{Test_b}.
+There is evidence to suggest that this actor navigation in Carla could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}. 
+%
+This behaviour is explored in Test 4 where a pedestrian collides with one of the vehicles at the crossing, triggering an \textit{event hit} callback, see Fig~\ref{Test_a}.  
+%
+% Unreal Engine is the underlying framework for the CARLA simulator which uses the A* search algorithm to find optimal paths for actor navigation~\cite{newton2016unreal}.
+
+Tests 5 \& 6 involve only pedestrians that either, do not collide (Test 5) and that do collide (Test 6), see Fig.~\ref{Test_b}.
 

### 2021-08-10 rebuttal_letter/rebuttal.tex
rebuttal letter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% LaTeX Template
% Version 2.0 (12/2/17)
%
% This template originates from:
% http://www.LaTeXTemplates.com
%
% Authors:
% Brian Moses
% Vel (vel@LaTeXTemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt, a4paper]{letter} % Set the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc)

\usepackage{graphicx} % Required for including pictures

\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage{gfsdidot} % Use the GFS Didot font: http://www.tug.dk/FontCatalogue/gfsdidot/
\usepackage{microtype} % Improves typography

% \input{structure.tex} % Include the file that specifies the document structure
\pagestyle{empty} % Suppress headers and footers

% \setlength\parindent{1cm} % Paragraph indentation

% Create a new command for the horizontal rule in the document which allows thickness specification
\makeatletter
\newcommand{\vhrulefill}[1]{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
\makeatother

%----------------------------------------------------------------------------------------
%	DOCUMENT MARGINS
%----------------------------------------------------------------------------------------

\usepackage{geometry} % Required for adjusting page dimensions

\geometry{
	top=1cm, % Top margin
	bottom=1.5cm, % Bottom margin
	left=3cm, % Left margin
	right=3cm, % Right margin
	%showframe, % Uncomment to show how the type block is set on the page
}

%----------------------------------------------------------------------------------------
%	DEFINE CUSTOM COMMANDS
%----------------------------------------------------------------------------------------

\newcommand{\logo}[1]{\renewcommand{\logo}{#1}}

\newcommand{\Who}[1]{\renewcommand{\Who}{#1}}
\newcommand{\Title}[1]{\renewcommand{\Title}{#1}}

\newcommand{\headerlineone}[1]{\renewcommand{\headerlineone}{#1}}
\newcommand{\headerlinetwo}[1]{\renewcommand{\headerlinetwo}{#1}}

\newcommand{\authordetails}[1]{\renewcommand{\authordetails}{#1}}

%----------------------------------------------------------------------------------------
%	AUTHOR DETAILS STRUCTURE
%----------------------------------------------------------------------------------------

\newcommand{\authordetailsblock}{
	\hspace{\fill} % Move the author details to the far right
	\parbox[t]{0.48\textwidth}{ % Box holding the author details; width value specifies where it starts and ends, increase to move details left
		\footnotesize % Use a smaller font size for the details
		\Who\\ % Author name
		\textit{\authordetails} % The author details text, all italicised
	}
}

%----------------------------------------------------------------------------------------
%	HEADER STRUCTURE
%----------------------------------------------------------------------------------------

\address{
	\includegraphics[width=2.1in]{\logo} % Include the logo of author institution
	\hspace{0.62\textwidth} % Position of the institution logo, increase to move left, decrease to move right
	\vskip -0.1\textheight~\\ % Position of the large header text in relation to the institution logo, increase to move down, decrease to move up
	% \Large\hspace{0.2\textwidth}\headerlineone\hfill ~\\[0.006\textheight] % First line of institution name, adjust hspace if your logo is wide
	\Large\hspace{0.4\textwidth}\headerlineone\hfill ~\\[0.006\textheight] % First line of institution name, adjust hspace if your logo is wide
	\hspace{0.4\textwidth}\headerlinetwo\hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
	\makebox[0ex][r]{\textbf{\Who\Title}}\hspace{0.01\textwidth} % Print author name and title with a little whitespace to the right
	~\\[-0.01\textheight] % Reduce the whitespace above the horizontal rule
	\hspace{0.4\textwidth}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
	\authordetailsblock % Include the letter author's details on the right side of the page under the horizontal rule
	\hspace{-0.25\textwidth} % Horizontal position of the author details block, increase to move left, decrease to move right
	\vspace{-0.1\textheight} % Move the date and letter content up for a more compact look
}

%----------------------------------------------------------------------------------------
%	COMPOSE THE ENTIRE HEADER
%----------------------------------------------------------------------------------------

\renewcommand{\opening}[1]{
	{\centering\fromaddress\vspace{0.05\textheight} \\ % Print the header and from address here, add whitespace to move date down
	\hspace*{\longindentation}\today\hspace*{\fill}\par} % Print today's date, remove \today to not display it
	{\raggedright \toname \\ \toaddress \par} % Print the to name and address
	\vspace{1cm} % White space after the to address
	\noindent #1 % Print the opening line
}
%----------------------------------------------------------------------------------------
%	SIGNATURE STRUCTURE
%----------------------------------------------------------------------------------------

\signature{\Who\Title} % The signature is a combination of the author's name and title

\renewcommand{\closing}[1]{
	\vspace{2.5mm} % Some whitespace after the letter content and before the signature
	\noindent % Stop paragraph indentation
	\hspace*{\longindentation} % Move the signature right to the value of \longindentation
	% \hspace*{-6.5cm}\includegraphics[width=1in]{signature.png}
	\parbox{\indentedwidth}{
		\raggedright
		#1 % Print the signature text
		\vskip 1.65cm % Whitespace between the closing text and author's name for a physical signature
		\fromsig % Prints the value of \signature{}, i.e. author name and title
	}
}
\longindentation=0pt % Un-commenting this line will push the closing "Sincerely," and date to the left of the page

%----------------------------------------------------------------------------------------
%	YOUR INFORMATION
%----------------------------------------------------------------------------------------

\Who{Dr. Greg Chance} % Your name

\Title{, PhD, CEng, MInstP} % Your title, leave blank for no title

\authordetails{
	Trustworthy Systems Lab\\ % Your department/institution
	University of Bristol\\
	1 Cathedral Square\\ % Your address
	Bristol, BS1 5DD, UK\\ % Your city, zip code, country, etc
	Email: greg.chance@bristol.ac.uk\\ % Your email address
	URL: \url{bristol.ac.uk}\\ % Your URL
	% URL: \url{bristol.ac.uk/engineering/research/trustworthy-systems-laboratory/}\\ % Your URL
	% YouTube: \url{https://youtu.be/1emO3-OYn5k}
}

%----------------------------------------------------------------------------------------
%	HEADER CONTENTS
%----------------------------------------------------------------------------------------

\logo{logo.png} % Logo filename, your logo should have square dimensions (i.e. roughly the same width and height), if it does not, you will need to adjust spacing within the HEADER STRUCTURE block in structure.tex (read the comments carefully!)

\headerlineone{} % Top header line, leave blank if you only want the bottom line
\headerlinetwo{} % Bottom header line

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TO ADDRESS
%----------------------------------------------------------------------------------------

\begin{letter}{
	Prof. Eskandarian\\
	Department Head\\ 
	Nicholas and Rebecca Des Champs Chair\\
	Mechanical Engineering Department, Virginia Tech\\
	635 Prices Fork Road, 449 Goodwin Hall (MC 0238)
	Blacksburg, VA 24061
}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

\opening{}

\textbf{Review of paper T-ITS-21-05-1111}.

Dear Prof Azim Eskandarian,

Thank you for taking the time to review our manuscript. We have given each comment careful consideration between all the authors and we have made changes to the document that make the document clearer and more correct. We give a full account of the changes and rebuttals below and we include a pdf diff to show the changes made.

\textbf{Reviewer 1}\\
\textit{Comment 1: "the whole manuscript seems written from the perspective of Game Engines testing, which lacks practical values in the transportation area."}.
Thank you for this comment, we believe this comment requires a considered response.

Carla is based on a game engine and Carla is currently a popular choice for simulation-based vehicle testing, so this would suggest that the underlying  game engine of Carla is of practical value to the community of intelligent transportation. The paper is not focused on game engines per se, but rather the implications of using those game engines for vehicle testing.

Furthermove, we state on pg.4 the differences between the requirements of game engines for gaming, and for that of AV testing. "Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences. Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision. The gamer neither needs a faithful representation of reality (i.e. gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e. gamer accepts low precision). In contrast, high accuracy and precision are necessary for AV development, testing and verification."

Carla, and other game engine based simulators, will be an entry point for many SME's and start-up companies looking to develop products and services in this area and we believe that this paper brings pertinent information to this community, many of whom may look to ITS for guidance.

Given the confusion to reviewer 1 of our message, we believe that the title of the document should be changed to better reflect the content of the paper, which is primarily concerned with the importance of determinism for simulation based verification of autonomous vehicles. Hence we have changed the paper title to: "On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine"

Thank you and the reviewers for your time and consideration into this paper.

\closing{Sincerely,}

% \begin{figure}
% 	\includegraphics[width=1in]{signature.png}
% \end{figure}

%----------------------------------------------------------------------------------------
%	OPTIONAL FOOTNOTE
%----------------------------------------------------------------------------------------

% Uncomment the 4 lines below to print a footnote with custom text
% \def\thefootnote{}
% \def\footnoterule{\hrule}
% \footnotetext{\hspace*{\fill}{\footnotesize\em Footnote text}}
% \def\thefootnote{\arabic{footnote}}

%----------------------------------------------------------------------------------------

\end{letter}

\end{document}

### 2021-08-10 rebuttal_letter/rebuttal.tex
rebuttal letter

@@
 \textit{Comment 1: "the whole manuscript seems written from the perspective of Game Engines testing, which lacks practical values in the transportation area."}.
-Thank you for this comment, we believe this comment requires a considered response.
 
+Thank you for this comment, we believe this comment requires a considered response to ensure the main message of the paper comes across to the reader.
 Carla is based on a game engine and Carla is currently a popular choice for simulation-based vehicle testing, so this would suggest that the underlying  game engine of Carla is of practical value to the community of intelligent transportation.
@@
 Given the confusion to reviewer 1 of our message, we believe that the title of the document should be changed to better reflect the content of the paper, which is primarily concerned with the importance of determinism for simulation based verification of autonomous vehicles.
-Hence we have changed the paper title to: "On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine"
+Hence we have changed the paper title to: "On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine".
+
+\textit{Comment 2: "One major concern is index selection.
+As we all know, the scale of deviation relies on the mean values of the investigated variables.
+Thus, I wonder why the authors pick maximum deviation to measure the performance of the simulation results.
+In my opinion, the average deviation seems better in measuring the overall performance for the whole simulation process."}.
+
+The issue we found with taking the average deviation over many (100's or 1000's) of repeated runs is that a single 'failure' can be hidden in an average.
+For example 1 error in 1000 would be an insignificant difference to an average, but a single simulation run that exceeds permisible variance could result in a false negative result or even fail to detect the presence of a serious fault with the system under test.
+The verification process needs to be aware of any failure, as even a single failure may coincide with a bug or error in the system that needs to be found and corrected.
+Hence, it is imperative that we discover any simulation that exceeds the simulation variance.
+We could reframe this as thinking less about simulation performance, and more about detecting any errors in the system.
+Verification requires the detection of edge cases (rare cases) not the average performance of the system.
+
+We have added (in italics) to the sentence on page 9:  "Initial testing [26] indicated an actor path deviation of1x10-13cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
+\textit{Due to this nature of fault rates, using an average simulation variance would 'hide' these errors and hence, this is the reason that we use maximum variance and not an average.}".
 

### 2021-08-11 rebuttal_letter/rebuttal.tex
rebuttal letter 2

@@
 
-\opening{}
+% \opening{}
 
@@
 We have added (in italics) to the sentence on page 9:  "Initial testing [26] indicated an actor path deviation of1x10-13cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
-\textit{Due to this nature of fault rates, using an average simulation variance would 'hide' these errors and hence, this is the reason that we use maximum variance and not an average.}".
+\textit{Due to the low probability of fault rates observed, using an average simulation variance would 'hide' these errors and hence this is the reason that we use maximum variance and not an average}".
+
+\textit{Comment 3: "Moreover, I also concern about the results from several tests.
+As stated in Section V-B, 1 cm considered sufficient for urban scenario assertion checking.
+However, the average vehicle speeds are setting as 20–35 km/h (i.e.
+5.56-9.72 m/s) [1].
+Thus, I doubt the practical value of this permissible variance (too accurate for general AV testing) in real-world autonomous vehicle verification.
+The authors should provide more support materials for the setting used in the whole experiment"}.
+
+The focus of this paper is not really the absolute values that we chose for permissible variance or the speed of the vehicles (which was based on UK urban speed limit of 20mph or 30kmph), but more about the method of detecting and controlling simulation variance in an end-user's system.
+The value of permissible variance chosen in the paper (1cm) was chosen as a good point between, say 1m (too coarse) and 1mm (too fine), to illustrate the violation of this limit given certain simulator and system conditions.
+
+The point is that repeated runs should give the exact same output, regardless of permissible variance or simulated vehicle speed.
+The reviewer suggests a paper that descibes the speed of vehicles in an urban environment similar to those in our case study.
+We have included this reference including a line to point interested readers whom wish to set the speed of their simulated vehicles to an apporpriate value.
+
+Please see the additional line on page 7: \textit{This tolerance may need to be chosen for the specific verification case and the speed of the vehicles within the environment [1]}.
+
+\textit{Comment 4: "The whole paper seems much more relevant to the UE4 engine than the CARLA platform.
+I understand the CARLA application is just a case study, but the first half part of the manuscript is less relevant to the AV testing.
+In particular, the authors said “the AV simulation domain introduces its unique challenges that were not considered in that paper.” in section IV"}.
+
+Carla uses Unreal Engine 4 (UE4) for vehicle testing, which is not the original purpose of UE4.
+The issues discussed in the paper do not have any negative implications for games, but (as the paper argues) they do have negative implications for vehicle testing.
+Thus, the issues are issues with Carla rather than UE4, and the same issues are likely to arise in any simulator based on a game engine.
+
+\textit{Comment 5: "The numbers of references are odd"}.
+Please let us know any format changes that are required specifically. If you are referring to the document hyperlinks that link to each reference, these can be switched off if required.
+
+\textit{Comment 6: "In section I, the sentence “without the need for millions of miles of costly on-road testing” is too arbitrary, the on-road testing is still necessary for AVs verification"}.
+We agree with the reviewer that the need for on-road testing is an essential component of the verification process.
+However, the previous sentence highlights the shortcomings of on-road testing where human fatalities have occured due to a lack of exploration of the parameter space.
+The point of this sentence  was in the context of parameter space exploration and the fact that this space could never be explored entirely with on-road testing alone and hence the neccessity of simulation for verification.
+
+\textit{Comment 7: "No need to provide such a detailed description of floating-point arithmetic since it does not cause non-zero simulation variance"}.
+Floating point arithmetic does not contribute to simulation variance, but it is a common misconception that people often jumpt to when considering or discussing deterministic computation.
+We therefore believe it is important to fully explain the issue to ensure that there is no doubt going forward from this point.
+
+To make this clearer we have modified this section to indicate this misconception. We have also made this section more concise. Please see page 4-5.
 

### 2021-08-11 document/ieee_trans/det_ieeetrans.tex
single corresponding author
rebuttal letter 2

@@
 % ***************************************************
-\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
+% \title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
+\title{On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine}
 % \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
@@
 
-\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper and are both corresponding authors.}
+\textit{Greg Chance and Abanoub Ghobrial contributed equally to this paper, Greg Chance is the corresponding author.}
 %
@@
 \input{../source/body.tex}
-
-% ***************************************************
-%  Thanks
-% ***************************************************
-% \section*{Acknowledgement}
-% This research has in part been funded by the ROBOPILOT and CAPRI projects.
-Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
 

### 2021-08-11 document/source/body.tex
rebuttal letter 2

@@
 A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives.
-Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected.
-Thus any variance of less than $1$cm is permissible.
+Therefore, for this case study a tolerance of $1$cm has been selected and thus any variance of less than $1$cm is permissible.
 To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
+This tolerance may need to be chosen for the specific verification case and the speed of the vehicles within the environment~\cite{jiang2014intercultural}.
 
@@
 %
-There is evidence to suggest that this actor navigation in Carla could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}. 
+There is evidence to suggest that this actor navigation in CARLA could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}. 
 %
@@
 Therefore each test was repeated 1000 times to provide a sufficient sample size.
+%
+Due to the low probability of observed fault rates, using an average simulation variance would 'hide' these errors and hence this is the reason that we use maximum variance and not an average. 
+%
 A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.

### 2021-08-11 rebuttal_letter/rebuttal.tex
rebuttal letter 2

@@
 
-To make this clearer we have modified this section to indicate this misconception. We have also made this section more concise. Please see page 4-5.
+To make this clearer we have modified this section to indicate this misconception.
+We have also made this section more concise.
+Please see page 4-5 starting \textit{A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation.
+This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here}.
+
+\textit{Comment 8: "There is no equal sign in Equation (1)."}.
+Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page ...
 

### 2021-08-11 document/source/body.tex
rebuttal letter 2

@@
 
+A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation.
+This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here.
+%
 Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
@@
 %
-Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
+% Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
+Execution order changes or aggressive optimisations~\cite{llvm-floating-point} can occur due to the use or settings of a compiler or as a consequence of parallelisation which can also introduce incorrectness in floating-point arithmetic, e.g.\ within the runtime environment or at the hardware level.
 %
@@
 
-However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
+% However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
 Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
 
-Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
+% Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
 However, the same executable should still return the same output for identical input.

### 2021-08-12 rebuttal_letter/rebuttal.tex
new fig 3

@@
 
-\textit{Comment 8: "There is no equal sign in Equation (1)."}.
-Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page ...
+\textit{Comment 8: "There is no equal sign in Equation (1)"}.
+Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page 8.
+
+\textit{Comment 8: "TABLE I and Fig 3 are not consistent, it would be better if there are three subfigures in Fig 3"}.
+Thank you we agree this would be clearer if there were three figures, this was done to save space in the document. Please see page XX for the new figure.
 

### 2021-08-12 document/source/body.tex
new fig 3

@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{../other/figures/TestCasesDiagramV2_a.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/det_fig3a}
         \caption{}
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{../other/figures/TestCasesDiagramV2_b.pdf}
+        \includegraphics[width=1\textwidth]{../other/figures/det_fig3b}
         \caption{}
@@
     \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{../other/figures/det_fig3c}
+        \caption{}
+        \label{Test_c}
+    \end{subfigure}
+    \caption{Schematic of test scenarios for (a) Tests 1-2, (b) Tests 3-4, (c) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
 \end{figure}
@@
 Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
-The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
+The maximum variance ($\Psi$) over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
 \begin{equation} \label{eq:max_sigma}
-\max_{a,t}\sigma_a^{2}(t)
+\Psi = \max_{a,t}\sigma_a^{2}(t)
 \end{equation}

### 2021-08-17 rebuttal_letter/rebuttal.tex
det comments

@@
 \textbf{Reviewer 1}\\
-\textit{Comment 1: "the whole manuscript seems written from the perspective of Game Engines testing, which lacks practical values in the transportation area."}.
+\textit{Comment 1: "the whole manuscript seems written from the perspective of Game Engines testing, which lacks practical values in the transportation area"}.
 
@@
 Thus, I wonder why the authors pick maximum deviation to measure the performance of the simulation results.
-In my opinion, the average deviation seems better in measuring the overall performance for the whole simulation process."}.
+In my opinion, the average deviation seems better in measuring the overall performance for the whole simulation process"}.
 
@@
 
-\textit{Comment 8: "TABLE I and Fig 3 are not consistent, it would be better if there are three subfigures in Fig 3"}.
+\textit{Comment 9: "TABLE I and Fig 3 are not consistent, it would be better if there are three subfigures in Fig 3"}.
 Thank you we agree this would be clearer if there were three figures, this was done to save space in the document. Please see page XX for the new figure.
+
+\textit{Comment 10: "The size of simulation scenes should be pointed out, especially the boundary of (x, y, z)"}.
+Thank you, we agree this would help describe the scene better to the reader.
+Please see section V.C on page 7 for the additional text: \textit{The map size for the simulation test environment was 354x170$m$.
+The range of movement of the actors within the environment was up to 70m for vehicles and up to 25m for pedestrians}.
+
+\textit{Comment 11: "Which vehicle collided with the pedestrian in test 4, please label them in Fig 3"}. 
+We agree it would be clearer for the reader if this information was presented. We have added the collision area to Fig.3 (on page 7) indicating where the collision happens and the ID of the vehicles.
+
+\textbf{Reviewer 2}\\
+\textit{Comment 1: "The current paper is a little bit too long and could be shortened.
+For instance, several definitions in Section II.A could be introduced in a more concise way.
+Sections II and III could be combined.
+Sections V and VI could be combined"}.
+We agree the paper is a bit long, however if readers are familair with the definitions it is likely that they will skip over these but it would be prudent to keep them in.
+We think that section 5 and 6 are already long independently and combining them might become unweildy.
+
+We have attempted to reduce the paper size in different areas, please see the diff document.
+However, as the first reviewer wanted additional material added this may have resulted in a net gain in overall length.
+
+\textit{Comment 2: "In Section IV, several potential sources of non-determinism are reviewed.
+Through the case study, the authors should have established a general understanding of the relative significance of each potential source to the overall simulation variance (as least for CARLA).
+It would be good to comment on the relative significance alongside the review of these sources"}.
+
+Giving a relative comparison is not easily achieved as each source would be difficult to quantify.
+Hopefully it is clear from Table.1 that pedestrian-vehicle collisions result in high simulation variance in our case study, and also the role that resource utilisation plays should be clear to the reader from veiwing Fig.4.
+However, writing this as a general description would be difficult.
+However, if the reader follows the methodology in section 7 they should be able to discover the principle sources, quantify them and proritise those which may lead to simulation variance above the permissible level for the user.
+
+\textit{Comment 3: "It would be good if the authors could show the vehicle/pedestrian trajectories of a test case where a large maximum variance occur (such as the trajectories pre and post a collision).
+This could help the reader better understand the cause of the large trajectory variance"}.
+Thank you, this is a great idea and we take on board your idea that this would give the reader a better insight into the issue.
+We have added an X-Y position data plot showing high simulation variance on page XX...
 

### 2021-08-23 rebuttal_letter/rebuttal.tex
letter changes

@@
 	Email: greg.chance@bristol.ac.uk\\ % Your email address
-	URL: \url{bristol.ac.uk}\\ % Your URL
+	URL: \url{www.bristol.ac.uk}\\ % Your URL
 	% URL: \url{bristol.ac.uk/engineering/research/trustworthy-systems-laboratory/}\\ % Your URL
@@
 %----------------------------------------------------------------------------------------
+
+\usepackage{etoolbox}
+\AtBeginEnvironment{quote}{\par\singlespacing\small}
+% \usepackage{setspace} % for \onehalfspacing and \singlespacing macros
+% \onehalfspacing
 
@@
 
-% \opening{}
+\opening{}
 
@@
 
-Thank you for taking the time to review our manuscript.
-We have given each comment careful consideration between all the authors and we have made changes to the document that make the document clearer and more correct.
-We give a full account of the changes and rebuttals below and we include a pdf diff to show the changes made.
+Thank you and the reviewers for taking the time to give feedback our manuscript.
+We have given each comment careful consideration between all the authors and we give a full account of the changes and rebuttals below.
+We also include a diff.pdf to show the changes made between the current and originally submitted manuscripts.
+
+\bigskip
+\centerline{\rule{13cm}{0.4pt}}
+\bigskip
 
@@
 
-Thank you for this comment, we believe this comment requires a considered response to ensure the main message of the paper comes across to the reader.
+Thank you for this comment.
+We believe this comment requires a considered response to ensure that the main message of the paper is being properly conveyed to the reader.
 Carla is based on a game engine and Carla is currently a popular choice for simulation-based vehicle testing, so this would suggest that the underlying  game engine of Carla is of practical value to the community of intelligent transportation.
@@
 
-Furthermove, we state on pg.4 the differences between the requirements of game engines for gaming, and for that of AV testing.
-"Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
+Furthermore, we state on pg.4 the differences between the requirements of game engines for gaming, and for that of AV testing:
+
+\begin{quote}
+Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
@@
 gamer accepts low precision).
-In contrast, high accuracy and precision are necessary for AV development, testing and verification."
+In contrast, high accuracy and precision are necessary for AV development, testing and verification.
+\end{quote}
 
@@
 
-Given the confusion to reviewer 1 of our message, we believe that the title of the document should be changed to better reflect the content of the paper, which is primarily concerned with the importance of determinism for simulation based verification of autonomous vehicles.
-Hence we have changed the paper title to: "On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine".
+To make our message clearer, we propose that the title of the document should be changed to better reflect the content, which is primarily concerned with the importance of determinism for simulation based verification of autonomous vehicles.
+Hence we have changed the title to: "On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine".
+
+\bigskip
 
@@
 
-The issue we found with taking the average deviation over many (100's or 1000's) of repeated runs is that a single 'failure' can be hidden in an average.
-For example 1 error in 1000 would be an insignificant difference to an average, but a single simulation run that exceeds permisible variance could result in a false negative result or even fail to detect the presence of a serious fault with the system under test.
-The verification process needs to be aware of any failure, as even a single failure may coincide with a bug or error in the system that needs to be found and corrected.
+The issue we found with taking the average deviation over many (100's or 1000's) of repeated runs is that a single `failure' can be hidden in an average.
+For example 1 error in 1000 would be an insignificant difference to an average, but a single simulation run that exceeds permissible variance could result in a false negative result or even fail to detect the presence of a serious fault with the system under test.
+
+The verification process, therefore, needs to be aware of any non-deterministic event beyond the permissible tolerance, as even a single event may coincide with a bug or error in the system that needs to be found and corrected.
 Hence, it is imperative that we discover any simulation that exceeds the simulation variance.
-We could reframe this as thinking less about simulation performance, and more about detecting any errors in the system.
-Verification requires the detection of edge cases (rare cases) not the average performance of the system.
 
-We have added (in italics) to the sentence on page 9:  "Initial testing [26] indicated an actor path deviation of1x10-13cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
-\textit{Due to the low probability of fault rates observed, using an average simulation variance would 'hide' these errors and hence this is the reason that we use maximum variance and not an average}".
+We could reframe this as thinking less about simulation performance, and more about error detection.
+It is important to know that any failure truly existed and if necessary can be repeated so that the bug can be tracked down and resolved.
+If the simulator is not deterministic then re-running the same test and hoping to observe the same error may not happen.
+
+We have added (in italics) to the sentence on page 9 to clarify this point:  
+\begin{quote} 
+Initial testing [26] indicated an actor path deviation of 1x10-13cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
+\textit{Due to the low probability of a simulation trace exceeding permissible variance, using an average would 'hide' these errors and hence this is the reason that we use maximum variance and not an average}.
+\end{quote}
+
+\bigskip
 
@@
 
-The focus of this paper is not really the absolute values that we chose for permissible variance or the speed of the vehicles (which was based on UK urban speed limit of 20mph or 30kmph), but more about the method of detecting and controlling simulation variance in an end-user's system.
+Thank you we take the point you are making.
+This seems to be less about the results quoted but more about what value of variance is appropriate for the given case study or application.
+The focus of this paper is not really the absolute values that we chose for permissible variance, indeed we state this value should be set by the user in the methodology (see section VII.A.3), or the speed of the vehicles (which was based on UK urban speed limit of 20mph or 30kmph), but more about the method of detecting and controlling simulation variance in an end-user's system.
 The value of permissible variance chosen in the paper (1cm) was chosen as a good point between, say 1m (too coarse) and 1mm (too fine), to illustrate the violation of this limit given certain simulator and system conditions.
@@
 The point is that repeated runs should give the exact same output, regardless of permissible variance or simulated vehicle speed.
-The reviewer suggests a paper that descibes the speed of vehicles in an urban environment similar to those in our case study.
-We have included this reference including a line to point interested readers whom wish to set the speed of their simulated vehicles to an apporpriate value.
+The reviewer suggests a paper that describes the speed of vehicles in an urban environment similar to those in our case study.
+We have included this reference including a line to point interested readers whom wish to set the speed of their simulated vehicles to an appropriate value.
 
-Please see the additional line on page 7: \textit{This tolerance may need to be chosen for the specific verification case and the speed of the vehicles within the environment [1]}.
+Please see the additional line (in italics) add to page 7: 
+\begin{quote}
+To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
+\textit{This tolerance may need to be chosen for the specific verification scenario and the speed of the vehicles within the environment [1]}.
+\end{quote}
+
+\bigskip
 
@@
 
+\bigskip
+
 \textit{Comment 5: "The numbers of references are odd"}.
@@
 
+\bigskip
+
 \textit{Comment 6: "In section I, the sentence “without the need for millions of miles of costly on-road testing” is too arbitrary, the on-road testing is still necessary for AVs verification"}.
 We agree with the reviewer that the need for on-road testing is an essential component of the verification process.
-However, the previous sentence highlights the shortcomings of on-road testing where human fatalities have occured due to a lack of exploration of the parameter space.
-The point of this sentence  was in the context of parameter space exploration and the fact that this space could never be explored entirely with on-road testing alone and hence the neccessity of simulation for verification.
+However, the previous sentence highlights the shortcomings of on-road testing where human fatalities have occurred due to a lack of verification which may result from insufficient exploration of the parameter space.
+
+The point of this sentence  was in the context of parameter space exploration and the fact that this space could never be explored sufficiently with on-road testing alone and hence the necessity of simulation for verification.
+We have modified the sentence to sound less arbitrary, and highlighted the complementary nature of both techniques:
+\begin{quote}
+While on-road testing is an essential and complementary part of AV verification, simulation environments offer a means to explore the vast parameter space safely and efficiently~[32] whilst limiting the scope for costly road trials~[29].
+\end{quote}
+
+\bigskip
 
 \textit{Comment 7: "No need to provide such a detailed description of floating-point arithmetic since it does not cause non-zero simulation variance"}.
-Floating point arithmetic does not contribute to simulation variance, but it is a common misconception that people often jumpt to when considering or discussing deterministic computation.
+Floating point arithmetic does not contribute to simulation variance, but it is a common misconception that people often jump to when considering or discussing deterministic computation.
 We therefore believe it is important to fully explain the issue to ensure that there is no doubt going forward from this point.
 
-To make this clearer we have modified this section to indicate this misconception.
-We have also made this section more concise.
-Please see page 4-5 starting \textit{A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation.
+To make this clearer we have modified this section to indicate this misconception. We have also made this section more concise. Please see page 4-5 starting:
+\begin{quote}
+\textit{A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation.
 This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here}.
+\end{quote}
+
+\bigskip
 
 \textit{Comment 8: "There is no equal sign in Equation (1)"}.
-Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page 8.
+Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page 8:
+
+\begin{equation} \label{eq:max_sigma}
+\Psi = \max_{a,t}\sigma_a^{2}(t)
+\end{equation}
+
+\bigskip
 
 \textit{Comment 9: "TABLE I and Fig 3 are not consistent, it would be better if there are three subfigures in Fig 3"}.
-Thank you we agree this would be clearer if there were three figures, this was done to save space in the document. Please see page XX for the new figure.
+Thank you we agree this would be clearer if there were three figures, this was done to save space in the document. Please see page 7 for the new figure.
+
+\bigskip
 
 \textit{Comment 10: "The size of simulation scenes should be pointed out, especially the boundary of (x, y, z)"}.
-Thank you, we agree this would help describe the scene better to the reader.
-Please see section V.C on page 7 for the additional text: \textit{The map size for the simulation test environment was 354x170$m$.
-The range of movement of the actors within the environment was up to 70m for vehicles and up to 25m for pedestrians}.
+Thank you, we agree this would help describe the scene better to the reader. Please see section V.C on page 7 for the additional text:
+\begin{quote}
+\textit{The map size for the simulation test environment was 354x170$m$. The range of movement of the actors within the environment was up to 70m for vehicles and up to 25m for pedestrians}.
+\end{quote}
+
+\bigskip
 
@@
 We agree it would be clearer for the reader if this information was presented. We have added the collision area to Fig.3 (on page 7) indicating where the collision happens and the ID of the vehicles.
+
+\bigskip
+\centerline{\rule{13cm}{0.4pt}}
+\bigskip
 
@@
 Sections V and VI could be combined"}.
-We agree the paper is a bit long, however if readers are familair with the definitions it is likely that they will skip over these but it would be prudent to keep them in.
-We think that section 5 and 6 are already long independently and combining them might become unweildy.
+We agree the paper is a bit long, however if readers are familiar with the definitions it is likely that they will skip over these but it would be prudent to keep them in.
+We think that section V and IV are already long independently and combining them might become unwieldy.
 
 We have attempted to reduce the paper size in different areas, please see the diff document.
-However, as the first reviewer wanted additional material added this may have resulted in a net gain in overall length.
+However, as the first reviewer wanted additional material added, including figures, this may have resulted in a net increase in overall manuscript length.
+
+\bigskip
 
@@
 Giving a relative comparison is not easily achieved as each source would be difficult to quantify.
-Hopefully it is clear from Table.1 that pedestrian-vehicle collisions result in high simulation variance in our case study, and also the role that resource utilisation plays should be clear to the reader from veiwing Fig.4.
+Hopefully it is clear from Table.1 that pedestrian-vehicle collisions result in high simulation variance in our case study, and also the role that resource utilisation plays should be clear to the reader from viewing Fig.4.
 However, writing this as a general description would be difficult.
-However, if the reader follows the methodology in section 7 they should be able to discover the principle sources, quantify them and proritise those which may lead to simulation variance above the permissible level for the user.
+
+If the reader follows the methodology in section VII they should be able to discover the principle sources, quantify them and prioritise those which may lead to simulation variance above the permissible level for the user.
+
+\bigskip
 
@@
 This could help the reader better understand the cause of the large trajectory variance"}.
-Thank you, this is a great idea and we take on board your idea that this would give the reader a better insight into the issue.
-We have added an X-Y position data plot showing high simulation variance on page XX...
+Thank you, this is a great idea and we agree that this would give the reader a better insight into the issue.
+We have added an X-Y plane data plot showing the high simulation variance that occurs during test 4.
+In this plot you can see the area that vehicle 2 and the pedestrian collide within and the subsequent path variance (see Fig.5 on page 9).
 
-Thank you and the reviewers for your time and consideration into this paper.
+This is referenced in the following text (additional in italics):
+\begin{quote}
+However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as 59cm.
+\textit{A plot of actor position in the X-Y plane (plan view, units m) is shown in Fig.5, where the inset shows the divergence of the path of vehicle 2 post-collision with the pedestrian.} Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.
+\end{quote}
 
-\closing{Sincerely,}
+\bigskip
 
-% \begin{figure}
-% 	\includegraphics[width=1in]{signature.png}
-% \end{figure}
+Thank you and the reviewers for your time and consideration to this paper. We hope given these changes you will consider this manuscript suitable for publishing.
 
-%----------------------------------------------------------------------------------------
-%	OPTIONAL FOOTNOTE
-%----------------------------------------------------------------------------------------
+\bigskip
 
-% Uncomment the 4 lines below to print a footnote with custom text
-% \def\thefootnote{}
-% \def\footnoterule{\hrule}
-% \footnotetext{\hspace*{\fill}{\footnotesize\em Footnote text}}
-% \def\thefootnote{\arabic{footnote}}
+Sincerely,
 
-%----------------------------------------------------------------------------------------
+\bigskip
+
+\includegraphics[width=1in]{signature.png}
+
+Dr. Greg Chance, PhD, CEng, MInstP
 

### 2021-08-23 document/source/body.tex
det comments
letter changes

@@
 There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-Simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
+While on-road testing is an essential and complementary part of AV verification, simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst limiting the need for costly road trials~\cite{kalra2016driving}.
 In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{../other/figures/det_fig3a}
+        \includegraphics[width=1\textwidth]{../other/figures/fig3a}
         \caption{}
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{../other/figures/det_fig3b}
+        \includegraphics[width=1\textwidth]{../other/figures/det_fig3bb}
         \caption{}
@@
 
-The tests used for the case study are listed in Table~\ref{TableOfExperiments}. They cover a range of interactions between actor types. 
+The tests used for the case study are listed in Table~\ref{TableOfExperiments}.
+They cover a range of interactions between actor types.
+The map size for the simulation test environment was 354x170$m$.
+The range of movement of the actors within the environment was up to 70$m$ for vehicles and up to 25$m$ for pedestrians.
 %
@@
 %
-Due to the low probability of observed fault rates, using an average simulation variance would 'hide' these errors and hence this is the reason that we use maximum variance and not an average. 
+Due to the low probability of a simulation trace exceeding permissible variance, using an average would 'hide' these errors and hence this is the reason that we use maximum variance and not an average.
 %
@@
 \section{Results and Discussion}\label{s:FinalResultsSection}
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{../other/figures/actor_path_plot.jpg}
+    \caption{Actor path plot in XY plane for Test 4 with 95\% resource utilisation.}
+    \label{actorPathPlot}
+\end{figure}
 
@@
 However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
+%
+A plot of actor position in the X-Y plane (plan view, units $m$) is shown in Fig.\ref{actorPathPlot}, where the inset clearly shows the divergence of the path of vehicle 2 post-collision with the pedestrian.
+%
 Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.

### 2021-09-02 rebuttal_letter/rebuttal.tex
changes rebuttal question 1

@@
 We believe this comment requires a considered response to ensure that the main message of the paper is being properly conveyed to the reader.
-Carla is based on a game engine and Carla is currently a popular choice for simulation-based vehicle testing, so this would suggest that the underlying  game engine of Carla is of practical value to the community of intelligent transportation.
+Carla is based on a game engine and Carla is currently a popular choice for simulation-based vehicle testing, so this
+is strong evidence that %
+% would suggest that 
+the underlying  game engine of Carla is of practical value to the community of intelligent transportation.
 The paper is not focused on game engines per se, but rather the implications of using those game engines for vehicle testing.
 
-Furthermore, we state on pg.4 the differences between the requirements of game engines for gaming, and for that of AV testing:
+We think there must be a misunderstanding here. We clearly state the importance of needing determinism for verification:
+
+\begin{quote}
+If the simulation is non-deterministic, e.g.
+it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
+In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV’s control software.
+\end{quote}
+
+We state that game engines are the underlying physics framework that support some driving simulators, such as Carla.
+Therefore, if you are using a driving simulator that is based on a game engine, this may lead to non-determinism and compromise your verification analysis.
+We agree with the reviewer that determinism for game engines may not be an important research field by itself, but we go on to state that this is not the case for AV verification.
+We believe this point is sufficiently clear to the reader and hope these examples bring this point to bear:
+
+\begin{quote}
+When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
+However, the situation is very different for AV development and testing.
+Thus, our main research questions are: How can one assess whether a simulation environment is deterministic?
+and How can one determine and control the simulation variance?
+\end{quote}
+
+We also state on pg.4 the differences between the requirements of game engines for gaming, and for that of AV testing.
+We state that there is no need for determinism in gaming (it may even benefit gameplay) but that this is not the case for verification:
 
@@
 
-To make our message clearer, we propose that the title of the document should be changed to better reflect the content, which is primarily concerned with the importance of determinism for simulation based verification of autonomous vehicles.
+% To make our message clearer, we propose that the title of the document should be changed to better reflect the content, which is primarily concerned with the importance of determinism for simulation based verification of autonomous vehicles.
 Hence we have changed the title to: "On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine".

### 2021-09-23 rebuttal_letter/rebuttal.tex
comment 4

@@
 \usepackage{microtype} % Improves typography
+\usepackage{hyperref}
+\hypersetup{hidelinks = true,}
+\usepackage{xurl}
 
@@
 
-We think there must be a misunderstanding here. We clearly state the importance of needing determinism for verification:
+We think that there must just be a misunderstanding here, we clearly state the importance of needing determinism for verification (from page 3):
 
@@
 Therefore, if you are using a driving simulator that is based on a game engine, this may lead to non-determinism and compromise your verification analysis.
-We agree with the reviewer that determinism for game engines may not be an important research field by itself, but we go on to state that this is not the case for AV verification.
-We believe this point is sufficiently clear to the reader and hope these examples bring this point to bear:
+We agree with the reviewer that determinism for game engines may not be an important research field by itself, but we go on to state that this is not the case for AV verification (from page 3):
 
@@
 Carla, and other game engine based simulators, will be an entry point for many SME's and start-up companies looking to develop products and services in this area and we believe that this paper brings pertinent information to this community, many of whom may look to ITS for guidance.
+We believe this point is sufficiently clear to the reader and hope these examples bring this point to bear.
 
@@
 \begin{quote} 
-Initial testing [26] indicated an actor path deviation of 1x10-13cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
-\textit{Due to the low probability of a simulation trace exceeding permissible variance, using an average would 'hide' these errors and hence this is the reason that we use maximum variance and not an average}.
+Initial testing [26] indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
+\textit{Due to the low probability of a simulation trace exceeding permissible variance, using an average would `hide' these errors and hence this is the reason that we use maximum variance and not an average}.
 \end{quote}
@@
 
+There are examples of other simulators that run on game engines.
+AirSim developed by Microsoft for autonomous drone and vehicle development (see footnote\footnote{\url{https://microsoft.github.io/AirSim/}}) is based on the same UE4 game engine as Carla.
+Unreal is also the simulator backbone for the research efforts at the University of Warwick where simulation based safety testing is done using a simulation cave and a real vehicle (see footnote\footnote{\url{https://www.unrealengine.com/en-US/spotlights/meet-the-hybrid-real-time-simulator-for-testing-autonomous-vehicles}}).
+The Unity game engine has also been used for development of the apollo driving simulator (see footnote\footnote{\url{https://apollo.auto/gamesim.html}}) which has partnerships with many vehicle manufacturers (see footnote\footnote{\url{https://apollo.auto/about/ecological\_partner.html#taboem}}).
+
 \bigskip
@@
 \textit{Comment 5: "The numbers of references are odd"}.
+
 Please let us know any format changes that are required specifically. If you are referring to the document hyperlinks that link to each reference, these can be switched off if required.
@@
 \textit{Comment 6: "In section I, the sentence “without the need for millions of miles of costly on-road testing” is too arbitrary, the on-road testing is still necessary for AVs verification"}.
+
 We agree with the reviewer that the need for on-road testing is an essential component of the verification process.
@@
 \textit{Comment 7: "No need to provide such a detailed description of floating-point arithmetic since it does not cause non-zero simulation variance"}.
+
 Floating point arithmetic does not contribute to simulation variance, but it is a common misconception that people often jump to when considering or discussing deterministic computation.
@@
 \textit{Comment 8: "There is no equal sign in Equation (1)"}.
+
 Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page 8:
@@
 \textit{Comment 9: "TABLE I and Fig 3 are not consistent, it would be better if there are three subfigures in Fig 3"}.
+
 Thank you we agree this would be clearer if there were three figures, this was done to save space in the document. Please see page 7 for the new figure.
+
+% \includegraphics[width=1in]{../document/other/figures/fig3a}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}
+        \includegraphics[width=4cm]{../document/other/figures/fig3a}
+    \end{subfigure}
+    \begin{subfigure}
+        \includegraphics[width=4cm]{../document/other/figures/det_fig3bb}
+    \end{subfigure}
+    \begin{subfigure}
+        \includegraphics[width=4cm]{../document/other/figures/det_fig3c}
+    \end{subfigure}
+    \caption{\\Schematic of test scenarios for (a) Tests 1-2, (b) Tests 3-4, (c) Tests 5-6. Descriptions are given in Table~1.}
+\end{figure}
 
@@
 \textit{Comment 10: "The size of simulation scenes should be pointed out, especially the boundary of (x, y, z)"}.
+
 Thank you, we agree this would help describe the scene better to the reader. Please see section V.C on page 7 for the additional text:
 \begin{quote}
-\textit{The map size for the simulation test environment was 354x170$m$. The range of movement of the actors within the environment was up to 70m for vehicles and up to 25m for pedestrians}.
+\textit{The map size for the simulation test environment was 354$m$ $\times$ 170$m$.
+The range of movement of the actors within the environment was up to 70m for vehicles and up to 25m for pedestrians}.
 \end{quote}
@@
 \textit{Comment 11: "Which vehicle collided with the pedestrian in test 4, please label them in Fig 3"}.
+
 We agree it would be clearer for the reader if this information was presented. We have added the collision area to Fig.3 (on page 7) indicating where the collision happens and the ID of the vehicles.
@@
 Sections V and VI could be combined"}.
+
 We agree the paper is a bit long, however if readers are familiar with the definitions it is likely that they will skip over these but it would be prudent to keep them in.
-We think that section V and IV are already long independently and combining them might become unwieldy.
+We think that section V and VI are already long independently and combining them might become unwieldy.
 
@@
 This could help the reader better understand the cause of the large trajectory variance"}.
+
 Thank you, this is a great idea and we agree that this would give the reader a better insight into the issue.
@@
 
-This is referenced in the following text (additional in italics):
+\begin{figure}
+    \centering
+    \includegraphics[width=12cm]{../document/other/figures/actor_path_plot.jpg}
+    \caption{\\Actor path plot in XY plane for Test 4 with 95\% resource utilisation.}
+\end{figure}
+
+This new figure is referenced in the following text (additional in italics):
 \begin{quote}

### 2021-09-23 rebuttal_letter/rebuttal.tex
comment 6

@@
 \begin{quote}
-While on-road testing is an essential and complementary part of AV verification, simulation environments offer a means to explore the vast parameter space safely and efficiently~[32] whilst limiting the scope for costly road trials~[29].
+While on-road testing is an essential part of AV verification, it can be complementary to simulation, offering a means to explore the vast parameter space safely and efficiently~[32] whilst reducing the amount of costly road trials~[29].
 \end{quote}

### 2021-09-23 document/source/body.tex
comment 6

@@
 There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-While on-road testing is an essential and complementary part of AV verification, simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst limiting the need for costly road trials~\cite{kalra2016driving}.
+
+While on-road testing is an essential part of AV verification, it can be complementary to simulation, offering a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst reducing the amount of costly road trials~\cite{kalra2016driving}.
+%
+%While on-road testing is an essential and complementary part of AV verification, simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst limiting the need for costly road trials~\cite{kalra2016driving}.
 In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
@@
 The gamer neither needs a faithful representation of reality (i.e.
-gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
+gamer accepts low accuracy) nor requires repeated actions to result in the same outcome (i.e.
 gamer accepts low precision).
@@
 They cover a range of interactions between actor types.
-The map size for the simulation test environment was 354x170$m$.
+The map size for the simulation test environment was 354$m$ $\times$ 170$m$.
 The range of movement of the actors within the environment was up to 70$m$ for vehicles and up to 25$m$ for pedestrians.
@@
 %
-Due to the low probability of a simulation trace exceeding permissible variance, using an average would 'hide' these errors and hence this is the reason that we use maximum variance and not an average.
+Due to the low probability of a simulation trace exceeding permissible variance, using an average would `hide' these errors and hence this is the reason that we use maximum variance and not an average.
 %

### 2021-09-23 rebuttal_letter/rebuttal.tex
reviewed and revised version of rebuttal letter with some TODO notes

@@
 \usepackage{xurl}
+\usepackage{todonotes}
 
@@
 
-Thank you and the reviewers for taking the time to give feedback our manuscript.
+Thank you and the reviewers for taking the time to give feedback on our manuscript.
 We have given each comment careful consideration between all the authors and we give a full account of the changes and rebuttals below.
@@
 We believe this comment requires a considered response to ensure that the main message of the paper is being properly conveyed to the reader.
-Carla is based on a game engine and Carla is currently a popular choice for simulation-based vehicle testing, so this
+Carla is based on a game engine and Carla is currently a popular choice for simulation-based vehicle testing.
+This
 is strong evidence that %
@@
 the underlying  game engine of Carla is of practical value to the community of intelligent transportation.
-The paper is not focused on game engines per se, but rather the implications of using those game engines for vehicle testing.
+The paper is not focused on game engines per se, but rather addresses the implications of using those game engines for vehicle testing.
 
@@
 
-We state that game engines are the underlying physics framework that support some driving simulators, such as Carla.
+We state that game engines provide the underlying physics framework that supports some driving simulators, such as Carla.
 Therefore, if you are using a driving simulator that is based on a game engine, this may lead to non-determinism and compromise your verification analysis.
@@
 
-We also state on pg.4 the differences between the requirements of game engines for gaming, and for that of AV testing.
+We also state on page 4 the differences between the requirements of game engines for gaming and for AV testing.
 We state that there is no need for determinism in gaming (it may even benefit gameplay) but that this is not the case for verification:
@@
 Carla, and other game engine based simulators, will be an entry point for many SME's and start-up companies looking to develop products and services in this area and we believe that this paper brings pertinent information to this community, many of whom may look to ITS for guidance.
-We believe this point is sufficiently clear to the reader and hope these examples bring this point to bear.
+We believe this point should be sufficiently clear to the attentive reader and hope these examples bring this point to bear.
 
@@
 
-The issue we found with taking the average deviation over many (100's or 1000's) of repeated runs is that a single `failure' can be hidden in an average.
+The issue we found with taking the average deviation over many (100's or 1000's of) repeated runs is that a single `failure' can be hidden in an average.
 For example 1 error in 1000 would be an insignificant difference to an average, but a single simulation run that exceeds permissible variance could result in a false negative result or even fail to detect the presence of a serious fault with the system under test.
 
-The verification process, therefore, needs to be aware of any non-deterministic event beyond the permissible tolerance, as even a single event may coincide with a bug or error in the system that needs to be found and corrected.
-Hence, it is imperative that we discover any simulation that exceeds the simulation variance.
+The verification process, therefore, needs to be aware of any deviation beyond the permissible tolerance, as even a single event may coincide with a bug or error in the system that needs to be found and corrected.
+Hence, it is imperative that we discover any simulation run that exceeds the permissible tolerance.
 
@@
 It is important to know that any failure truly existed and if necessary can be repeated so that the bug can be tracked down and resolved.
-If the simulator is not deterministic then re-running the same test and hoping to observe the same error may not happen.
+If the simulator is not deterministic, then by re-running the same test we may not be able to observe the same behaviour, i.e.\ we may not be able to recreate a previously seen error.
 
@@
 Initial testing [26] indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
-\textit{Due to the low probability of a simulation trace exceeding permissible variance, using an average would `hide' these errors and hence this is the reason that we use maximum variance and not an average}.
+\textit{Due to the low probability of a simulation trace exceeding the permissible variance, using an average would `hide' these errors; hence this is the reason for using maximum variance and not an average}.
 \end{quote}
@@
 This seems to be less about the results quoted but more about what value of variance is appropriate for the given case study or application.
-The focus of this paper is not really the absolute values that we chose for permissible variance, indeed we state this value should be set by the user in the methodology (see section VII.A.3), or the speed of the vehicles (which was based on UK urban speed limit of 20mph or 30kmph), but more about the method of detecting and controlling simulation variance in an end-user's system.
-The value of permissible variance chosen in the paper (1cm) was chosen as a good point between, say 1m (too coarse) and 1mm (too fine), to illustrate the violation of this limit given certain simulator and system conditions.
+The focus of this paper is not really on the absolute values that we chose for permissible variance, indeed we state this value should be set by the user in the methodology (see section VII.A.3), or the speed of the vehicles (which was based on the UK urban speed limit of 20mph or 30kmph), but more about the method of detecting and controlling simulation variance in an end-user's system.
+The value of permissible variance chosen in the paper (1cm) was considered an appropriate compromise between, say 1m (too coarse) and 1mm (too fine), to illustrate the violation of this limit given certain simulator and system conditions.
 
@@
 The reviewer suggests a paper that describes the speed of vehicles in an urban environment similar to those in our case study.
-We have included this reference including a line to point interested readers whom wish to set the speed of their simulated vehicles to an appropriate value.
+We have included this reference together with a line to point interested readers who wish to set the speed of their simulated vehicles to an appropriate value.
 
-Please see the additional line (in italics) add to page 7: 
+Please see the additional line (in italics) added to page 7: 
 \begin{quote}
@@
 Unreal is also the simulator backbone for the research efforts at the University of Warwick where simulation based safety testing is done using a simulation cave and a real vehicle (see footnote\footnote{\url{https://www.unrealengine.com/en-US/spotlights/meet-the-hybrid-real-time-simulator-for-testing-autonomous-vehicles}}).
-The Unity game engine has also been used for development of the apollo driving simulator (see footnote\footnote{\url{https://apollo.auto/gamesim.html}}) which has partnerships with many vehicle manufacturers (see footnote\footnote{\url{https://apollo.auto/about/ecological\_partner.html#taboem}}).
+The Unity game engine has also been used for development of the apollo driving simulator (see footnote\footnote{\url{https://apollo.auto/gamesim.html}}) which has partnerships with many vehicle manufacturers (see footnote\footnote{\url{https://apollo.auto/about/ecological\_partner.html#taboem}}).\todo{This needs a concluding sentence that brings it back to the comment.
+;)}
 
@@
 
-The point of this sentence  was in the context of parameter space exploration and the fact that this space could never be explored sufficiently with on-road testing alone and hence the necessity of simulation for verification.
-We have modified the sentence to sound less arbitrary, and highlighted the complementary nature of both techniques:
+The point of this sentence  was in the context of parameter space exploration and the fact that this space could never be explored sufficiently with on-road testing alone and hence the necessity of simulation-based verification.
+We have modified the sentence to clarify our intent, and highlighted the complementary nature of both techniques:
 \begin{quote}
-While on-road testing is an essential part of AV verification, it can be complementary to simulation, offering a means to explore the vast parameter space safely and efficiently~[32] whilst reducing the amount of costly road trials~[29].
+While on-road testing is an essential part of AV verification, it can be complemented by simulation-based testing, which offers a means to explore the vast parameter space safely and efficiently~[32] whilst reducing the amount of costly road trials~[29].
 \end{quote}
@@
 Floating point arithmetic does not contribute to simulation variance, but it is a common misconception that people often jump to when considering or discussing deterministic computation.
-We therefore believe it is important to fully explain the issue to ensure that there is no doubt going forward from this point.
+We therefore believe it is critically important to fully explain the issue to ensure that there is no doubt going forward from this point.
 
@@
 
-Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page 8:
+Thank you, we have created a symbol for this to turn this into an equation. Please see the amended equation on page 8:
 
@@
 
-Thank you we agree this would be clearer if there were three figures, this was done to save space in the document. Please see page 7 for the new figure.
+Thank you, we agree this would be clearer if there were three figures. Please see page 7 for the new figure.
 
@@
 
-We agree the paper is a bit long, however if readers are familiar with the definitions it is likely that they will skip over these but it would be prudent to keep them in.
+We agree the paper is a bit long, however if readers are familiar with the definitions it is likely that they will skip over these but it would be prudent to keep them in for the paper to be sufficiently self-contained.
 We think that section V and VI are already long independently and combining them might become unwieldy.
@@
 
-Giving a relative comparison is not easily achieved as each source would be difficult to quantify.
-Hopefully it is clear from Table.1 that pedestrian-vehicle collisions result in high simulation variance in our case study, and also the role that resource utilisation plays should be clear to the reader from viewing Fig.4.
-However, writing this as a general description would be difficult.
+We do not consider it feasible to provide a general understanding for the relative significance of each potential source of non-determinism due to the fact that these depend on the verification requirements, the simulation setup, the actual software/hardware configuration used, etc.
+These factors have been identified in our methodology in section VII.
+If the reader follows this methodology, then they should be able to discover the principle sources, quantify them and prioritise those which may lead to simulation variance above the permissible level for their use case.
 
-If the reader follows the methodology in section VII they should be able to discover the principle sources, quantify them and prioritise those which may lead to simulation variance above the permissible level for the user.
+We have illustrated in Table.1 that pedestrian-vehicle collisions result in high simulation variance for our case study, and also the role that resource utilisation plays should be clear to the reader from viewing Fig.4.
+\todo{Check this, please.
+Stating that something is ``difficult'' is not a good nor a scientifically justifiable reason to dismiss a suggestion.
+;)}
 
@@
 
-This new figure is referenced in the following text (additional in italics):
+This new figure is referenced in the following text (additional in italics):\todo{Ensure that the x and y axis in the graph show with units.}
 \begin{quote}
@@
 
-Thank you and the reviewers for your time and consideration to this paper. We hope given these changes you will consider this manuscript suitable for publishing.
+Thank you and the reviewers for your time and consideration to this paper. We hope, given the above changes and explanations, you will consider this manuscript suitable for publishing.
 

### 2021-09-23 document/source/body.tex
adjusted text to match rebuttal letter

@@
 
-While on-road testing is an essential part of AV verification, it can be complementary to simulation, offering a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst reducing the amount of costly road trials~\cite{kalra2016driving}.
+While on-road testing is an essential part of AV verification, it can be complemented by  simulation-based testing, which offers a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst reducing the amount of costly road trials~\cite{kalra2016driving}.
 %
@@
 %
-Due to the low probability of a simulation trace exceeding permissible variance, using an average would `hide' these errors and hence this is the reason that we use maximum variance and not an average.
+Due to the low probability of a simulation trace exceeding the permissible variance, using an average would `hide' these errors; hence this is the reason for using maximum variance and not an average.
 %

### 2021-09-24 rebuttal_letter/rebuttal.tex
rebuttal KE comments complete

@@
 There are examples of other simulators that run on game engines.
-AirSim developed by Microsoft for autonomous drone and vehicle development (see footnote\footnote{\url{https://microsoft.github.io/AirSim/}}) is based on the same UE4 game engine as Carla.
+For example, AirSim developed by Microsoft for autonomous drone and vehicle development (see footnote\footnote{\url{https://microsoft.github.io/AirSim/}}) is based on the same UE4 game engine as Carla.
 Unreal is also the simulator backbone for the research efforts at the University of Warwick where simulation based safety testing is done using a simulation cave and a real vehicle (see footnote\footnote{\url{https://www.unrealengine.com/en-US/spotlights/meet-the-hybrid-real-time-simulator-for-testing-autonomous-vehicles}}).
-The Unity game engine has also been used for development of the apollo driving simulator (see footnote\footnote{\url{https://apollo.auto/gamesim.html}}) which has partnerships with many vehicle manufacturers (see footnote\footnote{\url{https://apollo.auto/about/ecological\_partner.html#taboem}}).\todo{This needs a concluding sentence that brings it back to the comment.
-;)}
+Unity is another popular game engine and has been used for development of the Apollo driving simulator (see footnote\footnote{\url{https://apollo.auto/gamesim.html}}) which has partnerships with many vehicle manufacturers and development teams (see footnote\footnote{\url{https://apollo.auto/about/ecological\_partner.html#taboem}}).
+So from this, we can say that the use of UE4 and, more broadly, game engines are actively being used in the research and development of autonomous vehicles and is therefore relevant to AV testing and verification.
 
@@
 
+\newpage
 \textit{Comment 5: "The numbers of references are odd"}.
@@
 We have illustrated in Table.1 that pedestrian-vehicle collisions result in high simulation variance for our case study, and also the role that resource utilisation plays should be clear to the reader from viewing Fig.4.
-\todo{Check this, please.
-Stating that something is ``difficult'' is not a good nor a scientifically justifiable reason to dismiss a suggestion.
-;)}
 
@@
 
-This new figure is referenced in the following text (additional in italics):\todo{Ensure that the x and y axis in the graph show with units.}
+This new figure is referenced in the following text (additional in italics):
 \begin{quote}

### 2021-10-17 document/ieee_trans/det_ieeetrans.tex
changes rebuttal question 1
KIE review but not yet finished

@@
 % ***************************************************
-% \title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-\title{On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine}
+\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
+% \title{On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine}
 % \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
@@
 \begin{IEEEkeywords}
-Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
+Autonomous Driving, Autonomous Vehicles, Determinism, Game Engines, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
 \end{IEEEkeywords}
@@
 His research interest primarily concerns the socio-cognitive aspects of human-robot interaction and he has been focusing his recent experimental work on child-robot interaction and human-in-the-loop machine learning for social robots.
-Pr.
-S\'everin Lemaignan has been involved in several European projects related to social and cognitive robotics (CHRIS, DREAM, L2TOR) and was  awarded in 2015 a EU H2020 Marie Sk\l{}odowska-Curie Individual Fellowship for his project on robots and theory of mind.
+Pr.\ S\'everin Lemaignan has been involved in several European projects related to social and cognitive robotics (CHRIS, DREAM, L2TOR) and was  awarded in 2015 a EU H2020 Marie Sk\l{}odowska-Curie Individual Fellowship for his project on robots and theory of mind.
 \end{IEEEbiography}
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/kie}}]{{Kerstin Eder}} 
-is Professor of Computer Science and heads the Trustworthy Systems Laboratory at the University of Bristol.
+is Professor of Computer Science and heads the Trustworthy Systems Laboratory at the University of Bristol, UK.
 She also leads the Verification and Validation for Safety in Robots research theme at the  Bristol Robotics Laboratory.
@@
 Her research is focused on specification, verification and analysis techniques to verify or explore a system's behaviour in terms of functional correctness, safety, security, performance and energy efficiency.
-Prof.\ Eder's most recent contributions include agent-based testing, how to use assertions and theorem proving to verify control system designs, energy modeling of software and static  analysis to predict energy consumption of programs.
+Prof.\ Eder's most recent contributions include agent-based testing, how to use assertions and theorem proving to verify control system designs, energy modeling and static resource analysis techniques to predict energy consumption of software.
 She holds a PhD in Computational Logic and an MSc in Artificial Intelligence from the University of Bristol, UK, as well as an MEng in Informatics (Dipl.\,Inf.) from the Technical University Dresden, Germany.
-In 2007 she was awarded an ``Excellence in Engineering'' prize from the Royal Academy of Engineering.
+In 2007 she was awarded an ``Excellence in Engineering'' prize from the Royal Academy of Engineering, UK.
 

### 2021-10-17 document/source/abstract.tex
KIE review but not yet finished

@@
-Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments.
+Game engines are increasingly used as simulation platforms by the autonomous vehicle community to develop vehicle control systems and test environments.
 A key requirement for simulation-based development and verification is determinism, since a deterministic process will always produce the same output given the same initial conditions and event history.

### 2021-10-17 document/source/body.tex
rebuttal KE comments complete
KIE review but not yet finished

@@
 
-In this paper, we investigate non-determinism and how it affects simulation results using the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
+In this paper, we investigate non-determinism and how it affects simulation results using the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal Engine.
 %
 In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
-We find that the CARLA simulator is non-deterministic.
-Actor path variance was found to be non-zero and, under certain conditions, the deviation from the mean was observed up to $59$cm.
-We would consider a deviation of $1$cm to be permissible for autonomous vehicle verification tasks in an urban environment.
-However, we found that CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
 %
+We establish that the CARLA simulator is non-deterministic.  Actor path variance was found to be non-zero and, under certain conditions, a deviation from the mean was observed of up to $59$cm.
+%
+In such an urban environment, we consider a deviation of up to $1$cm to be permissible for AV verification.
+%
+However, in our experiments, CARLA only exhibits this permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected.
 
@@
 %
-An empirical case study of simulation variance for a number of scenarios involving pedestrian and vehicles is given in Section~\ref{s:case-study} including internal and external setting and system screening tests.
+An empirical case study of simulation variance for a number of scenarios involving pedestrians and vehicles is given in Section~\ref{s:case-study} including internal and external setting and system screening tests.
 %
@@
 \subsubsection{Determinism}
-Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
-However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
+Schumann et al.\ describe determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.\TODO{KIE: I can't find that quote in that paper, please point me to it.} However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
 Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
@@
 In the case of a driving simulator, kinematics will describe future states of actors given the current conditions and actions taken, thereby generating new data.
-If a simulation is deterministic, Fig.~\ref{variance_description}~(b), then there will be no variation in the generated output data, i.e.
+If a simulation is deterministic, Fig.~\ref{variance_description_b}, then there will be no variation in the generated output data, i.e.
 all future states are perfectly reproducible from prior states and actions.
-However, if a simulation is non-deterministic, Fig.~\ref{variance_description}~(a), then there will be a variation in the generated output data.
+However, if a simulation is non-deterministic, Fig.~\ref{variance_description_a}, then there will be a variation in the generated output data.
 \\
@@
     \centering
+
     \begin{subfigure}{.48\textwidth}
         \includegraphics[width=1\textwidth]{../other/figures/Variance_predicition_tolerance_definition_diagram_a.pdf}
-        \caption{Non-deterministic}
+        \caption{Non-determinism}
         \label{variance_description_a}
@@
         \includegraphics[width=1\textwidth]{../other/figures/Variance_predicition_tolerance_definition_diagram_b.pdf}
-        \caption{Deterministic}
+        \caption{Determinism}
         \label{variance_description_b}
     \end{subfigure}
-    \caption{Demonstration of variance, precision, tolerance and determinism}
+
+    \caption{Illustration of accuracy, variance, precision and tolerance in the context of non-determinism and determinism}
     \label{variance_description}
@@
 Those differences describe the spread or variance in the process output.
-A hard limit on the variance can then be defined, Fig.~\ref{variance_description}~(a), beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
+A hard limit on the variance can then be defined, Fig.~\ref{variance_description_a}, beyond which the output fails to meet the required tolerance, e.g.\ the output is rejected by quality control.
 Real manufacturing fails to achieve absolute precision.
@@
 
-If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description}~(b), and hence will be within tolerance by design.
+If a simulator is deterministic then it will produce results with absolute precision or zero variance, Fig.~\ref{variance_description_b}, and hence will be within tolerance by design.
 If the simulator is non-deterministic then there will be a measurable, non-zero variance in the output data.\\
@@
 
-Considering the objectives for gaming and comparing them to these for AV development and testing, there are fundamental differences.
+Considering the objectives for gaming and comparing them to those for AV development and testing, there are fundamental differences.
 Providing game players with a responsive real-time experience is often achieved at the cost of simulation accuracy and precision.
@@
 
-A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation.
-This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here.
+A common misconception is to attribute non-determinism to the use of floating point number representation. This is an erroneous conflation with non-deterministic computational execution.  
 %
@@
 %
-% Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
-Execution order changes or aggressive optimisations~\cite{llvm-floating-point} can occur due to the use or settings of a compiler or as a consequence of parallelisation which can also introduce incorrectness in floating-point arithmetic, e.g.\ within the runtime environment or at the hardware level.
-%
-Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
-
 % Translating to the usage of AV simulations in physics engines,
@@
 However, the same executable should still return the same output for identical input.
+
+% Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
+However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, and 
+parallelisation within the runtime environment or at the hardware level can change the execution order, which can cause simulation results to differ between runs when floating-point operations are affected.
+%
+Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
 
@@
 
+\section{Results and Discussion}\label{s:FinalResultsSection}
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.48\textwidth]{../other/figures/actor_path_plot.jpg}
+    \caption{Actor path plot in XY plane for Test 4 with 95\% resource utilisation.}
+    \label{actorPathPlot}
+\end{figure}
+
 \begin{figure}[t]
@@
 % ***************************************************
-\section{Results and Discussion}\label{s:FinalResultsSection}
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{../other/figures/actor_path_plot.jpg}
-    \caption{Actor path plot in XY plane for Test 4 with 95\% resource utilisation.}
-    \label{actorPathPlot}
-\end{figure}
 

### 2021-10-20 rebuttal_letter/rebuttal.tex
minor changes to rebuttal letter

@@
 \usepackage{microtype} % Improves typography
+
 \usepackage{hyperref}
@@
 \begin{quote} 
-Initial testing [26] indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10cm.
+Initial testing [26] indicated an actor path deviation of $1\times10^{-13}$cm for 997 out of 1000 tests, with three tests reporting a deviation of over 10 $cm$.
 \textit{Due to the low probability of a simulation trace exceeding the permissible variance, using an average would `hide' these errors; hence this is the reason for using maximum variance and not an average}.
@@
 
-Thank you we take the point you are making.
+Thank you we understand the point you are making.
 This seems to be less about the results quoted but more about what value of variance is appropriate for the given case study or application.
-The focus of this paper is not really on the absolute values that we chose for permissible variance, indeed we state this value should be set by the user in the methodology (see section VII.A.3), or the speed of the vehicles (which was based on the UK urban speed limit of 20mph or 30kmph), but more about the method of detecting and controlling simulation variance in an end-user's system.
-The value of permissible variance chosen in the paper (1cm) was considered an appropriate compromise between, say 1m (too coarse) and 1mm (too fine), to illustrate the violation of this limit given certain simulator and system conditions.
+The focus of this paper is not really on the absolute values that we chose for permissible variance, indeed we state this value should be set by the user in the methodology (see section VII.A.3), or the speed of the vehicles (which was based on the UK urban speed limit of 20 $mph$ or 30 $kmph$), but more about the method of detecting and controlling simulation variance in an end-user's system.
+The value of permissible variance chosen in the paper (1 $cm$) was considered an appropriate compromise between, say 1 $m$ (too coarse) and 1 $mm$ (too fine), to illustrate the violation of this limit given certain simulator and system conditions.
 
-The point is that repeated runs should give the exact same output, regardless of permissible variance or simulated vehicle speed.
+The point is that repeated runs should give the exact same output, regardless of the designated permissible variance or simulated vehicle speed.
 The reviewer suggests a paper that describes the speed of vehicles in an urban environment similar to those in our case study.

### 2021-10-20 document/source/first_submit_body.tex
copy of first submitted body source

--- 2021-09-23 document/source/body.tex
@@
 There have been several fatalities with AVs, some of which could be attributed to insufficient verification and validation (V\&V), e.g.~\cite{FatalityExample}.
-
-While on-road testing is an essential part of AV verification, it can be complementary to simulation, offering a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst reducing the amount of costly road trials~\cite{kalra2016driving}.
-%
-%While on-road testing is an essential and complementary part of AV verification, simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} whilst limiting the need for costly road trials~\cite{kalra2016driving}.
+Simulation environments offer a means to explore the vast parameter space safely and efficiently~\cite{korosec2019waymo} without the need for millions of miles of costly on-road testing~\cite{kalra2016driving}.
 In particular, simulations can be biased to increase the frequency at which otherwise rare events occur~\cite{Koopman2018}; this includes testing how the AV reacts to unexpected behaviour of the environment~\cite{RobustnessAutonomy}.
@@
 The gamer neither needs a faithful representation of reality (i.e.
-gamer accepts low accuracy) nor requires repeated actions to result in the same outcome (i.e.
+gamer accepts low accuracy) nor require repeated actions to result in the same outcome (i.e.
 gamer accepts low precision).
@@
 
-A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation.
-This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here.
-%
 Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
@@
 %
-% Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
-Execution order changes or aggressive optimisations~\cite{llvm-floating-point} can occur due to the use or settings of a compiler or as a consequence of parallelisation which can also introduce incorrectness in floating-point arithmetic, e.g.\ within the runtime environment or at the hardware level.
+Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
 %
@@
 
-% However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
+However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
 Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
 
-% Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
+Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
 However, the same executable should still return the same output for identical input.
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{../other/figures/fig3a}
+        \includegraphics[width=1\textwidth]{../other/figures/TestCasesDiagramV2_a.pdf}
         \caption{}
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{../other/figures/det_fig3bb}
+        \includegraphics[width=1\textwidth]{../other/figures/TestCasesDiagramV2_b.pdf}
         \caption{}
@@
     \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{../other/figures/det_fig3c}
-        \caption{}
-        \label{Test_c}
-    \end{subfigure}
-    \caption{Schematic of test scenarios for (a) Tests 1-2, (b) Tests 3-4, (c) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
+    \caption{Schematic of test scenarios for (a) Tests 1-4, (b) Tests 5-6. Descriptions are given in Table~\ref{TableOfExperiments}.}\label{f:test_a_and_b}
 \end{figure}
@@
 A very small value, e.g.\ $1\times10^{-12}m$, may be overly-sensitive to minor computational perturbations and generate false positives.
-Therefore, for this case study a tolerance of $1$cm has been selected and thus any variance of less than $1$cm is permissible.
+Therefore, for this case study and, more broadly, for our general verification requirements, a tolerance of $1$cm has been selected.
+Thus any variance of less than $1$cm is permissible.
 To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-This tolerance may need to be chosen for the specific verification case and the speed of the vehicles within the environment~\cite{jiang2014intercultural}.
 
@@
 
-The tests used for the case study are listed in Table~\ref{TableOfExperiments}.
-They cover a range of interactions between actor types.
-The map size for the simulation test environment was 354$m$ $\times$ 170$m$.
-The range of movement of the actors within the environment was up to 70$m$ for vehicles and up to 25$m$ for pedestrians.
+The tests used for the case study are listed in Table~\ref{TableOfExperiments}. They cover a range of interactions between actor types. 
 %
@@
 Instead of using variance, herein the results are given in terms of the deviation, $\sigma_a(t)$, which indicates the dispersion of the actor path relative to the mean and is helpfully in the same units as actor position, i.e.\ metres ($m$), for ease of interpretation.
-The maximum variance ($\Psi$) over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
+The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}.
 \begin{equation} \label{eq:max_sigma}
-\Psi = \max_{a,t}\sigma_a^{2}(t)
+\max_{a,t}\sigma_a^{2}(t)
 \end{equation}
@@
 Therefore each test was repeated 1000 times to provide a sufficient sample size.
-%
-Due to the low probability of a simulation trace exceeding permissible variance, using an average would `hide' these errors and hence this is the reason that we use maximum variance and not an average.
-%
 A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
@@
 \section{Results and Discussion}\label{s:FinalResultsSection}
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.48\textwidth]{../other/figures/actor_path_plot.jpg}
-    \caption{Actor path plot in XY plane for Test 4 with 95\% resource utilisation.}
-    \label{actorPathPlot}
-\end{figure}
 
@@
 However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
-%
-A plot of actor position in the X-Y plane (plan view, units $m$) is shown in Fig.\ref{actorPathPlot}, where the inset clearly shows the divergence of the path of vehicle 2 post-collision with the pedestrian.
-%
 Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.

### 2021-10-28 rebuttal_letter/rebuttal.tex
update letter with grammar correction question 4

@@
 Unity is another popular game engine and has been used for development of the Apollo driving simulator (see footnote\footnote{\url{https://apollo.auto/gamesim.html}}) which has partnerships with many vehicle manufacturers and development teams (see footnote\footnote{\url{https://apollo.auto/about/ecological\_partner.html#taboem}}).
-So from this, we can say that the use of UE4 and, more broadly, game engines are actively being used in the research and development of autonomous vehicles and is therefore relevant to AV testing and verification.
+So from this, we can say that the use of UE4 and, more broadly, game engines are actively being used in the research and development of autonomous vehicles and are therefore relevant to AV testing and verification.
 
@@
 \begin{quote}
-\textit{A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation.
-This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here}.
+\textit{A common misconception is to attribute non-determinism to the use of floating point number representation. This is an erroneous conflation with non-deterministic computational execution}.
 \end{quote}

### 2021-10-28 document/source/body.tex
review done

@@
 Selection then reduces to a dequeue operation on the frontier.
-In general the heuristic does not guarantee a uniquely optimal node in the frontier and thus ties may be encountered.
+In general, the heuristic does not guarantee a uniquely optimal node in the frontier and thus ties may be encountered.
 The question of how these ties are broken in a priority queue is related to the broader notion of sorting stability~\cite{Sedgewick2011}.
@@
 Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
-To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
+To ensure precise simulation outcomes, the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
 %
@@
 To put this another way, we can accept a precision with a tolerance of $\leq$$\pm$1$cm$.
-This tolerance may need to be chosen for the specific verification case and the speed of the vehicles within the environment~\cite{jiang2014intercultural}.
+In practice, this tolerance may need to be chosen for each specific verification case and the speed of the vehicles within the environment~\cite{jiang2014intercultural}.
 
@@
 However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values.
-Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.
+Details on this previous investigation can be found on the Trustworthy Systems github~\cite{TSLUnrealEngineTesting}.
 
@@
     \includegraphics[width=0.48\textwidth]{../other/figures/actor_path_plot.jpg}
-    \caption{Actor path plot in XY plane for Test 4 with 95\% resource utilisation.}
+    \caption{Actor path plot in X--Y plane for Test 4 with 95\% resource utilisation.}
     \label{actorPathPlot}
@@
 
-A summary of the main results are shown in Table~\ref{TableOfExperiments}.
+A summary of the main results is shown in Table~\ref{TableOfExperiments}.
 In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.
@@
 %
-A plot of actor position in the X-Y plane (plan view, units $m$) is shown in Fig.\ref{actorPathPlot}, where the inset clearly shows the divergence of the path of vehicle 2 post-collision with the pedestrian.
+A plot of actor position in the X--Y plane (plan view, units $m$) is shown in Fig.\ref{actorPathPlot}, where the inset clearly shows the divergence of the path of vehicle 2 post-collision with the pedestrian.
 %
@@
 \subsubsection{Sample Size}
-It is recommended that the chosen sample size, i.e.
-the number of repeated tests, is determined empirically.
+It is recommended that the chosen sample size, i.e.\ the number of repeated tests, is determined empirically.
 We recommend monitoring variance while increasing the sample size in orders of magnitude until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process.
-Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
+Note that we suggest monitoring the maximum value of $\sigma$, not the average,
+%
+%because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
+%
+because if even a single simulation run is outside of the permissible variance it would be critically important to detect this.
+Using an average may lead to false confidence in the verification result.
 
@@
 To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
-Alternatively there are commercially available driving simulators that claim to be fully deterministic, e.g.
-RFPro~\cite{rFpro2021}, that may be more suitable if using a game engine does not provide a simulation variance sufficient for your verification requirements.
+Alternatively, there are commercially available driving simulators that claim to be fully deterministic, e.g.\ RFPro~\cite{rFpro2021}, that may be more suitable if using a game engine does not provide a simulation variance sufficient for the verification requirements.
 
@@
 % ***************************************************
-\section{Acknowledgement}
+\section*{Acknowledgement}
 This research has in part been funded by the ROBOPILOT and CAPRI projects.
 Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-Special thanks to David May, Roger Shepherd, CFMS, Auroch Digital and Fortelix for their discussions and support on this topic.
+Special thanks to David May, Roger Shepherd, CFMS, Auroch Digital and Fortelix for productive discussions and support.

### 2021-10-28 document/source/body.tex
final changes

--- 2021-10-17 document/source/body.tex
@@
 \subsubsection{Determinism}
-Schumann et al.\ describe determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.\TODO{KIE: I can't find that quote in that paper, please point me to it.} However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
+Schumann et al.\ describe determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
+However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors.
 Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.

### 2021-11-01 paper_arxiv/arxiv_cav_determinisim_main.tex
update arxiv version

@@
 \usepackage[font=footnotesize,labelfont=bf]{subcaption}
-\usepackage[hidelinks]{hyperref}
+% \usepackage[hidelinks]{hyperref}
 \usepackage[capitalize]{cleveref}
@@
 Tony Pipe (e-mail: tony.pipe@brl.ac.uk), 
-are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom. }.
+are with the Bristol Robotics Lab, T Block, University of the West of England,Frenchay, Coldharbour Ln, Bristol, BS34 8QZ, United Kingdom.}
 \makeatletter

### 2021-11-01 document/source/body.tex
errant full-stop

@@
 Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-Special thanks to David May, Roger Shepherd, CFMS, Auroch Digital and Fortelix for productive discussions and support.
+This research was also supported in part by the “UKRI Trustworthy Autonomous Systems Node in Functionality” under grant number EP/V026518/1.Special thanks to David May, Roger Shepherd, CFMS, Auroch Digital and Fortelix for productive discussions and support.

### 2022-01-24 document/source/body.tex
shortened FP and scheduling

@@
 
-A common misconception is to attribute non-determinism to the use of floating point number representation. This is an erroneous conflation with non-deterministic computational execution.  
+% NEW TEXT V3
+It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which limited by the fixed bit width available and potentially resulting rouding errors~\cite{FloatingPointsBook,goldberg1991every}.
+As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}.
 %
+In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions but we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
+So, even if the result of a operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
+%
+However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
+parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU (which may have different register widths~\cite{Whitehead2011}) may all affect the execution order.
+%
+In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.
+
+% PREVIOUS TEXT V2
+% A common misconception is to attribute non-determinism to the use of floating point number representation.
+This is an erroneous conflation with non-deterministic computational execution.
 Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation.
@@
 This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
-%
 
-As a consequence, floating-point arithmetic is not associative.
+% As a consequence, floating-point arithmetic is not associative.
 Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}.
 This is because rounding is being performed at different stages in the computation, depending on the respective execution order.
-%
-% Translating to the usage of AV simulations in physics engines,
 In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
-%
 Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability.
@@
 
-In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-%
+% In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
 It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic.
-%
 Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern.
-%
 So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
 
-% However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results.
-Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.
-
-% Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}.
-However, the same executable should still return the same output for identical input.
-
-% Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
-However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, and 
-parallelisation within the runtime environment or at the hardware level can change the execution order, which can cause simulation results to differ between runs when floating-point operations are affected.
-%
+% However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, and 
+% parallelisation within the runtime environment or at the hardware level can change the execution order, which can cause simulation results to differ between runs when floating-point operations are affected.
 Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.
 
-In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.
+% In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.
 
 \subsection{Scheduling, Concurrency and Parallelisation}
-Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy.
+
+Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority depending on the operating system's scheduler policy.
 A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
-%
 In principle, changing the scheduling policy and thread priorities may increase simulation variance.
 
-It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
-Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
-
-However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
-This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
-%
-Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
-%
-
-Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
+It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should give the same output.
+However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading} or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.
 
@@
 This may be decided based on factors such as throughput, latency or CPU utilisation.
-%
+Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or if executing across parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
+Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
+
+Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}.
+
+% PREVIOUS TEXT V2
+% Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy.
+A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}.
+In principle, changing the scheduling policy and thread priorities may increase simulation variance.
+
+% It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration.
+Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.
+
+% However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events.
+This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
+Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
+Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.
+
+% Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes.
+This may be decided based on factors such as throughput, latency or CPU utilisation.
 If due to the CPU utilisation policy the same single-threaded script executes multiple times across different physical cores of the same CPU type,  then execution should still produce the same output.
-%
 This is because the processor cores are identical and any impurities across the bulk silicon and minor perturbations in the semiconductor processing across the chip that may exist should have been accommodated for in design and manufacturing tolerances.
-%
 However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
 
-Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
+% Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
-%
 
-Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously.
+% Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously.
 While this would be faster than a serial execution, the order of execution arising from program-level concurrency is often not guaranteed.
 
-Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}.
+% Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}.
 
@@
 \subsection{Actor Navigation}
-% There is evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
-Unreal Engine is the underlying framework for the CARLA simulator which uses the A* search algorithm to find optimal paths for actor navigation~\cite{newton2016unreal}.
-%
 In game engines, the A* search algorithm is often used for navigation to find optimal paths for actors. 
@@
 This also assumes the simulation environment itself is deterministic as sources of non-determinism that affect the simulator or the system more generally may cause changes to the operation of an A* implementation.
-
-% The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.
-A navigation mesh is a fixed area of the environment where actors are free to navigate.
-Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in then environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.
-However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment.
-These cannot be unknowingly modified and therefore can be considered fixed.
-
-% Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.
 

### 2022-01-25 document/source/body.tex
shortened: when is determinsim needed, external settings, conclusion

@@
 \subsubsection{Determinism}
+
 Schumann et al.\ describe determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
@@
 \subsection{When is Determinism needed?}
-Determinism is a key requirement for simulation during AV development and testing.
-A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
-Thus, a deterministic simulator has zero \textit{variance}.
 
-A simulator with non-zero variance can no longer be considered deterministic.
-Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}.
+% NEW TEXT
+Determinism is a key requirement for simulation during AV development and testing, it ensures repeated runs have the same output and therefore has zero variance.
+A simulator with non-zero variance is non-deterministic but may be sufficient for some applications as long as variance is permissible, i.e.
+\textit{within tolerance}.
 Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations.
@@
 
-If the simulation is non-deterministic, e.g.\ it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
-In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
-
+Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may result in intermittent assertion failure and failing to detect bugs and leading to false confidence in the system.
 When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
@@
 {\em How can one determine and control the simulation variance?}
+
+% OLD TEXT V2
+% Determinism is a key requirement for simulation during AV development and testing.
+A deterministic simulation environment guarantees that tests are repeatable, i.e.\ repeated runs of a test, given the same initial conditions and event history, produce the same output data.
+Thus, a deterministic simulator has zero \textit{variance}.
+
+% A simulator with non-zero variance can no longer be considered deterministic.
+Non-deterministic simulators may be sufficient for certain applications as long as their variance is permissible, i.e.\ \textit{within tolerance}.
+Therefore, \textit{tolerance} is the acceptable degree of variability between repeated simulations.
+When the simulation output is within tolerance, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated.
+This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy.
+
+% If the simulation is non-deterministic, e.g.\ it has a non-zero variance in, for example, actor positions, then this may, in the best case, lead to intermittent assertion failures, making it difficult to reproduce, understand and remove bugs and rendering verification results unstable.
+In the worst case, however, bugs that could have been identified in simulation remain undetected, leading to false confidence in the safety of the AV's control software.
+
+% When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
+It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience.
+However, the situation is very different for AV development and testing.
+Thus, our main research questions are:
+% {\em How can one assess whether a simulation environment is deterministic?} and 
+% {\em How can one determine and control the simulation variance?}
 
@@
 % NEW TEXT V3
-It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which limited by the fixed bit width available and potentially resulting rouding errors~\cite{FloatingPointsBook,goldberg1991every}.
+It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which limited by the fixed bit width available and potentially resulting rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
 As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}.
@@
 \subsection{Actor Navigation}
+
 In game engines, the A* search algorithm is often used for navigation to find optimal paths for actors. 
-%
 The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.
@@
 
-After examining controls the internal game engine, external settings were explored.
+Performing physics calculations and running the simulator may consume a significant proportion of system resources (e.g.
+CPU and GPU processors) for higher performing simulations.
+An initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance} which was corroborated with some initial investigations~\cite{TSLUnrealEngineTesting} and explored more fully in this work using the Carla platform.
+
+% PREVIOUS TEXT V2
+% After examining controls the internal game engine, external settings were explored.
 Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment.
@@
 
-Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load.
+% Some initial exploratory work was undertaken~\cite{TSLUnrealEngineTesting} that suggested computational resource utilisation was positively correlated with simulation variance, leading to high simulation variance when a system is under high load.
 This initial observation was explored more fully in this work using the CARLA platform.
@@
 In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency-induced imprecision~\cite{matlab_parallel_computing}.
+
+% ***************************************************
+%  Results and Discussion
+% ***************************************************
 
@@
 % ***************************************************
+
 \subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
@@
 \end{itemize}
-
-% ***************************************************
-%  Results and Discussion
-% ***************************************************
 
@@
 
+% ***************************************************
+%  Process Scheduling Priority
+% ***************************************************
+
 \subsection{Process Scheduling Priority} \label{r:process_scheduling}
@@
 
+% ***************************************************
+%  Non-uniform memory access (NUMA)
+% ***************************************************
+
 \subsection{Non-uniform memory access (NUMA)} \label{r:numa}
@@
 However, practitioners that are aiming to minimise simulation variance should use a technique such as this to minimise data transfer cost and potentially minimise any potential effect on process execution order that may lead to increased simulation variance.
+
+% ***************************************************
+%  Investigation Summary
+% ***************************************************
 
@@
 % ***************************************************
+
 \section{Method to determine the Variance of a Simulation} \label{s:methodology}
@@
 However, gaining fine control over the scheduling policy and thread execution order is non-trivial~\cite{acm-q-rr-interview}.
-
 The operating system schedules threads according to a specified scheduling policy, potentially based on equal thread priority.
 Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation.
-
 Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
@@
 We recommend monitoring variance while increasing the sample size in orders of magnitude until there is confidence that $\max\sigma$ will not exceed the permissible variance for the verification process.
-Note that we suggest monitoring the maximum value of $\sigma$, not the average,
-%
-%because if even a single simulation run is outside of the permissible variance this may lead to false confidence in the verification result.
-%
-because if even a single simulation run is outside of the permissible variance it would be critically important to detect this.
+Note that we suggest monitoring the maximum value of $\sigma$, not the average, because if even a single simulation run is outside of the permissible variance it would be critically important to detect this.
 Using an average may lead to false confidence in the verification result.
@@
 We then researched, identified and discussed potential sources for non-determinism in this context.
-A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution.
-In particular, actor collisions and system-level resource utilisation were identified as key contributors.
-To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
-Alternatively, there are commercially available driving simulators that claim to be fully deterministic, e.g.\ RFPro~\cite{rFpro2021}, that may be more suitable if using a game engine does not provide a simulation variance sufficient for the verification requirements.
+%
+% A systematic case study of the CARLA simulator uncovered the actual factors that contribute towards greater simulation variance, giving rise to non-deterministic execution. 
+%
+In particular, actor collisions and system-level resource utilisation were identified as key contributors to increased simulation variance when using Carla and we recommend monitoring these during simulation execution or using a commercial alternative, e.g.~\cite{rFpro2021}.
+%
+% To meet the recommended simulation variance for urban autonomous vehicle verification demonstrated in our case study, we recommend being aware of actor collisions and resource utilisation levels during simulation execution.
+%
+% Alternatively, there are commercially available driving simulators that claim to be fully deterministic, e.g.\ RFPro~\cite{rFpro2021}, that may be more suitable if using a game engine does not provide a simulation variance sufficient for the verification requirements.
 

### 2022-01-25 document/source/body.tex
Shortened Section 4F
Shortened Section 4F on Actor Navigation by 50%.

@@
 
-In game engines, the A* search algorithm is often used for navigation to find optimal paths for actors. 
-The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.
+The A* algorithm is commonly used for actor navigation in game engines.
+A* explores a search graph through a series of node-selection-and-expansion steps that continue until a goal node is found.
+A heuristic is used at each step to select the most promising node for expansion from a set of candidate nodes called the \textit{frontier}.
+The heuristic is not required to guarantee a uniquely preferred node in the frontier, so ties may be found during selection and then broken arbitrarily, meaning that an implementation of A* can be non-deterministic.
 
-To briefly describe the A* process, at each stage of the search, the algorithm selects one node for expansion from a set of nodes known as the \textit{frontier} (also fringe or open set) using a heuristic to select this node.
-For reasons of efficiency, A* represents the frontier as a priority queue, with nodes prioritised by the heuristic.
+For reasons of efficiency, A* represents the frontier as a priority queue with nodes prioritised by the heuristic.
 Selection then reduces to a dequeue operation on the frontier.
-In general, the heuristic does not guarantee a uniquely optimal node in the frontier and thus ties may be encountered.
-The question of how these ties are broken in a priority queue is related to the broader notion of sorting stability~\cite{Sedgewick2011}.
-An implementation of a priority queue is said to be stable if it preserves insertion order, otherwise it is said to be unstable.
-If an implementation of A* uses a stable priority queue, then it will behave as a deterministic algorithm which always returns the optimal node that was first added to the frontier.
-Stable priority queues are typically less computationally efficient than unstable priority queues, so the requirement on stability is often dropped in practice.
+How to break ties in priority queues is related to a broader question of sorting stability~\cite{Sedgewick2011} where a priority queue is stable if it breaks ties based on insertion order, and is unstable otherwise.
+Typically unstable priority queues are more computationally efficient than stable priority queues and the most common method (using a binary heap)~\cite{boost2021,priorityQueue} is unstable, breaking ties based on internal heap order rather than insertion order.
+While this priority queue is unstable, it is still deterministic, since it always breaks ties in the same way (based on heap order).
+An example of an unstable priority queue that is non-deterministic is one that breaks ties at random.
+An implementation of A* is deterministic if it uses either a stable priority queue or an unstable but deterministic priority queue.
 
-For example, one of the most common implementations of a priority queue relies on a binary heap and is unstable; it will always return the first optimal node according to its internal heap order rather than insertion order.
-The default priority queue implementations in the standard Java and Boost libraries are based on binary heaps~\cite{boost2021,priorityQueue}, while the Unreal documentation and source code suggests that this is also true of CARLA~\cite{FGraphAStar,TArray,GraphAStar}.
-Thus, an unstable but deterministic priority queue is an unstable priority queue that preserves some order, but not necessarily insertion order.
-If an implementation of A* uses an unstable but deterministic priority queue, then it will behave as a deterministic algorithm which always returns the optimal node according the ordering of its priority queue.
-
-Finally, an unstable but deterministic priority queue suggests a third kind of priority queue that we can refer to as an unstable and non-deterministic priority queue; that is, a priority queue that always returns an optimal node selected at random (i.e.
-breaking ties arbitrarily).
-If an implementation of A* uses an unstable and non-deterministic priority queue, then it will behave as a non-deterministic algorithm.
-Overall, an implementation of A* is much more likely to use a priority queue that is unstable but deterministic rather than one that is stable, and is very unlikely to use one that is unstable and non-deterministic.
-
-An implementation of A* will behave as a deterministic algorithm if it uses a stable priority queue or an unstable but deterministic priority queue.
-However, apparent determinism of A* will be sensitive to the ordering preserved by the priority queue (i.e.
-insertion order or e.g.
-heap order).
-If we are not aware of changes to this ordering, then a deterministic implementation of A* may appear to behave as a non-deterministic algorithm, since we will not have accounted for changes to the ordering.
-For example, suppose actions are iterated over in order of hash value (where order of iteration over actions often determines insertion order), but between runs there is some untracked change to the software/hardware stack that changes the calculation of the hash value, then this may change insertion order of nodes to the frontier, which could eventually materialise as a different optimal solution found by A*.
+The Unreal documentation and source code suggests that CARLA uses a priority queue based on a binary heap~\cite{FGraphAStar,TArray,GraphAStar}.
+Theoretically this means that use of A* in CARLA will be determinstic as long as the heap is always constructed in the same way.
+However, if some unknown change causes the heap to be constructed in a different way, then the use of A* may then appear non-deterministic.
+This can be a complex problem in practice.
+For example, suppose A* iterates over actions in order of hash value, but between runs there is some untracked change to the hardware/software stack that alters the hash function, then this may change insertion order of nodes to the frontier, which could eventually lead to a different binary heap and thus a different optimal solution found by A*.
 This suggests that while a simulated environment may behave deterministically, factors outside the simulator may cause changes to the operation of an A* implementation, which would then materialise as non-deterministic runs in the simulator.
-The difference between the use of a stable priority queue or an unstable but deterministic priority is simply that it may be easier to detect such changes under a stable priority queue because insertion order is typically more meaningful with respect to the actual implementation of A*.
-
-To summarise, documentation suggests that CARLA uses a binary heap for the priority queue~\cite{FGraphAStar,TArray,GraphAStar} which preserves some order of the frontier nodes, but not necessarily the insertion order, and hence is unstable.
-Therefore, determinism of the A* algorithm for actor navigation will be sensitive to the ordering preserved by the priority queue, in this case the heap order.
-This also assumes the simulation environment itself is deterministic as sources of non-determinism that affect the simulator or the system more generally may cause changes to the operation of an A* implementation.
+The difference between use of a stable priority queue and an unstable but deterministic priority is simply that it may be easier to detect such changes under the former because insertion order is typically more meaningful with respect to the actual implementation of A*.
 

### 2022-01-26 document/ieee_trans/det_ieeetrans.tex
final changes
trim to 15 pages, web links to footnotes, trimmed bios

@@
 % \input{archive/main.config.tex}
-\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
-% \addbibresource{IROS.bib}
+\usepackage[maxnames=2,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
 \addbibresource{../source/cav_determinism.bib}
@@
 \usepackage{enumitem}
-\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
+% \usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
 % \usepackage{subfig}
@@
 % \pagenumbering{gobble} %suppresses page numbers
-\usepackage{balance} %balance columns
+% \usepackage{balance} %balance columns
 %\usepackage{todonotes} 
@@
 Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
+%
+This research was also supported in part by the “UKRI Trustworthy Autonomous Systems Node in Functionality” under grant number EP/V026518/1.
 % The Associate Editor for this paper was ....
@@
 % ***************************************************
+
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/greg.png}}]{Greg Chance}
@@
 He has co-authored over 200 international refereed publications, supervised 35 PhDs, and created a BRL income of \pounds10M since 2013.
-He has acted as BRL-UWE PI on the EPSRC grant ‘Trustworthy Robotic Assistants’, the Programme/Hub grants ‘Robotics for Nuclear Environments’ and ‘National Centre for Nuclear Robotics’, and the Centre for Doctoral Training in Robotics and Autonomous Systems.
+%He has acted as BRL-UWE PI on the EPSRC grant ‘Trustworthy Robotic Assistants’, the Programme/Hub grants ‘Robotics for Nuclear Environments’ and ‘National Centre for Nuclear Robotics’, and the Centre for Doctoral Training in Robotics and Autonomous Systems.
 Prof.\ Pipe has been involved with Connected Autonomous Vehicle research and development since 2015 via the Innovate UK/CCAV funded projects VENTURER, FLOURISH, MultiCAV, CAPRI, ROBOPILOT and CAV-Forth.
@@
 She also leads the Verification and Validation for Safety in Robots research theme at the  Bristol Robotics Laboratory.
-She has gained extensive experience of verifying complex microelectronic designs while working with leading semiconductor design and Electronic Design Automation companies worldwide.
+%She has gained extensive experience of verifying complex microelectronic designs while working with leading semiconductor design and Electronic Design Automation companies worldwide.
 Her research is focused on specification, verification and analysis techniques to verify or explore a system's behaviour in terms of functional correctness, safety, security, performance and energy efficiency.
@@
 In 2007 she was awarded an ``Excellence in Engineering'' prize from the Royal Academy of Engineering, UK.
-
 \end{IEEEbiography}

### 2022-01-26 document/source/body.tex
trim to 15 pages, web links to footnotes, trimmed bios

@@
 %
-CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.
+CARLA~\cite{CARLA_paper}% ~\cite{carla_main_website}
+, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine\footnote{\url{https://www.unrealengine.com/}}
+% ~\cite{UE4_main_website}
+, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors.
 
@@
 
-There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}.
-Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
+There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine
+, Unity, CryEngine.\footnote{\url{https://www.cryengine.com/}} 
+% ~\cite{CryEngine_main_website}
+Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{CARLA_paper}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
+%~\cite{UE4_main_website}
+% ~\cite{Unity_main_website}
+%~\cite{carla_main_website}
+% ~\cite{AirSim_main_website}
+% ~\cite{Apollo_main_website}
+% ~\cite{nvidia_constellation}
 
@@
 The game loop is responsible for the interaction between the physics and rendering engines.
-Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed~\cite{unity_event_execution}.
+Fig.~\ref{GameEngineLoopDiagram} depicts a simplified representation of the process flow in a game engine loop, where  initialisation, game logic and decommissioning have been removed\footnote{\url{https://docs.unity3d.com/Manual/ExecutionOrder.html}}.
+% ~\cite{unity_event_execution}.
 A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering)~\cite{GameEngineArchBook}.
@@
 When the frame is rendered the game loop cycle returns to processing inputs.
-An intuitive and  more detailed description of the interplay between the physics and render cycles is given in~\cite{JohnAustinUnity}.
+An intuitive and  more detailed description of the interplay between the physics and render cycles, see\footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given in~\cite{JohnAustinUnity}.
 
@@
 Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$.
-However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}.
+However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations\footnote{\url{https://gafferongames.com/post/fix_your_timestep/}}.%~\cite{gaffer}.
 %
@@
 It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should give the same output.
-However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading} or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.
+However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}
+% ~\cite{unity_multithreading} 
+or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.
 
@@
 
-The Unreal documentation and source code suggests that CARLA uses a priority queue based on a binary heap~\cite{FGraphAStar,TArray,GraphAStar}.
+The Unreal documentation and source code suggests that CARLA uses a priority queue based on a binary heap 
+% e.g.
+see\footnote{\url{https://docs.unrealengine.com/en-US/API/Runtime/AIModule/FGraphAStar/index.html}, \url{https://github.com/EpicGames/UnrealEngine/blob/release/Engine/Source/Runtime/AIModule/Public/GraphAStar.h} and \url{https://docs.unrealengine.com/en-US/ProgrammingAndScripting/ProgrammingWithCPP/UnrealArchitecture/TArrays/index.html}}.
+~\cite{FGraphAStar,TArray,GraphAStar}. 
 Theoretically this means that use of A* in CARLA will be determinstic as long as the heap is always constructed in the same way.
@@
 Thus, in such a case, all tasks non-essential to the simulator should be terminated to prevent interference with the simulation.
-Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.
+Assigning a higher priority to the simulator process may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}}
+% ~\cite{TaskSettingWindows} 
+or \texttt{NICE} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.
+%~\cite{Nice_linux}.
 

### 2022-01-28 document/source/body.tex
has have

@@
 % NEW TEXT
-Determinism is a key requirement for simulation during AV development and testing, it ensures repeated runs have the same output and therefore has zero variance.
+Determinism is a key requirement for simulation during AV development and testing, it ensures repeated runs have the same output and therefore have zero variance.
 A simulator with non-zero variance is non-deterministic but may be sufficient for some applications as long as variance is permissible, i.e.

### 2022-01-31 document/source/body.tex
proof read GC

@@
 
-Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may result in intermittent assertion failure and failing to detect bugs and leading to false confidence in the system.
+Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may result in intermittent assertion failure, failing to detect bugs and potentially leading to false confidence in the system under test.
 When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
@@
 There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine
-, Unity, CryEngine.\footnote{\url{https://www.cryengine.com/}} 
+, Unity\footnote{\url{https://unity.com/}} and CryEngine.\footnote{\url{https://www.cryengine.com/}} 
 % ~\cite{CryEngine_main_website}
@@
 Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$.
-However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations\footnote{\url{https://gafferongames.com/post/fix_your_timestep/}}.%~\cite{gaffer}.
+However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations\footnote{\url{https://gafferongames.com/post/fix_your_timestep/}}.
+%~\cite{gaffer}.
 %
@@
 % NEW TEXT V3
-It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which limited by the fixed bit width available and potentially resulting rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
+It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which is limited by the fixed bit width available and can potentially result rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
 As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}.
@@
 In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions but we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-So, even if the result of a operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
+So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
 %
 However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
-parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU (which may have different register widths~\cite{Whitehead2011}) may all affect the execution order.
+parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order.
 %
@@
 
-It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should give the same output.
+It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output.
 However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}
@@
 This may be decided based on factors such as throughput, latency or CPU utilisation.
-Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or if executing across parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
+Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or executing across a parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
 Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.

### 2022-01-31 document/source/body.tex
GC KE meeting

@@
 
-Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may result in intermittent assertion failure, failing to detect bugs and potentially leading to false confidence in the system under test.
+Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may render the outcome unstable producing incorrect results potentially leading false confidence in the system under test.
+%in intermittent assertion failure, failing to detect bugs and potentially leading to false confidence in the system under test. 
 When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers.
@@
 When the frame is rendered the game loop cycle returns to processing inputs.
-An intuitive and  more detailed description of the interplay between the physics and render cycles, see\footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given in~\cite{JohnAustinUnity}.
+For an intuitive and  more detailed description of the interplay between the physics and render cycles, see~\cite{JohnAustinUnity}.
+% \footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given
 
@@
 % NEW TEXT V3
-It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which is limited by the fixed bit width available and can potentially result rounding errors~\cite{FloatingPointsBook,goldberg1991every}.
-As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}.
+It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which necessitates rounding due to a fixed bit width~\cite{FloatingPointsBook,goldberg1991every}.
+% available and can potentially result in rounding errors~\cite{FloatingPointsBook,goldberg1991every}. 
+As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and results may differ depending on execution order.
+% Some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
 %
-In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions but we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
+In the context of AV simulation, this
+% finite-precision of floating point number representation 
+% such rounding errors 
+could result in accuracy issues of, for example, actor positions. While some authors suggest
+avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
 So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
@@
 However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
-parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order.
+parallelisation within the runtime environment or at the hardware level and performing the execution on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order.
 %
@@
 
-It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output.
+It is responsibility of the tester
+% It is unlikely 
+It's unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output.
 However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}

### 2022-02-02 document/source/body.tex
final det paper edits

@@
 \subsection{Definitions}
-Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
+Refer to Fig.~\ref{variance_description} for the definitions introduced in this section.
+% Several definitions are introduced in this section. These are used in the subsequent discussion. Refer to Fig.~\ref{variance_description} throughout this section.
 \\
@@
 all future states are perfectly reproducible from prior states and actions.
-However, if a simulation is non-deterministic, Fig.~\ref{variance_description_a}, then there will be a variation in the generated output data.
+However, if a simulation is non-deterministic, Fig.~\ref{variance_description_a}, then there will be a variation in the output data.
 \\
@@
 Catching intermittent failures, or flaky tests~\cite{intermittently-failing-tests}, in a test suite that cannot be replayed makes the debugging process equally difficult~\cite{acm-q-rr-interview}.
-This section gives an overview of the internal structure of a game engine and what sources or settings in the engine may affect \textit{simulation variance}.
+This section gives an overview of the internal structure of a game engine and what sources or settings may affect \textit{simulation variance}.
 
@@
 When the frame is rendered the game loop cycle returns to processing inputs.
-For an intuitive and  more detailed description of the interplay between the physics and render cycles, see~\cite{JohnAustinUnity}.
+For an intuitive and  more detailed description of the interplay between the physics and render cycles see~\cite{JohnAustinUnity}.
 % \footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given
@@
 % NEW TEXT V3
-It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which necessitates rounding due to a fixed bit width~\cite{FloatingPointsBook,goldberg1991every}.
+It is a common misconception to attribute non-deterministic computational execution with the use of floating-point number representation, which necessitates rounding due to a fixed bit width~\cite{FloatingPointsBook,goldberg1991every}.
 % available and can potentially result in rounding errors~\cite{FloatingPointsBook,goldberg1991every}. 
@@
 could result in accuracy issues of, for example, actor positions. While some authors suggest
-avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
-So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that conform to the IEEE floating-point standard~\cite{8766229}.
+avoiding floating-point representation entirely~\cite{empirical-analysis-of-flaky-tests}, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se.
+So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that meet the IEEE floating-point standard~\cite{8766229}.
 %
@@
 In principle, changing the scheduling policy and thread priorities may increase simulation variance.
-
-It is responsibility of the tester
-% It is unlikely 
-It's unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output.
+%
+It would therefore be important to ensure these remain stable between repeated runs. 
+% Changes to thread priorities or scheduling policy during repeated runs on the same hardware and should therefore give the same output. 
 However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} 
@@
 This may be decided based on factors such as throughput, latency or CPU utilisation.
-Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or executing across a parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
+Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.
 Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory.
 This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
+%
+Likewise, simulation variance may be observed for game engines that use GPU parallelisation  by offloading time-critical calculations to several dedicated cores simultaneously.
+While this would be faster than a serial execution, the order of execution arising from program-level concurrency cannot, in general, be guaranteed.
 
@@
 
-The Unreal documentation and source code suggests that CARLA uses a priority queue based on a binary heap 
+The Unreal documentation and source code suggest that CARLA uses a priority queue based on a binary heap 
 % e.g.
@@
 ~\cite{FGraphAStar,TArray,GraphAStar}. 
-Theoretically this means that use of A* in CARLA will be determinstic as long as the heap is always constructed in the same way.
-However, if some unknown change causes the heap to be constructed in a different way, then the use of A* may then appear non-deterministic.
+Theoretically this means that use of A* in CARLA will be deterministic as long as the heap is always constructed in the same way.
+However, if some unknown change causes the heap to be constructed in a different way, then the use of A* may appear non-deterministic.
 This can be a complex problem in practice.
@@
 
-Performing physics calculations and running the simulator may consume a significant proportion of system resources (e.g.
-CPU and GPU processors) for higher performing simulations.
-An initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance} which was corroborated with some initial investigations~\cite{TSLUnrealEngineTesting} and explored more fully in this work using the Carla platform.
+Executing physics calculations during simulation may consume a significant proportion of system resources (e.g.\ CPU and GPU processors).
+This suggests that as resource utilisation of the simulation increases, so does the \textit{simulation variance} which was corroborated with some initial investigations~\cite{TSLUnrealEngineTesting} and explored more fully in this work using CARLA.
 
@@
 
-CPU utilisation software, such as the linux \texttt{stress} tool, which is a workload generator program, can be used to spawn workers on any number of cores or virtual threads on a system.
+CPU utilisation software, such as the linux workload generator \texttt{stress}, %which is a workload generator program, 
+can be used to spawn workers on any number of cores or virtual threads on a system.
 This can be used to artificially increase the load on the system.
@@
 \subsubsection{Scheduling}
-We hypothesise that thread scheduling may be a major contributor to the non-deterministic results found in the empirical study.
+We hypothesise that thread scheduling may be a major contributor to the non-deterministic results of the empirical study.
 However, gaining fine control over the scheduling policy and thread execution order is non-trivial~\cite{acm-q-rr-interview}.
@@
 %
-In particular, actor collisions and system-level resource utilisation were identified as key contributors to increased simulation variance when using Carla and we recommend monitoring these during simulation execution or using a commercial alternative, e.g.~\cite{rFpro2021}.
+In particular, actor collisions and system-level resource utilisation were identified as key contributors to increased simulation variance when using CARLA and we recommend monitoring these during simulation.
+Alternatively some commercial driving simulators claim to be fully deterministic, for instance RFpro~\cite{rFpro2021}.
 %
@@
 %
-% Alternatively, there are commercially available driving simulators that claim to be fully deterministic, e.g.\ RFPro~\cite{rFpro2021}, that may be more suitable if using a game engine does not provide a simulation variance sufficient for the verification requirements.
+% Alternatively, there are commercially available driving simulators that claim to be fully deterministic, e.g.\ RFPro~\cite{rFpro2021}, 
+These may be more suitable if using a game engine does not provide a simulation variance sufficient for the verification requirements.
 

### 2022-02-02 document/ieee_trans/det_ieeetrans.tex
update bios

@@
 received the B.Sc.
-(Hons.) and Ph.D.
-degrees in physics from the University of Bath, Bath, U.K., in 2001 and 2005, respectively.
-He was previously at the Bristol Robotics Lab and is currently a Senior Research Associate with the Trustworthy Systems Laboratory, University of Bristol, Bristol, U.K.
+(Hons.) and Ph.D.\ degrees in physics from the University of Bath, in 2001 and 2005, respectively.
+He was previously at the Bristol Robotics Lab and is currently a Senior Research Associate with the Trustworthy Systems Laboratory, University of Bristol.
 His current research interests are simulation-based verification for autonomous systems and cybersecurity.
-He has ten years industrial experience researching for Oxford Instruments, U.K., and BAE Systems, U.K.
+He has ten years industrial experience researching for Oxford Instruments, and BAE Systems.
 Dr Chance is a Chartered Engineer and a member of the Institute of Physics and holds a patent for a novel plasma control method.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/abanoub}}]{Abanoub Ghobrial}
-received the MEng degree in mechanical engineering from the University of Bristol, Bristol,  U.K., in 2018.
-He is currently pursuing the Ph.D.
-degree in computer science at the University of Bristol and is a part-time Research Associate with the Trustworthy Systems Lab, Bristol,  U.K.
-From 2018 to 2020, he was a full-time Research Associate with the Trustworthy Systems Lab.
-His current research interests are techniques to allow self-managing of autonomous safety-critical systems via continual learning during operation; and the development of simulation-based verification techniques for autonomous systems.
+received an MEng degree in mechanical engineering from the University of Bristol in 2018.
+He is currently pursuing a Ph.D.\ degree in computer science at the University of Bristol and is a part-time Research Associate with the Trustworthy Systems Lab, where he was a full-time Research Associate from 2018 to 2020.
+His current research interests are techniques to allow self-managing of autonomous safety-critical systems via continual learning during operation and the development of simulation-based verification techniques for autonomous systems.
 \end{IEEEbiography}

### 2022-02-02 document/ieee_trans/det_ieeetrans.tex
KIE bio update

@@
 Her research is focused on specification, verification and analysis techniques to verify or explore a system's behaviour in terms of functional correctness, safety, security, performance and energy efficiency.
-Prof.\ Eder's most recent contributions include agent-based testing, how to use assertions and theorem proving to verify control system designs, energy modeling and static resource analysis techniques to predict energy consumption of software.
-She holds a PhD in Computational Logic and an MSc in Artificial Intelligence from the University of Bristol, UK, as well as an MEng in Informatics (Dipl.\,Inf.) from the Technical University Dresden, Germany.
+Prof.\ Eder's most recent contributions include agent-based testing, how to use assertions and theorem proving to verify control system designs as well as novel energy modeling and static resource analysis techniques to predict energy consumption of software.
+She holds a PhD in Computational Logic and an MSc in Artificial Intelligence from the University of Bristol, UK, in addition to an MEng in Informatics (Dipl.\,Inf.) from the Technical University Dresden, Germany.
 In 2007 she was awarded an ``Excellence in Engineering'' prize from the Royal Academy of Engineering, UK.

### 2022-04-22 document/ieee_trans/det_ieeetrans_v1.tex
matlab code for scatter plot

--- 2022-01-26 document/ieee_trans/det_ieeetrans.tex
@@
 % \input{archive/main.config.tex}
-\usepackage[maxnames=2,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
+\usepackage[maxnames=6,firstinits=true,doi=false,url=true,isbn=false]{biblatex}
+% \addbibresource{IROS.bib}
 \addbibresource{../source/cav_determinism.bib}
@@
 \usepackage{enumitem}
-% \usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
+\usepackage{hyperref} % Add  [ocgcolorlinks,pdfusetitle] before hyperref for colored links
 % \usepackage{subfig}
@@
 % \pagenumbering{gobble} %suppresses page numbers
-% \usepackage{balance} %balance columns
+\usepackage{balance} %balance columns
 %\usepackage{todonotes} 
@@
 \title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
-% \title{On the Importance of Determinism for Simulation-based Autonomous Vehicle Verification using a Game Engine}
 % \title{On the Precision of Game Engines used for Simulation-based Autonomous Vehicle Verification}
@@
 Both projects are part-funded by the Centre for Connected and Autonomous Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers 103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
-%
-This research was also supported in part by the “UKRI Trustworthy Autonomous Systems Node in Functionality” under grant number EP/V026518/1.
 % The Associate Editor for this paper was ....
@@
 \begin{IEEEkeywords}
-Autonomous Driving, Autonomous Vehicles, Determinism, Game Engines, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
+Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
 \end{IEEEkeywords}
@@
 % ***************************************************
-
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/greg.png}}]{Greg Chance}
@@
 His research interest primarily concerns the socio-cognitive aspects of human-robot interaction and he has been focusing his recent experimental work on child-robot interaction and human-in-the-loop machine learning for social robots.
-Pr.\ S\'everin Lemaignan has been involved in several European projects related to social and cognitive robotics (CHRIS, DREAM, L2TOR) and was  awarded in 2015 a EU H2020 Marie Sk\l{}odowska-Curie Individual Fellowship for his project on robots and theory of mind.
+Pr.
+S\'everin Lemaignan has been involved in several European projects related to social and cognitive robotics (CHRIS, DREAM, L2TOR) and was  awarded in 2015 a EU H2020 Marie Sk\l{}odowska-Curie Individual Fellowship for his project on robots and theory of mind.
 \end{IEEEbiography}
@@
 He has co-authored over 200 international refereed publications, supervised 35 PhDs, and created a BRL income of \pounds10M since 2013.
-%He has acted as BRL-UWE PI on the EPSRC grant ‘Trustworthy Robotic Assistants’, the Programme/Hub grants ‘Robotics for Nuclear Environments’ and ‘National Centre for Nuclear Robotics’, and the Centre for Doctoral Training in Robotics and Autonomous Systems.
+He has acted as BRL-UWE PI on the EPSRC grant ‘Trustworthy Robotic Assistants’, the Programme/Hub grants ‘Robotics for Nuclear Environments’ and ‘National Centre for Nuclear Robotics’, and the Centre for Doctoral Training in Robotics and Autonomous Systems.
 Prof.\ Pipe has been involved with Connected Autonomous Vehicle research and development since 2015 via the Innovate UK/CCAV funded projects VENTURER, FLOURISH, MultiCAV, CAPRI, ROBOPILOT and CAV-Forth.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{../other/mugs/kie}}]{{Kerstin Eder}} 
-is Professor of Computer Science and heads the Trustworthy Systems Laboratory at the University of Bristol, UK.
+is Professor of Computer Science and heads the Trustworthy Systems Laboratory at the University of Bristol.
 She also leads the Verification and Validation for Safety in Robots research theme at the  Bristol Robotics Laboratory.
-%She has gained extensive experience of verifying complex microelectronic designs while working with leading semiconductor design and Electronic Design Automation companies worldwide.
+She has gained extensive experience of verifying complex microelectronic designs while working with leading semiconductor design and Electronic Design Automation companies worldwide.
 Her research is focused on specification, verification and analysis techniques to verify or explore a system's behaviour in terms of functional correctness, safety, security, performance and energy efficiency.
-Prof.\ Eder's most recent contributions include agent-based testing, how to use assertions and theorem proving to verify control system designs, energy modeling and static resource analysis techniques to predict energy consumption of software.
+Prof.\ Eder's most recent contributions include agent-based testing, how to use assertions and theorem proving to verify control system designs, energy modeling of software and static  analysis to predict energy consumption of programs.
 She holds a PhD in Computational Logic and an MSc in Artificial Intelligence from the University  of Bristol, UK, as well as an MEng in Informatics (Dipl.\,Inf.) from the Technical University Dresden, Germany.
-In 2007 she was awarded an ``Excellence in Engineering'' prize from the Royal Academy of Engineering, UK.
+In 2007 she was awarded an ``Excellence in Engineering'' prize from the Royal Academy of Engineering.
+
 \end{IEEEbiography}

### 2022-04-22 Results_matlab_code/CarlaReadLog_v1_GC.m

@@ -39,11 +39,11 @@ PCC=false; CC=false;
 % file = "TEST_CarsPeople_CG75.txt";
 % file = "TEST_CarsPeople_CG95.txt";
 % 
-file = "TEST_CarsPeopleCollision_CG0.txt";PCC=true;
+% file = "TEST_CarsPeopleCollision_CG0.txt";PCC=true;
 % file = "TEST_CarsPeopleCollision_CG25.txt";PCC=true;
 % file = "TEST_CarsPeopleCollision_CG50.txt";PCC=true;
 % file = "TEST_CarsPeopleCollision_CG75.txt"; PCC=true;
-% file = "TEST_CarsPeopleCollision_CG95.txt"; PCC=true;
+file = "TEST_CarsPeopleCollision_CG95.txt"; PCC=true;
 % 
 % file = "TEST_People_CG0.txt";
 % file = "TEST_People_CG25.txt";
@@ -85,9 +85,12 @@ rawData = zeros(nRepeats,nAgents,3,round(maxTime/timeStep));
 variance = zeros(nRepeats,nAgents,2);
 avgVar = zeros(nRepeats,1);
 
+
 %% For each agent get (T,X,Y) and interpolate, then variance
+data2D=[];
 for i=1:nRepeats
     for j=1:nAgents
+        A = data2D;
         sel1 = data.repeatNo==i; %select data for each repeat
         sel2 = data.agentNo==j; %select data for each agent
         sel = sel1 & sel2;
@@ -115,15 +118,53 @@ for i=1:nRepeats
         
         rawData = rawData(:,:,:,1:shortest_tempT); % chopping data to the shortest
         rawData(i,j,:,:) = [tempT, tempX, tempY]';
+        actor = j.* ones(length(squeeze(tempY)),1);
+        temp2D = [squeeze(tempX), squeeze(tempY), actor];
+        data2D = vertcat(A,temp2D);
 %         a=0
     end
 end
 
 %% Interpolate data if you need to?
 % % interpolate data
-% regT = min(tempT):0.1:max(tempT);
-% regX=interp1(tempT,tempX,regT,'linear','extrap');
-% regY=interp1(tempT,tempY,regT,'linear','extrap');
+regT = min(tempT):0.1:max(tempT);
[…]
